% Source: Lee, "Introduction to Riemannian Manifolds" (2nd ed.), Chapter 2 "Riemannian Metrics", PDF pages 23-68 of /home/axel/MathBooks/Introduction to Riemannian Manifolds (Second Edition) 2(1).pdf.
% Transcribed page-by-page by vision subagents (haiku), then merged and restructured into
% leanblueprint conventions (semantic \label, bracketed titles, \uses dependency edges, \dcref to Lee's own numbering) below.
% (Merged in 3 chunks due to chapter length.)

\section{Definitions}

Everything we know about the Euclidean geometry of $\mathbb{R}^n$ can be derived from its \textbf{dot product}, which is defined for $v = (v^1, \ldots, v^n)$ and $w = (w^1, \ldots, w^n)$ by
\[
v \cdot w = \sum_{i=1}^{n} v^i w^i.
\]

The dot product has a natural generalization to arbitrary vector spaces. Given a vector space $V$ (which we always assume to be real), an \textbf{inner product on $V$} is a map $V \times V \to \mathbb{R}$, typically written $(v, w) \mapsto \langle v, w \rangle$, that satisfies the following properties for all $u, v, x \in V$ and $a, b \in \mathbb{R}$:

\begin{enumerate}
\item[\textmd{(i)}] \textbf{Symmetry:} $\langle v, w \rangle = \langle w, v \rangle$.
\item[\textmd{(ii)}] \textbf{Bilinearity:} $\langle av + bw, x \rangle = a\langle v, x \rangle + b\langle w, x \rangle = \langle x, av + bw \rangle$.
\item[\textmd{(iii)}] \textbf{Positive Definiteness:} $\langle v,v \rangle \geq 0$, with equality if and only if $v = 0$.
\end{enumerate}

A vector space endowed with a specific inner product is called an \textit{inner product space}.

An inner product on $V$ allows us to make sense of geometric quantities such as lengths of vectors and angles between vectors. First, we define the \textit{length} or \textit{norm} of a vector $v \in V$ as
$$|v| = \langle v,v \rangle^{1/2}.$$

The following identity shows that an inner product is completely determined by knowledge of the lengths of all vectors.

\begin{lemma}[Polarization Identity]
\label{lem:polarization-identity}
\lean{LeeLib.Ch02.real_inner_eq_inner_add_add_self_sub_inner_sub_sub_self_div_four}
\leanok
\dcref{ch2:2.1}
Suppose $\langle \cdot,\cdot \rangle$ is an inner product on a vector space $V$. Then for all $v, w \in V$,
$$\langle v,w \rangle = \frac{1}{4}\bigl[\langle v+w, v+w \rangle - \langle v-w, v-w \rangle\bigr].$$
\end{lemma}

\begin{proof}
\leanok
By bilinearity and symmetry,
\[
\langle v+w, v+w \rangle = \langle v,v \rangle + 2\langle v,w \rangle + \langle w,w \rangle,
\qquad
\langle v-w, v-w \rangle = \langle v,v \rangle - 2\langle v,w \rangle + \langle w,w \rangle .
\]
Subtracting the second identity from the first cancels $\langle v,v\rangle$ and $\langle w,w\rangle$ and leaves $4\langle v,w\rangle$. Dividing by $4$ gives the stated formula.
\end{proof}

\begin{exercise}
\label{exer:polarization-proof}
\lean{LeeLib.Ch02.real_inner_eq_inner_add_add_self_sub_inner_sub_sub_self_div_four}
\leanok
\uses{lem:polarization-identity}
\dcref{ch2:2.2}
Prove the preceding lemma.
\end{exercise}

The \textit{angle} between two nonzero vectors $v, w \in V$ is defined as the unique $\theta \in [0,\pi]$ satisfying
$$\cos \theta = \frac{\langle v,w \rangle}{|v||w|}.$$

Two vectors $v, w \in V$ are said to be \textit{orthogonal} if $\langle v,w \rangle = 0$, which means that either their angle is $\pi/2$ or one of the vectors is zero. If $S \subseteq V$ is a linear subspace, the set $S^\perp \subseteq V$, consisting of all vectors in $V$ that are orthogonal to every vector in $S$, is also a linear subspace, called the \textit{orthogonal complement} of $S$. Vectors $v_1,\ldots,v_k$ are called \textit{orthonormal} if they are of length $1$ and pairwise orthogonal, or equivalently if $\langle v_i, v_j \rangle = \delta_{ij}$ (where $\delta_{ij}$ is the \textit{Kronecker delta symbol} defined in Appendix B; see (B.1)). The following well-known proposition shows that every finite-dimensional inner product space has an orthonormal basis.

\begin{proposition}[Gram-Schmidt Algorithm]
\label{prop:gram-schmidt}
\lean{LeeLib.Ch02.exists_orthonormalBasis_span_image_Iic_eq}
\leanok
\dcref{ch2:2.3}
Let $V$ be an $n$-dimensional inner product space, and suppose $(v_1,\ldots,v_n)$ is any ordered basis for $V$. Then there is an orthonormal ordered basis $(b_1,\ldots,b_n)$ satisfying the following conditions:
$$\text{span}(b_1,\ldots,b_k) = \text{span}(v_1,\ldots,v_k) \quad \text{for each } k = 1,\ldots,n.$$
\end{proposition}

\begin{proof}
\leanok
The basis vectors $b_1,\ldots,b_n$ are defined recursively by
$$b_1 = \frac{v_1}{|v_1|},$$
$$b_j = \frac{v_j - \sum_{i=1}^{j-1}\langle v_j, b_i \rangle b_i}{\left|v_j - \sum_{i=1}^{j-1}\langle v_j, b_i \rangle b_i\right|}, \quad 2 \le j \le n.$$

Because $v_1 \ne 0$ and $v_j \notin \text{span}(b_1,\ldots,b_{j-1})$ for each $j \ge 2$, the denominators are all nonzero. These vectors satisfy the span conditions by construction, and are orthonormal by direct computation.
\end{proof}

If two vector spaces $V$ and $W$ are both equipped with inner products, denoted by $\langle \cdot, \cdot \rangle_V$ and $\langle \cdot, \cdot \rangle_W$, respectively, then a map $F : V \to W$ is called a \emph{linear isometry} if it is a vector space isomorphism that preserves inner products: $\langle F(v), F(v') \rangle_W = \langle v, v' \rangle_V$.

\begin{proposition}[Classification of Inner Product Spaces]
\label{prop:inner-product-spaces-linearly-isometric}
\lean{LeeLib.Ch02.nonempty_linearIsometryEquiv_of_finrank_eq}
\leanok
\uses{prop:gram-schmidt}
All inner product spaces of the same finite dimension are linearly isometric to each other.
\end{proposition}

\begin{proof}
\leanok
Let $V$ and $W$ be inner product spaces of dimension $n$. Each has an orthonormal basis: apply Proposition \ref{prop:gram-schmidt} to any ordered basis. Choose orthonormal bases $(v_1,\ldots,v_n)$ for $V$ and $(w_1,\ldots,w_n)$ for $W$, and let $F : V \to W$ be the unique linear map with $F(v_i) = w_i$; it is an isomorphism, since it carries a basis to a basis. Writing $v = \sum_i a^i v_i$ and $v' = \sum_j b^j v_j$, orthonormality of the two bases gives
\[
\langle F(v), F(v') \rangle_W = \Big\langle \sum_i a^i w_i, \sum_j b^j w_j \Big\rangle_W = \sum_i a^i b^i = \Big\langle \sum_i a^i v_i, \sum_j b^j v_j \Big\rangle_V = \langle v, v' \rangle_V ,
\]
so $F$ preserves inner products and is therefore a linear isometry.
\end{proof}

\section{Riemannian Metrics}

To extend these geometric ideas to abstract smooth manifolds, we define a structure that amounts to a smoothly varying choice of inner product on each tangent space.

\begin{definition}[Riemannian Metric]
\label{def:riemannian-metric}
\lean{LeeLib.Ch02.RiemannianMetric}
\leanok
Let $M$ be a smooth manifold. A \textbf{Riemannian metric} on $M$ is a smooth covariant $2$-tensor field $g$ on $M$ whose value $g_p$ at each $p \in M$ is an inner product on $T_p M$; thus $g$ is a symmetric $2$-tensor field that is positive definite, in the sense that $g_p(v,v) \geq 0$ for each $p \in M$ and each $v \in T_p M$, with equality if and only if $v = 0$. A \textbf{Riemannian manifold} is a pair $(M,g)$ consisting of a smooth manifold $M$ together with a Riemannian metric $g$ on $M$.
\end{definition}

If $M$ is understood to be endowed with a specific Riemannian metric, we sometimes say ``$M$ is a Riemannian manifold.''

The next proposition shows that Riemannian metrics exist in great abundance.

\begin{lemma}[Convexity of the set of inner products]
\label{lem:convex-inner-products}
\lean{LeeLib.Ch02.convex_symm_posDef}
\leanok
Let $M$ be a smooth manifold and let $p \in M$. The set of inner products on $T_p M$ is a convex subset of the space of bilinear forms on $T_p M$: if $q_1, q_2$ are inner products and $a, b \geq 0$ satisfy $a + b = 1$, then $a q_1 + b q_2$ is an inner product.
\end{lemma}

\begin{proof}
\leanok
Symmetry is immediate from the symmetry of $q_1$ and $q_2$, since
\[
(a q_1 + b q_2)(u,v) = a q_1(u,v) + b q_2(u,v) = a q_1(v,u) + b q_2(v,u) = (a q_1 + b q_2)(v,u).
\]
For positive definiteness, let $v \neq 0$. Since $a + b = 1$ and $a, b \geq 0$, at least one of $a, b$ is strictly positive; say $a > 0$ (the case $b > 0$ is symmetric). Then $a q_1(v,v) > 0$ because $q_1$ is positive definite, and $b q_2(v,v) \geq 0$ because $b \geq 0$ and $q_2(v,v) > 0$. Hence
\[
(a q_1 + b q_2)(v,v) = a q_1(v,v) + b q_2(v,v) > 0. \qedhere
\]
\end{proof}

\begin{lemma}[Positive definite forms are coercive]
\label{lem:posdef-coercive}
\lean{LeeLib.Ch02.isVonNBounded_of_posDef}
\leanok
Let $V$ be a finite-dimensional normed real vector space and let $q$ be a bilinear form on $V$ with $q(v,v) > 0$ for every $v \neq 0$. Then the sublevel set $\{v \in V : q(v,v) < 1\}$ is bounded. The reason is that $q$ is coercive: there is a constant $c > 0$ with $q(v,v) \geq c |v|^2$ for all $v \in V$.
\end{lemma}

\begin{proof}
\leanok
If $V = \{0\}$ the statement is vacuous, so assume $V \neq \{0\}$. The unit sphere $S = \{v : |v| = 1\}$ is compact, because $V$ is finite dimensional, and nonempty. The function $v \mapsto q(v,v)$ is continuous, so it attains a minimum $c = q(v_0, v_0)$ at some $v_0 \in S$. Since $|v_0| = 1$ we have $v_0 \neq 0$, so $c > 0$ by positive definiteness.

Now let $v \in V$. If $v = 0$ both sides vanish. Otherwise $u = |v|^{-1} v$ lies in $S$, so $q(u,u) \geq c$, and bilinearity gives
\[
q(v,v) = q(|v| u, |v| u) = |v|^2 q(u,u) \geq c |v|^2 .
\]
Finally, if $q(v,v) < 1$ then $c |v|^2 < 1$, so $|v| < c^{-1/2}$; thus the sublevel set is contained in the ball of radius $c^{-1/2}$.
\end{proof}

\begin{proposition}[Every smooth manifold admits a Riemannian metric]
\label{prop:existence-riemannian-metric}
\lean{LeeLib.Ch02.exists_riemannianMetric}
\leanok
\uses{def:riemannian-metric}
\dcref{ch2:2.4}
Every smooth manifold admits a Riemannian metric.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-metric,lem:convex-inner-products,lem:posdef-coercive}
Let $M$ be a smooth $n$-manifold, and let $\{U_\alpha\}$ be the cover of $M$ by chart domains, with charts $\varphi_\alpha : U_\alpha \to \mathbb{R}^n$.

On each $U_\alpha$ define a Riemannian metric $g_\alpha$ by pulling back the Euclidean metric $\bar{g}$ of Example \ref{ex:euclidean-metric} along the chart:
\[
g_\alpha = \varphi_\alpha^* \bar{g}, \qquad\text{that is}\qquad
(g_\alpha)_p(v,w) = \langle d(\varphi_\alpha)_p (v), d(\varphi_\alpha)_p (w)\rangle
\]
for $p \in U_\alpha$ and $v, w \in T_p M$. This is a smooth symmetric $2$-tensor field on $U_\alpha$, and it is positive definite because $d(\varphi_\alpha)_p$ is a linear isomorphism, so $d(\varphi_\alpha)_p(v) = 0$ only for $v = 0$.

Let $\{\psi_\alpha\}$ be a smooth partition of unity subordinate to $\{U_\alpha\}$, which exists because $M$ is a smooth manifold, hence second countable and Hausdorff. Define
\[
g = \sum_\alpha \psi_\alpha \, g_\alpha ,
\]
where each summand $\psi_\alpha g_\alpha$ is extended by zero outside $\operatorname{supp}\psi_\alpha \subseteq U_\alpha$. The family $\{\operatorname{supp}\psi_\alpha\}$ is locally finite, so the sum has finitely many nonzero terms near each point and $g$ is a smooth $2$-tensor field.

It remains to check that $g_p$ is an inner product on $T_p M$ for each $p$. Fix $p$ and let $A = \{\alpha : \psi_\alpha(p) \neq 0\}$, a finite nonempty set. The weights $\{\psi_\alpha(p)\}_{\alpha \in A}$ are nonnegative and sum to $1$, and for $\alpha \in A$ we have $p \in U_\alpha$, so $(g_\alpha)_p$ is an inner product on $T_p M$. Hence $g_p = \sum_{\alpha \in A} \psi_\alpha(p) (g_\alpha)_p$ is a convex combination of inner products on $T_p M$, and is therefore an inner product by Lemma \ref{lem:convex-inner-products}. By Lemma \ref{lem:posdef-coercive} the inner product $g_p$ induces the given topology on $T_p M$. Thus $g$ is a Riemannian metric on $M$.
\end{proof}

\begin{exercise}[Use a partition of unity to prove the preceding proposition]
\label{exer:partition-unity-proof}
\lean{LeeLib.Ch02.exists_riemannianMetric}
\leanok
\uses{prop:existence-riemannian-metric}
\dcref{ch2:2.5}
Use a partition of unity to prove the preceding proposition.
\end{exercise}

We will give a number of examples of Riemannian metrics, along with several systematic methods for constructing them, later in this chapter and in the next.

If $M$ is a smooth manifold with boundary, a Riemannian metric on $M$ is defined in exactly the same way: a smooth symmetric $2$-tensor field $g$ that is positive definite everywhere. A \textbf{Riemannian manifold with boundary} is a pair $(M, g)$, where $M$ is a smooth manifold with boundary and $g$ is a Riemannian metric on $M$. Many of the results we will discuss in this book work equally well for manifolds with or without boundary, with the same proofs, and in such cases we will state them in that generality. But when the treatment of a boundary would involve additional difficulties, we will generally restrict attention to the case of manifolds without boundary, since that is our primary interest. Many problems involving Riemannian manifolds with boundary can be addressed by embedding into a larger manifold without boundary and extending the Riemannian metric arbitrarily to the larger manifold; see Proposition A.31 in Appendix A.

A Riemannian metric is not the same as a metric in the sense of metric spaces (though, as we will see later in this chapter, the two concepts are related). In this book, when we use the word ``metric'' without further qualification, it always refers to a Riemannian metric.

Let $g$ be a Riemannian metric on a smooth manifold $M$ with or without boundary. Because $g_p$ is an inner product on $T_p M$ for each $p \in M$, we often use the following angle-bracket notation for $v, w \in T_p M$:
\[
\langle v, w \rangle_g = g_p(v, w).
\]

Using this inner product, we can define lengths of tangent vectors, angles between nonzero tangent vectors, and orthogonality of tangent vectors as described above. The length of a vector $v \in T_p M$ is denoted by $|v|_g = \langle v, v \rangle_g^{1/2}$. If the metric is understood, we sometimes omit it from the notation, and write $\langle v, w \rangle$ and $|v|$ in place of $\langle v, w \rangle_g$ and $|v|_g$, respectively.

The starting point for Riemannian geometry is the following fundamental example.

\begin{definition}[The Euclidean Metric]
\label{ex:euclidean-metric}
\lean{LeeLib.Ch02.euclideanMetric}
\leanok
\uses{def:riemannian-metric}
\dcref{ch2:2.6}
The \emph{Euclidean metric} is the Riemannian metric $\bar{g}$ on $\mathbb{R}^n$ whose value at each $x \in \mathbb{R}^n$ is the usual dot product on $T_x\mathbb{R}^n$ under the natural identification $T_x \mathbb{R}^n \cong \mathbb{R}^n$. This means that for $v,w \in T_x \mathbb{R}^n$ written in standard coordinates $(x^1, \ldots, x^n)$ as $v = \sum_i v^i \partial_i|_x$, $w = \sum_j w^j \partial_j|_x$, we have
\[
\langle v, w \rangle_{\bar{g}} = \sum_{i=1}^n v^i w^i.
\]
More generally, any inner product space carries the Riemannian metric whose value at each point is that inner product; the Euclidean metric on $\mathbb{R}^n$ is the case of the dot product.
\end{definition}

When working with $\mathbb{R}^n$ as a Riemannian manifold, we always assume we are using the Euclidean metric unless otherwise specified.

\section{Isometries}

Suppose $(M, g)$ and $(\widetilde{M}, \widetilde{g})$ are Riemannian manifolds with or without boundary. An \emph{isometry from} $(M,g)$ \emph{to} $(\widetilde{M},\widetilde{g})$ is a diffeomorphism $\varphi: M \to \widetilde{M}$ such that $\varphi^* \widetilde{g} = g$. Unwinding the definitions shows that this is equivalent to the requirement that $\varphi$ be a smooth bijection and each differential $d\varphi_p: T_p M \to T_{\varphi(p)} \widetilde{M}$ be a linear isometry. We say $(M, g)$ and $(\widetilde{M}, \widetilde{g})$ are \emph{isometric} if there exists an isometry between them.

\begin{definition}[Metric-Preserving Maps]
\label{def:metric-preserving}
\lean{LeeLib.Ch02.IsMetricPreserving}
\leanok
\uses{def:riemannian-metric}
Let $(M,g)$ and $(\widetilde{M}, \widetilde{g})$ be Riemannian manifolds with or without boundary. A map $\varphi: M \to \widetilde{M}$ \emph{preserves the metric} if $\varphi^* \widetilde{g} = g$, that is, if
\[
  \widetilde{g}_{\varphi(p)}\bigl(d\varphi_p(v), d\varphi_p(w)\bigr) \;=\; g_p(v,w)
  \qquad \text{for all } p \in M \text{ and } v, w \in T_p M.
\]
Equivalently, each differential $d\varphi_p : T_p M \to T_{\varphi(p)} \widetilde{M}$ is a linear isometry onto its image.
\end{definition}

\begin{definition}[Isometry]
\label{def:isometry}
\lean{LeeLib.Ch02.IsIsometry}
\leanok
\uses{def:metric-preserving,def:riemannian-metric}
An \emph{isometry} from $(M,g)$ to $(\widetilde{M},\widetilde{g})$ is a diffeomorphism $\varphi: M \to \widetilde{M}$ that preserves the metric.
\end{definition}

\begin{lemma}[Metric-Preserving Maps Are Immersions]
\label{lem:metric-preserving-immersion}
\lean{LeeLib.Ch02.IsMetricPreserving.injective_mfderiv}
\leanok
\uses{def:metric-preserving,lem:pullback-immersion-metric}
If $\varphi: M \to \widetilde{M}$ preserves the metric, then $d\varphi_p$ is injective for every $p \in M$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:pullback-immersion-metric}
Since $\varphi^* \widetilde{g} = g$ and $g$ is positive definite, the pullback $\varphi^* \widetilde{g}$ is positive definite. By Lemma \ref{lem:pullback-immersion-metric} this holds if and only if $d\varphi_p$ is injective at every $p$.
\end{proof}

A composition of isometries and the inverse of an isometry are again isometries, so being isometric is an equivalence relation on the class of Riemannian manifolds with or without boundary. Our subject, \emph{Riemannian geometry}, is concerned primarily with properties of Riemannian manifolds that are preserved by isometries.

\begin{lemma}[The Inverse of an Isometry]
\label{lem:isometry-symm}
\lean{LeeLib.Ch02.IsIsometry.symm}
\leanok
\uses{def:isometry}
If $\varphi$ is an isometry from $(M,g)$ to $(\widetilde{M}, \widetilde{g})$, then $\varphi^{-1}$ is an isometry from $(\widetilde{M}, \widetilde{g})$ to $(M,g)$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:isometry,def:metric-preserving}
The inverse of a diffeomorphism is a diffeomorphism, so only metric preservation is at issue. Fix $q \in \widetilde{M}$ and put $p = \varphi^{-1}(q)$. Differentiating $\varphi \circ \varphi^{-1} = \mathrm{id}_{\widetilde{M}}$ at $q$ gives $d\varphi_p \circ d(\varphi^{-1})_q = \mathrm{id}_{T_q \widetilde{M}}$. Hence for $v, w \in T_q\widetilde{M}$, applying $\varphi^* \widetilde{g} = g$ at $p$ to the vectors $d(\varphi^{-1})_q(v)$ and $d(\varphi^{-1})_q(w)$ yields
\[
  g_p\bigl(d(\varphi^{-1})_q v,\, d(\varphi^{-1})_q w\bigr)
  = \widetilde{g}_{\varphi(p)}\bigl(d\varphi_p d(\varphi^{-1})_q v,\, d\varphi_p d(\varphi^{-1})_q w\bigr)
  = \widetilde{g}_q(v, w),
\]
using $\varphi(p) = q$. This is exactly $(\varphi^{-1})^* g = \widetilde{g}$.
\end{proof}

\begin{definition}[Local Isometry]
\label{def:local-isometry}
\lean{LeeLib.Ch02.IsLocalIsometry}
\leanok
\uses{def:isometry,def:metric-preserving,def:riemannian-metric}
If $(M, g)$ and $(\widetilde{M}, \widetilde{g})$ are Riemannian manifolds, a map $\varphi: M \to \widetilde{M}$ is a \emph{local isometry} if each point $p \in M$ has a neighborhood $U$ such that $\varphi|_U$ is an isometry onto an open subset of $\widetilde{M}$.

Equivalently, $\varphi$ is a local diffeomorphism that preserves the metric: the two conditions agree because preserving the metric is a condition on the differentials $d\varphi_p$ separately, so it holds on a neighborhood of every point if and only if it holds at every point.
\end{definition}

The substance of the following characterization is its converse half, which produces the local diffeomorphism out of nothing but the metric condition and a count of dimensions.

\begin{lemma}[Metric-Preserving Maps in Equal Dimensions Are Local Diffeomorphisms]
\label{lem:metric-preserving-equidim-local-diffeomorphism}
\lean{LeeLib.Ch02.IsMetricPreserving.isLocalDiffeomorph}
\leanok
\uses{def:metric-preserving,lem:metric-preserving-immersion,thm:inverse-function-theorem-manifolds}
Let $(M,g)$ and $(\widetilde{M},\widetilde{g})$ be Riemannian manifolds of the same dimension and let $\varphi: M \to \widetilde{M}$ be a smooth map with $\varphi^* \widetilde{g} = g$. Then $\varphi$ is a local diffeomorphism.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:metric-preserving-immersion,thm:inverse-function-theorem-manifolds}
Fix $p \in M$. By Lemma \ref{lem:metric-preserving-immersion} the differential $d\varphi_p : T_p M \to T_{\varphi(p)}\widetilde{M}$ is injective. Since $\dim M = \dim \widetilde{M}$, these tangent spaces are vector spaces of the same finite dimension, so an injective linear map between them is bijective, and $d\varphi_p$ is therefore invertible. By the inverse function theorem for manifolds, Theorem \ref{thm:inverse-function-theorem-manifolds}, $\varphi$ restricts to a diffeomorphism from a neighborhood of $p$ onto an open subset of $\widetilde{M}$. As $p$ was arbitrary, $\varphi$ is a local diffeomorphism.
\end{proof}

\begin{exercise}
\label{exer:characterize-local-isometry}
\dcref{ch2:2.7}
\lean{LeeLib.Ch02.isLocalIsometry_iff_isMetricPreserving}
\leanok
\uses{def:local-isometry,def:metric-preserving,def:isometry,lem:metric-preserving-equidim-local-diffeomorphism}
Prove that if $(M,g)$ and $(\widetilde{M}, \widetilde{g})$ are Riemannian manifolds of the same dimension, a smooth map $\varphi: M \to \widetilde{M}$ is a local isometry if and only if $\varphi^* \widetilde{g} = g$.
\end{exercise}

\begin{proof}
\leanok
\uses{def:local-isometry,lem:metric-preserving-equidim-local-diffeomorphism}
Suppose first that $\varphi$ is a local isometry, and let $p \in M$. Then $\varphi$ restricts, on some neighborhood $U$ of $p$, to an isometry onto an open subset of $\widetilde{M}$, so in particular $\widetilde{g}_{\varphi(p)}(d\varphi_p v, d\varphi_p w) = g_p(v,w)$ for all $v,w \in T_pM$. Since every point of $M$ lies in such a $U$, this identity holds at every point, which is $\varphi^*\widetilde{g} = g$. This half uses neither the smoothness hypothesis nor the equality of dimensions.

Conversely, suppose $\varphi^* \widetilde{g} = g$. By Lemma \ref{lem:metric-preserving-equidim-local-diffeomorphism}, $\varphi$ is a local diffeomorphism, so each $p \in M$ has a neighborhood $U$ on which $\varphi$ restricts to a diffeomorphism onto an open subset of $\widetilde{M}$. That restriction preserves the metric, because $\varphi$ does at every point of $M$ and metric preservation is a pointwise condition on the differential. Hence $\varphi|_U$ is an isometry onto an open subset, and $\varphi$ is a local isometry.
\end{proof}

A Riemannian $n$-manifold is said to be \emph{flat} if it is locally isometric to a Euclidean space, that is, if every point has a neighborhood that is isometric to an open set in $\mathbb{R}^n$ with its Euclidean metric. Problem 2-1 shows that all Riemannian 1-manifolds are flat; but we will see later that this is far from the case in higher dimensions.

An isometry from $(M,g)$ to itself is called an \emph{isometry of $(M,g)$}. The set of all isometries of $(M,g)$ is a group under composition, called the \emph{isometry group of $(M,g)$}; it is denoted by $\text{Iso}(M,g)$, or sometimes just $\text{Iso}(M)$ if the metric is understood.

A deep theorem of Sumner B. Myers and Norman E. Steenrod [MS39] shows that if $M$ has finitely many components, then $\text{Iso}(M,g)$ has a topology and smooth structure making it into a finite-dimensional Lie group acting smoothly on $M$. We will neither prove nor use the Myers--Steenrod theorem, but if you are interested, a good source for the proof is [Kob72].

\section{Local Representations for Metrics}

Suppose $(M,g)$ is a Riemannian manifold with or without boundary. If $(x^1,\ldots,x^n)$ are any smooth local coordinates on an open subset $U \subseteq M$, then $g$ can be written locally in $U$ as
\begin{equation*}
g = g_{ij} dx^i \otimes dx^j \tag{2.7}
\end{equation*}
for some collection of $n^2$ smooth functions $g_{ij}$ for $i,j = 1,\ldots,n$. (Here and throughout the book, we use the Einstein summation convention; see p. 375.) The component functions of this tensor field constitute a matrix-valued function $(g_{ij})$, characterized by $g_{ij}(p) = \langle \partial_i|_p, \partial_j|_p \rangle$, where $\partial_i = \partial/\partial x^i$ is the $i$th coordinate vector field; this matrix is symmetric in $i$ and $j$ and depends smoothly on $p \in U$. If $v = v^i \partial_i|_p$ is a vector in $T_p M$ such that $g_{ij}(p)v^i = 0$, it follows that $\langle v,v \rangle = g_{ij}(p) v^i v^j = 0$, which implies $v = 0$; thus the matrix $(g_{ij}(p))$ is always nonsingular. The notation for $g$ can be shortened by expressing it in terms of the symmetric product (see Appendix B): using the symmetry of $g_{ij}$, we compute
\begin{align*}
g &= g_{ij} dx^i \otimes dx^j \\
  &= \tfrac{1}{2}(g_{ij} dx^i \otimes dx^j + g_{ji} dx^i \otimes dx^j) \\
  &= \tfrac{1}{2}(g_{ij} dx^i \otimes dx^j + g_{ij} dx^j \otimes dx^i) \\
  &= g_{ij} dx^i dx^j.
\end{align*}

For example, the Euclidean metric on $\mathbb{R}^n$ (Example 2.6) can be expressed in standard coordinates in several ways:
\begin{equation*}
\bar{g} = \sum_i dx^i dx^i = \sum_i (dx^i)^2 = \delta_{ij} dx^i dx^j. \tag{2.8}
\end{equation*}

The matrix of $\bar{g}$ in these coordinates is thus $\bar{g}_{ij} = \delta_{ij}$.

More generally, if $(E_1,\ldots,E_n)$ is any smooth local frame for $TM$ on an open subset $U \subseteq M$ and $(e^1,\ldots,e^n)$ is its dual coframe, we can write $g$ locally in $U$ as
\begin{equation*}
g = g_{ij} e^i \otimes e^j, \tag{2.9}
\end{equation*}
where $g_{ij}(p) = \langle E_i|_p, E_j \rangle_p$, and the matrix-valued function $(g_{ij})$ is symmetric and smooth as before.

A Riemannian metric $g$ acts on smooth vector fields $X, Y \in \mathfrak{X}(M)$ to yield a real-valued function $\langle X,Y \rangle$. In terms of any smooth local frame, this function is expressed locally by $\langle X,Y \rangle = g_{ij} X^i Y^j$ and therefore is smooth. Similarly, we obtain a nonnegative real-valued function $|X| = \langle X,X \rangle^{1/2}$, which is continuous everywhere and smooth on the open subset where $X \neq 0$.

\begin{lemma}[Smoothness of the pairing of vector fields]
\label{lem:inner-product-vector-fields-smooth}
\lean{LeeLib.Ch02.RiemannianMetric.contMDiff_innerAt}
\leanok
\uses{def:riemannian-metric}
\dcref{ch2:2.2}
Let $(M,g)$ be a Riemannian manifold with or without boundary. If $X, Y \in \mathfrak{X}(M)$ are smooth vector fields, then $\langle X, Y \rangle : M \to \mathbb{R}$ is smooth.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric}
Smoothness is local, so it suffices to check it on the domain $U$ of an arbitrary smooth local frame $(E_i)$. Write $X = X^i E_i$ and $Y = Y^j E_j$ on $U$; the component functions $X^i, Y^j$ are smooth because $X$ and $Y$ are smooth sections and $(E_i)$ is a smooth frame. By bilinearity,
\[
\langle X, Y\rangle = g_{ij} X^i Y^j , \qquad g_{ij} = \langle E_i, E_j\rangle ,
\]
and the $g_{ij}$ are smooth on $U$ because $g$ is a smooth $2$-tensor field and the $E_i$ are smooth. Thus $\langle X, Y\rangle$ is a finite sum of products of smooth real-valued functions on $U$, hence smooth on $U$.
\end{proof}

\begin{lemma}[Continuity of the length of a vector field]
\label{lem:length-vector-field-continuous}
\lean{LeeLib.Ch02.RiemannianMetric.continuous_normAt}
\leanok
\uses{lem:inner-product-vector-fields-smooth}
\dcref{ch2:2.2}
Let $(M,g)$ be a Riemannian manifold with or without boundary. If $X \in \mathfrak{X}(M)$ is a smooth vector field, then $|X| = \langle X, X\rangle^{1/2} : M \to \mathbb{R}$ is continuous.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:inner-product-vector-fields-smooth}
By Lemma \ref{lem:inner-product-vector-fields-smooth} the function $\langle X, X\rangle$ is smooth, hence continuous, and it is nonnegative by positive semidefiniteness of $g$. The square root $t \mapsto \sqrt{t}$ is continuous on $[0,\infty)$ — including at $t = 0$, where it fails only to be differentiable. So $|X| = \sqrt{\cdot} \circ \langle X, X\rangle$ is a composition of continuous functions, hence continuous.
\end{proof}

\begin{lemma}[Smoothness of the length away from the zero set]
\label{lem:length-vector-field-smooth}
\lean{LeeLib.Ch02.RiemannianMetric.contMDiffAt_normAt}
\leanok
\uses{lem:inner-product-vector-fields-smooth}
\dcref{ch2:2.2}
Let $(M,g)$ be a Riemannian manifold with or without boundary and let $X \in \mathfrak{X}(M)$ be a smooth vector field. Then $|X|$ is smooth at every $p \in M$ with $X_p \neq 0$; that is, $|X|$ is smooth on the open subset of $M$ where $X$ does not vanish.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:inner-product-vector-fields-smooth}
Let $p \in M$ with $X_p \neq 0$. By positive definiteness of $g_p$ we have $\langle X_p, X_p\rangle > 0$, and by Lemma \ref{lem:inner-product-vector-fields-smooth} the function $\langle X, X\rangle$ is smooth. Since $t \mapsto \sqrt{t}$ is smooth on $(0,\infty)$, it is smooth at $\langle X_p, X_p \rangle$, so the composition $|X| = \sqrt{\cdot} \circ \langle X, X\rangle$ is smooth at $p$. As this holds at every point of the set $\{q : X_q \neq 0\}$, which is open by continuity of $X$, the function $|X|$ is smooth there.
\end{proof}

A local frame $\{E_i\}$ for $M$ on an open set $U$ is said to be an \textbf{orthonormal frame} if the vectors $E_1|_p, \ldots, E_n|_p$ are an orthonormal basis for $T_p M$ at each $p \in U$. Equivalently, $(E_i)$ is an orthonormal frame if and only if
\[
\langle E_i, E_j \rangle = \delta_{ij},
\]
in which case $g$ has the local expression
\[
g = (e^1)^2 + \cdots + (e^n)^2,
\]
where $(e^i)^2$ denotes the symmetric product $e^i e^i = e^i \otimes e^i$.

\begin{proposition}[Existence of Orthonormal Frames]
\label{prop:existence-orthonormal-frames}
\lean{LeeLib.Ch02.exists_orthonormalFrame}
\leanok
\dcref{ch2:2.8}
\uses{prop:gram-schmidt,def:riemannian-metric}

Let $(M, g)$ be a Riemannian $n$-manifold with or without boundary. If $(X_i)$ is any smooth local frame for $TM$ over an open subset $U \subseteq M$, then there is a smooth orthonormal frame $(E_j)$ over $U$ such that $\text{span}\{E_1|_p, \ldots, E_k|_p\} = \text{span}\{X_1|_p, \ldots, X_k|_p\}$ for each $k = 1, \ldots, n$ and each $p \in U$. In particular, for every $p \in M$, there is a smooth orthonormal frame $(E_j)$ defined on some neighborhood of $p$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:gram-schmidt}
Applying the Gram–Schmidt algorithm to the vectors $(X_1|_p, \ldots, X_n|_p)$ at each $p \in U$, we obtain an ordered $n$-tuple of rough orthonormal vector fields $(E_1, \ldots, E_n)$ over $U$ satisfying the span conditions. Because the vectors whose norms appear in the denominators of (2.5)–(2.6) are nowhere vanishing, those formulas show that each vector field $E_j$ is smooth. The last statement of the proposition follows by applying this construction to any smooth local frame in a neighborhood of $p$.

In more detail, write $u_j$ for the unnormalized Gram–Schmidt vectors, defined recursively by
\[
u_j = X_j - \sum_{i < j} \frac{\langle u_i, X_j\rangle}{\langle u_i, u_i\rangle}\, u_i ,
\qquad E_j = \frac{u_j}{|u_j|} .
\]
Fix $p \in U$. Since $(X_1|_p, \ldots, X_n|_p)$ is a basis of $T_pM$, it is linearly independent, so $u_j|_p \neq 0$ for every $j$, and hence $\langle u_j|_p, u_j|_p\rangle > 0$ by positive definiteness of $g_p$. Therefore every denominator appearing above is nonvanishing on all of $U$.

Smoothness of $E_j$ now follows by induction on $j$. The function $p \mapsto \langle v|_p, w|_p\rangle_g$ is smooth whenever $v, w$ are smooth vector fields, because $g$ is a smooth $2$-tensor field. So, assuming inductively that $u_i$ is smooth for every $i < j$, each coefficient $\langle u_i, X_j\rangle / \langle u_i, u_i\rangle$ is a quotient of smooth functions with nowhere-vanishing denominator, hence smooth; therefore $u_j$, being $X_j$ minus a finite sum of smooth multiples of smooth vector fields, is smooth. Finally $|u_j| = \langle u_j, u_j\rangle^{1/2}$ is smooth and nonvanishing on $U$, because $t \mapsto \sqrt{t}$ is smooth away from $t = 0$ and $\langle u_j, u_j\rangle > 0$ there; so $E_j = |u_j|^{-1} u_j$ is smooth.

Orthonormality of $(E_j)$ and the span conditions $\text{span}\{E_1|_p, \ldots, E_k|_p\} = \text{span}\{X_1|_p, \ldots, X_k|_p\}$ are the standard properties of the Gram–Schmidt algorithm of Proposition \ref{prop:gram-schmidt}, applied in the inner product space $(T_pM, g_p)$ at each $p \in U$.
\end{proof}

\begin{remark}
\label{rem:coordinate-frame-warning}
\uses{prop:existence-orthonormal-frames}

\textbf{Warning:} A common mistake made by beginners is to assume that one can find coordinates near $p$ such that the coordinate frame $(\partial_i)$ is orthonormal. Proposition 2.8 does not show this. In fact, as we will see in Chapter 7, this is possible only when the metric is flat, that is, locally isometric to the Euclidean metric.
\end{remark}

For a Riemannian manifold $(M, g)$ with or without boundary, we define the \textbf{unit tangent bundle} to be the subset $UTM \subseteq TM$ consisting of unit vectors:
\begin{equation}
UTM = \{(p, v) \in TM : |v|_g = 1\}.
\tag{2.10}
\end{equation}

\begin{proposition}[Properties of the Unit Tangent Bundle]
\label{prop:unit-tangent-bundle-properties}
\dcref{ch2:2.9}

If $(M, g)$ is a Riemannian manifold with or without boundary, its unit tangent bundle $UTM$ is a smooth, properly embedded codimension-1 submanifold with boundary in $TM$, with $\partial(UTM) = \pi^{-1}(\partial M)$ (where $\pi : UTM \to M$ is the canonical projection). The unit tangent bundle is connected if and only if $M$ is connected, and compact if and only if $M$ is compact.
\end{proposition}

\begin{exercise}
\label{exer:orthonormal-frames-proposition}
\dcref{ch2:2.10}
Use local orthonormal frames to prove the preceding proposition.
\end{exercise}

\section{Methods for Constructing Riemannian Metrics}

Many examples of Riemannian manifolds arise naturally as submanifolds, products, and quotients of other Riemannian manifolds. In this section, we introduce some of the tools for constructing such metrics.

\subsection*{Riemannian Submanifolds}

Every submanifold of a Riemannian manifold automatically inherits a Riemannian metric, and many interesting Riemannian metrics are defined in this way. The key fact is the following lemma.

\begin{lemma}
\label{lem:pullback-immersion-metric}
\lean{LeeLib.Ch02.pullbackForm_posDef_iff_immersion}
\leanok
\uses{def:riemannian-metric}
\dcref{ch2:2.11}
Suppose $(\tilde{M}, \tilde{g})$ is a Riemannian manifold with or without boundary, $M$ is a smooth manifold with or without boundary, and $F : M \to \tilde{M}$ is a smooth map. The smooth 2-tensor field $g = F^*\tilde{g}$ is a Riemannian metric on $M$ if and only if $F$ is an immersion.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric}
By definition the pullback is the $2$-tensor field
\[
(F^*\tilde{g})_p(v,w) = \tilde{g}_{F(p)}(dF_p v,\, dF_p w), \qquad v, w \in T_p M .
\]
It is symmetric at every $p$, whatever $F$ is, because $\tilde{g}_{F(p)}$ is symmetric, and it is a smooth $2$-tensor field because $F$ is smooth and $\tilde{g}$ is a smooth $2$-tensor field. So $F^*\tilde{g}$ is a Riemannian metric if and only if it is positive definite at every point, and it suffices to prove that this is equivalent to injectivity of every $dF_p$.

Suppose first that $F$ is an immersion, and let $p \in M$ and $v \in T_p M$ with $v \neq 0$. Then $dF_p v \neq 0$, since $dF_p$ is injective, so positive definiteness of $\tilde{g}_{F(p)}$ gives
\[
(F^*\tilde{g})_p(v,v) = \tilde{g}_{F(p)}(dF_p v,\, dF_p v) > 0 .
\]
Hence $F^*\tilde{g}$ is positive definite.

Conversely, suppose $F^*\tilde{g}$ is positive definite, and fix $p \in M$. Since $dF_p$ is linear, it suffices to show its kernel is trivial. Let $v \in T_p M$ with $dF_p v = 0$, and suppose for contradiction that $v \neq 0$. Then
\[
0 < (F^*\tilde{g})_p(v,v) = \tilde{g}_{F(p)}(dF_p v,\, dF_p v) = \tilde{g}_{F(p)}(0,0) = 0,
\]
a contradiction. So $v = 0$, and $dF_p$ is injective. As $p$ was arbitrary, $F$ is an immersion.
\end{proof}

\begin{exercise}
\label{exer:prove-lemma-2-11}
\lean{LeeLib.Ch02.pullbackForm_posDef_iff_immersion}
\leanok
\dcref{ch2:2.12}
\uses{lem:pullback-immersion-metric}
Prove Lemma 2.11.
\end{exercise}

\begin{definition}[The metric induced by an immersion]
\label{def:induced-metric}
\lean{LeeLib.Ch02.pullbackMetric}
\leanok
\uses{lem:pullback-immersion-metric,def:riemannian-metric}
\dcref{ch2:2.11}
Suppose $(\tilde{M}, \tilde{g})$ is a Riemannian manifold with or without boundary and $F : M \to \tilde{M}$ is a smooth immersion. The \emph{metric induced by $F$} is the Riemannian metric $F^*\tilde{g}$ on $M$, which is a Riemannian metric by Lemma \ref{lem:pullback-immersion-metric}.
\end{definition}

On the other hand, if $M$ is already endowed with a given Riemannian metric $g$, an immersion or embedding $F : M \to \tilde{M}$ satisfying $F^*\tilde{g} = g$ is called an \emph{isometric immersion} or \emph{isometric embedding}, respectively. Which terminology is used depends on whether the metric on $M$ is considered to be given independently of the immersion or not.

The most important examples of induced metrics occur on submanifolds. Suppose $M \subseteq \tilde{M}$ is an (immersed or embedded) submanifold, with or without boundary. The \emph{induced metric on $M$} is the metric $g = \iota^*\tilde{g}$ induced by the inclusion map $\iota : M \hookrightarrow \tilde{M}$. With this metric, $M$ is called a \emph{Riemannian submanifold} (or Riemannian submanifold with boundary) of $\tilde{M}$. We always consider submanifolds (with or without boundary) of Riemannian manifolds to be endowed with the induced metrics unless otherwise specified.

If $(M, g)$ is a Riemannian submanifold of $(\tilde{M}, \tilde{g})$, then for every $p \in M$ and $v, w \in T_p M$, the definition of the induced metric reads
\[
g_p(v,w) = \tilde{g}_p(d\iota_p(v), d\iota_p(w)).
\]

Because we usually identify $T_p M$ with its image in $T_p \tilde{M}$ under $d\iota_p$, and think of $d\iota_p$ as an inclusion map, what this really amounts to is $g_p(v,w) = \tilde{g}(v,w)$ for $v,w \in T_p M$. In other words, the induced metric $g$ is just the restriction of $\tilde{g}$ to vectors tangent to $M$. Many of the examples of Riemannian metrics that we will encounter are obtained in this way, starting with the following.

\begin{definition}[Spheres]
\label{ex:spheres}
\lean{LeeLib.Ch02.roundMetric}
\leanok
\uses{def:induced-metric,ex:euclidean-metric}
\dcref{ch2:2.13}
For each positive integer $n$, the unit $n$-sphere $S^n \subseteq \mathbb{R}^{n+1}$ is an embedded $n$-dimensional submanifold. The Riemannian metric induced on $S^n$ by the Euclidean metric is denoted by $\bar{g}$ and known as the \textit{round metric} or \textit{standard metric} on $S^n$.

Concretely, the round metric is the pullback $\iota^* \bar{g}$ of the Euclidean metric along the inclusion $\iota : S^n \hookrightarrow \mathbb{R}^{n+1}$. This is a Riemannian metric by Lemma \ref{lem:pullback-immersion-metric}, because $\iota$ is a smooth immersion: it is smooth, and its differential $d\iota_p : T_p S^n \to T_p\mathbb{R}^{n+1}$ is injective at every $p$, with image the orthogonal complement $(\mathbb{R}p)^\perp$ of the radial direction.
\end{definition}

The next lemma describes one of the most important tools for studying Riemannian submanifolds. If $(\tilde{M}, \tilde{g})$ is an $m$-dimensional smooth Riemannian manifold and $M \subseteq \tilde{M}$ is an $n$-dimensional submanifold (both with or without boundary), a local frame $(E_1, \ldots, E_m)$ for $\tilde{M}$ on an open subset $U \subseteq \tilde{M}$ is said to be \textit{adapted to} $M$ if the first $n$ vector fields $(E_1, \ldots, E_n)$ are tangent to $M$. In case $\tilde{M}$ has empty boundary (so that slice coordinates are available), adapted local orthonormal frames are easy to find.

\begin{proposition}[Existence of Adapted Orthonormal Frames]
\label{prop:adapted-orthonormal-frames}
\dcref{ch2:2.14}
\uses{def:riemannian-metric,lem:partial-frame-extension,lem:pushforward-smooth,lem:ambient-tangent-bundle-riemannian,prop:existence-orthonormal-frames,lem:pullback-immersion-metric}
Let $(\tilde{M}, \tilde{g})$ be a Riemannian manifold (without boundary), and let $M \subseteq \tilde{M}$ be an embedded smooth submanifold with or without boundary. Given $p \in M$, there exist a neighborhood $\tilde{U}$ of $p$ in $\tilde{M}$ and a smooth orthonormal frame for $\tilde{M}$ on $\tilde{U}$ that is adapted to $M$.
\end{proposition}

\begin{exercise}
\label{exer:gram-schmidt-adapted-frame}
\dcref{ch2:2.15}
\uses{prop:adapted-orthonormal-frames,prop:slice-coordinates,prop:existence-orthonormal-frames}
Prove the preceding proposition. [Hint: Apply the Gram--Schmidt algorithm to a coordinate frame in slice coordinates (see Prop. A.22).]
\end{exercise}

Slice coordinates are only one way to produce a frame whose first $n$ members are tangent to $M$. Presenting the submanifold by an immersion $F : M \to \tilde M$ instead, the tangent directions are supplied by the pushforward of a frame of $TM$, and the frame lives on the ambient tangent bundle $F^* T\tilde M$ of Lemma \ref{lem:ambient-tangent-bundle-riemannian}. The following four lemmas carry out that route; they use no slice coordinates and so apply to immersed as well as embedded submanifolds. Throughout, a \textit{local $k$-frame} for a vector bundle $V \to B$ over $u \subseteq B$ means a $k$-tuple of smooth sections of $V$ over $u$ that is linearly independent at every point of $u$ — not necessarily spanning.

\begin{lemma}[Ordered extension of a linearly independent family]
\label{lem:ordered-linear-extension}
\lean{LeeLib.Ch02.exists_linearIndependent_castAdd}
\leanok
Let $V$ be a vector space with $\dim V = m$, let $k, d \geq 0$ satisfy $k + d \leq m$, and let $(c_1, \ldots, c_k)$ be a linearly independent family in $V$. Then there is a linearly independent family $(w_1, \ldots, w_{k+d})$ in $V$ with $w_i = c_i$ for $1 \leq i \leq k$.
\end{lemma}

\begin{proof}
\leanok
Induct on $d$. If $d = 0$, take $w = c$.

Suppose the claim holds for $d$, and let $k + d + 1 \leq m$. Then $\dim \mathrm{span}(c_1, \ldots, c_k) = k < m$, so $\mathrm{span}(c_1, \ldots, c_k) \neq V$ and there is $y \in V \setminus \mathrm{span}(c_1, \ldots, c_k)$. The family $(c_1, \ldots, c_k, y)$ is linearly independent: in a vanishing linear combination $\sum_i a_i c_i + b y = 0$, a nonzero $b$ would give $y = -b^{-1}\sum_i a_i c_i \in \mathrm{span}(c_1,\ldots,c_k)$, contrary to the choice of $y$; so $b = 0$, and then all $a_i$ vanish by independence of the $c_i$. Applying the inductive hypothesis to this family of $k+1$ vectors and to $d$ (note $(k+1) + d \le m$) yields a linearly independent family of $k + 1 + d = k + (d+1)$ vectors whose first $k+1$ members are $c_1, \ldots, c_k, y$; in particular its first $k$ members are $c_1, \ldots, c_k$.
\end{proof}

\begin{lemma}[Linear independence of smooth sections is an open condition]
\label{lem:open-linear-independence}
\lean{LeeLib.Ch02.exists_isLocalIndepOn_nhds}
\leanok
Let $V \to B$ be a smooth vector bundle with model fibre $F$, let $u \subseteq B$ be open, and let $s_1, \ldots, s_k$ be smooth sections of $V$ over $u$. If $(s_1|_{x_0}, \ldots, s_k|_{x_0})$ is linearly independent in $V_{x_0}$ for some $x_0 \in u$, then there is an open set $v$ with $x_0 \in v \subseteq u$ such that $(s_1|_x, \ldots, s_k|_x)$ is linearly independent in $V_x$ for every $x \in v$.
\end{lemma}

\begin{proof}
\leanok
First, the set of linearly independent $k$-tuples is open in $F^k$. Let $(v_1, \ldots, v_k)$ be linearly independent and let $L : \mathbb{R}^k \to F$ be the injective linear map $a \mapsto \sum_i a_i v_i$. As $\mathbb{R}^k$ is finite dimensional, injectivity of $L$ gives a $\kappa > 0$ with $\|La\| \geq \kappa \|a\|$ for all $a$. Let $(v_1', \ldots, v_k')$ satisfy $\delta := \sum_i \|v_i' - v_i\| < \kappa$, and let $L'$ be the corresponding map. Then
\[
\|L'a - La\| = \Big\| \sum_i a_i (v_i' - v_i) \Big\| \leq \max_i |a_i| \sum_i \|v_i' - v_i\| \leq \delta \|a\|,
\]
so $\|L'a\| \geq \|La\| - \delta\|a\| \geq (\kappa - \delta)\|a\| > 0$ for $a \neq 0$. Hence $L'$ is injective, i.e. $(v_i')$ is linearly independent.

Now choose a local trivialization $\Phi$ of $V$ over an open set containing $x_0$, and write $\Phi_x : V_x \to F$ for its fibrewise linear isomorphism. Put $c(x) = (\Phi_x(s_1|_x), \ldots, \Phi_x(s_k|_x)) \in F^k$. Each component is continuous at $x_0$ because $s_i$ is a smooth section, so $c$ is continuous at $x_0$; and $c(x_0)$ is linearly independent, being the image of a linearly independent tuple under the isomorphism $\Phi_{x_0}$. By the previous paragraph linear independence holds on a neighbourhood of $c(x_0)$, so by continuity there is an open $v$ with $x_0 \in v$, contained in $u$ and in the domain of $\Phi$, such that $c(x)$ is linearly independent for every $x \in v$. For such $x$, $(s_i|_x)_i$ is the image of $c(x)$ under the isomorphism $\Phi_x^{-1}$ and is therefore linearly independent.
\end{proof}

\begin{lemma}[Extension of a partial frame]
\label{lem:partial-frame-extension}
\lean{LeeLib.Ch02.exists_isLocalFrameOn_extend}
\leanok
\uses{lem:ordered-linear-extension,lem:open-linear-independence}
Let $V \to B$ be a smooth vector bundle of rank $m$, let $u \subseteq B$ be open, and let $(s_1, \ldots, s_k)$ be a local $k$-frame for $V$ over $u$. Then $k \leq m$, and for each $x_0 \in u$ there are an open set $v$ with $x_0 \in v \subseteq u$ and a smooth local frame $(E_1, \ldots, E_m)$ for $V$ over $v$ with $E_i = s_i$ on $v$ for $1 \leq i \leq k$.
\end{lemma}

\begin{proof}
\leanok
Fix a local trivialization $\Phi$ of $V$ over an open set containing $x_0$, with fibrewise isomorphisms $\Phi_x : V_x \to F$. The vectors $c_i := \Phi_{x_0}(s_i|_{x_0})$, $1 \leq i \leq k$, are linearly independent in $F$, whence $k \leq \dim F = m$. By Lemma \ref{lem:ordered-linear-extension} with $d = m - k$ they extend to a linearly independent family $(w_1, \ldots, w_m)$ in $F$ with $w_i = c_i$ for $i \leq k$; being $m$ independent vectors in the $m$-dimensional space $F$, it is a basis.

Over $u \cap \mathrm{dom}\,\Phi$ define
\[
E_i = s_i \quad (1 \leq i \leq k), \qquad E_j|_x = \Phi_x^{-1}(w_j) \quad (k < j \leq m).
\]
For $i \leq k$ the section $E_i$ is smooth by hypothesis; for $j > k$ it is smooth because it is constant in the trivialization $\Phi$.

At $x_0$ we have $E_j|_{x_0} = \Phi_{x_0}^{-1}(w_j)$ for \emph{every} $j$: for $j > k$ by definition, and for $j \leq k$ because $w_j = c_j = \Phi_{x_0}(s_j|_{x_0})$. So $(E_1|_{x_0}, \ldots, E_m|_{x_0})$ is the image of the basis $(w_j)_j$ under the isomorphism $\Phi_{x_0}^{-1}$, hence a basis of $V_{x_0}$; in particular it is linearly independent. By Lemma \ref{lem:open-linear-independence} there is an open $v$ with $x_0 \in v \subseteq u \cap \mathrm{dom}\,\Phi$ on which $(E_1|_x, \ldots, E_m|_x)$ is linearly independent at every point. Since $\dim V_x = m$, a linearly independent family of $m$ vectors in $V_x$ spans $V_x$; so $(E_1, \ldots, E_m)$ is a smooth local frame for $V$ over $v$, and it agrees with $s$ in its first $k$ members by construction.
\end{proof}

\begin{lemma}[Smoothness of the pushforward of a vector field]
\label{lem:pushforward-smooth}
\lean{LeeLib.Ch02.contMDiffOn_pushforward}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian}
Let $F : M \to \tilde{M}$ be a smooth map and let $X$ be a smooth vector field on an open set $u \subseteq M$. Then $x \mapsto dF_x(X_x)$ is a smooth section of the ambient tangent bundle $F^* T\tilde{M}$ over $u$.
\end{lemma}

\begin{proof}
\leanok
Fix $x_0 \in u$, a trivialization $\Sigma$ of $TM$ around $x_0$ and a trivialization $\Theta$ of $T\tilde{M}$ around $F(x_0)$, with fibrewise isomorphisms $\Sigma_x$ and $\Theta_y$. By construction a trivialization of $F^* T\tilde M$ at $x_0$ is $\Theta$ read along $F$: its fibre map at $x$ is $\Theta_{F(x)}$. Hence, for $x$ in the neighbourhood of $x_0$ on which $x \in \mathrm{dom}\,\Sigma$ and $F(x) \in \mathrm{dom}\,\Theta$, the section $x \mapsto dF_x(X_x)$ read in that trivialization is
\[
x \longmapsto \Theta_{F(x)}\big(dF_x(X_x)\big) = D(x)\,\xi(x), \qquad
D(x) := \Theta_{F(x)} \circ dF_x \circ \Sigma_x^{-1}, \quad \xi(x) := \Sigma_x(X_x),
\]
the equality holding because $\Sigma_x^{-1}(\xi(x)) = X_x$.

The map $\xi$ is smooth because $X$ is a smooth section read in $\Sigma$, and $D$ — the differential of $F$ in tangent coordinates — is smooth because $F$ is smooth. Evaluating a smooth family of continuous linear maps on a smooth family of vectors is smooth, so the coordinate representation is smooth at $x_0$. As $x_0 \in u$ was arbitrary, the section is smooth on $u$.
\end{proof}

Suppose $(\tilde{M}, \tilde{g})$ is a Riemannian manifold and $M \subseteq \tilde{M}$ is a smooth submanifold with or without boundary in $\tilde{M}$. Given $p \in M$, a vector $v \in T_p \tilde M$ is said to be \textit{normal to} $M$ if $\langle v,w\rangle = 0$ for every $w \in T_p M$. The space of all vectors normal to $M$ at $p$ is a subspace of $T_p \tilde{M}$, called the \textit{normal space at} $p$ and denoted by $N_p M = (T_p M)^\perp$. At each $p \in M$, the ambient tangent space $T_p \tilde{M}$ splits as an orthogonal direct sum $T_p \tilde{M} = T_p M \oplus N_p M$. A section $N$ of the ambient tangent bundle $T\tilde{M}|_M$ is called a \textit{normal vector field along} $M$ if $N_p \in N_p M$ for each $p \in M$. The set
\[
NM = \bigsqcup_{p \in M} N_p M
\]
is called the \textit{normal bundle of} $M$.

\begin{lemma}[The Ambient Tangent Bundle Along a Submanifold]
\label{lem:ambient-tangent-bundle-riemannian}
\lean{Bundle.ContMDiffRiemannianMetric.pullback}
\leanok
\uses{def:riemannian-metric,def:induced-metric}
Let $(\tilde M, \tilde g)$ be a Riemannian manifold and let $F : M \to \tilde M$ be a smooth map. Then the pullback bundle $F^* T\tilde M$ over $M$, whose fibre at $p$ is $T_{F(p)}\tilde M$, carries the ambient inner product $\tilde g_{F(p)}$ on each fibre, and this inner product varies smoothly with $p$; that is, $F^* T\tilde M$ is a Riemannian vector bundle over $M$.

When $F$ is the inclusion of a submanifold $M \subseteq \tilde M$, this bundle is the \emph{ambient tangent bundle along $M$}, written $T\tilde M|_M$: the bundle in which the splitting $T_p\tilde M = T_p M \oplus N_p M$ and the projections $\pi^\top, \pi^\perp$ of Proposition \ref{prop:normal-bundle} live.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric}
The fibre of $F^* T\tilde M$ at $p$ \emph{is} $T_{F(p)}\tilde M$, so it carries $\tilde g_{F(p)}$ with no choice involved; symmetry and positive definiteness are those of $\tilde g$ at the point $F(p)$. Only the smooth dependence on $p$ requires an argument.

Fix $p_0 \in M$. Smoothness of a metric on a vector bundle means that, read through a local trivialization of the bundle near $p_0$, the metric is a smooth map into the space of bilinear forms on the model fibre. By construction, a trivialization of $F^* T\tilde M$ over $p_0$ is the pullback along $F$ of a trivialization $\Phi$ of $T\tilde M$ over $F(p_0)$: its fibre map at $p$ is the fibre map of $\Phi$ at $F(p)$. Hence, for $p$ in the neighbourhood of $p_0$ on which $F(p)$ stays in the domain of $\Phi$, the metric of $F^* T\tilde M$ read through the pulled-back trivialization at $p$ equals the metric $\tilde g$ read through $\Phi$ at $F(p)$.

The map $p \mapsto (\tilde g \text{ read through } \Phi)(F(p))$ is a composition of $F$, which is smooth, with the coordinate representation of $\tilde g$, which is smooth because $\tilde g$ is a smooth metric. So the coordinate representation of the pulled-back metric is smooth at $p_0$, and $p_0$ was arbitrary.
\end{proof}

\begin{proposition}[Existence of Adapted Orthonormal Frames, immersed form]
\label{prop:adapted-orthonormal-frames-immersed}
\lean{LeeLib.Ch02.exists_adapted_orthonormalFrame}
\leanok
\uses{def:riemannian-metric,lem:partial-frame-extension,lem:pushforward-smooth,lem:ambient-tangent-bundle-riemannian,prop:existence-orthonormal-frames,lem:pullback-immersion-metric}
Let $(\tilde M, \tilde g)$ be a Riemannian $m$-manifold, let $M$ be an $n$-manifold and let
$F : M \to \tilde M$ be a smooth immersion. Then $n \le m$, and every $p \in M$ has a
neighborhood $v$ carrying a smooth orthonormal frame $(E_1, \ldots, E_m)$ for the ambient
tangent bundle $F^* T\tilde M$ of Lemma \ref{lem:ambient-tangent-bundle-riemannian} such that,
at every $x \in v$,
\[
  \operatorname{span}(E_1|_x, \ldots, E_n|_x) \;=\; \operatorname{im} dF_x .
\]
Since $\operatorname{im} dF_x$ is Lee's $T_x M$ viewed inside $T_{F(x)}\tilde M$, this says
exactly that the frame is adapted to $M$. It is Proposition \ref{prop:adapted-orthonormal-frames}
with the submanifold presented by an immersion rather than by an embedding, so it applies to
immersed submanifolds as well; the price is that the frame is a frame of $F^* T\tilde M$ over
$M$ rather than of $T\tilde M$ over an open subset of $\tilde M$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:partial-frame-extension,lem:pushforward-smooth,lem:ambient-tangent-bundle-riemannian,prop:existence-orthonormal-frames}
Fix $p \in M$ and let $(X_1, \ldots, X_n)$ be any smooth local frame for $TM$ on an open
$u \ni p$ — for instance the frame induced on the base set of a trivialization of $TM$ at $p$ by
a basis of the model space. Put
\[
  Z_i|_x := dF_x(X_i|_x), \qquad i = 1, \ldots, n, \quad x \in u .
\]
By Lemma \ref{lem:pushforward-smooth} each $Z_i$ is a smooth section of $F^* T\tilde M$ over $u$.
At each $x \in u$ the family $(Z_i|_x)_i$ is linearly independent: $(X_i|_x)_i$ is, and $dF_x$ is
injective because $F$ is an immersion. So $(Z_1, \ldots, Z_n)$ is a local $n$-frame for
$F^* T\tilde M$ over $u$.

By Lemma \ref{lem:partial-frame-extension} applied to this $n$-frame — the rank of
$F^* T\tilde M$ is $m$ — we get $n \le m$, an open $v$ with $p \in v \subseteq u$, and a smooth
local frame $(Y_1, \ldots, Y_m)$ for $F^* T\tilde M$ over $v$ with $Y_i = Z_i$ on $v$ for
$i \le n$.

By Lemma \ref{lem:ambient-tangent-bundle-riemannian} the ambient metric $\tilde g$ makes
$F^* T\tilde M$ a Riemannian bundle, so Proposition \ref{prop:existence-orthonormal-frames}
applies to it: the fibrewise Gram--Schmidt orthonormalization $(E_1, \ldots, E_m)$ of
$(Y_1, \ldots, Y_m)$ is again a smooth local frame over $v$, is orthonormal with respect to
$\tilde g$ at every point of $v$, and satisfies the flag condition
$\operatorname{span}(E_1|_x, \ldots, E_k|_x) = \operatorname{span}(Y_1|_x, \ldots, Y_k|_x)$ for
every $k$ and every $x \in v$.

It remains to identify the span of the first $n$ members. Fix $x \in v$ and $i \le n$. The flag
condition at $k = i$ gives
$E_i|_x \in \operatorname{span}(Y_1|_x, \ldots, Y_i|_x) = \operatorname{span}(Z_1|_x, \ldots, Z_i|_x)$,
and each $Z_j|_x = dF_x(X_j|_x)$ lies in $\operatorname{im} dF_x$, which is a subspace; hence
$E_i|_x \in \operatorname{im} dF_x$. Therefore
$\operatorname{span}(E_1|_x, \ldots, E_n|_x) \subseteq \operatorname{im} dF_x$. Both sides have
dimension $n$: the left because $(E_1|_x, \ldots, E_n|_x)$ is orthonormal, hence linearly
independent, and the right because $dF_x$ is injective on the $n$-dimensional $T_x M$. A subspace
of a space of the same finite dimension is the whole space, so the inclusion is an equality.
\end{proof}

\begin{definition}[The Normal Space]
\label{def:normal-space}
\lean{LeeLib.Ch02.normalSpace}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian}
Let $(\tilde M, \tilde g)$ be a Riemannian $m$-manifold and let $F : M \to \tilde M$ present an
$n$-manifold $M$ as a smooth immersion. For $x \in M$, the \emph{normal space to $M$ at $x$} is
the subspace of the ambient tangent space $T_{F(x)}\tilde M$ given by
\[
  N_x M = \bigl\{\, v \in T_{F(x)}\tilde M \;:\; \tilde g_{F(x)}(v, w) = 0
    \ \text{ for every } w \in T_x M \,\bigr\},
\]
where, as elsewhere in this chapter, $T_x M$ denotes the image $dF_x(T_x M) \subseteq
T_{F(x)}\tilde M$. It is a linear subspace, being the intersection of the kernels of the linear
functionals $\tilde g_{F(x)}(\cdot, w)$ for $w \in T_x M$.
\end{definition}

\begin{lemma}[Local Orthonormal Frames for the Normal Space]
\label{lem:normal-space-orthonormal-frame}
\lean{LeeLib.Ch02.exists_orthonormalFrame_normalSpace}
\leanok
\uses{def:normal-space,prop:adapted-orthonormal-frames-immersed}
Let $(\tilde M, \tilde g)$ be a Riemannian $m$-manifold and let $F : M \to \tilde M$ present an
$n$-manifold $M$ as a smooth immersion. Then $n \le m$, and every $p \in M$ has a neighbourhood
$v$ carrying smooth sections $N_1, \ldots, N_{m-n}$ of the ambient tangent bundle $F^*T\tilde M$
over $v$ whose values at each $x \in v$ form an orthonormal basis of $N_x M$.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:adapted-orthonormal-frames-immersed}
By Proposition \ref{prop:adapted-orthonormal-frames-immersed} there are $n \le m$, a
neighbourhood $v$ of $p$, and a smooth local frame $(E_1, \ldots, E_m)$ for $F^*T\tilde M$ over
$v$ that is orthonormal with respect to $\tilde g$ at every $x \in v$ and whose first $n$ members
span $T_x M$ at every $x \in v$. Put $N_j = E_{n+j}$ for $1 \le j \le m-n$; these are smooth
sections over $v$, and they are orthonormal at each $x \in v$ because $(E_a)_a$ is.

Fix $x \in v$. Each $N_j|_x$ is normal: for $i \le n$ we have $n + j \ne i$, so
$\tilde g(N_j|_x, E_i|_x) = 0$ by orthonormality; since $E_1|_x, \ldots, E_n|_x$ span $T_x M$ and
$\tilde g$ is linear in its second argument, $\tilde g(N_j|_x, w) = 0$ for every $w \in T_x M$.
Hence $\operatorname{span}(N_1|_x, \ldots, N_{m-n}|_x) \subseteq N_x M$, as $N_x M$ is a subspace.

Conversely let $w \in N_x M$. Since $(E_a|_x)_{a=1}^m$ is a basis of $T_{F(x)}\tilde M$, write
$w = \sum_{a=1}^m c_a E_a|_x$. Pairing with $E_b|_x$ and using orthonormality gives
\[
  \tilde g(E_b|_x, w) = \sum_{a=1}^m c_a \, \tilde g(E_b|_x, E_a|_x) = c_b .
\]
For $b \le n$ we have $E_b|_x \in T_x M$, and $w$ is normal, so $c_b = \tilde g(E_b|_x, w) = 0$.
Therefore $w = \sum_{j=1}^{m-n} c_{n+j} N_j|_x$ lies in
$\operatorname{span}(N_1|_x, \ldots, N_{m-n}|_x)$.

The two inclusions give $\operatorname{span}(N_1|_x, \ldots, N_{m-n}|_x) = N_x M$; together with
orthonormality, $(N_j|_x)_j$ is an orthonormal basis of $N_x M$.
\end{proof}

\begin{proposition}[The Normal Bundle]
\label{prop:normal-bundle}
\lean{LeeLib.Ch02.contMDiffVectorBundle_normalSpace}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian,def:normal-space,lem:normal-space-orthonormal-frame,lem:local-frame-criterion-for-subbundles}
\dcref{ch2:2.16}
If $\tilde{M}$ is a Riemannian $m$-manifold and $M \subseteq \tilde{M}$ is an immersed or embedded $n$-dimensional submanifold with or without boundary, then $NM$ is a smooth rank-$(m-n)$ vector subbundle of the ambient tangent bundle $T\tilde{M}|_M$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:normal-space-orthonormal-frame,lem:local-frame-criterion-for-subbundles,lem:ambient-tangent-bundle-riemannian}
By Lemma \ref{lem:ambient-tangent-bundle-riemannian} the ambient tangent bundle $T\tilde M|_M = F^*T\tilde M$ is a smooth vector bundle over $M$ carrying the fibre metric $\tilde g$, so Lemma \ref{lem:local-frame-criterion-for-subbundles} applies to it. Take the family of subspaces $D_p = N_p M \subseteq T_p\tilde M$, each of dimension $m - n$.

Its hypothesis is Lemma \ref{lem:normal-space-orthonormal-frame}: every $p \in M$ has a neighbourhood $v$ carrying smooth sections $N_1, \ldots, N_{m-n}$ of $T\tilde M|_M$ whose values at each $x \in v$ form an orthonormal basis --- in particular a basis --- of $N_x M$. Lemma \ref{lem:local-frame-criterion-for-subbundles} therefore makes $NM = \bigcup_{p \in M} N_p M$ a smooth rank-$(m-n)$ subbundle of $T\tilde M|_M$.
\end{proof}

\begin{remark}
\label{rem:normal-bundle-projections}
\uses{prop:normal-bundle}
Lee's statement of the preceding proposition also asserts the existence of smooth bundle homomorphisms
\[
\pi^\top : T\tilde{M}|_M \to TM, \quad \pi^\perp : T\tilde{M}|_M \to NM,
\]
the \emph{tangential and normal projections}, restricting at each $p \in M$ to the orthogonal projections of $T_p\tilde M$ onto $T_p M$ and $N_p M$. These are the fibrewise orthogonal projections of the splitting $T_p\tilde M = T_p M \oplus N_p M$; their smoothness follows from the same adapted orthonormal frame, in which $\pi^\top$ and $\pi^\perp$ are the truncations of the coordinate expansion to the first $n$ and the last $m-n$ members. They are not part of the node above.
\end{remark}

\begin{proof}
\uses{thm:local-embedding-theorem,lem:local-frame-criterion-for-subbundles,prop:adapted-orthonormal-frames-immersed,lem:normal-space-orthonormal-frame}
Given any point $p \in M$, Theorem A.16 shows that there is a neighborhood $U$ of $p$ in $\widetilde{M}$ that is embedded in $\widetilde{M}$, and then Proposition 2.14 shows that there is a smooth orthonormal frame $(E_1, \ldots, E_m)$ that is adapted to $U$ on some neighborhood $\widetilde{U}$ of $p$ in $\widetilde{M}$. This means that the restrictions of $(E_1, \ldots, E_n)$ to $U \cap \widetilde{U}$ form a local orthonormal frame for $M$. Given such an adapted frame, the restrictions of the last $m - n$ vector fields $(E_{n+1}, \ldots, E_m)$ to $M$ form a smooth local frame for $NM$, so it follows from Lemma A.34 that $NM$ is a smooth subbundle.

The bundle homomorphisms $\pi^\top$ and $\pi^\perp$ are defined pointwise as orthogonal projections onto the tangent and normal spaces, respectively, which shows that they are uniquely defined. In terms of an adapted orthonormal frame, they can be written
\[
\pi^\top(X^1 E_1 + \cdots + X^m E_m) = X^1 E_1 + \cdots + X^n E_n,
\]
\[
\pi^\perp(X^1 E_1 + \cdots + X^m E_m) = X^{n+1} E_{n+1} + \cdots + X^m E_m,
\]
which shows that they are smooth.
\end{proof}

In case $\widetilde{M}$ is a manifold with boundary, the preceding constructions do not always work, because there is not a fully general construction of slice coordinates in that case. However, there is a satisfactory result in case the submanifold is the boundary itself, using boundary coordinates in place of slice coordinates.

Suppose $(M, g)$ is a Riemannian manifold with boundary. We will always consider $\partial M$ to be a Riemannian submanifold with the induced metric.

\begin{proposition}[Existence of Outward-Pointing Normal]
\label{prop:outward-pointing-normal}
\dcref{ch2:2.17}
If $(M, g)$ is a smooth Riemannian manifold with boundary, the normal bundle to $\partial M$ is a smooth rank-1 vector bundle over $\partial M$, and there is a unique smooth outward-pointing unit normal vector field along all of $\partial M$.
\end{proposition}

\begin{exercise}
\label{exer:prove-outward-normal}
\dcref{ch2:2.18}
\uses{prop:induced-orientation-boundary}
Prove this proposition. [Hint: Use the paragraph preceding Prop. B.17 as a starting point.]
\end{exercise}

Computations on a submanifold $M \subset \widetilde{M}$ are usually carried out most conveniently in terms of a \emph{smooth local parametrization}: this is a smooth map $X : U \to \widetilde{M}$, where $U$ is an open subset of $\mathbb{R}^n$ (or $\mathbb{R}^n_+$ in case $M$ has a boundary), such that $X(U)$ is an open subset of $M$, and such that $X$, regarded as a map from $U$ into $M$, is a diffeomorphism onto its image. Note that we can think of $X$ either as a map into $M$ or as a map into $\widetilde{M}$; both maps are typically denoted by the same symbol $X$. If we put $V = X(U) \subseteq M$ and $\varphi = X^{-1} : V \to U$, then $(V, \varphi)$ is a smooth coordinate chart on $M$.

Suppose $(M, g)$ is a Riemannian submanifold of $(\widetilde{M}, \widetilde{g})$ and $X : U \to \widetilde{M}$ is a smooth local parametrization of $M$. The coordinate representation of $g$ in these coordinates is given by the following 2-tensor field on $U$:
\[
(\varphi^{-1})^* g = X^* g = X^* \iota^* \widetilde{g} = (\iota \circ X)^* \widetilde{g}.
\]
Since $\iota \circ X$ is just the map $X$ itself, regarded as a map into $\tilde{M}$, this is really just $X^*\tilde{g}$. The simplicity of the formula for the pullback of a tensor field makes this expression exceedingly easy to compute, once a coordinate expression for $\tilde{g}$ is known. For example, if $M$ is an immersed $n$-dimensional Riemannian submanifold of $\mathbb{R}^m$ and $X : U \to \mathbb{R}^m$ is a smooth local parametrization of $M$, the induced metric on $U$ is just
\[
g = X^*\tilde{g} = \sum_{i=1}^m (dX^i)^2 = \sum_{i=1}^m \left(\sum_{j=1}^n \frac{\partial X^i}{\partial u^j} du^j\right)^2 = \sum_{i=1}^m \sum_{j,k=1}^n \frac{\partial X^i}{\partial u^j}\frac{\partial X^i}{\partial u^k} du^j du^k.
\]

\begin{example}[Metrics in Graph Coordinates]
\label{ex:metrics-in-graph-coordinates}
\dcref{ch2:2.19}

If $U \subseteq \mathbb{R}^n$ is an open set and $f : U \to \mathbb{R}$ is a smooth function, then the \textbf{graph of $f$} is the subset $\Gamma(f) = \{(x, f(x)) : x \in U\} \subseteq \mathbb{R}^{n+1}$, which is an embedded submanifold of dimension $n$. It has a global parametrization $X : U \to \mathbb{R}^{n+1}$ called a \textbf{graph parametrization}, given by $X(u) = (u, f(u))$; the corresponding coordinates $(u^1,\ldots,u^n)$ on $M$ are called \textbf{graph coordinates}. In graph coordinates, the induced metric of $\Gamma(f)$ is
\[
X^*\tilde{g} = X^*\left((dx^1)^2 + \cdots + (dx^{n+1})^2\right) = (du^1)^2 + \cdots + (du^n)^2 + df^2.
\]

Applying this to the upper hemisphere of $S^2$ with the parametrization $X : \mathbb{R}^2 \to \mathbb{R}^3$ given by
\[
X(u,v) = \left(u, v, \sqrt{1-u^2-v^2}\right),
\]
we see that the round metric on $S^2$ can be written locally as
\[
\hat{g} = X^*\bar{g} = du^2 + dv^2 + \left(\frac{u\,du + v\,dv}{\sqrt{1-u^2-v^2}}\right)^2
\]
\[
= \frac{(1-v^2) du^2 + (1-u^2) dv^2 + 2uv\,du\,dv}{1-u^2-v^2}.
\]
\end{example}

\begin{example}[Surfaces of Revolution]
\label{ex:surfaces-of-revolution}
\dcref{ch2:2.20}

Let $H$ be the half-plane $\{(r,z) : r > 0\}$, and suppose $C \subseteq H$ is an embedded 1-dimensional submanifold. The \textbf{surface of revolution} determined by $C$ is the subset $S_C \subseteq \mathbb{R}^3$ given by
\[
S_C = \left\{(x,y,z) : \left(\sqrt{x^2 + y^2}, z\right) \in C\right\}.
\]

The set $C$ is called its \textbf{generating curve} (see Fig. 2.1). Every smooth local parametrization $\gamma(t) = (a(t), b(t))$ for $C$ yields a smooth local parametrization for $S_C$ of the form
\begin{equation}
X(t, \theta) = (a(t) \cos \theta, a(t) \sin \theta, b(t)), \tag{2.11}
\end{equation}

provided that $(t, \theta)$ is restricted to a sufficiently small open set in the plane. The $t$-coordinate curves $t \mapsto X(t, \theta_0)$ are called \textbf{meridians}, and the $\theta$-coordinate curves $\theta \mapsto X(t_0, \theta)$ are called \textbf{latitude circles}. The induced metric on $S_C$ is
% [FIGURE: Two diagrams showing a surface of revolution — left diagram shows a curve C in the rz-plane, right diagram shows the 3D surface obtained by rotating C around the z-axis]
\begin{align*}
X^*\bar{g} &= d(a(t)\cos\theta)^2 + d(a(t)\sin\theta)^2 + d(b(t))^2 \\
&= (a'(t)\cos\theta \, dt - a(t)\sin\theta \, d\theta)^2 + (a'(t)\sin\theta \, dt + a(t)\cos\theta \, d\theta)^2 + (b'(t)dt)^2 \\
&= (a'(t)^2 + b'(t)^2)dt^2 + a(t)^2 d\theta^2.
\end{align*}

In particular, if $\gamma$ is a \emph{unit-speed curve} (meaning that $|\gamma'(t)|^2 = a'(t)^2 + b'(t)^2 \equiv 1$), this reduces to $dt^2 + a(t)^2 d\theta^2$.

Here are some examples of surfaces of revolution and their induced metrics.

\begin{itemize}
\item If $C$ is the semicircle $r^2 + z^2 = 1$, parametrized by $\gamma(\varphi)=(\sin\varphi, \cos\varphi)$ for $0 < \varphi < \pi$, then $S_C$ is the unit sphere (minus the north and south poles). The map $X(\varphi, \theta) = (\sin\varphi\cos\theta, \sin\varphi\sin\theta, \cos\varphi)$ constructed above is called the \emph{spherical coordinate parametrization}, and the induced metric is $d\varphi^2 + \sin^2\varphi \, d\theta^2$. (This example is the source of the terminology for meridians and latitude circles.)

\item If $C$ is the circle $(r-2)^2 + z^2 = 1$, parametrized by $\gamma(t) = (2+\cos t, \sin t)$, we obtain a torus of revolution, whose induced metric is $dt^2 + (2+\cos t)^2 d\theta^2$.

\item If $C$ is a vertical line parametrized by $\gamma(t) = (1,t)$, then $S_C$ is the unit cylinder $x^2 + y^2 = 1$, and the induced metric is $dt^2 + d\theta^2$. Note that this means that the parametrization $X: \mathbb{R}^2 \to \mathbb{R}^3$ is an isometric immersion. \quad //
\end{itemize}
\end{example}

\begin{example}[The n-Torus as a Riemannian Submanifold]
\label{ex:n-torus-riemannian-submanifold}
\dcref{ch2:2.21}
\uses{ex:covering-n-torus}

The smooth covering map $X: \mathbb{R}^n \to \mathbb{T}^n$ described in Example A.52 restricts to a smooth local parametrization on any sufficiently small open subset of $\mathbb{R}^n$, and the induced metric is equal to the Euclidean metric in $(u^i)$ coordinates, and therefore the induced metric on $\mathbb{T}^n$ is flat.
\end{example}

\begin{exercise}
\label{exer:verify-surface-revolution-examples}
\dcref{ch2:2.22}
\uses{ex:metrics-in-graph-coordinates,ex:surfaces-of-revolution,ex:n-torus-riemannian-submanifold}

Verify the claims in Examples 2.19–2.21.
\end{exercise}

\section{Riemannian Products}

Next we consider products. If $(M_1, g_1)$ and $(M_2, g_2)$ are Riemannian manifolds, the product manifold $M_1 \times M_2$ has a natural Riemannian metric $g = g_1 \oplus g_2$, called the \emph{product metric}, defined by
\begin{equation}
g_{(p_1,p_2)}((v_1, v_2), (w_1, w_2)) = g_1\big|_{p_1}(v_1, w_1) + g_2\big|_{p_2}(v_2, w_2). \tag{2.12}
\end{equation}

where $(v_1, v_2)$ and $(w_1, w_2)$ are elements of $T_{p_1}M_1 \oplus T_{p_2}M_2$, which is naturally identified with $T_{(p_1,p_2)}(M_1 \times M_2)$. Smooth local coordinates $(x^1, \ldots, x^n)$ for $M_1$ and $(x^{n+1}, \ldots, x^{n+m})$ for $M_2$ give coordinates $(x^1, \ldots, x^{n+m})$ for $M_1 \times M_2$. In terms of these coordinates, the product metric has the local expression $g = g_{ij} \, dx^i \, dx^j$, where $(g_{ij})$ is the block diagonal matrix
\[
(g_{ij}) = \begin{pmatrix} (g_1)_{ab} & 0 \\ 0 & (g_2)_{cd} \end{pmatrix};
\]
here the indices $a, b$ run from $1$ to $n$, and $c, d$ run from $n+1$ to $n+m$. Product metrics on products of three or more Riemannian manifolds are defined similarly.

The two blocks of (2.12) are the pullbacks of $g_1$ and $g_2$ along the two projections. Indeed, $d(\pi_1)_{(p_1,p_2)}(v_1,v_2) = v_1$ and $d(\pi_2)_{(p_1,p_2)}(v_1,v_2) = v_2$ under the identification $T_{(p_1,p_2)}(M_1 \times M_2) \cong T_{p_1}M_1 \oplus T_{p_2}M_2$, so
\begin{equation}
g_1 \oplus g_2 = \pi_1^* g_1 + \pi_2^* g_2 . \tag{2.12'}
\end{equation}
Each summand is a smooth $2$-tensor field, being the pullback of one along a smooth map; neither is positive definite, since a projection is not an immersion; but the sum is, which is the content of Lemma \ref{lem:product-form-positive-definite}. Assembling (2.12') from its blocks uses only the two standard facts that sections of a smooth vector bundle may be added, and rescaled by a smooth function, without losing smoothness.

\begin{lemma}[Sums of smooth sections]
\label{lem:smooth-tensor-field-add}
\lean{ContMDiff.add_section}
\mathlibok
Let $V$ be a smooth vector bundle over a smooth manifold $B$, and let $s, t$ be smooth sections of $V$. Then $s + t$ is a smooth section of $V$.
\end{lemma}

\begin{lemma}[Smooth rescalings of smooth sections]
\label{lem:smooth-tensor-field-smul}
\lean{ContMDiff.smul_section}
\mathlibok
Let $V$ be a smooth vector bundle over a smooth manifold $B$, let $s$ be a smooth section of $V$, and let $c : B \to \mathbb{R}$ be smooth. Then $b \mapsto c(b)\,s_b$ is a smooth section of $V$.
\end{lemma}

\begin{lemma}[Positive definiteness of the product form]
\label{lem:product-form-positive-definite}
\lean{LeeLib.Ch02.prodForm_pos}
\leanok
\uses{def:riemannian-metric}
Let $(M_1,g_1)$ and $(M_2,g_2)$ be Riemannian manifolds, and let $g_1 \oplus g_2$ be the symmetric $2$-tensor field on $M_1 \times M_2$ defined by (2.12). Then $g_1 \oplus g_2$ is positive definite at every point.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric}
Fix $(p_1,p_2) \in M_1 \times M_2$ and let $v = (v_1,v_2) \in T_{p_1}M_1 \oplus T_{p_2}M_2$ be nonzero. Then $v_1 \neq 0$ or $v_2 \neq 0$. By (2.12),
\[
(g_1 \oplus g_2)_{(p_1,p_2)}(v,v) = g_1|_{p_1}(v_1,v_1) + g_2|_{p_2}(v_2,v_2),
\]
and both summands are nonnegative because $g_1$ and $g_2$ are positive definite. If $v_1 \neq 0$ the first summand is strictly positive and the second is nonnegative; if $v_2 \neq 0$ the second is strictly positive and the first is nonnegative. In either case the sum is strictly positive.
\end{proof}

\begin{definition}[The Product Metric]
\label{def:product-metric}
\lean{LeeLib.Ch02.prodMetric}
\leanok
\uses{def:riemannian-metric,lem:product-form-positive-definite,lem:smooth-tensor-field-add,lem:pullback-immersion-metric}
Let $(M_1,g_1)$ and $(M_2,g_2)$ be Riemannian manifolds. The \emph{product metric} $g_1 \oplus g_2$ on $M_1 \times M_2$ is the Riemannian metric defined by (2.12). It is symmetric because $g_1$ and $g_2$ are, it is positive definite by Lemma \ref{lem:product-form-positive-definite}, and it is a smooth $2$-tensor field by (2.12') together with Lemma \ref{lem:smooth-tensor-field-add}, each summand being a pullback of a smooth $2$-tensor field along a smooth map.
\end{definition}

\begin{exercise}[Induced Metric on $T^n$]
\label{exer:product-metric-torus}
\dcref{ch2:2.23}
\uses{ex:n-torus-riemannian-submanifold,def:product-metric}
Show that the induced metric on $T^n$ described in Exercise 2.21 is equal to the product metric obtained from the usual induced metric on $S^1 \subseteq \mathbb{R}^2$.
\end{exercise}

Here is an important generalization of product metrics. Suppose $(M_1, g_1)$ and $(M_2, g_2)$ are two Riemannian manifolds, and $f : M_1 \to \mathbb{R}^+$ is a strictly positive smooth function. The \emph{warped product} $M_1 \times_f M_2$ is the product manifold $M_1 \times M_2$ endowed with the Riemannian metric $g = g_1 \oplus f^2 g_2$, defined by
\begin{equation}
g_{(p_1,p_2)}((v_1, v_2), (w_1, w_2)) = g_1\big|_{p_1}(v_1, w_1) + f(p_1)^2 \, g_2\big|_{p_2}(v_2, w_2),
\end{equation}

where $(v_1, v_2), (w_1, w_2) \in T_{p_1}M_1 \oplus T_{p_2}M_2$ as before. (Despite the similarity with the notation for product metrics, $g_1 \oplus f^2 g_2$ is generally not a product metric unless $f$ is constant.) A wide variety of metrics can be constructed in this way; here are just a few examples.

\begin{lemma}[Positive definiteness of the warped product form]
\label{lem:warped-product-form-positive-definite}
\lean{LeeLib.Ch02.warpedProductForm_pos}
\leanok
\uses{def:riemannian-metric}
Let $(M_1,g_1)$ and $(M_2,g_2)$ be Riemannian manifolds and let $f : M_1 \to \mathbb{R}$ be nowhere vanishing. Then $g_1 \oplus f^2 g_2$ is positive definite at every point.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric,lem:product-form-positive-definite}
Fix $(p_1,p_2)$ and a nonzero $v = (v_1,v_2) \in T_{p_1}M_1 \oplus T_{p_2}M_2$, so that $v_1 \neq 0$ or $v_2 \neq 0$. Since $f(p_1) \neq 0$ we have $f(p_1)^2 > 0$, and
\[
(g_1 \oplus f^2 g_2)_{(p_1,p_2)}(v,v) = g_1|_{p_1}(v_1,v_1) + f(p_1)^2\, g_2|_{p_2}(v_2,v_2).
\]
Both summands are nonnegative, by positive definiteness of $g_1$ and $g_2$ and $f(p_1)^2 > 0$. If $v_1 \neq 0$ the first is strictly positive; if $v_2 \neq 0$ then $g_2|_{p_2}(v_2,v_2) > 0$ and hence the second is strictly positive, being a product of two strictly positive numbers. In either case the sum is strictly positive.
\end{proof}

\begin{definition}[Warped Products]
\label{def:warped-product}
\lean{LeeLib.Ch02.warpedProductMetric}
\leanok
\uses{def:product-metric,def:riemannian-metric,lem:warped-product-form-positive-definite,lem:smooth-tensor-field-add,lem:smooth-tensor-field-smul}
Let $(M_1,g_1)$ and $(M_2,g_2)$ be Riemannian manifolds and let $f : M_1 \to \mathbb{R}^+$ be a strictly positive smooth function. The \emph{warped product} $M_1 \times_f M_2$ is the product manifold $M_1 \times M_2$ endowed with the Riemannian metric $g_1 \oplus f^2 g_2$.

That $g_1 \oplus f^2 g_2$ is a Riemannian metric follows as for the product metric: it is symmetric because $g_1$ and $g_2$ are, positive definite by Lemma \ref{lem:warped-product-form-positive-definite}, and smooth because
\[
g_1 \oplus f^2 g_2 = \pi_1^* g_1 + (f \circ \pi_1)^2\, \pi_2^* g_2,
\]
whose first summand is smooth as in Definition \ref{def:product-metric} and whose second is the smooth function $(f \circ \pi_1)^2$ times a smooth $2$-tensor field, hence smooth by Lemma \ref{lem:smooth-tensor-field-smul}; the sum is then smooth by Lemma \ref{lem:smooth-tensor-field-add}.

Note that $f$ enters the metric only through $f^2$, so the definition makes sense for any nowhere-vanishing smooth $f$, and only $f(p_1) \neq 0$ is used.
\end{definition}

\begin{proposition}[Constant warping gives the product metric]
\label{prop:warped-product-constant-one}
\lean{LeeLib.Ch02.warpedProductMetric_one}
\leanok
\uses{def:warped-product,def:product-metric}
With $f \equiv 1$, the warped product $M_1 \times_f M_2$ is the product manifold $M_1 \times M_2$ with the product metric $g_1 \oplus g_2$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:warped-product,def:product-metric}
Both metrics are defined on the same underlying manifold $M_1 \times M_2$, so it suffices to compare them at each point. For every $(p_1,p_2)$ and all $(v_1,v_2), (w_1,w_2) \in T_{p_1}M_1 \oplus T_{p_2}M_2$,
\[
(g_1 \oplus f^2 g_2)_{(p_1,p_2)}((v_1,v_2),(w_1,w_2)) = g_1|_{p_1}(v_1,w_1) + 1^2\, g_2|_{p_2}(v_2,w_2),
\]
which is exactly the right-hand side of (2.12). A Riemannian metric is determined by the bilinear form it assigns to each point, so the two metrics are equal.
\end{proof}

\begin{example}[Warped Products]
\label{ex:warped-products}
\dcref{ch2:2.24}
\uses{ex:surfaces-of-revolution,REF:ch2-3,def:warped-product,prop:warped-product-constant-one}

\noindent (a) With $f \equiv 1$, the warped product $M_1 \times_f M_2$ is just the space $M_1 \times M_2$ with the product metric; this is Proposition \ref{prop:warped-product-constant-one}.

\noindent (b) Every surface of revolution can be expressed as a warped product, as follows. Let $H$ be the half-plane $\{(r,z) : r > 0\}$, let $C \subseteq H$ be an embedded smooth 1-dimensional submanifold, and let $S_C \subseteq \mathbb{R}^3$ denote the corresponding surface of revolution as in Example 2.20. Endow $C$ with the Riemannian metric induced from the Euclidean metric on $H$, and let $S^1$ be endowed with its standard metric. Let $f : C \to \mathbb{R}$ be the distance to the $z$-axis: $f(r,z) = r$. Then Problem 2-3 shows that $S_C$ is isometric to the warped product $C \times_f S^1$.

(c) If we let $\rho$ denote the standard coordinate function on $\mathbb{R}^+ \subset \mathbb{R}$, then the map $\Phi(\rho,\omega) = \rho\omega$ gives an isometry from the warped product $\mathbb{R}^+ \times_\rho S^{n-1}$ to $\mathbb{R}^n \setminus \{0\}$ with its Euclidean metric (see Problem 2-4). \hfill //
\end{example}

\section{Riemannian Submersions}

Unlike submanifolds and products of Riemannian manifolds, which automatically inherit Riemannian metrics of their own, quotients of Riemannian manifolds inherit Riemannian metrics only under very special circumstances. In this section, we see what those circumstances are.

Suppose $\tilde{M}$ and $M$ are smooth manifolds, $\pi : \tilde{M} \to M$ is a smooth submersion, and $\tilde{g}$ is a Riemannian metric on $\tilde{M}$. By the submersion level set theorem (Corollary A.25), each fiber $\tilde{M}_y = \pi^{-1}(y)$ is a properly embedded smooth submanifold of $\tilde{M}$. At each point $x \in \tilde{M}$, we define two subspaces of the tangent space $T_x \tilde{M}$ as follows: the \textbf{vertical tangent space at $x$} is
\[
V_x = \ker d\pi_x = T_x(\tilde{M}_{\pi(x)})
\]
(that is, the tangent space to the fiber containing $x$), and the \textbf{horizontal tangent space at $x$} is its orthogonal complement:
\[
H_x = (V_x)^\perp.
\]
Then the tangent space $T_x \tilde{M}$ decomposes as an orthogonal direct sum $T_x \tilde{M} = H_x \oplus V_x$. Note that the vertical space is well defined for every submersion, because it does not refer to the metric; but the horizontal space depends on the metric.

A vector field on $\tilde{M}$ is said to be a \textbf{horizontal vector field} if its value at each point lies in the horizontal space at that point; a \textbf{vertical vector field} is defined similarly. Given a vector field $X$ on $M$, a vector field $\tilde{X}$ on $\tilde{M}$ is called a \textbf{horizontal lift of $X$} if $\tilde{X}$ is horizontal and $\pi$-related to $X$. (The latter property means that $d\pi_x(\tilde{X}_x) = X_{\pi(x)}$ for each $x \in \tilde{M}$.)

The next proposition is the principal tool for doing computations on Riemannian submersions.

Proposition \ref{prop:horizontal-vector-field-properties} below is part (a) of
Lee's Proposition 2.25; part (b) is Proposition
\ref{prop:horizontal-lift-of-vector-field} and part (c) is Proposition
\ref{prop:horizontal-lift-realizes-horizontal-vector}, both stated once the
horizontal lift has been constructed.

Lee derives all three parts from the rank theorem: it puts $\pi$ into the normal
form $(x^1,\ldots,x^m) \mapsto (x^1,\ldots,x^n)$, so that $V_q$ is spanned by
$\partial_{n+1}|_q,\ldots,\partial_m|_q$ on a chart, and Gram--Schmidt applied to
the reordered coordinate frame $(\partial_{n+1},\ldots,\partial_m,\partial_1,\ldots,\partial_n)$
produces a smooth orthonormal frame $(E_1,\ldots,E_m)$ whose first $m-n$ entries
span $V_q$ and whose last $n$ span $H_q$; the components of $W$ in that frame are
smooth, which gives the smoothness of $W^H$ and $W^V$. The treatment below
replaces that frame by an explicit formula for the horizontal lift. This costs
nothing in generality, and it buys the smoothness assertions directly: the lift is
a rational expression in $\tilde{g}$ and $d\pi$, and inversion of invertible
operators is smooth.

\begin{proposition}[The Smooth Horizontal--Vertical Splitting of a Vector Field]
\label{prop:horizontal-vector-field-properties}
\dcref{ch2:2.25}
\lean{LeeLib.Ch02.existsUnique_horizontal_add_vertical_field}
\leanok
\uses{def:submersion,def:vertical-space,def:horizontal-space,def:riemannian-metric,lem:tangent-space-horizontal-vertical-splitting,lem:horizontal-projection-field-smooth}

Let $\tilde{M}$ and $M$ be smooth manifolds, let $\pi : \tilde{M} \to M$ be a smooth submersion, and let $\tilde{g}$ be a Riemannian metric on $\tilde{M}$. Every smooth vector field $W$ on $\tilde{M}$ can be expressed uniquely in the form $W = W^H + W^V$, where $W^H$ is horizontal, $W^V$ is vertical, and both $W^H$ and $W^V$ are smooth.
\end{proposition}

\begin{proof}
\leanok
At each point $x \in \tilde{M}$ the tangent vector $W_x$ splits uniquely as a horizontal plus a vertical vector, by Lemma \ref{lem:tangent-space-horizontal-vertical-splitting}; explicitly the two summands are $W^H_x = L_x(d\pi_x(W_x))$ and $W^V_x = W_x - W^H_x$, where $L_x$ is the horizontal lift of Definition \ref{def:horizontal-lift-tangent}. This already proves existence and uniqueness of the decomposition into rough vector fields, since uniqueness is a pointwise assertion and involves no smoothness.

It remains to show that the two rough fields so defined are smooth. The field $W^H$ is smooth by Lemma \ref{lem:horizontal-projection-field-smooth}, and $W^V = W - W^H$ is then smooth as a difference of smooth vector fields.
\end{proof}

The smoothness of $W^H$ is the only analytic content of the proposition, and it is
isolated in the next two lemmas. The first observes that $x \mapsto d\pi_x(W_x)$ is
a smooth section along $\pi$ — a section of $\pi^* TM$ — and the second lifts such
a section back to $\tilde{M}$. The generality matters: $d\pi_x(W_x)$ depends on $x$
and not merely on $\pi(x)$, since two points of a fibre may carry unrelated values
of $W$, so $W^H$ is \emph{not} the lift of any vector field on $M$ and part (b)
does not apply to it.

\begin{lemma}[The Pushforward of a Vector Field is a Smooth Section Along the Map]
\label{lem:pushforward-along-map-smooth}
\lean{LeeLib.Ch02.contMDiff_mfderivAlong}
\leanok
\uses{def:submersion}
Let $F : \tilde{M} \to M$ be a smooth map and let $W$ be a smooth vector field on $\tilde{M}$. Then $x \mapsto dF_x(W_x)$ is a smooth section of $F^*TM$ along $F$, i.e.\ the map $x \mapsto (F(x), dF_x(W_x))$ is a smooth map $\tilde{M} \to TM$.
\end{lemma}

\begin{proof}
\leanok
The bundled differential $dF : T\tilde{M} \to TM$, $(x, v) \mapsto (F(x), dF_x(v))$, is a smooth map of tangent bundles because $F$ is smooth. The map in question is its composite with the smooth map $x \mapsto (x, W_x)$, which is smooth because $W$ is a smooth vector field.
\end{proof}

\begin{lemma}[The Horizontal Lift of a Smooth Section Along $\pi$ is Smooth]
\label{lem:horizontal-lift-along-smooth}
\lean{LeeLib.Ch02.contMDiffAt_horizontalLiftAlong}
\leanok
\uses{def:horizontal-lift-tangent,lem:horizontal-lift-smooth-in-data,def:submersion}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. Let $X$ be a smooth section along $\pi$, that is, a family $X_x \in T_{\pi(x)}M$ depending smoothly on $x \in \tilde{M}$. Then $x \mapsto L_x(X_x)$ is a smooth vector field on $\tilde{M}$.
\end{lemma}

\begin{proof}
\leanok
Smoothness is local, so fix $x_0 \in \tilde{M}$ and read both tangent bundles in trivializations around $x_0$ and $\pi(x_0)$. In those trivializations $\tilde{g}$ and $d\pi$ become smooth families $B(x)$ of positive definite bilinear forms and $A(x)$ of surjective linear maps, and $X$ becomes a smooth $E'$-valued function $u(x)$; by Lemma \ref{lem:horizontal-lift-linear-spec} the trivialized lift of the trivialized data is the trivialized lift, so near $x_0$ the field in question reads as $x \mapsto \mathcal{L}(B(x), A(x))\, u(x)$, where $\mathcal{L}(B,A)$ is the linear-algebraic horizontal lift of Definition \ref{def:horizontal-lift-linear}. This is smooth by Lemma \ref{lem:horizontal-lift-smooth-in-data} together with the smoothness of $u$, since evaluation of a linear map on a vector is a continuous bilinear operation.
\end{proof}

\begin{lemma}[Smoothness of the Horizontal Projection of a Vector Field]
\label{lem:horizontal-projection-field-smooth}
\lean{LeeLib.Ch02.contMDiff_horizontalProjField}
\leanok
\uses{lem:pushforward-along-map-smooth,lem:horizontal-lift-along-smooth,def:horizontal-space}
Let $\pi : \tilde{M} \to M$ be a smooth submersion, let $\tilde{g}$ be a Riemannian metric on $\tilde{M}$, and let $W$ be a smooth vector field on $\tilde{M}$. Then the horizontal part $W^H$, defined pointwise by $W^H_x = L_x(d\pi_x(W_x))$, is a smooth vector field on $\tilde{M}$.
\end{lemma}

\begin{proof}
\leanok
By Lemma \ref{lem:pushforward-along-map-smooth} the family $x \mapsto d\pi_x(W_x)$ is a smooth section along $\pi$, and by Lemma \ref{lem:horizontal-lift-along-smooth} the horizontal lift of any such section is a smooth vector field. Composing the two gives the claim.
\end{proof}

\begin{definition}[Smooth submersion]
\label{def:submersion}
\lean{LeeLib.Ch02.IsSubmersion}
\leanok
A smooth map $\pi : \tilde{M} \to M$ between smooth manifolds is a \emph{submersion} if its differential $d\pi_x : T_x\tilde{M} \to T_{\pi(x)}M$ is surjective at every $x \in \tilde{M}$.
\end{definition}

\begin{definition}[The horizontal lift of a surjection]
\label{def:horizontal-lift-linear}
\lean{LeeLib.Ch02.horizontalLift}
\leanok
Let $B$ be a positive definite symmetric bilinear form on a finite-dimensional real vector space $E$, let $E'$ be a finite-dimensional real vector space, and let $A : E \to E'$ be surjective. Writing $A^t : (E')^* \to E^*$ for the transpose and identifying $B$ with the isomorphism $E \to E^*$, $v \mapsto B(v, \cdot)$, the \emph{horizontal lift} of $A$ relative to $B$ is
\[
L \;=\; B^{-1}A^t\bigl(AB^{-1}A^t\bigr)^{-1} \;:\; E' \to E ,
\]
the $B$-weighted Moore--Penrose right inverse of $A$. Lemma \ref{lem:horizontal-lift-linear-spec} shows that the two inverses exist and that $L$ is characterized by the properties appearing there.
\end{definition}

\begin{lemma}[The horizontal lift is the unique $B$-orthogonal right inverse]
\label{lem:horizontal-lift-linear-spec}
\lean{LeeLib.Ch02.horizontalLift_unique}
\leanok
\uses{def:horizontal-lift-linear}
With $B$, $E$, $E'$ and $A$ as in Definition \ref{def:horizontal-lift-linear}, there is exactly one linear map $L : E' \to E$ such that
\begin{enumerate}
\item[(i)] $A(Lu) = u$ for every $u \in E'$, and
\item[(ii)] $B(Lu, w) = 0$ for every $u \in E'$ and every $w \in \ker A$,
\end{enumerate}
namely the map of Definition \ref{def:horizontal-lift-linear}.
\end{lemma}

\begin{proof}
\leanok
\uses{def:horizontal-lift-linear}
Since $B$ is positive definite it is nondegenerate, so $v \mapsto B(v,\cdot)$ is injective, hence an isomorphism $E \to E^*$ because $\dim E^* = \dim E$. Put $S = B^{-1}A^t : (E')^* \to E$, so that $S$ is characterized by
\begin{equation}
B(S\xi, w) = \xi(Aw) \qquad \text{for all } \xi \in (E')^*,\ w \in E. \tag{$*$}
\end{equation}

We claim $AS : (E')^* \to E'$ is an isomorphism. It is injective: if $A(S\xi) = 0$ then, taking $w = S\xi$ in $(*)$, $B(S\xi, S\xi) = \xi(A S\xi) = 0$, so $S\xi = 0$ by positive definiteness; then $(*)$ gives $\xi(Aw) = 0$ for every $w \in E$, and $A$ is surjective, so $\xi = 0$. Since $\dim (E')^* = \dim E'$, injectivity gives bijectivity.

Set $L = S(AS)^{-1}$. Then $A(Lu) = (AS)(AS)^{-1}u = u$, which is (i). For (ii), write $\xi = (AS)^{-1}u$, so $Lu = S\xi$; if $w \in \ker A$ then $(*)$ gives $B(Lu, w) = \xi(Aw) = 0$.

For uniqueness, suppose $L$ and $L'$ both satisfy (i) and (ii), and fix $u \in E'$. Put $d = Lu - L'u$. By (i), $Ad = u - u = 0$, so $d \in \ker A$; by (ii) applied to each of $L$ and $L'$ with $w = d$, we get $B(Lu, d) = B(L'u, d) = 0$, hence $B(d,d) = B(Lu,d) - B(L'u,d) = 0$. Positive definiteness forces $d = 0$, so $Lu = L'u$.
\end{proof}

\begin{lemma}[The horizontal lift depends smoothly on the data]
\label{lem:horizontal-lift-smooth-in-data}
\lean{LeeLib.Ch02.contMDiffAt_horizontalLift}
\leanok
\uses{def:horizontal-lift-linear}
Let $E$, $E'$ be finite-dimensional real vector spaces, let $N$ be a smooth manifold, and let $x \mapsto B_x$ and $x \mapsto A_x$ be smooth maps from $N$ into the bilinear forms on $E$ and into $\operatorname{Hom}(E,E')$ respectively. If $B_{x_0}$ is positive definite and $A_{x_0}$ is surjective, then $x \mapsto L_x$, the horizontal lift of $A_x$ relative to $B_x$, is smooth at $x_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:horizontal-lift-linear,lem:horizontal-lift-linear-spec}
By Definition \ref{def:horizontal-lift-linear}, $L_x = B_x^{-1}A_x^t(A_xB_x^{-1}A_x^t)^{-1}$. The map $A \mapsto A^t$ is linear, hence smooth. Inversion is smooth at every invertible operator, and composition of operators is bilinear, hence smooth. By Lemma \ref{lem:horizontal-lift-linear-spec} both $B_{x_0}$ and $A_{x_0}B_{x_0}^{-1}A_{x_0}^t$ are invertible, so each inversion in the formula is applied at an invertible operator at $x = x_0$, and the composite is smooth at $x_0$.
\end{proof}

\begin{definition}[The vertical space]
\label{def:vertical-space}
\lean{LeeLib.Ch02.verticalSpace}
\leanok
\uses{def:submersion}
Let $\pi : \tilde{M} \to M$ be a smooth submersion. The \emph{vertical space} at $x \in \tilde{M}$ is $V_x = \ker d\pi_x \subseteq T_x\tilde{M}$. It refers to no metric.
\end{definition}

\begin{definition}[The horizontal space]
\label{def:horizontal-space}
\lean{LeeLib.Ch02.horizontalSpace}
\leanok
\uses{def:vertical-space,def:riemannian-metric}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. The \emph{horizontal space} at $x \in \tilde{M}$ is
\[
H_x \;=\; \{\, v \in T_x\tilde{M} \;:\; \tilde{g}_x(v,w) = 0 \text{ for every } w \in V_x \,\} ,
\]
the $\tilde{g}_x$-orthogonal complement of $V_x$. Unlike $V_x$, it depends on $\tilde{g}$.
\end{definition}

\begin{definition}[The horizontal lift of a tangent vector]
\label{def:horizontal-lift-tangent}
\lean{LeeLib.Ch02.horizontalLiftAt}
\leanok
\uses{def:horizontal-lift-linear,def:horizontal-space,def:submersion}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. For $x \in \tilde{M}$, the \emph{horizontal lift at $x$} is the map $L_x : T_{\pi(x)}M \to T_x\tilde{M}$ given by Definition \ref{def:horizontal-lift-linear} applied to $B = \tilde{g}_x$ and $A = d\pi_x$. By Lemma \ref{lem:horizontal-lift-linear-spec} it is the unique linear map with $d\pi_x(L_x u) = u$ and $L_x u \in H_x$ for all $u$.
\end{definition}

\begin{lemma}[The tangent space splits]
\label{lem:tangent-space-horizontal-vertical-splitting}
\lean{LeeLib.Ch02.isCompl_horizontalSpace_verticalSpace}
\leanok
\uses{def:vertical-space,def:horizontal-space,def:horizontal-lift-tangent}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. For every $x \in \tilde{M}$,
\[
T_x\tilde{M} = H_x \oplus V_x .
\]
\end{lemma}

\begin{proof}
\leanok
\uses{def:vertical-space,def:horizontal-space,def:horizontal-lift-tangent,lem:horizontal-lift-linear-spec}
The two subspaces intersect trivially: if $v \in H_x \cap V_x$ then, taking $w = v$ in the definition of $H_x$, we get $\tilde{g}_x(v,v) = 0$, hence $v = 0$ because $\tilde{g}_x$ is positive definite.

They span: given $v \in T_x\tilde{M}$, set $v^H = L_x(d\pi_x v)$ and $v^V = v - v^H$, where $L_x$ is the horizontal lift of Definition \ref{def:horizontal-lift-tangent}. Then $v^H \in H_x$ by Lemma \ref{lem:horizontal-lift-linear-spec}(ii), and $d\pi_x(v^V) = d\pi_x v - d\pi_x(L_x(d\pi_x v)) = d\pi_x v - d\pi_x v = 0$ by Lemma \ref{lem:horizontal-lift-linear-spec}(i), so $v^V \in V_x$. Hence $v = v^H + v^V \in H_x + V_x$.
\end{proof}

\begin{proposition}[Existence and uniqueness of horizontal lifts]
\label{prop:horizontal-lift-of-vector-field}
\lean{LeeLib.Ch02.exists_unique_horizontalLift}
\leanok
\uses{def:horizontal-lift-tangent,def:horizontal-space,def:submersion,def:riemannian-metric,lem:horizontal-lift-linear-spec,lem:horizontal-lift-smooth-in-data}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. Every $X \in \mathfrak{X}(M)$ has a horizontal lift $\tilde{X} \in \mathfrak{X}(\tilde{M})$: a smooth vector field with $\tilde{X}_x \in H_x$ and $d\pi_x(\tilde{X}_x) = X_{\pi(x)}$ for every $x \in \tilde{M}$. It is the only vector field with those two properties. This is part (b) of Proposition \ref{prop:horizontal-vector-field-properties}.
\end{proposition}

\begin{proof}
\leanok
\uses{def:horizontal-lift-tangent,lem:horizontal-lift-linear-spec,lem:horizontal-lift-smooth-in-data,def:horizontal-space}
Define $\tilde{X}_x = L_x(X_{\pi(x)})$, with $L_x$ as in Definition \ref{def:horizontal-lift-tangent}. Then $\tilde{X}_x \in H_x$ and $d\pi_x(\tilde{X}_x) = X_{\pi(x)}$ by Lemma \ref{lem:horizontal-lift-linear-spec}, so only smoothness and uniqueness remain.

Smoothness is local, so fix $x_0 \in \tilde{M}$ and work in local trivializations of $T\tilde{M}$ near $x_0$ and of $TM$ near $\pi(x_0)$; write $\theta_x : T_x\tilde{M} \to E$ and $\iota_y : T_yM \to E'$ for the resulting isomorphisms onto the model spaces. Set
\[
B_x(a,b) = \tilde{g}_x(\theta_x^{-1}a, \theta_x^{-1}b), \qquad A_x = \iota_{\pi(x)} \circ d\pi_x \circ \theta_x^{-1} .
\]
Then $x \mapsto B_x$ is smooth because $\tilde{g}$ is a smooth $2$-tensor field, and $x \mapsto A_x$ is smooth because $\pi$ is smooth and the differential of a smooth map is smooth in coordinates. Each $B_x$ is positive definite and each $A_x$ is surjective, both properties being carried across the isomorphisms $\theta_x$, $\iota_{\pi(x)}$ from $\tilde{g}_x$ and $d\pi_x$. Transporting the two defining properties of Lemma \ref{lem:horizontal-lift-linear-spec} through $\theta_x$ and $\iota_{\pi(x)}$ shows that the horizontal lift of $A_x$ relative to $B_x$ is $\theta_x \circ L_x \circ \iota_{\pi(x)}^{-1}$; uniqueness in that lemma is what makes this an identity rather than a definition. Hence
\[
\theta_x(\tilde{X}_x) \;=\; \bigl(\theta_x \circ L_x \circ \iota_{\pi(x)}^{-1}\bigr)\bigl(\iota_{\pi(x)}(X_{\pi(x)})\bigr),
\]
which is smooth at $x_0$ by Lemma \ref{lem:horizontal-lift-smooth-in-data} together with smoothness of $x \mapsto \iota_{\pi(x)}(X_{\pi(x)})$, the latter because $X$ is smooth and $\pi$ is smooth. As $\theta$ is the trivialization, $\tilde{X}$ is a smooth section near $x_0$.

For uniqueness, let $Z$ be any vector field with $Z_x \in H_x$ and $d\pi_x(Z_x) = X_{\pi(x)}$ for all $x$. Then $Z_x$ and $\tilde{X}_x$ are both horizontal with the same image under $d\pi_x$, so they agree by the uniqueness in Lemma \ref{lem:horizontal-lift-linear-spec}. Hence $Z = \tilde{X}$.
\end{proof}

\begin{proposition}[Every Horizontal Vector is the Value of a Lift]
\label{prop:horizontal-lift-realizes-horizontal-vector}
\dcref{ch2:2.25}
\lean{LeeLib.Ch02.exists_horizontalLift_eq}
\leanok
\uses{def:horizontal-space,def:horizontal-lift-tangent,prop:horizontal-lift-of-vector-field,lem:vector-field-extension-point,lem:horizontal-lift-linear-spec}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. For every $x \in \tilde{M}$ and every $v \in H_x$, there is a vector field $X \in \mathfrak{X}(M)$ whose horizontal lift $\tilde{X}$ satisfies $\tilde{X}_x = v$.
\end{proposition}

\begin{proof}
\leanok
Push $v$ down to the tangent vector $d\pi_x(v) \in T_{\pi(x)}M$, and use Lemma \ref{lem:vector-field-extension-point} to choose a smooth vector field $X \in \mathfrak{X}(M)$ with $X_{\pi(x)} = d\pi_x(v)$. Let $\tilde{X}$ be its horizontal lift, which exists and is smooth by Proposition \ref{prop:horizontal-lift-of-vector-field}. At the point $x$,
\[
\tilde{X}_x \;=\; L_x\bigl(X_{\pi(x)}\bigr) \;=\; L_x\bigl(d\pi_x(v)\bigr) \;=\; v,
\]
the last equality by the uniqueness clause of Lemma \ref{lem:horizontal-lift-linear-spec}: $L_x$ is a right inverse of $d\pi_x$ whose image is $H_x$, so it inverts $d\pi_x$ on horizontal vectors. This is the only place the hypothesis $v \in H_x$ is used, and it cannot be dropped: every horizontal lift is horizontal at every point, so a vector outside $H_x$ is not the value of any lift.
\end{proof}

The fact that every horizontal vector at a point of $\tilde{M}$ can be extended to a horizontal lift on all of $\tilde{M}$ (part (c) of Proposition \ref{prop:horizontal-vector-field-properties}) is highly useful for computations. It is important to be aware, though, that not every horizontal vector field on $\tilde{M}$ is a horizontal lift, as the next exercise shows.

\begin{exercise}
\label{exer:horizontal-vector-field-not-lift}
\dcref{ch2:2.26}
Let $\pi: \mathbb{R}^2 \to \mathbb{R}$ be the projection map $\pi(x,y) = x$, and let $\tilde{W}$ be the smooth vector field $y\partial_x$ on $\mathbb{R}^2$. Show that $\tilde{W}$ is horizontal, but there is no vector field on $\mathbb{R}$ whose horizontal lift is equal to $\tilde{W}$.
\end{exercise}

Now we can identify some quotients of Riemannian manifolds that inherit metrics of their own. Let us begin by describing what such a metric should look like.

Suppose $(\tilde{M},\tilde{g})$ and $(M,g)$ are Riemannian manifolds, and $\pi: \tilde{M} \to M$ is a smooth submersion. Then $\pi$ is said to be a \textbf{Riemannian submersion} if for each $x \in \tilde{M}$, the differential $d\pi_x$ restricts to a linear isometry from $H_x$ onto $T_{\pi(x)}M$. In other words, $\tilde{g}_x(v,w) = g_{\pi(x)}(d\pi_x(v), d\pi_x(w))$ whenever $v,w \in H_x$.

\begin{example}[Riemannian Submersions]
\label{ex:riemannian-submersions}
\dcref{ch2:2.27}
\begin{enumerate}
\item[(a)] The projection $\pi: \mathbb{R}^{n+k} \to \mathbb{R}^n$ onto the first $n$ coordinates is a Riemannian submersion if $\mathbb{R}^{n+k}$ and $\mathbb{R}^n$ are both endowed with their Euclidean metrics.
\item[(b)] If $M$ and $N$ are Riemannian manifolds and $M \times N$ is endowed with the product metric, then both projections $\pi_M: M \times N \to M$ and $\pi_N: M \times N \to N$ are Riemannian submersions.
\item[(c)] If $M \times_f N$ is a warped product manifold, then the projection $\pi_M : M \times_f N \to M$ is a Riemannian submersion, but $\pi_N$ typically is not.
\end{enumerate}
\end{example}

Given a Riemannian manifold $(\tilde{M}, \tilde{g})$ and a surjective submersion $\pi: \tilde{M} \to M$, it is almost never the case that there is a metric on $M$ that makes $\pi$ into a Riemannian submersion. It is not hard to see why: for this to be the case, whenever $p, q \in M$ are two points in the same fiber $\pi^{-1}(y)$, the linear maps $(d\pi_p|_{H_p})^{-1}: T_y M \to H_p$ both have to pull $\tilde{g}$ back to the same inner product on $T_y M$.

There is, however, an important special case in which there is such a metric. Suppose $\pi: \tilde{M} \to M$ is a smooth surjective submersion, and $G$ is a group acting on $\tilde{M}$. (See Appendix C for a review of the basic definitions and terminology regarding group actions on manifolds.) We say that the action is \emph{vertical} if every element $\varphi \in G$ takes each fiber to itself, meaning that $\pi(\varphi \cdot p) = \pi(p)$ for all $p \in M$. The action is \emph{transitive on fibers} if for each $p, q \in \tilde{M}$ such that $\pi(p) = \pi(q)$, there exists $\varphi \in G$ such that $\varphi \cdot p = q$.

If in addition $\tilde{M}$ is endowed with a Riemannian metric, the action is said to be an \emph{isometric action} or an \emph{action by isometries}, and the metric is said to be \emph{invariant under $G$}, if the map $x \mapsto \varphi \cdot x$ is an isometry for each $\varphi \in G$. In that case, provided the action is effective (so that different elements of $G$ define different isometries of $\tilde{M}$), we can identify $G$ with a subgroup of $\text{Iso}(\tilde{M}, \tilde{g})$. Since an isometry is, in particular, a diffeomorphism, every isometric action is an action by diffeomorphisms.

\begin{definition}[Riemannian submersion]
\label{def:riemannian-submersion}
\lean{LeeLib.Ch02.IsRiemannianSubmersion}
\leanok
\uses{def:submersion,def:horizontal-space,def:riemannian-metric}
Let $(\tilde{M},\tilde{g})$ and $(M,g)$ be Riemannian manifolds. A smooth submersion $\pi : \tilde{M} \to M$ is a \emph{Riemannian submersion} if for every $x \in \tilde{M}$ and all $v,w \in H_x$,
\[
g_{\pi(x)}(d\pi_x v, d\pi_x w) = \tilde{g}_x(v,w),
\]
that is, $d\pi_x$ restricts to a linear isometry of $H_x$ onto $T_{\pi(x)}M$. Surjectivity of $d\pi_x|_{H_x}$ is automatic from Lemma \ref{lem:tangent-space-horizontal-vertical-splitting}.
\end{definition}

\begin{definition}[The inner product induced on the base]
\label{def:submersion-inner}
\lean{LeeLib.Ch02.submersionInner}
\leanok
\uses{def:horizontal-lift-tangent,def:riemannian-metric}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and $\tilde{g}$ a Riemannian metric on $\tilde{M}$. For $x \in \tilde{M}$, the \emph{induced form at $x$} is the bilinear form on $T_{\pi(x)}M$
\[
\beta_x(u,v) \;=\; \tilde{g}_x(L_x u, L_x v),
\]
where $L_x$ is the horizontal lift of Definition \ref{def:horizontal-lift-tangent}.
\end{definition}

\begin{lemma}[A Riemannian submersion's base metric is the induced form]
\label{lem:riemannian-submersion-inner-eq}
\lean{LeeLib.Ch02.IsRiemannianSubmersion.inner_eq}
\leanok
\uses{def:riemannian-submersion,def:submersion-inner}
If $\pi : (\tilde{M},\tilde{g}) \to (M,g)$ is a Riemannian submersion then $g_{\pi(x)} = \beta_x$ for every $x \in \tilde{M}$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-submersion,def:submersion-inner,lem:horizontal-lift-linear-spec}
Let $u,v \in T_{\pi(x)}M$. The vectors $L_x u$ and $L_x v$ are horizontal, so the Riemannian submersion condition gives $g_{\pi(x)}(d\pi_x L_x u, d\pi_x L_x v) = \tilde{g}_x(L_x u, L_x v) = \beta_x(u,v)$. But $d\pi_x L_x u = u$ and $d\pi_x L_x v = v$, so the left side is $g_{\pi(x)}(u,v)$.
\end{proof}

\begin{proposition}[Uniqueness of the submersion metric]
\label{prop:riemannian-submersion-metric-unique}
\lean{LeeLib.Ch02.isRiemannianSubmersion_unique}
\leanok
\uses{lem:riemannian-submersion-inner-eq}
Let $\tilde{g}$ be a Riemannian metric on $\tilde{M}$ and let $\pi : \tilde{M} \to M$ be a surjective smooth submersion. There is \emph{at most one} Riemannian metric $g$ on $M$ making $\pi$ a Riemannian submersion. This is the uniqueness half of Theorem \ref{thm:group-action-riemannian-submersion}, and it needs no group action.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:riemannian-submersion-inner-eq}
A Riemannian metric is determined by its inner products. Let $y \in M$ and, by surjectivity, write $y = \pi(x)$. If $g_1$ and $g_2$ both make $\pi$ a Riemannian submersion then Lemma \ref{lem:riemannian-submersion-inner-eq} gives $(g_1)_y = \beta_x = (g_2)_y$.
\end{proof}

\noindent The existence half is where the group action enters, through the following three lemmas.

\begin{lemma}[The action's differential is invertible]
\label{lem:mfderiv-smul-inv-smul}
\lean{LeeLib.Ch02.IsIsometricAction.mfderiv_smul_inv_smul}
\leanok
\uses{def:riemannian-metric}
Let $G$ act on $\tilde{M}$ by smooth maps. For $a \in G$ and $x \in \tilde{M}$, the differentials of $a \cdot {-}$ at $x$ and of $a^{-1} \cdot {-}$ at $a \cdot x$ are mutually inverse.
\end{lemma}

\begin{proof}
\leanok
Apply the chain rule to $(a \cdot {-}) \circ (a^{-1} \cdot {-}) = \mathrm{Id}$ at the point $a \cdot x$, noting $a^{-1} \cdot (a \cdot x) = x$, and use that the differential of the identity is the identity.
\end{proof}

\begin{lemma}[A vertical action preserves the splitting]
\label{lem:horizontal-space-smul}
\lean{LeeLib.Ch02.horizontalSpace_smul}
\leanok
\uses{lem:mfderiv-smul-inv-smul,def:vertical-space,def:horizontal-space}
Let $\pi : \tilde{M} \to M$ be a smooth submersion and let $G$ act on $\tilde{M}$ by smooth maps preserving $\tilde{g}$, with $\pi(a \cdot x) = \pi(x)$ for all $a, x$. Then $d(a \cdot {-})_x$ carries $V_x$ onto $V_{a \cdot x}$, and carries $H_x$ into $H_{a \cdot x}$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:mfderiv-smul-inv-smul,def:vertical-space,def:horizontal-space}
Differentiating $\pi \circ (a \cdot {-}) = \pi$ gives $d\pi_{a \cdot x} \circ d(a \cdot {-})_x = d\pi_x$. Hence if $d\pi_x z = 0$ then $d\pi_{a\cdot x}(d(a\cdot{-})_x z) = 0$, so $d(a \cdot {-})_x$ maps $V_x$ into $V_{a \cdot x}$. Conversely, given $w \in V_{a \cdot x}$, put $z = d(a^{-1} \cdot {-})_{a \cdot x} w$; by Lemma \ref{lem:mfderiv-smul-inv-smul}, $d(a\cdot{-})_x z = w$, and then $d\pi_x z = d\pi_{a \cdot x} w = 0$, so $z \in V_x$. Thus $V_{a \cdot x}$ is the image of $V_x$.

Now let $v \in H_x$ and $w \in V_{a \cdot x}$. Write $w = d(a \cdot {-})_x z$ with $z \in V_x$, as just shown. Since the action preserves $\tilde{g}$,
\[
\tilde{g}_{a \cdot x}(d(a\cdot{-})_x v, w) = \tilde{g}_{a \cdot x}(d(a\cdot{-})_x v, d(a\cdot{-})_x z) = \tilde{g}_x(v,z) = 0,
\]
the last step because $v$ is horizontal and $z$ is vertical. So $d(a\cdot{-})_x v \in H_{a \cdot x}$.
\end{proof}

\begin{lemma}[The induced form is constant on fibers]
\label{lem:submersion-inner-smul}
\lean{LeeLib.Ch02.submersionInner_smul}
\leanok
\uses{lem:horizontal-space-smul,def:submersion-inner,lem:horizontal-lift-linear-spec}
With hypotheses as in Lemma \ref{lem:horizontal-space-smul}, $\beta_{a \cdot x} = \beta_x$ for every $a \in G$ and $x \in \tilde{M}$ (an identity of bilinear forms on $T_{\pi(x)}M = T_{\pi(a \cdot x)}M$).
\end{lemma}

\begin{proof}
\leanok
\uses{lem:horizontal-space-smul,def:submersion-inner,lem:horizontal-lift-linear-spec}
Write $\Phi = d(a \cdot {-})_x$. We claim $L_{a \cdot x} = \Phi \circ L_x$. By the uniqueness in Lemma \ref{lem:horizontal-lift-linear-spec} it suffices to check that $\Phi \circ L_x$ is a horizontal right inverse of $d\pi_{a \cdot x}$. It is a right inverse: $d\pi_{a \cdot x}(\Phi(L_x u)) = d\pi_x(L_x u) = u$, using $d\pi_{a \cdot x} \circ \Phi = d\pi_x$ from the proof of Lemma \ref{lem:horizontal-space-smul}. It is horizontal: $L_x u \in H_x$, so $\Phi(L_x u) \in H_{a \cdot x}$ by Lemma \ref{lem:horizontal-space-smul}.

Therefore, for $u,v \in T_{\pi(x)}M$,
\[
\beta_{a \cdot x}(u,v) = \tilde{g}_{a \cdot x}(\Phi(L_x u), \Phi(L_x v)) = \tilde{g}_x(L_x u, L_x v) = \beta_x(u,v),
\]
the middle step because the action preserves $\tilde{g}$.
\end{proof}

\begin{lemma}[The induced form depends only on the fiber]
\label{lem:submersion-inner-congr}
\lean{LeeLib.Ch02.submersionInner_congr}
\leanok
\uses{lem:submersion-inner-smul,def:submersion-inner,lem:horizontal-space-smul}
With hypotheses as in Lemma \ref{lem:horizontal-space-smul}, suppose in addition that $G$ acts transitively on the fibers of $\pi$. If $\pi(x) = \pi(y)$ then $\beta_x = \beta_y$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:submersion-inner-smul,def:submersion-inner}
By transitivity on fibers there is $a \in G$ with $y = a \cdot x$, and then $\beta_y = \beta_{a \cdot x} = \beta_x$ by Lemma \ref{lem:submersion-inner-smul}.
\end{proof}

\begin{lemma}[The horizontal lift is smooth in tangent coordinates]
\label{lem:horizontal-lift-in-coordinates-smooth}
\lean{LeeLib.Ch02.contMDiffAt_horizontalLiftAt_inTangentCoordinates}
\leanok
\uses{lem:horizontal-lift-smooth-in-data,def:horizontal-lift-tangent,lem:horizontal-lift-linear-spec}
Let $\pi : \tilde{M} \to M$ be a smooth submersion of a Riemannian manifold $(\tilde{M},\tilde{g})$ and let $x_0 \in \tilde{M}$. The family of linear maps $x \mapsto L_x : T_{\pi(x)}M \to T_x\tilde{M}$, read in tangent coordinates around $x_0$ and around $\pi(x_0)$, is smooth at $x_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:horizontal-lift-smooth-in-data,def:horizontal-lift-tangent,lem:horizontal-lift-linear-spec}
Read in the trivializations $\theta$ of $T\tilde{M}$ at $x_0$ and $\iota$ of $TM$ at $\pi(x_0)$, the data of $L_x$ becomes the pair $(B(x), A(x))$, where $B(x)$ is $\tilde{g}_x$ and $A(x)$ is $d\pi_x$ so read; $B$ is smooth at $x_0$ because $\tilde{g}$ is a smooth $2$-tensor field, and $A$ is smooth at $x_0$ because $\pi$ is smooth. Moreover $B(x_0)$ is positive definite and $A(x_0)$ is surjective, since $\theta$ and $\iota$ are isomorphisms and $\pi$ is a submersion. By the naturality of the lift under a pair of linear isomorphisms, the lift of the transported data is the transported lift, that is $\theta \circ L_x \circ \iota^{-1} = L(B(x), A(x))$, which is exactly $L_x$ read in tangent coordinates. Lemma \ref{lem:horizontal-lift-smooth-in-data} says the right-hand side is smooth at $x_0$.
\end{proof}

\begin{theorem}
\label{thm:group-action-riemannian-submersion}
\dcref{ch2:2.28}
\lean{LeeLib.Ch02.existsUnique_isRiemannianSubmersion_metric}
\leanok
\uses{REF:ch2-6,prop:riemannian-submersion-metric-unique,lem:submersion-inner-congr,prop:local-section-manifold,def:riemannian-submersion,lem:bilinear-comp-of-family-smooth,lem:horizontal-lift-in-coordinates-smooth}
Let $(\tilde{M}, \tilde{g})$ be a Riemannian manifold, let $\pi: \tilde{M} \to M$ be a surjective smooth submersion, and let $G$ be a group acting on $\tilde{M}$. If the action is isometric, vertical, and transitive on fibers, then there is a unique Riemannian metric on $M$ such that $\pi$ is a Riemannian submersion.
\end{theorem}

\begin{proof}
\leanok
\uses{prop:riemannian-submersion-metric-unique,lem:submersion-inner-congr,prop:local-section-manifold,def:submersion-inner,def:riemannian-submersion,lem:bilinear-comp-of-family-smooth,lem:horizontal-lift-in-coordinates-smooth}
Uniqueness is Proposition \ref{prop:riemannian-submersion-metric-unique}. For existence, define $g_y = \beta_x$ for any $x$ with $\pi(x) = y$; this is legitimate because if $\pi(x) = \pi(x')$ then $\beta_x = \beta_{x'}$ by Lemma \ref{lem:submersion-inner-congr} (whose hypotheses hold: the action is isometric, vertical, and transitive on fibers).

$g_y$ is symmetric because $\tilde{g}_x$ is. It is positive definite: if $u \neq 0$ then $L_x u \neq 0$, since $d\pi_x(L_x u) = u \neq 0$, and hence $\beta_x(u,u) = \tilde{g}_x(L_x u, L_x u) > 0$.

For smoothness, fix $y_0 \in M$ and choose $x_0$ with $\pi(x_0) = y_0$. By Proposition \ref{prop:local-section-manifold} there is a smooth local section $\sigma$ of $\pi$ near $y_0$ with $\sigma(y_0) = x_0$. For $y$ near $y_0$ we have $\pi(\sigma(y)) = y$, so by well-definedness $g_y = \beta_{\sigma(y)}$, that is,
\[
g_y(u,v) = \tilde{g}_{\sigma(y)}(L_{\sigma(y)} u, L_{\sigma(y)} v).
\]
The right-hand side is the transport of the smooth field $\tilde{g}$ by the family of linear maps $y \mapsto L_{\sigma(y)} : T_yM \to T_{\sigma(y)}\tilde{M}$, in the sense of Definition \ref{def:bilinear-comp-of-family}, so Lemma \ref{lem:bilinear-comp-of-family-smooth} applies once its two hypotheses are checked. The base map $\sigma$ is smooth at $y_0$; and the family $y \mapsto L_{\sigma(y)}$ is smooth at $y_0$ in tangent coordinates, being the composite of $\sigma$ with $x \mapsto L_x$, which is smooth in tangent coordinates by Lemma \ref{lem:horizontal-lift-in-coordinates-smooth}. (Here the reference points match: $\sigma(y_0) = x_0$ upstairs and $\pi(x_0) = y_0$ downstairs.) Hence $g$ is smooth at $y_0$, and $y_0$ was arbitrary, so $g$ is a smooth $2$-tensor field.

Note that this is the step for which Lemma \ref{lem:bilinear-comp-of-family-smooth} is stated for an arbitrary family of linear maps rather than for a family of differentials: $L_{\sigma(y)}$ is not $d\sigma_y$, and is not the differential of any map --- $\sigma$ is a section of $\pi$, but not a horizontal one.

Finally $g$ makes $\pi$ a Riemannian submersion: for $v,w \in H_x$, $L_x(d\pi_x v) = v$ and $L_x(d\pi_x w) = w$, so $g_{\pi(x)}(d\pi_x v, d\pi_x w) = \beta_x(d\pi_x v, d\pi_x w) = \tilde{g}_x(v,w)$.
\end{proof}

The next corollary describes one important situation to which the preceding theorem applies.

\begin{corollary}
\label{cor:quotient-manifold-lie-group-riemannian-metric}
\dcref{ch2:2.29}
\uses{thm:quotient-manifold-theorem,thm:group-action-riemannian-submersion}
Suppose $(\tilde{M}, \tilde{g})$ is a Riemannian manifold, and $G$ is a Lie group acting smoothly, freely, properly, and isometrically on $\tilde{M}$. Then the orbit space $M = \tilde{M}/G$ has a unique smooth manifold structure and Riemannian metric such that $\pi$ is a Riemannian submersion.
\end{corollary}

\begin{proof}
Under the given hypotheses, the quotient manifold theorem (Thm. C.17) shows that $M$ has a unique smooth manifold structure such that the quotient map $\pi: \tilde{M} \to M$ is a smooth submersion. It follows easily from the definitions in that case that the given action of $G$ on $\tilde{M}$ is vertical and transitive on fibers. Since the action is also isometric, Theorem 2.28 shows that $M$ inherits a unique Riemannian metric making $\pi$ into a Riemannian submersion.
\end{proof}

Here is an important example of a Riemannian metric defined in this way. A larger class of such metrics is described in Problem 2-7.

\begin{example}[The Fubini--Study Metric]
\label{ex:fubini-study-metric}
\dcref{ch2:2.30}
\uses{ex:complex-projective-spaces,thm:local-section-theorem,thm:group-action-riemannian-submersion}

Let $n$ be a positive integer, and consider the complex projective space $\mathbb{CP}^n$ defined in Example C.19. That example shows that the map $\pi : \mathbb{C}^{n+1} \setminus \{0\} \to \mathbb{CP}^n$ sending each point in $\mathbb{C}^{n+1} \setminus \{0\}$ to its span is a surjective smooth submersion. Identifying $\mathbb{C}^{n+1}$ with $\mathbb{R}^{2n+2}$ endowed with its Euclidean metric, we can view the unit sphere $S^{2n+1}$ with its round metric $\bar{g}$ as an embedded Riemannian submanifold of $\mathbb{C}^{n+1} \setminus \{0\}$. Let $p: S^{2n+1} \to \mathbb{CP}^n$ denote the restriction of the map $\pi$. Then $p$ is smooth, and it is surjective because every 1-dimensional complex subspace contains elements of unit norm. We need to show that it is a submersion. Let $z_0 \in S^{2n+1}$ and set $\zeta_0 = p(z_0) \in \mathbb{CP}^n$. Since $\pi$ is a smooth submersion, it has a smooth local section $\sigma : U \to \mathbb{C}^{n+1}$ defined on a neighborhood $U$ of $\zeta_0$ and satisfying $\sigma(\zeta_0) = z_0$ (Thm. A.17). Let $\nu: \mathbb{C}^{n+1} \setminus \{0\} \to S^{2n+1}$ be the radial projection onto the sphere:
\[
\nu(z) = \frac{z}{|z|}.
\]

Since dividing an element of $\mathbb{C}^{n+1}$ by a nonzero scalar does not change its span, it follows that $p \circ \nu = \pi$. Therefore, if we set $\bar{\sigma} = \nu \circ \sigma$, we have $p \circ \bar{\sigma} = p \circ \nu \circ \sigma = \pi \circ \sigma = \text{Id}_U$, so $\bar{\sigma}$ is a local section of $p$. By Theorem A.17, this shows that $p$ is a submersion.

Define an action of $S^1$ on $S^{2n+1}$ by complex multiplication:
\[
\lambda \cdot (z^1,\ldots,z^{n+1}) = (\lambda z^1,\ldots,\lambda z^{n+1}),
\]
for $\lambda \in S^1$ (viewed as a complex number of norm 1) and $z = (z^1,\ldots,z^{n+1}) \in S^{2n+1}$. This is easily seen to be isometric, vertical, and transitive on fibers of $p$. By Theorem 2.28, therefore, there is a unique metric on $\mathbb{CP}^n$ such that the map $p: S^{2n+1} \to \mathbb{CP}^n$ is a Riemannian submersion. This metric is called the \textbf{Fubini--Study metric}; you will have a chance to study its geometric properties in Problems 3-19 and 8-13.
\end{example}

\section{Riemannian Coverings}

Another important special case of Riemannian submersions occurs in the context of covering maps. Suppose $(\tilde{M}, \tilde{g})$ and $(M, g)$ are Riemannian manifolds. A smooth covering map $\pi: \tilde{M} \to M$ is called a \textbf{Riemannian covering} if it is a local isometry.

\begin{definition}
\label{def:smooth-covering-map}
\lean{LeeLib.Ch02.IsSmoothCoveringMap}
\leanok
\uses{prop:elementary-covering-maps}
A smooth map $\pi: \tilde{M} \to M$ is a \emph{smooth covering map} if it is surjective, is a topological covering map, and is a local diffeomorphism.

By Proposition A.49(c) a topological covering map is a smooth covering map, in the sense of Appendix A, exactly when it is a local diffeomorphism, so this description agrees with the sheetwise one.
\end{definition}

\begin{definition}
\label{def:covering-automorphism-group}
\lean{LeeLib.Ch02.CoveringAut}
\leanok
\uses{def:smooth-covering-map}
Let $\pi: \tilde{M} \to M$ be a smooth covering map. A \emph{covering automorphism} of $\pi$ is a diffeomorphism $\varphi: \tilde{M} \to \tilde{M}$ satisfying $\pi \circ \varphi = \pi$. The covering automorphisms form a group $\operatorname{Aut}_\pi(\tilde{M})$ under composition, which acts on $\tilde{M}$ by evaluation.
\end{definition}

\begin{definition}
\label{def:normal-covering-fibre-transitive}
\lean{LeeLib.Ch02.IsNormalCovering}
\leanok
\uses{def:covering-automorphism-group,prop:normal-covering-transitive-action}
A smooth covering map $\pi: \tilde{M} \to M$ is a \emph{normal covering} if $\operatorname{Aut}_\pi(\tilde{M})$ acts transitively on each fiber of $\pi$.

By Proposition C.21 this holds if and only if the image of $\pi_*: \pi_1(\tilde{M}, \tilde{x}) \to \pi_1(M, \pi(\tilde{x}))$ is a normal subgroup of $\pi_1(M, \pi(\tilde{x}))$, which is the condition the name refers to.
\end{definition}

\begin{definition}
\label{def:covering-automorphism-invariant-metric}
\lean{LeeLib.Ch02.IsCoveringAutInvariant}
\leanok
\uses{def:covering-automorphism-group,def:metric-preserving}
Let $\pi: \tilde{M} \to M$ be a smooth covering map. A metric $\tilde{g}$ on $\tilde{M}$ is \emph{invariant under all covering automorphisms} if $\varphi^*\tilde{g} = \tilde{g}$ for every $\varphi \in \operatorname{Aut}_\pi(\tilde{M})$.
\end{definition}

\begin{definition}
\label{def:riemannian-covering}
\lean{LeeLib.Ch02.IsRiemannianCovering}
\leanok
\uses{def:smooth-covering-map,def:metric-preserving}
Let $(\tilde{M}, \tilde{g})$ and $(M, g)$ be Riemannian manifolds. A smooth covering map $\pi: \tilde{M} \to M$ is a \emph{Riemannian covering} if $\pi^*g = \tilde{g}$.
\end{definition}

Lee phrases the definition as ``a smooth covering map that is a local isometry''; the two agree, because a covering map is already a local diffeomorphism.

\begin{lemma}[A Riemannian Covering Is a Local Isometry]
\label{lem:riemannian-covering-local-isometry}
\lean{LeeLib.Ch02.IsRiemannianCovering.isLocalIsometry}
\leanok
\uses{def:riemannian-covering,def:local-isometry,def:smooth-covering-map}
Every Riemannian covering is a local isometry.
\end{lemma}

\begin{proof}
\leanok
\uses{def:local-isometry,def:smooth-covering-map}
A smooth covering map is a local diffeomorphism, and a Riemannian covering satisfies $\pi^*g = \tilde{g}$ by definition; a local diffeomorphism that preserves the metric is a local isometry. No hypothesis on dimensions is needed: unlike in Exercise \ref{exer:characterize-local-isometry}, the local diffeomorphism is given rather than deduced, so the inverse function theorem does not enter.
\end{proof}

\begin{proposition}
\label{prop:equidimensional-riemannian-submersion-local-isometry}
\dcref{ch2:2.31}
\lean{LeeLib.Ch02.isRiemannianSubmersion_iff_isMetricPreserving}
\leanok
\uses{def:riemannian-submersion,def:metric-preserving,def:submersion}
Let $\pi: \tilde{M} \to M$ be a smooth submersion whose differential $d\pi_x$ is injective at every $x \in \tilde{M}$, and let $\tilde{g}$ and $g$ be metrics on $\tilde{M}$ and $M$. Then $\pi$ is a Riemannian submersion if and only if $\pi^*g = \tilde{g}$.
\end{proposition}

\begin{proof}
\leanok
Suppose first that $\pi$ is a Riemannian submersion, and fix $x \in \tilde{M}$. The vertical space $V_x = \ker d\pi_x$ is trivial because $d\pi_x$ is injective, so the horizontal space $H_x$, consisting of the vectors $g$-orthogonal to every vertical vector, is all of $T_x\tilde{M}$. The defining condition of a Riemannian submersion therefore applies to every pair $v, w \in T_x\tilde{M}$ and reads $g_{\pi(x)}(d\pi_x v, d\pi_x w) = \tilde{g}_x(v, w)$, which is $(\pi^*g)_x = \tilde{g}_x$.

Conversely, suppose $\pi^*g = \tilde{g}$. Then $g_{\pi(x)}(d\pi_x v, d\pi_x w) = \tilde{g}_x(v, w)$ for all $v, w \in T_x\tilde{M}$, and in particular for horizontal $v$ and $w$; since $\pi$ is a submersion by hypothesis, $\pi$ is a Riemannian submersion.
\end{proof}

\begin{proposition}
\label{prop:riemannian-covering-metric}
\dcref{ch2:2.31}
\lean{LeeLib.Ch02.existsUnique_isRiemannianCovering_metric}
\leanok
\uses{def:smooth-covering-map,def:normal-covering-fibre-transitive,def:covering-automorphism-invariant-metric,def:riemannian-covering,prop:elementary-covering-maps,prop:normal-covering-transitive-action,thm:group-action-riemannian-submersion,prop:equidimensional-riemannian-submersion-local-isometry}

Suppose $\pi: \tilde{M} \to M$ is a smooth normal covering map, and $\tilde{g}$ is any metric on $\tilde{M}$ that is invariant under all covering automorphisms. Then there is a unique metric $g$ on $M$ such that $\pi$ is a Riemannian covering.
\end{proposition}

\begin{proof}
\leanok
Proposition A.49 shows that $\pi$ is a surjective smooth submersion. The automorphism group acts vertically by definition, and Proposition C.21 shows that it acts transitively on fibers when the covering is normal. It then follows from Theorem 2.28 that there is a unique metric $g$ on $M$ such that $\pi$ is a Riemannian submersion. Since $\pi$ is a local diffeomorphism, each $d\pi_x$ is injective, so by Proposition \ref{prop:equidimensional-riemannian-submersion-local-isometry} the metrics making $\pi$ a Riemannian submersion are exactly those making $\pi$ a Riemannian covering. Hence $g$ is the unique metric on $M$ for which $\pi$ is a Riemannian covering.
\end{proof}

\begin{proposition}
\label{prop:riemannian-quotient-metric}
\dcref{ch2:2.32}
\uses{prop:quotient-orbit-space-covering-map,prop:riemannian-covering-metric}
Suppose $(\widetilde{M}, \tilde{g})$ is a Riemannian manifold, and $\Gamma$ is a discrete Lie group acting smoothly, freely, properly, and isometrically on $\widetilde{M}$. Then $\widetilde{M}/\Gamma$ has a unique Riemannian metric such that the quotient map $\pi: \widetilde{M} \to \widetilde{M}/\Gamma$ is a normal Riemannian covering.
\end{proposition}

\begin{proof}
Proposition C.23 shows that $\pi$ is a smooth normal covering map, and Proposition 2.31 shows that $M = \widetilde{M}/\Gamma$ has a unique Riemannian metric such that $\pi$ is a Riemannian covering.
\end{proof}

\begin{corollary}
\label{cor:isometric-to-quotient}
\dcref{ch2:2.33}
\uses{prop:covering-automorphisms-discrete-lie-group,prop:quotient-orbit-space-covering-map,prop:uniqueness-smooth-quotients}
Suppose $(M, g)$ and $(\widetilde{M}, \tilde{g})$ are connected Riemannian manifolds, $\pi: \widetilde{M} \to M$ is a normal Riemannian covering map, and $\Gamma = \text{Aut}_\pi(\widetilde{M})$. Then $M$ is isometric to $\widetilde{M}/\Gamma$.
\end{corollary}

\begin{proof}
Proposition C.20 shows that with the discrete topology, $\Gamma$ is a discrete Lie group acting smoothly, freely, and properly on $\widetilde{M}$, and then Proposition C.23 shows that $\widetilde{M}/\Gamma$ is a smooth manifold and the quotient map $q: \widetilde{M} \to \widetilde{M}/\Gamma$ is a smooth normal covering map. The fact that both $\pi$ and $q$ are normal coverings implies that $\Gamma$ acts transitively on the fibers of both maps, so the two maps are constant on each other's fibers. Proposition A.19 then implies that there is a diffeomorphism $F: M \to \widetilde{M}/\Gamma$ that satisfies $q \circ F = \pi$. Because both $q$ and $\pi$ are local isometries, $F$ is too, and because it is bijective it is a global isometry.
\end{proof}

\begin{example}
\label{ex:real-projective-space-metric}
\dcref{ch2:2.34}
\uses{ex:universal-covering-projective-space,prop:riemannian-quotient-metric}
The two-element group $\Gamma = \{\pm1\}$ acts smoothly, freely, properly, and isometrically on $S^n$ by multiplication. Example C.24 shows that the quotient space is diffeomorphic to the real projective space $\mathbb{RP}^n$ and the quotient map $q: S^n \to \mathbb{RP}^n$ is a smooth normal covering map. Because the action is isometric, Proposition 2.32 shows that there is a unique metric on $\mathbb{RP}^n$ such that $q$ is a Riemannian covering.
\end{example}

\begin{example}[The Open Möbius Band]
\label{ex:open-mobius-band}
\dcref{ch2:2.35}
\uses{REF:ch2-8}
The open Möbius band is the quotient space $M = \mathbb{R}^2/\mathbb{Z}$, where $\mathbb{Z}$ acts on $\mathbb{R}^2$ by $n \cdot (x,y) = (x+n, (-1)^n y)$. This action is smooth, free, proper, and isometric, and therefore $M$ inherits a flat Riemannian metric such that the quotient map is a Riemannian covering. (See Problem 2-8.)
\end{example}

\begin{exercise}
\label{exer:torus-covering}
\dcref{ch2:2.36}
\uses{ex:n-torus-riemannian-submanifold}
Let $T^n \subseteq \mathbb{R}^{2n}$ be the $n$-torus with its induced metric. Show that the map $X: \mathbb{R}^n \to T^n$ of Example 2.21 is a Riemannian covering.
\end{exercise}

\section{Basic Constructions on Riemannian Manifolds}

Every Riemannian metric yields an abundance of useful constructions on manifolds, besides the obvious ones of lengths of vectors and angles between them. In this section we describe the most basic ones. Throughout this section $M$ is a smooth manifold with or without boundary.

\section{Raising and Lowering Indices}

One elementary but important property of Riemannian metrics is that they allow us to convert vectors to covectors and vice versa. Given a Riemannian metric $g$ on $M$, we define a bundle homomorphism $\tilde{g} : TM \to T^* M$ by setting
\[
\tilde{g}(v)(w) = g_p(v,w)
\]
for all $p \in M$ and $v,w \in T_p M$. If $X$ and $Y$ are smooth vector fields on $M$, this yields
\[
\tilde{g}(X)(Y) = g(X,Y),
\]
which implies, first, that $\tilde{g}(X)(Y)$ is linear over $C^\infty(M)$ in $Y$ and thus $\tilde{g}(X)$ is a smooth covector field by the tensor characterization lemma (Lemma B.6); and second, that the covector field $\tilde{g}(X)$ is linear over $C^\infty(M)$ as a function of $X$, and thus $\tilde{g}$ is a smooth bundle homomorphism.

Given a smooth local frame $(E_i)$ and its dual coframe $(e^i)$, let $g = g_{ij} e^i \otimes e^j$ be the local expression for $g$. If $X = X^i E_i$ is a smooth vector field, the covector field $\tilde{g}(X)$ has the coordinate expression
\[
\tilde{g}(X) = (g_{ij}X^i)e^j.
\]
Thus the matrix of $\tilde{g}$ in any local frame is the same as the matrix of $g$ itself. Given a vector field $X$, it is standard practice to denote the components of the covector field $\tilde{g}(X)$ by
\[
X_j = g_{ij}X^i,
\]
so that
\[
\tilde{g}(X) = X_j e^j,
\]
and we say that $\tilde{g}(X)$ is obtained from $X$ by \textit{lowering an index}. With this in mind, the covector field $g(X)$ is denoted by $X^\flat$ and called $X$ \textit{flat}, borrowing from the musical notation for lowering a tone.

Because the matrix $(g_{ij})$ is nonsingular at each point, the map $\tilde{g}$ is invertible, and the matrix of $\tilde{g}^{-1}$ is just the inverse matrix of $(g_{ij})$. We denote this inverse matrix by $(g^{ij})$, so that $g^{ij}g_{jk} = g_{kj}g^{jl} = \delta_k^l$. The symmetry of $g_{ij}$ easily implies that $(g^{ij})$ is also symmetric in $i$ and $j$. In terms of a local frame, the inverse map $\tilde{g}^{-1}$ is given by
\[
\tilde{g}^{-1}(\omega) = \omega^i E_i,
\]
where
\begin{equation}
\omega^i = g^{ij}\omega_j. \tag{2.13}
\end{equation}

If $\omega$ is a covector field, the vector field $\tilde{g}^{-1}(\omega)$ is called (what else?) $\omega$ \textit{sharp} and denoted by $\omega^\sharp$, and we say that it is obtained from $\omega$ by \textit{raising an index}. The two inverse isomorphisms $\flat$ and $\sharp$ are known as the \textit{musical isomorphisms}.

We record the fibrewise content of the preceding discussion: at each point, $\flat$ is the map $v \mapsto g_p(v,\cdot)$, it is a bijection, and $\sharp$ is its inverse.

\begin{definition}[Lowering an index]
\label{def:flat-map}
\lean{LeeLib.Ch02.flat}
\leanok
\uses{def:riemannian-metric}
Let $(M,g)$ be a Riemannian manifold and $p \in M$. The \emph{flat map} at $p$ is the linear map $\flat : T_p M \to T_p^* M$ given by $v^\flat = g_p(v, \cdot)$.
\end{definition}

\begin{lemma}[The flat map is a bijection]
\label{lem:flat-bijective}
\lean{LeeLib.Ch02.flat_bijective}
\leanok
\uses{def:flat-map,def:riemannian-metric}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$ and $p \in M$. Then $\flat : T_p M \to T_p^* M$ is bijective.
\end{lemma}

\begin{proof}
\leanok
\uses{def:flat-map,def:riemannian-metric}
Since $g_p$ is an inner product on the finite-dimensional real vector space $T_p M$, the space $T_p M$ is a finite-dimensional real inner product space, hence complete. The Riesz representation theorem states that the map sending $v \in T_p M$ to the covector $g_p(v,\cdot)$ is a linear isometric isomorphism of $T_p M$ onto its dual $T_p^* M$. That map is precisely $\flat$, so $\flat$ is bijective.
\end{proof}

\begin{definition}[Raising an index]
\label{def:sharp-map}
\lean{LeeLib.Ch02.sharp}
\leanok
\uses{def:flat-map,lem:flat-bijective}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$ and $p \in M$. The \emph{sharp map} at $p$ is the inverse $\sharp = \flat^{-1} : T_p^* M \to T_p M$, which exists by Lemma \ref{lem:flat-bijective}. Thus $\alpha^\sharp$ is the unique $v \in T_p M$ with $g_p(v,\cdot) = \alpha$, characterized by
\[
g_p(\alpha^\sharp, w) = \alpha(w) \qquad \text{for all } w \in T_p M .
\]
\end{definition}

Probably the most important application of the sharp operator is to extend the classical gradient operator to Riemannian manifolds. If $g$ is a Riemannian metric on $M$ and $f : M \to \mathbb{R}$ is a smooth function, the \emph{gradient of $f$} is the vector field
$\operatorname{grad} f = (df)^\#$ obtained from $df$ by raising an index. Unwinding the definitions,
we see that $\operatorname{grad} f$ is characterized by the fact that
\begin{equation}
df_p(w) = \langle \operatorname{grad} f|_p, w \rangle \quad \text{for all } p \in M, w \in T_p M.
\tag{2.14}
\end{equation}
and has the local basis expression
\[
\operatorname{grad} f = (g^{ij} E_i f) E_j.
\]

Thus if $(E_i)$ is an orthonormal frame, then $\operatorname{grad} f$ is the vector field whose
components are the same as the components of $df$; but in other frames, this will
not be the case.

\begin{definition}[The gradient]
\label{def:gradient}
\lean{LeeLib.Ch02.grad}
\leanok
\uses{def:sharp-map,def:riemannian-metric}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$ and $f : M \to \mathbb{R}$ smooth. The \emph{gradient} of $f$ is $\operatorname{grad} f = (df)^\sharp$, so that $\operatorname{grad} f|_p = (df_p)^\sharp \in T_p M$ for each $p \in M$.
\end{definition}

\begin{lemma}[Characterization of the gradient]
\label{lem:gradient-characterization}
\lean{LeeLib.Ch02.innerAt_grad}
\leanok
\uses{def:gradient,def:sharp-map}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$ and $f : M \to \mathbb{R}$ smooth. Then (2.14) holds: $df_p(w) = \langle \operatorname{grad} f|_p, w\rangle$ for all $p \in M$ and $w \in T_p M$. Moreover $\operatorname{grad} f|_p$ is the unique vector with this property.
\end{lemma}

\begin{proof}
\leanok
\uses{def:gradient,def:sharp-map,lem:flat-bijective}
By Definition \ref{def:sharp-map}, $\alpha^\sharp$ is characterized by $g_p(\alpha^\sharp, w) = \alpha(w)$ for all $w$. Taking $\alpha = df_p$ gives $\langle \operatorname{grad} f|_p, w\rangle = df_p(w)$, which is (2.14). For uniqueness, suppose $v \in T_p M$ also satisfies $\langle v, w\rangle = df_p(w)$ for all $w$. Then $v^\flat = (\operatorname{grad} f|_p)^\flat$, and $\flat$ is injective by Lemma \ref{lem:flat-bijective}, so $v = \operatorname{grad} f|_p$.
\end{proof}

\begin{lemma}[The gradient is orthogonal to the kernel of the differential]
\label{lem:gradient-orthogonal-kernel}
\lean{LeeLib.Ch02.innerAt_grad_eq_zero_of_mem_ker}
\leanok
\uses{def:gradient,lem:gradient-characterization}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$, let $f : M \to \mathbb{R}$ be smooth, and let $p \in M$. If $w \in T_p M$ satisfies $df_p(w) = 0$, then $\langle \operatorname{grad} f|_p, w \rangle = 0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gradient-characterization}
By (2.14), $\langle \operatorname{grad} f|_p, w\rangle = df_p(w) = 0$.
\end{proof}

\begin{lemma}[Critical points are the zeros of the gradient]
\label{lem:gradient-eq-zero-iff}
\lean{LeeLib.Ch02.grad_eq_zero_iff}
\leanok
\uses{def:gradient,lem:gradient-characterization,lem:flat-bijective}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$, let $f : M \to \mathbb{R}$ be smooth, and let $p \in M$. Then $\operatorname{grad} f|_p = 0$ if and only if $df_p = 0$. Hence the regular points of $f$ are exactly the points where $\operatorname{grad} f$ does not vanish.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gradient-characterization,lem:flat-bijective}
If $\operatorname{grad} f|_p = 0$ then for every $w \in T_p M$, (2.14) gives $df_p(w) = \langle 0, w\rangle = 0$, so $df_p = 0$. Conversely, if $df_p = 0$ then for every $w$ we have $\langle \operatorname{grad} f|_p, w\rangle = df_p(w) = 0 = \langle 0, w \rangle$, so $(\operatorname{grad} f|_p)^\flat = 0^\flat$, and $\flat$ is injective by Lemma \ref{lem:flat-bijective}, whence $\operatorname{grad} f|_p = 0$.
\end{proof}

Lemma \ref{lem:gradient-orthogonal-kernel} has a converse, which is what makes the gradient a faithful record of the differential: orthogonality to $\operatorname{grad} f|_p$ detects the kernel of $df_p$ exactly.

\begin{lemma}[Orthogonality to the gradient characterizes the kernel of the differential]
\label{lem:gradient-orthogonal-iff}
\lean{LeeLib.Ch02.innerAt_grad_eq_zero_iff}
\leanok
\uses{def:gradient,lem:gradient-characterization}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$, let $f : M \to \mathbb{R}$ be smooth, let $p \in M$, and let $w \in T_p M$. Then $\langle \operatorname{grad} f|_p, w\rangle = 0$ if and only if $df_p(w) = 0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gradient-characterization}
Both directions are immediate from (2.14): $\langle \operatorname{grad} f|_p, w\rangle = df_p(w)$, so one side vanishes precisely when the other does.
\end{proof}

The next proposition shows that the gradient has the same geometric interpretation
on a Riemannian manifold as it does in Euclidean space. If $f$ is a smooth
real-valued function on a smooth manifold $M$, recall that a point $p \in M$ is called a
\emph{regular point of $f$} if $df_p \neq 0$, and a \emph{critical point of $f$} otherwise; and a level set
$f^{-1}(c)$ is called a \emph{regular level set} if every point of $f^{-1}(c)$ is a regular point of $f$
(see Appendix A). Corollary A.26 shows that each regular level set is an embedded
smooth hypersurface in $M$.

\begin{definition}[The regular set]
\label{def:regular-set}
\lean{LeeLib.Ch02.regularSet}
\leanok
Let $M$ be a smooth manifold and $f : M \to \mathbb{R}$ smooth. The \emph{regular set} of $f$ is $\mathcal{R} = \{x \in M : df_x \neq 0\}$, the set of regular points of $f$.
\end{definition}

\begin{lemma}[The regular set is where the gradient does not vanish]
\label{lem:regular-set-eq-gradient-nonvanishing}
\lean{LeeLib.Ch02.mem_regularSet_iff_grad_ne_zero}
\leanok
\uses{def:regular-set,def:gradient,lem:gradient-eq-zero-iff}
Let $(M,g)$ be a Riemannian manifold with $\dim M < \infty$ and $f : M \to \mathbb{R}$ smooth. Then $x \in \mathcal{R}$ if and only if $\operatorname{grad} f|_x \neq 0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gradient-eq-zero-iff}
By Lemma \ref{lem:gradient-eq-zero-iff}, $\operatorname{grad} f|_x = 0$ if and only if $df_x = 0$. Negating both sides gives the claim.
\end{proof}

The reduction of Proposition \ref{prop:regular-level-set-hypersurface} to the regular-value case rests on $\mathcal{R}$ being open, which is what makes it an open submanifold on which $f$ has $c$ as a regular value.

\begin{lemma}[The non-vanishing locus of a differential is open]
\label{lem:mfderiv-nonvanishing-open}
\lean{LeeLib.Ch02.isOpen_setOf_mfderiv_ne_zero}
\leanok
Let $F : M \to N$ be a smooth map of smooth manifolds. Then $\{x \in M : dF_x \neq 0\}$ is open in $M$.
\end{lemma}

\begin{proof}
\leanok
Fix $x_0$ with $dF_{x_0} \neq 0$, and read $x \mapsto dF_x$ in tangent coordinates around $x_0$ and $F(x_0)$: choose trivializations $\Phi$ of $TM$ over a neighbourhood of $x_0$ and $\Psi$ of $TN$ over a neighbourhood of $F(x_0)$, and set
\[
  D(x) = \Psi_{F(x)} \circ dF_x \circ \Phi_x^{-1},
\]
a map into the fixed vector space $\operatorname{Hom}(E, E')$ of the two model spaces, defined for $x$ in the (open) set where both trivializations apply and $F(x)$ lies in the second base set. Since $F$ is smooth, $D$ is smooth, hence continuous, at $x_0$.

Because $\Phi_x$ and $\Psi_{F(x)}$ are linear isomorphisms on the relevant fibres, $D(x) = 0$ if and only if $dF_x = 0$. In particular $D(x_0) \neq 0$. As $\{L : L \neq 0\}$ is open in $\operatorname{Hom}(E,E')$ and $D$ is continuous at $x_0$, we get $D(x) \neq 0$, and hence $dF_x \neq 0$, for all $x$ in a neighbourhood of $x_0$. So every point of the set is interior to it.
\end{proof}

\begin{lemma}[The regular set is open]
\label{lem:regular-set-open}
\lean{LeeLib.Ch02.isOpen_regularSet}
\leanok
\uses{def:regular-set,lem:mfderiv-nonvanishing-open}
Let $M$ be a smooth manifold and $f : M \to \mathbb{R}$ smooth. Then the regular set $\mathcal{R}$ of $f$ is open in $M$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:mfderiv-nonvanishing-open}
This is Lemma \ref{lem:mfderiv-nonvanishing-open} applied to $F = f$ with $N = \mathbb{R}$, since $\mathcal{R} = \{x : df_x \neq 0\}$ by definition.
\end{proof}

For a regular \emph{value} $c$ — one for which every point of $f^{-1}(c)$ is regular — the level set is a hypersurface by Proposition \ref{prop:regular-level-set-hypersurface-codim-one}, and the gradient is normal to it.

\begin{proposition}[The gradient is normal to a regular level set]
\label{prop:gradient-normal-regular-level-set}
\lean{LeeLib.Ch02.innerAt_grad_mfderiv_levelSet_val_eq_zero}
\leanok
\uses{prop:regular-level-set-hypersurface-codim-one,prop:regular-level-set-tangent-kernel,def:gradient,lem:gradient-orthogonal-kernel}
Let $(M,g)$ be a Riemannian manifold of dimension $n+1$ without boundary, let $f \in C^\infty(M)$, and let $c$ be a regular value of $f$. Then for every $x_0 \in f^{-1}(c)$ and every $w \in T_{x_0} f^{-1}(c)$,
\[
  \bigl\langle \operatorname{grad} f|_{x_0},\, d(\iota_c)_{x_0}(w) \bigr\rangle = 0 ,
\]
where $\iota_c : f^{-1}(c) \hookrightarrow M$ is the inclusion. That is, $\operatorname{grad} f$ is normal to $f^{-1}(c)$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:regular-level-set-tangent-kernel,lem:gradient-orthogonal-kernel}
Since $f \circ \iota_c$ is constant, the chain rule gives $df_{x_0}(d(\iota_c)_{x_0}(w)) = 0$ for every $w$, which is the inclusion $\subseteq$ of Proposition \ref{prop:regular-level-set-tangent-kernel}. So $d(\iota_c)_{x_0}(w) \in \ker df_{x_0}$, and Lemma \ref{lem:gradient-orthogonal-kernel} applies.
\end{proof}

\begin{proposition}[The tangent space of a regular level set is the orthogonal complement of the gradient]
\label{prop:tangent-space-regular-level-set-gradient-orthogonal}
\lean{LeeLib.Ch02.exists_mfderiv_levelSet_val_eq_of_innerAt_grad_eq_zero}
\leanok
\uses{prop:regular-level-set-tangent-kernel,lem:gradient-orthogonal-iff,def:gradient}
In the situation of Proposition \ref{prop:gradient-normal-regular-level-set}, conversely, every $u \in T_{x_0} M$ with $\langle \operatorname{grad} f|_{x_0}, u\rangle = 0$ lies in the image of $d(\iota_c)_{x_0}$. Hence
\[
  d(\iota_c)_{x_0}\bigl(T_{x_0} f^{-1}(c)\bigr) = \bigl(\operatorname{grad} f|_{x_0}\bigr)^{\perp} ,
\]
so the normal space of $f^{-1}(c)$ at $x_0$ is the line spanned by $\operatorname{grad} f|_{x_0}$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:regular-level-set-tangent-kernel,lem:gradient-orthogonal-iff}
If $\langle \operatorname{grad} f|_{x_0}, u\rangle = 0$ then $df_{x_0}(u) = 0$ by Lemma \ref{lem:gradient-orthogonal-iff}, so $u \in \ker df_{x_0}$. By Proposition \ref{prop:regular-level-set-tangent-kernel} the image of $d(\iota_c)_{x_0}$ is all of $\ker df_{x_0}$, so $u$ is in that image. Combined with Proposition \ref{prop:gradient-normal-regular-level-set} this gives the displayed equality of subspaces; the normal space is then the orthogonal complement of a hyperplane, hence the line spanned by $\operatorname{grad} f|_{x_0}$, which is nonzero by Lemma \ref{lem:regular-set-eq-gradient-nonvanishing}.
\end{proof}

Proposition \ref{prop:regular-level-set-hypersurface} is the same statement for the possibly-critical level set $f^{-1}(c)$, of which only the regular part $M_c = f^{-1}(c) \cap \mathcal{R}$ is retained. It reduces to the regular-value case above once one knows that $\mathcal{R}$ is open, so that $M_c$ is a regular level set of the restriction $f|_{\mathcal{R}}$ on the open submanifold $\mathcal{R}$. Only the case $N = \mathbb{R}$ of the regular level set theorem is used, so Proposition \ref{prop:regular-level-set-hypersurface-codim-one} suffices and the general Corollary \ref{cor:regular-level-set} is not needed.

\subsection*{Open submanifolds and restriction}

An open subset $U \subseteq M$ is a smooth manifold of the same dimension, its charts being those of $M$ restricted. The two lemmas below say that this identification is invisible to differentials, which is what makes the reduction to the regular-value case work.

\begin{lemma}[The differential of an open-submanifold inclusion]
\label{lem:opens-inclusion-differential}
\lean{LeeLib.Ch02.mfderiv_opens_subtypeVal}
\leanok
Let $M$ be a smooth manifold and $U \subseteq M$ open, regarded as an open submanifold. Then for each $x \in U$ the differential of the inclusion $\iota : U \hookrightarrow M$ is the identity,
\[
  d\iota_x = \operatorname{id} : T_x U \to T_x M ,
\]
under the identification of $T_x U$ with $T_x M$ induced by the charts. In particular $d\iota_x$ is an isomorphism.
\end{lemma}

\begin{proof}
\leanok
The chart of $U$ at $x$ is the chart of $M$ at $x$ restricted to $U$, so the two charts agree at every point of $U$ near $x$. The coordinate representation of $\iota$ is therefore $\kappa \circ \iota \circ \kappa^{-1}$ for one and the same chart $\kappa$, which is the identity on the chart's image. Differentiating the identity gives $d\iota_x = \operatorname{id}$.
\end{proof}

\begin{lemma}[Restriction to an open set preserves the differential]
\label{lem:opens-restrict-differential}
\lean{LeeLib.Ch02.mfderiv_opens_restrict}
\leanok
\uses{lem:opens-inclusion-differential}
Let $F : M \to M'$ be a smooth map, $U \subseteq M$ open, and $x \in U$. Then
\[
  d(F|_U)_x = dF_x .
\]
\end{lemma}

\begin{proof}
\leanok
\uses{lem:opens-inclusion-differential}
$F|_U = F \circ \iota$ for the inclusion $\iota : U \hookrightarrow M$, so the chain rule gives $d(F|_U)_x = dF_x \circ d\iota_x$, and $d\iota_x = \operatorname{id}$ by Lemma \ref{lem:opens-inclusion-differential}.
\end{proof}

\begin{lemma}[Every value is a regular value of $f|_{\mathcal{R}}$]
\label{lem:regular-restrict-regular-value}
\lean{LeeLib.Ch02.regularRestrict_regular}
\leanok
\uses{def:regular-set,lem:regular-set-open,lem:opens-restrict-differential}
Let $M$ be a smooth manifold, $f \in C^\infty(M)$, and $\mathcal{R}$ the regular set of $f$, regarded as an open submanifold of $M$ (Lemma \ref{lem:regular-set-open}). Then $d(f|_{\mathcal{R}})_x \neq 0$ for every $x \in \mathcal{R}$; that is, every $c \in \mathbb{R}$ is a regular value of $f|_{\mathcal{R}}$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:opens-restrict-differential,def:regular-set}
By Lemma \ref{lem:opens-restrict-differential}, $d(f|_{\mathcal{R}})_x = df_x$, and $df_x \neq 0$ because $x \in \mathcal{R}$ by the definition of the regular set.
\end{proof}

\subsection*{The regular part of an arbitrary level set}

Throughout, $M_c$ denotes $(f|_{\mathcal{R}})^{-1}(c)$, the level set of the restriction inside the open submanifold $\mathcal{R}$; Lemma \ref{lem:regular-part-image} identifies it with Lee's subset $f^{-1}(c) \cap \mathcal{R}$ of $M$.

\begin{proposition}
\label{prop:regular-level-set-hypersurface}
\lean{LeeLib.Ch02.isManifold_regularLevelSet}
\leanok
\dcref{ch2:2.37}
\uses{prop:regular-level-set-hypersurface-codim-one,def:regular-set,lem:regular-set-open,lem:regular-restrict-regular-value}
Suppose $M$ is a smooth $(n+1)$-manifold without boundary, $f \in C^\infty(M)$, and
$\mathcal{R} \subseteq M$ is the set of regular points of $f$. For each $c \in \mathbb{R}$, the set $M_c = f^{-1}(c) \cap \mathcal{R}$ is a smooth $n$-manifold.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:regular-level-set-hypersurface-codim-one,lem:regular-restrict-regular-value,lem:regular-set-open}
By Lemma \ref{lem:regular-set-open} the set $\mathcal{R}$ is open, so it is an open submanifold of $M$ of the same dimension $n+1$. Then $M_c = (f|_{\mathcal{R}})^{-1}(c)$, and $c$ is a regular value of $f|_{\mathcal{R}}$ by Lemma \ref{lem:regular-restrict-regular-value}. Proposition \ref{prop:regular-level-set-hypersurface-codim-one}, applied to $f|_{\mathcal{R}}$ on the manifold $\mathcal{R}$, gives the slice-chart atlas making $M_c$ a smooth $n$-manifold.
\end{proof}

\begin{lemma}[The regular part, as a subset of $M$]
\label{lem:regular-part-image}
\lean{LeeLib.Ch02.range_regularLevelSetIncl}
\leanok
\uses{prop:regular-level-set-hypersurface,def:regular-set}
In the situation of Proposition \ref{prop:regular-level-set-hypersurface}, the image of $M_c = (f|_{\mathcal{R}})^{-1}(c)$ under the inclusion $M_c \hookrightarrow \mathcal{R} \hookrightarrow M$ is exactly $f^{-1}(c) \cap \mathcal{R}$.
\end{lemma}

\begin{proof}
\leanok
A point of $M$ lies in the image if and only if it lies in $\mathcal{R}$ and $f$ takes the value $c$ there, which is the definition of $f^{-1}(c) \cap \mathcal{R}$.
\end{proof}

\begin{lemma}[The regular part is embedded]
\label{lem:regular-part-embedded}
\lean{LeeLib.Ch02.isEmbedding_regularLevelSetIncl}
\leanok
\uses{prop:regular-level-set-hypersurface}
In the situation of Proposition \ref{prop:regular-level-set-hypersurface}, the inclusion $M_c \hookrightarrow M$ is a topological embedding.
\end{lemma}

\begin{proof}
\leanok
It is the composite of the inclusions $M_c \hookrightarrow \mathcal{R}$ and $\mathcal{R} \hookrightarrow M$, each of which carries the subspace topology and so is a topological embedding; a composite of embeddings is an embedding.
\end{proof}

\begin{lemma}[The inclusion of the regular part is smooth]
\label{lem:regular-part-inclusion-smooth}
\lean{LeeLib.Ch02.contMDiff_regularLevelSetIncl}
\leanok
\uses{prop:regular-level-set-hypersurface,prop:regular-level-set-inclusion-smooth}
In the situation of Proposition \ref{prop:regular-level-set-hypersurface}, the inclusion $M_c \hookrightarrow M$ is smooth.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:regular-level-set-inclusion-smooth}
It is the composite of $M_c \hookrightarrow \mathcal{R}$, smooth by Proposition \ref{prop:regular-level-set-inclusion-smooth} applied to $f|_{\mathcal{R}}$, with the inclusion $\mathcal{R} \hookrightarrow M$ of an open submanifold, which is smooth.
\end{proof}

\begin{lemma}[The inclusion of the regular part is an immersion]
\label{lem:regular-part-inclusion-immersion}
\lean{LeeLib.Ch02.injective_mfderiv_regularLevelSetIncl}
\leanok
\uses{prop:regular-level-set-hypersurface,lem:opens-inclusion-differential,prop:regular-level-set-inclusion-smooth}
In the situation of Proposition \ref{prop:regular-level-set-hypersurface}, the differential of the inclusion $M_c \hookrightarrow M$ is injective at every point. With Lemma \ref{lem:regular-part-embedded}, $M_c$ is an embedded smooth hypersurface in $M$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:opens-inclusion-differential,prop:regular-level-set-inclusion-smooth}
The differential of the composite $M_c \hookrightarrow \mathcal{R} \hookrightarrow M$ is $d(\mathcal{R} \hookrightarrow M) \circ d(M_c \hookrightarrow \mathcal{R})$, and the first factor is the identity by Lemma \ref{lem:opens-inclusion-differential}. So the differential equals that of $M_c \hookrightarrow \mathcal{R}$, which is injective by Proposition \ref{prop:regular-level-set-inclusion-smooth} applied to $f|_{\mathcal{R}}$.
\end{proof}

\begin{lemma}[The tangent space of the regular part]
\label{lem:regular-part-tangent-kernel}
\lean{LeeLib.Ch02.range_mfderiv_regularLevelSetIncl}
\leanok
\uses{prop:regular-level-set-hypersurface,prop:regular-level-set-tangent-kernel,lem:opens-restrict-differential}
In the situation of Proposition \ref{prop:regular-level-set-hypersurface}, for every $x \in M_c$ the image of the differential of the inclusion $M_c \hookrightarrow M$ is exactly $\ker df_x$; that is, $T_x M_c = \ker df_x$.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:regular-level-set-tangent-kernel,lem:opens-restrict-differential,lem:opens-inclusion-differential}
By Lemma \ref{lem:opens-inclusion-differential} the differential of $M_c \hookrightarrow M$ equals that of $M_c \hookrightarrow \mathcal{R}$, whose image is $\ker d(f|_{\mathcal{R}})_x$ by Proposition \ref{prop:regular-level-set-tangent-kernel} applied to $f|_{\mathcal{R}}$. Finally $d(f|_{\mathcal{R}})_x = df_x$ by Lemma \ref{lem:opens-restrict-differential}, so the two kernels coincide.
\end{proof}

\begin{proposition}[$\operatorname{grad} f$ is normal to the regular part]
\label{prop:gradient-normal-regular-part}
\lean{LeeLib.Ch02.innerAt_grad_mfderiv_regularLevelSetIncl_eq_zero}
\leanok
\uses{prop:regular-level-set-hypersurface,lem:regular-part-tangent-kernel,def:gradient,lem:gradient-orthogonal-kernel}
Let $(M,g)$ be a Riemannian manifold of dimension $n+1$ without boundary, $f \in C^\infty(M)$, and $c \in \mathbb{R}$. Then for every $x \in M_c$ and every $w \in T_x M_c$,
\[
  \bigl\langle \operatorname{grad} f|_{x},\, d\iota_{x}(w) \bigr\rangle = 0 ,
\]
where $\iota : M_c \hookrightarrow M$ is the inclusion. That is, $\operatorname{grad} f$ is everywhere normal to $M_c$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:regular-part-tangent-kernel,lem:gradient-orthogonal-kernel}
Since $f$ is constant on $M_c$, the chain rule gives $df_x(d\iota_x(w)) = 0$ for every $w \in T_x M_c$, so $d\iota_x(w) \in \ker df_x$. Lemma \ref{lem:gradient-orthogonal-kernel} then gives $\langle \operatorname{grad} f|_x, d\iota_x(w)\rangle = 0$.
\end{proof}

\begin{lemma}[$\operatorname{grad} f$ does not vanish on the regular part]
\label{lem:gradient-nonvanishing-regular-part}
\lean{LeeLib.Ch02.grad_ne_zero_regularLevelSet}
\leanok
\uses{prop:regular-level-set-hypersurface,lem:regular-set-eq-gradient-nonvanishing,def:gradient}
In the situation of Proposition \ref{prop:gradient-normal-regular-part}, $\operatorname{grad} f|_x \neq 0$ for every $x \in M_c$. Together with Proposition \ref{prop:gradient-normal-regular-part} and Lemma \ref{lem:regular-part-tangent-kernel}, $\operatorname{grad} f|_x$ spans the normal line of the hypersurface $M_c$ at $x$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:regular-set-eq-gradient-nonvanishing}
Every point of $M_c$ lies in $\mathcal{R}$, so this is Lemma \ref{lem:regular-set-eq-gradient-nonvanishing}.
\end{proof}

The flat and sharp operators can be applied to tensors of any rank, in any index
position, to convert tensors from covariant to contravariant or vice versa. Formally,
this operation is defined as follows: if $F$ is any $(k,l)$-tensor and $i \in \{1, \ldots, k + l\}$ is
any covariant index position for $F$ (meaning that the $i$th argument is a vector, not a
covector), we can form a new tensor $F^\#$ of type $(k + 1, l - 1)$ by setting
\[
F^\#(\alpha_1, \ldots, \alpha_{k+l}) = F(\alpha_1, \ldots, \alpha_{i-1}, \alpha_i^*, \alpha_{i+1}, \ldots, \alpha_{k+l})
\]
whenever $\alpha_1, \ldots, \alpha_{k+l}$ are vectors or covectors as appropriate. In any local frame,
the components of $F^\#$ are obtained by multiplying the components of $F$ by $g^{kl}$ and
contracting one of the indices of $g^{kl}$ with the $i$th index of $F$. Similarly, if $i$ is a
contravariant index position, we can define a $(k - 1, l + 1)$-tensor $F^b$ by
\[
F^b(\alpha_1, \ldots, \alpha_{k+l}) = F(\alpha_1, \ldots, \alpha_{i-1}, \alpha_i^*, \alpha_{i+1}, \ldots, \alpha_{k+l}).
\]

In components, it is computed by multiplying by $g_{kl}$ and contracting.

For example, if $A$ is a mixed 3-tensor given in terms of a local frame by
\[
A = A_i^{jk} \varepsilon^i \otimes E_j \otimes \varepsilon^k
\]
(see (B.6)), we can lower its middle index to obtain a covariant 3-tensor $A^b$ with components
\[
A_{ijk} = g_{jl} A^l_{ik}.
\]

To avoid overly cumbersome notation, we use the symbols $F^\#$ and $F^b$ without explicitly specifying which index position the sharp or flat operator is to be applied to; when there is more than one choice, we will always stipulate in words what is meant.

Another important application of the flat and sharp operators is to extend the trace operator introduced in Appendix B to covariant tensors. If $h$ is any covariant $k$-tensor field on a Riemannian manifold with $k \ge 2$, we can raise one of its indices (say the last one for definiteness) and obtain a $(1, k-1)$-tensor $h^\#$. The trace of $h^\#$ is thus a well-defined covariant $(k-2)$-tensor field (see Exercise B.3). We define the \textbf{trace of $h$ with respect to $g$} as
\[
\text{tr}_g h = \text{tr}(h^\#).
\]

Sometimes we may wish to raise an index other than the last, or to take the trace on a pair of indices other than the last covariant and contravariant ones. In each such case, we will say in words what is meant.

The most important case is that of a covariant 2-tensor field. In this case, $h^\#$ is a $(1,1)$-tensor field, which can equivalently be regarded as an endomorphism field, and $\text{tr}_g h$ is just the ordinary trace of this endomorphism field. In terms of a basis, this is
\[
\text{tr}_g h = h_i^i = g^{ij} h_{ij}.
\]

In particular, in an orthonormal frame this is the ordinary trace of the matrix $(h_{ij})$ (the sum of its diagonal entries); but if the frame is not orthonormal, then this trace is different from the ordinary trace.

\begin{exercise}
\label{exer:inverse-metric-ambiguity}
\dcref{ch2:2.38}
If $g$ is a Riemannian metric on $M$ and $(E_i)$ is a local frame on $M$, there is a potential ambiguity about what the expression $(g^{ij})$ represents: we have defined it to mean the inverse matrix of $(g_{ij})$, but one could also interpret it as the components of the contravariant 2-tensor field $g^{\#\#}$ obtained by raising both of the indices of $g$. Show that these two interpretations lead to the same result.
\end{exercise}

\section{Inner Products of Tensors}

A Riemannian metric yields, by definition, an inner product on tangent vectors at each point. Because of the musical isomorphisms between vectors and covectors, it is easy to carry the inner product over to covectors as well.

Suppose $g$ is a Riemannian metric on $M$, and $x \in M$. We can define an inner product on the cotangent space $T_x^* M$ by
\[
\langle \omega, \eta \rangle_g = \langle \omega^\#, \eta^\# \rangle_g.
\]
(Just as with inner products of vectors, we might sometimes omit $g$ from the notation when the metric is understood.) To see how to compute this, we just use the basis formula (2.13) for the sharp operator, together with the relation $g_{kl}g^{kl} = g_{ik}g^{kl} = \delta_l^i$, to obtain
\begin{align*}
\langle\omega, \eta\rangle &= g_{kl}(g^{ki}\omega_i)(g^{lj}\eta_j)\\
&= \delta_l^i g^{lj}\omega_i\eta_j\\
&= g^{ij}\omega_i\eta_j.
\end{align*}

In other words, the inner product on covectors is represented by the inverse matrix $(g^{ij})$. Using our conventions for raising and lowering indices, this can also be written
\begin{align*}
\langle\omega, \eta\rangle &= \omega_i\eta^i = \omega^j\eta_j.
\end{align*}

\begin{exercise}
\label{exer:orthonormal-frames-and-coframes}
\dcref{ch2:2.39}
Let $(M, g)$ be a Riemannian manifold with or without boundary, let $(E_i)$ be a local frame for $M$, and let $(e^i)$ be its dual coframe. Show that the following are equivalent:
\begin{enumerate}
\item[(a)] $(E_i)$ is orthonormal.
\item[(b)] $(e^i)$ is orthonormal.
\item[(c)] $(e^i)^{\sharp} = E_i$ for each $i$.
\end{enumerate}
\end{exercise}

This construction can be extended to tensor bundles of any rank, as the following proposition shows. First a bit of terminology: if $E \to M$ is a smooth vector bundle, a \textbf{smooth fiber metric} on $E$ is an inner product on each fiber $E_p$ that varies smoothly, in the sense that for any (local) smooth sections $\sigma, \tau$ of $E$, the inner product $\langle\sigma, \tau\rangle$ is a smooth function.

\begin{proposition}[Inner Products of Tensors]
\label{prop:inner-products-of-tensors}
\dcref{ch2:2.40}
Let $(M, g)$ be an $n$-dimensional Riemannian manifold with or without boundary. There is a unique smooth fiber metric on each tensor bundle $T^{(k,l)}TM$ with the property that if $\alpha_1, \ldots, \alpha_{k+l}, \beta_1, \ldots, \beta_{k+l}$ are vector or covector fields as appropriate, then
\begin{equation}
\langle\alpha_1 \otimes \cdots \otimes \alpha_{k+l}, \beta_1 \otimes \cdots \otimes \beta_{k+l}\rangle = \langle\alpha_1, \beta_1\rangle \cdots \langle\alpha_{k+l}, \beta_{k+l}\rangle. \tag{2.15}
\end{equation}

With this inner product, if $(E_1, \ldots, E_n)$ is a local orthonormal frame for $TM$ and $(e^1, \ldots, e^n)$ is the corresponding dual coframe, then the collection of tensor fields $E_{i_1} \otimes \cdots \otimes E_{i_k} \otimes e^{j_1} \otimes \cdots \otimes e^{j_l}$ as all the indices range from $1$ to $n$ forms a local orthonormal frame for $T^{(k,l)}(T_pM)$. In terms of any (not necessarily orthonormal) frame, this fiber metric satisfies
\begin{equation}
\langle F, G\rangle = g_{i_1r_1} \cdots g_{i_kr_k} g^{j_1s_1} \cdots g^{j_ls_l} F_{j_1 \cdots j_l}^{i_1 \cdots i_k} G_{s_1 \cdots s_l}^{r_1 \cdots r_k}. \tag{2.16}
\end{equation}

If $F$ and $G$ are both covariant, this can be written
\[
\langle F, G\rangle = F_{j_1 \cdots j_l} G^{j_1 \cdots j_l},
\]
where the last factor on the right represents the components of $G$ with all of its
indices raised:
\[
G^{j_1 \ldots j_i} = g^{i_1 s_1} \cdots g^{j_i s_I} G_{s_1 \ldots s_I}.
\]
\end{proposition}

\textit{Proof.} Problem 2-11. $\square$

\section{The Volume Form and Integration}

Another important construction provided by a metric on an oriented manifold is a
canonical volume form.

The construction needs one notion that the development so far has not: an orientation of $M$
that varies smoothly from point to point. A choice of orientation of each tangent space
already singles out a form $dV_g|_x$ on each $T_x M$; what makes $x \mapsto dV_g|_x$ a
\emph{differential form}, rather than an unrelated family of forms, is exactly the smoothness
of that choice.

\begin{definition}[Smooth Orientation]
\label{def:smooth-orientation}
\lean{LeeLib.Ch02.IsSmoothOrientation}
\leanok
A \emph{pointwise orientation} of a smooth $n$-manifold $M$ is a choice of orientation $o_x$ of
the vector space $T_x M$ for each $x \in M$. It is \emph{smooth} if every $p \in M$ has an open
neighbourhood $u$ carrying a smooth local frame $(Y_1, \ldots, Y_n)$ for $TM$ that is
positively oriented throughout $u$: for every $x \in u$ the basis $(Y_1|_x, \ldots, Y_n|_x)$ of
$T_x M$ induces $o_x$. An \emph{oriented} Riemannian manifold is a Riemannian manifold together
with a smooth orientation.
\end{definition}

\begin{lemma}[Gram--Schmidt Preserves Orientation]
\label{lem:gram-schmidt-preserves-orientation}
\lean{InnerProductSpace.gramSchmidtOrthonormalBasis_orientation}
\leanok
Let $V$ be an $n$-dimensional real inner product space and let $b = (b_1, \ldots, b_n)$ be a
basis of $V$. Then the Gram--Schmidt orthonormalization of $b$ induces the same orientation of
$V$ as $b$ does.
\end{lemma}

\begin{proof}
\leanok
Write $u_i$ for the unnormalized Gram--Schmidt vectors and $\hat u_i = u_i / |u_i|$ for their
normalizations, so that $(\hat u_i)$ is the orthonormalization of $b$. The change of basis from
$b$ to $(\hat u_i)$ is triangular, and its determinant equals
$\prod_{i=1}^{n} \langle \hat u_i, b_i \rangle$.

Each factor is positive. Indeed the recursion defining $u_i$ subtracts from $b_i$ a combination
of $u_1, \ldots, u_{i-1}$, each of which is orthogonal to $u_i$; hence
$\langle u_i, b_i \rangle = \langle u_i, u_i \rangle = |u_i|^2$, and therefore
$\langle \hat u_i, b_i \rangle = |u_i|$. Since $b$ is linearly independent, $u_i \ne 0$ and so
$|u_i| > 0$.

The determinant of the change of basis is thus positive, and a change of basis with positive
determinant preserves orientation.
\end{proof}

\begin{proposition}[Existence of Oriented Orthonormal Frames]
\label{prop:oriented-orthonormal-frames}
\lean{LeeLib.Ch02.exists_orientedOrthonormalFrame_nhds}
\leanok
\uses{def:smooth-orientation,lem:gram-schmidt-preserves-orientation,prop:existence-orthonormal-frames,def:riemannian-metric}
Let $(M,g)$ be a Riemannian $n$-manifold with a smooth orientation $o$. Then every $p \in M$ has
an open neighbourhood $u$ carrying a smooth local frame $(Z_1, \ldots, Z_n)$ for $TM$ that is
both orthonormal and positively oriented at every $x \in u$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:smooth-orientation,lem:gram-schmidt-preserves-orientation,prop:existence-orthonormal-frames}
By Definition \ref{def:smooth-orientation} there are an open $u \ni p$ and a smooth local frame
$(Y_1, \ldots, Y_n)$ on $u$ that is positively oriented at every $x \in u$. Orthonormalize it
fibrewise: by Proposition \ref{prop:existence-orthonormal-frames} the resulting family
$(Z_1, \ldots, Z_n)$ is again a smooth local frame on $u$, orthonormal at every $x \in u$. By
Lemma \ref{lem:gram-schmidt-preserves-orientation}, applied in the inner product space
$(T_x M, g_x)$, the basis $(Z_1|_x, \ldots, Z_n|_x)$ induces the same orientation as
$(Y_1|_x, \ldots, Y_n|_x)$, which is $o_x$.

The initial-span condition of Proposition \ref{prop:existence-orthonormal-frames} is not by
itself enough here: $\operatorname{span}(Z_1|_x, \ldots, Z_k|_x) =
\operatorname{span}(Y_1|_x, \ldots, Y_k|_x)$ for every $k$ does not determine the orientation of
$(Z_i|_x)$, as replacing $Z_1$ by $-Z_1$ shows. This is why the orientation must be tracked
through the orthonormalization, by Lemma \ref{lem:gram-schmidt-preserves-orientation}.
\end{proof}

\begin{definition}[The Riemannian Volume Form]
\label{def:volume-form-pointwise}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm}
\leanok
\uses{def:riemannian-metric,def:smooth-orientation}
Let $(M,g)$ be a Riemannian $n$-manifold and let $o$ be a pointwise orientation of $M$. For each
$x \in M$ let $dV_g|_x$ be the volume form of the oriented inner product space
$(T_x M, g_x, o_x)$, that is, the unique alternating $n$-form on $T_x M$ taking the value $1$ on
every orthonormal basis of $T_x M$ inducing $o_x$. The family $dV_g : x \mapsto dV_g|_x$ is the
\emph{Riemannian volume form} of $(M, g, o)$.
\end{definition}

\begin{proposition}[The Determinant Formula]
\label{prop:volume-form-determinant}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_apply_eq_det}
\leanok
\uses{def:volume-form-pointwise,prop:oriented-orthonormal-frames}
Let $(M,g)$ be a Riemannian $n$-manifold with pointwise orientation $o$, let $(E_1,\ldots,E_n)$
be a local orthonormal frame on an open set $u$ that is positively oriented at $x \in u$, and let
$v_1, \ldots, v_n \in T_x M$. Then
\[
dV_g|_x(v_1, \ldots, v_n) = \det \bigl( \langle E_i|_x, v_j \rangle_g \bigr)_{i,j=1}^{n}.
\]
\end{proposition}

\begin{proof}
\leanok
\uses{def:volume-form-pointwise}
The tuple $(E_1|_x, \ldots, E_n|_x)$ is an orthonormal basis of $(T_x M, g_x)$ inducing $o_x$, so
by Definition \ref{def:volume-form-pointwise} $dV_g|_x$ is the determinant form of that basis.
The determinant form of an orthonormal basis $E$ sends $(v_1,\ldots,v_n)$ to the determinant of
the matrix expressing the $v_j$ in the basis $E$, whose $(i,j)$ entry is the coefficient
$\langle E_i|_x, v_j \rangle_g$ of $v_j$ along $E_i|_x$.
\end{proof}

\begin{proposition}[The Characterizing Property of $dV_g$]
\label{prop:volume-form-value-one}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_apply_eq_one}
\leanok
\uses{def:volume-form-pointwise,prop:oriented-orthonormal-frames}
Let $(M,g)$ be a Riemannian $n$-manifold with pointwise orientation $o$ and let
$(E_1, \ldots, E_n)$ be a local orthonormal frame on an open set $u$ that is positively oriented
at $x \in u$. Then $dV_g|_x(E_1|_x, \ldots, E_n|_x) = 1$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:volume-form-pointwise}
Immediate from Definition \ref{def:volume-form-pointwise}: $(E_1|_x, \ldots, E_n|_x)$ is an
orthonormal basis of $T_x M$ inducing $o_x$, and $dV_g|_x$ takes the value $1$ on every such
basis. Equivalently, the matrix of Proposition \ref{prop:volume-form-determinant} is the
identity, whose determinant is $1$.
\end{proof}

\begin{proposition}[Uniqueness of $dV_g$]
\label{prop:volume-form-uniqueness}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_unique}
\leanok
\uses{def:volume-form-pointwise,prop:volume-form-value-one}
Let $(M,g)$ be a Riemannian $n$-manifold with pointwise orientation $o$, let $(E_1,\ldots,E_n)$
be a local orthonormal frame on an open set $u$ that is positively oriented at $x \in u$, and let
$\theta$ be an alternating $n$-form on $T_x M$ with $\theta(E_1|_x, \ldots, E_n|_x) = 1$. Then
$\theta = dV_g|_x$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:volume-form-value-one}
An alternating $n$-form on an $n$-dimensional space is determined by its value on a single
basis: every such form $\phi$ satisfies $\phi = \phi(E_1|_x,\ldots,E_n|_x) \det\nolimits_E$,
where $\det_E$ is the determinant form of the basis $(E_i|_x)$. Both $\theta$ and $dV_g|_x$ take
the value $1$ there, the latter by Proposition \ref{prop:volume-form-value-one}, so both equal
$\det_E$.
\end{proof}

\begin{lemma}[Top-Degree Forms are Proportional]
\label{lem:top-forms-proportional}
\lean{ContinuousAlternatingMap.eq_smul_of_apply_basis_ne_zero}
\leanok
Let $V$ be an $n$-dimensional real vector space, let $b = (b_1, \ldots, b_n)$ be a basis of $V$,
and let $\omega$ and $\theta$ be alternating $n$-forms on $V$ with
$\theta(b_1, \ldots, b_n) \ne 0$. Then
\[
\omega = \frac{\omega(b_1, \ldots, b_n)}{\theta(b_1, \ldots, b_n)} \, \theta .
\]
\end{lemma}

\begin{proof}
\leanok
Every alternating $n$-form $\phi$ on $V$ satisfies $\phi = \phi(b_1,\ldots,b_n) \det\nolimits_b$,
where $\det_b$ is the determinant form of $b$. Applying this to $\theta$ and dividing by
$\theta(b_1,\ldots,b_n) \ne 0$ expresses $\det_b$ as a multiple of $\theta$; substituting that
into the same identity for $\omega$ gives the claim.
\end{proof}

\begin{definition}[The Wedge Product of Covectors]
\label{def:wedge-covectors}
\lean{LeeLib.Ch02.wedgeCovectors}
\leanok
Let $V$ be a real topological vector space and let $f_1, \ldots, f_n$ be continuous linear
functionals on $V$. Their \emph{wedge product} is the alternating $n$-form
\[
(f_1 \wedge \cdots \wedge f_n)(v_1, \ldots, v_n) = \det \bigl( f_i(v_j) \bigr)_{i,j=1}^{n} .
\]
This is alternating and multilinear in $(v_1, \ldots, v_n)$ because the determinant is
alternating and multilinear in the columns of its matrix, and it is continuous because a
determinant is a polynomial in the entries $f_i(v_j)$, each of which depends continuously on
$v_j$. This is the determinant convention for the wedge product, as used throughout.
\end{definition}

\begin{proposition}[The Coframe Characterization of $dV_g$]
\label{prop:volume-form-coframe}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_eq_wedgeCovectors}
\leanok
\uses{def:wedge-covectors,def:volume-form-pointwise,prop:volume-form-determinant,prop:oriented-orthonormal-frames}
Let $(M,g)$ be a Riemannian $n$-manifold with pointwise orientation $o$, and let
$(E_1, \ldots, E_n)$ be a local orthonormal frame on an open set $u$ that is positively oriented
at $x \in u$. Let $(e^1, \ldots, e^n)$ be the dual coframe, $e^i = \langle E_i, \cdot \rangle_g$.
Then
\[
dV_g|_x = e^1 \wedge \cdots \wedge e^n .
\]
This is characterization (a) of Lee's Proposition 2.41.
\end{proposition}

\begin{proof}
\leanok
\uses{def:wedge-covectors,prop:volume-form-determinant}
Both sides are alternating $n$-forms on $T_x M$, so it suffices to compare them on an arbitrary
tuple $(v_1, \ldots, v_n)$. By Definition \ref{def:wedge-covectors},
\[
(e^1 \wedge \cdots \wedge e^n)(v_1, \ldots, v_n)
  = \det \bigl( e^i(v_j) \bigr)_{i,j}
  = \det \bigl( \langle E_i|_x, v_j \rangle_g \bigr)_{i,j},
\]
which is $dV_g|_x(v_1, \ldots, v_n)$ by Proposition \ref{prop:volume-form-determinant}.
\end{proof}

\begin{proposition}[The Riemannian Volume Form is Smooth]
\label{prop:volume-form-smooth}
\lean{LeeLib.Ch02.RiemannianMetric.contMDiff_volumeForm}
\leanok
\uses{def:volume-form-pointwise,def:smooth-orientation,prop:oriented-orthonormal-frames,lem:top-forms-proportional,prop:volume-form-determinant,prop:volume-form-value-one,lem:inner-product-vector-fields-smooth}
Let $(M,g)$ be a Riemannian $n$-manifold with a smooth orientation $o$. Then $dV_g$ is a smooth
$n$-form on $M$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:oriented-orthonormal-frames,lem:top-forms-proportional,prop:volume-form-determinant,prop:volume-form-value-one,lem:inner-product-vector-fields-smooth}
Smoothness is local, so fix $x_0 \in M$. Choose a local trivialization $e$ of $TM$ over an open
set containing $x_0$ and write $e^{-1}_x : \mathbb{R}^n \to T_x M$ for the induced isomorphism of
the fibre over $x$. The trivialization of $TM$ induces one of the bundle of $n$-forms, under
which the value of $dV_g$ at $x$ is read as the $n$-form
$\Phi(x) = dV_g|_x \circ e^{-1}_x$ on $\mathbb{R}^n$; thus $dV_g$ is smooth at $x_0$ if and only
if $x \mapsto \Phi(x)$ is.

Fix once and for all a basis $b = (b_1,\ldots,b_n)$ of $\mathbb{R}^n$. By Proposition
\ref{prop:oriented-orthonormal-frames} there are an open $u \ni x_0$ and a smooth local frame
$(Z_1, \ldots, Z_n)$ on $u$, orthonormal and positively oriented at every $x \in u$.

First, $\Phi(x_0)(b_1,\ldots,b_n) \ne 0$. Indeed $dV_g|_{x_0}(Z_1|_{x_0},\ldots,Z_n|_{x_0}) = 1$
by Proposition \ref{prop:volume-form-value-one}, so $dV_g|_{x_0} \ne 0$; and
$(e^{-1}_{x_0} b_1, \ldots, e^{-1}_{x_0} b_n)$ is a basis of $T_{x_0} M$, because $e^{-1}_{x_0}$
is an isomorphism; and a nonzero alternating $n$-form cannot vanish on a basis, since by Lemma
\ref{lem:top-forms-proportional} its value on a basis determines it.

Since the $n$-forms on $\mathbb{R}^n$ form a line, Lemma \ref{lem:top-forms-proportional} applied
with $\theta = \Phi(x_0)$ now gives, for every $x$,
\[
\Phi(x) = \frac{\Phi(x)(b_1,\ldots,b_n)}{\Phi(x_0)(b_1,\ldots,b_n)} \, \Phi(x_0) .
\]
The right-hand side is a scalar function of $x$ times the fixed form $\Phi(x_0)$, so it suffices
to prove that $x \mapsto \Phi(x)(b_1,\ldots,b_n)$ is smooth at $x_0$.

By Proposition \ref{prop:volume-form-determinant}, for every $x \in u$,
\[
\Phi(x)(b_1,\ldots,b_n) = dV_g|_x \bigl(e^{-1}_x b_1, \ldots, e^{-1}_x b_n\bigr)
  = \det \bigl( \langle Z_i|_x, e^{-1}_x b_j \rangle_g \bigr)_{i,j=1}^{n} .
\]
For each fixed $j$ the assignment $x \mapsto e^{-1}_x b_j$ is a smooth local section of $TM$: it
is the local frame induced on $TM$ by the trivialization $e$ and the basis $b$, which is constant
in that trivialization. Each $Z_i$ is smooth on $u$ by construction. Hence each entry
$x \mapsto \langle Z_i|_x, e^{-1}_x b_j \rangle_g$ is smooth on $u$ by Lemma
\ref{lem:inner-product-vector-fields-smooth}. A determinant is a polynomial in the entries of
its matrix, so $x \mapsto \Phi(x)(b_1,\ldots,b_n)$ is smooth on $u$, and in particular at $x_0$.
\end{proof}

Characterization (c) of Proposition \ref{prop:riemannian-volume-form} computes $dV_g$ in oriented
local coordinates. It is proved for an arbitrary oriented local frame rather than only for a
coordinate frame: a coordinate frame $(\partial/\partial x^i)$ is a local frame, its matrix
$\bigl(g(\partial_i,\partial_j)\bigr)$ is Lee's $(g_{ij})$, and its dual coframe is $(dx^i)$, so
the general statement contains Lee's verbatim.

\begin{definition}[The Matrix of $g$ in a Frame]
\label{def:frame-matrix}
\lean{LeeLib.Ch02.RiemannianMetric.frameMatrix}
\leanok
Let $(M,g)$ be a Riemannian manifold and $(Y_1,\ldots,Y_n)$ a local frame on an open set $u$. The
\emph{matrix of $g$ in the frame $(Y_i)$} at $x \in u$ is
\[
g_{ij}(x) = \langle Y_i|_x, Y_j|_x \rangle_g .
\]
For a coordinate frame $(\partial/\partial x^i)$ this is Lee's $(g_{ij})$.
\end{definition}

\begin{lemma}[The Gram Determinant of a Frame is Nonzero]
\label{lem:frame-matrix-nonsingular}
\lean{LeeLib.Ch02.RiemannianMetric.frameMatrix_det_ne_zero}
\leanok
\uses{def:frame-matrix}
With the notation of Definition \ref{def:frame-matrix}, $\det\bigl(g_{ij}(x)\bigr) \ne 0$ for
every $x \in u$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:frame-matrix}
The matrix $\bigl(g_{ij}(x)\bigr)$ is the Gram matrix of the family $(Y_1|_x,\ldots,Y_n|_x)$ with
respect to the inner product $\langle\cdot,\cdot\rangle_g$ on $T_x M$. A Gram matrix of a family
of vectors in an inner product space is nonsingular if and only if the family is linearly
independent. Since $(Y_i)$ is a frame, $(Y_i|_x)$ is a basis of $T_x M$, hence linearly
independent, and so $\det\bigl(g_{ij}(x)\bigr) \ne 0$.
\end{proof}

\begin{definition}[The Dual Coframe of a Frame]
\label{def:dual-coframe}
\lean{LeeLib.Ch02.RiemannianMetric.dualCoframe}
\leanok
\uses{def:frame-matrix,lem:frame-matrix-nonsingular}
With the notation of Definition \ref{def:frame-matrix}, write $\bigl(g^{ij}(x)\bigr)$ for the
inverse of the matrix $\bigl(g_{ij}(x)\bigr)$, which exists by Lemma
\ref{lem:frame-matrix-nonsingular}. The \emph{dual coframe} of $(Y_i)$ is the family of covectors
\[
\varepsilon^i|_x = \sum_{j} g^{ij}(x) \, \langle Y_j|_x, \cdot \rangle_g .
\]
For a coordinate frame this is the coordinate coframe $(dx^i)$.
\end{definition}

\begin{lemma}[The Dual Coframe is Dual]
\label{lem:dual-coframe-duality}
\lean{LeeLib.Ch02.RiemannianMetric.dualCoframe_apply_frame}
\leanok
\uses{def:dual-coframe,def:frame-matrix}
With the notation of Definition \ref{def:dual-coframe}, $\varepsilon^i|_x(Y_k|_x) = \delta^i_k$
for all $i,k$ and all $x \in u$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:dual-coframe,def:frame-matrix,lem:frame-matrix-nonsingular}
By definition,
\[
\varepsilon^i|_x(Y_k|_x) = \sum_j g^{ij}(x) \, \langle Y_j|_x, Y_k|_x \rangle_g
  = \sum_j g^{ij}(x) \, g_{jk}(x),
\]
which is the $(i,k)$ entry of the product of $\bigl(g^{ij}(x)\bigr)$ with $\bigl(g_{ij}(x)\bigr)$.
Since the first matrix is by definition the inverse of the second, that product is the identity,
so the entry is $\delta^i_k$.
\end{proof}

\begin{lemma}[The Gram Determinant Formula, Pointwise]
\label{lem:volume-form-gram-determinant}
\lean{LeeLib.Ch02.volumeFormL_apply_eq_sqrt_det_gram}
\leanok
\uses{lem:gram-schmidt-preserves-orientation}
Let $V$ be an oriented $n$-dimensional inner product space with volume form $dV$, and let
$(b_1,\ldots,b_n)$ be a positively oriented basis of $V$, not assumed orthonormal. Then
\[
dV(b_1,\ldots,b_n) = \sqrt{\det\bigl(\langle b_i, b_j \rangle\bigr)} .
\]
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gram-schmidt-preserves-orientation}
Let $(e_1,\ldots,e_n)$ be the Gram--Schmidt orthonormalization of $(b_i)$. By Lemma
\ref{lem:gram-schmidt-preserves-orientation} it is a positively oriented orthonormal basis. Let
$A$ be the matrix $A_{ij} = \langle e_i, b_j \rangle$, so that $b_j = \sum_i A_{ij} e_i$.

Expanding each $b_i$ in the orthonormal basis $(e_k)$ gives
\[
\langle b_i, b_j \rangle = \Bigl\langle \sum_k A_{ki} e_k, \sum_l A_{lj} e_l \Bigr\rangle
  = \sum_k A_{ki} A_{kj} = (A^{\mathsf T} A)_{ij},
\]
so $\det\bigl(\langle b_i, b_j\rangle\bigr) = \det(A^{\mathsf T}A) = (\det A)^2$.

By the determinant formula of Proposition \ref{prop:volume-form-determinant} applied to the
oriented orthonormal basis $(e_i)$, we have $dV(b_1,\ldots,b_n) = \det A$. Finally $\det A > 0$:
$\det A$ is the determinant of the change of basis from $(e_i)$ to $(b_j)$, and it is positive
exactly because the two bases induce the same orientation. Hence
$\sqrt{\det(\langle b_i,b_j\rangle)} = \sqrt{(\det A)^2} = |\det A| = \det A = dV(b_1,\ldots,b_n)$.
\end{proof}

\begin{proposition}[The Value of $dV_g$ on an Oriented Frame]
\label{prop:volume-form-frame-value}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_apply_eq_sqrt_det_frameMatrix}
\leanok
\uses{def:volume-form-pointwise,def:frame-matrix,lem:volume-form-gram-determinant}
Let $(M,g)$ be a Riemannian $n$-manifold with a pointwise orientation, and let $(Y_1,\ldots,Y_n)$
be a local frame on $u$ that is positively oriented at $x \in u$. Then
\[
dV_g|_x(Y_1|_x, \ldots, Y_n|_x) = \sqrt{\det\bigl(g_{ij}(x)\bigr)} .
\]
\end{proposition}

\begin{proof}
\leanok
\uses{lem:volume-form-gram-determinant,def:frame-matrix}
Apply Lemma \ref{lem:volume-form-gram-determinant} to the oriented inner product space $T_x M$ and
its positively oriented basis $(Y_i|_x)$; the Gram matrix $\bigl(\langle Y_i|_x, Y_j|_x\rangle_g\bigr)$
is by Definition \ref{def:frame-matrix} the matrix $\bigl(g_{ij}(x)\bigr)$.
\end{proof}

\begin{proposition}[Lee 2.41(c): the Coordinate Formula for $dV_g$]
\label{prop:volume-form-coordinate-formula}
\lean{LeeLib.Ch02.RiemannianMetric.volumeForm_eq_sqrt_det_smul_wedgeCovectors}
\leanok
\uses{def:volume-form-pointwise,def:frame-matrix,def:wedge-covectors,prop:volume-form-frame-value,lem:top-forms-proportional,lem:dual-coframe-duality}
Let $(M,g)$ be a Riemannian $n$-manifold with a pointwise orientation, let $(Y_1,\ldots,Y_n)$ be a
local frame on $u$ that is positively oriented at $x \in u$, and let $(\varepsilon^1,\ldots,
\varepsilon^n)$ be any coframe dual to it at $x$, that is $\varepsilon^i(Y_k|_x) = \delta^i_k$.
Then
\[
dV_g|_x = \sqrt{\det\bigl(g_{ij}(x)\bigr)} \; \varepsilon^1 \wedge \cdots \wedge \varepsilon^n .
\]
In particular, for oriented local coordinates $(x^1,\ldots,x^n)$, taking $(Y_i)$ to be the
coordinate frame $(\partial/\partial x^i)$ and $\varepsilon^i = dx^i$ gives
$dV_g = \sqrt{\det(g_{ij})}\, dx^1 \wedge \cdots \wedge dx^n$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:volume-form-frame-value,lem:top-forms-proportional,def:wedge-covectors,lem:dual-coframe-duality}
Such a dual coframe exists, by Definition \ref{def:dual-coframe} and Lemma
\ref{lem:dual-coframe-duality}.

Both sides are $n$-forms on the $n$-dimensional space $T_x M$, and by Lemma
\ref{lem:top-forms-proportional} such a form is determined by its value on the single basis
$(Y_i|_x)$; so it suffices to check they agree there.

By Definition \ref{def:wedge-covectors},
\[
(\varepsilon^1 \wedge \cdots \wedge \varepsilon^n)(Y_1|_x,\ldots,Y_n|_x)
  = \det\bigl(\varepsilon^i(Y_j|_x)\bigr) = \det(\delta^i_j) = 1,
\]
so the right-hand side takes the value $\sqrt{\det(g_{ij}(x))}$ on $(Y_i|_x)$. The left-hand side
takes the same value by Proposition \ref{prop:volume-form-frame-value}.
\end{proof}

Lee's Proposition 2.41 collects the preceding results: its characterizations (a), (b) and (c) are
Propositions \ref{prop:volume-form-coframe}, \ref{prop:volume-form-value-one} and
\ref{prop:volume-form-coordinate-formula} respectively.

\begin{proposition}[The Riemannian Volume Form]
\label{prop:riemannian-volume-form}
\dcref{ch2:2.41}
\lean{LeeLib.Ch02.RiemannianMetric.riemannian_volumeForm}
\leanok
\uses{def:volume-form-pointwise,prop:volume-form-smooth,prop:volume-form-value-one,prop:volume-form-uniqueness,prop:volume-form-determinant,prop:volume-form-coframe,def:wedge-covectors,prop:volume-form-coordinate-formula}
Let $(M, g)$ be an oriented Riemannian $n$-manifold with or without boundary. There is a unique $n$-form $dV_g$ on
$M$, called the \emph{Riemannian volume form}, characterized by any one of the following
three equivalent properties:

\begin{enumerate}[label=(\alph*)]
\item If $(e^1, \ldots, e^n)$ is any local oriented orthonormal coframe for $T^* M$, then
$$dV_g = e^1 \wedge \cdots \wedge e^n.$$

\item If $(E_1, \ldots, E_n)$ is any local oriented orthonormal frame for $TM$, then
$$dV_g(E_1, \ldots, E_n) = 1.$$

\item If $(x^1, \ldots, x^n)$ are any oriented local coordinates, then
$$dV_g = \sqrt{\det(g_{ij})} \, dx^1 \wedge \cdots \wedge dx^n.$$
\end{enumerate}
\end{proposition}

\begin{proof}
\leanok
\uses{def:volume-form-pointwise,prop:volume-form-smooth,prop:volume-form-uniqueness,prop:volume-form-coframe,prop:volume-form-value-one,prop:volume-form-coordinate-formula}
The form $dV_g$ is the one of Definition \ref{def:volume-form-pointwise}, which is defined at each
point as the volume form of the oriented inner product space $T_x M$; it is a smooth $n$-form by
Proposition \ref{prop:volume-form-smooth}, and it is the unique $n$-form with property (b) by
Proposition \ref{prop:volume-form-uniqueness}, since an $n$-form on an $n$-dimensional space is
determined by its value on one basis.

Properties (a), (b) and (c) hold for it by Propositions \ref{prop:volume-form-coframe},
\ref{prop:volume-form-value-one} and \ref{prop:volume-form-coordinate-formula} respectively.

The three are equivalent because each pins the form down on a single basis: a form satisfying (a)
or (c) takes the value $1$ on an oriented orthonormal frame, since by Definition
\ref{def:wedge-covectors} a wedge of a dual coframe evaluates on the dual frame to
$\det(\delta^i_j) = 1$, and $\det(g_{ij}) = 1$ for an orthonormal frame; so each of (a) and (c)
implies (b), and (b) determines the form uniquely by Proposition
\ref{prop:volume-form-uniqueness}.
\end{proof}

The significance of the Riemannian volume form is that it allows us to integrate
functions on an oriented Riemannian manifold, not just differential forms. If $f$ is
a continuous, compactly supported real-valued function on an oriented Riemannian
$n$-manifold $(M, g)$ with or without boundary, then $f \, dV_g$ is a compactly supported
$n$-form. Therefore, the integral $\int_M f \, dV_g$ makes sense, and we define it to be the
\emph{integral of $f$ over $M$}. Similarly, if $M$ is compact, the \emph{volume of $M$} is defined to
be
$$\operatorname{Vol}(M) = \int_M dV_g = \int_M 1 \, dV_g.$$

In particular, if $D \subseteq M$ is a \emph{regular domain} (a closed, embedded codimension-$0$
submanifold with boundary), we can apply these definitions to $D$ with its induced
metric and thereby make sense of the integral of $f$ over $D$ and, in case $D$ is compact, the volume of $D$.

The notation $dV_g$ is chosen to emphasize the similarity of the integral $\int_M f \, dV_g$
with the standard integral of a function over an open subset of $\mathbb{R}^n$. It is \emph{not} meant
to imply that $dV_g$ is an exact form; in fact, if $M$ is a compact oriented manifold
without boundary, then $dV_g$ is never exact, because its integral over $M$ is positive, and exact forms integrate to zero by Stokes's theorem.

Because there are two conventions in common use for the wedge product (see p.~401), it should be noted that properties (a) and (c) of Proposition 2.41 are the same regardless of which convention is used; but property (b) holds only for the determinant convention that we use. If the Alt convention is used, the number 1 should be replaced by $1/n!$ in that formula.

Stating Exercise \ref{exer:isometry-preserves-volume-form} requires the pullback of a differential
form along a smooth map.

\begin{definition}[The Pullback of a Differential Form]
\label{def:pullback-alternating}
\lean{LeeLib.Ch02.pullbackAlternating}
\leanok
Let $\varphi : M \to \tilde{M}$ be a smooth map and $w$ a $k$-form on $\tilde{M}$. The
\emph{pullback} $\varphi^* w$ is the $k$-form on $M$ given at $x$ by
\[
(\varphi^* w)|_x(v_1,\ldots,v_k) = w|_{\varphi(x)}\bigl(d\varphi_x(v_1), \ldots, d\varphi_x(v_k)\bigr).
\]
\end{definition}

\begin{lemma}[A Metric-Preserving Map Pushes Orthonormal Frames Forward]
\label{lem:orthonormal-pushforward}
\lean{LeeLib.Ch02.IsMetricPreserving.orthonormal_pushforward}
\leanok
\uses{def:metric-preserving}
Let $\varphi : (M,g) \to (\tilde{M},\tilde{g})$ be metric-preserving and let $(E_i)$ be a
$g$-orthonormal frame on an open set $u$. Then for every $x \in u$ the family
$\bigl(d\varphi_x(E_i|_x)\bigr)$ is $\tilde{g}$-orthonormal.
\end{lemma}

\begin{proof}
\leanok
\uses{def:metric-preserving}
By the defining property of a metric-preserving map,
$\langle d\varphi_x v, d\varphi_x w \rangle_{\tilde{g}} = \langle v, w \rangle_g$ for all
$v,w \in T_x M$. Applying this to $v = E_i|_x$ and $w = E_j|_x$ gives
$\langle d\varphi_x E_i|_x, d\varphi_x E_j|_x \rangle_{\tilde{g}} = \langle E_i|_x, E_j|_x\rangle_g
= \delta_{ij}$.
\end{proof}

\begin{lemma}[The Pushforward Frame is a Basis]
\label{lem:push-basis}
\lean{LeeLib.Ch02.IsMetricPreserving.pushBasis}
\leanok
\uses{lem:orthonormal-pushforward}
In the situation of Lemma \ref{lem:orthonormal-pushforward}, the family
$\bigl(d\varphi_x(E_i|_x)\bigr)$ is a basis of $T_{\varphi(x)}\tilde{M}$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthonormal-pushforward}
An orthonormal family is linearly independent, so by Lemma \ref{lem:orthonormal-pushforward} the
family $\bigl(d\varphi_x(E_i|_x)\bigr)$ is linearly independent. It has $n$ members, and
$\dim T_{\varphi(x)}\tilde{M} = n$, so a linearly independent family of $n$ vectors in an
$n$-dimensional space spans, hence is a basis.
\end{proof}

\begin{exercise}\label{exer:isometry-preserves-volume-form}\dcref{ch2:2.42}
\lean{LeeLib.Ch02.IsMetricPreserving.pullbackAlternating_volumeForm}
\leanok
\uses{def:pullback-alternating,def:volume-form-pointwise,prop:volume-form-uniqueness,prop:volume-form-value-one,lem:orthonormal-pushforward,lem:push-basis}
Suppose $(M,g)$ and $(\tilde{M},\tilde{g})$ are oriented Riemannian manifolds, and $\varphi: M \to \tilde{M}$ is an orientation-preserving isometry. Prove that $\varphi^* dV_{\tilde{g}} = dV_g$.
\end{exercise}

\begin{proof}
\leanok
\uses{def:pullback-alternating,prop:volume-form-uniqueness,prop:volume-form-value-one,lem:orthonormal-pushforward,lem:push-basis}
Fix $x \in M$ and let $(E_1,\ldots,E_n)$ be a local oriented orthonormal frame near $x$. By
Definition \ref{def:pullback-alternating},
\[
(\varphi^* dV_{\tilde{g}})|_x(E_1|_x,\ldots,E_n|_x)
  = dV_{\tilde{g}}|_{\varphi(x)}\bigl(d\varphi_x E_1|_x, \ldots, d\varphi_x E_n|_x\bigr).
\]
By Lemmas \ref{lem:orthonormal-pushforward} and \ref{lem:push-basis} the family
$\bigl(d\varphi_x E_i|_x\bigr)$ is an orthonormal basis of $T_{\varphi(x)}\tilde{M}$, and it is
positively oriented because $\varphi$ is orientation-preserving. So by Proposition
\ref{prop:volume-form-value-one} the right-hand side equals $1$.

Thus $(\varphi^* dV_{\tilde{g}})|_x$ is an $n$-form taking the value $1$ on the oriented
orthonormal frame $(E_i|_x)$. By Proposition \ref{prop:volume-form-uniqueness} the only such
$n$-form is $dV_g|_x$, so $(\varphi^* dV_{\tilde{g}})|_x = dV_g|_x$. As $x$ was arbitrary,
$\varphi^* dV_{\tilde{g}} = dV_g$.
\end{proof}

For Riemannian hypersurfaces, we have the following important characterization of the volume form on the hypersurface in terms of that of the ambient manifold. If $X$ is a vector field and $\mu$ is a differential form, recall that $X \lrcorner \mu$ denotes \emph{interior multiplication} of $\mu$ by $X$ (see p.~401).

\begin{definition}[Interior Multiplication]
\label{def:interior-multiplication}
\lean{LeeLib.Ch02.camContract}
\leanok
Let $V$ be a real topological vector space and let $\mu$ be an alternating $(k+1)$-form on $V$.
For $x \in V$, the \emph{interior multiplication} $x \lrcorner \mu$ is the alternating $k$-form
\[
  (x \lrcorner \mu)(v_1, \ldots, v_k) = \mu(x, v_1, \ldots, v_k).
\]
It is alternating because $\mu$ is: if $v_i = v_j$ with $i \ne j$, then the tuple
$(x, v_1, \ldots, v_k)$ has two equal entries. It is continuous because $\mu$ is and
$(v_1,\ldots,v_k) \mapsto (x, v_1, \ldots, v_k)$ is.
\end{definition}

\begin{definition}[Unit Normal]
\label{def:unit-normal}
\lean{LeeLib.Ch02.IsUnitNormalAt}
\leanok
\uses{def:normal-space}
Let $(\tilde M, \tilde g)$ be a Riemannian manifold and let $F : M \to \tilde M$ present $M$ as a
smooth immersion. A family $N$ assigning to each $x \in M$ a vector $N_x \in T_{F(x)}\tilde M$ is
a \emph{unit normal} at $x$ if
\[
  \tilde g_{F(x)}(N_x, dF_x v) = 0 \quad\text{for every } v \in T_x M,
  \qquad \tilde g_{F(x)}(N_x, N_x) = 1,
\]
that is, if $N_x \in N_x M$ and $|N_x|_{\tilde g} = 1$.
\end{definition}

\begin{lemma}[The Adapted Frame Along a Hypersurface]
\label{lem:adapted-frame-cons-normal}
\lean{LeeLib.Ch02.inner_consNormal}
\leanok
\uses{def:unit-normal,def:metric-preserving,def:normal-space}
Let $(\tilde M, \tilde g)$ be a Riemannian $(n+1)$-manifold, let $F : M \to \tilde M$ present the
$n$-manifold $M$ as a smooth immersion, and let $g$ be a metric on $M$ that $F$ preserves. Let
$N$ be a unit normal at $x \in M$ and let $(E_1, \ldots, E_n)$ be a $g$-orthonormal basis of
$T_x M$. Then
\[
  \bigl(N_x, dF_x E_1, \ldots, dF_x E_n\bigr)
\]
is a $\tilde g$-orthonormal family in $T_{F(x)}\tilde M$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:unit-normal,def:metric-preserving}
There are four cases. First, $\tilde g(N_x, N_x) = 1$ because $N$ is a unit normal. Second,
$\tilde g(N_x, dF_x E_i) = 0$ for each $i$, because $N_x$ is normal and $dF_x E_i$ is tangent to
$M$; the same holds with the arguments exchanged, by symmetry of $\tilde g$. Third, for $i, j$,
\[
  \tilde g_{F(x)}(dF_x E_i, dF_x E_j) = g_x(E_i, E_j) = \delta_{ij},
\]
the first equality because $F$ preserves the metric and the second because $(E_i)$ is
orthonormal. So the family is orthonormal.
\end{proof}

\begin{lemma}[The Adapted Frame is an Ambient Basis]
\label{lem:adapted-frame-ambient-basis}
\lean{LeeLib.Ch02.consNormalBasis}
\leanok
\uses{lem:adapted-frame-cons-normal}
In the situation of Lemma \ref{lem:adapted-frame-cons-normal}, the family
$\bigl(N_x, dF_x E_1, \ldots, dF_x E_n\bigr)$ is a basis of $T_{F(x)}\tilde M$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:adapted-frame-cons-normal}
By Lemma \ref{lem:adapted-frame-cons-normal} the family is orthonormal, hence linearly
independent. It has $n+1$ members and $\dim T_{F(x)}\tilde M = n+1$, so a linearly independent
family of $n+1$ vectors in an $(n+1)$-dimensional space spans, hence is a basis.
\end{proof}

For Riemannian hypersurfaces, we have the following important characterization of the volume form on the hypersurface in terms of that of the ambient manifold.

\begin{proposition}[Equation (2.17)]
\label{prop:hypersurface-volume-form-formula}
\lean{LeeLib.Ch02.volumeForm_eq_normalRestrict}
\leanok
\uses{def:interior-multiplication,def:unit-normal,lem:adapted-frame-cons-normal,lem:adapted-frame-ambient-basis,def:volume-form-pointwise,prop:volume-form-uniqueness,prop:volume-form-value-one,def:metric-preserving}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold, let $F : M \to \tilde M$
present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on $M$ that $F$
preserves, and let $N$ be a unit normal at $x \in M$. Suppose $(E_1, \ldots, E_n)$ is a local
oriented $g$-orthonormal frame near $x$ for which the adapted basis
$\bigl(N_x, dF_x E_1, \ldots, dF_x E_n\bigr)$ of $T_{F(x)}\tilde M$ is positively oriented. Then
\begin{equation}
dV_g|_x = \bigl((N \lrcorner dV_{\tilde{g}})|_M\bigr)_x, \tag{2.17}
\end{equation}
where the right-hand side is the $n$-form $(v_1, \ldots, v_n) \mapsto
dV_{\tilde g}|_{F(x)}(N_x, dF_x v_1, \ldots, dF_x v_n)$ on $T_x M$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:interior-multiplication,lem:adapted-frame-cons-normal,lem:adapted-frame-ambient-basis,prop:volume-form-uniqueness,prop:volume-form-value-one}
Evaluate the right-hand side on the frame $(E_1, \ldots, E_n)$: by definition of interior
multiplication and of the restriction, it equals
\[
  dV_{\tilde g}|_{F(x)}\bigl(N_x, dF_x E_1, \ldots, dF_x E_n\bigr).
\]
By Lemmas \ref{lem:adapted-frame-cons-normal} and \ref{lem:adapted-frame-ambient-basis} the
family $\bigl(N_x, dF_x E_1, \ldots, dF_x E_n\bigr)$ is an orthonormal basis of
$T_{F(x)}\tilde M$, and it is positively oriented by hypothesis. So by Proposition
\ref{prop:volume-form-value-one} the value is $1$.

Thus the right-hand side is an $n$-form on $T_x M$ taking the value $1$ on the oriented
orthonormal frame $(E_i|_x)$. By Proposition \ref{prop:volume-form-uniqueness} the only such
$n$-form is $dV_g|_x$, which is (2.17).
\end{proof}

The orientation hypothesis of Proposition \ref{prop:hypersurface-volume-form-formula} is not an
extra assumption on $M$: a unit normal produces, with no choices, the orientation that satisfies
it. The construction rests on the observation that an orientation of an $n$-dimensional space
\emph{is} a ray of nonzero $n$-forms, so a nonvanishing $n$-form orients the space outright.

\begin{lemma}[A Nonvanishing Top Form Orients the Space]
\label{lem:nonvanishing-top-form-orientation}
\lean{LeeLib.Ch02.rayOfNeZero_eq_basis_orientation_iff}
\leanok
Let $V$ be an $n$-dimensional real vector space and let $\omega$ be a nonzero alternating
$n$-form on $V$. Then $\omega$ determines an orientation of $V$, namely the ray of $\omega$ in
the one-dimensional space of $n$-forms, and a basis $(b_1, \ldots, b_n)$ of $V$ is positively
oriented for it if and only if $\omega(b_1, \ldots, b_n) > 0$.
\end{lemma}

\begin{proof}
\leanok
Every $n$-form on $V$ is a scalar multiple of the determinant form $\det{}_b$ of $b$; applying
the relation $\omega = c \cdot \det_b$ to $b$ itself and using $\det_b(b) = 1$ identifies the
scalar as $c = \omega(b)$, and $c \ne 0$ because $\omega \ne 0$. The orientation determined by
$b$ is the ray of $\det_b$, so the ray of $\omega = \omega(b) \cdot \det_b$ equals it exactly
when the scalar $\omega(b)$ is positive.
\end{proof}

\begin{definition}[The Orientation Induced by a Unit Normal]
\label{def:induced-orientation}
\lean{LeeLib.Ch02.inducedOrientation}
\leanok
\uses{def:unit-normal,def:interior-multiplication,lem:nonvanishing-top-form-orientation,lem:adapted-frame-ambient-basis,def:volume-form-pointwise}
In the situation of Proposition \ref{prop:hypersurface-volume-form-formula}, let $N$ be a unit
normal at every point of $M$. The \emph{induced orientation} of $M$ assigns to each $x \in M$ the
orientation of $T_x M$ determined, by Lemma \ref{lem:nonvanishing-top-form-orientation}, by the
$n$-form $\bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x$.

This is well defined: the form does not vanish, since its value on a $g$-orthonormal basis
$(E_i)$ of $T_x M$ is $dV_{\tilde g}(N_x, dF E_1, \ldots, dF E_n) = \pm 1$, the adapted family
being an orthonormal basis of $T_{F(x)}\tilde M$ by Lemma
\ref{lem:adapted-frame-ambient-basis}.
\end{definition}

\begin{lemma}[The Induced Orientation is Lee's]
\label{lem:induced-orientation-characterization}
\lean{LeeLib.Ch02.consNormalBasis_orientation_eq_iff}
\leanok
\uses{def:induced-orientation,lem:adapted-frame-ambient-basis,lem:nonvanishing-top-form-orientation,prop:volume-form-value-one}
Let $(E_1, \ldots, E_n)$ be a $g$-orthonormal basis of $T_x M$. Then $(E_i)$ is positively
oriented for the induced orientation of $M$ if and only if the adapted basis
$\bigl(N_x, dF E_1, \ldots, dF E_n\bigr)$ is positively oriented in $\tilde M$.

That is, the induced orientation is exactly the one Lee describes.
\end{lemma}

\begin{proof}
\leanok
\uses{def:induced-orientation,lem:nonvanishing-top-form-orientation,prop:volume-form-value-one}
By Lemma \ref{lem:nonvanishing-top-form-orientation}, $(E_i)$ is positively oriented for the
induced orientation if and only if
\[
  \bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1, \ldots, E_n)
    = dV_{\tilde g}\bigl(N_x, dF E_1, \ldots, dF E_n\bigr) > 0.
\]
The adapted family is an orthonormal basis of $T_{F(x)}\tilde M$, so by Proposition
\ref{prop:volume-form-value-one} the right-hand side is $1$ if that basis is positively oriented
and $-1$ if it is not. So the value is positive exactly when the adapted basis is positively
oriented.
\end{proof}

\begin{lemma}[Top Forms are Determined by One Basis]
\label{lem:top-form-eq-of-basis}
\lean{LeeLib.Ch02.eq_of_apply_basis}
\leanok
Let $V$ be an $n$-dimensional real vector space and let $b$ be a basis of $V$. If two alternating
$n$-forms on $V$ agree on $b$, they are equal.
\end{lemma}

\begin{proof}
\leanok
Each $n$-form $\theta$ satisfies $\theta = \theta(b) \cdot \det_b$, so two forms agreeing on $b$
are the same multiple of $\det_b$.
\end{proof}

\begin{proposition}[Equation (2.17) for the Induced Orientation]
\label{prop:hypersurface-volume-form-induced}
\lean{LeeLib.Ch02.volumeForm_inducedOrientation_eq_normalRestrict}
\leanok
\uses{def:induced-orientation,lem:induced-orientation-characterization,lem:top-form-eq-of-basis,prop:volume-form-value-one,def:interior-multiplication,def:unit-normal}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold, let $F : M \to \tilde M$
present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on $M$ that $F$
preserves, and let $N$ be a unit normal at every point of $M$. Then, for the induced orientation
of $M$,
\[
  dV_g = (N \lrcorner dV_{\tilde{g}})|_M
\]
at every point, with no compatibility hypothesis between the orientations of $M$ and $\tilde M$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:induced-orientation,lem:induced-orientation-characterization,lem:top-form-eq-of-basis,prop:volume-form-value-one}
Fix $x \in M$ and choose a $g$-orthonormal basis $(E_i)$ of $T_x M$. By Lemma
\ref{lem:top-form-eq-of-basis} it suffices to see that both sides agree on $(E_i)$.

Suppose first that $(E_i)$ is positively oriented for the induced orientation. Then
$dV_g(E_1,\ldots,E_n) = 1$ by Proposition \ref{prop:volume-form-value-one}, and by Lemma
\ref{lem:induced-orientation-characterization} the adapted basis is positively oriented in
$\tilde M$, so $(N \lrcorner dV_{\tilde g})|_M$ also takes the value $1$ on $(E_i)$.

Otherwise $(E_i)$ is negatively oriented, and $dV_g(E_1,\ldots,E_n) = -1$; by Lemma
\ref{lem:induced-orientation-characterization} the adapted basis is then negatively oriented in
$\tilde M$, so $(N \lrcorner dV_{\tilde g})|_M$ takes the value $-1$ on $(E_i)$ as well. In both
cases the two forms agree on $(E_i)$, hence are equal.
\end{proof}

\begin{lemma}[Smoothness of a Pulled-Back Section]
\label{lem:pullback-comp-section}
\lean{LeeLib.Ch02.contMDiffAt_comp_section}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian}
Let $F : M \to \tilde M$ be a smooth map and let $Z$ be a smooth section of $T\tilde M$ defined
near $F(x_0)$. Then $x \mapsto Z_{F(x)}$ is a smooth section of the ambient tangent bundle
$F^* T\tilde M$ near $x_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian}
The fibre of $F^* T\tilde M$ at $x$ is $T_{F(x)}\tilde M$, and a local trivialization of
$F^* T\tilde M$ near $x_0$ is by definition a local trivialization of $T\tilde M$ near $F(x_0)$
precomposed with $F$. Read in these trivializations the section in question is $Z$, read in the
trivialization of $T\tilde M$, composed with $F$. Both are smooth, hence so is the composite.
\end{proof}

\begin{definition}[Smoothness of an Ambient Family at a Point]
\label{def:smooth-normal-at}
\lean{LeeLib.Ch02.IsSmoothNormalAt}
\leanok
\uses{lem:ambient-tangent-bundle-riemannian}
Let $F : M \to \tilde M$ present $M$ as a smooth immersion in a Riemannian manifold
$(\tilde M, \tilde g)$, and let $N$ assign to each $x \in M$ a vector $N_x \in T_{F(x)}\tilde M$.
Then $N$ \emph{varies smoothly at} $x_0 \in M$ if it is a smooth section of the ambient tangent
bundle $F^* T\tilde M$ at $x_0$.

The notion is pointwise rather than global because the normal fields it is applied to are in
general only defined and smooth on a neighbourhood.
\end{definition}

\begin{definition}[Smooth Unit Normal Field]
\label{def:smooth-unit-normal-field}
\lean{LeeLib.Ch02.IsSmoothNormalField}
\leanok
\uses{def:unit-normal,def:smooth-normal-at,lem:ambient-tangent-bundle-riemannian,def:normal-space}
In the situation of Definition \ref{def:smooth-normal-at}, the family $N$ \emph{varies smoothly}
if it varies smoothly at every point of $M$. A \emph{smooth global unit normal field} for $M$ is
such a family that is in addition a unit normal at every point of $M$.
\end{definition}

\begin{lemma}[Negating One Member of a Local Frame]
\label{lem:frame-neg-member}
\lean{LeeLib.Ch02.isLocalFrameOn_negMember}
\leanok
Let $(Y_i)_{i \in \iota}$ be a smooth local frame for a smooth vector bundle $V \to B$ over an
open set $u \subseteq B$, and fix an index $i_0 \in \iota$. Then the family obtained from $(Y_i)$
by replacing $Y_{i_0}$ with $-Y_{i_0}$ is again a smooth local frame for $V$ over $u$.
\end{lemma}

\begin{proof}
\leanok
The modified family is smooth because negation is a smooth bundle map. At each $x \in u$ it is
obtained from the basis $(Y_i|_x)$ of $V_x$ by scaling one member by the unit $-1$, so it is
linearly independent. Finally $-Y_{i_0}|_x$ and $Y_{i_0}|_x$ span the same line, so the modified
family spans the same subspace as $(Y_i|_x)$, namely all of $V_x$. Being independent and
spanning at every $x \in u$, it is a local frame.
\end{proof}

\begin{lemma}[The Adapted Frame Varies Smoothly]
\label{lem:adapted-frame-smooth}
\lean{LeeLib.Ch02.contMDiffAt_consNormal}
\leanok
\uses{def:smooth-normal-at,lem:pushforward-smooth,def:unit-normal,lem:ambient-tangent-bundle-riemannian}
Let $F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion in the
$(n+1)$-manifold $\tilde M$, let $N$ be a family of ambient vectors along $M$ that varies
smoothly, and let $(E_1, \ldots, E_n)$ be a smooth local frame for $TM$ over an open set $u$.
Then every member of the adapted family $\bigl(N, dF E_1, \ldots, dF E_n\bigr)$ is a smooth
section of $F^* T\tilde M$ over $u$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:smooth-normal-at,lem:pushforward-smooth}
The zeroth member is $N$, which is a smooth section by hypothesis. Each remaining member is a
pushforward $dF(E_i)$ of a smooth vector field, hence a smooth section of $F^* T\tilde M$ by
Lemma \ref{lem:pushforward-smooth}.
\end{proof}

\begin{lemma}[Smoothness of the Value on a Frame]
\label{lem:normal-restrict-frame-smooth}
\lean{LeeLib.Ch02.contMDiffAt_normalRestrict_frame}
\leanok
\uses{def:smooth-normal-at,lem:adapted-frame-smooth,prop:oriented-orthonormal-frames,prop:volume-form-determinant,def:interior-multiplication,def:smooth-orientation,lem:ambient-tangent-bundle-riemannian,lem:pullback-comp-section}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold, let $F : M \to \tilde M$
present the $n$-manifold $M$ as a smooth immersion, let $N$ be a family of ambient vectors along
$M$ that varies smoothly, and let $(E_1, \ldots, E_n)$ be a smooth local frame for $TM$ over an
open set $u$. Then
\[
  x \longmapsto \bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1|_x, \ldots, E_n|_x)
    = dV_{\tilde g}|_{F(x)}\bigl(N_x, dF_x E_1|_x, \ldots, dF_x E_n|_x\bigr)
\]
is a smooth real-valued function on $u$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:adapted-frame-smooth,prop:oriented-orthonormal-frames,prop:volume-form-determinant,lem:pullback-comp-section,lem:ambient-tangent-bundle-riemannian}
Smoothness is local, so fix $x_0 \in u$. By Proposition \ref{prop:oriented-orthonormal-frames}
there is an open neighbourhood $w$ of $F(x_0)$ in $\tilde M$ carrying a smooth local frame
$(Z_0, \ldots, Z_n)$ for $T\tilde M$ that is orthonormal and positively oriented at every point
of $w$. Put $v = u \cap F^{-1}(w)$, an open neighbourhood of $x_0$.

For $x \in v$ the frame $(Z_a)$ is orthonormal and positively oriented at $F(x)$, so Proposition
\ref{prop:volume-form-determinant} gives, for any $v_0, \ldots, v_n \in T_{F(x)}\tilde M$,
\[
  dV_{\tilde g}|_{F(x)}(v_0, \ldots, v_n)
    = \det\bigl[\tilde g_{F(x)}(Z_a|_{F(x)}, v_b)\bigr]_{a,b}.
\]
Applying this to the adapted family $(W_0, \ldots, W_n) = \bigl(N, dF E_1, \ldots, dF E_n\bigr)$,
the function in question equals, on $v$,
\[
  x \longmapsto \det\bigl[\tilde g_{F(x)}\bigl(Z_a|_{F(x)}, W_b(x)\bigr)\bigr]_{a,b}.
\]
Each $x \mapsto Z_a|_{F(x)}$ is a smooth section of $F^* T\tilde M$ by Lemma
\ref{lem:pullback-comp-section}, and each $W_b$ is one by Lemma \ref{lem:adapted-frame-smooth}.
The metric of $F^* T\tilde M$ varies smoothly by Lemma
\ref{lem:ambient-tangent-bundle-riemannian}, so each matrix entry is a smooth function of
$x \in v$. A determinant is a polynomial in its entries, so the displayed function is smooth on
$v$, and in particular at $x_0$.
\end{proof}

\begin{lemma}[Negating a Frame Member Negates the Value]
\label{lem:normal-restrict-neg-member}
\lean{LeeLib.Ch02.normalRestrict_negMember}
\leanok
\uses{def:interior-multiplication,def:unit-normal}
In the situation of Proposition \ref{prop:hypersurface-volume-form-formula}, let
$(E_1, \ldots, E_n)$ be a family of vectors in $T_x M$ and let $i_0$ be an index. Then
\[
  \bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1, \ldots, -E_{i_0}, \ldots, E_n)
    = - \bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1, \ldots, E_{i_0}, \ldots, E_n).
\]
\end{lemma}

\begin{proof}
\leanok
$\bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x$ is an alternating $n$-form on $T_x M$, hence
linear in each argument separately.
\end{proof}

\begin{proposition}[The Induced Orientation is Smooth]
\label{prop:hypersurface-orientable-of-normal}
\lean{LeeLib.Ch02.isSmoothOrientation_inducedOrientation}
\leanok
\uses{def:induced-orientation,def:smooth-unit-normal-field,def:smooth-orientation,lem:normal-restrict-frame-smooth,lem:frame-neg-member,lem:normal-restrict-neg-member,lem:nonvanishing-top-form-orientation,prop:existence-orthonormal-frames,def:interior-multiplication,lem:adapted-frame-ambient-basis,lem:induced-orientation-characterization}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold with $n \geq 1$, let
$F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on
$M$ that $F$ preserves, and let $N$ be a smooth global unit normal field for $M$. Then the
induced orientation of $M$ is smooth. In particular, $M$ is orientable.
\end{proposition}

\begin{proof}
\leanok
\uses{def:induced-orientation,lem:normal-restrict-frame-smooth,lem:frame-neg-member,lem:normal-restrict-neg-member,lem:nonvanishing-top-form-orientation,prop:existence-orthonormal-frames}
Fix $p \in M$. By Proposition \ref{prop:existence-orthonormal-frames} there is an open
neighbourhood $u$ of $p$ carrying a smooth $g$-orthonormal local frame $(E_1, \ldots, E_n)$ for
$TM$. Consider
\[
  s(x) = \bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1|_x, \ldots, E_n|_x),
  \qquad x \in u .
\]
By Lemma \ref{lem:normal-restrict-frame-smooth} the function $s$ is smooth on $u$, hence
continuous. It never vanishes: by Definition \ref{def:induced-orientation} the form
$\bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x$ is a nonzero alternating $n$-form on $T_x M$, and
a nonzero alternating $n$-form on an $n$-dimensional space does not vanish on a basis, $(E_i|_x)$
being one. Moreover, by Lemma \ref{lem:nonvanishing-top-form-orientation}, $(E_i|_x)$ induces the
induced orientation of $M$ at $x$ if and only if $s(x) > 0$.

Since $s$ is continuous and $s(p) \neq 0$, exactly one of the following holds.

If $s(p) > 0$, then $u' = u \cap s^{-1}\bigl((0,\infty)\bigr)$ is an open neighbourhood of $p$, and
on it $(E_i)$ is a smooth local frame inducing the induced orientation at every point.

If $s(p) < 0$, then $u' = u \cap s^{-1}\bigl((-\infty,0)\bigr)$ is an open neighbourhood of $p$.
Because $n \geq 1$ there is an index $i_0$, and by Lemma \ref{lem:frame-neg-member} the family
$(E_i')$ obtained from $(E_i)$ by replacing $E_{i_0}$ with $-E_{i_0}$ is again a smooth local
frame over $u'$. By Lemma \ref{lem:normal-restrict-neg-member} its value satisfies
$\bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1'|_x, \ldots, E_n'|_x) = -s(x) > 0$ for every
$x \in u'$, so $(E_i')$ induces the induced orientation at every point of $u'$.

In both cases $p$ has an open neighbourhood carrying a smooth local frame that is positively
oriented for the induced orientation throughout. As $p$ was arbitrary, the induced orientation is
smooth.
\end{proof}

\begin{remark}
\label{rem:hypersurface-orientation-dimension}
\uses{prop:hypersurface-orientable-of-normal,def:smooth-orientation,def:induced-orientation}
The hypothesis $n \geq 1$ in Proposition \ref{prop:hypersurface-orientable-of-normal} cannot be
dropped. A smooth orientation is by Definition \ref{def:smooth-orientation} one realized near each
point by a positively oriented smooth local frame; a $0$-manifold admits only the empty frame,
whose orientation is the ray of the constant $1$. But $T_x M$ carries two orientations even when
$n = 0$, and for a $0$-manifold immersed in a $1$-manifold the induced orientation is the ray of
the number $dV_{\tilde g}(N_x) = \pm 1$, which may be $-1$. In that case the induced orientation
is not realized by any frame, though $M$ is of course still orientable.
\end{remark}

\begin{proposition}[Hypersurface Volume Form: the Sufficiency Half]
\label{prop:hypersurface-volume-form-if}
\lean{LeeLib.Ch02.exists_smoothOrientation_volumeForm_eq_normalRestrict}
\leanok
\uses{prop:hypersurface-orientable-of-normal,prop:hypersurface-volume-form-induced,def:induced-orientation,def:smooth-unit-normal-field,def:smooth-orientation,def:interior-multiplication}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold with $n \geq 1$, let
$F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on
$M$ that $F$ preserves, and suppose $M$ admits a smooth global unit normal field $N$. Then $M$ is
orientable, and it carries a smooth orientation for which
\[
  dV_g = (N \lrcorner dV_{\tilde{g}})|_M. \tag{2.17}
\]
\end{proposition}

\begin{proof}
\leanok
\uses{prop:hypersurface-orientable-of-normal,prop:hypersurface-volume-form-induced}
Take the induced orientation of $M$. It is smooth by Proposition
\ref{prop:hypersurface-orientable-of-normal}, so $M$ is orientable, and (2.17) holds for it by
Proposition \ref{prop:hypersurface-volume-form-induced}.
\end{proof}

Proposition \ref{prop:hypersurface-volume-form-if} produces an orientation from a normal. The
converse produces a normal from an orientation, and rests on the fact that a hypersurface has
\emph{exactly two} unit normals at each point, of which the orientation selects one.

\begin{lemma}[A Hypersurface Has Exactly Two Unit Normals at Each Point]
\label{lem:two-unit-normals}
\lean{LeeLib.Ch02.eq_or_eq_neg_of_isUnitNormalVecAt}
\leanok
\uses{def:unit-normal,lem:normal-space-orthonormal-frame,def:metric-preserving}
Let $F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion in the Riemannian
$(n+1)$-manifold $(\tilde M, \tilde g)$, let $x \in M$, and suppose the normal space $N_x M$ is
the line spanned by a unit normal $w_0$. Then every unit normal at $x$ is $w_0$ or $-w_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:unit-normal}
A unit normal $w$ lies in $N_x M = \mathbb{R} \cdot w_0$, so $w = c\, w_0$ for some
$c \in \mathbb{R}$. Then
\[
  1 = \langle w, w \rangle_{\tilde g} = c^2 \langle w_0, w_0 \rangle_{\tilde g} = c^2,
\]
so $c = \pm 1$.
\end{proof}

\begin{lemma}[The Normal Space of a Hypersurface Is a Line]
\label{lem:pointwise-unit-normal-span}
\lean{LeeLib.Ch02.exists_isUnitNormalVecAt_span}
\leanok
\uses{def:unit-normal,lem:normal-space-orthonormal-frame}
In the situation of Lemma \ref{lem:two-unit-normals}, every $x \in M$ admits a unit normal $w_0$
with $N_x M = \mathbb{R} \cdot w_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:normal-space-orthonormal-frame}
By Lemma \ref{lem:normal-space-orthonormal-frame} some neighbourhood of $x$ carries
$\dim \tilde M - \dim M$ smooth sections whose values at each of its points form an orthonormal
basis of the normal space. Here $\dim \tilde M - \dim M = 1$, so there is exactly one such
section; its value $w_0$ at $x$ is a unit normal and spans $N_x M$.
\end{proof}

\begin{lemma}[Negating the Normal Negates the Contracted Form]
\label{lem:normal-restrict-neg-normal}
\lean{LeeLib.Ch02.normalRestrictVec_neg}
\leanok
\uses{def:interior-multiplication,def:unit-normal,def:volume-form-pointwise}
In the situation of Lemma \ref{lem:two-unit-normals}, for every $w \in T_{F(x)}\tilde M$,
\[
  \bigl(((-w) \lrcorner dV_{\tilde g})|_M\bigr)_x = -\bigl((w \lrcorner dV_{\tilde g})|_M\bigr)_x .
\]
\end{lemma}

\begin{proof}
\leanok
The contracted form places $w$ in the first slot of $dV_{\tilde g}$, and $dV_{\tilde g}$ is
multilinear, so negating that slot negates the value. No orthogonality, unit length or continuity
is used.
\end{proof}

\begin{lemma}[A Basis Realizing a Prescribed Orientation]
\label{lem:basis-realizing-orientation}
\lean{LeeLib.Ch02.exists_basis_orientation_eq}
\leanok
\uses{lem:nonvanishing-top-form-orientation}
Let $M$ be an $n$-manifold with $n \geq 1$, let $o$ be a pointwise orientation of $M$ and let
$x \in M$. Then $T_x M$ has a basis that is positively oriented for $o(x)$.
\end{lemma}

\begin{proof}
\leanok
Take any basis of $T_x M$. Its orientation is either $o(x)$, in which case it serves, or the
opposite one, in which case negating one of its members — possible because $n \geq 1$ — reverses
its orientation and produces the required basis.
\end{proof}

\begin{definition}[A Unit Normal Inducing a Given Orientation]
\label{def:oriented-unit-normal}
\lean{LeeLib.Ch02.IsOrientedUnitNormalVecAt}
\leanok
\uses{def:unit-normal,def:interior-multiplication,def:induced-orientation,lem:nonvanishing-top-form-orientation}
Let $o$ be a pointwise orientation of $M$ and $x \in M$. A unit normal $w$ at $x$
\emph{induces $o$ at $x$} if
$\bigl((w \lrcorner dV_{\tilde g})|_M\bigr)_x(b_1,\ldots,b_n) > 0$ for every basis
$(b_1,\ldots,b_n)$ of $T_x M$ that is positively oriented for $o(x)$.
\end{definition}

\begin{lemma}[One Basis Suffices]
\label{lem:oriented-unit-normal-iff-pos}
\lean{LeeLib.Ch02.isOrientedUnitNormalVecAt_iff_pos}
\leanok
\uses{def:oriented-unit-normal,lem:nonvanishing-top-form-orientation,def:unit-normal,def:metric-preserving,lem:adapted-frame-ambient-basis}
In the situation of Definition \ref{def:oriented-unit-normal}, let $b$ be one basis of $T_x M$
that is positively oriented for $o(x)$. Then the unit normal $w$ induces $o$ at $x$ if and only if
$\bigl((w \lrcorner dV_{\tilde g})|_M\bigr)_x(b) > 0$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nonvanishing-top-form-orientation}
The form $\bigl((w \lrcorner dV_{\tilde g})|_M\bigr)_x$ is a nonzero $n$-form, so by Lemma
\ref{lem:nonvanishing-top-form-orientation} its positivity on $b$ says exactly that its ray equals
the orientation determined by $b$, which is $o(x)$. That is a statement about rays and mentions no
basis, so by the same lemma it implies positivity on every other basis positively oriented for
$o(x)$.
\end{proof}

\begin{proposition}[An Orientation Selects a Unit Normal]
\label{prop:existsunique-oriented-unit-normal}
\lean{LeeLib.Ch02.existsUnique_isOrientedUnitNormalVecAt}
\leanok
\uses{def:oriented-unit-normal,lem:two-unit-normals,lem:pointwise-unit-normal-span,lem:normal-restrict-neg-normal,lem:basis-realizing-orientation,lem:oriented-unit-normal-iff-pos,def:unit-normal,def:metric-preserving}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold with $n \geq 1$, let
$F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on
$M$ that $F$ preserves, and let $o$ be a pointwise orientation of $M$. Then each $x \in M$ has
exactly one unit normal inducing $o$ at $x$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:two-unit-normals,lem:pointwise-unit-normal-span,lem:normal-restrict-neg-normal,lem:basis-realizing-orientation,lem:oriented-unit-normal-iff-pos}
By Lemma \ref{lem:pointwise-unit-normal-span} there is a unit normal $w_0$ at $x$ spanning
$N_x M$, and by Lemma \ref{lem:two-unit-normals} the only unit normals at $x$ are $\pm w_0$. By
Lemma \ref{lem:basis-realizing-orientation} fix a basis $b$ of $T_x M$ positively oriented for
$o(x)$, and set $t = \bigl((w_0 \lrcorner dV_{\tilde g})|_M\bigr)_x(b)$. As in Proposition
\ref{prop:hypersurface-orientable-of-normal}, $t \neq 0$, because a nonzero $n$-form does not
vanish on a basis.

If $t > 0$ then $w_0$ induces $o$ at $x$ by Lemma \ref{lem:oriented-unit-normal-iff-pos}, and
$-w_0$ does not, since by Lemma \ref{lem:normal-restrict-neg-normal} its value on $b$ is $-t < 0$.
If $t < 0$ the roles are exchanged. Either way exactly one of $\pm w_0$ induces $o$ at $x$.
\end{proof}

\begin{definition}[The Unit Normal Determined by an Orientation]
\label{def:unit-normal-of-orientation}
\lean{LeeLib.Ch02.unitNormalOfOrientation}
\leanok
\uses{prop:existsunique-oriented-unit-normal,def:oriented-unit-normal}
In the situation of Proposition \ref{prop:existsunique-oriented-unit-normal}, the
\emph{unit normal determined by $o$} assigns to each $x \in M$ the unique unit normal at $x$
inducing $o$ at $x$. This is a global normal field, defined with no choices beyond the sign that
$o$ itself fixes.
\end{definition}

\begin{proposition}[The Normal Determined by an Orientation Is Smooth]
\label{prop:unit-normal-of-orientation-smooth}
\lean{LeeLib.Ch02.isSmoothNormalField_unitNormalOfOrientation}
\leanok
\uses{def:unit-normal-of-orientation,def:smooth-unit-normal-field,def:smooth-orientation,lem:normal-space-orthonormal-frame,lem:normal-restrict-frame-smooth,lem:oriented-unit-normal-iff-pos,lem:normal-restrict-neg-normal,prop:existsunique-oriented-unit-normal,def:interior-multiplication}
In the situation of Proposition \ref{prop:existsunique-oriented-unit-normal}, suppose in addition
that $o$ is a smooth orientation of $M$ and that the orientation of $\tilde M$ is smooth. Then the
unit normal determined by $o$ is a smooth global unit normal field.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:normal-space-orthonormal-frame,lem:normal-restrict-frame-smooth,lem:oriented-unit-normal-iff-pos,lem:normal-restrict-neg-normal,prop:existsunique-oriented-unit-normal}
Write $N$ for the unit normal determined by $o$ and fix $p \in M$. Because $o$ is smooth, some
open $u \ni p$ carries a smooth local frame $(E_1, \ldots, E_n)$ of $TM$ that is positively
oriented for $o$ at every point of $u$. By Lemma \ref{lem:normal-space-orthonormal-frame} some
open $v \ni p$ carries a smooth local unit normal $W$. Put $c = u \cap v$ and
\[
  s(x) = \bigl((W \lrcorner dV_{\tilde g})|_M\bigr)_x(E_1|_x, \ldots, E_n|_x), \qquad x \in c .
\]
By Lemma \ref{lem:normal-restrict-frame-smooth} $s$ is smooth on $c$, hence continuous, and it
never vanishes there, a nonzero $n$-form not vanishing on a basis.

By Lemma \ref{lem:oriented-unit-normal-iff-pos}, $W_x$ induces $o$ at $x \in c$ exactly when
$s(x) > 0$; and by Lemma \ref{lem:normal-restrict-neg-normal} the value of the form on $(E_i|_x)$
for $-W_x$ is $-s(x)$, so $-W_x$ induces $o$ at $x$ exactly when $s(x) < 0$. Since $s(p) \neq 0$,
one of these two cases holds on the open neighbourhood $c \cap s^{-1}\bigl((0,\infty)\bigr)$ or
$c \cap s^{-1}\bigl((-\infty,0)\bigr)$ of $p$; by the uniqueness in Proposition
\ref{prop:existsunique-oriented-unit-normal}, $N = W$ on the first and $N = -W$ on the second.
Both $W$ and $-W$ are smooth on $v$, so $N$ is smooth at $p$. As $p$ was arbitrary, $N$ is a
smooth global unit normal field.
\end{proof}

\begin{proposition}[Hypersurface Volume Form: the Necessity Half]
\label{prop:hypersurface-volume-form-only-if}
\lean{LeeLib.Ch02.exists_isSmoothNormalField_volumeForm_eq_normalRestrict}
\leanok
\uses{prop:unit-normal-of-orientation-smooth,def:unit-normal-of-orientation,prop:hypersurface-volume-form-induced,def:induced-orientation,lem:oriented-unit-normal-iff-pos,lem:basis-realizing-orientation,def:smooth-unit-normal-field,def:smooth-orientation,def:interior-multiplication}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold with $n \geq 1$, let
$F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion, let $g$ be a metric on
$M$ that $F$ preserves, and suppose $M$ carries a smooth orientation $o$. Then $M$ admits a
smooth global unit normal field $N$, and for it
\[
  dV_g = (N \lrcorner dV_{\tilde{g}})|_M. \tag{2.17}
\]
\end{proposition}

\begin{proof}
\leanok
\uses{prop:unit-normal-of-orientation-smooth,prop:hypersurface-volume-form-induced,lem:oriented-unit-normal-iff-pos,lem:basis-realizing-orientation}
Let $N$ be the unit normal determined by $o$; it is a smooth global unit normal field by
Proposition \ref{prop:unit-normal-of-orientation-smooth}. Its induced orientation is $o$ itself:
at each $x$, pick by Lemma \ref{lem:basis-realizing-orientation} a basis $b$ of $T_x M$ positively
oriented for $o(x)$; then $\bigl((N \lrcorner dV_{\tilde g})|_M\bigr)_x(b) > 0$ because $N_x$
induces $o$ at $x$, so by Lemma \ref{lem:nonvanishing-top-form-orientation} the ray of that form
— which is by definition the induced orientation at $x$ — equals the orientation of $b$, namely
$o(x)$. Now (2.17) is Proposition \ref{prop:hypersurface-volume-form-induced} applied to $N$.
\end{proof}

\begin{proposition}
\label{prop:hypersurface-volume-form}
\dcref{ch2:2.43}
\lean{LeeLib.Ch02.exists_isSmoothOrientation_iff_exists_isSmoothNormalField}
\leanok
\uses{prop:hypersurface-volume-form-formula,prop:hypersurface-volume-form-induced,prop:hypersurface-volume-form-if,prop:hypersurface-volume-form-only-if,prop:hypersurface-orientable-of-normal,prop:unit-normal-of-orientation-smooth,def:induced-orientation,def:smooth-unit-normal-field,def:unit-normal,def:interior-multiplication,lem:normal-space-orthonormal-frame,lem:induced-orientation-characterization}
Let $(\tilde M, \tilde g)$ be an oriented Riemannian $(n+1)$-manifold with $n \geq 1$, let
$F : M \to \tilde M$ present the $n$-manifold $M$ as a smooth immersion, and let $g$ be a metric
on $M$ that $F$ preserves — for instance the induced metric. Then $M$ is orientable if and only
if there exists a smooth global unit normal vector field $N$ for $M$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:hypersurface-volume-form-if,prop:hypersurface-volume-form-only-if,prop:hypersurface-orientable-of-normal,prop:unit-normal-of-orientation-smooth,def:induced-orientation,def:smooth-unit-normal-field}
If $N$ is a smooth global unit normal field, then the orientation it induces (Definition
\ref{def:induced-orientation}) is smooth by Proposition
\ref{prop:hypersurface-orientable-of-normal}, so $M$ is orientable. Conversely, if $M$ carries a
smooth orientation $o$, then the unit normal determined by $o$ is a smooth global unit normal
field by Proposition \ref{prop:unit-normal-of-orientation-smooth}.
\end{proof}

In either direction the volume form is the contraction of the ambient one against the normal:
equation (2.17),
\[
  dV_g = (N \lrcorner dV_{\tilde g})|_M ,
\]
holds for the orientation and normal that the respective direction produces. That is
Proposition \ref{prop:hypersurface-volume-form-if} for a given normal and Proposition
\ref{prop:hypersurface-volume-form-only-if} for a given orientation.

When $M$ is not orientable, we can still define integrals of functions, but now we have to use \emph{densities} instead of differential forms (see pp.~405--406).

\subsection{The density of an inner product space}

A density on a vector space is the unoriented analogue of a top-degree form: it assigns an
unsigned volume to an $n$-tuple of vectors, and consequently transforms by $|\det|$ rather than
by $\det$. Because no sign is involved, no orientation is needed anywhere in this subsection,
and that is precisely what makes Proposition \ref{prop:riemannian-density} available on every
Riemannian manifold rather than only on the orientable ones.

\begin{definition}[Density of a Vector Space]
\label{def:density-vector-space}
\lean{LeeLib.Ch02.IsDensity}
\leanok
Let $V$ be an $n$-dimensional real vector space. A \emph{density} on $V$ is a function
$\mu : V^n \to \mathbb{R}$ such that
\begin{equation}
\mu(f v_1, \ldots, f v_n) = |\det f| \, \mu(v_1, \ldots, v_n)
\tag{B.14}
\end{equation}
for every linear map $f : V \to V$ and all $v_1, \ldots, v_n \in V$. A density is
\emph{positive} if $\mu(v_1, \ldots, v_n) > 0$ whenever $(v_1, \ldots, v_n)$ is a basis of $V$.
\end{definition}

\begin{definition}[The Density of an Inner Product Space]
\label{def:density-pointwise}
\lean{LeeLib.Ch02.densityL}
\leanok
Let $V$ be an $n$-dimensional real inner product space. For $v_1, \ldots, v_n \in V$ set
\[
|v_1 \wedge \cdots \wedge v_n| := \sqrt{\det \bigl( \langle v_i, v_j \rangle \bigr)_{i,j=1}^{n}},
\]
the square root of the determinant of the Gram matrix of $(v_1, \ldots, v_n)$. The Gram matrix
is positive semidefinite, so its determinant is nonnegative and the square root is defined.
\end{definition}

\begin{proposition}[The Determinant Formula for the Density]
\label{prop:density-abs-det}
\lean{LeeLib.Ch02.densityL_eq_abs_det}
\leanok
\uses{def:density-pointwise}
Let $V$ be an $n$-dimensional real inner product space, let $(e_1, \ldots, e_n)$ be any
orthonormal basis of $V$, and let $v_1, \ldots, v_n \in V$. Then
\[
|v_1 \wedge \cdots \wedge v_n| = \bigl| \det \bigl( \langle e_i, v_j \rangle \bigr)_{i,j=1}^{n} \bigr|.
\]
\end{proposition}

\begin{proof}
\leanok
\uses{def:density-pointwise}
Write $M = (\langle e_i, v_j \rangle)_{i,j}$. Expanding each $v_j$ in the orthonormal basis gives
$v_j = \sum_k \langle e_k, v_j \rangle e_k$, so
\[
\langle v_i, v_j \rangle
= \Bigl\langle \sum_k \langle e_k, v_i \rangle e_k, \; \sum_l \langle e_l, v_j \rangle e_l \Bigr\rangle
= \sum_k \langle e_k, v_i \rangle \langle e_k, v_j \rangle,
\]
by orthonormality of $(e_k)$; that is, the Gram matrix of $(v_1, \ldots, v_n)$ equals
$M^{\mathsf{T}} M$. Hence
$\det (\langle v_i, v_j \rangle) = \det(M^{\mathsf{T}}) \det(M) = (\det M)^2$, and taking the
square root gives $\sqrt{(\det M)^2} = |\det M|$.
\end{proof}

\begin{proposition}[The Density of an Orthonormal Basis]
\label{prop:density-value-one}
\lean{LeeLib.Ch02.densityL_apply_orthonormalBasis}
\leanok
\uses{prop:density-abs-det}
Let $V$ be an $n$-dimensional real inner product space and let $(e_1, \ldots, e_n)$ be an
orthonormal basis of $V$. Then $|e_1 \wedge \cdots \wedge e_n| = 1$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:density-abs-det}
Apply Proposition \ref{prop:density-abs-det} with $v_i = e_i$. The matrix
$(\langle e_i, e_j \rangle)_{i,j}$ is the identity matrix by orthonormality, and its determinant
is $1$, so the density is $|1| = 1$.
\end{proof}

\begin{proposition}[The Transformation Law]
\label{prop:density-transformation-law}
\lean{LeeLib.Ch02.densityL_comp}
\leanok
\uses{prop:density-abs-det,def:density-vector-space}
Let $V$ be an $n$-dimensional real inner product space, let $f : V \to V$ be linear and let
$v_1, \ldots, v_n \in V$. Then
\[
|f v_1 \wedge \cdots \wedge f v_n| = |\det f| \cdot |v_1 \wedge \cdots \wedge v_n|.
\]
In particular $|{\cdot} \wedge \cdots \wedge {\cdot}|$ is a density on $V$ in the sense of
Definition \ref{def:density-vector-space}.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:density-abs-det}
Fix an orthonormal basis $(e_1, \ldots, e_n)$ of $V$ and let $A = (\langle e_i, f e_k \rangle)_{i,k}$
be the matrix of $f$ in that basis, so that $\det A = \det f$. Expanding $v_j$ in the basis and
using linearity of $f$,
\[
\langle e_i, f v_j \rangle
= \Bigl\langle e_i, \; f \Bigl( \sum_k \langle e_k, v_j \rangle e_k \Bigr) \Bigr\rangle
= \sum_k \langle e_i, f e_k \rangle \langle e_k, v_j \rangle,
\]
so the matrix of $(f v_1, \ldots, f v_n)$ against $(e_i)$ is $A \cdot M$, where
$M = (\langle e_i, v_j \rangle)_{i,j}$. By Proposition \ref{prop:density-abs-det} applied twice,
\[
|f v_1 \wedge \cdots \wedge f v_n| = |\det(AM)| = |\det A| \, |\det M|
= |\det f| \cdot |v_1 \wedge \cdots \wedge v_n|. \qedhere
\]
\end{proof}

\begin{proposition}[Positivity of the Density]
\label{prop:density-positive}
\lean{LeeLib.Ch02.densityL_pos}
\leanok
\uses{def:density-pointwise}
Let $V$ be an $n$-dimensional real inner product space and let $(v_1, \ldots, v_n)$ be linearly
independent. Then $|v_1 \wedge \cdots \wedge v_n| > 0$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:density-pointwise}
The Gram matrix of a linearly independent family is positive definite: for $c \neq 0$ the vector
$w = \sum_j c_j v_j$ is nonzero, and $c^{\mathsf{T}} (\langle v_i, v_j \rangle) c = \langle w, w \rangle > 0$.
A positive definite matrix has positive determinant, so the square root defining
$|v_1 \wedge \cdots \wedge v_n|$ is a square root of a positive number.
\end{proof}

\begin{proposition}[Fibrewise Uniqueness of the Density]
\label{prop:density-uniqueness-pointwise}
\lean{LeeLib.Ch02.densityL_unique}
\leanok
\uses{def:density-vector-space,prop:density-transformation-law,prop:density-value-one}
Let $V$ be an $n$-dimensional real inner product space, let $(e_1, \ldots, e_n)$ be an orthonormal
basis of $V$, and let $\mu$ be a density on $V$ with $\mu(e_1, \ldots, e_n) = 1$. Then
$\mu(v_1, \ldots, v_n) = |v_1 \wedge \cdots \wedge v_n|$ for all $v_1, \ldots, v_n \in V$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:density-transformation-law,prop:density-value-one}
Let $v_1, \ldots, v_n \in V$ and let $f : V \to V$ be the unique linear map with $f e_i = v_i$ for
every $i$. By the transformation law (B.14) for $\mu$ and the hypothesis $\mu(e_1,\ldots,e_n) = 1$,
\[
\mu(v_1, \ldots, v_n) = \mu(f e_1, \ldots, f e_n) = |\det f| \, \mu(e_1, \ldots, e_n) = |\det f|.
\]
By Proposition \ref{prop:density-transformation-law} and Proposition \ref{prop:density-value-one}
the same computation gives
$|v_1 \wedge \cdots \wedge v_n| = |\det f| \cdot |e_1 \wedge \cdots \wedge e_n| = |\det f|$.
The two agree.
\end{proof}

\begin{proposition}[The Density Is the Absolute Value of the Volume Form]
\label{prop:density-eq-abs-volume-form}
\lean{LeeLib.Ch02.densityL_eq_abs_volumeForm}
\leanok
\uses{def:density-pointwise,prop:volume-form-determinant}
Let $V$ be an $n$-dimensional real inner product space with $n > 0$, let $o$ be an orientation of
$V$, and let $\omega$ be the volume form of $(V, o)$. Then
$|v_1 \wedge \cdots \wedge v_n| = |\omega(v_1, \ldots, v_n)|$ for all $v_1, \ldots, v_n \in V$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:density-abs-det}
Choose an orthonormal basis $(e_1, \ldots, e_n)$ of $V$ inducing $o$; this is possible because
$n > 0$. The determinant formula for the volume form gives
$\omega(v_1, \ldots, v_n) = \det (\langle e_i, v_j \rangle)_{i,j}$, while Proposition
\ref{prop:density-abs-det} gives
$|v_1 \wedge \cdots \wedge v_n| = |\det (\langle e_i, v_j \rangle)_{i,j}|$. Taking absolute values
of the former yields the latter.
\end{proof}

\subsection{Smooth densities and the Riemannian density}

\begin{definition}[Smooth Density]
\label{def:smooth-density}
\lean{LeeLib.Ch02.SmoothDensity}
\leanok
\uses{def:density-vector-space}
Let $M$ be a smooth $n$-manifold. A \emph{smooth density} on $M$ is a family $\mu$ assigning to
each $x \in M$ a density $\mu_x$ on $T_x M$, such that for every smooth local frame
$(Y_1, \ldots, Y_n)$ on an open set $u \subseteq M$ the function
$x \mapsto \mu_x(Y_1|_x, \ldots, Y_n|_x)$ is smooth on $u$.
\end{definition}

\begin{definition}[The Riemannian Density, Pointwise]
\label{def:riemannian-density-pointwise}
\lean{LeeLib.Ch02.RiemannianMetric.densityAt}
\leanok
\uses{def:density-pointwise,def:riemannian-metric}
Let $(M,g)$ be a Riemannian $n$-manifold. For $x \in M$ and $v_1, \ldots, v_n \in T_x M$ let
\[
\mu_x(v_1, \ldots, v_n) := \sqrt{\det \bigl( \langle v_i, v_j \rangle_g \bigr)_{i,j=1}^{n}}
\]
be the density of the inner product space $(T_x M, g_x)$ of Definition
\ref{def:density-pointwise}. No orientation of $M$ enters this definition.
\end{definition}

\begin{proposition}[Smoothness of the Riemannian Density]
\label{prop:density-smooth}
\lean{LeeLib.Ch02.RiemannianMetric.contMDiffOn_densityAt}
\leanok
\uses{def:riemannian-density-pointwise,prop:density-positive,lem:inner-product-vector-fields-smooth}
Let $(M,g)$ be a Riemannian $n$-manifold and let $(Y_1, \ldots, Y_n)$ be a smooth local frame on
an open set $u \subseteq M$. Then $x \mapsto \mu_x(Y_1|_x, \ldots, Y_n|_x)$ is smooth on $u$; that
is, $\mu$ is a smooth density.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-density-pointwise,prop:density-positive,lem:inner-product-vector-fields-smooth}
The entries $x \mapsto \langle Y_i|_x, Y_j|_x \rangle_g$ of the Gram matrix are smooth on $u$,
being pairings of smooth vector fields under $g$. A determinant is a polynomial in the matrix
entries, so $x \mapsto \det (\langle Y_i|_x, Y_j|_x \rangle_g)$ is smooth on $u$. At every $x \in u$
the vectors $(Y_1|_x, \ldots, Y_n|_x)$ are linearly independent, so this determinant is strictly
positive by Proposition \ref{prop:density-positive}; since the square root is smooth away from
$0$, the composite $x \mapsto \sqrt{\det (\langle Y_i|_x, Y_j|_x \rangle_g)}$ is smooth on $u$.
\end{proof}

\begin{proposition}[The Riemannian Density]\label{prop:riemannian-density}\dcref{ch2:2.44}
\lean{LeeLib.Ch02.riemannian_density_existsUnique}
\leanok
\uses{def:smooth-density,def:riemannian-density-pointwise,prop:density-smooth,prop:density-transformation-law,prop:density-value-one,prop:density-uniqueness-pointwise,prop:density-positive,prop:existence-orthonormal-frames}
If $(M,g)$ is any Riemannian manifold, then there is a unique smooth positive density $\mu$ on $M$, called the \emph{Riemannian density}, with the property that
\begin{equation}
\mu(E_1,\ldots,E_n) = 1 \tag{2.18}
\end{equation}
for every local orthonormal frame $(E_i)$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-density-pointwise,prop:density-smooth,prop:density-transformation-law,prop:density-value-one,prop:density-uniqueness-pointwise,prop:density-positive,prop:existence-orthonormal-frames}
\emph{Existence.} Take $\mu$ to be the family of Definition \ref{def:riemannian-density-pointwise}.
Each $\mu_x$ is a density on $T_x M$ by Proposition \ref{prop:density-transformation-law}, and
$\mu$ is smooth by Proposition \ref{prop:density-smooth}, so $\mu$ is a smooth density. If
$(E_1, \ldots, E_n)$ is a local orthonormal frame on $u$ then $(E_1|_x, \ldots, E_n|_x)$ is an
orthonormal basis of $(T_x M, g_x)$ for each $x \in u$, so $\mu_x(E_1|_x, \ldots, E_n|_x) = 1$ by
Proposition \ref{prop:density-value-one}; this is (2.18). It is positive by Proposition
\ref{prop:density-positive}, since the values of a frame are linearly independent.

\emph{Uniqueness.} Let $\nu$ be a smooth density satisfying (2.18) and let $x \in M$. By
Proposition \ref{prop:existence-orthonormal-frames} there are a neighbourhood $u$ of $x$ and a
local orthonormal frame $(E_1, \ldots, E_n)$ on $u$. Then $(E_1|_x, \ldots, E_n|_x)$ is an
orthonormal basis of $T_x M$ and $\nu_x(E_1|_x, \ldots, E_n|_x) = 1$ by (2.18). Since $\nu_x$ is a
density on $T_x M$ taking the value $1$ on an orthonormal basis, Proposition
\ref{prop:density-uniqueness-pointwise} gives $\nu_x = \mu_x$. As $x$ was arbitrary, $\nu = \mu$.
\end{proof}

\begin{exercise}\label{exer:riemannian-density-formula}\dcref{ch2:2.45}
\uses{prop:riemannian-density}
Prove this proposition by showing that $\mu$ can be defined in terms of any local orthonormal frame by
\[
\mu = |\varepsilon^1 \wedge \cdots \wedge \varepsilon^n|.
\]
\end{exercise}

Let $(M,g)$ be a Riemannian manifold (with or without boundary). If $M$ is oriented and $dV_g$ is its Riemannian volume form, then its Riemannian density is easily seen to be equal to $|dV_g|$. On the other hand, the Riemannian density is defined whether $M$ is oriented or not. It is customary to denote the Riemannian density by the same notation $dV_g$ that we use for the Riemannian volume form, and to specify when necessary whether the notation refers to a density or a form. In either case, we can define the integral of a compactly supported smooth function $f: M \to \mathbb{R}$ as $\int_M f\, dV_g$. This is to be interpreted as the integral of a density when $M$ is nonorientable; when $M$ is orientable, it can be interpreted either as the integral of a density
or as the integral of an $n$-form (with respect to some choice of orientation), because
both give the same result.

\section{The Divergence and the Laplacian}

In advanced calculus, you have undoubtedly been introduced to three important
differential operators involving vector fields on $\mathbb{R}^3$: the \emph{gradient} (which takes real-valued functions to vector fields), \emph{divergence} (vector fields to functions), and \emph{curl}
(vector fields to vector fields). We have already described how the gradient operator
can be generalized to Riemannian manifolds (see equation (2.14)); now we can show
that the divergence operator also generalizes easily to that setting. Problem 2-27
describes a similar, but more limited, generalization of the curl.

Suppose $(M, g)$ is an oriented Riemannian $n$-manifold with or without boundary,
and $dV_g$ is its volume form. If $X$ is a smooth vector field on $M$, then $X \lrcorner dV_g$ is
an $(n-1)$-form. The exterior derivative of this $(n-1)$-form is a smooth $n$-form, so
it can be expressed as a smooth function multiplied by $dV_g$. That function is called
the \emph{divergence of } $X$, and denoted by $\operatorname{div} X$; thus it is characterized by the following
formula:
\begin{equation}
d(X \lrcorner dV_g) = (\operatorname{div} X)dV_g. \tag{2.19}
\end{equation}

Even if $M$ is nonorientable, in a neighborhood of each point we can choose an
orientation and define the divergence by (2.19), and then note that reversing the orientation changes the sign of $dV_g$ on both sides of the equation, so $\operatorname{div} X$ is well
defined, independently of the choice of orientation. In this way, we can define the
divergence operator on \emph{any} Riemannian manifold with or without boundary, by
requiring that it satisfy (2.19) for any choice of orientation in a neighborhood of
each point.

The most important application of the divergence operator is the \emph{divergence theorem}, which you will be asked to prove in Problem 2-22.

Using the divergence operator, we can define another important operator, this one
acting on real-valued functions. The \emph{Laplacian} (or \emph{Laplace--Beltrami operator}) is
the linear operator $\Delta: C^\infty(M) \to C^\infty(M)$ defined by
\begin{equation}
\Delta u = \operatorname{div}(\operatorname{grad} u). \tag{2.20}
\end{equation}

(Note that many books, including [LeeSM] and the first edition of this book, define
the Laplacian as $-\operatorname{div}(\operatorname{grad} u)$. The main reason for choosing the negative sign is
so that the operator will have nonnegative eigenvalues; see Problem 2-24. But the
definition we give here is much more common in Riemannian geometry.)

The next proposition gives alternative formulas for these operators.

\begin{proposition}
\label{prop:coordinate-representations-divergence-laplacian}
\dcref{ch2:2.46}

Let $(M, g)$ be a Riemannian manifold with or without boundary,
and let $(x^i)$ be any smooth local coordinates on an open set $U \subseteq M$. The coordinate
representations of the divergence and Laplacian are as follows:
\[
\operatorname{div}\left(X^i \frac{\partial}{\partial x^i}\right) = \frac{1}{\sqrt{\det g}} \frac{\partial}{\partial x^i}\left(\sqrt{\det g}\, X^i\right),
\]
\[
\Delta u = \frac{1}{\sqrt{\det g}} \frac{\partial}{\partial x^i}\left(\sqrt{\det g}\, g^{ij} \frac{\partial u}{\partial x^j}\right),
\]
where $\det g = \det(g_{kl})$ is the determinant of the component matrix of $g$ in these coordinates. On $\mathbb{R}^n$ with the Euclidean metric and standard coordinates, these reduce to

\[
\operatorname{div}\left(X^i \frac{\partial}{\partial x^i}\right) = \sum_{i=1}^n \frac{\partial X^i}{\partial x^i},
\]

\[
\Delta u = \sum_{i=1}^n \frac{\partial^2 u}{(\partial x^i)^2}.
\]
\end{proposition}

\textit{Proof.} Problem 2-21. \hfill $\square$

\section{Lengths and Distances}

Perhaps the most important tool that a Riemannian metric gives us is the ability to measure lengths of curves and distances between points. Throughout this section, $(M, g)$ denotes a Riemannian manifold with or without boundary.

Without further qualification, a \textbf{curve} in $M$ always means a \textbf{parametrized curve}, that is, a continuous map $\gamma: I \to M$, where $I \subseteq \mathbb{R}$ is some interval. Unless otherwise specified, we will not worry about whether the interval is bounded or unbounded, or whether it includes endpoints or not. To say that $\gamma$ is a \textbf{smooth curve} is to say that it is smooth as a map from the manifold (with boundary) $I$ to $M$. If $I$ has one or two endpoints and $M$ has empty boundary, then $\gamma$ is smooth if and only if it extends to a smooth curve defined on some open interval containing $I$. (If $\partial M \neq \emptyset$, then smoothness of $\gamma$ has to be interpreted as meaning that each coordinate representation of $\gamma$ has a smooth extension to an open interval.) A \textbf{curve segment} is a curve whose domain is a compact interval.

A smooth curve $\gamma: I \to M$ has a well-defined velocity $\gamma'(t) \in T_{\gamma(t)} M$ for each $t \in I$. We say that $\gamma$ is a \textbf{regular curve} if $\gamma'(t) \neq 0$ for $t \in I$. This implies that $\gamma$ is an immersion, so its image has no \textquotedblleft corners\textquotedblright\ or \textquotedblleft kinks.\textquotedblright

We wish to use curve segments as \textquotedblleft measuring tapes\textquotedblright\ to define distances between points in a connected Riemannian manifold. Many aspects of the theory become technically much simpler if we work with a slightly larger class of curve segments instead of just the regular ones. We now describe the appropriate class of curves.

If $[a, b] \subseteq \mathbb{R}$ is a closed bounded interval, a \textbf{partition of $[a,b]$} is a finite sequence $(a_0, \ldots, a_k)$ of real numbers such that $a = a_0 < a_1 < \cdots < a_k = b$. Each interval $[a_{i-1}, a_i]$ is called a \textbf{subinterval of the partition}. If $M$ is a smooth manifold with or without boundary, a (continuous) curve segment $\gamma: [a, b] \to M$ is said to be
\emph{piecewise regular} if there exists a partition $(a_0,\ldots,a_k)$ of $[a,b]$ such that $\gamma|_{[a_{i-1},a_i]}$ is a regular curve segment (meaning it is smooth with nonvanishing velocity) for $i=1,\ldots,k$. For brevity, we refer to a piecewise regular curve segment as an \textbf{admissible curve}, and any partition $(a_0,\ldots,a_k)$ such that $\gamma|_{[a_{i-1},a_i]}$ is smooth for each $i$ an \textbf{admissible partition for $\gamma$}. (There are many admissible partitions for a given admissible curve, because we can always add more points to the partition.) It is also convenient to consider any map $\gamma: \{a\} \to M$ whose domain is a single real number to be an admissible curve.

Suppose $\gamma$ is an admissible curve and $(a_0,\ldots,a_k)$ is an admissible partition for it. At each of the intermediate partition points $a_1,\ldots,a_{k-1}$, there are two one-sided velocity vectors, which we denote by
\[
\gamma'(a_i^-) = \lim_{t \nearrow a_i} \gamma'(t),
\]
\[
\gamma'(a_i^+) = \lim_{t \searrow a_i} \gamma'(t).
\]
They are both nonzero, but they need not be equal.

If $\gamma: I \to M$ is a smooth curve, we define a \emph{reparametrization} of $\gamma$ to be a curve of the form $\tilde{\gamma} = \gamma \circ \varphi: I' \to M$, where $I' \subseteq \mathbb{R}$ is another interval and $\varphi: I' \to I$ is a diffeomorphism. Because intervals are connected, $\varphi$ is either strictly increasing or strictly decreasing on $I'$. We say that $\tilde{\gamma}$ is a \emph{forward reparametrization} if $\varphi$ is increasing, and a \emph{backward reparametrization} if it is decreasing.

For an admissible curve $\gamma: [a,b] \to M$, we define a \emph{reparametrization of $\gamma$} a little more broadly, as a curve of the form $\tilde{\gamma} = \gamma \circ \varphi$, where $\varphi: [c,d] \to [a,b]$ is a homeomorphism for which there is a partition $(c_0,\ldots,c_k)$ of $[c,d]$ such that the restriction of $\varphi$ to each subinterval $[c_{i-1},c_i]$ is a diffeomorphism onto its image.

If $\gamma: [a,b] \to M$ is an admissible curve, we define the \emph{length of $\gamma$} to be
\[
L_g(\gamma) = \int_a^b |\gamma'(t)|_g \, dt.
\]
The integrand is bounded and continuous everywhere on $[a,b]$ except possibly at the finitely many points where $\gamma$ is not smooth, so this integral is well defined.

\begin{definition}[The Length of a Curve, extended form]
\label{def:length-of-curve}
\lean{Manifold.pathELength}
\mathlibok
\uses{def:riemannian-metric}
For a curve $\gamma$ into $M$ and $a \le b$, the \emph{length} of $\gamma$ on $[a,b]$ is
\[
  L_g(\gamma|_{[a,b]}) \;=\; \int_{[a,b]} |\gamma'(t)|_g \, dt \;\in\; [0,\infty],
\]
the integral being taken in the Lebesgue sense, so that it is defined for every curve and takes values in $[0,\infty]$. For an admissible curve on a compact interval the integrand is bounded, so the value is finite and agrees with $L_g(\gamma)$ above.
\end{definition}

\begin{proposition}[Additivity of Length]
\label{prop:length-additive}
\lean{Manifold.pathELength_add}
\mathlibok
\uses{def:length-of-curve}
\dcref{ch2:2.47}
Suppose $(M,g)$ is a Riemannian manifold with or without boundary and $\gamma$ is an admissible curve. If $a \le c \le b$, then
\[
  L_g(\gamma|_{[a,c]}) + L_g(\gamma|_{[c,b]}) = L_g(\gamma|_{[a,b]}).
\]
\end{proposition}

\begin{proposition}[Parameter Independence of Length]
\label{prop:length-reparam-invariant}
\lean{Manifold.pathELength_comp_of_monotoneOn}
\mathlibok
\uses{def:length-of-curve}
\dcref{ch2:2.47}
Suppose $(M,g)$ is a Riemannian manifold with or without boundary and $\gamma$ is an admissible curve on $[a,b]$. If $\tilde{\gamma} = \gamma \circ \varphi$ is a forward reparametrization of $\gamma$, then $L_g(\tilde{\gamma}) = L_g(\gamma)$.
\end{proposition}

\begin{proposition}[Isometry Invariance of Length]
\label{prop:length-isometry-invariant}
\lean{LeeLib.Ch02.IsMetricPreserving.pathELength_comp}
\leanok
\uses{def:length-of-curve,def:metric-preserving}
\dcref{ch2:2.47}
Suppose $(M,g)$ and $(\tilde{M},\tilde{g})$ are Riemannian manifolds with or without boundary, $\varphi: M \to \tilde{M}$ is a smooth metric-preserving map, and $\gamma : [a,b] \to M$ is an admissible curve. Then
\[
  L_{\tilde{g}}(\varphi \circ \gamma) = L_g(\gamma).
\]
In particular this holds when $\varphi$ is a local isometry, since a local isometry preserves the metric.
\end{proposition}

\begin{proof}
\leanok
\uses{def:metric-preserving,def:length-of-curve}
Since $\varphi$ preserves the metric, each differential $d\varphi_p$ is a linear isometry, so for every $p \in M$ and $v \in T_p M$,
\[
  |d\varphi_p(v)|_{\tilde{g}}^2 = \tilde{g}_{\varphi(p)}(d\varphi_p v, d\varphi_p v) = g_p(v,v) = |v|_g^2 .
\]
At every $t$ in the open interval $(a,b)$ the curve $\gamma$ is differentiable, so the chain rule gives $(\varphi \circ \gamma)'(t) = d\varphi_{\gamma(t)}(\gamma'(t))$, and therefore
\[
  |(\varphi \circ \gamma)'(t)|_{\tilde{g}} = |d\varphi_{\gamma(t)}(\gamma'(t))|_{\tilde{g}} = |\gamma'(t)|_g .
\]
The two integrands agree on $(a,b)$, which differs from $[a,b]$ by a set of measure zero, so the integrals agree.
\end{proof}

Lee's Proposition 2.47 collects three properties of the length functional: additivity, parameter independence, and isometry invariance. They are recorded above as Propositions \ref{prop:length-additive}, \ref{prop:length-reparam-invariant} and \ref{prop:length-isometry-invariant} respectively, each stated separately because each is proved by a different argument: the first two are properties of the integral defining the length and hold on any manifold whose tangent spaces are normed, while only the third refers to a metric at all.

\begin{exercise}[Prove Proposition 2.47]
\label{exer:prove-prop-247}
\leanok
\dcref{ch2:2.48}
\uses{prop:length-additive,prop:length-reparam-invariant,prop:length-isometry-invariant}
\end{exercise}

Suppose $\gamma : [a, b] \to M$ is an admissible curve. The \textbf{arc-length function} of $\gamma$ is the function $s : [a, b] \to \mathbb{R}$ defined by
\[s(t) = L_g(\gamma|_{[a,t]}) = \int_a^t |\gamma'(u)|_g \, du.\]

It is continuous everywhere, and it follows from the fundamental theorem of calculus that it is smooth wherever $\gamma$ is, with derivative $s'(t) = |\gamma'(t)|$.

For this reason, if $\gamma: I \to M$ is any smooth curve (not necessarily a curve segment), we define the \textbf{speed} of $\gamma$ at any time $t \in I$ to be the scalar $|\gamma'(t)|$. We say that $\gamma$ is a \textbf{unit-speed curve} if $|\gamma'(t)| = 1$ for all $t$, and a \textbf{constant-speed curve} if $|\gamma'(t)|$ is constant. If $\gamma$ is a piecewise smooth curve, we say that $\gamma$ has \textbf{unit speed} if $|\gamma'(t)| = 1$ wherever $\gamma$ is smooth.

If $\gamma : [a, b] \to M$ is a unit-speed admissible curve, then its arc-length function has the simple form $s(t) = t - a$. If, in addition, its parameter interval is of the form $[0, b]$ for some $b > 0$, then the arc-length function is $s(t) = t$. For this reason, a unit-speed admissible curve whose parameter interval is of the form $[0, b]$ is said to be \textbf{parametrized by arc length}.

\begin{proposition}
\label{prop:reparametrization-by-arc-length}
\dcref{ch2:2.49}
Suppose $(M, g)$ is a Riemannian manifold with or without boundary.
\begin{enumerate}
\item[(a)] Every regular curve in $M$ has a unit-speed forward reparametrization.
\item[(b)] Every admissible curve in $M$ has a unique forward reparametrization by arc length.
\end{enumerate}
\end{proposition}

\begin{proof}
Suppose $\gamma : I \to M$ is a regular curve. Choose an arbitrary $t_0 \in I$, and define $s : I \to \mathbb{R}$ by
\[s(t) = \int_{t_0}^t |\gamma'(u)|_g \, du.\]

Since $s'(t) = |\gamma'(t)|_g > 0$, it follows that $s$ is a strictly increasing local diffeomorphism, and thus maps $I$ diffeomorphically onto an interval $I' \subseteq \mathbb{R}$. If we let $\varphi = s^{-1} : I' \to I$, then $\tilde{\gamma} = \gamma \circ \varphi$ is a forward reparametrization of $\gamma$, and the chain rule gives
\[|\tilde{\gamma}'(t)|_g = |\varphi'(t)\gamma'(\varphi(t))|_g = \frac{1}{s'(\varphi(t))}|\gamma'(\varphi(t))|_g = 1.\]

Thus $\tilde{\gamma}$ is a unit-speed reparametrization of $\gamma$.

Now let $\gamma : [a, b] \to M$ be an admissible curve. We prove the existence statement in part (b) by induction on the number of smooth segments in an admissible partition. If $\gamma$ has only one smooth segment, then it is a regular curve segment, and (b) follows by applying (a) in the special case $I = [a, b]$ and choosing $t_0 = a$. Assuming that the result is true for admissible curves with $k$ smooth segments, suppose $\gamma$
has an admissible partition $(a_0, \ldots, a_{k+1})$. The inductive hypothesis gives piecewise regular homeomorphisms $\varphi : [0,c] \to [a, a_k]$ and $\psi : [0,d] \to [a_k, b]$ such that $\gamma \circ \varphi$ is an arc-length reparametrization of $\gamma|_{[a,a_k]}$ and $\gamma \circ \psi$ is an arc-length reparametrization of $\gamma|_{[a_k,b]}$. If we define $\bar{\varphi}: [0, c+d] \to [a,b]$ by
\[
\bar{\varphi}(s) = \begin{cases}
\varphi(s), & s \in [0,c], \\
\psi(s-c), & s \in [c, c+d],
\end{cases}
\]
then $\gamma \circ \bar{\varphi}$ is a reparametrization of $\gamma$ by arc length.

To prove uniqueness, suppose that $\tilde{\gamma} = \gamma \circ \varphi$ and $\hat{\gamma} = \gamma \circ \psi$ are both forward reparametrizations of $\gamma$ by arc length. Since both have the same length, it follows that $\varphi$ and $\psi$ both have the same parameter domain $[0,c]$, and thus both are piecewise regular homeomorphisms from $[0,c]$ to $[a,b]$. If we define $\eta = \varphi^{-1} \circ \psi: [0,c] \to [0,c]$, then $\eta$ is a piecewise regular increasing homeomorphism satisfying $\tilde{\gamma} = \gamma \circ \varphi \circ \eta = \tilde{\gamma} \circ \eta$. The fact that both $\tilde{\gamma}$ and $\hat{\gamma}$ are of unit speed implies the following equality for all $s \in [0,c]$ except the finitely many values at which $\tilde{\gamma}$ or $\eta$ is not smooth:
\[
1 = |\dot{\tilde{\gamma}}(s)|_g = |\dot{\tilde{\gamma}}(\eta(s))\eta'(s)|_g = |\dot{\tilde{\gamma}}(\eta(s))|_g \eta'(s) = \eta'(s).
\]

Since $\eta$ is continuous and $\eta(0) = 0$, it follows that $\eta(s) = s$ for all $s \in [0,c]$, and thus $\tilde{\gamma} = \hat{\gamma}$.
\end{proof}

\section{The Riemannian Distance Function}

We are now in a position to introduce one of the most important concepts from classical geometry into the Riemannian setting: distances between points.

Suppose $(M, g)$ is a connected Riemannian manifold with or without boundary. For each pair of points $p, q \in M$, we define the \textbf{Riemannian distance from $p$ to $q$}, denoted by $d_g(p,q)$, to be the infimum of the lengths of all admissible curves from $p$ to $q$. The following proposition guarantees that $d_g(p,q)$ is a well-defined nonnegative real number for each $p, q \in M$.

\begin{proposition}
\label{prop:connected-manifold-admissible-curve}
\lean{LeeLib.Ch02.exists_contMDiffOn_path_of_preconnected}
\leanok
\uses{prop:existence-riemannian-metric,prop:riemannian-distance-finite,def:riemannian-distance}
\dcref{ch2:2.50}
If $M$ is a connected smooth manifold (with or without boundary), then any two points of $M$ can be joined by an admissible curve.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:existence-riemannian-metric,prop:riemannian-distance-finite,def:riemannian-distance}
Lee argues directly: since a connected manifold is path-connected, $p$ and $q$ are joined by a continuous path $c: [a,b] \to M$; by compactness there is a partition of $[a,b]$ such that each $c([a_{i-1}, a_i])$ lies in a single smooth coordinate ball (or half-ball if $\partial M \neq \emptyset$), and replacing each segment by a straight-line path in coordinates yields an admissible curve between the same points (Fig.~2.2).

The following route reaches the same conclusion from results already established. The statement mentions no metric, but nothing prevents introducing one: by Proposition \ref{prop:existence-riemannian-metric} choose a Riemannian metric $g$ on $M$. By Proposition \ref{prop:riemannian-distance-finite}, $d_g(p,q) < \infty$, so $d_g(p,q) < d_g(p,q) + 1$. Since $d_g(p,q)$ is by definition the infimum of the lengths of admissible curves from $p$ to $q$, a bound strictly exceeding the infimum is attained by some competitor: there is an admissible curve $\gamma$ from $p$ to $q$ with $L_g(\gamma) < d_g(p,q) + 1$. In particular such a $\gamma$ exists, which is the assertion; the metric $g$ and the length bound play no part in it and are discarded.
\end{proof}

For convenience, if $(M, g)$ is a disconnected Riemannian manifold, we also let $d_g(p,q)$ denote the Riemannian distance from $p$ to $q$, provided that $p$ and $q$ lie in the same connected component of $M$. (See also Problem 2-30.)

It is convenient to have a form of the definition that makes sense before one knows
that $p$ and $q$ can be joined at all, and that ranges over a class of curves closed
under the operations used below.  We therefore take the following as the working
definition, with values in $[0,\infty]$.

\begin{definition}[The Riemannian Distance, extended form]
\label{def:riemannian-distance}
\lean{Manifold.riemannianEDist}
\mathlibok
\uses{def:riemannian-metric,def:length-of-curve}
Let $(M,g)$ be a Riemannian manifold. For $p, q \in M$ set
\[
  d_g(p,q) \;=\; \inf \bigl\{\, L_g(\gamma) \;:\; \gamma : [0,1] \to M \text{ is } C^1,\;
    \gamma(0) = p,\; \gamma(1) = q \,\bigr\} \;\in\; [0, \infty],
\]
the infimum of the empty set being $\infty$.
\end{definition}

\begin{remark}
\label{rem:riemannian-distance-c1-vs-admissible}
\uses{def:riemannian-distance}
Definition \ref{def:riemannian-distance} takes the infimum over $C^1$ curves, whereas
Lee takes it over admissible (piecewise regular) curves.  Neither class contains the
other: a $C^1$ curve need not be regular, and an admissible curve need not be $C^1$ at
its corners.  The two infima nevertheless agree, because an admissible curve may be
reparametrized so as to become $C^1$ with vanishing velocity at the corners without
changing its length, and conversely a $C^1$ curve may be approximated in length by
regular ones.  That equality is not needed for anything below and is not proved here;
every statement in this section refers to Definition \ref{def:riemannian-distance}.
\end{remark}

\begin{proposition}[Connected Riemannian Manifolds Are $C^1$-Path Connected]
\label{prop:connected-manifold-c1-curve}
\lean{LeeLib.Ch02.exists_contMDiffOn_path}
\leanok
\uses{def:riemannian-distance,prop:riemannian-distance-finite}
Let $(M,g)$ be a connected Riemannian manifold. Then any two points of $M$ can be joined
by a $C^1$ curve $\gamma : [0,1] \to M$.

This is Proposition \ref{prop:connected-manifold-admissible-curve} in the $C^1$ formulation
of Definition \ref{def:riemannian-distance}. Lee assumes only that $M$ be a connected smooth
manifold, with no metric; the statement here carries a metric because the proof measures with
it. That costs no generality, since every smooth manifold admits a Riemannian metric by
Proposition \ref{prop:existence-riemannian-metric}.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-distance,prop:riemannian-distance-finite}
Let $p, q \in M$. By Proposition \ref{prop:riemannian-distance-finite}, $d_g(p,q) < \infty$,
so $d_g(p,q) < d_g(p,q) + 1 < \infty$. By Definition \ref{def:riemannian-distance},
$d_g(p,q)$ is the infimum of $L_g(\gamma)$ over $C^1$ curves $\gamma$ from $p$ to $q$; since
$d_g(p,q) + 1$ is strictly greater than that infimum, it is not a lower bound for the set of
such lengths, so the set is nonempty and contains an element $L_g(\gamma) < d_g(p,q)+1$. The
corresponding $\gamma$ is a $C^1$ curve from $p$ to $q$.
\end{proof}

\begin{lemma}[Symmetry of the Riemannian Distance]
\label{lem:riemannian-distance-symm}
\lean{Manifold.riemannianEDist_comm}
\mathlibok
\uses{def:riemannian-distance}
$d_g(p,q) = d_g(q,p)$ for all $p, q$ in a Riemannian manifold $(M,g)$.
\end{lemma}

\begin{lemma}[Triangle Inequality for the Riemannian Distance]
\label{lem:riemannian-distance-triangle}
\lean{Manifold.riemannianEDist_triangle}
\mathlibok
\uses{def:riemannian-distance}
$d_g(p,r) \le d_g(p,q) + d_g(q,r)$ for all $p, q, r$ in a Riemannian manifold $(M,g)$.
\end{lemma}

\begin{lemma}[Nearby Points Are at Small Distance]
\label{lem:riemannian-distance-locally-small}
\lean{eventually_riemannianEDist_lt}
\mathlibok
\uses{def:riemannian-distance}
Let $(M,g)$ be a Riemannian manifold, let $p \in M$ and let $c > 0$. Then $p$ has a
neighborhood $U$ such that $d_g(p,q) < c$ for every $q \in U$.

This is the content of Lemma \ref{lem:distance-comparison-neighborhood}(a): a point
close to $p$ in a coordinate neighborhood is joined to $p$ by the coordinate segment,
whose $g$-length is controlled by its Euclidean length.
\end{lemma}

\begin{lemma}[Points at Small Distance Are Nearby]
\label{lem:riemannian-distance-small-implies-nearby}
\lean{setOf_riemannianEDist_lt_subset_nhds'}
\mathlibok
\uses{def:riemannian-distance}
Let $(M,g)$ be a Riemannian manifold and let $U$ be a neighborhood of $p \in M$. Then
there is $c > 0$ such that $\{q \in M : d_g(p,q) < c\} \subseteq U$.

This is the content of Lemma \ref{lem:distance-comparison-neighborhood}(b): a curve
from $p$ that leaves $U$ must cross a coordinate annulus around $p$, and so has
$g$-length bounded below.
\end{lemma}

\begin{proposition}[Finiteness of the Riemannian Distance]
\label{prop:riemannian-distance-finite}
\lean{LeeLib.Ch02.riemannianEDist_ne_top}
\leanok
\uses{def:riemannian-distance,lem:riemannian-distance-locally-small,lem:riemannian-distance-symm,lem:riemannian-distance-triangle}
Let $(M,g)$ be a connected Riemannian manifold. Then $d_g(p,q) < \infty$ for all
$p, q \in M$.

This is the conclusion that Proposition \ref{prop:connected-manifold-admissible-curve}
exists to supply: it is exactly the assertion that some curve of finite length joins
$p$ to $q$, so that the infimum defining $d_g(p,q)$ is taken over a nonempty set of
finite numbers.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:riemannian-distance-locally-small,lem:riemannian-distance-symm,lem:riemannian-distance-triangle}
Fix $p \in M$ and put $S = \{q \in M : d_g(p,q) < \infty\}$. We show $S$ is open, closed
and nonempty.

\emph{$S$ is open.} Let $q_0 \in S$. By Lemma \ref{lem:riemannian-distance-locally-small}
applied at $q_0$ with $c = 1$, there is a neighborhood $U$ of $q_0$ with $d_g(q_0, q) < 1$
for every $q \in U$. For such $q$, Lemma \ref{lem:riemannian-distance-triangle} gives
\[
  d_g(p,q) \;\le\; d_g(p,q_0) + d_g(q_0,q) \;<\; \infty,
\]
both summands being finite. Hence $U \subseteq S$.

\emph{$S$ is closed.} We show $M \setminus S$ is open. Let $q_0 \in M \setminus S$, so
$d_g(p,q_0) = \infty$. Again there is a neighborhood $U$ of $q_0$ with $d_g(q_0,q) < 1$
for every $q \in U$. Suppose some $q \in U$ had $d_g(p,q) < \infty$. Then by
Lemmas \ref{lem:riemannian-distance-triangle} and \ref{lem:riemannian-distance-symm},
\[
  d_g(p,q_0) \;\le\; d_g(p,q) + d_g(q,q_0) \;=\; d_g(p,q) + d_g(q_0,q) \;<\; \infty,
\]
contradicting $d_g(p,q_0) = \infty$. Hence $U \subseteq M \setminus S$.

\emph{$S$ is nonempty.} The constant curve at $p$ has length $0$, so $d_g(p,p) = 0 < \infty$
and $p \in S$.

Since $M$ is connected and $S$ is a nonempty subset that is both open and closed, $S = M$.
As $p \in M$ was arbitrary, $d_g(p,q) < \infty$ for all $p, q \in M$.
\end{proof}

% [FIGURE: Any two points can be connected by an admissible curve]

\begin{lemma}[Metric-Preserving Maps Do Not Increase Distance]
\label{lem:metric-preserving-distance-le}
\lean{LeeLib.Ch02.IsMetricPreserving.riemannianEDist_le}
\leanok
\uses{def:metric-preserving,def:riemannian-distance,prop:length-isometry-invariant}
Suppose $(M,g)$ and $(\tilde{M},\tilde{g})$ are Riemannian manifolds with or without boundary and $\varphi : M \to \tilde{M}$ is a smooth metric-preserving map. Then
\[
  d_{\tilde{g}}(\varphi(x), \varphi(y)) \;\le\; d_g(x,y) \qquad \text{for all } x, y \in M.
\]
\end{lemma}

\begin{proof}
\leanok
\uses{prop:length-isometry-invariant,def:riemannian-distance}
Let $r > d_g(x,y)$. By definition of $d_g$ as an infimum there is an admissible curve $\gamma : [0,1] \to M$ from $x$ to $y$ with $L_g(\gamma) < r$. Then $\varphi \circ \gamma$ is an admissible curve from $\varphi(x)$ to $\varphi(y)$, and by Proposition \ref{prop:length-isometry-invariant} it has $L_{\tilde{g}}(\varphi \circ \gamma) = L_g(\gamma) < r$. Since $d_{\tilde{g}}(\varphi(x),\varphi(y))$ is an infimum over such curves, $d_{\tilde{g}}(\varphi(x),\varphi(y)) < r$. As $r > d_g(x,y)$ was arbitrary and the order is dense, $d_{\tilde{g}}(\varphi(x),\varphi(y)) \le d_g(x,y)$.
\end{proof}

The inequality cannot be improved to an equality for metric-preserving maps in general: $\varphi$ carries curves of $M$ to curves of the same length in $\tilde{M}$, but $\tilde{M}$ may contain shorter curves between $\varphi(x)$ and $\varphi(y)$ that do not lie in the image of $\varphi$. Equality requires $\varphi$ to be invertible.

\begin{proposition}[Isometry Invariance of the Riemannian Distance Function]
\label{prop:isometry-invariance-riemannian-distance}
\lean{LeeLib.Ch02.IsIsometry.riemannianEDist_comp}
\leanok
\uses{def:isometry,def:riemannian-distance,lem:metric-preserving-distance-le,lem:isometry-symm}
\dcref{ch2:2.51}

Suppose $(M, g)$ and $(\tilde{M}, \tilde{g})$ are connected Riemannian manifolds with or without boundary, and $\varphi: M \to \tilde{M}$ is an isometry. Then $d_{\tilde{g}}(\varphi(x), \varphi(y)) = d_g(x, y)$ for all $x, y \in M$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:metric-preserving-distance-le,lem:isometry-symm}
Applying Lemma \ref{lem:metric-preserving-distance-le} to $\varphi$ gives $d_{\tilde{g}}(\varphi(x),\varphi(y)) \le d_g(x,y)$. By Lemma \ref{lem:isometry-symm}, $\varphi^{-1}$ is an isometry, so the same lemma applied to $\varphi^{-1}$ at the points $\varphi(x), \varphi(y)$ gives
\[
  d_g(x,y) = d_g\bigl(\varphi^{-1}(\varphi(x)), \varphi^{-1}(\varphi(y))\bigr) \le d_{\tilde{g}}(\varphi(x), \varphi(y)).
\]
The two inequalities give equality.
\end{proof}

\begin{exercise}
\label{exer:prove-isometry-invariance}
\leanok
\dcref{ch2:2.52}
\uses{prop:isometry-invariance-riemannian-distance}

Prove the preceding proposition.
\end{exercise}

\begin{proposition}[Linear Isometries of Euclidean Space]
\label{prop:euclidean-linear-isometry-distance}
\lean{LeeLib.Ch02.isIsometry_euclideanMetric_riemannianEDist}
\leanok
\uses{ex:euclidean-metric,def:isometry,prop:isometry-invariance-riemannian-distance,prop:euclidean-riemannian-distance}
Let $F$ and $\tilde{F}$ be inner product spaces with their Euclidean metrics, and let $e : F \to \tilde{F}$ be a linear isometry of $F$ onto $\tilde{F}$. Then $e$ is an isometry of the corresponding Riemannian manifolds, and
\[
  d_{\bar{g}}(e(x), e(y)) = |x - y| \qquad \text{for all } x, y \in F.
\]
\end{proposition}

\begin{proof}
\leanok
\uses{prop:isometry-invariance-riemannian-distance,prop:euclidean-riemannian-distance,ex:euclidean-metric}
A linear map is its own differential, so $de_x = e$ for every $x$, and the condition $e^* \bar{g} = \bar{g}$ reads $\langle e(v), e(w)\rangle = \langle v, w \rangle$, which holds because $e$ is a linear isometry. Being in addition a linear homeomorphism, $e$ is a diffeomorphism, so it is an isometry. Proposition \ref{prop:isometry-invariance-riemannian-distance} then gives $d_{\bar{g}}(e(x), e(y)) = d_{\bar{g}}(x,y)$, and $d_{\bar{g}}(x,y) = |x-y|$ by Proposition \ref{prop:euclidean-riemannian-distance}.
\end{proof}

(Note that unlike lengths of curves, Riemannian distances are not necessarily preserved by \textit{local} isometries; see Problem 2-31.)
\uses{prob:local-isometry-distance}

We wish to show that the Riemannian distance function turns $M$ into a metric space, whose metric topology is the same as its original manifold topology. To do so, we need the following lemmas.

The comparison below rests on two ingredients, isolated here. The first is the joint
smoothness of the metric in the base point and the tangent vector together, which is what
makes the function $(x,v) \mapsto |v|_g$ on the tangent bundle continuous.

\begin{lemma}[Smoothness of the Energy Function]
\label{lem:metric-energy-smooth}
\lean{LeeLib.Ch02.RiemannianMetric.contMDiff_innerAt_totalSpace}
\leanok
\uses{def:riemannian-metric}
Let $(M,g)$ be a Riemannian manifold. The energy function $E : TM \to \mathbb{R}$, defined on
the total space of the tangent bundle by
\[
E(x,v) = \langle v, v \rangle_g,
\]
is smooth. Consequently $(x,v) \mapsto |v|_g = E(x,v)^{1/2}$ is continuous on $TM$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-metric}
Smoothness of $g$ says that for any smooth map $b : N \to M$ and any two smooth sections
$X, Y$ of the pullback $b^*TM$, the function $p \mapsto \langle X_p, Y_p \rangle_{g_{b(p)}}$ is
smooth on $N$. Apply this with $N = TM$, with $b : TM \to M$ the bundle projection, and with
$X = Y$ the tautological section of $b^*TM$ assigning to a point $(x,v) \in TM$ the vector
$v \in T_x M$ itself. The resulting function is $E$. The final claim holds because
$t \mapsto t^{1/2}$ is continuous on $[0,\infty)$ and $E \geq 0$.
\end{proof}

The second ingredient is the analytic heart of the matter, stated with no reference to
manifolds: a family of inner products that varies continuously over a compact set is
comparable to a fixed norm, \emph{uniformly} over that set. The hypotheses are imposed only
on $K$, so the lemma applies to a family defined on any set containing $K$ --- which is what
Lemma \ref{lem:distance-comparison-neighborhood} needs, since there the family is a metric
read in a chart and is undefined outside the chart's image.

\begin{lemma}[Uniform Comparison over a Compact Set]
\label{lem:norm-comparison-compact-core}
\lean{LeeLib.Ch02.exists_forall_sqrt_comparison}
\leanok
Let $V$ be a finite-dimensional real normed vector space, let $q$ assign to each $x \in V$ a
bilinear form $q_x$ on $V$, and let $K \subseteq V$ be compact. Write $S = \{v \in V : |v| = 1\}$
and suppose
\begin{enumerate}
\item[(i)] $(x,v) \mapsto q_x(v,v)$ is continuous on $K \times S$;
\item[(ii)] $q_x(v,v) > 0$ for every $x \in K$ and every $v \neq 0$.
\end{enumerate}
Then there are positive constants $c, C$ such that
\[
c|v| \leq q_x(v,v)^{1/2} \leq C|v| \qquad \text{for all } x \in K \text{ and all } v \in V.
\]
\end{lemma}

\begin{proof}
\leanok
The set $K \times S$ is compact, being a product of compact sets ($S$ is compact because $V$ is
finite-dimensional). Every $v \in S$ is nonzero, so by (ii) the function $(x,v) \mapsto q_x(v,v)$
is strictly positive on $K \times S$; being continuous there by (i), it is bounded below by some
$c_0 > 0$ and above by some $C_0$. Set $c = c_0^{1/2}$ and $C = \max(C_0,1)^{1/2}$, both positive.

Let $x \in K$ and $v \in V$. If $v = 0$ both sides vanish. Otherwise put $u = v/|v| \in S$, so
that $v = |v|\,u$. Bilinearity gives the homogeneity relation $q_x(\lambda u, \lambda u) =
\lambda^2 q_x(u,u)$, whence $q_x(v,v) = |v|^2 q_x(u,u)$ and therefore
\[
q_x(v,v)^{1/2} = |v| \, q_x(u,u)^{1/2}.
\]
Since $c_0 \leq q_x(u,u) \leq C_0 \leq \max(C_0,1)$, taking square roots and multiplying by
$|v| \geq 0$ gives $c|v| \leq q_x(v,v)^{1/2} \leq C|v|$.

If $K$ or $S$ is empty the two bounds are produced vacuously, so no nondegeneracy assumption is
needed beyond (ii).
\end{proof}

Lee's lemma follows for a metric defined on all of $\mathbb{R}^n$; this is the case in which
the tangent bundle is canonically a product, so that the compact set $L$ appearing in the
argument literally is $K \times S^{n-1}$.

\begin{lemma}[Norm Comparison on $\mathbb{R}^n$]
\label{lem:riemannian-euclidean-norm-comparison-flat}
\lean{LeeLib.Ch02.RiemannianMetric.exists_forall_normAt_le_euclidean}
\leanok
\uses{lem:norm-comparison-compact-core, lem:metric-energy-smooth, def:riemannian-metric, ex:euclidean-metric}
Let $g$ be a Riemannian metric on $\mathbb{R}^n$ and let $\tilde{g}$ denote the Euclidean
metric. For every compact subset $K \subseteq \mathbb{R}^n$, there are positive constants
$c, C$ such that for all $x \in K$ and all $v \in T_x \mathbb{R}^n$,
\[
c|v|_{\tilde{g}} \leq |v|_g \leq C|v|_{\tilde{g}}.
\]
\end{lemma}

\begin{proof}
\leanok
\uses{lem:norm-comparison-compact-core, lem:metric-energy-smooth, ex:euclidean-metric}
Under the canonical identification of $T\mathbb{R}^n$ with $\mathbb{R}^n \times \mathbb{R}^n$,
Lemma \ref{lem:metric-energy-smooth} says exactly that $(x,v) \mapsto \langle v,v\rangle_g$ is
continuous on $\mathbb{R}^n \times \mathbb{R}^n$, so hypothesis (i) of Lemma
\ref{lem:norm-comparison-compact-core} holds for $q_x = g_x$ on $K \times S^{n-1}$; hypothesis
(ii) is the positive definiteness of $g$. That lemma yields $c, C > 0$ with
$c|v| \leq |v|_g \leq C|v|$ for $x \in K$, and $|v|_{\tilde{g}} = |v|$ because the Euclidean
metric assigns to a tangent vector its norm.
\end{proof}

Lee states the comparison for a metric defined only on an open subset, which is the form in
which it is applied to a metric read in a chart.

\begin{lemma}
\label{lem:riemannian-euclidean-norm-comparison}
\dcref{ch2:2.53}
\lean{LeeLib.Ch02.RiemannianMetric.exists_forall_normAt_le_euclidean_opens}
\leanok
\uses{def:riemannian-metric, ex:euclidean-metric, lem:norm-comparison-compact-core, lem:metric-energy-smooth, lem:riemannian-euclidean-norm-comparison-flat}

Let $g$ be a Riemannian metric on an open subset $W \subseteq \mathbb{R}^n$ or $\mathbb{R}^n_+$, and let $\tilde{g}$ denote the Euclidean metric on $W$. For every compact subset $K \subseteq W$, there are positive constants $c, C$ such that for all $x \in K$ and all $v \in T_x \mathbb{R}^n$,
\begin{equation}
c|v|_{\tilde{g}} \leq |v|_g \leq C|v|_{\tilde{g}}.
\tag{2.21}
\end{equation}

\end{lemma}

\begin{proof}
\leanok
\uses{lem:norm-comparison-compact-core, lem:metric-energy-smooth, ex:euclidean-metric}
Define a continuous function $F: TW \to \mathbb{R}$ by $F(x, v) = |v|_g = g_x(v, v)^{1/2}$ for $x \in W$ and $v \in T_x \mathbb{R}^n$; it is continuous by Lemma \ref{lem:metric-energy-smooth}. Let $K \subseteq W$ be any compact subset, and define $L \subseteq TW$ by
\[
L = \{(x, v) \in TW : x \in K, |v|_{\tilde{g}} = 1\}.
\]

Because $W$ is open in $\mathbb{R}^n$, every chart of $W$ is the inclusion $W \hookrightarrow \mathbb{R}^n$; the tangent bundle $TW$ therefore admits the single global trivialization $TW \cong W \times \mathbb{R}^n$, under which $L$ is the product set $K \times S^{n-1}$, which is compact. Since $F$ is positive on $L$, there are positive constants $c, C$ such that
\[
c \leq |v|_g \leq C \quad \text{for all } (x, v) \in L.
\]

For each $x \in K$ and $v \in T_x \mathbb{R}^n$ with $v \neq 0$, we can write $v = \lambda \tilde{v}$, where $\lambda = |v|_{\tilde{g}}$ and $\tilde{v} = v/\lambda$. Then $(x, \tilde{v}) \in L$, and it follows from the homogeneity of $|\cdot|_g$ that
\[
|v|_g = \lambda |\tilde{v}|_g \leq \lambda C = C|v|_{\tilde{g}}.
\]

The same inequality holds trivially when $v = 0$. Arguing similarly in the other direction, we conclude that (2.21) holds for all $x \in K$ and $v \in T_x \mathbb{R}^n$. The two displayed bounds on $L$, together with the homogeneity step, are the content of Lemma \ref{lem:norm-comparison-compact-core} applied to $q_x = g_x$ and the compact set $K$.
\end{proof}

% [FIGURE: Two overlapping regions with one labeled V, a point p inside, and a point q with an arrow labeled  \gamma (t₀), illustrating the neighborhood structure for the distance comparison proof]

% [FIGURE: A curved path with labeled points p,  \gamma ₁, q,  \gamma ₂, and r, demonstrating the triangle inequality concept]

The next lemma shows how to transfer this result to Riemannian manifolds.

\begin{lemma}\label{lem:distance-comparison-neighborhood}\dcref{ch2:2.54}
\lean{LeeLib.Ch02.RiemannianMetric.exists_coordinate_nbhd_riemannianEDist_comparison}
\leanok
\uses{def:riemannian-metric,def:riemannian-distance,lem:riemannian-distance-locally-small,lem:riemannian-distance-small-implies-nearby}
Let $(M, g)$ be a Riemannian manifold with or without boundary and
let $d_g$ be its Riemannian distance function. Suppose $U$ is an open subset of $M$ and
$p \in U$. Then $p$ has a coordinate neighborhood $V \subset U$ with the property that there
are positive constants $C, D$ satisfying the following inequalities:
\begin{enumerate}
\item[(a)] If $q \in V$, then $d_g(p,q) \leq Cd_{\bar{g}}(p,q)$, where $\bar{g}$ is the Euclidean metric in the
given coordinates on $\mathbb{R}^n$ or $\mathbb{R}^n_+$.
\item[(b)] If $q \in M \setminus V$, then $d_g(p,q) \geq D$.
\end{enumerate}

\begin{proof}
\leanok
Let $W$ be any neighborhood of $p$ contained in $U$ on which there exist smooth
coordinates $(x^i)$. Using these coordinates, we can identify $W$ with an open subset
of $\mathbb{R}^n$ or $\mathbb{R}^n_+$. Let $K$ be a compact subset of $W$ containing a neighborhood $V$ of
$p$, chosen as follows: If $p$ is an interior point of $M$, let $K$ be the closed Euclidean
ball $\bar{B}_\epsilon(0)$, with $\epsilon > 0$ chosen small enough that $K \subseteq W$, and let $V =
B_\epsilon(0)$. If $p \in \partial M$, let $K$ be a closed half-ball $\bar{B}_\epsilon(0) \cap \mathbb{R}^n_+$ with $\epsilon$ chosen similarly, and let $V =
B_\epsilon(0) \cap \mathbb{R}^n_+$. Let $\bar{g} = \sum_i (dx^i)^2$ denote the Euclidean metric in these coordinates,
and let $c, C$ be constants satisfying (2.21). If $\gamma$ is any admissible curve whose image
lies entirely in $V$, it follows that

$$cL_{\bar{g}}(\gamma) \leq L_g(\gamma) \leq CL_{\bar{g}}(\gamma). \quad (2.22)$$

To prove (a), suppose $q \in V$. Letting $\gamma : [a,b] \to V$ be a parametrization of the
line segment from $p$ to $q$, we conclude from (2.22) that

$$d_g(p,q) \leq L_g(\gamma) \leq CL_{\bar{g}}(\gamma) = Cd_{\bar{g}}(p,q).$$

To prove (b), suppose $q \in M \setminus V$. If $\gamma : [a,b] \to M$ is any admissible curve from
$p$ to $q$, let $t_0$ denote the infimum of times $t \in [a,b]$ such that $\gamma(t) \notin V$. It follows
that $\gamma([a, t_0]) \subseteq K$ and $d_{\bar{g}}(p, \gamma(t_0)) = \epsilon$ by continuity (Fig. 2.3), so (2.22) implies

$$L_g(\gamma) \geq L_g(\gamma|_{[a,t_0]}) \geq cL_{\bar{g}}(\gamma|_{[a,t_0]}) \geq cd_{\bar{g}}(p, \gamma(t_0)) = c\epsilon.$$

Taking the infimum over all such curves $\gamma$, we obtain (b) with $D = c\varepsilon$. \qed
\end{proof}
\end{lemma}

\begin{theorem}[Riemannian Manifolds as Metric Spaces]
\label{thm:riemannian-manifolds-metric-spaces}
\dcref{ch2:2.55}
\lean{LeeLib.Ch02.MetricSpace.ofRiemannianMetric}
\leanok
\uses{def:riemannian-distance,prop:riemannian-distance-finite,lem:riemannian-distance-symm,lem:riemannian-distance-triangle,lem:riemannian-distance-locally-small,lem:riemannian-distance-small-implies-nearby}

Let $(M,g)$ be a connected Riemannian manifold with or without boundary. Then $M$ carries a
metric space structure.

That its distance function is $d_g$ is Lemma \ref{lem:riemannian-metric-space-dist}, and that
its topology is the manifold topology is Theorem \ref{thm:riemannian-distance-topology}; the
three together are Lee's Theorem 2.55.
\end{theorem}

\begin{proof}
\leanok
\uses{prop:riemannian-distance-finite,lem:riemannian-distance-symm,lem:riemannian-distance-triangle,lem:riemannian-distance-locally-small,lem:riemannian-distance-small-implies-nearby}
It is immediate from the definition of $d_g$ that $d_g(p,q) \ge 0$ and $d_g(p,p) = 0$, and
$d_g(p,q) = d_g(q,p)$ is Lemma \ref{lem:riemannian-distance-symm}. The triangle inequality
is Lemma \ref{lem:riemannian-distance-triangle}; it holds because a curve from $p$ to $q$
can be concatenated with one from $q$ to $r$ to yield one from $p$ to $r$ whose length is
the sum of the two lengths.

For separation, suppose $p, q \in M$ are distinct. Since $M$ is Hausdorff, there is an open
set $U$ containing $p$ but not $q$. By Lemma \ref{lem:riemannian-distance-small-implies-nearby}
there is $c > 0$ with $\{r : d_g(p,r) < c\} \subseteq U$. As $q \notin U$, we get
$d_g(p,q) \ge c > 0$.

Finally, $d_g$ takes finite values by Proposition \ref{prop:riemannian-distance-finite},
which is where connectedness of $M$ is used, so $d_g$ is a genuine real-valued metric rather
than merely an extended one. Together these say that $d_g$ turns $M$ into a metric space.
\end{proof}

The theorem as stated asserts only that some metric exists. What makes it Lee's Theorem 2.55
is that the metric in question is $d_g$, which is the following lemma.

\begin{lemma}[The Metric Is the Riemannian Distance]
\label{lem:riemannian-metric-space-dist}
\lean{LeeLib.Ch02.MetricSpace.ofRiemannianMetric_dist}
\leanok
\uses{thm:riemannian-manifolds-metric-spaces,def:riemannian-distance,prop:riemannian-distance-finite}
The distance function of the metric space of Theorem \ref{thm:riemannian-manifolds-metric-spaces}
is the Riemannian distance: for all $p, q \in M$,
\[
  \operatorname{dist}(p,q) \;=\; d_g(p,q),
\]
where the right-hand side, a priori an element of $[0,\infty]$ by
Definition \ref{def:riemannian-distance}, is read as a real number using
Proposition \ref{prop:riemannian-distance-finite}.
\end{lemma}

\begin{proof}
\leanok
\uses{def:riemannian-distance,prop:riemannian-distance-finite}
The metric space of Theorem \ref{thm:riemannian-manifolds-metric-spaces} is obtained from the
extended distance $d_g$ of Definition \ref{def:riemannian-distance} by taking real values, which
is legitimate because $d_g$ is finite by Proposition \ref{prop:riemannian-distance-finite}. Its
distance function is therefore $d_g$ itself, and the identity holds by construction.
\end{proof}

The remaining assertion of Lee's Theorem 2.55 is that this metric induces the topology one
started with; it is separated out here because it is a statement about the metric
constructed above rather than part of its construction.

\begin{theorem}[The Metric Topology Is the Manifold Topology]
\label{thm:riemannian-distance-topology}
\lean{LeeLib.Ch02.MetricSpace.ofRiemannianMetric_topology}
\leanok
\uses{thm:riemannian-manifolds-metric-spaces,lem:riemannian-distance-locally-small,lem:riemannian-distance-small-implies-nearby}
Let $(M,g)$ be a connected Riemannian manifold with or without boundary. The topology
induced by the metric $d_g$ of Theorem \ref{thm:riemannian-manifolds-metric-spaces} is the
given manifold topology.
\end{theorem}

\begin{proof}
\leanok
\uses{lem:riemannian-distance-locally-small,lem:riemannian-distance-small-implies-nearby}
Suppose first that $U \subseteq M$ is open in the manifold topology and let $p \in U$. By
Lemma \ref{lem:riemannian-distance-small-implies-nearby} there is $c > 0$ with
$\{q : d_g(p,q) < c\} \subseteq U$, i.e. the $d_g$-ball of radius $c$ about $p$ is contained
in $U$. Thus $U$ is open in the metric topology.

Conversely, suppose $U'$ is open in the metric topology and let $p \in U'$. Choose $\delta > 0$
with the $d_g$-ball of radius $\delta$ about $p$ contained in $U'$. By
Lemma \ref{lem:riemannian-distance-locally-small} applied at $p$ with $c = \delta$, there is a
neighborhood $V$ of $p$ in the manifold topology with $d_g(p,q) < \delta$ for every $q \in V$;
hence $V \subseteq U'$. So $U'$ is a neighborhood of each of its points in the manifold
topology, and is therefore open in it.

The two inclusions of neighborhood filters say exactly that the two topologies coincide.
\end{proof}

\begin{proposition}[The Euclidean Distance]
\label{prop:euclidean-riemannian-distance}
\lean{LeeLib.Ch02.euclideanMetric_riemannianEDist}
\leanok
\uses{def:riemannian-distance,ex:euclidean-metric}
On $\mathbb{R}^n$ with the Euclidean metric $\bar g$ of Example \ref{ex:euclidean-metric},
the Riemannian distance is the Euclidean distance:
\[
  d_{\bar g}(x,y) \;=\; |x - y| \qquad \text{for all } x, y \in \mathbb{R}^n .
\]
This is the one case in which the answer is known in advance, so it is the natural check on
Definition \ref{def:riemannian-distance}. The real-valued form $d_{\bar g}(x,y) = |x-y|$ is
Lemma \ref{lem:euclidean-riemannian-distance-real}.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-distance,ex:euclidean-metric}
For the upper bound, let $\gamma(t) = x + t(y-x)$ for $t \in [0,1]$. This curve is $C^1$, joins
$x$ to $y$, and has $\gamma'(t) = y - x$, so under the identification of each tangent space of
$\mathbb{R}^n$ with $\mathbb{R}^n$ its $\bar g$-length is
\[
  L_{\bar g}(\gamma) \;=\; \int_0^1 |\gamma'(t)| \, dt \;=\; \int_0^1 |y-x| \, dt \;=\; |y-x| .
\]
Hence $d_{\bar g}(x,y) \le |x-y|$.

For the lower bound, let $\gamma : [0,1] \to \mathbb{R}^n$ be any $C^1$ curve with
$\gamma(0) = x$ and $\gamma(1) = y$. By the fundamental theorem of calculus and the triangle
inequality for the integral,
\[
  |y - x| \;=\; \Bigl| \int_0^1 \gamma'(t) \, dt \Bigr| \;\le\; \int_0^1 |\gamma'(t)| \, dt
    \;=\; L_{\bar g}(\gamma) .
\]
So $|x-y|$ is a lower bound for the set of lengths whose infimum is $d_{\bar g}(x,y)$, giving
$|x-y| \le d_{\bar g}(x,y)$.

The two inequalities give $d_{\bar g}(x,y) = |x-y|$.
\end{proof}

\begin{lemma}[The Euclidean Distance, real-valued form]
\label{lem:euclidean-riemannian-distance-real}
\lean{LeeLib.Ch02.euclideanMetric_riemannianEDist_toReal}
\leanok
\uses{prop:euclidean-riemannian-distance}
With the notation of Proposition \ref{prop:euclidean-riemannian-distance}, $d_{\bar g}(x,y)$ is
finite and equals $|x-y|$ as a real number.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:euclidean-riemannian-distance}
Immediate from Proposition \ref{prop:euclidean-riemannian-distance}, since $|x-y|$ is a real
number and hence $d_{\bar g}(x,y) \ne \infty$.
\end{proof}

Thanks to the preceding theorem, it makes sense to apply all the concepts of the theory of metric spaces to a connected Riemannian manifold $(M,g)$. For example, we say that $M$ is \textit{(metrically) complete} if every Cauchy sequence in $M$ converges. A subset $A \subseteq M$ is \textit{bounded} if there is a constant $C$ such that $d_g(p,q) \leq C$ for all $p,q \in A$; if this is the case, the \textit{diameter of } $A$ is the smallest such constant:
\[
\operatorname{diam}(A) = \sup\{d_g(p,q) : p,q \in A\}.
\]

Since every compact metric space is bounded, every compact connected Riemannian manifold with or without boundary has finite diameter. (Note that the unit sphere with the Riemannian distance determined by the round metric has diameter $\pi$, not $2$, since the Riemannian distance between antipodal points is $\pi$. See Problem 6-2.)

\section{Pseudo-Riemannian Metrics}

From the point of view of geometry, Riemannian metrics are by far the most important structures that manifolds carry. However, there is a generalization of Riemannian metrics that has become especially important because of its application to physics.

Before defining this generalization, we begin with some linear-algebraic preliminaries. Suppose $V$ is a finite-dimensional vector space, and $q$ is a symmetric covariant 2-tensor on $V$ (also called a \textbf{symmetric bilinear form}). Just as for an inner product, there is a linear map $\tilde{q}: V \to V^*$ defined by
\[
\tilde{q}(v)(w) = q(v,w) \text{ for all } v, w \in V.
\]

We say that $q$ is \textbf{nondegenerate} if $\tilde{q}$ is an isomorphism.

\begin{lemma}
\label{lem:nondegenerate-symmetric-bilinear}
\lean{LeeLib.Ch02.nondegenerate_tfae}
\leanok
\dcref{ch2:2.56}
Suppose $q$ is a symmetric covariant 2-tensor on a finite-dimensional vector space $V$. The following are equivalent:
\begin{enumerate}
\item[(a)] $q$ is nondegenerate.
\item[(b)] For every nonzero $v \in V$, there is some $w \in V$ such that $q(v,w) \neq 0$.
\item[(c)] If $q = q_{ij} e^i e^j$ in terms of some basis $(e^j)$ for $V^*$, then the matrix $(q_{ij})$ is invertible.
\end{enumerate}
\end{lemma}

\begin{proof}
\leanok
(a) $\Rightarrow$ (b): if $\tilde q$ is an isomorphism and $v \neq 0$, then $\tilde q(v) \neq 0$, so there is some $w \in V$ with $q(v,w) = \tilde q(v)(w) \neq 0$.

(b) $\Rightarrow$ (a): condition (b) says exactly that $\tilde q$ is injective, since $\tilde q(v) = 0$ means $q(v,w) = 0$ for every $w$, which by (b) forces $v = 0$. As $\dim V^* = \dim V < \infty$, an injective linear map $V \to V^*$ is an isomorphism.

(a) $\Leftrightarrow$ (c): let $(e_i)$ be the basis of $V$ dual to $(e^j)$. Then $\tilde q(e_i)(e_j) = q(e_i, e_j) = q_{ij}$, so $(q_{ij})$ is precisely the matrix of $\tilde q : V \to V^*$ with respect to the bases $(e_i)$ and $(e^j)$. A linear map between spaces of equal finite dimension is an isomorphism if and only if its matrix in any pair of bases is invertible.
\end{proof}

\begin{exercise}
\label{exer:prove-nondegenerate-equivalent}
\leanok
\dcref{ch2:2.57}
\uses{lem:nondegenerate-symmetric-bilinear}
Prove the preceding lemma.
\end{exercise}

\begin{proof}
\leanok
This is the content of Lemma \ref{lem:nondegenerate-symmetric-bilinear}, proved above.
\end{proof}

Every inner product is a nondegenerate symmetric bilinear form, as is every symmetric bilinear form that is \textbf{negative definite} (which is defined by obvious analogy with positive definite). But there are others that are neither positive definite nor negative definite, as we will see below.

We use the term \textbf{scalar product} to denote any nondegenerate symmetric bilinear form on a finite-dimensional vector space $V$, and reserve the term \textbf{inner product} for the special case of a positive definite scalar product. A \textbf{scalar product space} is a finite-dimensional vector space endowed with a scalar product. When convenient, we will often use a notation like $\langle \cdot, \cdot \rangle$ to denote a scalar product. We say that vectors $v, w \in V$ are \textbf{orthogonal} if $\langle v, w \rangle = 0$, just as in the case of an inner product. Given a vector $v \in V$, we define the \textbf{norm} of $v$ to be $|v| = \sqrt{\langle v, v \rangle}$. Note that in the indefinite case, it is possible for a nonzero vector to be orthogonal to itself, and thus to have norm zero. Thus $|v|$ is not technically a norm in the sense defined on page 47 below, but it is customary to call it ``the norm of $v$'' anyway.

\begin{exercise}
\label{exer:polarization-scalar-product}
\lean{LeeLib.Ch02.BilinForm.IsSymm.inner_eq_apply_add_sub_apply_sub_div_four}
\leanok
\dcref{ch2:2.58}
\uses{exer:polarization-proof}
Prove that the polarization identity (2.2) holds for every scalar product.
\end{exercise}

\begin{proof}
\leanok
The proof of Lemma \ref{lem:polarization-identity} used only bilinearity and symmetry of $\langle \cdot, \cdot \rangle$, never positive definiteness: expanding $\langle v \pm w, v \pm w \rangle = \langle v,v \rangle \pm 2\langle v,w \rangle + \langle w,w \rangle$ and subtracting gives $4\langle v,w \rangle$. Since a scalar product is by definition a symmetric bilinear form, the identity holds verbatim.
\end{proof}

If $S \subseteq V$ is any linear subspace, the set of vectors in $V$ that are orthogonal to every vector in $S$ is a linear subspace denoted by $S^{\perp}$.

\begin{lemma}
\label{lem:orthogonal-complement}
\lean{LeeLib.Ch02.finrank_add_finrank_orthogonal_eq_finrank}
\leanok
\dcref{ch2:2.59}
Suppose $(V, q)$ is a finite-dimensional scalar product space, and $S \subseteq V$ is a linear subspace. Then $\dim S + \dim S^{\perp} = \dim V$.
\end{lemma}

\begin{proof}
\leanok
Define a linear map $\Phi : V \to S^*$ by $\Phi(v) = \tilde{q}(v)|_S$. Note that $v \in \Ker\Phi$ if and only if $q(v,x) = \tilde{q}(v)(x) = 0$ for all $x \in S$, so $\Ker\Phi = S^\perp$. If $\varphi \in S^*$ is arbitrary, there is a covector $\tilde{\varphi} \in V^*$ whose restriction to $S$ is equal to $\varphi$. (For example, such a covector is easily constructed after choosing a basis for $S$ and extending it to a basis for $V$.) Since $\tilde{q}$ is an isomorphism, there exists $v \in V$ such that $\tilde{q}(v) = \tilde{\varphi}$. It follows that $\Phi(v) = \varphi$, and therefore $\Phi$ is surjective. By the rank--nullity theorem, the dimension of $S^\perp = \Ker\Phi$ is equal to $\dim V - \dim S^* = \dim V - \dim S$.
\end{proof}

\begin{lemma}
\label{lem:orthogonal-complement-involution}
\lean{LeeLib.Ch02.orthogonal_orthogonal_eq_self}
\leanok
\dcref{ch2:2.59}
\uses{lem:orthogonal-complement}
Suppose $(V, q)$ is a finite-dimensional scalar product space, and $S \subseteq V$ is a linear subspace. Then $(S^{\perp})^{\perp} = S$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthogonal-complement}
Every $v \in S$ is orthogonal to every element of $S^\perp$ by definition, so $S \subseteq (S^\perp)^\perp$. Applying Lemma \ref{lem:orthogonal-complement} twice gives $\dim (S^\perp)^\perp = \dim V - \dim S^\perp = \dim S$. Because these finite-dimensional vector spaces are nested and have the same dimension, they are equal.
\end{proof}

An ordered $k$-tuple $(v_1,\ldots,v_k)$ of elements of $V$ is said to be \textbf{orthonormal} if $|v_i| = 1$ for each $i$ and $\langle v_i, v_j \rangle = 0$ for $i \neq j$, or equivalently, if $\langle v_i, v_j \rangle = \pm\delta_{ij}$ for each $i$ and $j$. We wish to prove that every scalar product space has an orthonormal basis. Note that the usual Gram--Schmidt algorithm does not always work in this situation, because the vectors that appear in the denominators in (2.5)--(2.6) might have vanishing norms. In order to get around this problem, we introduce the following definitions. If $(V,q)$ is a finite-dimensional scalar product space, a subspace $S \subseteq V$ is said to be \textbf{nondegenerate} if the restriction of $q$ to $S \times S$ is nondegenerate. An ordered $k$-tuple of vectors $(v_1,\ldots,v_k)$ in $V$ is said to be \textbf{nondegenerate} if for each $j = 1,\ldots,k$, the vectors $(v_1,\ldots,v_j)$ span a nondegenerate $j$-dimensional subspace of $V$. For example, every orthonormal basis is nondegenerate.

\begin{lemma}\label{lem:nondegenerate-subspace}
\lean{LeeLib.Ch02.restrict_nondegenerate_tfae}
\leanok
\dcref{ch2:2.60}
\uses{lem:orthogonal-complement-involution}
Suppose $(V,q)$ is a finite-dimensional scalar product space, and $S \subseteq V$ is a linear subspace. The following are equivalent:
\begin{enumerate}[label=(\alph*)]
\item $S$ is nondegenerate.
\item $S^\perp$ is nondegenerate.
\item $S \cap S^\perp = \{0\}$.
\item $V = S \oplus S^\perp$.
\end{enumerate}
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthogonal-complement,lem:orthogonal-complement-involution}
(c) $\Rightarrow$ (a): suppose $S \cap S^\perp = \{0\}$ and let $x \in S$ satisfy $q(x,y) = 0$ for all $y \in S$. Then $x \in S^\perp$, so $x \in S \cap S^\perp = \{0\}$; thus $q|_{S \times S}$ is nondegenerate by Lemma \ref{lem:nondegenerate-symmetric-bilinear}.

(a) $\Rightarrow$ (d): suppose $S$ is nondegenerate. Then $S \cap S^\perp = \{0\}$, since any $x$ in the intersection is orthogonal to all of $S$ and lies in $S$, hence vanishes by nondegeneracy of $q|_{S \times S}$. By Lemma \ref{lem:orthogonal-complement}, $\dim S + \dim S^\perp = \dim V$, so $S + S^\perp$ has dimension $\dim S + \dim S^\perp - \dim(S \cap S^\perp) = \dim V$, whence $S + S^\perp = V$ and the sum is direct.

(d) $\Rightarrow$ (c): immediate, since a direct sum decomposition has trivial intersection.

(a) $\Leftrightarrow$ (b): by Lemma \ref{lem:orthogonal-complement-involution}, $(S^\perp)^\perp = S$, so applying the equivalence (a) $\Leftrightarrow$ (d) to the subspace $S^\perp$ shows that $S^\perp$ is nondegenerate if and only if $V = S^\perp \oplus (S^\perp)^\perp = S^\perp \oplus S$. Since a direct sum decomposition is symmetric in its two summands, this holds if and only if $V = S \oplus S^\perp$, which by (a) $\Leftrightarrow$ (d) holds if and only if $S$ is nondegenerate.
\end{proof}

\begin{exercise}\label{exer:preceding-lemma-nondegenerate}
\leanok
\dcref{ch2:2.61}
\uses{lem:nondegenerate-subspace}
Prove the preceding lemma.
\end{exercise}

\begin{proof}
\leanok
This is the content of Lemma \ref{lem:nondegenerate-subspace}, proved above.
\end{proof}

\begin{lemma}[Completion of Nondegenerate Bases]\label{lem:completion-nondegenerate-bases}
\lean{LeeLib.Ch02.exists_isNondegenerateTuple_basis}
\leanok
\dcref{ch2:2.62}
\uses{lem:nondegenerate-subspace}
Suppose $(V,q)$ is an $n$-dimensional scalar product space, and $(v_1,\ldots,v_k)$ is a nondegenerate $k$-tuple in $V$ with $0 \leq k < n$. Then there exist vectors $v_{k+1},\ldots,v_n \in V$ such that $(v_1,\ldots,v_n)$ is a nondegenerate basis for $V$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-subspace,lem:orthogonal-complement,exer:polarization-scalar-product}
It suffices to adjoin one vector, and then to repeat.

Let $S = \operatorname{span}(v_1,\ldots,v_k) \subseteq V$, a nondegenerate subspace of dimension $k$. Because $k < n$, Lemma \ref{lem:orthogonal-complement} gives $\dim S^\perp = n - k > 0$, so $S^\perp$ is a nontrivial subspace of $V$, and Lemma \ref{lem:nondegenerate-subspace} shows that $S^\perp$ is nondegenerate and $V = S \oplus S^\perp$.

There must be a vector in $S^\perp$ with nonzero length. Otherwise $\langle x,x\rangle = 0$ for every $x \in S^\perp$, and then the polarization identity of Exercise \ref{exer:polarization-scalar-product},
\[
\langle x,y \rangle = \tfrac14\left(\langle x+y,x+y\rangle - \langle x-y,x-y\rangle\right),
\]
applied to $x, y \in S^\perp$ (which is a subspace, so $x \pm y \in S^\perp$) would give $\langle x,y\rangle = 0$ for all $x,y \in S^\perp$; as $S^\perp \neq \{0\}$, this contradicts the nondegeneracy of $S^\perp$.

Let $v_{k+1} \in S^\perp$ have $\langle v_{k+1},v_{k+1}\rangle \neq 0$. Then $v_{k+1} \notin S$, since $S \cap S^\perp = \{0\}$ and $v_{k+1} \neq 0$; hence $\dim(S \oplus \langle v_{k+1}\rangle) = k+1$. This span is nondegenerate: suppose $y = s + c\,v_{k+1}$ with $s \in S$ is orthogonal to all of $S \oplus \langle v_{k+1}\rangle$. Testing against any $w \in S$ gives $\langle s,w\rangle + c\langle v_{k+1},w\rangle = \langle s,w\rangle = 0$, since $v_{k+1} \in S^\perp$; as $S$ is nondegenerate, $s = 0$. Testing $y = c\,v_{k+1}$ against $v_{k+1}$ gives $c\langle v_{k+1},v_{k+1}\rangle = 0$, and $\langle v_{k+1},v_{k+1}\rangle \neq 0$ forces $c = 0$. So $y = 0$.

Thus $(v_1,\ldots,v_{k+1})$ is a nondegenerate $(k+1)$-tuple. Repeating this argument for $v_{k+2},\ldots,v_n$, which terminates after $n-k$ steps, completes the proof.
\end{proof}


\begin{proposition}[Gram--Schmidt Algorithm for Scalar Products]
\label{prop:gram-schmidt-algorithm-scalar-products}
\lean{LeeLib.Ch02.exists_basis_isOrthonormal_prefixSpan_eq}
\leanok
\dcref{ch2:2.63}
Suppose $(V, q)$ is an $n$-dimensional scalar product space. If $(v_i)$ is a nondegenerate basis for $V$, then there is an orthonormal basis $(b_i)$ with the property that $\operatorname{span}(b_1, \ldots, b_k) = \operatorname{span}(v_1, \ldots, v_k)$ for each $k = 1, \ldots, n$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
As in the positive definite case the tuple is constructed recursively, the inductive hypothesis at stage $k$ being that $(b_1,\ldots,b_k)$ is orthonormal and satisfies $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq k$. For $k = 0$ both conditions are vacuous.

Assume $(b_1,\ldots,b_k)$ has been constructed and write $S = \operatorname{span}(b_1,\ldots,b_k) = \operatorname{span}(v_1,\ldots,v_k)$. Set
\[
z = v_{k+1} - \sum_{i=1}^k \frac{\langle v_{k+1}, b_i \rangle}{\langle b_i, b_i \rangle} b_i.
\]
Each denominator is $\langle b_i, b_i \rangle = \pm 1 \neq 0$, so $z$ is well defined; this is the step at which positive definiteness is dispensable but nondegeneracy of the prefixes is not. For each $j \leq k$, orthogonality of the $b_i$ leaves a single surviving term in the sum:
\[
\langle z, b_j \rangle = \langle v_{k+1}, b_j \rangle - \frac{\langle v_{k+1}, b_j \rangle}{\langle b_j, b_j \rangle}\langle b_j, b_j \rangle = 0,
\]
so $z$ is orthogonal to every element of $S$.

Since $z - v_{k+1} \in S$, adjoining $z$ and adjoining $v_{k+1}$ to $S$ give the same subspace:
\[
S + \langle z \rangle = S + \langle v_{k+1} \rangle = \operatorname{span}(v_1,\ldots,v_{k+1}) =: T .
\]
Moreover $z \neq 0$; otherwise $v_{k+1} \in S$ and hence $T = S$, contradicting $\dim T = k+1 > k = \dim S$, both dimensions being part of the hypothesis that $(v_i)$ is nondegenerate.

Next, $\langle z, z \rangle \neq 0$. This is the one point at which the indefinite case genuinely differs: $z \neq 0$ no longer suffices. Suppose $\langle z,z\rangle = 0$. Every $w \in T$ can be written $w = s + cz$ with $s \in S$ and $c \in \mathbb{R}$, whence
\[
\langle z, w \rangle = \langle z, s \rangle + c\langle z, z \rangle = 0 ,
\]
so $z$ is orthogonal to all of $T$. As $T$ is nondegenerate, Lemma \ref{lem:nondegenerate-symmetric-bilinear} forces $z = 0$, a contradiction.

We may therefore complete the inductive step by putting
\[
b_{k+1} = \frac{z}{\sqrt{|\langle z, z \rangle|}} ,
\]
dividing by $\sqrt{|\langle z,z\rangle|}$ rather than by $|z|$, which need not be a real number in the indefinite case. Then $\langle b_{k+1}, b_{k+1} \rangle = \langle z,z\rangle/|\langle z,z\rangle| = \pm 1$, and $b_{k+1}$ is still orthogonal to $b_1,\ldots,b_k$, so $(b_1,\ldots,b_{k+1})$ is orthonormal. Rescaling by the nonzero scalar $|\langle z,z\rangle|^{-1/2}$ does not change the span, so
\[
\operatorname{span}(b_1,\ldots,b_{k+1}) = S + \langle z \rangle = T = \operatorname{span}(v_1,\ldots,v_{k+1}),
\]
which completes the induction.

Taking $k = n$: the tuple $(b_1,\ldots,b_n)$ is orthonormal, hence linearly independent, and it spans $\operatorname{span}(v_1,\ldots,v_n) = V$. So it is an orthonormal basis with the required property.
\end{proof}

\begin{proposition}
\label{prop:symmetric-bilinear-orthogonal-basis}
\lean{LinearMap.BilinForm.exists_orthogonal_basis}
\mathlibok
Let $q$ be a symmetric bilinear form on a finite-dimensional vector space $V$ over a field in which $2$ is invertible. Then $V$ has a basis $(b_i)$ that is $q$-orthogonal, that is, $q(b_i, b_j) = 0$ whenever $i \neq j$.
\end{proposition}

\begin{proposition}
\label{prop:scalar-product-orthonormal-basis}
\lean{LeeLib.Ch02.exists_basis_isOrthonormal}
\leanok
\uses{prop:symmetric-bilinear-orthogonal-basis}
Every scalar product space $(V,q)$ has an orthonormal basis.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:symmetric-bilinear-orthogonal-basis,lem:nondegenerate-symmetric-bilinear}
Since $q$ is symmetric and $\dim V < \infty$, Proposition \ref{prop:symmetric-bilinear-orthogonal-basis} provides a $q$-orthogonal basis $(b_i)$ of $V$. No $b_i$ has vanishing norm: if $\langle b_i, b_i \rangle = 0$, then, $b_i$ being orthogonal to every $b_j$ with $j \neq i$ as well, $b_i$ would be orthogonal to a basis of $V$ and hence to all of $V$, contradicting nondegeneracy by Lemma \ref{lem:nondegenerate-symmetric-bilinear}. Set
\[
e_i = \frac{b_i}{\sqrt{|\langle b_i, b_i \rangle|}}.
\]
This is legitimate because $|\langle b_i, b_i \rangle| > 0$; note that one divides by $\sqrt{|\langle b_i,b_i\rangle|}$ rather than by $|b_i|$, which need not be a real number in the indefinite case. Rescaling each vector of a basis by a nonzero scalar again yields a basis and preserves orthogonality, and
\[
\langle e_i, e_i \rangle = \frac{\langle b_i, b_i \rangle}{|\langle b_i, b_i \rangle|} = \pm 1 .
\]
Hence $(e_i)$ is an orthonormal basis.
\end{proof}

\begin{corollary}
\label{cor:scalar-product-signature-form}
\lean{LeeLib.Ch02.exists_basis_isOrthonormal_signature}
\leanok
\dcref{ch2:2.64}
\uses{prop:scalar-product-orthonormal-basis}
Suppose $(V, q)$ is an $n$-dimensional scalar product space. There is a basis $(\beta^i)$ for $V^*$ with respect to which $q$ has the expression
\[
q = (\beta^1)^2 + \cdots + (\beta^r)^2 - (\beta^{r+1})^2 - \cdots - (\beta^{r+s})^2,
\]
for some nonnegative integers $r, s$ with $r + s = n$.
\end{corollary}

\begin{proof}
\leanok
\uses{prop:scalar-product-orthonormal-basis}
Let $(b_i)$ be an orthonormal basis for $V$, which exists by Proposition \ref{prop:scalar-product-orthonormal-basis}, and let $(\beta^i)$ be the dual basis for $V^*$. For $v \in V$ write $v = \sum_i \beta^i(v) \, b_i$; bilinearity and orthogonality give
\[
q(v,v) = \sum_{i,j} \beta^i(v)\beta^j(v) \langle b_i, b_j\rangle = \sum_i \langle b_i, b_i \rangle \, (\beta^i(v))^2 ,
\]
so $q = \sum_i \langle b_i, b_i\rangle (\beta^i)^2$ with each coefficient $\langle b_i, b_i \rangle = \pm 1$. This is a basis expression of the required form, but perhaps with the positive and negative terms in a different order. Let $r$ be the number of indices with $\langle b_i, b_i \rangle = 1$ and $s$ the number with $\langle b_i, b_i \rangle = -1$; then $r + s = n$, since every index is of exactly one of the two kinds. Permuting the index set so that the $r$ indices of the first kind come first leaves the basis orthonormal, since orthonormality constrains only the pairwise values $\langle b_i, b_j \rangle$, and yields the stated expression.
\end{proof}

It turns out that the numbers $r$ and $s$ in (2.23) are independent of the choice of basis. The key to proving this is the following classical result from linear algebra.

\begin{proposition}[Sylvester's Law of Inertia]
\label{prop:sylvesters-law-inertia}
\lean{LeeLib.Ch02.IsOrthonormal.isGreatest_ncard}
\leanok
\dcref{ch2:2.65}
\uses{cor:scalar-product-signature-form}
Suppose $(V, q)$ is a finite-dimensional scalar product space. If $q$ has the representation (2.23) in some basis, then the number $r$ is the maximum dimension among all subspaces on which the restriction of $q$ is positive definite, and thus $r$ and $s$ are independent of the choice of basis.
\end{proposition}

\begin{proof}
\leanok
\uses{cor:scalar-product-signature-form}
Let $(b_i)$ be the orthonormal basis dual to $(\beta^i)$, so that $\langle b_i, b_i\rangle = 1$ for $i \leq r$ and $\langle b_i, b_i \rangle = -1$ for $i > r$, and $q(v,v) = \sum_i \langle b_i,b_i\rangle (\beta^i(v))^2$.

First, $r$ is attained. On $P = \operatorname{span}(b_1,\ldots,b_r)$ every $v \neq 0$ has $q(v,v) = \sum_{i \leq r} (\beta^i(v))^2 > 0$, since some coordinate $\beta^i(v)$ with $i \le r$ is nonzero. So $q|_P$ is positive definite and $\dim P = r$.

Second, $r$ is an upper bound. Let $N = \operatorname{span}(b_{r+1},\ldots,b_{r+s})$, so $\dim N = s$ and every $v \in N$ has $q(v,v) = -\sum_{i > r} (\beta^i(v))^2 \leq 0$. Let $W \subseteq V$ be any subspace on which $q$ is positive definite. Then $W \cap N = \{0\}$: a nonzero $v$ in the intersection would satisfy both $q(v,v) > 0$ and $q(v,v) \leq 0$. Hence $\dim W + \dim N \leq \dim V$, that is, $\dim W + s \leq r + s$, so $\dim W \leq r$.

Therefore $r$ is the maximum of $\dim W$ over all subspaces $W$ on which $q$ is positive definite. This characterization makes no reference to the basis, so $r$ is independent of the choice of basis; and then so is $s = n - r$.
\end{proof}

Proposition \ref{prop:sylvesters-law-inertia} characterizes $r$, and hence $s = n - r$, without reference to a basis. It is worth recording the relation $r + s = n$ of Corollary \ref{cor:scalar-product-signature-form} in those basis-free terms as well, since that is the form in which a signature can be manipulated arithmetically. The radical of a scalar product is the obstruction to doing so, and it vanishes.

\begin{lemma}[A Scalar Product Has Trivial Radical]
\label{lem:radical-scalar-product-trivial}
\lean{LeeLib.Ch02.radical_toQuadraticMap_eq_bot}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Let $(V,q)$ be a scalar product space. The radical of $q$, that is, the set of $v \in V$ with $q(v,v) = 0$ whose associated polar form $w \mapsto q(v,w) + q(w,v)$ also vanishes identically, is $\{0\}$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Let $v$ lie in the radical. Since $q$ is symmetric, the polar form at $v$ is $w \mapsto 2\langle v,w\rangle$, so its vanishing gives $2\langle v,w\rangle = 0$, hence $\langle v,w \rangle = 0$, for every $w \in V$. Thus $v$ is orthogonal to all of $V$, and Lemma \ref{lem:nondegenerate-symmetric-bilinear} forces $v = 0$.
\end{proof}

\begin{corollary}[$r + s = n$, without a basis]
\label{cor:scalar-product-signature-sum}
\lean{LeeLib.Ch02.sigPos_add_sigNeg_eq_finrank}
\leanok
\uses{lem:radical-scalar-product-trivial,prop:sylvesters-law-inertia}
Let $(V,q)$ be an $n$-dimensional scalar product space of signature $(r,s)$. Then $r + s = n$.
\end{corollary}

\begin{proof}
\leanok
\uses{lem:radical-scalar-product-trivial,prop:sylvesters-law-inertia}
For any quadratic form on a finite-dimensional space, the maximal dimension of a subspace on which it is positive definite, plus that on which it is negative definite, plus the dimension of its radical, is $\dim V$. By Proposition \ref{prop:sylvesters-law-inertia} the first two quantities are $r$ and $s$, and by Lemma \ref{lem:radical-scalar-product-trivial} the radical is trivial, so its dimension is $0$. Hence $r + s = n$.
\end{proof}

The integer $s$ in the expression (2.23) (the number of negative terms) is called the \textit{index of $q$}, and the ordered pair $(r, s)$ is called the \textit{signature of $q$}.

Now suppose $M$ is a smooth manifold. The generalization we are after replaces the inner product on each tangent space by a scalar product; that is, it drops positivity and keeps nondegeneracy. It costs nothing to state the underlying structure for an arbitrary smooth vector bundle, and there is a reason to do so: the orthonormalization results below are proved in that generality, and the normal bundle of Corollary \ref{cor:normal-bundle-pseudo-riemannian-submanifold} is not a tangent bundle.

\begin{definition}[Smooth Field of Scalar Products]
\label{def:smooth-field-scalar-products}
\lean{Bundle.ContMDiffPseudoMetric}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Let $\pi : V \to B$ be a smooth real vector bundle over a smooth manifold $B$. A \emph{smooth field of scalar products} on $V$ is a choice, for each $b \in B$, of a bilinear form $g_b$ on the fibre $V_b$ such that
\begin{enumerate}[label=(\alph*)]
\item $g_b$ is symmetric: $g_b(v,w) = g_b(w,v)$ for all $v,w \in V_b$;
\item $g_b$ is nondegenerate: for every nonzero $v \in V_b$ there is a $w \in V_b$ with $g_b(v,w) \neq 0$;
\item $b \mapsto g_b$ is a smooth section of the bundle of bilinear forms on $V$.
\end{enumerate}
Condition (b) is stated in the witness form (b) of Lemma \ref{lem:nondegenerate-symmetric-bilinear}, which is the form a construction can discharge directly.
\end{definition}

\begin{lemma}[The Fibres are Scalar Product Spaces]
\label{lem:pseudo-metric-fibre-scalar-product}
\lean{Bundle.ContMDiffPseudoMetric.bilin_nondegenerate}
\leanok
\uses{def:smooth-field-scalar-products,lem:nondegenerate-symmetric-bilinear}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$ whose fibres are finite-dimensional. Then $(V_b, g_b)$ is a scalar product space for every $b \in B$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:smooth-field-scalar-products,lem:nondegenerate-symmetric-bilinear}
The form $g_b$ is a symmetric bilinear form on the finite-dimensional space $V_b$ by conditions (a) and (c) of Definition \ref{def:smooth-field-scalar-products}. Condition (b) of that definition is verbatim condition (b) of Lemma \ref{lem:nondegenerate-symmetric-bilinear}, which is there shown equivalent to nondegeneracy of $g_b$. So $g_b$ is a nondegenerate symmetric bilinear form on a finite-dimensional vector space, which is what a scalar product is.
\end{proof}

The pairing of two smooth sections against such a field is a smooth function. This is the analytic workhorse of everything below, and the point to notice is that its proof appeals only to the smoothness of $b \mapsto g_b$: positivity plays no part, so Lemma \ref{lem:inner-product-vector-fields-smooth} carries over unchanged.

\begin{lemma}[Smoothness of the Pairing of Smooth Sections]
\label{lem:pseudo-metric-pairing-smooth}
\lean{Bundle.ContMDiffPseudoMetric.contMDiffOn_pairing, Bundle.ContMDiffPseudoMetric.contMDiff_pairing}
\leanok
\uses{def:smooth-field-scalar-products}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$, let $N$ be a smooth manifold, and let $s \subseteq N$. Let $f : N \to B$ be smooth on $s$, and let $v, w$ be sections of $V$ along $f$ — that is, assignments $x \mapsto v_x \in V_{f(x)}$ — that are smooth on $s$. Then
\[
x \longmapsto g_{f(x)}(v_x, w_x)
\]
is a smooth real-valued function on $s$. In particular, taking $s = N$, the pairing of two smooth sections along a smooth map is smooth.
\end{lemma}

\begin{proof}
\leanok
\uses{def:smooth-field-scalar-products}
Smoothness on $s$ is checked at each point of $s$, so fix $x_0 \in s$ and put $b_0 = f(x_0)$. Choose a local trivialization $\Phi$ of $V$ over an open set $U \ni b_0$, with fibrewise linear isomorphisms $\Phi_b : V_b \to F$ onto the model fibre. It induces a trivialization of the bundle of bilinear forms on $V$ over $U$, carrying a bilinear form $\beta$ on $V_b$ to the bilinear form
\[
\hat\beta = \beta\bigl(\Phi_b^{-1}\,\cdot\,, \Phi_b^{-1}\,\cdot\,\bigr)
\]
on $F$; this is again a linear isomorphism on each fibre. Since $f$ is continuous at $x_0$ there is a neighbourhood $W$ of $x_0$ in $N$ with $f(W) \subseteq U$, and on $W$
\[
g_{f(x)}(v_x,w_x) = \hat g_{f(x)}\bigl(\Phi_{f(x)}v_x,\ \Phi_{f(x)}w_x\bigr).
\]
Read in the trivialization, the three maps
\[
x \mapsto \hat g_{f(x)}, \qquad x \mapsto \Phi_{f(x)}v_x, \qquad x \mapsto \Phi_{f(x)}w_x
\]
are smooth on $W \cap s$ with values in the fixed spaces of bilinear forms on $F$, in $F$, and in $F$: the first because $b \mapsto g_b$ is a smooth section and $f$ is smooth on $s$, the second and third because $v$ and $w$ are smooth sections along $f$ on $s$. Finally, evaluation $(\beta, y, z) \mapsto \beta(y,z)$ is a continuous multilinear map, hence smooth. The displayed function is the composite of these, so it is smooth at $x_0$ within $s$. As $x_0 \in s$ was arbitrary, it is smooth on $s$.
\end{proof}

\begin{definition}[Pseudo-Riemannian Metric]
\label{def:pseudo-riemannian-metric}
\lean{LeeLib.Ch02.PseudoRiemannianMetric}
\leanok
\dcref{ch2:2.7}
\uses{def:smooth-field-scalar-products}
Let $M$ be a smooth manifold. A \emph{pseudo-Riemannian metric} on $M$ (called by some authors a \emph{semi-Riemannian metric}) is a smooth field of scalar products on the tangent bundle of $M$; that is, a smooth symmetric $2$-tensor field $g$ whose value $g_p$ at each $p \in M$ is nondegenerate on $T_pM$, so that $(T_pM, g_p)$ is a scalar product space at every point. A \emph{pseudo-Riemannian manifold} is a pair $(M,g)$ consisting of a smooth manifold together with a pseudo-Riemannian metric on it.
\end{definition}

Lee's definition asks in addition that $g$ have \emph{the same signature everywhere}. That requirement is carried separately here, by Definition \ref{def:pseudo-metric-has-signature} below, and a ``pseudo-Riemannian manifold of signature $(r,s)$'' is a pseudo-Riemannian manifold together with the assertion that its signature is constantly $(r,s)$. The two are kept apart because Propositions \ref{prop:pseudo-riemannian-hypersurface-signature} and \ref{cor:pseudo-riemannian-embedding-criterion} each exhibit a metric first and then \emph{conclude} that its signature is a particular constant; that conclusion is only sayable if a metric can be spoken of before its signature is known to be constant.

\begin{proposition}[Every Riemannian Metric is Pseudo-Riemannian]
\label{prop:riemannian-is-pseudo-riemannian}
\lean{LeeLib.Ch02.RiemannianMetric.toPseudoRiemannianMetric}
\leanok
\dcref{ch2:2.7}
\uses{def:riemannian-metric,def:pseudo-riemannian-metric}
Every Riemannian metric on a smooth manifold $M$ is a pseudo-Riemannian metric on $M$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:riemannian-metric,def:pseudo-riemannian-metric,lem:nondegenerate-symmetric-bilinear}
Let $g$ be a Riemannian metric. By Definition \ref{def:riemannian-metric} it is a smooth symmetric $2$-tensor field, so conditions (a) and (c) of Definition \ref{def:smooth-field-scalar-products} hold and only nondegeneracy at each point requires argument. Fix $p \in M$ and let $v \in T_pM$ be nonzero. Take $w = v$: positive definiteness of $g_p$ gives $g_p(v,v) > 0$, so in particular $g_p(v,w) \neq 0$. This is condition (b) of Lemma \ref{lem:nondegenerate-symmetric-bilinear}, hence $g_p$ is nondegenerate.
\end{proof}

The witness $w = v$ used in that proof is exactly the step that fails for an indefinite scalar product: there a nonzero vector may be null, that is, orthogonal to itself, so $\langle v,v\rangle \neq 0$ does not follow from $v \neq 0$. Every difficulty in this section traces back to that one line.

\begin{definition}[Signature of a Pseudo-Riemannian Metric at a Point]
\label{def:pseudo-metric-signature-at}
\lean{LeeLib.Ch02.PseudoRiemannianMetric.sigPosAt, LeeLib.Ch02.PseudoRiemannianMetric.sigNegAt}
\leanok
\uses{def:pseudo-riemannian-metric,lem:pseudo-metric-fibre-scalar-product,cor:scalar-product-signature-form,prop:sylvesters-law-inertia}
Let $(M,g)$ be a pseudo-Riemannian manifold and let $p \in M$. By Lemma \ref{lem:pseudo-metric-fibre-scalar-product} the pair $(T_pM, g_p)$ is a scalar product space, so by Corollary \ref{cor:scalar-product-signature-form} it has a signature $(r_p, s_p)$, which by Proposition \ref{prop:sylvesters-law-inertia} does not depend on the basis used to compute it. We call $(r_p,s_p)$ the \emph{signature of $g$ at $p$}, and $s_p$ the \emph{index of $g$ at $p$}. Equivalently, by Proposition \ref{prop:sylvesters-law-inertia}, $r_p$ is the largest dimension of a subspace of $T_pM$ on which $g_p$ is positive definite, and $s_p$ the largest dimension of one on which $g_p$ is negative definite.
\end{definition}

\begin{proposition}[The Signature at a Point Sums to the Dimension]
\label{prop:pseudo-metric-signature-sum}
\lean{LeeLib.Ch02.PseudoRiemannianMetric.sigPosAt_add_sigNegAt}
\leanok
\uses{def:pseudo-metric-signature-at,lem:pseudo-metric-fibre-scalar-product,cor:scalar-product-signature-sum}
Let $(M,g)$ be a pseudo-Riemannian manifold, let $p \in M$, and let $(r_p,s_p)$ be the signature of $g$ at $p$. Then $r_p + s_p = \dim T_pM$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:pseudo-metric-fibre-scalar-product,cor:scalar-product-signature-sum}
By Lemma \ref{lem:pseudo-metric-fibre-scalar-product} the pair $(T_pM, g_p)$ is a scalar product space, and $(r_p,s_p)$ is by Definition \ref{def:pseudo-metric-signature-at} its signature. Corollary \ref{cor:scalar-product-signature-sum} applied to it gives $r_p + s_p = \dim T_pM$.
\end{proof}

\begin{definition}[Constant Signature]
\label{def:pseudo-metric-has-signature}
\lean{LeeLib.Ch02.HasSignature}
\leanok
\uses{def:pseudo-metric-signature-at}
Let $(M,g)$ be a pseudo-Riemannian manifold and let $r,s \geq 0$ be integers. We say that $g$ \emph{has signature $(r,s)$} if the signature of $g$ at $p$ is $(r,s)$ for every $p \in M$. A \emph{pseudo-Riemannian manifold of signature $(r,s)$} is a pseudo-Riemannian manifold $(M,g)$ whose metric has signature $(r,s)$.
\end{definition}

By Proposition \ref{prop:pseudo-metric-signature-sum}, a metric of signature $(r,s)$ on a manifold of dimension $n$ satisfies $r + s = n$, so on a fixed manifold a signature has only one degree of freedom.

\subsection{The Gram--Schmidt Algorithm as an Explicit Recursion}

Proposition \ref{prop:gram-schmidt-algorithm-scalar-products} orthonormalizes a nondegenerate basis, but it does so as an \emph{existence} statement: its induction produces \emph{some} orthonormal basis, and asserts nothing about how that basis depends on the input. For the frames of Proposition \ref{prop:orthonormal-frames-pseudo-riemannian} that is not enough. There the input tuple varies with a point of a manifold and the output must be shown to vary \emph{smoothly} with it; smoothness is a property of a function, and there is no function to speak of until the orthonormalization is exhibited as a formula in its input. We therefore rerun the algorithm of Proposition \ref{prop:gram-schmidt-algorithm-scalar-products} as an explicit recursion and reprove its properties for that formula.

\begin{definition}[The Gram--Schmidt Recursion for a Bilinear Form]
\label{def:gram-schmidt-bilin}
\lean{LeeLib.Ch02.gramSchmidtBilin}
\leanok
Let $V$ be a real vector space, let $B$ be a bilinear form on $V$, and let $(v_1,\ldots,v_n)$ be an $n$-tuple in $V$. Define the $n$-tuple $u = \mathrm{GS}_B(v)$ by the recursion
\[
u_j = v_j - \sum_{i < j} \frac{B(u_i, v_j)}{B(u_i,u_i)} \, u_i .
\]
This is well founded, $u_j$ depending only on those $u_i$ with $i < j$. The formula is used only under the hypothesis of Proposition \ref{prop:gram-schmidt-bilin-spec}, which guarantees that no denominator $B(u_i,u_i)$ vanishes.
\end{definition}

\begin{definition}[The Normalized Gram--Schmidt Recursion]
\label{def:gram-schmidt-bilin-normed}
\lean{LeeLib.Ch02.gramSchmidtBilinNormed}
\leanok
\uses{def:gram-schmidt-bilin}
With $V$, $B$ and $(v_1,\ldots,v_n)$ as in Definition \ref{def:gram-schmidt-bilin} and $u = \mathrm{GS}_B(v)$, define the $n$-tuple $b = \widehat{\mathrm{GS}}_B(v)$ by
\[
b_j = \frac{u_j}{\sqrt{\,|B(u_j,u_j)|\,}} .
\]
As in Proposition \ref{prop:gram-schmidt-algorithm-scalar-products}, one divides by $\sqrt{|B(u_j,u_j)|}$ rather than by $|u_j|$, which need not be a real number in the indefinite case. The effect is that $B(b_j,b_j) = \pm 1$ rather than $= 1$.
\end{definition}

The next lemma is the computation at the heart of the algorithm, isolated from the induction that uses it.

\begin{lemma}[The New Vector is Orthogonal to its Predecessors]
\label{lem:gram-schmidt-bilin-orthogonal-earlier}
\lean{LeeLib.Ch02.gramSchmidtBilin_apply_of_lt}
\leanok
\uses{def:gram-schmidt-bilin}
Let $B$ be a bilinear form on a real vector space $V$, let $(v_1,\ldots,v_n)$ be an $n$-tuple in $V$, and write $u = \mathrm{GS}_B(v)$. Fix $k$ and suppose
\[
B(u_i,u_i) \neq 0 \quad\text{for every } i \leq k, \qquad B(u_i,u_{i'}) = 0 \quad\text{for all } i \neq i' \text{ with } i,i' \leq k .
\]
Then, writing $j = k+1$, we have $B(u_c, u_j) = 0$ for every $c < j$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:gram-schmidt-bilin}
Let $c < j$, so that $c \leq k$. Expanding the recursion of Definition \ref{def:gram-schmidt-bilin} at $j$ and using linearity of $B$ in its second argument,
\[
B(u_c, u_j) = B(u_c, v_j) - \sum_{i < j} \frac{B(u_i,v_j)}{B(u_i,u_i)} \, B(u_c, u_i) .
\]
Every index $i < j$ satisfies $i \leq k$, as does $c$; so by the second hypothesis $B(u_c,u_i) = 0$ for every such $i$ other than $i = c$, and the sum collapses to its single term $i = c$. Hence
\[
B(u_c,u_j) = B(u_c,v_j) - \frac{B(u_c,v_j)}{B(u_c,u_c)}\, B(u_c,u_c) = B(u_c,v_j) - B(u_c,v_j) = 0 ,
\]
the cancellation being legitimate because $B(u_c,u_c) \neq 0$ by the first hypothesis.
\end{proof}

The nonvanishing of the denominators, the pairwise orthogonality of the $u_j$, and the reproduction of the flag are proved together, by a single induction. They cannot be separated: orthogonality of $u_j$ against its predecessors needs their denominators to be nonzero, the denominator $B(u_j,u_j) \neq 0$ needs $\operatorname{span}(u_1,\ldots,u_j) = \operatorname{span}(v_1,\ldots,v_j)$, and that identity needs $u_j \equiv v_j$ modulo the earlier span.

\begin{proposition}[The Gram--Schmidt Recursion: Simultaneous Induction]
\label{prop:gram-schmidt-bilin-spec}
\lean{LeeLib.Ch02.gramSchmidtBilin_spec}
\leanok
\uses{def:gram-schmidt-bilin,lem:gram-schmidt-bilin-orthogonal-earlier,lem:nondegenerate-symmetric-bilinear}
Let $V$ be a real vector space, let $B$ be a symmetric bilinear form on $V$, and let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$. Write $u = \mathrm{GS}_B(v)$. Then for every $k \leq n$ the following three statements hold:
\begin{enumerate}[label=(\alph*)]
\item $B(u_i,u_i) \neq 0$ for every $i \leq k$;
\item $B(u_i,u_{i'}) = 0$ for all $i \neq i'$ with $i,i' \leq k$;
\item $\operatorname{span}(u_1,\ldots,u_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq k$.
\end{enumerate}
\end{proposition}

\begin{proof}
\leanok
\uses{def:gram-schmidt-bilin,lem:gram-schmidt-bilin-orthogonal-earlier,lem:nondegenerate-symmetric-bilinear}
Induct on $k$. For $k = 0$, statements (a) and (b) are vacuous, and (c) holds for the only admissible index $j = 0$, both spans being the span of the empty family, namely $\{0\}$.

Assume (a), (b), (c) hold for some $k$ with $k + 1 \leq n$, and put $j = k+1$ and $S = \operatorname{span}(v_1,\ldots,v_k)$. By the inductive hypothesis (c), $S = \operatorname{span}(u_1,\ldots,u_k)$, and $\dim S = k$ because the tuple $(v_i)$ is nondegenerate.

\emph{The vector $u_j$ is orthogonal to $S$.} The inductive hypotheses (a) and (b) are exactly the hypotheses of Lemma \ref{lem:gram-schmidt-bilin-orthogonal-earlier}, which therefore gives $B(u_c,u_j) = 0$ for every $c < j$, that is, for every $c \leq k$. By symmetry of $B$ also $B(u_j,u_c) = 0$ for every $c \leq k$; since $B$ is linear in its second argument and the $u_c$ with $c \leq k$ span $S$, it follows that $B(u_j, y) = 0$ for every $y \in S$.

\emph{The vector $u_j$ agrees with $v_j$ modulo $S$.} Write $\sigma = \sum_{i<j} \frac{B(u_i,v_j)}{B(u_i,u_i)} u_i$, so that $u_j = v_j - \sigma$ by Definition \ref{def:gram-schmidt-bilin}. Each index $i < j$ satisfies $i \leq k$, so $\sigma$ is a linear combination of $u_1,\ldots,u_k$ and hence lies in $S$.

\emph{The vector $v_j$ is not in $S$.} If it were, then $\operatorname{span}(v_1,\ldots,v_j) \subseteq S$; the reverse inclusion always holds, so $\operatorname{span}(v_1,\ldots,v_j) = S$. But nondegeneracy of the tuple $(v_i)$ gives $\dim \operatorname{span}(v_1,\ldots,v_j) = j = k+1$, while $\dim S = k$, a contradiction.

\emph{Hence $u_j \neq 0$}: otherwise $v_j = \sigma \in S$, contrary to the previous paragraph.

\emph{The span identity.} Put $T = \operatorname{span}(v_1,\ldots,v_j)$. Since $u_j = v_j - \sigma$ with $\sigma \in S \subseteq T$ and $v_j \in T$, we get $u_j \in T$, whence $S + \langle u_j\rangle \subseteq T$. Conversely $v_j = u_j + \sigma \in S + \langle u_j \rangle$ and $v_1,\ldots,v_k \in S$, so $T \subseteq S + \langle u_j\rangle$. Therefore
\[
T = S + \langle u_j \rangle .
\]

\emph{The denominator does not vanish:} $B(u_j,u_j) \neq 0$. This is the one point at which the indefinite case genuinely differs, $u_j \neq 0$ no longer sufficing. Suppose $B(u_j,u_j) = 0$. By the span identity every $w \in T$ can be written $w = y + c\,u_j$ with $y \in S$ and $c \in \mathbb{R}$, and then
\[
B(u_j, w) = B(u_j, y) + c\,B(u_j,u_j) = 0 ,
\]
the first term vanishing because $u_j$ is orthogonal to $S$. So $u_j$ is an element of $T$ orthogonal to all of $T$. As $T$ is nondegenerate, being a member of the flag of the nondegenerate tuple $(v_i)$, Lemma \ref{lem:nondegenerate-symmetric-bilinear} forces $u_j = 0$, contradicting the previous paragraph.

\emph{Assembling the three statements at $k+1$.} For (a), let $i \leq k+1$: if $i \leq k$ this is the inductive hypothesis, and if $i = j$ it is the previous paragraph. For (b), let $i \neq i'$ with $i,i' \leq k+1$; not both can equal $j$, so after using the symmetry of $B$ we may assume $i' \leq k$. If also $i \leq k$ this is the inductive hypothesis; if $i = j$ it is the first paragraph. For (c), let $m \leq k+1$: if $m \leq k$ this is the inductive hypothesis, and if $m = k+1$ then, by the inductive hypothesis (c) and the span identity,
\[
\operatorname{span}(u_1,\ldots,u_{k+1}) = \operatorname{span}(u_1,\ldots,u_k) + \langle u_j\rangle = S + \langle u_j \rangle = T = \operatorname{span}(v_1,\ldots,v_{k+1}) . \qedhere
\]
\end{proof}

Taking $k = n$ in Proposition \ref{prop:gram-schmidt-bilin-spec} gives the three statements in the form in which they are used.

\begin{lemma}[The Denominators Never Vanish]
\label{lem:gram-schmidt-bilin-nonvanishing}
\lean{LeeLib.Ch02.gramSchmidtBilin_apply_self_ne_zero}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
Let $B$ be a symmetric bilinear form on a real vector space $V$, let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$, and let $u = \mathrm{GS}_B(v)$. Then $B(u_j,u_j) \neq 0$ for every $j$.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
This is part (a) of Proposition \ref{prop:gram-schmidt-bilin-spec} with $k = n$.
\end{proof}

This statement has no positive definite counterpart: for an inner product, $B(u_j,u_j) \neq 0$ follows from $u_j \neq 0$ alone, and $u_j \neq 0$ follows from linear independence. It is the reason nondegeneracy of the whole flag, and not mere linear independence, is the right hypothesis for an indefinite Gram--Schmidt.

\begin{lemma}[The Unnormalized Vectors are Orthogonal]
\label{lem:gram-schmidt-bilin-orthogonal}
\lean{LeeLib.Ch02.gramSchmidtBilin_apply_ne}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
With $B$, $(v_i)$ and $u$ as in Lemma \ref{lem:gram-schmidt-bilin-nonvanishing}, $B(u_i,u_j) = 0$ whenever $i \neq j$.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
This is part (b) of Proposition \ref{prop:gram-schmidt-bilin-spec} with $k = n$.
\end{proof}

\begin{lemma}[The Unnormalized Vectors Reproduce the Flag]
\label{lem:gram-schmidt-bilin-flag}
\lean{LeeLib.Ch02.prefixSpan_gramSchmidtBilin}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
With $B$, $(v_i)$ and $u$ as in Lemma \ref{lem:gram-schmidt-bilin-nonvanishing}, $\operatorname{span}(u_1,\ldots,u_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq n$.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:gram-schmidt-bilin-spec}
This is part (c) of Proposition \ref{prop:gram-schmidt-bilin-spec} with $k = n$.
\end{proof}

\begin{lemma}[The Normalizing Scalar is Nonzero]
\label{lem:gram-schmidt-bilin-normalizer-ne-zero}
\lean{LeeLib.Ch02.gramSchmidtBilin_normalizer_ne_zero}
\leanok
\uses{lem:gram-schmidt-bilin-nonvanishing,def:gram-schmidt-bilin-normed}
With $B$, $(v_i)$ and $u$ as in Lemma \ref{lem:gram-schmidt-bilin-nonvanishing}, the scalar $\bigl(\sqrt{|B(u_j,u_j)|}\bigr)^{-1}$ is nonzero for every $j$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gram-schmidt-bilin-nonvanishing}
By Lemma \ref{lem:gram-schmidt-bilin-nonvanishing}, $B(u_j,u_j) \neq 0$, so $|B(u_j,u_j)| > 0$ and therefore $\sqrt{|B(u_j,u_j)|} > 0$. The inverse of a nonzero real number is nonzero.
\end{proof}

\begin{proposition}[The Recursion Produces an Orthonormal Tuple]
\label{prop:gram-schmidt-bilin-orthonormal}
\lean{LeeLib.Ch02.gramSchmidtBilin_isOrthonormal}
\leanok
\uses{def:gram-schmidt-bilin-normed,lem:gram-schmidt-bilin-nonvanishing,lem:gram-schmidt-bilin-orthogonal}
Let $B$ be a symmetric bilinear form on a real vector space $V$, let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$, and let $b = \widehat{\mathrm{GS}}_B(v)$. Then $(b_1,\ldots,b_n)$ is orthonormal: $B(b_i,b_j) = 0$ for $i \neq j$, and $B(b_j,b_j) = \pm 1$ for every $j$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:gram-schmidt-bilin-normed,lem:gram-schmidt-bilin-nonvanishing,lem:gram-schmidt-bilin-orthogonal}
Write $u = \mathrm{GS}_B(v)$ and $\lambda_j = \bigl(\sqrt{|B(u_j,u_j)|}\bigr)^{-1}$, so that $b_j = \lambda_j u_j$ by Definition \ref{def:gram-schmidt-bilin-normed}. By Lemma \ref{lem:gram-schmidt-bilin-nonvanishing} we have $B(u_j,u_j) \neq 0$, hence $|B(u_j,u_j)| > 0$ and $\lambda_j^2 = |B(u_j,u_j)|^{-1}$.

For $i \neq j$, bilinearity and Lemma \ref{lem:gram-schmidt-bilin-orthogonal} give
\[
B(b_i,b_j) = \lambda_i \lambda_j \, B(u_i,u_j) = 0 .
\]
For each $j$, bilinearity gives
\[
B(b_j,b_j) = \lambda_j^2 \, B(u_j,u_j) = \frac{B(u_j,u_j)}{|B(u_j,u_j)|} = \pm 1 ,
\]
the sign being $+1$ if $B(u_j,u_j) > 0$ and $-1$ if $B(u_j,u_j) < 0$.
\end{proof}

\begin{lemma}[The Normalized Tuple Reproduces the Flag]
\label{lem:gram-schmidt-bilin-normed-flag}
\lean{LeeLib.Ch02.prefixSpan_gramSchmidtBilinNormed}
\leanok
\uses{def:gram-schmidt-bilin-normed,lem:gram-schmidt-bilin-flag,lem:gram-schmidt-bilin-normalizer-ne-zero}
With $B$, $(v_i)$ and $b = \widehat{\mathrm{GS}}_B(v)$ as in Proposition \ref{prop:gram-schmidt-bilin-orthonormal}, $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq n$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gram-schmidt-bilin-flag,lem:gram-schmidt-bilin-normalizer-ne-zero}
By Lemma \ref{lem:gram-schmidt-bilin-flag} it suffices to show $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(u_1,\ldots,u_j)$, where $u = \mathrm{GS}_B(v)$. Each $b_i = \lambda_i u_i$ with $\lambda_i \neq 0$ by Lemma \ref{lem:gram-schmidt-bilin-normalizer-ne-zero}, so $b_i \in \operatorname{span}(u_1,\ldots,u_j)$ for $i \leq j$, giving one inclusion; and $u_i = \lambda_i^{-1} b_i \in \operatorname{span}(b_1,\ldots,b_j)$ for $i \leq j$, giving the other. Rescaling the members of a family by nonzero scalars therefore leaves every span unchanged.
\end{proof}

\begin{corollary}[Proposition \ref{prop:gram-schmidt-algorithm-scalar-products}, Recovered from the Recursion]
\label{cor:gram-schmidt-bilin-recovers}
\lean{LeeLib.Ch02.exists_isOrthonormal_prefixSpan_eq_of_gramSchmidt}
\leanok
\uses{prop:gram-schmidt-bilin-orthonormal,lem:gram-schmidt-bilin-normed-flag}
Let $B$ be a symmetric bilinear form on a real vector space $V$ and let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$. Then there is an orthonormal $n$-tuple $(b_1,\ldots,b_n)$ in $V$ with $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq n$.
\end{corollary}

\begin{proof}
\leanok
\uses{prop:gram-schmidt-bilin-orthonormal,lem:gram-schmidt-bilin-normed-flag}
Take $b = \widehat{\mathrm{GS}}_B(v)$. It is orthonormal by Proposition \ref{prop:gram-schmidt-bilin-orthonormal} and reproduces the flag by Lemma \ref{lem:gram-schmidt-bilin-normed-flag}.
\end{proof}

This confirms that the recursion computes what the existence proof of Proposition \ref{prop:gram-schmidt-algorithm-scalar-products} asserts; the point of the recursion is that, unlike that proof, it has an output one can differentiate.

\subsection{Smoothness of the Fibrewise Gram--Schmidt Algorithm}

Two ingredients are needed to run the recursion of Definition \ref{def:gram-schmidt-bilin} fibrewise and conclude that the output is smooth. The first is Lemma \ref{lem:pseudo-metric-pairing-smooth}, which makes the pairings appearing in the recursion smooth functions of the base point. The second concerns normalization: the Riemannian proof of Proposition \ref{prop:existence-orthonormal-frames} divides by $\sqrt{\langle u_j,u_j\rangle}$, whose argument is positive, whereas here one divides by $\sqrt{|B(u_j,u_j)|}$, and $|\cdot|$ is not differentiable at $0$. It does not need to be.

\begin{lemma}[$(\sqrt{|\cdot|})^{-1}$ is Smooth Away from the Origin]
\label{lem:inv-sqrt-abs-smooth}
\lean{LeeLib.Ch02.contDiffAt_inv_sqrt_abs}
\leanok
Let $t \in \mathbb{R}$ with $t \neq 0$. Then $s \mapsto \bigl(\sqrt{|s|}\bigr)^{-1}$ is smooth at $t$.
\end{lemma}

\begin{proof}
\leanok
First, $s \mapsto |s|$ is smooth at $t$. If $t < 0$ then $|s| = -s$ for every $s$ in the neighbourhood $(-\infty,0)$ of $t$; if $t > 0$ then $|s| = s$ for every $s$ in the neighbourhood $(0,\infty)$ of $t$. In either case $|\cdot|$ agrees near $t$ with a linear function, and smoothness at a point depends only on the germ there.

Next, $|t| > 0$ and $r \mapsto \sqrt{r}$ is smooth on $(0,\infty)$, hence smooth at $|t|$; so the composite $s \mapsto \sqrt{|s|}$ is smooth at $t$.

Finally $\sqrt{|t|} > 0$, in particular $\sqrt{|t|} \neq 0$, and $r \mapsto r^{-1}$ is smooth at every nonzero real number. Composing once more, $s \mapsto (\sqrt{|s|})^{-1}$ is smooth at $t$.
\end{proof}

\begin{definition}[Pointwise Nondegenerate Smooth Family]
\label{def:local-nondegenerate-family}
\lean{LeeLib.Ch02.IsLocalNondegenerateOn}
\leanok
\uses{def:smooth-field-scalar-products,lem:pseudo-metric-fibre-scalar-product}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$ with finite-dimensional fibres, and let $u \subseteq B$. A family $(X_1,\ldots,X_m)$ of sections of $V$ is \emph{nondegenerate on $u$} if
\begin{enumerate}[label=(\alph*)]
\item each $X_i$ is smooth on $u$; and
\item for every $x \in u$ the tuple $(X_1|_x,\ldots,X_m|_x)$ is a nondegenerate $m$-tuple in the scalar product space $(V_x, g_x)$.
\end{enumerate}
\end{definition}

Condition (b) is strictly stronger than pointwise linear independence: it asks each $\operatorname{span}(X_1|_x,\ldots,X_k|_x)$ to be a nondegenerate subspace of $V_x$, not merely of dimension $k$. This is the indefinite replacement for the hypothesis ``$(X_i)$ is a local frame'' of Proposition \ref{prop:existence-orthonormal-frames}. The strengthening is forced: a linearly independent tuple of null vectors is impossible when $g_x$ is an inner product but routine for an indefinite scalar product, and on such a tuple the recursion of Definition \ref{def:gram-schmidt-bilin} divides by zero.

\begin{definition}[The Fibrewise Gram--Schmidt Frame]
\label{def:pseudo-gram-schmidt-frame}
\lean{LeeLib.Ch02.pseudoGramSchmidtFrameAux, LeeLib.Ch02.pseudoGramSchmidtFrame}
\leanok
\uses{def:gram-schmidt-bilin,def:gram-schmidt-bilin-normed,def:smooth-field-scalar-products}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$ and let $(X_1,\ldots,X_m)$ be a family of sections of $V$. Applying Definitions \ref{def:gram-schmidt-bilin} and \ref{def:gram-schmidt-bilin-normed} in each fibre against the form $g_x$, define families $(U_1,\ldots,U_m)$ and $(E_1,\ldots,E_m)$ of sections of $V$ by
\[
U_j|_x = \mathrm{GS}_{g_x}\bigl(X_1|_x,\ldots,X_m|_x\bigr)_j , \qquad
E_j|_x = \widehat{\mathrm{GS}}_{g_x}\bigl(X_1|_x,\ldots,X_m|_x\bigr)_j = \frac{U_j|_x}{\sqrt{\,|g_x(U_j|_x, U_j|_x)|\,}} .
\]
\end{definition}

\begin{lemma}[Nonvanishing of the Fibrewise Denominators]
\label{lem:pseudo-gram-schmidt-denominators}
\lean{LeeLib.Ch02.pseudoGramSchmidtFrameAux_apply_self_ne_zero}
\leanok
\uses{def:pseudo-gram-schmidt-frame,def:local-nondegenerate-family,lem:gram-schmidt-bilin-nonvanishing,lem:pseudo-metric-fibre-scalar-product}
Let $g$ be a smooth field of scalar products on $V \to B$ with finite-dimensional fibres, and let $(X_1,\ldots,X_m)$ be nondegenerate on $u \subseteq B$. Then $g_x(U_j|_x, U_j|_x) \neq 0$ for every $x \in u$ and every $j$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gram-schmidt-bilin-nonvanishing,lem:pseudo-metric-fibre-scalar-product}
Fix $x \in u$. By Lemma \ref{lem:pseudo-metric-fibre-scalar-product} the form $g_x$ is a symmetric bilinear form on $V_x$, and by condition (b) of Definition \ref{def:local-nondegenerate-family} the tuple $(X_1|_x,\ldots,X_m|_x)$ is nondegenerate. Lemma \ref{lem:gram-schmidt-bilin-nonvanishing}, applied in $V_x$ to $B = g_x$ and this tuple, gives the claim.
\end{proof}

\begin{proposition}[Orthonormality of the Gram--Schmidt Frame]
\label{prop:pseudo-gram-schmidt-frame-orthonormal}
\lean{LeeLib.Ch02.pseudoGramSchmidtFrame_isOrthonormal}
\leanok
\uses{def:pseudo-gram-schmidt-frame,def:local-nondegenerate-family,prop:gram-schmidt-bilin-orthonormal,lem:pseudo-metric-fibre-scalar-product}
With $g$ and $(X_i)$ as in Lemma \ref{lem:pseudo-gram-schmidt-denominators}, the tuple $(E_1|_x,\ldots,E_m|_x)$ is orthonormal in $(V_x, g_x)$ for every $x \in u$; that is, $g_x(E_i|_x, E_j|_x) = \pm\delta_{ij}$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:gram-schmidt-bilin-orthonormal,lem:pseudo-metric-fibre-scalar-product}
Fix $x \in u$. As in the proof of Lemma \ref{lem:pseudo-gram-schmidt-denominators}, $g_x$ is a symmetric bilinear form on $V_x$ and $(X_1|_x,\ldots,X_m|_x)$ is a nondegenerate tuple. By Definition \ref{def:pseudo-gram-schmidt-frame}, $(E_1|_x,\ldots,E_m|_x)$ is exactly the tuple $\widehat{\mathrm{GS}}_{g_x}(X_1|_x,\ldots,X_m|_x)$, which Proposition \ref{prop:gram-schmidt-bilin-orthonormal} shows to be orthonormal.
\end{proof}

\begin{lemma}[The Gram--Schmidt Frame Reproduces the Flag]
\label{lem:pseudo-gram-schmidt-frame-flag}
\lean{LeeLib.Ch02.prefixSpan_pseudoGramSchmidtFrame}
\leanok
\uses{def:pseudo-gram-schmidt-frame,def:local-nondegenerate-family,lem:gram-schmidt-bilin-normed-flag,lem:pseudo-metric-fibre-scalar-product}
With $g$ and $(X_i)$ as in Lemma \ref{lem:pseudo-gram-schmidt-denominators}, for every $x \in u$ and every $k \leq m$,
\[
\operatorname{span}\bigl(E_1|_x,\ldots,E_k|_x\bigr) = \operatorname{span}\bigl(X_1|_x,\ldots,X_k|_x\bigr) .
\]
\end{lemma}

\begin{proof}
\leanok
\uses{lem:gram-schmidt-bilin-normed-flag,lem:pseudo-metric-fibre-scalar-product}
Fix $x \in u$. As above, $g_x$ is a symmetric bilinear form on $V_x$ and $(X_1|_x,\ldots,X_m|_x)$ is a nondegenerate tuple, and $(E_j|_x)_j$ is $\widehat{\mathrm{GS}}_{g_x}$ of it. Lemma \ref{lem:gram-schmidt-bilin-normed-flag} applied in $V_x$ gives the claim.
\end{proof}

\begin{proposition}[Smoothness of the Unnormalized Gram--Schmidt Sections]
\label{prop:pseudo-gram-schmidt-frame-aux-smooth}
\lean{LeeLib.Ch02.contMDiffOn_pseudoGramSchmidtFrameAux}
\leanok
\uses{def:pseudo-gram-schmidt-frame,def:local-nondegenerate-family,lem:pseudo-metric-pairing-smooth,lem:pseudo-gram-schmidt-denominators}
With $g$ and $(X_i)$ as in Lemma \ref{lem:pseudo-gram-schmidt-denominators}, each section $U_j$ is smooth on $u$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:pseudo-gram-schmidt-frame,lem:pseudo-metric-pairing-smooth,lem:pseudo-gram-schmidt-denominators}
By strong induction on $j$: assume $U_i$ is smooth on $u$ for every $i < j$. By Definition \ref{def:pseudo-gram-schmidt-frame} and Definition \ref{def:gram-schmidt-bilin},
\[
U_j|_x = X_j|_x - \sum_{i<j} \frac{g_x(U_i|_x, X_j|_x)}{g_x(U_i|_x, U_i|_x)} \, U_i|_x .
\]
Fix $i < j$. The functions
\[
x \mapsto g_x(U_i|_x, X_j|_x) \qquad\text{and}\qquad x \mapsto g_x(U_i|_x, U_i|_x)
\]
are smooth on $u$ by Lemma \ref{lem:pseudo-metric-pairing-smooth}, applied with $f$ the identity of $B$: its hypotheses hold because $U_i$ is smooth on $u$ by the inductive hypothesis and $X_j$ is smooth on $u$ by condition (a) of Definition \ref{def:local-nondegenerate-family}. By Lemma \ref{lem:pseudo-gram-schmidt-denominators} the second of them is nowhere zero on $u$, so the quotient
\[
c_i : x \mapsto \frac{g_x(U_i|_x, X_j|_x)}{g_x(U_i|_x, U_i|_x)}
\]
is smooth on $u$, a quotient of smooth real-valued functions with nonvanishing denominator.

Therefore each $c_i \cdot U_i$ is a smooth section on $u$, being a smooth scalar multiple of a smooth section; the finite sum $\sum_{i<j} c_i U_i$ is a smooth section on $u$; and $U_j = X_j - \sum_{i<j} c_i U_i$ is smooth on $u$, being a difference of smooth sections.
\end{proof}

\begin{proposition}[Smoothness of the Gram--Schmidt Frame]
\label{prop:pseudo-gram-schmidt-frame-smooth}
\lean{LeeLib.Ch02.contMDiffOn_pseudoGramSchmidtFrame}
\leanok
\uses{prop:pseudo-gram-schmidt-frame-aux-smooth,lem:inv-sqrt-abs-smooth,lem:pseudo-metric-pairing-smooth,lem:pseudo-gram-schmidt-denominators,def:pseudo-gram-schmidt-frame}
With $g$ and $(X_i)$ as in Lemma \ref{lem:pseudo-gram-schmidt-denominators}, each section $E_j$ is smooth on $u$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:pseudo-gram-schmidt-frame-aux-smooth,lem:inv-sqrt-abs-smooth,lem:pseudo-metric-pairing-smooth,lem:pseudo-gram-schmidt-denominators}
By Proposition \ref{prop:pseudo-gram-schmidt-frame-aux-smooth} the section $U_j$ is smooth on $u$, so by Lemma \ref{lem:pseudo-metric-pairing-smooth} the real-valued function
\[
q : x \mapsto g_x(U_j|_x, U_j|_x)
\]
is smooth on $u$, and by Lemma \ref{lem:pseudo-gram-schmidt-denominators} it is nowhere zero there. Lemma \ref{lem:inv-sqrt-abs-smooth} shows that $s \mapsto (\sqrt{|s|})^{-1}$ is smooth at every nonzero real number, hence at $q(x)$ for every $x \in u$. Therefore the composite
\[
x \mapsto \bigl(\sqrt{|q(x)|}\bigr)^{-1}
\]
is smooth on $u$. By Definition \ref{def:pseudo-gram-schmidt-frame}, $E_j$ is the product of this smooth real-valued function with the smooth section $U_j$, hence is a smooth section on $u$.
\end{proof}

Note where positivity was and was not missed. The pairings are smooth without it, by Lemma \ref{lem:pseudo-metric-pairing-smooth}; the quotients are smooth because their denominators are \emph{nonvanishing}, which is Lemma \ref{lem:pseudo-gram-schmidt-denominators}, and not because they are positive. Only the normalization notices the difference, and Lemma \ref{lem:inv-sqrt-abs-smooth} absorbs it.

\subsection{Orthonormal Frames}

The results above orthonormalize a family that is \emph{given} to be nondegenerate at every point of $u$. Proposition \ref{prop:orthonormal-frames-pseudo-riemannian} additionally requires such a family to exist near each point, which the Riemannian case obtains for free — any local frame will do there — but which here is a genuine step. It rests on three lemmas.

\begin{lemma}[Nondegeneracy of a Tuple via its Gram Matrix]
\label{lem:nondegenerate-tuple-gram-matrix}
\lean{LeeLib.Ch02.isNondegenerate_span_iff_det_gramMatrix_ne_zero}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Let $(V,q)$ be a scalar product space and let $(v_1,\ldots,v_k)$ be a $k$-tuple in $V$. Then $\operatorname{span}(v_1,\ldots,v_k)$ is a nondegenerate subspace of dimension $k$ if and only if the Gram matrix $G = \bigl(\langle v_i, v_j\rangle\bigr)_{i,j=1}^{k}$ is invertible.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Suppose $G$ is invertible, and write $S = \operatorname{span}(v_1,\ldots,v_k)$. If $\sum_i a_i v_i = 0$ then pairing with $v_j$ gives $\sum_i a_i \langle v_j,v_i\rangle = 0$ for each $j$, that is, $Ga = 0$, whence $a = 0$; so the $v_i$ are linearly independent and $\dim S = k$. If $y = \sum_i a_i v_i \in S$ is orthogonal to all of $S$, then in particular $\langle v_j, y\rangle = 0$ for each $j$, which is again $Ga = 0$, so $a = 0$ and $y = 0$. By Lemma \ref{lem:nondegenerate-symmetric-bilinear} the restriction $q|_{S\times S}$ is therefore nondegenerate.

Conversely, suppose $S$ is nondegenerate of dimension $k$. Then $(v_1,\ldots,v_k)$ is a basis of $S$, and $G$ is precisely the matrix of the map $\widetilde{q|_S} : S \to S^*$ with respect to that basis and its dual, since $\widetilde{q|_S}(v_i)(v_j) = \langle v_i,v_j\rangle$. Nondegeneracy of $q|_{S \times S}$ says that this map is an isomorphism, so by Lemma \ref{lem:nondegenerate-symmetric-bilinear} its matrix $G$ is invertible.
\end{proof}

\begin{lemma}[Nondegeneracy of a Tuple via its Leading Principal Minors]
\label{lem:nondegenerate-tuple-leading-minors}
\lean{LeeLib.Ch02.isNondegenerateTuple_iff_forall_det_gramMatrix_ne_zero}
\leanok
\uses{lem:nondegenerate-tuple-gram-matrix}
Let $(V,q)$ be a scalar product space and let $(v_1,\ldots,v_n)$ be an $n$-tuple in $V$. For $j \leq n$ write $G_j = \bigl(\langle v_i, v_l\rangle\bigr)_{i,l=1}^{j}$ for the Gram matrix of the prefix $(v_1,\ldots,v_j)$. Then $(v_1,\ldots,v_n)$ is a nondegenerate tuple if and only if $\det G_j \neq 0$ for every $j \leq n$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-tuple-gram-matrix}
By definition $(v_1,\ldots,v_n)$ is a nondegenerate tuple exactly when, for every $j \leq n$, the subspace $\operatorname{span}(v_1,\ldots,v_j)$ is nondegenerate of dimension $j$. For each fixed $j$ that span is the span of the $j$-tuple $(v_1,\ldots,v_j)$, whose Gram matrix is $G_j$, so Lemma \ref{lem:nondegenerate-tuple-gram-matrix} applied to that $j$-tuple says precisely that the condition at $j$ holds if and only if $G_j$ is invertible, that is, $\det G_j \neq 0$. Conjoining over $j \leq n$ gives the claim.
\end{proof}

\begin{lemma}[A Nondegenerate Tuple is Linearly Independent]
\label{lem:nondegenerate-tuple-independent}
\lean{LeeLib.Ch02.IsNondegenerateTuple.linearIndependent}
\leanok
Let $(V,q)$ be a scalar product space and let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$. Then $v_1,\ldots,v_n$ are linearly independent.
\end{lemma}

\begin{proof}
\leanok
By definition the last step of the flag, $\operatorname{span}(v_1,\ldots,v_n)$, has dimension $n$. A spanning set of $n$ vectors for an $n$-dimensional space is a basis, hence linearly independent.
\end{proof}

\begin{lemma}[A Nondegenerate Tuple of Full Length is a Basis]
\label{lem:nondegenerate-tuple-basis}
\lean{LeeLib.Ch02.IsNondegenerateTuple.basis}
\leanok
\uses{lem:nondegenerate-tuple-independent}
Let $(V,q)$ be an $n$-dimensional scalar product space and let $(v_1,\ldots,v_n)$ be a nondegenerate $n$-tuple in $V$. Then $(v_1,\ldots,v_n)$ is a basis of $V$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-tuple-independent}
By Lemma \ref{lem:nondegenerate-tuple-independent} the tuple is linearly independent, and it consists of $n = \dim V$ vectors; a linearly independent family of $\dim V$ vectors spans, hence is a basis.
\end{proof}

\begin{lemma}[An Orthonormal Tuple is Nondegenerate]
\label{lem:orthonormal-tuple-nondegenerate}
\lean{LeeLib.Ch02.IsOrthonormal.isNondegenerateTuple}
\leanok
\uses{lem:nondegenerate-tuple-leading-minors}
Let $(V,q)$ be a scalar product space and let $(b_1,\ldots,b_n)$ be an orthonormal $n$-tuple in $V$. Then $(b_1,\ldots,b_n)$ is a nondegenerate tuple.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-tuple-leading-minors}
Fix $j \leq n$. The prefix $(b_1,\ldots,b_j)$ is again orthonormal, orthonormality being a condition on pairs of members. Its Gram matrix $G_j$ therefore satisfies $(G_j)_{il} = \langle b_i, b_l \rangle = 0$ for $i \neq l$, so $G_j$ is diagonal, with diagonal entries $\langle b_i,b_i\rangle = \pm 1$. Hence $\det G_j = \prod_{i=1}^{j} \langle b_i, b_i \rangle$ is a product of numbers $\pm 1$ and so is nonzero. As $j \leq n$ was arbitrary, Lemma \ref{lem:nondegenerate-tuple-leading-minors} shows the tuple is nondegenerate.
\end{proof}

\begin{lemma}[Smoothness of the Leading Gram Determinants]
\label{lem:gram-det-smooth}
\lean{LeeLib.Ch02.contMDiffOn_det_gramMatrix}
\leanok
\uses{lem:pseudo-metric-pairing-smooth}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$, let $u \subseteq B$, and let $Y_1,\ldots,Y_k$ be sections of $V$ that are smooth on $u$. Then
\[
x \longmapsto \det \bigl(g_x(Y_i|_x, Y_l|_x)\bigr)_{i,l=1}^{k}
\]
is a smooth real-valued function on $u$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:pseudo-metric-pairing-smooth}
By Lemma \ref{lem:pseudo-metric-pairing-smooth} each entry $x \mapsto g_x(Y_i|_x, Y_l|_x)$ is smooth on $u$. The determinant of a $k \times k$ matrix is the finite sum, over the permutations $\sigma$ of $\{1,\ldots,k\}$, of $\operatorname{sgn}(\sigma)$ times the finite product of the entries $(\sigma(i), i)$; finite sums and finite products of smooth real-valued functions are smooth, so the composite is smooth on $u$.
\end{proof}

\begin{lemma}[Nondegeneracy of a Tuple of Smooth Sections is an Open Condition]
\label{lem:open-nondegenerate-tuple}
\lean{LeeLib.Ch02.exists_isLocalNondegenerateOn_nhds}
\leanok
\uses{def:local-nondegenerate-family,lem:nondegenerate-tuple-leading-minors,lem:gram-det-smooth,lem:pseudo-metric-fibre-scalar-product}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$ with finite-dimensional fibres, let $u \subseteq B$ be open, and let $X_1,\ldots,X_m$ be sections of $V$ that are smooth on $u$. If $(X_1|_{x_0},\ldots,X_m|_{x_0})$ is a nondegenerate $m$-tuple in $(V_{x_0}, g_{x_0})$ for some $x_0 \in u$, then there is an open set $w$ with $x_0 \in w \subseteq u$ such that $(X_1,\ldots,X_m)$ is nondegenerate on $w$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-tuple-leading-minors,lem:gram-det-smooth,lem:pseudo-metric-fibre-scalar-product}
For $k \leq m$ and $x \in u$ let $G_k(x) = \bigl(g_x(X_i|_x, X_j|_x)\bigr)_{i,j=1}^{k}$ be the $k$-th leading principal submatrix of the Gram matrix of the tuple at $x$. By Lemma \ref{lem:gram-det-smooth} the function $x \mapsto \det G_k(x)$ is smooth, in particular continuous, on $u$.

By Lemma \ref{lem:pseudo-metric-fibre-scalar-product} each $(V_x, g_x)$ is a scalar product space, so Lemma \ref{lem:nondegenerate-tuple-leading-minors} applies in each fibre. Nondegeneracy of the tuple $(X_1|_{x_0},\ldots,X_m|_{x_0})$ therefore says exactly that $\det G_k(x_0) \neq 0$ for every $k \leq m$.

For each $k \leq m$ continuity therefore yields an open set $w_k$ with $x_0 \in w_k \subseteq u$ on which $\det G_k$ does not vanish. Put $w = \bigcap_{k \leq m} w_k$, an intersection of finitely many open sets, so open, and containing $x_0$. For $x \in w$ every $\det G_k(x)$ is nonzero, so by Lemma \ref{lem:nondegenerate-tuple-leading-minors} again the tuple $(X_1|_x,\ldots,X_m|_x)$ is nondegenerate. Together with the smoothness of the $X_i$ on $w \subseteq u$, this is Definition \ref{def:local-nondegenerate-family}.
\end{proof}

\begin{lemma}[Smooth Sections Through a Prescribed Basis]
\label{lem:sections-through-prescribed-basis}
\lean{LeeLib.AppendixA.exists_contMDiffOn_section_eq_basis}
\leanok
Let $V \to B$ be a smooth vector bundle of rank $m$, let $x_0 \in B$, and let $(w_1,\ldots,w_m)$ be a basis of the fibre $V_{x_0}$. Then there are an open set $v \ni x_0$ and sections $Y_1,\ldots,Y_m$ of $V$, smooth on $v$, with $Y_j|_{x_0} = w_j$ for every $j$.
\end{lemma}

\begin{proof}
\leanok
Choose a local trivialization $\Phi$ of $V$ over an open set $v \ni x_0$, with fibrewise linear isomorphisms $\Phi_x : V_x \to F$, and set
\[
Y_j|_x = \Phi_x^{-1}\bigl(\Phi_{x_0}(w_j)\bigr), \qquad x \in v .
\]
Each $Y_j$ is smooth on $v$, being constant in the trivialization $\Phi$, and $Y_j|_{x_0} = \Phi_{x_0}^{-1}(\Phi_{x_0}(w_j)) = w_j$.
\end{proof}

\begin{proposition}[Orthonormalization of a Nondegenerate Family of Vector Fields]
\label{prop:orthonormalize-nondegenerate-family}
\lean{LeeLib.Ch02.exists_pseudo_orthonormalFrame}
\leanok
\uses{def:pseudo-riemannian-metric,def:local-nondegenerate-family,def:pseudo-gram-schmidt-frame,prop:pseudo-gram-schmidt-frame-smooth,prop:pseudo-gram-schmidt-frame-orthonormal,lem:pseudo-gram-schmidt-frame-flag}
Let $(M,g)$ be a pseudo-Riemannian manifold, let $u \subseteq M$, and let $(X_1,\ldots,X_m)$ be a family of vector fields that is nondegenerate on $u$. Then there is a family $(Y_1,\ldots,Y_m)$ of vector fields, smooth on $u$, such that for every $x \in u$ the tuple $(Y_1|_x,\ldots,Y_m|_x)$ is orthonormal in $(T_xM,g_x)$ and
\[
\operatorname{span}(Y_1|_x,\ldots,Y_k|_x) = \operatorname{span}(X_1|_x,\ldots,X_k|_x) \qquad \text{for every } k \leq m .
\]
\end{proposition}

\begin{proof}
\leanok
\uses{def:pseudo-gram-schmidt-frame,prop:pseudo-gram-schmidt-frame-smooth,prop:pseudo-gram-schmidt-frame-orthonormal,lem:pseudo-gram-schmidt-frame-flag}
Take $(Y_1,\ldots,Y_m)$ to be the fibrewise Gram--Schmidt frame of $(X_1,\ldots,X_m)$ of Definition \ref{def:pseudo-gram-schmidt-frame}. Proposition \ref{prop:pseudo-gram-schmidt-frame-smooth} gives its smoothness on $u$, Proposition \ref{prop:pseudo-gram-schmidt-frame-orthonormal} its orthonormality at each point of $u$, and Lemma \ref{lem:pseudo-gram-schmidt-frame-flag} the equality of spans.
\end{proof}

\begin{proposition}[Orthonormal Frames for Pseudo-Riemannian Manifolds]
\label{prop:orthonormal-frames-pseudo-riemannian}
\lean{LeeLib.Ch02.exists_pseudo_orthonormalFrame_nhds}
\leanok
\dcref{ch2:2.66}
\uses{def:pseudo-riemannian-metric,lem:completion-nondegenerate-bases,lem:sections-through-prescribed-basis,lem:open-nondegenerate-tuple,prop:orthonormalize-nondegenerate-family}
Let $(M,g)$ be a pseudo-Riemannian manifold. For each $p \in M$, there exists a smooth orthonormal frame on a neighborhood of $p$ in $M$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:completion-nondegenerate-bases,lem:sections-through-prescribed-basis,lem:open-nondegenerate-tuple,prop:orthonormalize-nondegenerate-family,lem:pseudo-metric-fibre-scalar-product}
Write $n = \dim M$ and fix $p \in M$. By Definition \ref{def:pseudo-riemannian-metric} and Lemma \ref{lem:pseudo-metric-fibre-scalar-product}, $(T_pM, g_p)$ is a scalar product space.

\emph{A nondegenerate basis at $p$.} The empty tuple is nondegenerate, so Lemma \ref{lem:completion-nondegenerate-bases}, applied with $k = 0$, produces a nondegenerate basis $(w_1,\ldots,w_n)$ of $T_pM$.

\emph{Extension to smooth vector fields.} By Lemma \ref{lem:sections-through-prescribed-basis} applied to the tangent bundle there are an open set $v \ni p$ and vector fields $Y_1,\ldots,Y_n$, smooth on $v$, with $Y_j|_p = w_j$.

\emph{Shrinking to a nondegenerate family.} The tuple $(Y_1|_p,\ldots,Y_n|_p) = (w_1,\ldots,w_n)$ is nondegenerate in $(T_pM,g_p)$, so Lemma \ref{lem:open-nondegenerate-tuple} yields an open set $v'$ with $p \in v' \subseteq v$ such that $(Y_1,\ldots,Y_n)$ is nondegenerate on $v'$ in the sense of Definition \ref{def:local-nondegenerate-family}.

\emph{Orthonormalization.} Proposition \ref{prop:orthonormalize-nondegenerate-family}, applied to $(Y_1,\ldots,Y_n)$ on $v'$, yields vector fields $(E_1,\ldots,E_n)$, smooth on $v'$, with $(E_1|_x,\ldots,E_n|_x)$ orthonormal in $(T_xM, g_x)$ for every $x \in v'$. An orthonormal tuple is linearly independent, and $n$ linearly independent vectors in the $n$-dimensional space $T_xM$ form a basis; so $(E_1,\ldots,E_n)$ is a smooth orthonormal frame on the neighborhood $v'$ of $p$.
\end{proof}

The argument splits into two halves of unequal cost. The orthonormalization half — Propositions \ref{prop:pseudo-gram-schmidt-frame-smooth} and \ref{prop:pseudo-gram-schmidt-frame-orthonormal}, resting on the recursion of Definition \ref{def:gram-schmidt-bilin} and its properties — smoothly orthonormalizes any family that is nondegenerate on a set, and is the substance of the proposition. The other half produces such a family near a given point: Lemma \ref{lem:completion-nondegenerate-bases} supplies a nondegenerate basis of the single tangent space $T_pM$, Lemma \ref{lem:sections-through-prescribed-basis} spreads it to smooth vector fields near $p$, and Lemma \ref{lem:open-nondegenerate-tuple} shrinks the neighborhood until nondegeneracy holds at every point of it, not only at $p$. That last step is where the Riemannian case is free and this one is not: there, linear independence is both what Gram--Schmidt needs and what an arbitrary local frame already supplies, whereas here nondegeneracy of the whole flag must be propagated off the point $p$ by an open condition on the leading principal minors of the Gram matrix.

\begin{exercise}
\label{exer:prove-orthonormal-frames}
\leanok
\dcref{ch2:2.67}
\uses{prop:orthonormal-frames-pseudo-riemannian}
Prove the preceding proposition.
\end{exercise}

\begin{proof}
\leanok
\uses{prop:orthonormal-frames-pseudo-riemannian}
This is the content of Proposition \ref{prop:orthonormal-frames-pseudo-riemannian}, proved above.
\end{proof}

\begin{exercise}
\label{exer:product-pseudo-riemannian}
\dcref{ch2:2.68}
\uses{def:pseudo-riemannian-metric,def:pseudo-metric-has-signature,def:product-metric}

Suppose $(M_1, g_1)$ and $(M_2, g_2)$ are pseudo-Riemannian manifolds of signatures $(r_1, s_1)$ and $(r_2, s_2)$, respectively. Show that $(M_1 \times M_2, g_1 \oplus g_2)$ is a pseudo-Riemannian manifold of signature $(r_1 + r_2, s_1 + s_2)$.
\end{exercise}

For nonnegative integers $r$ and $s$, we define the \textbf{pseudo-Euclidean space of signature $(r,s)$}, denoted by $\mathbb{R}^{r,s}$, to be the manifold $\mathbb{R}^{r+s}$, with standard coordinates denoted by $(\xi^1, \ldots, \xi^r, \tau^1, \ldots, \tau^s)$, and with the pseudo-Riemannian metric $\bar{g}^{(r,s)}$ defined by
\begin{equation}
\bar{g}^{(r,s)} = (d\xi^1)^2 + \cdots + (d\xi^r)^2 - (d\tau^1)^2 - \cdots - (d\tau^s)^2.
\tag{2.24}
\end{equation}

By far the most important pseudo-Riemannian metrics (other than the Riemannian ones) are the Lorentz metrics.

\begin{definition}[Lorentz Metric]
\label{def:lorentz-metric}
\lean{LeeLib.Ch02.IsLorentzMetric}
\leanok
\uses{def:pseudo-metric-has-signature}
A \emph{Lorentz metric} is a pseudo-Riemannian metric of index $1$; that is, one having signature $(r,1)$ for some $r \geq 0$. A \emph{Lorentz manifold} is a pseudo-Riemannian manifold whose metric is a Lorentz metric.
\end{definition}

By Proposition \ref{prop:pseudo-metric-signature-sum} the integer $r$ is determined by the manifold: a Lorentz metric on an $n$-manifold has signature $(n-1,1)$. (Some authors, especially in the physics literature, prefer to use signature $(1, r)$; either one can be converted to the other by multiplying the metric by $-1$, so there is no significant difference.)

The pseudo-Euclidean metric $\bar{g}^{(r,1)}$ is a Lorentz metric called the \textbf{Minkowski metric}, and the Lorentz manifold $\mathbb{R}^{r,1}$ is called $(r+1)$-\textbf{dimensional Minkowski space}. If we denote standard coordinates on $\mathbb{R}^{r,1}$ by $(\xi^1, \ldots, \xi^r, \tau)$, then the Minkowski metric is given by
\begin{equation}
\bar{g}^{(r,1)} = (d\xi^1)^2 + \cdots + (d\xi^r)^2 - (d\tau)^2.
\tag{2.25}
\end{equation}

In the special case of $\mathbb{R}^{3,1}$, the Minkowski metric is the fundamental invariant of Albert Einstein's \textit{special theory of relativity}, which can be expressed succinctly by saying that if gravity is ignored, then spacetime is accurately modeled by $(3+1)$-dimensional Minkowski space, and the laws of physics have the same form in every coordinate system in which the Minkowski metric has the expression (2.25). The differing physical characteristics of ``space'' (the $\xi^i$ directions) and ``time'' (the $\tau$ direction) arise from the fact that they are subspaces on which the metric is positive definite and negative definite, respectively. The \textit{general theory of relativity} includes gravitational effects by allowing the Lorentz metric to vary from point to point.

Many, but not all, results from the theory of Riemannian metrics apply equally well to pseudo-Riemannian metrics. Throughout this book, we will attempt to point out which results carry over directly to pseudo-Riemannian metrics, which ones can be adapted with minor modifications, and which ones do not carry over at all. As a rule of thumb, proofs that depend only on the nondegeneracy of the metric tensor, such as properties of the musical isomorphisms and existence and uniqueness of geodesics, work fine in the pseudo-Riemannian setting, while proofs that use positivity in an essential way, such as those involving lengths of curves, do not.

One notable result that does \emph{not} carry over to the pseudo-Riemannian case is Proposition~2.4, about the existence of metrics. For example, the following result characterizes those manifolds that admit Lorentz metrics.

The two halves of Theorem \ref{thm:lorentz-metrics} are of very different character, so they are recorded separately. Sufficiency is constructive: a rank-1 distribution and an auxiliary Riemannian metric produce a Lorentz metric by flipping the sign of the metric along the distribution. Necessity runs the other way: recovering a distribution from a Lorentz metric $\bar g$ means selecting, smoothly in $x$, the negative-eigenvalue eigenline of $\bar g_x$ relative to an auxiliary Riemannian metric.

That such a selection is smooth is usually quoted as a perturbation-theoretic fact about simple eigenvalues. It need not be. The eigen-equation $Sv = \lambda v$, together with a normalization $\langle v_0, v\rangle = 1$ pinning down the scale, is a system of $\dim V + 1$ equations in the $\dim V + 1$ unknowns $(v,\lambda)$, depending on the operator $S$ as a parameter; Definition \ref{def:augmented-eigen-equation} packages it as a single map $\Phi_{v_0}$ carrying $S$ along unchanged. Simplicity of the eigenvalue is exactly the invertibility of the derivative of that map (Lemma \ref{lem:eigen-linearisation-injective}), so the inverse function theorem alone yields the selection and its smoothness (Proposition \ref{prop:smooth-eigen-selection}). No spectral projection is constructed, and no operator square root, contour integral or functional calculus is needed.

The pointwise content of sufficiency is the following sign flip, which is Lee's own recipe in Problem \ref{prob:lorentz-metric-existence}(a) written invariantly: for a $g$-orthonormal frame with $E_1$ spanning $D$ and dual coframe $(\varepsilon^i)$, the form $-(\varepsilon^1)^2 + (\varepsilon^2)^2 + \cdots + (\varepsilon^n)^2$ equals $\langle v,w\rangle - 2\langle v,E_1\rangle\langle w,E_1\rangle$.

\begin{lemma}[Flipping the sign of a scalar product along a unit vector]
\label{lem:lorentz-form-pointwise}
\lean{LeeLib.Ch02.lorentzForm_isSymm_nondegenerate_signature}
\leanok
\uses{prop:sylvesters-law-inertia,cor:scalar-product-signature-sum}
Let $(V, \langle\cdot,\cdot\rangle)$ be an inner product space with $\dim V = n+1$, and let $u \in V$ be a unit vector. Then
\[
  B(v,w) = \langle v,w\rangle - 2\langle v,u\rangle\langle w,u\rangle
\]
is symmetric and nondegenerate, and has signature $(n,1)$. Moreover $B$ depends on $u$ only through $\operatorname{span}(u)$: replacing $u$ by $-u$ leaves $B$ unchanged.
\end{lemma}

\begin{proof}
\leanok
\uses{prop:sylvesters-law-inertia,cor:scalar-product-signature-sum}
Symmetry is immediate from symmetry of $\langle\cdot,\cdot\rangle$, and the invariance under $u \mapsto -u$ holds because $u$ occurs exactly twice in the formula, so the two sign changes cancel.

For nondegeneracy, write $R w = w - 2\langle w,u\rangle u$, so that $B(v,w) = \langle v, Rw\rangle$. Since $\langle u,u\rangle = 1$, we get $\langle Rw, u\rangle = \langle w,u\rangle - 2\langle w,u\rangle = -\langle w,u\rangle$, whence
\[
  R(Rw) = Rw - 2\langle Rw,u\rangle u = w - 2\langle w,u\rangle u + 2\langle w,u\rangle u = w,
\]
so $R$ is an involution. If $B(v,w) = 0$ for all $w$, then taking $w = Rv$ gives $0 = \langle v, R(Rv)\rangle = \langle v,v\rangle$, so $v = 0$.

For the signature, extend $u$ to an orthonormal basis $(e_0, \ldots, e_n)$ of $V$ with $e_0 = u$. Then $\langle e_i, u\rangle = \delta_{i0}$, so
\[
  B(e_i, e_j) = \delta_{ij} - 2\delta_{i0}\delta_{j0},
\]
which is $-1$ when $i = j = 0$, is $1$ when $i = j \neq 0$, and vanishes when $i \neq j$. So $(e_i)$ is an orthonormal basis for $B$ in the indefinite sense, with exactly one diagonal entry equal to $-1$. By Sylvester's law of inertia (Proposition \ref{prop:sylvesters-law-inertia}) the index of $B$ is therefore $1$, and since $B$ is nondegenerate its radical is trivial, so the number of positive terms is $\dim V - 1 = n$. Hence $B$ has signature $(n,1)$.
\end{proof}

\begin{definition}[Rank-1 tangent distribution]
\label{def:rank-one-distribution}
\lean{LeeLib.Ch02.IsRankOneDistribution}
\leanok
A \emph{rank-1 tangent distribution} on a smooth manifold $M$ is a family $D = (D_x)_{x \in M}$ of subspaces $D_x \subseteq T_xM$ such that every point of $M$ has a neighbourhood $U$ carrying a smooth vector field $X$ with $X_x \neq 0$ and $D_x = \operatorname{span}(X_x)$ for every $x \in U$. Each $D_x$ is then a line, so $D$ is a field of lines in $TM$.
\end{definition}

\begin{proposition}[A rank-1 distribution is a rank-1 subbundle]
\label{prop:rank-one-distribution-is-subbundle}
\lean{LeeLib.Ch02.hasLocalSubframes_of_isRankOneDistribution}
\leanok
\uses{def:rank-one-distribution,lem:local-frame-criterion-for-subbundles,prop:existence-riemannian-metric}
Let $D$ be a rank-1 tangent distribution on a smooth manifold $M$. Then $D$ satisfies the hypothesis of the subbundle criterion (Lemma \ref{lem:local-frame-criterion-for-subbundles}) with $k = 1$, and is therefore a smooth rank-1 subbundle of $TM$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:rank-one-distribution,lem:local-frame-criterion-for-subbundles}
The criterion asks that every point have a neighbourhood on which $D$ is spanned by $k$ smooth sections of $TM$ that are linearly independent at each point. Take $k=1$ and let $U$ and $X$ be as in Definition \ref{def:rank-one-distribution}. A family consisting of a single vector is linearly independent exactly when that vector is nonzero, which holds on $U$; and the span of the range of the one-element family $(X)$ is $\operatorname{span}(X_x) = D_x$ for $x \in U$. So the hypothesis holds, and the criterion makes $D$ a smooth rank-1 vector bundle over $M$.
\end{proof}

\begin{definition}[Tensor product of two $1$-forms]
\label{def:tensor-product-one-forms}
\lean{Bundle.formProduct}
\leanok
For covectors $\alpha, \beta \in V^*$, the bilinear form $\alpha \otimes \beta$ on $V$ is $(\alpha \otimes \beta)(v,w) = \alpha(v)\,\beta(w)$. For $1$-forms $\alpha, \beta$ on $M$ this is applied fibrewise, giving a field of bilinear forms $x \mapsto \alpha_x \otimes \beta_x$.
\end{definition}

\begin{lemma}[The tensor product of two smooth $1$-forms is smooth]
\label{lem:tensor-product-one-forms-smooth}
\lean{Bundle.contMDiffAt_formProduct}
\leanok
\uses{def:tensor-product-one-forms}
Let $E \to M$ be a smooth vector bundle and let $\alpha, \beta$ be smooth sections of its dual bundle. Then $x \mapsto \alpha_x \otimes \beta_x$ is a smooth section of the bundle of bilinear forms on $E$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:tensor-product-one-forms}
Smoothness is local, so it may be checked in a local trivialization $\Lambda$ of $E$ over a neighbourhood of a given point. In that trivialization a bilinear form $\varphi_x$ on $E_x$ is represented by $(\xi,\eta) \mapsto \varphi_x(\Lambda_x^{-1}\xi, \Lambda_x^{-1}\eta)$ and a covector $\alpha_x$ by $\hat\alpha_x \colon \xi \mapsto \alpha_x(\Lambda_x^{-1}\xi)$. Since
\[
  (\alpha_x \otimes \beta_x)(\Lambda_x^{-1}\xi, \Lambda_x^{-1}\eta) = \alpha_x(\Lambda_x^{-1}\xi)\,\beta_x(\Lambda_x^{-1}\eta) = (\hat\alpha_x \otimes \hat\beta_x)(\xi,\eta),
\]
the representation of $\alpha \otimes \beta$ is the tensor product $\hat\alpha \otimes \hat\beta$ of the representations of $\alpha$ and $\beta$. By hypothesis $\hat\alpha$ and $\hat\beta$ are smooth maps into the fixed model dual space, and the tensor product $(a,b) \mapsto a \otimes b$ is a bounded bilinear map between fixed normed spaces, hence smooth; so $\hat\alpha \otimes \hat\beta$ is smooth, and therefore so is $\alpha \otimes \beta$.
\end{proof}

\begin{definition}[The metric with its sign flipped along a vector field]
\label{def:sign-flipped-form}
\lean{Bundle.flipFormOf}
\leanok
\uses{def:tensor-product-one-forms}
Let $b$ be a field of bilinear forms on a vector bundle $E \to M$ and $u$ a section of $E$. Write $b(u,\cdot)$ for the $1$-form $v \mapsto b(u,v)$. The \emph{sign flip of $b$ along $u$} is the field of bilinear forms
\[
  \flip(b,u) = b - 2\, b(u,\cdot) \otimes b(u,\cdot), \qquad \flip(b,u)_x(v,w) = b_x(v,w) - 2\,b_x(u_x,v)\,b_x(u_x,w).
\]
\end{definition}

\begin{lemma}[The sign flip of a smooth form along a smooth section is smooth]
\label{lem:sign-flipped-form-smooth}
\lean{Bundle.contMDiffAt_flipFormOf}
\leanok
\uses{def:sign-flipped-form,lem:tensor-product-one-forms-smooth}
If $b$ is a smooth field of bilinear forms on $E \to M$ and $u$ is a smooth section of $E$, then $\flip(b,u)$ is a smooth field of bilinear forms.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:tensor-product-one-forms-smooth}
The $1$-form $b(u,\cdot)$ is smooth, being a smooth field of bilinear forms applied to a smooth section. Its tensor square is smooth by Lemma \ref{lem:tensor-product-one-forms-smooth}, and the smooth sections of a vector bundle are closed under subtraction and multiplication by constants, so $b - 2\,b(u,\cdot)\otimes b(u,\cdot)$ is smooth.
\end{proof}

\begin{lemma}[A line contains exactly two unit vectors]
\label{lem:two-unit-vectors-of-a-line}
\lean{LeeLib.Ch02.eq_or_eq_neg_of_span_eq}
\leanok
Let $(V, \langle\cdot,\cdot\rangle)$ be an inner product space and let $u, u' \in V$ satisfy $\langle u,u\rangle = \langle u',u'\rangle = 1$ and $\operatorname{span}(u') = \operatorname{span}(u)$. Then $u' = u$ or $u' = -u$.
\end{lemma}

\begin{proof}
\leanok
Since $u' \in \operatorname{span}(u)$ we have $u' = cu$ for some $c \in \mathbb{R}$. Then $1 = \langle u',u'\rangle = c^2\langle u,u\rangle = c^2$, so $c = \pm 1$.
\end{proof}

\begin{lemma}[Smooth local unit sections of a rank-1 distribution]
\label{lem:local-unit-section}
\lean{LeeLib.Ch02.IsRankOneDistribution.exists_localUnitSection}
\leanok
\uses{def:rank-one-distribution,def:riemannian-metric}
Let $(M,g)$ be a Riemannian manifold and $D$ a rank-1 tangent distribution on $M$. Every $p \in M$ has a neighbourhood $U$ carrying a smooth vector field $E$ with $\langle E_x, E_x\rangle_g = 1$ and $\operatorname{span}(E_x) = D_x$ for every $x \in U$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:rank-one-distribution}
Let $U$ and $X$ be as in Definition \ref{def:rank-one-distribution}, so $X$ is smooth and nowhere zero on $U$ and spans $D$ there. The function $x \mapsto \langle X_x, X_x\rangle_g$ is smooth on $U$, and strictly positive because $g$ is positive definite and $X$ is nowhere zero; hence $x \mapsto \langle X_x, X_x\rangle_g^{-1/2}$ is smooth on $U$. Therefore $E = X/|X|_g$ is smooth on $U$, and $\langle E,E\rangle_g = \langle X,X\rangle_g / |X|_g^2 = 1$. Finally $E_x$ is a nonzero scalar multiple of $X_x$, so $\operatorname{span}(E_x) = \operatorname{span}(X_x) = D_x$.
\end{proof}

\begin{proposition}[Sufficiency in Theorem \ref{thm:lorentz-metrics}]
\label{prop:lorentz-metric-of-rank-one-distribution}
\lean{LeeLib.Ch02.exists_isLorentzMetric_of_isRankOneDistribution}
\leanok
\uses{def:lorentz-metric,def:rank-one-distribution,prop:existence-riemannian-metric,lem:lorentz-form-pointwise}
If a smooth manifold $M$ admits a rank-1 tangent distribution $D \subseteq TM$, then $M$ admits a Lorentz metric.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:lorentz-form-pointwise,prop:existence-riemannian-metric,def:sign-flipped-form,lem:sign-flipped-form-smooth,lem:two-unit-vectors-of-a-line,lem:local-unit-section}
Choose a Riemannian metric $g$ on $M$ (Proposition \ref{prop:existence-riemannian-metric}). For each $x \in M$ the fibre $D_x$ is a line in $T_xM$, so it contains exactly two $g$-unit vectors, $\pm u_x$ (Lemma \ref{lem:two-unit-vectors-of-a-line}); choose one of them for each $x$, with no continuity required of the choice. Define
\[
  \bar{g}_x(v,w) = g_x(v,w) - 2\,g_x(v,u_x)\,g_x(w,u_x),
\]
that is, $\bar g = \flip(g,u)$ in the notation of Definition \ref{def:sign-flipped-form}.

By Lemma \ref{lem:lorentz-form-pointwise} the value of $\bar g_x$ does not depend on which of the two unit vectors is called $u_x$, so $\bar{g}$ is independent of the choice; and each $\bar{g}_x$ is symmetric, nondegenerate, and of signature $(n,1)$, where $n+1 = \dim M$. Note $\dim M \geq 1$, since $T_xM$ contains the line $D_x$. In particular $\bar{g}$ has index $1$ at every point.

It remains to check that $\bar{g}$ is smooth, which is a local question. Near any $x_0$, Lemma \ref{lem:local-unit-section} provides a smooth $g$-unit section $E$ of $D$ on a neighbourhood $U$ of $x_0$. For every $x \in U$ both $E_x$ and $u_x$ are $g$-unit vectors spanning the line $D_x$, so $E_x = \pm u_x$ by Lemma \ref{lem:two-unit-vectors-of-a-line}, and hence $\bar g_x = \flip(g,E)_x$ by the sign-independence just established. Thus $\bar g$ agrees on $U$ with $\flip(g,E)$, which is smooth by Lemma \ref{lem:sign-flipped-form-smooth}; so $\bar g$ is smooth at $x_0$. Hence $\bar{g}$ is a Lorentz metric.
\end{proof}

\begin{definition}[Simple eigenpair]
\label{def:simple-eigenpair}
\lean{LeeLib.Ch02.IsSimpleEigenpair}
\leanok
Let $(V,\langle\cdot,\cdot\rangle)$ be a finite-dimensional real inner product space, $S : V \to V$ linear, $v \in V$ and $\lambda \in \mathbb{R}$. The triple $(S,v,\lambda)$ is a \emph{simple eigenpair} if $S$ is symmetric, $v \neq 0$, $Sv = \lambda v$, and every $w \in V$ with $Sw = \lambda w$ lies in $\operatorname{span}(v)$; that is, $\ker(S - \lambda\,\mathrm{id}) = \operatorname{span}(v)$.
\end{definition}

\begin{definition}[Linearisation of the eigen-equation]
\label{def:eigen-linearisation}
\lean{LeeLib.Ch02.eigenLin}
\leanok
For $S : V \to V$ linear, $v, v_0 \in V$ and $\lambda \in \mathbb{R}$, let $L_{S,v,\lambda,v_0} : V \times \mathbb{R} \to V \times \mathbb{R}$ be the linear map
\[
  L_{S,v,\lambda,v_0}(w,\mu) = \bigl(Sw - \lambda w - \mu v,\ \langle v_0, w\rangle\bigr).
\]
\end{definition}

\begin{lemma}[Simplicity makes the linearisation injective]
\label{lem:eigen-linearisation-injective}
\lean{LeeLib.Ch02.eigenLin_injective}
\leanok
\uses{def:simple-eigenpair,def:eigen-linearisation}
Let $(S,v,\lambda)$ be a simple eigenpair and let $v_0 \in V$ satisfy $\langle v_0, v\rangle = 1$. Then $L_{S,v,\lambda,v_0}$ is injective, hence bijective.
\end{lemma}

\begin{proof}
\leanok
\uses{def:simple-eigenpair,def:eigen-linearisation}
Suppose $L_{S,v,\lambda,v_0}(w,\mu) = 0$, that is $Sw - \lambda w = \mu v$ and $\langle v_0, w\rangle = 0$. Pair the first equation with $v$: since $S$ is symmetric and $Sv = \lambda v$,
\[
  \langle Sw - \lambda w, v\rangle = \langle w, Sv\rangle - \lambda\langle w, v\rangle = \lambda\langle w,v\rangle - \lambda\langle w,v\rangle = 0 ,
\]
while the right-hand side gives $\langle \mu v, v\rangle = \mu\langle v,v\rangle$. As $v \neq 0$ we have $\langle v,v\rangle \neq 0$, so $\mu = 0$. Then $Sw = \lambda w$, so simplicity puts $w = cv$ for some $c \in \mathbb{R}$, and $0 = \langle v_0, w\rangle = c\langle v_0, v\rangle = c$. Hence $w = 0$ and $(w,\mu) = 0$. Injectivity of a linear self-map of the finite-dimensional space $V \times \mathbb{R}$ gives bijectivity.
\end{proof}

\begin{definition}[Augmented eigen-equation]
\label{def:augmented-eigen-equation}
\lean{LeeLib.Ch02.eigenAug}
\leanok
For $v_0 \in V$ let $\Phi_{v_0} : \operatorname{End}(V) \times V \times \mathbb{R} \to \operatorname{End}(V) \times V \times \mathbb{R}$ be
\[
  \Phi_{v_0}(S, v, \lambda) = \bigl(S,\ Sv - \lambda v,\ \langle v_0, v\rangle - 1\bigr).
\]
Thus $\Phi_{v_0}(S,v,\lambda) = (S,0,0)$ says exactly that $v$ is an eigenvector of $S$ with eigenvalue $\lambda$, normalized by $\langle v_0, v\rangle = 1$.
\end{definition}

\begin{lemma}[The augmented equation has an invertible derivative at a simple eigenpair]
\label{lem:augmented-eigen-equation-derivative}
\lean{LeeLib.Ch02.hasFDerivAt_eigenAug}
\leanok
\uses{def:augmented-eigen-equation,def:eigen-linearisation,def:simple-eigenpair,lem:eigen-linearisation-injective}
$\Phi_{v_0}$ is smooth, and its derivative at $(S,v,\lambda)$ is
\[
  D\Phi_{v_0}(S,v,\lambda)\,(T,w,\mu) = \bigl(T,\ Tv + Sw - \lambda w - \mu v,\ \langle v_0, w\rangle\bigr) .
\]
If $(S,v,\lambda)$ is a simple eigenpair and $\langle v_0, v\rangle = 1$, this derivative is a linear isomorphism.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:eigen-linearisation-injective}
Each component of $\Phi_{v_0}$ is a polynomial expression in $(S,v,\lambda)$ built from the bilinear evaluation map $(S,v) \mapsto Sv$, the scalar multiplication $(\lambda,v)\mapsto \lambda v$ and the linear functional $v \mapsto \langle v_0,v\rangle$, so $\Phi_{v_0}$ is smooth and the displayed formula is its derivative.

For invertibility, note the derivative is block lower triangular with respect to the splitting $\operatorname{End}(V) \oplus (V \times \mathbb{R})$: it is the identity on the first block, it is $L_{S,v,\lambda,v_0}$ on the second, and the remaining block is the map $T \mapsto (Tv, 0)$. A block triangular map with invertible diagonal blocks is invertible, and $L_{S,v,\lambda,v_0}$ is bijective by Lemma \ref{lem:eigen-linearisation-injective}.
\end{proof}

\begin{proposition}[Smooth selection of a simple eigenpair]
\label{prop:smooth-eigen-selection}
\lean{LeeLib.Ch02.exists_eigenSelection}
\leanok
\uses{def:simple-eigenpair,def:augmented-eigen-equation,lem:augmented-eigen-equation-derivative}
Let $(S_0, v_0, \lambda_0)$ be a simple eigenpair with $\langle v_0, v_0\rangle = 1$. Then there are an open set $W \subseteq \operatorname{End}(V)$ containing $S_0$ and maps $\mathcal{V} : W \to V$, $\Lambda : W \to \mathbb{R}$ such that
\begin{enumerate}
  \item $\mathcal{V}(S_0) = v_0$ and $\Lambda(S_0) = \lambda_0$;
  \item $S\,\mathcal{V}(S) = \Lambda(S)\,\mathcal{V}(S)$ and $\langle v_0, \mathcal{V}(S)\rangle = 1$ for every $S \in W$;
  \item $\mathcal{V}$ and $\Lambda$ are continuous on $W$;
  \item $\mathcal{V}$ and $\Lambda$ are smooth at every $S \in W$ for which $(S, \mathcal{V}(S), \Lambda(S))$ is again a simple eigenpair.
\end{enumerate}
\end{proposition}

\begin{proof}
\leanok
\uses{lem:augmented-eigen-equation-derivative,def:augmented-eigen-equation}
Since $S_0 v_0 = \lambda_0 v_0$ and $\langle v_0,v_0\rangle = 1$, we have $\Phi_{v_0}(S_0,v_0,\lambda_0) = (S_0,0,0)$. By Lemma \ref{lem:augmented-eigen-equation-derivative}, $\Phi_{v_0}$ is smooth with invertible derivative at $(S_0,v_0,\lambda_0)$, so by the inverse function theorem it restricts to a diffeomorphism $\Phi$ from an open neighbourhood $\mathcal{U}$ of $(S_0,v_0,\lambda_0)$ onto an open neighbourhood $\mathcal{W}$ of $(S_0,0,0)$.

Let $W = \{S : (S,0,0) \in \mathcal{W}\}$, an open set containing $S_0$ because $S \mapsto (S,0,0)$ is continuous, and define $(\mathcal{V}(S), \Lambda(S))$ to be the last two components of $\Phi^{-1}(S,0,0)$.

For (2), apply $\Phi_{v_0}$ to $\Phi^{-1}(S,0,0)$: the result is $(S,0,0)$, and reading off the three components of the definition of $\Phi_{v_0}$ gives that the first component of $\Phi^{-1}(S,0,0)$ is $S$ itself, that $S\,\mathcal{V}(S) - \Lambda(S)\,\mathcal{V}(S) = 0$, and that $\langle v_0, \mathcal{V}(S)\rangle - 1 = 0$. For (1), $\Phi^{-1}(S_0,0,0) = (S_0,v_0,\lambda_0)$ because $\Phi(S_0,v_0,\lambda_0) = (S_0,0,0)$ and $\Phi$ is injective on $\mathcal{U}$. Claim (3) is continuity of $\Phi^{-1}$ composed with the continuous map $S \mapsto (S,0,0)$.

For (4), fix $S \in W$ such that $(S,\mathcal{V}(S),\Lambda(S))$ is a simple eigenpair. By (2) we have $\langle v_0, \mathcal{V}(S)\rangle = 1$, so Lemma \ref{lem:augmented-eigen-equation-derivative} applies at the point $\Phi^{-1}(S,0,0) = (S, \mathcal{V}(S), \Lambda(S))$ and shows the derivative of $\Phi_{v_0}$ there is invertible. Since $\Phi_{v_0}$ is smooth, $\Phi^{-1}$ is therefore smooth at $(S,0,0)$; composing with $S \mapsto (S,0,0)$ and projecting gives smoothness of $\mathcal{V}$ and $\Lambda$ at $S$.
\end{proof}

\begin{lemma}[The index counts the negative eigenvalues]
\label{lem:signature-and-eigenvalue-signs}
\lean{LeeLib.Ch02.sigNeg_eq_ncard_neg_eigenvalues}
\leanok
\uses{cor:scalar-product-signature-form,prop:sylvesters-law-inertia}
Let $(V,\langle\cdot,\cdot\rangle)$ be a finite-dimensional real inner product space and let $B$ be a symmetric nondegenerate bilinear form on $V$, so that $(V,B)$ is a scalar product space of some signature $(r,s)$. Let $A : V \to V$ be a linear operator with $B(v,w) = \langle Av, w\rangle$ for all $v,w$; such an $A$ exists and is unique because $\langle\cdot,\cdot\rangle$ is nondegenerate, and it is symmetric. Then the index $s$ is the number of negative eigenvalues of $A$, counted with multiplicity.
\end{lemma}

\begin{proof}
\leanok
\uses{cor:scalar-product-signature-form,prop:sylvesters-law-inertia}
Symmetry of $A$ is immediate: $\langle Av,w\rangle = B(v,w) = B(w,v) = \langle Aw,v\rangle = \langle v, Aw\rangle$. By the spectral theorem for symmetric operators there is a $\langle\cdot,\cdot\rangle$-orthonormal basis $u_1,\dots,u_n$ of $V$ with $Au_i = \mu_i u_i$, the eigenvalues listed with multiplicity. Then
\[
  B(u_i,u_j) = \langle Au_i, u_j\rangle = \mu_i\langle u_i,u_j\rangle = \mu_i\delta_{ij},
\]
so $B$ is diagonal in this basis with the eigenvalues on the diagonal. Each $\mu_i \neq 0$: otherwise $B(u_i, w) = \langle \mu_i u_i, w\rangle = 0$ for every $w$, putting $u_i \neq 0$ in the radical of $B$ and contradicting nondegeneracy.

Set $e_i = |\mu_i|^{-1/2}u_i$, which is again a basis since the scalars are nonzero. Then $B(e_i,e_j) = |\mu_i|^{-1/2}|\mu_j|^{-1/2}\mu_i\delta_{ij}$, which vanishes for $i \neq j$ and equals $\mu_i/|\mu_i| = \pm 1$ for $i = j$. So $(e_i)$ is an orthonormal basis for $B$ in the indefinite sense of Corollary \ref{cor:scalar-product-signature-form}, and $B(e_i,e_i) = -1$ exactly at the indices with $\mu_i < 0$. By Sylvester's law of inertia (Proposition \ref{prop:sylvesters-law-inertia}) the number of $-1$'s is the index $s$, so $s$ is the number of indices with $\mu_i < 0$, which is the number of negative eigenvalues counted with multiplicity.
\end{proof}

\begin{lemma}[Index one yields a simple eigenpair]
\label{lem:simple-eigenpair-of-index-one}
\lean{LeeLib.Ch02.exists_isSimpleEigenpair_of_sigNeg_eq_one}
\leanok
\uses{lem:signature-and-eigenvalue-signs,def:simple-eigenpair}
In the situation of Lemma \ref{lem:signature-and-eigenvalue-signs}, suppose the index of $B$ is $1$. Then $A$ has exactly one negative eigenvalue $\lambda$; the eigenspace $\ker(A - \lambda\,\mathrm{id})$ is a line, spanned by an eigenvector $u$; and $(A, u, \lambda)$ is a simple eigenpair.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:signature-and-eigenvalue-signs,def:simple-eigenpair}
Keep the eigenbasis $(u_i)$ and eigenvalues $(\mu_i)$ of the previous proof. By Lemma \ref{lem:signature-and-eigenvalue-signs} exactly one index $i_0$ has $\mu_{i_0} < 0$; write $\lambda = \mu_{i_0}$ and $u = u_{i_0}$, so $u \neq 0$ and $Au = \lambda u$.

For a symmetric operator the dimension of the $\lambda$-eigenspace equals the number of indices $i$ with $\mu_i = \lambda$. If $\mu_i = \lambda$ then $\mu_i < 0$, so $i = i_0$ by uniqueness of that index; hence there is exactly one such index and $\dim\ker(A - \lambda\,\mathrm{id}) = 1$. Now $\operatorname{span}(u) \subseteq \ker(A-\lambda\,\mathrm{id})$ because $Au = \lambda u$, and $\dim\operatorname{span}(u) = 1$ because $u \neq 0$; two subspaces of the same finite dimension with one contained in the other are equal, so $\operatorname{span}(u) = \ker(A - \lambda\,\mathrm{id})$. Thus every $w$ with $Aw = \lambda w$ lies in $\operatorname{span}(u)$, which together with symmetry of $A$ and $u \neq 0$ says exactly that $(A,u,\lambda)$ is a simple eigenpair.
\end{proof}

\begin{proposition}[Necessity in Theorem \ref{thm:lorentz-metrics}]
\label{prop:rank-one-distribution-of-lorentz-metric}
\lean{LeeLib.Ch02.exists_isRankOneDistribution_of_isLorentzMetric}
\leanok
\uses{def:lorentz-metric,def:rank-one-distribution,prop:existence-riemannian-metric,lem:simple-eigenpair-of-index-one,prop:smooth-eigen-selection,prop:existence-orthonormal-frames}
If a smooth manifold $M$ admits a Lorentz metric, then $M$ admits a rank-1 tangent distribution.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:existence-riemannian-metric,lem:simple-eigenpair-of-index-one,prop:smooth-eigen-selection,prop:existence-orthonormal-frames,def:rank-one-distribution}
Let $\bar{g}$ be a Lorentz metric and choose an auxiliary Riemannian metric $g$, which exists by Proposition \ref{prop:existence-riemannian-metric}. For each $x$ let $A_x : T_xM \to T_xM$ be the $g$-self-adjoint endomorphism determined by $\bar{g}_x(v,w) = g_x(A_xv, w)$. Define
\[
  D_x = \bigsqcup_{\mu < 0} \ker\bigl(A_x - \mu\,\mathrm{id}\bigr),
\]
the sum of all the negative eigenspaces of $A_x$. This mentions no eigenvalue of $A_x$, so $D$ is a well-defined field of subspaces depending on nothing but $g$.

We claim $D_x$ is a line, spanned by any nonzero $v$ with $A_xv = \mu v$ and $\mu < 0$. Since $\bar{g}$ is Lorentz, its index at every $x$ is $1$, so by Lemma \ref{lem:simple-eigenpair-of-index-one} the operator $A_x$ has exactly one negative eigenvalue $\lambda(x)$, and its eigenspace is the line spanned by the corresponding spectral eigenvector. Any negative eigenvalue of $A_x$ occurs in the spectrum of $A_x$, hence equals $\lambda(x)$; so every summand $\ker(A_x - \mu)$ with $\mu < 0$ is either $0$ or that same line, and $D_x$ is exactly it. In particular $(A_x, v, \lambda(x))$ is a simple eigenpair for every nonzero $v \in D_x$. It remains to produce smooth local spanning sections, which is a local question.

Fix $x_0 \in M$. By Proposition \ref{prop:existence-orthonormal-frames} there is a smooth $g$-orthonormal frame $E_1,\dots,E_n$ on a neighbourhood $U$ of $x_0$. Let $\iota_x : \mathbb{R}^n \to T_xM$, $\iota_x(c) = \sum_i c_i E_i(x)$, be its coefficient map; it is linear and bijective because a frame is a pointwise basis. Let $S(x) \in \operatorname{End}(\mathbb{R}^n)$ be the operator with matrix $S(x)_{ij} = \bar{g}_x(E_i, E_j)$. These entries are $\bar{g}$-pairings of smooth sections, hence smooth functions of $x$ on $U$, so $x \mapsto S(x)$ is a smooth map from $U$ into $\operatorname{End}(\mathbb{R}^n)$, and each $S(x)$ is symmetric because $\bar g$ is. Because the frame is $g$-orthonormal, $S(x)$ is conjugate to $A_x$:
\[
  A_x \circ \iota_x = \iota_x \circ S(x),
\]
since both sides have the same $g$-inner products against every $E_j(x)$, namely $\sum_i c_i \bar{g}_x(E_i,E_j)$. Consequently $\iota_x$ matches the $\mu$-eigenvectors of $S(x)$ with those of $A_x$; combined with the claim above, $(S(x), c, \mu)$ is a simple eigenpair whenever $c \neq 0$, $S(x)c = \mu c$ and $\mu < 0$. Note only the injectivity of $\iota_x$ is used here, not that it is an isometry.

Let $v_0$ be a unit vector spanning the $\lambda(x_0)$-eigenspace of $S(x_0)$, obtained by pulling a negative eigenvector of $A_{x_0}$ back through $\iota_{x_0}$ and normalizing. Proposition \ref{prop:smooth-eigen-selection} applied to the simple eigenpair $(S(x_0), v_0, \lambda(x_0))$ gives an open $W \ni S(x_0)$ and maps $\mathcal{V}, \Lambda$ on $W$ selecting an eigenpair of each $S \in W$. Since $\Lambda$ is continuous and $\Lambda(S(x_0)) = \lambda(x_0) < 0$, the set
\[
  U' = \{x \in U : S(x) \in W \text{ and } \Lambda(S(x)) < 0\}
\]
is open and contains $x_0$. Set $X_x = \iota_x\,\mathcal{V}(S(x))$.

For $x \in U'$ the vector $\mathcal{V}(S(x))$ is an eigenvector of $S(x)$ with eigenvalue $\Lambda(S(x)) < 0$, and it is nonzero because $\langle v_0, \mathcal{V}(S(x))\rangle = 1$; so $X_x \neq 0$ by injectivity of $\iota_x$, and $X_x$ is an eigenvector of $A_x$ with the negative eigenvalue $\Lambda(S(x))$ by the conjugation relation. Hence $\operatorname{span}(X_x) = D_x$ by the claim.

Smoothness of $X$ on $U'$ is smoothness of $x \mapsto \mathcal{V}(S(x))$ together with smoothness of the frame, since $X_x = \sum_i \mathcal{V}(S(x))_i\,E_i(x)$. By the previous paragraph $(S(x), \mathcal{V}(S(x)), \Lambda(S(x)))$ is a simple eigenpair for each $x \in U'$. So clause (4) of Proposition \ref{prop:smooth-eigen-selection} applies at every $S(x)$, making $\mathcal{V}$ smooth there; composing with the smooth map $x \mapsto S(x)$ shows the coefficients of $X$ are smooth on $U'$, hence so is $X$.

Thus every $x_0 \in M$ has a neighbourhood carrying a smooth nowhere-zero section of $D$, so $D$ is a rank-1 tangent distribution in the sense of Definition \ref{def:rank-one-distribution}.
\end{proof}

\begin{theorem}
\label{thm:lorentz-metrics}
\dcref{ch2:2.69}
\lean{LeeLib.Ch02.exists_isLorentzMetric_iff_exists_isRankOneDistribution}
\leanok
\uses{def:lorentz-metric,prop:existence-riemannian-metric,prop:lorentz-metric-of-rank-one-distribution,prop:rank-one-distribution-of-lorentz-metric}
A smooth manifold $M$ admits a Lorentz metric if and only if it admits a rank-1 tangent distribution (i.e., a rank-1 subbundle of $TM$).
\end{theorem}

\begin{proof}
\leanok
\uses{prop:lorentz-metric-of-rank-one-distribution,prop:rank-one-distribution-of-lorentz-metric}
Sufficiency is Proposition \ref{prop:lorentz-metric-of-rank-one-distribution} and necessity is Proposition \ref{prop:rank-one-distribution-of-lorentz-metric}.
\end{proof}

\emph{Proof.} Problem~2-34. \qed

With some more sophisticated tools from algebraic topology, it can be shown that every noncompact connected smooth manifold admits a Lorentz metric, and a compact connected smooth manifold admits a Lorentz metric if and only if its Euler characteristic is zero (see \cite{ON83}, p.~149). It follows that no even-dimensional sphere admits a Lorentz metric, because $S^{2n}$ has Euler characteristic equal to 2. For a thorough treatment of pseudo-Riemannian metrics from a mathematical point of view, see the excellent book \cite{ON83}; a more physical treatment can be found in \cite{HE73}.

\section{Pseudo-Riemannian Submanifolds}

The theory of submanifolds is only slightly more complicated in the pseudo-Riemannian case. If $(\tilde{M}, \tilde{g})$ is a pseudo-Riemannian manifold, a smooth submanifold $\iota: M \hookrightarrow \tilde{M}$ is called a \emph{pseudo-Riemannian submanifold} of $(\tilde{M}, \tilde{g})$ if $\iota^* \tilde{g}$ is nondegenerate with constant signature. If this is the case, we always consider $M$ to be endowed with the induced pseudo-Riemannian metric $\iota^* \tilde{g}$. In the special case in which $\iota^* \tilde{g}$ is positive definite, $M$ is called a \emph{Riemannian submanifold}.

The nondegeneracy hypothesis is not automatically satisfied; for example, if $M \subseteq \mathbb{R}^{1,1}$ is the submanifold $\{(\xi, \tau) : \xi = \tau\}$ and $\iota: M \to \mathbb{R}^{1,1}$ is inclusion, then the pullback tensor $\iota^* \tilde{g}^{(1,1)}$ is identically zero on $M$.

For hypersurfaces (submanifolds of codimension 1), the nondegeneracy condition is easy to check. If $M \subseteq \tilde{M}$ is a smooth submanifold and $p \in M$, then a vector $v \in T_p \tilde{M}$ is said to be \emph{normal} to $M$ if $(v,x) = 0$ for all $x \in T_p M$, just as in the Riemannian case. The space of all normal vectors at $p$ is a subspace of $T_p \tilde{M}$ denoted by $N_p M$.

The proofs of Propositions \ref{prop:pseudo-riemannian-hypersurface-signature} and \ref{prop:pseudo-riemannian-adapted-frame} both begin with the same purely linear-algebraic step, carried out pointwise on a tangent space: a nondegenerate tuple is absorbed into an orthonormal basis inducing the same flag. We isolate it once.

\begin{lemma}[Orthonormal Bases Adapted to a Nondegenerate Flag]
\label{lem:adapted-orthonormal-basis}
\lean{LeeLib.Ch02.exists_isOrthonormal_prefixSpan_eq_of_isNondegenerateTuple}
\leanok
\uses{lem:completion-nondegenerate-bases,prop:gram-schmidt-algorithm-scalar-products}
Suppose $(V,q)$ is an $n$-dimensional scalar product space and $(v_1,\ldots,v_k)$ is a nondegenerate $k$-tuple in $V$, with $k \leq n$. Then there is an orthonormal basis $(b_1,\ldots,b_n)$ for $V$ such that $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(v_1,\ldots,v_j)$ for every $j \leq k$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:completion-nondegenerate-bases,prop:gram-schmidt-algorithm-scalar-products}
By Lemma \ref{lem:completion-nondegenerate-bases} the tuple extends to a nondegenerate basis $(w_1,\ldots,w_n)$ of $V$ with $w_i = v_i$ for every $i \leq k$. Proposition \ref{prop:gram-schmidt-algorithm-scalar-products} applied to $(w_i)$ yields an orthonormal basis $(b_i)$ with $\operatorname{span}(b_1,\ldots,b_j) = \operatorname{span}(w_1,\ldots,w_j)$ for every $j \leq n$. For $j \leq k$ the tuples $(w_i)$ and $(v_i)$ have the same first $j$ entries, so $\operatorname{span}(w_1,\ldots,w_j) = \operatorname{span}(v_1,\ldots,v_j)$, which gives the claim.
\end{proof}

\begin{lemma}[Orthonormal Basis Adapted to a Non-Null Line]
\label{lem:adapted-orthonormal-basis-line}
\lean{LeeLib.Ch02.exists_isOrthonormal_prefixSpan_one_eq_span}
\leanok
\uses{lem:adapted-orthonormal-basis}
Suppose $(V,q)$ is an $n$-dimensional scalar product space with $n \geq 1$, and $x \in V$ satisfies $\langle x, x \rangle \neq 0$. Then there is an orthonormal basis $(b_1,\ldots,b_n)$ for $V$ with $\operatorname{span}(b_1) = \operatorname{span}(x)$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:adapted-orthonormal-basis}
The line $\operatorname{span}(x)$ is $1$-dimensional, since $\langle x,x\rangle \neq 0$ forces $x \neq 0$. It is nondegenerate: an element $cx$ orthogonal to every element of $\operatorname{span}(x)$ satisfies in particular $c\langle x,x \rangle = \langle cx, x\rangle = 0$, and $\langle x,x\rangle \neq 0$ gives $c = 0$. Hence the $1$-tuple $(x)$ is nondegenerate, and Lemma \ref{lem:adapted-orthonormal-basis} applied with $k = 1$ produces an orthonormal basis $(b_i)$ with $\operatorname{span}(b_1) = \operatorname{span}(x)$.
\end{proof}

The rest of the linear algebra behind Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} is pointwise as well, and involves the manifold nowhere: it takes place entirely in the scalar product space $(T_p\tilde M, \tilde g_p)$, in which the normal space $N_pM$ is a line and the tangent space $T_pM$ is that line's orthogonal complement. We isolate it here, in four steps: nondegeneracy is decided on the normal line; the tail of an adapted orthonormal basis spans the complement of its head; that tail is an orthonormal basis of the complement; and counting its signs gives the signature.

\begin{lemma}[Nondegeneracy of a Line]
\label{lem:nondegenerate-line}
\lean{LeeLib.Ch02.restrict_nondegenerate_iff_of_finrank_eq_one}
\leanok
Let $q$ be a bilinear form on a finite-dimensional vector space $V$ and let $L \subseteq V$ be a subspace with $\dim L = 1$. Then $q|_{L \times L}$ is nondegenerate if and only if $\langle v,v \rangle \neq 0$ for every nonzero $v \in L$.
\end{lemma}

\begin{proof}
\leanok
Suppose first that $q|_{L\times L}$ is nondegenerate, and let $v \in L$ be nonzero with $\langle v,v\rangle = 0$. Then $\operatorname{span}(v) \subseteq L$ and both are $1$-dimensional, so $\operatorname{span}(v) = L$. Every $w \in L$ is therefore of the form $cv$, whence $\langle v,w\rangle = c\langle v,v\rangle = 0$. Thus $v$ is orthogonal to all of $L$, and nondegeneracy forces $v = 0$, a contradiction.

Conversely, suppose every nonzero $v \in L$ satisfies $\langle v,v\rangle \neq 0$, and let $v \in L$ be orthogonal to all of $L$ on either side. Taking the other argument to be $v$ itself gives $\langle v,v \rangle = 0$, so $v = 0$ by hypothesis.
\end{proof}

This equivalence is special to dimension $1$. In dimension $\geq 2$ the condition that no nonzero vector be null says that the restriction is \emph{definite}, which is strictly stronger than nondegenerate: the form $(\xi,\tau) \mapsto \xi^2 - \tau^2$ on $\mathbb{R}^2$ is nondegenerate, yet every vector with $\xi = \pm\tau$ is null. This is why Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} is stated for hypersurfaces, whose normal spaces are lines.

\begin{lemma}[Nondegeneracy of a Hyperplane]
\label{lem:hyperplane-nondegenerate}
\lean{LeeLib.Ch02.restrict_nondegenerate_iff_forall_orthogonal}
\leanok
\uses{lem:nondegenerate-subspace,lem:nondegenerate-line,lem:orthogonal-complement}
Let $(V,q)$ be a scalar product space and let $W \subseteq V$ be a subspace with $\dim W + 1 = \dim V$. Then $q|_{W\times W}$ is nondegenerate if and only if $\langle v,v\rangle \neq 0$ for every nonzero $v \in W^\perp$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-subspace,lem:nondegenerate-line,lem:orthogonal-complement}
By Lemma \ref{lem:nondegenerate-subspace}, $q|_{W\times W}$ is nondegenerate if and only if $q|_{W^\perp \times W^\perp}$ is. By Lemma \ref{lem:orthogonal-complement}, $\dim W^\perp = \dim V - \dim W = 1$, so Lemma \ref{lem:nondegenerate-line} applies to $W^\perp$ and identifies the latter condition with the requirement that every nonzero $v \in W^\perp$ satisfy $\langle v,v \rangle \neq 0$.
\end{proof}

\begin{lemma}[The Tail of an Orthonormal Basis Spans the Complement of its Head]
\label{lem:orthonormal-tail-spans-orthogonal}
\lean{LeeLib.Ch02.span_range_succ_eq_orthogonal}
\leanok
\uses{lem:orthogonal-complement}
Let $q$ be a nondegenerate bilinear form on a finite-dimensional vector space $V$ and let $(b_1,\ldots,b_n)$ be an orthonormal basis for $V$. Then $\operatorname{span}(b_2,\ldots,b_n) = \operatorname{span}(b_1)^\perp$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthogonal-complement}
Each $b_j$ with $j \geq 2$ is orthogonal to $b_1$, hence to every element $cb_1$ of $\operatorname{span}(b_1)$; as $\operatorname{span}(b_1)^\perp$ is a subspace, this gives $\operatorname{span}(b_2,\ldots,b_n) \subseteq \operatorname{span}(b_1)^\perp$. The reverse inclusion is a matter of dimension. The vectors $b_2,\ldots,b_n$ are linearly independent, being part of a basis, so $\dim\operatorname{span}(b_2,\ldots,b_n) = n-1$. On the other hand $\langle b_1,b_1\rangle = \pm 1 \neq 0$ forces $b_1 \neq 0$, so $\dim \operatorname{span}(b_1) = 1$, and Lemma \ref{lem:orthogonal-complement} gives $\dim \operatorname{span}(b_1)^\perp = n-1$. A subspace contained in another of the same finite dimension equals it.
\end{proof}

\begin{lemma}[Orthonormal Basis of the Complement of a Line]
\label{lem:orthonormal-basis-orthogonal-complement}
\lean{LeeLib.Ch02.exists_basis_isOrthonormal_orthogonal_span_singleton}
\leanok
\uses{lem:orthonormal-tail-spans-orthogonal}
Let $q$ be a nondegenerate bilinear form on a finite-dimensional vector space $V$ and let $(b_1,\ldots,b_n)$ be an orthonormal basis for $V$. Then $(b_2,\ldots,b_n)$ is an orthonormal basis for $\operatorname{span}(b_1)^\perp$ equipped with the restriction of $q$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthonormal-tail-spans-orthogonal}
By Lemma \ref{lem:orthonormal-tail-spans-orthogonal} the vectors $b_2,\ldots,b_n$ lie in $\operatorname{span}(b_1)^\perp$ and span it. They are linearly independent in $V$, being part of a basis, and hence linearly independent as a family in the subspace, the inclusion being injective; so they form a basis of $\operatorname{span}(b_1)^\perp$. Orthonormality is inherited, the restricted form taking the same values on them as $q$: $\langle b_i,b_j\rangle = 0$ for $i \neq j$, and $\langle b_i,b_i\rangle = \pm 1$.
\end{proof}

\begin{lemma}[The Complement of a Non-Null Line is a Hyperplane]
\label{lem:orthogonal-complement-non-null-line-hyperplane}
\lean{LeeLib.Ch02.finrank_orthogonal_span_singleton}
\leanok
\uses{lem:orthogonal-complement}
Let $(V,q)$ be a scalar product space of dimension $n$ and let $x \in V$ be nonzero. Then $\dim \operatorname{span}(x)^\perp = n - 1$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthogonal-complement}
Since $x \neq 0$, the line $\operatorname{span}(x)$ has dimension $1$, and Lemma \ref{lem:orthogonal-complement} gives $\dim \operatorname{span}(x) + \dim \operatorname{span}(x)^\perp = n$.
\end{proof}

This is the pointwise form of $\dim T_pM = \dim T_p\tilde M - 1$, and it is what makes Lemma \ref{lem:hyperplane-nondegenerate} applicable to the tangent space of a hypersurface.

\begin{lemma}[The Complement of a Non-Null Line is a Scalar Product Space]
\label{lem:orthogonal-complement-non-null-line-nondegenerate}
\lean{LeeLib.Ch02.restrict_orthogonal_span_singleton_nondegenerate}
\leanok
\uses{lem:nondegenerate-subspace,lem:adapted-orthonormal-basis-line}
Let $(V,q)$ be a scalar product space and let $x \in V$ satisfy $\langle x,x\rangle \neq 0$. Then the restriction of $q$ to $\operatorname{span}(x)^\perp$ is nondegenerate.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:nondegenerate-subspace,lem:adapted-orthonormal-basis-line}
The line $\operatorname{span}(x)$ is nondegenerate, as recorded in the proof of Lemma \ref{lem:adapted-orthonormal-basis-line}: an element $cx$ orthogonal to all of $\operatorname{span}(x)$ satisfies $c\langle x,x\rangle = 0$, and $\langle x,x \rangle \neq 0$ gives $c = 0$. Lemma \ref{lem:nondegenerate-subspace} then shows that $\operatorname{span}(x)^\perp$ is nondegenerate as well.
\end{proof}

\begin{proposition}[Signature of the Complement of a Non-Null Line]
\label{prop:hypersurface-signature-pointwise}
\lean{LeeLib.Ch02.sigPos_sigNeg_restrict_orthogonal_span_singleton}
\leanok
\uses{lem:adapted-orthonormal-basis-line,lem:orthonormal-basis-orthogonal-complement,lem:orthogonal-complement-non-null-line-nondegenerate,prop:sylvesters-law-inertia}
Let $(V,q)$ be a scalar product space of signature $(r,s)$ and let $x \in V$ satisfy $\langle x,x\rangle \neq 0$. Let $(r',s')$ be the signature of $\operatorname{span}(x)^\perp$ with the restriction of $q$, which is a scalar product space by Lemma \ref{lem:orthogonal-complement-non-null-line-nondegenerate}. If $\langle x,x\rangle > 0$, then $r' + 1 = r$ and $s' = s$. If $\langle x,x \rangle < 0$, then $r' = r$ and $s' + 1 = s$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:adapted-orthonormal-basis-line,lem:orthonormal-basis-orthogonal-complement,prop:sylvesters-law-inertia}
By Lemma \ref{lem:adapted-orthonormal-basis-line} there is an orthonormal basis $(b_1,\ldots,b_n)$ for $V$ with $\operatorname{span}(b_1) = \operatorname{span}(x)$, so that $\operatorname{span}(x)^\perp = \operatorname{span}(b_1)^\perp$; and by Lemma \ref{lem:orthonormal-basis-orthogonal-complement} the tail $(b_2,\ldots,b_n)$ is an orthonormal basis of that complement.

By Proposition \ref{prop:sylvesters-law-inertia}, $r$ is the number of indices $i$ with $\langle b_i,b_i\rangle = 1$ and $s$ the number with $\langle b_i,b_i \rangle = -1$; likewise $r'$ and $s'$ are the corresponding counts over $i \geq 2$. The two pairs of counts therefore differ exactly by the contribution of the index $1$, which contributes to $r$ if $\langle b_1,b_1\rangle = 1$ and to $s$ if $\langle b_1,b_1 \rangle = -1$.

It remains to determine the sign of $\langle b_1, b_1 \rangle$. Since $\operatorname{span}(b_1) = \operatorname{span}(x)$, we may write $b_1 = cx$ with $c \neq 0$, and then $\langle b_1,b_1\rangle = c^2 \langle x,x\rangle$ has the same sign as $\langle x,x \rangle$. So if $\langle x,x\rangle > 0$ then $\langle b_1,b_1\rangle = 1$, giving $r' + 1 = r$ and $s' = s$; and if $\langle x,x\rangle < 0$ then $\langle b_1,b_1\rangle = -1$, giving $r' = r$ and $s' + 1 = s$.
\end{proof}

Proposition \ref{prop:hypersurface-signature-pointwise} is stated about the complement of a line, whereas Lee states his conclusion about the hyperplane $T_pM$ itself, and Lemma \ref{lem:hyperplane-nondegenerate} is about that hyperplane too. The two are identified by Lemma \ref{lem:orthogonal-complement-involution}: a hyperplane is the orthogonal complement of its own normal line. Making that identification puts both halves of the proposition in the same terms.

\begin{lemma}[A Hyperplane is the Complement of its Normal Line]
\label{lem:hyperplane-eq-complement-normal-line}
\lean{LeeLib.Ch02.orthogonal_span_singleton_eq}
\leanok
\uses{lem:orthogonal-complement,lem:orthogonal-complement-involution}
Let $(V,q)$ be a scalar product space, let $W \subseteq V$ satisfy $\dim W + 1 = \dim V$, and let $x \in W^\perp$ be nonzero. Then $\operatorname{span}(x)^\perp = W$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:orthogonal-complement,lem:orthogonal-complement-involution}
By Lemma \ref{lem:orthogonal-complement}, $\dim W^\perp = \dim V - \dim W = 1$. Since $x$ is a nonzero element of $W^\perp$, the line $\operatorname{span}(x)$ is a $1$-dimensional subspace of the $1$-dimensional space $W^\perp$, so $\operatorname{span}(x) = W^\perp$. Hence $\operatorname{span}(x)^\perp = (W^\perp)^\perp = W$, the last step by Lemma \ref{lem:orthogonal-complement-involution}.
\end{proof}

\begin{proposition}[Signature of a Hypersurface's Tangent Space]
\label{prop:hypersurface-signature-pointwise-hyperplane}
\lean{LeeLib.Ch02.sigPos_sigNeg_restrict_of_mem_orthogonal}
\leanok
\uses{prop:hypersurface-signature-pointwise,lem:hyperplane-eq-complement-normal-line}
Let $(V,q)$ be a scalar product space of signature $(r,s)$, let $W \subseteq V$ satisfy $\dim W + 1 = \dim V$, and let $x \in W^\perp$ be nonzero. Let $(r',s')$ be the signature of $W$ with the restriction of $q$. If $\langle x,x\rangle > 0$, then $r' + 1 = r$ and $s' = s$. If $\langle x,x\rangle < 0$, then $r' = r$ and $s' + 1 = s$.
\end{proposition}

\begin{proof}
\leanok
\uses{prop:hypersurface-signature-pointwise,lem:hyperplane-eq-complement-normal-line}
By Lemma \ref{lem:hyperplane-eq-complement-normal-line}, $W = \operatorname{span}(x)^\perp$, so the claim is exactly Proposition \ref{prop:hypersurface-signature-pointwise} applied to $x$.
\end{proof}

Lemma \ref{lem:hyperplane-nondegenerate} and Proposition \ref{prop:hypersurface-signature-pointwise-hyperplane} together are Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} at a single point: taking $V = T_p\tilde M$, $q = \tilde g_p$ and $W = T_pM$, whose orthogonal complement is $N_pM$, the first says that $\iota^*\tilde g$ is nondegenerate at $p$ exactly when every nonzero normal vector at $p$ is non-null, and the second says that the signature of $T_pM$ is then $(r-1,s)$ or $(r,s-1)$ according to the sign of the normal vectors. What Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} adds is the passage from a point to the manifold: that these pointwise conditions, holding at every $p$, make $\iota^*\tilde g$ a pseudo-Riemannian metric of constant signature.

That passage has exactly three steps. First, $T_pM$ and $N_pM$ must be a hyperplane and its complement \emph{in one and the same space}; they already are, both being subspaces of the ambient tangent space $T_{F(p)}\tilde{M}$, which is the fibre at $p$ of the pullback bundle $F^*T\tilde{M}$, so nothing need be transported. Second, $N_pM$, defined by orthogonality against the tangent space, must be recognized as the orthogonal complement of that tangent space. Third, the signature of $g$ at $p$ is that of a form on $T_pM$, whereas the pointwise results compute the signature of $\tilde{g}_{F(p)}$ restricted to $dF_p(T_pM)$, a form on a subspace of the ambient tangent space; these are different vector spaces, and $dF_p$ is an isometry between them.

\begin{lemma}[The Normal Space is an Orthogonal Complement]
\label{lem:pseudo-normal-space-eq-orthogonal}
\lean{LeeLib.Ch02.pseudoNormalSpace_eq_orthogonal}
\leanok
\uses{def:pseudo-normal-space,def:pseudo-riemannian-metric}
Let $(\tilde{M},\tilde{g})$ be a pseudo-Riemannian manifold and let $F : M \to \tilde{M}$ be smooth. For every $x \in M$,
\[
  N_xM \;=\; \bigl(dF_x(T_xM)\bigr)^{\perp},
\]
the orthogonal complement being taken in the scalar product space $(T_{F(x)}\tilde{M}, \tilde{g}_{F(x)})$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-normal-space,def:pseudo-riemannian-metric}
Definition \ref{def:pseudo-normal-space} places $v$ in $N_xM$ when $\tilde{g}_{F(x)}(v,w) = 0$ for every $w \in dF_x(T_xM)$, whereas $v$ lies in $(dF_x(T_xM))^\perp$ when $\tilde{g}_{F(x)}(w,v) = 0$ for every such $w$. The two conditions differ only in which argument the tested vector occupies, and $\tilde{g}_{F(x)}$ is symmetric, so each is equivalent to the other term by term.
\end{proof}

\begin{lemma}[A Hypersurface's Tangent Space is a Hyperplane]
\label{lem:hypersurface-tangent-hyperplane}
\lean{LeeLib.Ch02.finrank_tangentRange_add_one}
\leanok
Let $F : M \to \tilde{M}$ be smooth, let $x \in M$ be a point at which $dF_x$ is injective, and suppose $\dim M + 1 = \dim \tilde{M}$. Then
\[
  \dim dF_x(T_xM) + 1 \;=\; \dim T_{F(x)}\tilde{M} .
\]
\end{lemma}

\begin{proof}
\leanok
Since $dF_x$ is injective, it is a linear isomorphism onto its image, so $\dim dF_x(T_xM) = \dim T_xM = \dim M$. The ambient tangent space $T_{F(x)}\tilde{M}$ has dimension $\dim\tilde{M}$. Substituting both into the hypothesis $\dim M + 1 = \dim \tilde{M}$ gives the claim.
\end{proof}

This is the codimension hypothesis in the form the pointwise results of this section ask for it, and it is the only place where $\dim M = \dim\tilde{M} - 1$ is used.

\begin{lemma}[The Normal Space of a Hypersurface is a Line]
\label{lem:pseudo-normal-space-line}
\lean{LeeLib.Ch02.finrank_pseudoNormalSpace}
\leanok
\uses{def:pseudo-normal-space,lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:orthogonal-complement}
Let $(\tilde{M},\tilde{g})$ be a pseudo-Riemannian manifold, let $F : M \to \tilde{M}$ be smooth with $\dim M + 1 = \dim\tilde{M}$, and let $x \in M$ be a point at which $dF_x$ is injective. Then $\dim N_xM = 1$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:orthogonal-complement}
Write $W = dF_x(T_xM)$. By Lemma \ref{lem:pseudo-normal-space-eq-orthogonal}, $N_xM = W^\perp$, so Lemma \ref{lem:orthogonal-complement} applied in the scalar product space $(T_{F(x)}\tilde{M}, \tilde{g}_{F(x)})$ gives $\dim W + \dim N_xM = \dim T_{F(x)}\tilde{M}$. Lemma \ref{lem:hypersurface-tangent-hyperplane} gives $\dim W + 1 = \dim T_{F(x)}\tilde{M}$. Cancelling $\dim W$ leaves $\dim N_xM = 1$.
\end{proof}

This is why a \emph{single} non-null normal vector decides Proposition \ref{prop:pseudo-riemannian-hypersurface-signature}: any nonzero $v \in N_xM$ spans the line, so the sign of $\tilde{g}(v,v)$ cannot depend on which nonzero normal is chosen, rescaling $v \mapsto cv$ multiplying that number by $c^2 > 0$. We record this reduction, first as a statement about an arbitrary line and then in the form Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} consumes.

\begin{lemma}[The Sign of a Line is Decided by One of its Vectors]
\label{lem:line-sign-one-vector}
\lean{LeeLib.Ch02.pos_of_mem_of_finrank_eq_one}
\leanok
Let $V$ be a real vector space, $B$ a bilinear form on $V$, and $N \leq V$ a subspace with
$\dim N = 1$. Suppose $x \in N$ is nonzero with $B(x,x) > 0$. Then $B(v,v) > 0$ for every
nonzero $v \in N$.
\end{lemma}

\begin{proof}
\leanok
Since $\dim N = 1$ and $x \in N$ is nonzero, the line $\mathbb{R}x$ is contained in $N$ and has the
same dimension, so $N = \mathbb{R}x$. A nonzero $v \in N$ is therefore $v = ax$ for some
$a \in \mathbb{R}$, and $a \neq 0$ because $v \neq 0$. Bilinearity gives
$B(v,v) = B(ax, ax) = a^2 B(x,x)$, a product of the positive numbers $a^2$ and $B(x,x)$.
\end{proof}

\begin{lemma}[One Positive Normal Makes Every Normal Positive]
\label{lem:one-normal-decides-sign}
\lean{LeeLib.Ch02.forall_pos_of_pos_mem_pseudoNormalSpace}
\leanok
\uses{def:pseudo-normal-space,lem:pseudo-normal-space-line,lem:line-sign-one-vector}
Let $(\tilde{M},\tilde{g})$, $F$ and $x$ be as in Lemma \ref{lem:pseudo-normal-space-line}, and
suppose some nonzero $v_0 \in N_xM$ satisfies $\tilde{g}(v_0,v_0) > 0$. Then
$\tilde{g}(v,v) > 0$ for every nonzero $v \in N_xM$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:pseudo-normal-space-line,lem:line-sign-one-vector}
By Lemma \ref{lem:pseudo-normal-space-line}, $\dim N_xM = 1$. Apply Lemma
\ref{lem:line-sign-one-vector} to the subspace $N_xM$ of the ambient tangent space, the form
$\tilde{g}_{F(x)}$, and the vector $v_0$.
\end{proof}

\begin{lemma}[One Negative Normal Makes Every Normal Negative]
\label{lem:one-normal-decides-sign-neg}
\lean{LeeLib.Ch02.forall_neg_of_neg_mem_pseudoNormalSpace}
\leanok
\uses{def:pseudo-normal-space,lem:pseudo-normal-space-line,lem:line-sign-one-vector}
Let $(\tilde{M},\tilde{g})$, $F$ and $x$ be as in Lemma \ref{lem:pseudo-normal-space-line}, and
suppose some nonzero $v_0 \in N_xM$ satisfies $\tilde{g}(v_0,v_0) < 0$. Then
$\tilde{g}(v,v) < 0$ for every nonzero $v \in N_xM$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:one-normal-decides-sign}
Apply Lemma \ref{lem:one-normal-decides-sign} to the form $-\tilde{g}$, for which $v_0$ and every
nonzero normal satisfy the positivity hypothesis and conclusion respectively; note $N_xM$ is the
same subspace for $\tilde{g}$ and $-\tilde{g}$.
\end{proof}

This is the form in which the hypothesis of Proposition
\ref{prop:pseudo-riemannian-hypersurface-signature} is met in practice: a caller does not verify a
sign for every normal vector, it exhibits one distinguished normal — for a regular level set, the
gradient of the defining function — and checks the sign there.

\begin{lemma}[A Hypersurface has a Nonzero Normal Vector at Every Point]
\label{lem:exists-nonzero-normal}
\lean{LeeLib.Ch02.exists_ne_zero_mem_pseudoNormalSpace}
\leanok
\uses{def:pseudo-normal-space,lem:pseudo-normal-space-line}
Let $(\tilde{M},\tilde{g})$, $F$ and $x$ be as in Lemma \ref{lem:pseudo-normal-space-line}. Then there is a nonzero $v \in N_xM$.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:pseudo-normal-space-line}
By Lemma \ref{lem:pseudo-normal-space-line}, $\dim N_xM = 1 > 0$, so $N_xM$ is not the zero subspace and therefore contains a nonzero vector.
\end{proof}

Slight as it is, this is what allows the hypothesis ``$\tilde{g}(v,v) > 0$ for every nonzero normal vector $v$'' to be \emph{used} rather than merely assumed: were $N_xM$ trivial, that hypothesis would hold vacuously and yield nothing.

\begin{lemma}[The Differential is an Isometry onto the Tangent Space]
\label{lem:differential-isometry-tangent-range}
\lean{LeeLib.Ch02.isometryEquivRestrictTangentRange}
\leanok
\uses{def:induced-tensor-along-map,lem:pseudo-submanifold-immersion}
Let $F : M \to \tilde{M}$ be smooth and suppose the pseudo-Riemannian metric $g$ on $M$ is induced by $\tilde{g}$ along $F$. Then for every $x \in M$ the differential $dF_x$ is an isometry of scalar product spaces from $(T_xM, g_x)$ onto $\bigl(dF_x(T_xM),\, \tilde{g}_{F(x)}|_{dF_x(T_xM)}\bigr)$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:induced-tensor-along-map,lem:pseudo-submanifold-immersion}
By Lemma \ref{lem:pseudo-submanifold-immersion} the map $dF_x$ is injective, hence a linear isomorphism onto its image $dF_x(T_xM)$. It preserves the forms: for $v,w \in T_xM$, Definition \ref{def:induced-tensor-along-map} gives $g_x(v,w) = \tilde{g}_{F(x)}(dF_xv,\, dF_xw)$, and since $dF_xv$ and $dF_xw$ both lie in $dF_x(T_xM)$, the right-hand side is the restriction of $\tilde{g}_{F(x)}$ to that subspace evaluated at them. A form-preserving linear isomorphism is an isometry.
\end{proof}

This is the formal content of Lee's ``we usually identify $T_pM$ with its image in $T_p\tilde{M}$ under $d\iota_p$''. The identification is not free of charge: $g_x$ is a form on $T_xM$, whereas the pointwise results compute the signature of $\tilde{g}_{F(x)}$ restricted to $dF_x(T_xM)$, a form on a subspace of a different vector space. The isometry is what makes the two signatures the same number.

\begin{lemma}[The Signature of a Submanifold is Computed in the Ambient Tangent Space]
\label{lem:submanifold-signature-in-ambient-fibre}
\lean{LeeLib.Ch02.sigPosAt_eq_sigPos_restrict_tangentRange, LeeLib.Ch02.sigNegAt_eq_sigNeg_restrict_tangentRange}
\leanok
\uses{def:pseudo-metric-signature-at,lem:differential-isometry-tangent-range,prop:sylvesters-law-inertia}
Let $F : M \to \tilde{M}$ be smooth and suppose the pseudo-Riemannian metric $g$ on $M$ is induced by $\tilde{g}$ along $F$. Then for every $x \in M$ the signature of $g$ at $x$ is the signature of the scalar product space $\bigl(dF_x(T_xM),\, \tilde{g}_{F(x)}|_{dF_x(T_xM)}\bigr)$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-metric-signature-at,lem:differential-isometry-tangent-range,prop:sylvesters-law-inertia}
By Definition \ref{def:pseudo-metric-signature-at} the signature of $g$ at $x$ is the signature of the scalar product space $(T_xM, g_x)$, and by Lemma \ref{lem:differential-isometry-tangent-range} the map $dF_x$ is an isometry from that space onto the displayed one. It therefore suffices that isometric scalar product spaces have the same signature. An isometry carries an orthonormal basis $(b_1,\ldots,b_n)$ of the source to a basis of the target, being a linear isomorphism, and that image basis is orthonormal with $\langle dF_xb_i, dF_xb_i\rangle = \langle b_i,b_i\rangle$ for every $i$, the forms being preserved. So the two bases have the same number of members of square $+1$ and the same number of square $-1$, and by Proposition \ref{prop:sylvesters-law-inertia} those two counts are the signature of the respective space.
\end{proof}

\begin{proposition}[Nondegeneracy of a Hypersurface]
\label{prop:pseudo-riemannian-hypersurface-nondegenerate}
\lean{LeeLib.Ch02.restrict_tangentRange_nondegenerate_iff}
\leanok
\dcref{ch2:2.70}
\uses{def:pseudo-riemannian-metric,def:pseudo-normal-space,lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:hyperplane-nondegenerate}
Suppose $(\tilde{M},\tilde{g})$ is a pseudo-Riemannian manifold and $F : M \to \tilde{M}$ is smooth with $\dim M + 1 = \dim\tilde{M}$, and let $x \in M$ be a point at which $dF_x$ is injective. Then the restriction of $\tilde{g}_{F(x)}$ to $dF_x(T_xM)$ is nondegenerate if and only if $\tilde{g}_{F(x)}(v,v) \neq 0$ for every nonzero $v \in N_xM$.
\end{proposition}

\begin{proof}
\leanok
\uses{lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:hyperplane-nondegenerate}
Apply Lemma \ref{lem:hyperplane-nondegenerate} in the scalar product space $V = (T_{F(x)}\tilde{M}, \tilde{g}_{F(x)})$ with $W = dF_x(T_xM)$. Its hypothesis $\dim W + 1 = \dim V$ is Lemma \ref{lem:hypersurface-tangent-hyperplane}, available because $dF_x$ is injective. Its conclusion is that $\tilde{g}_{F(x)}|_W$ is nondegenerate if and only if every nonzero $v \in W^\perp$ satisfies $\tilde{g}_{F(x)}(v,v) \neq 0$; and $W^\perp = N_xM$ by Lemma \ref{lem:pseudo-normal-space-eq-orthogonal}.
\end{proof}

Here the induced tensor appears as $\tilde{g}$ restricted to $dF_x(T_xM)$ rather than as a metric on $M$, and $F$ is assumed only to be an immersion. This is forced: Definition \ref{def:induced-tensor-along-map} presupposes a \emph{nondegenerate} $g$ on $M$, and nondegeneracy of the induced tensor is precisely what this proposition decides, so it cannot appear among its hypotheses.

\begin{proposition}[A Single Normal Vector Decides the Signature at a Point]
\label{prop:hypersurface-signature-at-a-point}
\lean{LeeLib.Ch02.sigPosAt_sigNegAt_of_mem_pseudoNormalSpace}
\leanok
\uses{def:pseudo-metric-has-signature,def:pseudo-normal-space,def:induced-tensor-along-map,lem:submanifold-signature-in-ambient-fibre,prop:hypersurface-signature-pointwise-hyperplane,lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:pseudo-submanifold-immersion}
Suppose $(\tilde{M},\tilde{g})$ is a pseudo-Riemannian manifold of signature $(r,s)$ and $F : M \to \tilde{M}$ presents $M$ as a pseudo-Riemannian submanifold with induced metric $g$, where $\dim M + 1 = \dim\tilde{M}$. Let $x \in M$, let $(r_x,s_x)$ be the signature of $g$ at $x$, and let $v \in N_xM$ be nonzero. If $\tilde{g}_{F(x)}(v,v) > 0$, then $r_x + 1 = r$ and $s_x = s$. If $\tilde{g}_{F(x)}(v,v) < 0$, then $r_x = r$ and $s_x + 1 = s$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:pseudo-metric-has-signature,lem:submanifold-signature-in-ambient-fibre,prop:hypersurface-signature-pointwise-hyperplane,lem:pseudo-normal-space-eq-orthogonal,lem:hypersurface-tangent-hyperplane,lem:pseudo-submanifold-immersion}
Put $V = (T_{F(x)}\tilde{M}, \tilde{g}_{F(x)})$ and $W = dF_x(T_xM)$. By Definition \ref{def:pseudo-metric-has-signature} applied at the point $F(x)$, the scalar product space $V$ has signature $(r,s)$. By Lemma \ref{lem:pseudo-submanifold-immersion} the differential $dF_x$ is injective, so Lemma \ref{lem:hypersurface-tangent-hyperplane} gives $\dim W + 1 = \dim V$; and $v$ is a nonzero element of $W^\perp$ by Lemma \ref{lem:pseudo-normal-space-eq-orthogonal}. Proposition \ref{prop:hypersurface-signature-pointwise-hyperplane}, applied to $V$, $W$ and $v$, therefore computes the signature $(r',s')$ of $W$ with the restriction of $\tilde{g}_{F(x)}$: it satisfies $r' + 1 = r$ and $s' = s$ when $\tilde{g}_{F(x)}(v,v) > 0$, and $r' = r$ and $s' + 1 = s$ when $\tilde{g}_{F(x)}(v,v) < 0$. Finally $(r',s') = (r_x,s_x)$ by Lemma \ref{lem:submanifold-signature-in-ambient-fibre}.
\end{proof}

The conclusion is recorded as $r_x + 1 = r$ rather than $r_x = r - 1$. The two are equivalent, but the first carries the additional information that $r \geq 1$, which is what makes the subtraction in the statement below meaningful.

\begin{proposition}
\label{prop:pseudo-riemannian-hypersurface-signature}
\lean{LeeLib.Ch02.hasSignature_of_forall_normal_pos, LeeLib.Ch02.hasSignature_of_forall_normal_neg}
\leanok
\dcref{ch2:2.70}
\uses{def:pseudo-riemannian-metric,def:pseudo-metric-has-signature,def:induced-tensor-along-map,def:pseudo-normal-space,lem:exists-nonzero-normal,prop:hypersurface-signature-at-a-point,lem:pseudo-submanifold-immersion}
Suppose $(\tilde{M}, \tilde{g})$ is a pseudo-Riemannian manifold of signature $(r,s)$, and $F : M \to \tilde{M}$ presents $M$ as a pseudo-Riemannian submanifold with induced metric $g$, where $\dim M + 1 = \dim\tilde{M}$; the case Lee has in mind is that of a smooth hypersurface $M \subseteq \tilde{M}$ and its inclusion $\iota$, with $g = \iota^*\tilde{g}$. If $\tilde{g}(v,v) > 0$ for every $p \in M$ and every nonzero normal vector $v \in N_pM$, then $g$ has signature $(r-1,s)$; and if $\tilde{g}(v,v) < 0$ for every such vector, then $g$ has signature $(r,s-1)$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:pseudo-metric-has-signature,lem:exists-nonzero-normal,prop:hypersurface-signature-at-a-point,lem:pseudo-submanifold-immersion}
Take the first case, and fix $p \in M$; let $(r_p,s_p)$ be the signature of $g$ at $p$. By Lemma \ref{lem:pseudo-submanifold-immersion} the differential $dF_p$ is injective, so Lemma \ref{lem:exists-nonzero-normal} supplies a nonzero $v \in N_pM$. By hypothesis $\tilde{g}_{F(p)}(v,v) > 0$, so Proposition \ref{prop:hypersurface-signature-at-a-point} gives $r_p + 1 = r$ and $s_p = s$; in particular $r \geq 1$ and $(r_p,s_p) = (r-1,s)$. As $p$ was arbitrary, $g$ has signature $(r-1,s)$ in the sense of Definition \ref{def:pseudo-metric-has-signature}. The second case is identical, using instead the alternative $r_p = r$, $s_p + 1 = s$ of Proposition \ref{prop:hypersurface-signature-at-a-point}, which gives $s \geq 1$ and $(r_p,s_p) = (r,s-1)$.
\end{proof}

\begin{lemma}[The Induced Tensor of a Non-Null Hypersurface is Nondegenerate]
\label{lem:pseudo-hypersurface-induced-tensor-nondegenerate}
\lean{LeeLib.Ch02.exists_pseudoPullbackForm_ne_zero}
\leanok
\uses{def:pseudo-pullback-form,def:pseudo-normal-space,prop:pseudo-riemannian-hypersurface-nondegenerate}
Suppose $(\tilde{M},\tilde{g})$ is a pseudo-Riemannian manifold and $F : M \to \tilde{M}$ is a smooth immersion with $\dim M + 1 = \dim\tilde{M}$, and suppose $\tilde{g}_{F(p)}(v,v) \neq 0$ for every $p \in M$ and every nonzero $v \in N_pM$. Then the pullback tensor $F^*\tilde{g}$ of Definition \ref{def:pseudo-pullback-form} is nondegenerate at every point: for every $p \in M$ and every nonzero $v \in T_pM$ there is a $w \in T_pM$ with $(F^*\tilde{g})_p(v,w) \neq 0$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-pullback-form,prop:pseudo-riemannian-hypersurface-nondegenerate}
Fix $p \in M$ and a nonzero $v \in T_pM$, and put $W = dF_p(T_pM)$. By Proposition \ref{prop:pseudo-riemannian-hypersurface-nondegenerate} the restriction of $\tilde{g}_{F(p)}$ to $W$ is nondegenerate. The vector $u = dF_p(v)$ lies in $W$, and $u \neq 0$ because $dF_p$ is injective and $v \neq 0$. Nondegeneracy of the restriction therefore supplies a $u' \in W$ with $\tilde{g}_{F(p)}(u,u') \neq 0$. Since $W$ is the image of $dF_p$, we may write $u' = dF_p(w)$ for some $w \in T_pM$, and then
\[
  (F^*\tilde{g})_p(v,w) = \tilde{g}_{F(p)}\bigl(dF_pv,\, dF_pw\bigr) = \tilde{g}_{F(p)}(u,u') \neq 0 .
\]
\end{proof}

This is what turns Proposition \ref{prop:pseudo-riemannian-hypersurface-nondegenerate} from a criterion into a construction, and with it the two halves of Lee's proposition combine into a single statement in which the induced metric is produced rather than presupposed.

\begin{corollary}[Signature of a Pseudo-Riemannian Hypersurface]
\label{cor:pseudo-riemannian-hypersurface-submanifold}
\lean{LeeLib.Ch02.hasSignature_pseudoPullbackMetric_of_forall_normal_pos, LeeLib.Ch02.hasSignature_pseudoPullbackMetric_of_forall_normal_neg}
\leanok
\dcref{ch2:2.70}
\uses{def:pseudo-metric-has-signature,def:pseudo-normal-space,def:pseudo-induced-metric,lem:pseudo-hypersurface-induced-tensor-nondegenerate,lem:pseudo-riemannian-submanifold-induced-tensor,prop:pseudo-riemannian-hypersurface-signature}
Suppose $(\tilde{M},\tilde{g})$ is a pseudo-Riemannian manifold of signature $(r,s)$ and $F : M \to \tilde{M}$ is a smooth immersion with $\dim M + 1 = \dim\tilde{M}$. If $\tilde{g}_{F(p)}(v,v) > 0$ for every $p \in M$ and every nonzero $v \in N_pM$, then the pullback tensor $F^*\tilde{g}$ is a pseudo-Riemannian metric on $M$ of signature $(r-1,s)$, so that $F$ presents $M$ as a pseudo-Riemannian submanifold of that signature. If instead $\tilde{g}_{F(p)}(v,v) < 0$ for every such $v$, then $F^*\tilde{g}$ is a pseudo-Riemannian metric of signature $(r,s-1)$.
\end{corollary}

\begin{proof}
\leanok
\uses{def:pseudo-induced-metric,lem:pseudo-hypersurface-induced-tensor-nondegenerate,lem:pseudo-riemannian-submanifold-induced-tensor,prop:pseudo-riemannian-hypersurface-signature}
In either case every nonzero normal vector is non-null, so Lemma \ref{lem:pseudo-hypersurface-induced-tensor-nondegenerate} shows that $F^*\tilde{g}$ is nondegenerate at every point of $M$, and Definition \ref{def:pseudo-induced-metric} accordingly makes $g = F^*\tilde{g}$ a pseudo-Riemannian metric on $M$. By Lemma \ref{lem:pseudo-riemannian-submanifold-induced-tensor} this $g$ is induced by $\tilde{g}$ along $F$, so $F$ presents $M$ as a pseudo-Riemannian submanifold. Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} applied to $g$ then gives signature $(r-1,s)$ in the positive case and $(r,s-1)$ in the negative one.
\end{proof}

Every hypothesis of this corollary is one a reader has in hand: an ambient metric, a codimension-one immersion, and the sign of the normal vectors. Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} takes the induced metric $g$ as given, and Definition \ref{def:pseudo-induced-metric} cannot produce it without nondegeneracy; for an indefinite $\tilde{g}$ that nondegeneracy is available from no source other than the non-nullity of the normals.

If $(\bar{M}, \bar{g})$ is a pseudo-Riemannian manifold and $f \in C^\infty(M)$, then the \emph{gradient} of $f$ is defined as the smooth vector field $\text{grad} f = (df)^\#$ just as in the Riemannian case.

The gradient of the preceding paragraph makes sense on any pseudo-Riemannian manifold, and its fibrewise construction is the following. On a vector space carrying a nondegenerate symmetric bilinear form there is no Riesz representation to appeal to, since the form need not be positive definite; nondegeneracy alone suffices.

\begin{definition}[The gradient of a functional relative to a scalar product]
\label{def:bilinear-gradient}
\lean{LeeLib.Ch02.bilinGrad}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Let $V$ be a finite-dimensional real vector space, let $B$ be a nondegenerate symmetric bilinear form on $V$, and let $a \in V^*$. The \emph{$B$-gradient} $a^\sharp$ of $a$ is the unique $v \in V$ with $B(v, w) = a(w)$ for all $w \in V$.
\end{definition}

\begin{proof}
\leanok
\uses{lem:nondegenerate-symmetric-bilinear}
Consider the linear map $\Phi : V \to V^*$, $\Phi(v) = B(v,\cdot)$. Its kernel is $\{v : B(v,w) = 0 \text{ for all } w\}$, which is trivial exactly because $B$ is nondegenerate, so $\Phi$ is injective. Since $\dim V^* = \dim V < \infty$, $\Phi$ is therefore bijective, and $a^\sharp := \Phi^{-1}(a)$ is the unique vector with $B(a^\sharp, w) = a(w)$ for all $w$.
\end{proof}

\begin{lemma}[The $B$-gradient is orthogonal to the kernel of the functional]
\label{lem:bilinear-gradient-orthogonal-ker}
\lean{LeeLib.Ch02.bilinGrad_mem_orthogonal_ker}
\leanok
\uses{def:bilinear-gradient}
With $V$, $B$ and $a$ as in Definition \ref{def:bilinear-gradient}, $a^\sharp \in (\ker a)^\perp$, where $\perp$ is taken with respect to $B$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:bilinear-gradient}
Let $w \in \ker a$. By symmetry of $B$ and the defining property of $a^\sharp$, $B(w, a^\sharp) = B(a^\sharp, w) = a(w) = 0$. As $w$ was arbitrary, $a^\sharp \in (\ker a)^\perp$.
\end{proof}

\begin{proposition}[Signature drop on the kernel of a functional]
\label{prop:embedding-criterion-pointwise}
\lean{LeeLib.Ch02.sigPos_sigNeg_restrict_ker_bilinGrad}
\leanok
\uses{def:bilinear-gradient,lem:bilinear-gradient-orthogonal-ker,prop:hypersurface-signature-pointwise-hyperplane}
Let $V$ be a finite-dimensional real vector space, let $B$ be a nondegenerate symmetric bilinear form on $V$ of signature $(r,s)$, and let $a \in V^*$ be nonzero. If $B(a^\sharp, a^\sharp) > 0$ then $B|_{\ker a}$ has signature $(r-1, s)$; and if $B(a^\sharp, a^\sharp) < 0$ then $B|_{\ker a}$ has signature $(r, s-1)$.

This is the fibrewise content of Corollary \ref{cor:pseudo-riemannian-embedding-criterion}, with $V = T_p \bar{M}$, $B = \bar{g}_p$, $a = df_p$, $a^\sharp = \operatorname{grad} f|_p$ and $\ker a = T_p M$.
\end{proposition}

\begin{proof}
\leanok
\uses{def:bilinear-gradient,lem:bilinear-gradient-orthogonal-ker,prop:hypersurface-signature-pointwise-hyperplane}
Since $a \neq 0$ it is surjective onto $\mathbb{R}$, so by rank--nullity $\dim \ker a + 1 = \dim V$; that is, $W = \ker a$ is a hyperplane. The vector $x = a^\sharp$ lies in $W^\perp$ by Lemma \ref{lem:bilinear-gradient-orthogonal-ker}, and $x \neq 0$: if $a^\sharp = 0$ then $a(w) = B(a^\sharp, w) = B(0,w) = 0$ for every $w$, contradicting $a \neq 0$. Proposition \ref{prop:hypersurface-signature-pointwise-hyperplane} applied to $W$ and $x$ now gives exactly the two claimed signature drops, according to the sign of $B(x,x) = B(a^\sharp, a^\sharp)$.
\end{proof}

The signature computation above does not by itself say that $B|_{\ker a}$ is \emph{nondegenerate}, which is what "pseudo-Riemannian submanifold" asserts and what constructing the induced metric requires. It comes from the same source.

\begin{lemma}[Nondegeneracy on the kernel of a functional]
\label{lem:embedding-criterion-nondegenerate-pointwise}
\lean{LeeLib.Ch02.nondegenerate_restrict_ker_bilinGrad}
\leanok
\uses{def:bilinear-gradient,lem:bilinear-gradient-orthogonal-ker,lem:hyperplane-eq-complement-normal-line,lem:orthogonal-complement-non-null-line-nondegenerate}
Let $V$ be a finite-dimensional real vector space, let $B$ be a nondegenerate symmetric bilinear form on $V$, and let $a \in V^*$ be nonzero. If $B(a^\sharp, a^\sharp) \neq 0$ then $B|_{\ker a}$ is nondegenerate.
\end{lemma}

\begin{proof}
\leanok
\uses{lem:bilinear-gradient-orthogonal-ker,lem:hyperplane-eq-complement-normal-line,lem:orthogonal-complement-non-null-line-nondegenerate}
Since $a \neq 0$, rank--nullity makes $\ker a$ a hyperplane, and $a^\sharp$ is a nonzero element of $(\ker a)^\perp$ by Lemma \ref{lem:bilinear-gradient-orthogonal-ker}. Hence $\ker a = \langle a^\sharp \rangle^{\perp}$ by Lemma \ref{lem:hyperplane-eq-complement-normal-line}. The hypothesis $B(a^\sharp, a^\sharp) \neq 0$ says the line $\langle a^\sharp\rangle$ is non-null, and the $B$-orthogonal complement of a non-null line is nondegenerate by Lemma \ref{lem:orthogonal-complement-non-null-line-nondegenerate}.
\end{proof}

\begin{corollary}
\label{cor:pseudo-riemannian-embedding-criterion}
\lean{LeeLib.Ch02.hasSignature_pseudoPullbackMetric_levelSet_of_pseudoGrad_pos, LeeLib.Ch02.hasSignature_pseudoPullbackMetric_levelSet_of_pseudoGrad_neg, LeeLib.Ch02.form_pseudoGrad_mfderiv_levelSet_val_eq_zero}
\leanok
\dcref{ch2:2.71}
\uses{def:pseudo-riemannian-metric,def:pseudo-metric-has-signature,prop:embedding-criterion-pointwise,def:bilinear-gradient,prop:regular-level-set-hypersurface-codim-one,prop:regular-level-set-tangent-kernel}
Suppose $(\bar{M}, \bar{g})$ is a pseudo-Riemannian manifold of signature $(r, s)$, $f \in C^\infty(\bar{M})$, and $M = f^{-1}(c)$ for some $c \in \mathbb{R}$. If $\bar{g}(\text{grad} f, \text{grad} f) > 0$ everywhere on $M$, then $M$ is an embedded pseudo-Riemannian submanifold of $\bar{M}$ of signature $(r-1, s)$; and if $\bar{g}(\text{grad} f, \text{grad} f) < 0$ everywhere on $M$, then $M$ is an embedded pseudo-Riemannian submanifold of $\bar{M}$ of signature $(r, s-1)$. In either case, $\text{grad} f$ is everywhere normal to $M$.
\end{corollary}

\begin{proof}
\leanok
\uses{prop:embedding-criterion-pointwise,prop:regular-level-set-hypersurface-codim-one,prop:regular-level-set-tangent-kernel}
The hypothesis $\bar{g}(\operatorname{grad} f, \operatorname{grad} f) \neq 0$ on $M$ forces $df_p \neq 0$ at each $p \in M$, so $c$ is a regular value of $f$; since $f$ is real-valued, the codimension-one case suffices, and Propositions \ref{prop:regular-level-set-hypersurface-codim-one} and \ref{prop:regular-level-set-tangent-kernel} show that $M = f^{-1}(c)$ is an embedded smooth hypersurface with $T_p M = \ker df_p$ for each $p \in M$. Fix such a $p$ and apply Proposition \ref{prop:embedding-criterion-pointwise} with $V = T_p \bar{M}$, $B = \bar{g}_p$ and $a = df_p$: since $\operatorname{grad} f|_p = (df_p)^\sharp$, the signature of $\bar{g}_p|_{T_p M}$ is $(r-1,s)$ when $\bar{g}_p(\operatorname{grad} f|_p, \operatorname{grad} f|_p) > 0$ and $(r,s-1)$ when it is negative. In particular $\bar{g}_p|_{T_p M}$ is nondegenerate at every $p$, so $M$ is a pseudo-Riemannian submanifold of the stated signature. That $\operatorname{grad} f$ is normal to $M$ is Lemma \ref{lem:bilinear-gradient-orthogonal-ker} applied at each $p$, since $T_p M = \ker df_p$.
\end{proof}

As in the Riemannian case (Proposition \ref{prop:adapted-orthonormal-frames-immersed}), a submanifold is presented by a smooth immersion $F : M \to \bar{M}$, and the bundle in which an adapted frame lives is the ambient tangent bundle along $M$, that is, the pullback $F^* T\bar{M}$, whose fibre at $p$ is $T_{F(p)}\bar{M}$. The ambient metric restricts to each such fibre, making $F^* T\bar{M}$ a bundle of scalar product spaces.

\begin{lemma}[The Ambient Metric Along a Submanifold]
\label{lem:pullback-pseudo-metric-bundle}
\lean{Bundle.ContMDiffPseudoMetric.pullback}
\leanok
\uses{def:smooth-field-scalar-products}
Let $g$ be a smooth field of scalar products on a vector bundle $V \to B$ and let $F : B' \to B$ be smooth. Then $(F^* g)_x := g_{F(x)}$ is a smooth field of scalar products on the pullback bundle $F^* V \to B'$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:smooth-field-scalar-products}
Fibrewise there is nothing to prove: $(F^*V)_x$ \emph{is} $V_{F(x)}$, so $g_{F(x)}$ is already a bilinear form on it, and symmetry and nondegeneracy are those of $g$ at $F(x)$ verbatim. It remains to see that the form varies smoothly with $x \in B'$. Read in a trivialization of $F^*V$ at $x_0$, it is the field $g$ read in the trivialization of $V$ at $F(x_0)$ and precomposed with $F$; the first is smooth because $g$ is, the second because $F$ is, and a composite of smooth maps is smooth.
\end{proof}

\begin{definition}[The Tensor Induced Along a Map]
\label{def:induced-tensor-along-map}
\lean{LeeLib.Ch02.IsPullbackAlong}
\leanok
\uses{def:pseudo-riemannian-metric}
Let $F : M \to \bar{M}$ be smooth, let $\bar{g}$ be a pseudo-Riemannian metric on $\bar{M}$, and let $g$ be a pseudo-Riemannian metric on $M$. We say $g$ is \emph{induced by $\bar{g}$ along $F$}, and write $g = F^*\bar{g}$, if
\[
  g_p(v,w) \;=\; \bar{g}_{F(p)}\bigl(dF_p(v), dF_p(w)\bigr)
  \qquad\text{for all } p \in M \text{ and } v,w \in T_pM.
\]
That such a $g$ exists is precisely Lee's condition that $M$ be a pseudo-Riemannian submanifold: it says the pullback tensor $F^*\bar{g}$ is itself nondegenerate.
\end{definition}

\begin{lemma}[A Pseudo-Riemannian Submanifold is Immersed]
\label{lem:pseudo-submanifold-immersion}
\lean{LeeLib.Ch02.injective_mfderiv_of_isPullbackAlong}
\leanok
\uses{def:induced-tensor-along-map}
Let $F : M \to \bar{M}$ be smooth and suppose the pseudo-Riemannian metric $g$ on $M$ is induced by $\bar{g}$ along $F$. Then $dF_p$ is injective for every $p \in M$; that is, $F$ is an immersion.
\end{lemma}

\begin{proof}
\leanok
\uses{def:induced-tensor-along-map}
Let $v \in T_pM$ satisfy $dF_p(v) = 0$. For every $w \in T_pM$,
\[
  g_p(v,w) = \bar{g}_{F(p)}\bigl(dF_p(v), dF_p(w)\bigr) = \bar{g}_{F(p)}\bigl(0, dF_p(w)\bigr) = 0 .
\]
So $v$ is orthogonal to all of $T_pM$ with respect to $g_p$, and nondegeneracy of $g_p$ forces $v = 0$. Since $dF_p$ is linear with trivial kernel, it is injective.
\end{proof}

This is the precise sense in which nondegeneracy of the induced tensor, and not immersivity, is the substantive hypothesis: it implies the latter. It does not come for free — the diagonal $\{\xi = \tau\}$ in $\mathbb{R}^{1,1}$ is an embedded line on which the induced tensor vanishes identically — whereas in the Riemannian case the two conditions coincide, by Lemma \ref{lem:pullback-immersion-metric}.

\begin{lemma}[A Riemannian Submanifold is a Pseudo-Riemannian Submanifold]
\label{lem:riemannian-submanifold-induced-tensor}
\lean{LeeLib.Ch02.isPullbackAlong_pullbackMetric}
\leanok
\uses{def:induced-tensor-along-map,lem:pullback-immersion-metric,def:induced-metric}
Let $F : M \to \bar{M}$ be a smooth immersion and let $\bar{g}$ be a Riemannian metric on $\bar{M}$. Then the induced metric $F^*\bar{g}$ of Definition \ref{def:induced-metric} is induced by $\bar{g}$ along $F$ in the sense of Definition \ref{def:induced-tensor-along-map}.
\end{lemma}

\begin{proof}
\leanok
\uses{def:induced-tensor-along-map,lem:pullback-immersion-metric,def:induced-metric}
By Lemma \ref{lem:pullback-immersion-metric} the tensor $F^*\bar{g}$ is positive definite, hence nondegenerate, so it is a pseudo-Riemannian metric on $M$. Its defining formula is $(F^*\bar{g})_p(v,w) = \bar{g}_{F(p)}(dF_p v, dF_p w)$, which is the condition of Definition \ref{def:induced-tensor-along-map}.
\end{proof}

For an indefinite ambient metric that route is closed, and the pullback must be built by hand. It is worth isolating first what pulling back actually uses, which is nothing beyond bilinearity --- and indeed nothing beyond \emph{linearity of the transporting maps}, which need not be differentials at all.

\begin{definition}[Transport of a Field of Bilinear Forms by a Family of Linear Maps]
\label{def:bilinear-comp-of-family}
\lean{LeeLib.Ch02.bilinearCompOf}
\leanok
Let $F : M \to \tilde{M}$ be a map, let $b$ assign to each $y \in \tilde{M}$ a bilinear form $b_y$ on $T_y\tilde{M}$, and let $A$ assign to each $p \in M$ a linear map $A_p : T_pM \to T_{F(p)}\tilde{M}$. The \emph{transport of $b$ by $A$} is the field of bilinear forms on $M$ defined by
\[
  (A^*b)_p(v,w) \;=\; b_{F(p)}\bigl(A_p v,\, A_p w\bigr), \qquad v,w \in T_pM .
\]
Each $(A^*b)_p$ is indeed bilinear, being the composite of the bilinear $b_{F(p)}$ with the linear map $A_p$ in each of its two arguments.
\end{definition}

\begin{lemma}[Transport by a Smooth Family of Linear Maps is Smooth]
\label{lem:bilinear-comp-of-family-smooth}
\lean{LeeLib.Ch02.contMDiffAt_bilinearCompOf}
\leanok
\uses{def:bilinear-comp-of-family}
Let $b$ be a smooth field of bilinear forms on $T\tilde{M}$, let $p_0 \in M$, let $F : M \to \tilde{M}$ be smooth at $p_0$, and let $A_p : T_pM \to T_{F(p)}\tilde{M}$ be a family of linear maps whose representation in tangent coordinates is smooth at $p_0$. Then the section $p \mapsto (A^*b)_p$ is smooth at $p_0$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:bilinear-comp-of-family}
Smoothness is local, so read the section $p \mapsto (A^*b)_p$ in tangent coordinates around $p_0$ and around $F(p_0)$. It becomes the map
\[
  p \;\longmapsto\; B(F(p))\bigl(D(p)\,\cdot,\, D(p)\,\cdot\bigr),
\]
where $D(p)$ is $A_p$ read in tangent coordinates and $B$ is $b$ read in coordinates, both with values in fixed model spaces. Now $p \mapsto D(p)$ is smooth at $p_0$ by hypothesis, and $p \mapsto B(F(p))$ is smooth at $p_0$ as a composite of the smooth section $b$ with $F$. Finally the map $(\beta, A) \mapsto \beta(A\,\cdot, A\,\cdot)$, from pairs of a form on the model space of $T\tilde{M}$ and an operator into it, to forms on the model space of $TM$, is continuous and polynomial in its arguments --- linear in $\beta$ and quadratic in $A$ --- hence smooth. The coordinate representation is the composite of these, so it is smooth at $p_0$.
\end{proof}

The transporting family is left arbitrary because the two consumers need exactly that. Pulling back along $F$ is the case $A_p = dF_p$, immediately below; but the quotient metric of Theorem \ref{thm:group-action-riemannian-submersion} is the case where $A_p$ is the horizontal lift $L_{\sigma(p)}$ of a Riemannian submersion along a local section, and that family is \emph{not} the differential of any map.

\begin{definition}[The Pullback of a Field of Bilinear Forms]
\label{def:pullback-form-of-family}
\lean{LeeLib.Ch02.pullbackFormOf}
\leanok
\uses{def:bilinear-comp-of-family}
Let $F : M \to \tilde{M}$ be a map and let $b$ assign to each $y \in \tilde{M}$ a bilinear form $b_y$ on $T_y\tilde{M}$. The \emph{pullback of $b$ along $F$} is the transport of $b$ by the family of differentials $A_p = dF_p$ in the sense of Definition \ref{def:bilinear-comp-of-family}:
\[
  (F^*b)_p(v,w) \;=\; b_{F(p)}\bigl(dF_p v,\, dF_p w\bigr), \qquad v,w \in T_pM .
\]
Each $(F^*b)_p$ is indeed bilinear, being the composite of the bilinear $b_{F(p)}$ with the linear map $dF_p$ in each of its two arguments.
\end{definition}

Neither positivity, nor nondegeneracy, nor even symmetry of $b$ plays any part here, so both the Riemannian pullback of Definition \ref{def:induced-metric} and the indefinite one of Definition \ref{def:pseudo-pullback-form} below are instances of this definition, and both inherit its smoothness.

\begin{lemma}[The Pullback of a Smooth Field of Forms is Smooth]
\label{lem:pullback-form-of-family-smooth}
\lean{LeeLib.Ch02.contMDiff_pullbackFormOf}
\leanok
\uses{def:pullback-form-of-family,lem:bilinear-comp-of-family-smooth}
Let $F : M \to \tilde{M}$ be smooth and let $b$ be a smooth field of bilinear forms on $T\tilde{M}$, that is, a smooth section $y \mapsto b_y$ of the bundle of bilinear forms. Then $F^*b$ is a smooth field of bilinear forms on $TM$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pullback-form-of-family,lem:bilinear-comp-of-family-smooth}
This is the case $A_p = dF_p$ of Lemma \ref{lem:bilinear-comp-of-family-smooth}, applied at each $p_0 \in M$. Its hypothesis on $A$ holds: the differential of a smooth map, read in tangent coordinates, is smooth in the base point, which is the standard smoothness of $x \mapsto dF_x$ as a section of the bundle $\mathrm{Hom}(TM, F^*T\tilde{M})$.
\end{proof}

\begin{definition}[The Pullback of an Indefinite Metric]
\label{def:pseudo-pullback-form}
\lean{LeeLib.Ch02.pseudoPullbackForm}
\leanok
\uses{def:pullback-form-of-family,def:pseudo-riemannian-metric}
Let $(\tilde{M},\tilde{g})$ be a pseudo-Riemannian manifold and let $F : M \to \tilde{M}$ be a map. The \emph{pullback tensor} $F^*\tilde{g}$ is the pullback of the field $y \mapsto \tilde{g}_y$ along $F$ in the sense of Definition \ref{def:pullback-form-of-family}:
\[
  (F^*\tilde{g})_p(v,w) \;=\; \tilde{g}_{F(p)}\bigl(dF_p v,\, dF_p w\bigr), \qquad v,w \in T_pM .
\]
\end{definition}

This is Lee's $\iota^*\tilde{g}$. It is a field of bilinear forms for every $F$ whatever; whether it is a \emph{metric} is a separate question, and the one Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} answers.

\begin{lemma}[The Pullback Tensor is Symmetric]
\label{lem:pseudo-pullback-form-symm}
\lean{LeeLib.Ch02.pseudoPullbackForm_symm}
\leanok
\uses{def:pseudo-pullback-form,def:pseudo-riemannian-metric}
With $(\tilde{M},\tilde{g})$ and $F$ as in Definition \ref{def:pseudo-pullback-form}, $(F^*\tilde{g})_p(v,w) = (F^*\tilde{g})_p(w,v)$ for every $p \in M$ and $v,w \in T_pM$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-pullback-form,def:pseudo-riemannian-metric}
Unwinding Definition \ref{def:pseudo-pullback-form} twice and using symmetry of $\tilde{g}_{F(p)}$ in the middle step,
\[
  (F^*\tilde{g})_p(v,w) = \tilde{g}_{F(p)}(dF_pv, dF_pw) = \tilde{g}_{F(p)}(dF_pw, dF_pv) = (F^*\tilde{g})_p(w,v) .
\]
\end{proof}

\begin{lemma}[The Pullback Tensor is Smooth]
\label{lem:pseudo-pullback-form-smooth}
\lean{LeeLib.Ch02.pseudoPullbackForm_contMDiff}
\leanok
\uses{def:pseudo-pullback-form,lem:pullback-form-of-family-smooth,def:smooth-field-scalar-products}
With $(\tilde{M},\tilde{g})$ as in Definition \ref{def:pseudo-pullback-form} and $F : M \to \tilde{M}$ smooth, the field $p \mapsto (F^*\tilde{g})_p$ is smooth.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-pullback-form,lem:pullback-form-of-family-smooth,def:smooth-field-scalar-products}
By condition (c) of Definition \ref{def:smooth-field-scalar-products}, $y \mapsto \tilde{g}_y$ is a smooth section of the bundle of bilinear forms on $T\tilde{M}$, so Lemma \ref{lem:pullback-form-of-family-smooth} applies to it verbatim. Nondegeneracy of $\tilde{g}$ plays no part.
\end{proof}

\begin{definition}[The Metric Induced by an Indefinite Ambient Metric]
\label{def:pseudo-induced-metric}
\lean{LeeLib.Ch02.pseudoPullbackMetric}
\leanok
\uses{def:pseudo-pullback-form,lem:pseudo-pullback-form-symm,lem:pseudo-pullback-form-smooth,def:pseudo-riemannian-metric,def:smooth-field-scalar-products}
Let $(\tilde{M},\tilde{g})$ be a pseudo-Riemannian manifold and let $F : M \to \tilde{M}$ be smooth. Suppose the pullback tensor $F^*\tilde{g}$ is nondegenerate at every point, that is, for every $p \in M$ and every nonzero $v \in T_pM$ there is a $w \in T_pM$ with $(F^*\tilde{g})_p(v,w) \neq 0$. Then $F^*\tilde{g}$ is a pseudo-Riemannian metric on $M$, the \emph{metric induced by $F$}.
\end{definition}

\begin{proof}
\leanok
\uses{lem:pseudo-pullback-form-symm,lem:pseudo-pullback-form-smooth,def:pseudo-riemannian-metric,def:smooth-field-scalar-products}
The three conditions of Definition \ref{def:smooth-field-scalar-products} for the tangent bundle of $M$ are met in turn: symmetry at each point is Lemma \ref{lem:pseudo-pullback-form-symm}, nondegeneracy at each point is the hypothesis, in exactly the witness form (b) requires, and smoothness of the section is Lemma \ref{lem:pseudo-pullback-form-smooth}. So $F^*\tilde{g}$ is a smooth field of scalar products on $TM$, which by Definition \ref{def:pseudo-riemannian-metric} is a pseudo-Riemannian metric.
\end{proof}

Nondegeneracy is a genuine hypothesis here, and this is where the indefinite case parts company with the Riemannian one. By Lemma \ref{lem:pullback-immersion-metric}, positive definiteness of a pullback is \emph{equivalent} to $F$ being an immersion, so in the Riemannian case the immersion hypothesis alone builds the metric, as Definition \ref{def:induced-metric} does. Nondegeneracy of a pullback is not implied by immersion: a nondegenerate form can restrict to a degenerate one on a subspace, as the diagonal $\{\xi = \tau\}$ in $\mathbb{R}^{1,1}$ already shows. Deciding when it does not is the content of Proposition \ref{prop:pseudo-riemannian-hypersurface-signature}, and that is why the proposition exists at all.

\begin{lemma}[A Pseudo-Riemannian Pullback is an Induced Tensor]
\label{lem:pseudo-riemannian-submanifold-induced-tensor}
\lean{LeeLib.Ch02.isPullbackAlong_pseudoPullbackMetric}
\leanok
\uses{def:pseudo-induced-metric,def:pseudo-pullback-form,def:induced-tensor-along-map}
Let $(\tilde{M},\tilde{g})$, $F$ and the nondegeneracy hypothesis be as in Definition \ref{def:pseudo-induced-metric}. Then the induced metric $F^*\tilde{g}$ is induced by $\tilde{g}$ along $F$ in the sense of Definition \ref{def:induced-tensor-along-map}.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-induced-metric,def:pseudo-pullback-form,def:induced-tensor-along-map}
By Definition \ref{def:pseudo-induced-metric} the tensor $g = F^*\tilde{g}$ is a pseudo-Riemannian metric on $M$, so it is eligible. The condition of Definition \ref{def:induced-tensor-along-map} is $g_p(v,w) = \tilde{g}_{F(p)}(dF_pv, dF_pw)$ for all $p$, $v$ and $w$, and for $g = F^*\tilde{g}$ that is the defining formula of Definition \ref{def:pseudo-pullback-form}.
\end{proof}

This is the indefinite counterpart of Lemma \ref{lem:riemannian-submanifold-induced-tensor}, and it is what makes the hypothesis of Definition \ref{def:induced-tensor-along-map} --- assumed by Propositions \ref{prop:pseudo-riemannian-hypersurface-signature} and \ref{prop:pseudo-riemannian-adapted-frame} and by Corollary \ref{cor:normal-bundle-pseudo-riemannian-submanifold} alike --- available for a genuinely indefinite ambient metric. Lemma \ref{lem:riemannian-submanifold-induced-tensor} supplies that hypothesis only when $\tilde{g}$ is Riemannian, and in that case Proposition \ref{prop:pseudo-riemannian-hypersurface-signature} says no more than that a hypersurface of a Riemannian manifold is Riemannian.

\begin{lemma}[Extension of a Nondegenerate Family to a Nondegenerate Frame]
\label{lem:extend-nondegenerate-family}
\lean{LeeLib.Ch02.exists_isLocalNondegenerateOn_extend}
\leanok
\uses{def:local-nondegenerate-family,lem:completion-nondegenerate-bases,lem:nondegenerate-tuple-independent,lem:nondegenerate-tuple-basis,lem:sections-through-prescribed-basis,lem:open-nondegenerate-tuple}
Let $g$ be a smooth field of scalar products on a rank-$m$ vector bundle $V \to B$, let $u \subseteq B$ be open, let $x_0 \in u$, and let $X_1,\ldots,X_k$ be sections of $V$ that are nondegenerate on $u$. Then $k \leq m$, and there are an open set $v$ with $x_0 \in v \subseteq u$ and sections $Y_1,\ldots,Y_m$ of $V$, nondegenerate on $v$, with $Y_i = X_i$ on $v$ for every $i \leq k$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:local-nondegenerate-family,lem:completion-nondegenerate-bases,lem:nondegenerate-tuple-independent,lem:nondegenerate-tuple-basis,lem:sections-through-prescribed-basis,lem:open-nondegenerate-tuple}
The tuple $(X_1|_{x_0},\ldots,X_k|_{x_0})$ is nondegenerate in the $m$-dimensional scalar product space $(V_{x_0}, g_{x_0})$, hence linearly independent by Lemma \ref{lem:nondegenerate-tuple-independent}; so $k \leq m$. By Lemma \ref{lem:completion-nondegenerate-bases} it extends to a nondegenerate $m$-tuple $(w_1,\ldots,w_m)$ of $V_{x_0}$ with $w_i = X_i|_{x_0}$ for $i \leq k$, and by Lemma \ref{lem:nondegenerate-tuple-basis} that tuple is a basis of $V_{x_0}$.

Lemma \ref{lem:sections-through-prescribed-basis} applied to this basis yields an open $v_1 \ni x_0$ and sections $W_1,\ldots,W_m$, smooth on $v_1$, with $W_j|_{x_0} = w_j$. Define
\[
  Y_j \;=\; \begin{cases} X_j, & j \leq k,\\ W_j, & j > k,\end{cases}
\]
so that $Y_j$ is smooth on $u \cap v_1$ for every $j$, and $Y_j|_{x_0} = w_j$ for every $j$: for $j \leq k$ because $w_j = X_j|_{x_0}$, and for $j > k$ by the choice of $W_j$. Hence $(Y_1|_{x_0},\ldots,Y_m|_{x_0}) = (w_1,\ldots,w_m)$ is a nondegenerate $m$-tuple, and Lemma \ref{lem:open-nondegenerate-tuple} produces an open $v$ with $x_0 \in v \subseteq u \cap v_1$ on which $(Y_1,\ldots,Y_m)$ is nondegenerate. Finally $Y_i = X_i$ on $v$ for $i \leq k$ by construction.
\end{proof}

\begin{proposition}[Pseudo-Riemannian Adapted Orthonormal Frames]
\label{prop:pseudo-riemannian-adapted-frame}
\lean{LeeLib.Ch02.exists_adapted_pseudo_orthonormalFrame}
\leanok
\dcref{ch2:2.72}
\uses{def:pseudo-riemannian-metric,def:induced-tensor-along-map,lem:pullback-pseudo-metric-bundle,prop:orthonormal-frames-pseudo-riemannian,lem:extend-nondegenerate-family,prop:orthonormalize-nondegenerate-family,lem:orthonormal-tuple-nondegenerate,lem:pseudo-submanifold-immersion}
Suppose $(\bar{M}, \bar{g})$ is a pseudo-Riemannian manifold, and $F : M \to \bar{M}$ presents $M$ as a pseudo-Riemannian or Riemannian submanifold, the induced metric being $g$. For each $p \in M$, there exists a smooth orthonormal frame for $F^* T\bar{M}$ on a neighborhood of $p$ that is adapted to $M$, in the sense that its first $n = \dim M$ members span $dF_x(T_xM)$ at every point $x$ of that neighborhood.
\end{proposition}

\begin{proof}
\leanok
\uses{def:induced-tensor-along-map,lem:pullback-pseudo-metric-bundle,prop:orthonormal-frames-pseudo-riemannian,lem:extend-nondegenerate-family,prop:orthonormalize-nondegenerate-family,lem:orthonormal-tuple-nondegenerate,lem:pseudo-submanifold-immersion}
Write $m = \dim\bar{M}$ and $n = \dim M$, and equip $F^*T\bar{M}$ with the ambient metric $F^*\bar{g}$ of Lemma \ref{lem:pullback-pseudo-metric-bundle}. By Lemma \ref{lem:pseudo-submanifold-immersion}, $F$ is an immersion, so $dF_x$ is injective and $\dim dF_x(T_xM) = n$ for every $x$.

\emph{The tangential half.} By Proposition \ref{prop:orthonormal-frames-pseudo-riemannian} applied to $(M,g)$ there are an open $u \ni p$ and a smooth orthonormal frame $(E_1,\ldots,E_n)$ for $TM$ on $u$. Put $Z_i = dF(E_i)$, a smooth section of $F^*T\bar{M}$ on $u$. Because $g$ is induced by $\bar{g}$ along $F$,
\[
  (F^*\bar{g})_x(Z_i|_x, Z_j|_x) = \bar{g}_{F(x)}\bigl(dF_x(E_i|_x), dF_x(E_j|_x)\bigr) = g_x(E_i|_x, E_j|_x),
\]
so $(Z_1,\ldots,Z_n)$ is orthonormal at every $x \in u$, and hence nondegenerate there by Lemma \ref{lem:orthonormal-tuple-nondegenerate}. This is the step the Riemannian argument does not need and cannot be replaced by a coordinate frame: an arbitrary frame of $\bar{M}$ pushed to the fibres need not be nondegenerate.

\emph{Extension and orthonormalization.} By Lemma \ref{lem:extend-nondegenerate-family} there are an open $v$ with $p \in v \subseteq u$ and sections $(Y_1,\ldots,Y_m)$ of $F^*T\bar{M}$, nondegenerate on $v$, with $Y_i = Z_i$ on $v$ for $i \leq n$. Proposition \ref{prop:orthonormalize-nondegenerate-family} applied to $(Y_j)$ yields sections $(b_1,\ldots,b_m)$, smooth and orthonormal at every point of $v$, with $\operatorname{span}(b_1|_x,\ldots,b_k|_x) = \operatorname{span}(Y_1|_x,\ldots,Y_k|_x)$ for every $k \leq m$ and $x \in v$. Being orthonormal, $(b_1|_x,\ldots,b_m|_x)$ is linearly independent, and $m$ independent vectors in the $m$-dimensional fibre form a basis; so $(b_j)$ is a frame on $v$.

\emph{Adaptedness.} Fix $x \in v$ and $k \leq n$. Then
\[
  \operatorname{span}(b_1|_x,\ldots,b_k|_x) = \operatorname{span}(Y_1|_x,\ldots,Y_k|_x) = \operatorname{span}(Z_1|_x,\ldots,Z_k|_x) \subseteq dF_x(T_xM),
\]
the last inclusion because each $Z_i|_x = dF_x(E_i|_x)$ lies in the image of $dF_x$, which is a subspace. Taking $k = n$, the left-hand side is spanned by $n$ linearly independent vectors, so has dimension $n$; as it is contained in $dF_x(T_xM)$, which also has dimension $n$, the two are equal. Thus $(b_1,\ldots,b_n)$ spans $dF_x(T_xM)$ at every $x \in v$, which is the assertion that $(b_1,\ldots,b_m)$ is adapted to $M$.
\end{proof}

The next corollary is proved in the same way as Proposition 2.16.

\begin{definition}[The Normal Space of a Pseudo-Riemannian Submanifold]
\label{def:pseudo-normal-space}
\lean{LeeLib.Ch02.pseudoNormalSpace}
\leanok
\uses{def:pseudo-riemannian-metric,lem:pullback-pseudo-metric-bundle}
Let $(\bar{M},\bar{g})$ be a pseudo-Riemannian $m$-manifold and let $F : M \to \bar{M}$ present an $n$-manifold $M$ as a smooth immersion. For $x \in M$, the \emph{normal space to $M$ at $x$} is the subspace of the ambient tangent space $T_{F(x)}\bar{M}$ given by
\[
  N_x M \;=\; \bigl\{\, v \in T_{F(x)}\bar{M} \;:\; \bar{g}_{F(x)}(v,w) = 0 \text{ for every } w \in dF_x(T_xM) \,\bigr\}.
\]
\end{definition}

\begin{lemma}[Local Frames for the Normal Space]
\label{lem:pseudo-normal-space-frame}
\lean{LeeLib.Ch02.exists_pseudo_localFrame_normalSpace}
\leanok
\uses{def:pseudo-normal-space,prop:pseudo-riemannian-adapted-frame,lem:orthonormal-tuple-nondegenerate}
Let $(\bar{M},\bar{g})$ be a pseudo-Riemannian $m$-manifold and let $F : M \to \bar{M}$ present $M$ as an $n$-dimensional pseudo-Riemannian or Riemannian submanifold. Every $p \in M$ has a neighborhood $v$ carrying $m-n$ sections of $F^* T\bar{M}$, smooth on $v$, whose values at each $x \in v$ are orthonormal and span $N_x M$.
\end{lemma}

\begin{proof}
\leanok
\uses{def:pseudo-normal-space,prop:pseudo-riemannian-adapted-frame,lem:orthonormal-tuple-nondegenerate}
Let $(b_1,\ldots,b_m)$ be an adapted orthonormal frame on a neighborhood $v$ of $p$, as furnished by Proposition \ref{prop:pseudo-riemannian-adapted-frame}, and take $N_j = b_{n+j}$ for $j = 1,\ldots,m-n$. These are smooth on $v$, and orthonormal at each $x \in v$, both being inherited from $(b_i)$.

\emph{They are normal.} Fix $x \in v$ and $j$. For every $i \leq n$ we have $\langle N_j|_x, b_i|_x \rangle = 0$, the indices $n+j$ and $i$ being distinct; so $N_j|_x$ is orthogonal to each member of $(b_1|_x,\ldots,b_n|_x)$, and hence to their span, which is $dF_x(T_xM)$ by adaptedness. Thus $N_j|_x \in N_xM$.

\emph{They span.} The previous paragraph gives $\operatorname{span}(N_1|_x,\ldots,N_{m-n}|_x) \subseteq N_xM$. For the reverse, let $w \in N_xM$. Being orthonormal, $(b_1|_x,\ldots,b_m|_x)$ is linearly independent and so a basis of the $m$-dimensional fibre; write $w = \sum_a c_a b_a|_x$. Pairing with $b_a|_x$ and using orthogonality of the distinct members,
\[
  \langle b_a|_x, w\rangle = c_a \langle b_a|_x, b_a|_x\rangle .
\]
For $a \leq n$ the vector $b_a|_x$ lies in $dF_x(T_xM)$, so $\langle b_a|_x, w \rangle = 0$ because $w$ is normal; and $\langle b_a|_x,b_a|_x\rangle = \pm 1 \neq 0$ by orthonormality, whence $c_a = 0$. (It is here that the indefinite case needs orthonormality rather than mere independence: a null $b_a$ would make the displayed equation vacuous and leave $c_a$ undetermined.) So $w = \sum_{a > n} c_a b_a|_x$ lies in $\operatorname{span}(N_1|_x,\ldots,N_{m-n}|_x)$.
\end{proof}

Lemma \ref{lem:pseudo-normal-space-frame} is the hypothesis of Lemma \ref{lem:local-frame-criterion-for-subbundles}, but that lemma additionally requires the ambient bundle to carry a Riemannian fibre metric, which it uses only to normalize the given frames. An indefinite $\bar{g}$ is not such a metric, and Proposition \ref{prop:normal-bundle}'s move of supplying the ambient metric itself is therefore unavailable. The remedy is that the fibre metric of Lemma \ref{lem:local-frame-criterion-for-subbundles} is required to bear no relation to the family of subspaces: an arbitrary Riemannian metric on $\bar{M}$ serves, and one exists by Proposition \ref{prop:existence-riemannian-metric}.

\begin{corollary}[Normal Bundle to a Pseudo-Riemannian Submanifold]
\label{cor:normal-bundle-pseudo-riemannian-submanifold}
\lean{LeeLib.Ch02.contMDiffVectorBundle_pseudoNormalSpace}
\leanok
\dcref{ch2:2.73}
\uses{def:pseudo-riemannian-metric,def:pseudo-normal-space,prop:pseudo-riemannian-adapted-frame,lem:pseudo-normal-space-frame,lem:local-frame-criterion-for-subbundles,prop:existence-riemannian-metric}
Suppose $(\bar{M}, \bar{g})$ is a pseudo-Riemannian $m$-manifold, and $F : M \to \bar{M}$ presents $M$ as an $n$-dimensional embedded or immersed pseudo-Riemannian or Riemannian submanifold. The set of vectors normal to $M$ is a smooth rank-$(m-n)$ vector subbundle of $T\bar{M}|_M$, called the \emph{normal bundle} to $M$.
\end{corollary}

\begin{proof}
\leanok
\uses{def:pseudo-normal-space,lem:pseudo-normal-space-frame,lem:local-frame-criterion-for-subbundles,prop:existence-riemannian-metric}
By Proposition \ref{prop:existence-riemannian-metric} choose a Riemannian metric $h$ on $\bar{M}$; by Lemma \ref{lem:ambient-tangent-bundle-riemannian} it makes $F^*T\bar{M}$ a Riemannian vector bundle over $M$. Lemma \ref{lem:pseudo-normal-space-frame} provides, near every point of $M$, sections of $F^*T\bar{M}$ smooth there and forming a basis of $N_xM$ at each point, orthonormality for $\bar{g}$ giving in particular linear independence. This is exactly the hypothesis of Lemma \ref{lem:local-frame-criterion-for-subbundles} applied to the bundle $F^*T\bar{M}$ with fibre metric $h$ and the family of subspaces $D_x = N_xM$, which therefore exhibits $NM$ as a smooth rank-$(m-n)$ subbundle of $F^*T\bar{M}$. The metric $h$ enters only through Lemma \ref{lem:local-frame-criterion-for-subbundles}'s normalization of frames and is unrelated to $\bar{g}$; the subspaces $N_xM$ and the frames spanning them are defined by $\bar{g}$ alone.
\end{proof}

\section{Other Generalizations of Riemannian Metrics}

Pseudo-Riemannian metrics are obtained by relaxing the positivity requirement for Riemannian metrics. In addition, there are other useful generalizations that result when we relax other requirements. In this section we touch briefly on two of those generalizations. We will not treat these anywhere else in the book, but it is useful to know the definitions.

\subsection{Sub-Riemannian Metrics}

The first generalization arises when we relax the requirement that the metric be defined on the whole tangent space.

A \emph{sub-Riemannian metric} (also sometimes known as a \emph{singular Riemannian metric} or \emph{Carnot--Carathéodory metric}) on a smooth manifold $M$ is a smooth fiber metric on a smooth tangent distribution $S \subseteq TM$ (i.e., a vector subbundle of $TM$). Since lengths make sense only for vectors in $S$, the only curves whose lengths can be measured are those whose velocity vectors lie everywhere in $S$. Therefore, one usually imposes some condition on $S$ that guarantees that any two nearby points can be connected by such a curve. This is, in a sense, the opposite of the Frobenius integrability condition, which would restrict every such curve to lie in a single leaf of a foliation.

Sub-Riemannian metrics arise naturally in the study of the abstract models of real submanifolds of complex space $\mathbb{C}^n$, called \emph{CR manifolds} (short for \emph{Cauchy--Riemann manifolds}). CR manifolds are real manifolds endowed with a tangent distribution $S \subseteq TM$ whose fibers carry the structure of complex vector spaces (with an additional integrability condition that need not concern us here). In the model case of a submanifold $M \subseteq \mathbb{C}^n$, $S$ is the set of vectors tangent to $M$ that remain tangent after multiplication by $i = \sqrt{-1}$ in the ambient complex coordinates. If $S$ is sufficiently far from being integrable, choosing a fiber metric on $S$ results in a sub-Riemannian metric whose geometric properties closely reflect the complex-analytic properties of $M$ as a subset of $\mathbb{C}^n$.

Another motivation for studying sub-Riemannian metrics arises from \emph{control theory}. In this subject, one is given a manifold with a vector field depending on parameters called \emph{controls}, with the goal being to vary the controls so as to obtain a solution curve with desired properties, often one that minimizes some function such as arc length. If the vector field is constrained to be everywhere tangent to a distribution $S$ on the manifold (for example, in the case of a robot arm whose motion is restricted by the orientations of its hinges), then the function can often be modeled as a sub-Riemannian metric and optimal solutions modeled as sub-Riemannian geodesics.

A useful introduction to the geometry of sub-Riemannian metrics is provided in the article [Str86].

\subsection{Finsler Metrics}

Another important generalization arises from relaxing the requirement that norms of vectors be defined in terms of an inner product on each tangent space.

In general, a \emph{norm} on a vector space $V$ is a real-valued function on $V$, usually written $v \mapsto |v|$, that satisfies
\begin{enumerate}
\item \textsc{Homogeneity}: $|cv| = |c||v|$ for $v \in V$ and $c \in \mathbb{R}$;
\item \textsc{Positivity}: $|v| \geq 0$ for $v \in V$, with equality if and only if $v = 0$;
\item \textsc{Triangle Inequality}: $|v + w| \leq |v| + |w|$ for $v, w \in V$.
\end{enumerate}

For example, the length function associated with an inner product is a norm. Now suppose $M$ is a smooth manifold. A \emph{Finsler metric} on $M$ is a continuous function $F: TM \to \mathbb{R}$, smooth on the set of nonzero vectors, whose restriction to each tangent space $T_pM$ is a norm. Again, the norm function associated with a Riemannian metric is a special case.

The inventor of Riemannian geometry himself, Bernhard Riemann, clearly envisioned an important role in $n$-dimensional geometry for what we now call Finsler metrics; he restricted his investigations to the ``Riemannian'' case purely for simplicity (see [Spi79, Vol. 2]). However, it was not until the late twentieth century that Finsler metrics began to be studied seriously from a geometric point of view.

The recent upsurge of interest in Finsler metrics has been motivated in part by the fact that two different Finsler metrics appear very naturally in the theory of several complex variables. For certain bounded open sets in $\mathbb{C}^n$ (the ones with smooth, strictly convex boundaries, for example), the \emph{Kobayashi metric} and the \emph{Carathéodory metric} are intrinsically defined, biholomorphically invariant Finsler metrics. Combining differential-geometric and complex-analytic methods has led to striking new insights into both the function theory and the geometry of such domains. We do not treat Finsler metrics further in this book, but you can consult one of the recent books on the subject [AP94, BCS00, JP13].

\section{Problems}

\begin{exercise}\label{prob:riemannian-1-manifold-flat}\dcref{ch2:2-1}
Show that every Riemannian $1$-manifold is flat. (Used on pp. 13, 193.)
\end{exercise}

\begin{exercise}\label{prob:distance-preserving-map-linear-isometry}\dcref{ch2:2-2}
Suppose $V$ and $W$ are finite-dimensional real inner product spaces of the same dimension, and $F: V \to W$ is any map (not assumed to be linear or even continuous) that preserves the origin and all distances: $F(0) = 0$ and $|F(x) - F(y)| = |x - y|$ for all $x, y \in V$. Prove that $F$ is a linear isometry. [Hint: First show that $F$ preserves inner products, and then show that it is linear.] (Used on p. 187.)
\end{exercise}

\begin{exercise}\label{prob:surface-of-revolution-warped-product}\dcref{ch2:2-3}
\uses{ex:warped-products}
Given a smooth embedded $1$-dimensional submanifold $C \subseteq H$ as in Example 2.24(b), show that the surface of revolution $S_C \subseteq \mathbb{R}^3$ with its induced metric is isometric to the warped product $C \times_a S^1$, where $a: C \to \mathbb{R}$ is the distance to the $z$-axis.
\end{exercise}

\begin{definition}[The open submanifold metric]
\label{def:open-submanifold-metric}
\lean{LeeLib.Ch02.openSubmanifoldMetric}
\leanok
\uses{def:riemannian-metric,def:induced-metric,lem:mfderiv-open-inclusion}
Let $W \subseteq \mathbb{R}^n$ be open and let $g$ be a Riemannian metric on $\mathbb{R}^n$. The metric \emph{induced on $W$} is $\iota^* g$, where $\iota : W \hookrightarrow \mathbb{R}^n$ is the inclusion. By Lemma \ref{lem:mfderiv-open-inclusion} the inclusion is an immersion, so Lemma \ref{lem:pullback-immersion-metric} applies and $\iota^* g$ is a Riemannian metric.
\end{definition}

\begin{lemma}[The differential of the inclusion of an open subset]
\label{lem:mfderiv-open-inclusion}
\lean{LeeLib.Ch02.mfderiv_opens_subtypeVal}
\leanok
An open set $W \subseteq \mathbb{R}^n$ is a smooth manifold whose single chart is the inclusion $\iota : W \hookrightarrow \mathbb{R}^n$, and every tangent space $T_x W$ is canonically $\mathbb{R}^n$. Under these identifications $d\iota_x = \operatorname{id}$ for every $x \in W$. In particular $\iota$ is an immersion.
\end{lemma}

\begin{proof}
\leanok
The differential is computed from the coordinate representation of $\iota$, which is $\iota \circ \iota^{-1}$ read through the chart. Since the chart of $W$ is the inclusion itself, this composite is the identity map of $\mathbb{R}^n$ \emph{restricted to $W$} — it is undefined outside $W$, so the two maps are not equal globally. As $W$ is open it is a neighborhood of $x$, and a derivative depends only on the germ at the point, so $d\iota_x$ is the derivative of the identity, namely $\operatorname{id}$.
\end{proof}

\begin{lemma}[Tangent directions to the sphere are orthogonal to the radius]
\label{lem:sphere-tangent-orthogonal-radius}
\lean{LeeLib.Ch02.inner_coe_sphere_mfderiv_eq_zero}
\leanok
\uses{ex:spheres}
Let $\iota : S^{n-1} \hookrightarrow \mathbb{R}^n$ be the inclusion. For every $\omega \in S^{n-1}$ and every $w \in T_\omega S^{n-1}$,
\[
\langle \omega, d\iota_\omega(w)\rangle = 0.
\]
\end{lemma}

\begin{proof}
\leanok
The image of $d\iota_\omega$ is the orthogonal complement $(\mathbb{R}\omega)^\perp$ of the line spanned by $\omega$. Since $\omega$ lies in $\mathbb{R}\omega$, every element of the image is orthogonal to $\omega$.
\end{proof}

\begin{definition}[The polar map]
\label{def:polar-map}
\lean{LeeLib.Ch02.polarMap}
\leanok
\uses{def:warped-product,ex:spheres,def:open-submanifold-metric}
$\Phi : \mathbb{R}^+ \times S^{n-1} \to \mathbb{R}^n$ is defined by $\Phi(\rho, \omega) = \rho\,\omega$.
\end{definition}

\begin{lemma}[The differential of the polar map]
\label{lem:mfderiv-polar-map}
\lean{LeeLib.Ch02.mfderiv_polarMap_apply}
\leanok
\uses{def:polar-map,lem:mfderiv-open-inclusion}
For $(r, \omega) \in \mathbb{R}^+ \times S^{n-1}$ and $(a, w) \in T_r\mathbb{R}^+ \oplus T_\omega S^{n-1}$,
\[
d\Phi_{(r,\omega)}(a, w) = a\,\omega + r\,d\iota_\omega(w).
\]
\end{lemma}

\begin{proof}
\leanok
$\Phi$ is a map on a product manifold, so its differential is the sum of its two partial differentials. Holding $\omega$ fixed, $r \mapsto r\,\omega$ is linear in $r$ precomposed with the inclusion $\mathbb{R}^+ \hookrightarrow \mathbb{R}$, whose differential is the identity by Lemma \ref{lem:mfderiv-open-inclusion}; its differential therefore sends $a$ to $a\,\omega$. Holding $r$ fixed, $\omega \mapsto r\,\omega$ is the inclusion $\iota$ followed by the fixed scaling $u \mapsto r\,u$, which is linear, so its differential sends $w$ to $r\,d\iota_\omega(w)$.
\end{proof}

\begin{exercise}\label{prob:warped-product-isometry}
\dcref{ch2:2-4}
\lean{LeeLib.Ch02.isMetricPreserving_polarMap}
\leanok
\uses{ex:warped-products,def:polar-map,lem:mfderiv-polar-map,lem:sphere-tangent-orthogonal-radius,def:metric-preserving,def:open-submanifold-metric,prop:warped-product-constant-one}
Let $\rho: \mathbb{R}^+ \to \mathbb{R}$ be the restriction of the standard coordinate function, and let $\mathbb{R}^+ \times_\rho S^{n-1}$ denote the resulting warped product (see Example 2.24(c)). Define $\Phi: \mathbb{R}^+ \times_\rho S^{n-1} \to \mathbb{R}^n \setminus \{0\}$ by $\Phi(\rho,\omega) = \rho\omega$. Show that $\Phi$ is an isometry between the warped product metric and the Euclidean metric on $\mathbb{R}^n \setminus \{0\}$. (Used on p.~293.)
\end{exercise}

\begin{proof}
\leanok
\uses{def:polar-map,lem:mfderiv-polar-map,lem:sphere-tangent-orthogonal-radius,def:metric-preserving}
We verify $\Phi^*\bar{g} = g_{\mathbb{R}^+} \oplus \rho^2 \mathring{g}$ at each point; that $\Phi$ is a bijection onto $\mathbb{R}^n \setminus \{0\}$ is checked separately below.

Fix $(r,\omega)$ and let $(a,w), (a',w')$ be tangent vectors there. By Lemma \ref{lem:mfderiv-polar-map},
\[
\langle d\Phi(a,w), d\Phi(a',w')\rangle
 = \langle a\omega + r\,d\iota(w),\ a'\omega + r\,d\iota(w')\rangle,
\]
which expands into four terms:
\[
a a' \langle\omega,\omega\rangle + a r \langle \omega, d\iota(w')\rangle + a' r \langle d\iota(w), \omega\rangle + r^2 \langle d\iota(w), d\iota(w')\rangle.
\]
Now $\langle\omega,\omega\rangle = |\omega|^2 = 1$ because $\omega \in S^{n-1}$, and the two middle terms vanish by Lemma \ref{lem:sphere-tangent-orthogonal-radius}. Since $\mathring{g}_\omega(w,w') = \langle d\iota(w), d\iota(w')\rangle$ by the definition of the round metric as $\iota^*\bar{g}$, what remains is
\[
a a' + r^2\,\mathring{g}_\omega(w,w'),
\]
which is exactly the warped product metric $g_{\mathbb{R}^+} \oplus \rho^2\mathring{g}$ evaluated at $(r,\omega)$, the radial factor $g_{\mathbb{R}^+}$ being $\iota^*\bar{g} = d\rho^2$ by Lemma \ref{lem:mfderiv-open-inclusion} and the warping function being $\rho(r) = r$.

For the bijection: $|\Phi(r,\omega)| = r|\omega| = r > 0$, so $\Phi$ never takes the value $0$; if $\Phi(r,\omega) = \Phi(r',\omega')$ then taking norms gives $r = r'$ and then cancelling the nonzero scalar $r$ gives $\omega = \omega'$, so $\Phi$ is injective; and any $x \neq 0$ equals $\Phi(|x|, x/|x|)$, so $\Phi$ is onto $\mathbb{R}^n \setminus \{0\}$.
\end{proof}

\begin{exercise}\label{prob:horizontal-vector-fields}
\dcref{ch2:2-5}
\lean{LeeLib.Ch02.exists_horizontalLift_eq}
\leanok
\uses{prop:horizontal-lift-of-vector-field,prop:horizontal-lift-realizes-horizontal-vector}
Prove parts (b) and (c) of Proposition 2.25 (properties of horizontal vector fields). (Used on p.~146.)
\end{exercise}

\begin{exercise}\label{prob:riemannian-metric-submersion}
\lean{LeeLib.Ch02.existsUnique_isRiemannianSubmersion_metric}
\leanok
\dcref{ch2:2-6}
\uses{thm:group-action-riemannian-submersion}
Prove Theorem 2.28 (if $\pi: \tilde{M} \to M$ is a surjective smooth submersion, and a group acts on $\tilde{M}$ isometrically, vertically, and transitively on fibers, then $M$ inherits a unique Riemannian metric such that $\pi$ is a Riemannian submersion).
\end{exercise}

\begin{exercise}\label{prob:grassmann-stiefel-manifolds}
\dcref{ch2:2-7}
For $0 \le k \le n$, the set $G_k(\mathbb{R}^n)$ of $k$-dimensional linear subspaces of $\mathbb{R}^n$ is called a \textbf{Grassmann manifold} or \textbf{Grassmannian}. The group $GL(n,\mathbb{R})$ acts transitively on $G_k(\mathbb{R}^n)$ in an obvious way, and $G_k(\mathbb{R}^n)$ has a unique smooth manifold structure making this action smooth (see [LeeSM, Example 21.21]).

(a) Let $V_k(\mathbb{R}^n)$ denote the set of orthonormal ordered $k$-tuples of vectors in $\mathbb{R}^n$. By arranging the vectors in $k$ columns, we can view $V_k(\mathbb{R}^n)$ as a subset of the vector space $M(n \times k, \mathbb{R})$ of all $n \times k$ real matrices. Prove that $V_k(\mathbb{R}^n)$ is a smooth submanifold of $M(n \times k, \mathbb{R})$ of dimension $k(2n - k - 1)/2$, called a \textbf{Stiefel manifold}. [Hint: Consider the map $\Phi: M(n \times k, \mathbb{R}) \to M(k \times k, \mathbb{R})$ given by $\Phi(A) = A^T A$.]

(b) Show that the map $\pi: V_k(\mathbb{R}^n) \to G_k(\mathbb{R}^n)$ that sends a $k$-tuple to its span is a surjective smooth submersion.

(c) Give $V_k(\mathbb{R}^n)$ the Riemannian metric induced from the Euclidean metric on $M(n \times k, \mathbb{R})$. Show that the right action of $O(k)$ on $V_k(\mathbb{R}^n)$ by matrix multiplication on the right is isometric, vertical, and transitive on fibers of $\pi$, and thus there is a unique metric on $G_k(\mathbb{R}^n)$ such that $\pi$ is a Riemannian submersion. [Hint: It might help to note that the Euclidean inner product on $M(n \times k, \mathbb{R})$ can be written in the form $(A,B) = \operatorname{tr}(A^T B)$.]
\end{exercise}

\begin{exercise}\label{prob:mobius-band-metric}
\dcref{ch2:2-8}
\uses{ex:open-mobius-band}
Prove that the action of $\mathbb{Z}$ on $\mathbb{R}^2$ defined in Example 2.35 is smooth, free, proper, and isometric, and therefore the open Möbius band inherits a flat Riemannian metric such that the quotient map is a Riemannian covering.
\end{exercise}

\begin{exercise}\label{prob:gradient-orthogonal-level-sets}
\lean{LeeLib.Ch02.innerAt_grad_mfderiv_regularLevelSetIncl_eq_zero}
\leanok
\dcref{ch2:2-9}
\uses{prop:regular-level-set-hypersurface,prop:gradient-normal-regular-part,lem:gradient-nonvanishing-regular-part}
Prove Proposition 2.37 (the gradient is orthogonal to regular level sets).
\end{exercise}

\begin{exercise}\label{prob:gradient-characterization}
\dcref{ch2:2-10}
\lean{LeeLib.Ch02.eq_grad_iff_of_forall_ne_zero}
\leanok
\uses{def:gradient,def:regular-set,lem:gradient-characterization,lem:transversal-representative-eq-gradient}
Suppose $(M,g)$ is a Riemannian manifold, $f \in C^\infty(M)$, and $X \in \mathfrak{X}(M)$ is a nowhere-vanishing vector field. Prove that $X = \operatorname{grad} f$ if and only if $Xf = |X|_g^2$ and $X$ is orthogonal to the level sets of $f$ at all regular points of $f$. (Used on pp.~161, 180.)
\end{exercise}

\begin{proof}
\leanok
Throughout, the tangent space to the level set of $f$ through a regular point $p$ is $\ker df_p$, so ``$X$ is orthogonal to the level sets at regular points'' means $\langle X_p, w\rangle_g = 0$ for every regular $p$ and every $w \in \ker df_p$; and $Xf = df(X)$, $|X|_g^2 = \langle X, X\rangle_g$.

Suppose first that $X = \operatorname{grad} f$. Then $Xf = df(X) = \langle \operatorname{grad} f, X\rangle_g = \langle X, X\rangle_g = |X|_g^2$ by Lemma \ref{lem:gradient-characterization}, and for $w \in \ker df_p$ the same lemma gives $\langle X_p, w\rangle_g = df_p(w) = 0$. Neither implication used the hypothesis that $X$ vanishes nowhere.

Conversely, suppose $Xf = |X|_g^2$ and $X$ is orthogonal to the level sets at regular points. Fix $p \in M$. Since $X_p \neq 0$ and $g$ is positive definite, $\langle X_p, X_p\rangle_g > 0$, so
\[
df_p(X_p) = \langle X_p, X_p \rangle_g > 0 .
\]
In particular $df_p \neq 0$, so \emph{every} point of $M$ is a regular point of $f$ and the orthogonality hypothesis is available at $p$. The claim $X_p = \operatorname{grad} f|_p$ is then Lemma \ref{lem:transversal-representative-eq-gradient}.
\end{proof}

\begin{lemma}[A transversal orthogonal representative is the gradient]
\label{lem:transversal-representative-eq-gradient}
\lean{LeeLib.Ch02.eq_grad_of_ne_zero_of_innerAt_self_of_orthogonal_ker}
\leanok
\uses{def:gradient,lem:gradient-characterization}
Let $(M,g)$ be a Riemannian manifold, $f \in C^\infty(M)$, $p \in M$, and $v \in T_pM$ with $v \neq 0$. Suppose $\langle v, v\rangle_g = df_p(v)$ and $\langle v, w\rangle_g = 0$ for every $w \in \ker df_p$. Then $v = \operatorname{grad} f|_p$.
\end{lemma}

\begin{proof}
\leanok
Since $v \neq 0$ and $g$ is positive definite, $df_p(v) = \langle v,v\rangle_g > 0$; in particular $df_p(v) \neq 0$, so $v \notin \ker df_p$.

Let $w \in T_pM$ be arbitrary and set $t = df_p(w)/df_p(v)$, which is defined because $df_p(v) \neq 0$. Then
\[
df_p(w - t\,v) = df_p(w) - t\,df_p(v) = df_p(w) - df_p(w) = 0,
\]
so $w - t\,v \in \ker df_p$ and the orthogonality hypothesis gives $\langle v, w - t\,v\rangle_g = 0$. Expanding by bilinearity,
\[
\langle v, w \rangle_g \;=\; t \langle v, v\rangle_g \;=\; t\, df_p(v) \;=\; df_p(w).
\]
As $w$ was arbitrary, $v$ represents $df_p$ under $g$, and Lemma \ref{lem:gradient-characterization} identifies the unique such vector as $\operatorname{grad} f|_p$. Hence $v = \operatorname{grad} f|_p$.

Note that no dimension count is used: the decomposition $w = (w - t\,v) + t\,v$ is explicit, so the argument does not need $\ker df_p$ to be a hyperplane.
\end{proof}

\begin{exercise}\label{prob:inner-products-tensor-bundles}
\dcref{ch2:2-11}
\uses{prop:inner-products-of-tensors}
Prove Proposition 2.40 (inner products on tensor bundles).
\end{exercise}

\begin{exercise}
\label{prob:riemannian-volume-form-existence}
\lean{LeeLib.Ch02.RiemannianMetric.riemannian_volumeForm}
\leanok
\dcref{ch2:2-12}
\uses{prop:riemannian-volume-form}
Prove Proposition 2.41 (existence and uniqueness of the Riemannian volume form).
\end{exercise}

\begin{exercise}
\label{prob:hypersurface-volume-form}
\lean{LeeLib.Ch02.exists_isSmoothOrientation_iff_exists_isSmoothNormalField}
\leanok
\dcref{ch2:2-13}
\uses{prop:hypersurface-volume-form,prop:hypersurface-volume-form-formula,prop:hypersurface-volume-form-if,prop:hypersurface-volume-form-only-if}
Prove Proposition 2.43 (characterizing the volume form of a Riemannian hypersurface). [Hint: To prove (2.17), use an adapted orthonormal frame.]
\end{exercise}

\begin{exercise}
\label{prob:covering-volume-form}
\dcref{ch2:2-14}
Suppose $(\tilde{M}, \tilde{g})$ and $(M, g)$ are compact connected Riemannian manifolds, and $\pi : \tilde{M} \to M$ is a $k$-sheeted Riemannian covering. Prove that $\text{Vol}(\tilde{M}) = k \cdot \text{Vol}(M)$. (Used on p. 363.)
\end{exercise}

\begin{exercise}
\label{prob:warped-product-volume-form}
\dcref{ch2:2-15}
Suppose $(M_1, g_1)$ and $(M_2, g_2)$ are oriented Riemannian manifolds of dimensions $k_1$ and $k_2$, respectively. Let $f : M_1 \to \mathbb{R}^+$ be a smooth function, and let $g = g_1 \oplus f^2 g_2$ be the corresponding warped product metric on $M_1 \times_f M_2$. Prove that the Riemannian volume form of $g$ is given by
\[
dV_g = f^{k_2} dV_{g_1} \wedge dV_{g_2},
\]
where $f$, $dV_{g_1}$, and $dV_{g_2}$ are understood to be pulled back to $M_1 \times M_2$ by the projection maps. (Used on p. 295.)
\end{exercise}

The fiber metric on $\Lambda^k T^* M$ requested in Problem 2-16 below is built pointwise first. The following sequence carries out the linear algebra on a single inner product space $V$; the fibers of $\Lambda^k T^*M$ are the spaces of alternating $k$-tensors on the tangent spaces $T_pM$, each of which carries the inner product $g_p$.

Throughout, $V$ is a finite-dimensional real inner product space, $k \ge 1$, and $\Lambda^k(V^*)$ denotes the space of alternating $k$-tensors on $V$. For a covector $a \in V^*$ we write $a^\sharp$ for its Riesz representative, the unique vector with $\langle a^\sharp, \cdot\rangle = a$, and we equip $V^*$ with the inner product
\begin{equation}
\langle a, b \rangle = \langle a^\sharp, b^\sharp \rangle, \tag{2.26a}
\end{equation}
which is the coordinate-free form of Lee's $g^{ij}a_i b_j$.

\begin{definition}
\label{def:inner-k-covectors}
\lean{LeeLib.Ch02.innerForms}
\leanok
Let $(e_m)_{m \in \iota'}$ be an orthonormal basis of $V$. For $\omega, \eta \in \Lambda^k(V^*)$ set
\[
\langle \omega, \eta \rangle_e = \frac{1}{k!} \sum_{s \colon \{1,\ldots,k\} \to \iota'} \omega\big(e_{s(1)}, \ldots, e_{s(k)}\big)\, \eta\big(e_{s(1)}, \ldots, e_{s(k)}\big),
\]
the sum running over all maps $s$.
\end{definition}

The sum is over all index maps rather than over increasing multi-indices $i_1 < \cdots < i_k$. The two prescriptions agree: a map $s$ that repeats a value contributes $0$ because $\omega$ and $\eta$ are alternating, and each increasing multi-index is counted once for each of the $k!$ maps with the same image, all contributing equally since the two sign changes cancel. Written this way, symmetry, bilinearity and positive definiteness are visible directly from the formula.

\begin{lemma}
\label{lem:inner-k-covectors-posdef}
\lean{LeeLib.Ch02.innerForms_self_pos}
\leanok
\uses{def:inner-k-covectors}
$\langle \cdot, \cdot \rangle_e$ is a symmetric bilinear form on $\Lambda^k(V^*)$, and $\langle \omega, \omega \rangle_e > 0$ for every $\omega \ne 0$. Thus it is an inner product.
\end{lemma}
\begin{proof}
\leanok
Symmetry and bilinearity are immediate from the formula, each summand being a product of two values that are linear in $\omega$ and in $\eta$ respectively. For positivity, $\langle \omega, \omega \rangle_e = \frac{1}{k!}\sum_s \omega(e_{s(1)},\ldots,e_{s(k)})^2 \ge 0$. If $\omega \ne 0$ then, since an alternating map is determined by its values on tuples of basis vectors, some $\omega(e_{s(1)},\ldots,e_{s(k)})$ is nonzero, so at least one summand is positive.
\end{proof}

\begin{lemma}
\label{lem:inner-k-covectors-wedge}
\lean{LeeLib.Ch02.innerForms_wedgeCovectors}
\leanok
\uses{def:inner-k-covectors,lem:alternating-cauchy-binet}
For all covectors $\omega^1, \ldots, \omega^k, \eta^1, \ldots, \eta^k \in V^*$,
\[
\langle \omega^1 \wedge \cdots \wedge \omega^k, \eta^1 \wedge \cdots \wedge \eta^k \rangle_e = \det\big(\langle \omega^i, \eta^j \rangle\big).
\]
\end{lemma}
\begin{proof}
\leanok
\uses{lem:alternating-cauchy-binet}
Let $A_{im} = \omega^i(e_m)$. Apply Lemma~\ref{lem:alternating-cauchy-binet} with $\theta = \eta^1 \wedge \cdots \wedge \eta^k$, the family $(e_m)$, and the matrix $A$. By the determinant convention for the wedge product, $\det(A[\,\cdot\,, s]) = (\omega^1 \wedge \cdots \wedge \omega^k)(e_{s(1)},\ldots,e_{s(k)})$, so the left-hand side of that lemma is $k!$ times the sum defining $\langle \omega^1 \wedge \cdots \wedge \omega^k, \eta^1 \wedge \cdots \wedge \eta^k \rangle_e$. On the right-hand side, $\sum_m \omega^i(e_m)e_m = (\omega^i)^\sharp$ by the expansion of a vector in an orthonormal basis, so it is
\[
k!\,(\eta^1 \wedge \cdots \wedge \eta^k)\big((\omega^1)^\sharp, \ldots, (\omega^k)^\sharp\big) = k!\,\det\big(\eta^i((\omega^j)^\sharp)\big) = k!\,\det\big(\langle \eta^i, \omega^j\rangle\big),
\]
using (2.26a). Cancelling $k!$ and transposing the matrix gives the claim.
\end{proof}

\begin{lemma}
\label{lem:k-covector-wedge-expansion}
\lean{LeeLib.Ch02.factorial_smul_eq_sum_wedgeCovectors}
\leanok
\uses{lem:alternating-cauchy-binet}
For every $\omega \in \Lambda^k(V^*)$,
\[
k!\,\omega = \sum_{u \colon \{1,\ldots,k\} \to \iota'} \omega\big(e_{u(1)}, \ldots, e_{u(k)}\big)\; e_{u(1)}^\flat \wedge \cdots \wedge e_{u(k)}^\flat,
\]
where $v^\flat = \langle v, \cdot \rangle$.
\end{lemma}
\begin{proof}
\leanok
\uses{lem:alternating-cauchy-binet}
Both sides are alternating $k$-tensors, so it suffices to check that they agree on every tuple $(e_{v(1)}, \ldots, e_{v(k)})$ of basis vectors. Apply Lemma~\ref{lem:alternating-cauchy-binet} with $\theta = \omega$, the family $(e_m)$, and the matrix $A_{im} = \langle e_{v(i)}, e_m \rangle$. Since $\sum_m \langle e_{v(i)}, e_m\rangle e_m = e_{v(i)}$, its right-hand side is $k!\,\omega(e_{v(1)},\ldots,e_{v(k)})$, which is the left-hand side of the claim evaluated at that tuple. Its left-hand side is $\sum_{u} \det\big(\langle e_{v(i)}, e_{u(j)}\rangle\big)\,\omega(e_{u(1)},\ldots,e_{u(k)})$, and by the determinant convention for the wedge product together with the symmetry of the inner product, $\det\big(\langle e_{v(i)}, e_{u(j)}\rangle\big) = \big(e_{u(1)}^\flat \wedge \cdots \wedge e_{u(k)}^\flat\big)(e_{v(1)},\ldots,e_{v(k)})$. So this is the right-hand side of the claim evaluated at that tuple.
\end{proof}

\begin{lemma}
\label{lem:span-wedge-covectors}
\lean{LeeLib.Ch02.span_range_wedgeCovectors}
\leanok
\uses{lem:k-covector-wedge-expansion}
The wedge products $\omega^1 \wedge \cdots \wedge \omega^k$ of covectors span $\Lambda^k(V^*)$.
\end{lemma}
\begin{proof}
\leanok
\uses{lem:k-covector-wedge-expansion}
Immediate from Lemma~\ref{lem:k-covector-wedge-expansion}, since $k!$ is invertible in $\mathbb{R}$: every $\omega$ is $\frac{1}{k!}$ times a linear combination of wedges of covectors.
\end{proof}

\begin{proposition}
\label{prop:inner-k-covectors-unique}
\lean{LeeLib.Ch02.eq_innerForms_of_wedgeCovectors}
\leanok
\uses{def:inner-k-covectors,lem:span-wedge-covectors,lem:inner-k-covectors-wedge}
$\langle \cdot, \cdot \rangle_e$ is the unique bilinear form $B$ on $\Lambda^k(V^*)$ satisfying
\[
B(\omega^1 \wedge \cdots \wedge \omega^k, \eta^1 \wedge \cdots \wedge \eta^k) = \det\big(\langle \omega^i, \eta^j\rangle\big)
\]
for all covectors $\omega^i, \eta^j$.
\end{proposition}
\begin{proof}
\leanok
\uses{lem:span-wedge-covectors,lem:inner-k-covectors-wedge}
Lemma~\ref{lem:inner-k-covectors-wedge} shows $\langle \cdot, \cdot \rangle_e$ has the stated property. Conversely, a bilinear form is determined by its values on a spanning set of each argument, and the wedges of covectors span $\Lambda^k(V^*)$ by Lemma~\ref{lem:span-wedge-covectors}; the displayed identity prescribes exactly those values.
\end{proof}

\begin{proposition}
\label{prop:inner-k-covectors-frame-independent}
\lean{LeeLib.Ch02.innerForms_eq_innerForms}
\leanok
\uses{prop:inner-k-covectors-unique,lem:inner-k-covectors-wedge}
$\langle \cdot, \cdot \rangle_e$ does not depend on the orthonormal basis $e$ of $V$ used to define it.
\end{proposition}
\begin{proof}
\leanok
\uses{prop:inner-k-covectors-unique}
If $e$ and $f$ are orthonormal bases of $V$, then by Lemma~\ref{lem:inner-k-covectors-wedge}, applied once to each, both $\langle \cdot, \cdot\rangle_e$ and $\langle \cdot, \cdot\rangle_f$ satisfy the identity of Proposition~\ref{prop:inner-k-covectors-unique}, which determines the form uniquely. Hence they are equal.
\end{proof}

\begin{exercise}
\label{prob:fiber-metric-exterior-powers}
\uses{prop:inner-k-covectors-unique,prop:inner-k-covectors-frame-independent,lem:inner-k-covectors-posdef}
\dcref{ch2:2-16}
Let $(M, g)$ be a Riemannian $n$-manifold. Show that for each $k = 1, \ldots, n$, there is a unique fiber metric $\langle \cdot, \cdot \rangle_g$ on the bundle $\Lambda^k T^* M$ that satisfies
\[
\langle \omega^1 \wedge \cdots \wedge \omega^k, \eta^1 \wedge \cdots \wedge \eta^k \rangle_g = \det\left(\langle \omega^i, \eta^j \rangle_g\right) \quad (2.26)
\]
whenever $\omega^1, \ldots, \omega^k, \eta^1, \ldots, \eta^k$ are covectors at a point $p \in M$. [Hint: Define the inner product locally by declaring the set of $k$-covectors
\[
\{e^{i_1} \wedge \cdots \wedge e^{i_k} |_p : i_1 < \cdots < i_k\}
\]
to be an orthonormal basis for $\Lambda^k(T_p^* M)$ whenever $(e^i)$ is a local orthonormal coframe for $T^* M$, and then prove that the resulting inner product satisfies (2.26) and is independent of the choice of frame.]
\end{exercise}

\begin{exercise}
\label{prob:inner-products-k-forms}
\dcref{ch2:2-17}
\uses{prob:fiber-metric-exterior-powers,prop:inner-products-of-tensors}
Because we regard the bundle $\Lambda^k T^* M$ of $k$-forms as a subbundle of the bundle $T^k T^* M$ of covariant $k$-tensors, we have two inner products to choose from on $k$-forms: the one defined in Problem 2-16, and the restriction of the tensor inner product defined in Proposition 2.40. For this problem, we use the notation $\langle \cdot, \cdot \rangle$ to denote the inner product of Problem 2-16, and $\langle\langle \cdot, \cdot \rangle\rangle$ to denote the restriction of the tensor inner product.

(a) Using the convention for the wedge product that we use in this book (see p. 400), prove that
\[
\langle \cdot, \cdot \rangle_g = \frac{1}{k!} \langle\langle \cdot, \cdot \rangle\rangle_g.
\]

(b) Show that if the Alt convention is used for the wedge product (p. 401), the formula becomes
\[
\langle\langle \cdot, \cdot \rangle\rangle_g = k! \langle \cdot, \cdot \rangle_g.
\]
\end{exercise}

\begin{exercise}\label{prob:hodge-star-operator}\dcref{ch2:2-18}
Let $(M,g)$ be an oriented Riemannian $n$-manifold.

(a) For each $k = 0, \ldots, n$, show that there is a unique smooth bundle homomorphism $*: \Lambda^k T^* M \to \Lambda^{n-k}T^*M$, called the \textbf{Hodge star operator}, satisfying
\[
\omega \wedge *\eta = \langle\omega,\eta\rangle_g dV_g
\]
for all smooth $k$-forms $\omega, \eta$, where $\langle\cdot,\cdot\rangle_g$ is the inner product on $k$-forms defined in Problem~2-16. (For $k = 0$, interpret the inner product as ordinary multiplication.) [Hint: First prove uniqueness, and then define $*$ locally by setting
\[
*(e^{i_1} \wedge \cdots \wedge e^{i_k}) = \pm e^{j_1} \wedge \cdots \wedge e^{j_{n-k}}
\]
in terms of an orthonormal coframe $(e^i)$, where the indices $j_1,\ldots,j_{n-k}$ are chosen such that $(i_1,\ldots,i_k,j_1,\ldots,j_{n-k})$ is some permutation of $(1,\ldots,n)$.]

(b) Show that $*: \Lambda^0 T^*M \to \Lambda^n T^*M$ is given by $*f = f dV_g$.

(c) Show that $* * \omega = (-1)^{k(n-k)}\omega$ if $\omega$ is a $k$-form.
\end{exercise}

\begin{exercise}\label{prob:hodge-star-euclidean}\dcref{ch2:2-19}
Regard $\mathbb{R}^n$ as a Riemannian manifold with the Euclidean metric and the standard orientation, and let $*$ denote the Hodge star operator defined in Problem~2-18.

(a) Calculate $* dx^i$ for $i = 1, \ldots, n$.

(b) Calculate $*(dx^i \wedge dx^j)$ in the case $n = 4$.
\end{exercise}

\begin{exercise}\label{prob:self-dual-anti-self-dual}\dcref{ch2:2-20}
Let $M$ be an oriented Riemannian 4-manifold. A 2-form $\omega$ on $M$ is said to be \textbf{self-dual} if $*\omega = \omega$, and \textbf{anti-self-dual} if $*\omega = -\omega$.

(a) Show that every 2-form $\omega$ on $M$ can be written uniquely as a sum of a self-dual form and an anti-self-dual form.

(b) On $M = \mathbb{R}^4$ with the Euclidean metric, determine the self-dual and anti-self-dual 2-forms in standard coordinates.
\end{exercise}

\begin{exercise}\label{prob:divergence-laplacian-formulas-ch2}\dcref{ch2:2-21}
Prove Proposition~2.46 (the coordinate formulas for the divergence and the Laplacian).
\uses{prop:coordinate-representations-divergence-laplacian}
\end{exercise}

\begin{exercise}\label{prob:divergence-theorem}\dcref{ch2:2-22}
Suppose $(M,g)$ is a compact Riemannian manifold with boundary.

(a) Prove the following \textbf{divergence theorem} for $X \in \mathfrak{X}(M)$:
\[
\int_M (\operatorname{div} X) dV_g = \int_{\partial M} \langle X, N\rangle_g d\bar{g},
\]
where $N$ is the outward unit normal to $\partial M$ and $\bar{g}$ is the induced metric on $\partial M$. [Hint: Prove it first in the case that $M$ is orientable, and then apply that case to the orientation covering of $M$ (Prop.~B.18).]
\uses{prop:orientation-covering-theorem}

(b) Show that the divergence operator satisfies the following product rule for
$u \in C^\infty(M)$ and $X \in \mathfrak{X}(M)$:
\[
\operatorname{div}(uX) = u \operatorname{div} X + \langle \operatorname{grad} u, X \rangle_g,
\]
and deduce the following ``integration by parts'' formula:
\[
\int_M \langle \operatorname{grad} u, X \rangle_g \, dV_g = \int_{\partial M} u(X,N)_g \, d\tilde{V}_g - \int_M u \operatorname{div} X \, dV_g.
\]
What does this say when $M$ is a compact interval in $\mathbb{R}$?

(Used on p.\ 149.)
\end{exercise}

\begin{exercise}\label{prob:harmonic-functions}\dcref{ch2:2-23}
Let $(M,g)$ be a compact Riemannian manifold with or without boundary.
A function $u \in C^\infty(M)$ is said to be \emph{harmonic} if $\Delta u = 0$, where $\Delta$ is the
Laplacian defined on page~32.

(a) Prove \emph{Green's identities}:
\[
\int_M u\Delta v \, dV_g = \int_{\partial M} u N v \, d\tilde{V}_g - \int_M \langle \operatorname{grad} u, \operatorname{grad} v \rangle_g \, dV_g,
\]
\[
\int_M (u\Delta v - v\Delta u) \, dV_g = \int_{\partial M} (u N v - v N u) \, d\tilde{V}_g,
\]
where $N$ is the outward unit normal vector field on $\partial M$ and $\tilde{g}$ is the
induced metric on $\partial M$.

(b) Show that if $M$ is connected, $\partial M \ne \emptyset$, and $u, v$ are harmonic functions
on $M$ whose restrictions to $\partial M$ agree, then $u \equiv v$.

(c) Show that if $M$ is connected and $\partial M = \emptyset$, then the only harmonic
functions on $M$ are the constants, and every smooth function $u$ satisfies
$\int_M \Delta u \, dV_g = 0$.

(Used on pp.\ 149, 223.)
\end{exercise}

\begin{exercise}\label{prob:eigenvalues}\dcref{ch2:2-24}
Let $(M,g)$ be a compact Riemannian manifold (without boundary). A real
number $\lambda$ is called an \emph{eigenvalue of $M$} if there exists a smooth function $u$
on $M$, not identically zero, such that $-\Delta u = \lambda u$. In this case, $u$ is called an
\emph{eigenfunction} corresponding to $\lambda$.

(a) Prove that $0$ is an eigenvalue of $M$, and that all other eigenvalues are
strictly positive.

(b) Show that if $u$ and $v$ are eigenfunctions corresponding to distinct eigen-
values, then $\int_M uv \, dV_g = 0$.

(Used on p.\ 149.)
\end{exercise}

\begin{exercise}\label{prob:dirichlet-neumann-eigenvalues}\dcref{ch2:2-25}
Let $(M,g)$ be a compact connected Riemannian $n$-manifold with nonempty
boundary. A number $\lambda \in \mathbb{R}$ is called a \emph{Dirichlet eigenvalue for $M$} if there
exists a smooth real-valued function $u$ on $M$, not identically zero, such that
$-\Delta u = \lambda u$ and $u|_{\partial M} = 0$. Similarly, $\lambda$ is called a \emph{Neumann eigenvalue} if there exists a smooth real-valued function $u$ on $M$, not identically zero, such that $-\Delta u = \lambda u$ and $Nu|_{\partial M} = 0$, where $N$ is the outward unit normal vector field along $\partial M$.

(a) Show that every Dirichlet eigenvalue is strictly positive.
(b) Show that 0 is a Neumann eigenvalue, and all other Neumann eigenvalues are strictly positive.
\end{exercise}

\begin{exercise}[Dirichlet's Principle]
\label{prob:dirichlet-principle}
\dcref{ch2:2-26}
Suppose $(M,g)$ is a compact connected Riemannian $n$-manifold with nonempty boundary. Prove that a function $u \in C^{\infty}(M)$ is harmonic if and only if it minimizes $\int_M |\operatorname{grad} u|^2 \, dV_g$ among all smooth functions with the same boundary values. [Hint: For any function $f \in C^{\infty}(M)$ that vanishes on $\partial M$, expand $\int_M |\operatorname{grad}(u + \varepsilon f)|^2 \, dV_g$ and use Problem 2-22.]
\uses{prob:divergence-theorem}
\end{exercise}

\begin{exercise}
\label{prob:curl-vector-field}
\dcref{ch2:2-27}
Suppose $(M,g)$ is an oriented Riemannian 3-manifold.

(a) Define $\beta : TM \to \Lambda^2 T^* M$ by $\beta(X) = X \mathbin{\lrcorner} dV_g$. Show that $\beta$ is a smooth bundle isomorphism, and thus we can define the \emph{curl} of a vector field $X \in \mathfrak{X}(M)$ by
\[
\operatorname{curl} X = \beta^{-1} d(X^b).
\]

(b) Show that the following diagram commutes:
\[
\begin{array}{ccccccc}
C^{\infty}(M) & \xrightarrow{\operatorname{grad}} & \mathfrak{X}(M) & \xrightarrow{\operatorname{curl}} & \mathfrak{X}(M) & \xrightarrow{\operatorname{div}} & C^{\infty}(M) \\
\downarrow \text{Id} & & \downarrow b & & \downarrow \beta & & \downarrow * \\
\Omega^0(M) & \xrightarrow{d} & \Omega^1(M) & \xrightarrow{d} & \Omega^2(M) & \xrightarrow{d} & \Omega^3(M)
\end{array}
\]
where $*(f) = f \, dV_g$, and use this to prove that $\operatorname{curl}(\operatorname{grad} f) = 0$ for every $f \in C^{\infty}(M)$, and $\operatorname{div}(\operatorname{curl} X) = 0$ for every $X \in \mathfrak{X}(M)$.

(c) Compute the formula for the curl in standard coordinates on $\mathbb{R}^3$ with the Euclidean metric.
\end{exercise}

\begin{exercise}
\label{prob:hodge-star-exterior-derivative}
\dcref{ch2:2-28}
Let $(M,g)$ be an oriented Riemannian manifold and let $*$ denote its Hodge star operator (Problem 2-18). Show that for every $X \in \mathfrak{X}(M)$,
\[
X \mathbin{\lrcorner} dV_g = *(X^b),
\]
\[
\operatorname{div} X = *d*(X^b),
\]
and, when $\dim M = 3$,
\[
\operatorname{curl} X = (*d(X^b))^{\#},
\]
where the curl of a vector field is defined as in Problem 2-27.
\uses{prob:hodge-star-operator,prob:curl-vector-field}
\end{exercise}

\begin{exercise}
\label{prob:codifferential-operator}
\dcref{ch2:2-29}
Let $(M,g)$ be a compact oriented Riemannian $n$-manifold. For $1 \leq k \leq n$, define a map $d^* : \Omega^k(M) \to \Omega^{k-1}(M)$ by $d^* \omega = (-1)^{n(k+1)+1} * d * \omega$, where $*$ is the Hodge star operator defined in Problem 2-18. Extend this definition to 0-forms by defining $d^*\omega = 0$ for $\omega \in \Omega^0(M)$.

\begin{enumerate}
\item[(a)] Show that $d^* \circ d^* = 0$.
\item[(b)] Show that the formula
\[
(\omega, \eta) = \int_M \langle \omega, \eta \rangle_g \, dV_g
\]
defines an inner product on $\Omega^k(M)$ for each $k$, where $(\cdot, \cdot)_g$ is the inner product on forms defined in Problem 2-16.
\item[(c)] Show that $(d^*\omega, \eta) = (\omega, d\eta)$ for all $\omega \in \Omega^k(M)$ and $\eta \in \Omega^{k-1}(M)$.
\end{enumerate}
\end{exercise}

\begin{exercise}\label{prob:distance-disconnected-riemannian}\dcref{ch2:2-30}
Suppose $(M,g)$ is a (not necessarily connected) Riemannian manifold. Show that there is a distance function $d$ on $M$ that induces the given topology and restricts to the Riemannian distance on each component of $M$. (Used on p. 187.)
\end{exercise}

\begin{exercise}\label{prob:local-isometry-distance}\dcref{ch2:2-31}
Suppose $(M,g)$ and $(\tilde{M}, \tilde{g})$ are connected Riemannian manifolds, and $\varphi: M \to \tilde{M}$ is a local isometry. Show that $d_{\tilde{g}}(\varphi(x), \varphi(y)) \leq d_g(x,y)$ for all $x, y \in M$. Give an example to show that equality need not hold. (Used on p. 37.)
\end{exercise}

\begin{exercise}\label{prob:arc-length-integral}\dcref{ch2:2-32}
Let $(M,g)$ be a Riemannian manifold and $\gamma: [a,b] \to M$ a smooth curve segment. For each continuous function $f: [a,b] \to \mathbb{R}$, we define the \textit{integral of $f$ with respect to arc length}, denoted by $\int_\gamma f \, ds$, by
\[
\int_\gamma f \, ds = \int_a^b f(t) |\gamma'(t)|_g \, dt.
\]

\begin{enumerate}
\item[(a)] Show that $\int_\gamma f \, ds$ is independent of parametrization in the following sense: if $\varphi: [c,d] \to [a,b]$ is a diffeomorphism, then
\[
\int_a^b f(t) |\gamma'(t)|_g \, dt = \int_c^d \tilde{f}(u) |\tilde{\gamma}'(u)|_g \, du,
\]
where $\tilde{f} = f \circ \varphi$ and $\tilde{\gamma} = \gamma \circ \varphi$.
\item[(b)] Suppose now that $\gamma$ is a smooth embedding, so that $C = \gamma([a,b])$ is an embedded submanifold with boundary in $M$. Show that
\[
\int_\gamma f \, ds = \int_C (f \circ \gamma^{-1}) \, d\tilde{V},
\]
where $d\tilde{V}$ is the Riemannian volume element on $C$ associated with the induced metric and the orientation determined by $\gamma$.
\end{enumerate}

(Used on p. 273.)
\end{exercise}

\begin{exercise}\label{prob:sylvesters-law-inertia}\dcref{ch2:2-33}
\leanok
\uses{prop:sylvesters-law-inertia}

Prove Proposition 2.65 (Sylvester's law of inertia).
\end{exercise}

\begin{exercise}\label{prob:lorentz-metric-existence}\dcref{ch2:2-34}
\lean{LeeLib.Ch02.exists_isLorentzMetric_iff_exists_isRankOneDistribution}
\leanok
\uses{thm:lorentz-metrics}
Prove Theorem 2.69 (existence of Lorentz metrics), as follows.

\begin{enumerate}
\item[(a)] For sufficiency, assume that $D \subseteq TM$ is a 1-dimensional distribution, and choose any Riemannian metric $g$ on $M$. Show that locally it is possible to choose a $g$-orthonormal frame $(E_i)$ and dual coframe $(\varepsilon^i)$ such that $E_1$ spans $D$; and then show that the Lorentz metric $-(\varepsilon^1)^2 + (\varepsilon^2)^2 + \cdots + (\varepsilon^n)^2$ is independent of the choice of frame.

\item[(b)] To prove necessity, suppose that $g$ is a Lorentz metric on $M$, and let $g_0$ be any Riemannian metric. Show that for each $p \in M$, there are exactly two $g_0$-unit vectors $v_0, -v_0$ on which the function $v \mapsto g(v, v)$ takes its minimum among all unit vectors in $T_p M$, and use Lagrange multipliers to conclude that there exists a number $\lambda(p) < 0$ such that $g(v_0, w) = \lambda(p)g_0(v_0, w)$ for all $w \in T_p M$. You may use the following standard result from perturbation theory: if $U$ is an open subset of $\mathbb{R}^n$ and $A: U \to GL(n, \mathbb{R})$ is a smooth matrix-valued function such that $A(x)$ is symmetric and has exactly one negative eigenvalue for each $x \in U$, then there exist smooth functions $\lambda: U \to (-\infty, 0)$ and $X: U \to \mathbb{R}^n \setminus \{0\}$ such that $A(x)X(x) = \lambda(x)X(x)$ for all $x \in U$.
\end{enumerate}
\end{exercise}

\begin{exercise}\label{prob:level-sets-pseudo-riemannian}\dcref{ch2:2-35}
\lean{LeeLib.Ch02.hasSignature_pseudoPullbackMetric_levelSet_of_pseudoGrad_pos, LeeLib.Ch02.hasSignature_pseudoPullbackMetric_levelSet_of_pseudoGrad_neg, LeeLib.Ch02.form_pseudoGrad_mfderiv_levelSet_val_eq_zero}
\leanok
\uses{cor:pseudo-riemannian-embedding-criterion}
Prove Corollary 2.71 (about level sets in pseudo-Riemannian manifolds). (Used on p.\ 63.)
\end{exercise}