\chapter{Jacobi Fields}
\label{chap:dc-jacobi-fields}

Faithful transcription of do Carmo, \emph{Riemannian Geometry}. Each numbered
statement keeps do Carmo's number, kind and (name); proofs keep his explicit
cross-references ("by Lemma 4.1 of Chap. 4", "from \cref{cor:dc-ch5-2-5}") so the
dependency graph is recoverable from the text.

\section{Introduction}

In this chapter we derive a first relation between the two basic concepts
introduced previously, namely geodesics and curvature. As we shall see (cf.
\cref{rem:dc-ch5-2-11}), the curvature $K(p,\sigma)$, $\sigma\subset T_p M$, determines how
fast the geodesics that start from $p$ and are tangent to $\sigma$ spread apart.
To formalize precisely this velocity of deviation of the geodesics, it is
necessary to introduce the so-called \emph{Jacobi fields}. Jacobi fields are vector
fields along geodesics, defined by means of a differential equation that arises
naturally in the study of the exponential mapping (cf. Sec. 2). Besides
furnishing the relation mentioned above, the Jacobi fields allow us to obtain a
simple characterization of the singularities of the exponential map (cf.
\cref{prop:dc-ch5-3-5}).

\section{The Jacobi equation}

Let $M$ be a Riemannian manifold and let $p\in M$. In the proof of the Gauss
Lemma we saw that if $\exp_p$ is defined at $v\in T_p M$, and if
$w\in T_v(T_p M)$, then

$$
(d\exp_p)_v w=\frac{\partial f}{\partial s}(1,0),
$$

where $f$ is a parametrized surface given by

$$
f(t,s)=\exp_p tv(s),\quad 0\le t\le 1,\quad -\varepsilon\le s\le\varepsilon,
$$

and $v(s)$ is a curve in $T_p M$ with $v(0)=v$, $v'(0)=w$.

We would like to obtain information on $|(d\exp_p)_v(w)|$, which denotes,
intuitively, the rate of spreading of the geodesics $t\mapsto\exp_p tv(s)$ that
start from $p$. This spreading is associated with the value of the sectional
curvature at $p$ with respect to the plane generated by $v$ and $w$. Also, if
$|(d\exp_p)_v(w)|=0$ with $w\neq 0$, then $v$ is a critical point of $\exp_p$. It
is convenient to study the field

$$
(d\exp_p)_{tv}(tw)=\frac{\partial f}{\partial s}(t,0)
$$

along the geodesic $\gamma(t)=\exp_p(tv)$, $0\le t\le 1$.

The basic remark is that $\frac{\partial f}{\partial s}$ satisfies a differential
equation. Since $\gamma$ is a geodesic, we have for all $(t,s)$,
$\frac{D}{\partial t}\frac{\partial f}{\partial t}=0$. Thus, from Lemma 4.1 of
Chap. 4,

$$
0=\frac{D}{\partial s}\Big(\frac{D}{\partial t}\frac{\partial f}{\partial t}\Big)
=\frac{D}{\partial t}\frac{D}{\partial s}\frac{\partial f}{\partial t}
-R\Big(\frac{\partial f}{\partial s},\frac{\partial f}{\partial t}\Big)\frac{\partial f}{\partial t}
=\frac{D}{\partial t}\frac{D}{\partial t}\frac{\partial f}{\partial s}
+R\Big(\frac{\partial f}{\partial t},\frac{\partial f}{\partial s}\Big)\frac{\partial f}{\partial t}.
$$

Putting $\frac{\partial f}{\partial s}(t,0)=J(t)$, we obtain that $J$ satisfies
the equation

$$
\frac{D^2 J}{dt^2}+R(\gamma'(t),J(t))\gamma'(t)=0.\qquad(1)
$$

The equation above is called the \emph{Jacobi equation}.

\begin{definition}[Jacobi field]
  \label{def:dc-ch5-2-1}
  \dcref{ch5:2.1}
  Let $\gamma:[0,a]\to M$ be a geodesic in $M$. A vector field $J$ along $\gamma$ is
  said to be a \emph{Jacobi field} if it satisfies the Jacobi equation (1), for all
  $t\in[0,a]$.
  
  A Jacobi field is determined by its initial conditions $J(0)$,
  $\frac{DJ}{dt}(0)$. Indeed, let $e_1(t),\dots,e_n(t)$ be parallel, orthonormal
  fields along $\gamma$. Writing
  
  $$
  J(t)=\sum_i f_i(t)e_i(t),\quad a_{ij}=\langle R(\gamma'(t),e_i(t))\gamma'(t),e_j(t)\rangle,\quad i,j=1,\dots,n=\dim M,
  $$
  
  one has $\frac{D^2 J}{dt^2}=\sum_i f_i''(t)e_i(t)$ and
  $R(\gamma',J)\gamma'=\sum_{ij}f_i a_{ij}e_j$. Therefore equation (1) is equivalent
  to the system
  
  $$
  f_j''(t)+\sum_i a_{ij}(t)f_i(t)=0,\quad j=1,\dots,n,
  $$
  
  which is a linear system of the second order. Hence, given the initial
  conditions $J(0)$, $\frac{DJ}{dt}(0)$, there exists a $C^\infty$ solution defined
  on $[0,a]$. There exist, therefore, $2n$ linearly independent Jacobi fields along
  $\gamma$.
\end{definition}

\begin{remark}[Remark 2.2]
  \label{rem:dc-ch5-2-2}
  \dcref{ch5:2.2}
  It is worth noting that $\gamma'(t)$ and $t\gamma'(t)$ are Jacobi fields along
  $\gamma$. The first field has derivative zero and vanishes nowhere; the second is
  zero if and only if $t=0$. Due to these facts, we shall consider Jacobi fields
  along $\gamma$ that are normal to $\gamma'$.
\end{remark}

\begin{example}[Jacobi fields on manifolds of constant curvature]
  \label{ex:dc-ch5-2-3}
  \dcref{ch5:2.3}
  \uses{lem:dc-ch4-3-4}
  Let $M$ be a Riemannian manifold of constant sectional curvature $K$, and let
  $\gamma:[0,\ell]\to M$ be a normalized geodesic on $M$. Let $J$ be a Jacobi field
  along $\gamma$, normal to $\gamma'$. We claim that from $|\gamma'|=1$ and from
  \cref{lem:dc-ch4-3-4}, it follows that
  
  $$
  R(\gamma',J)\gamma'=KJ.
  $$
  
  Indeed, for all vector fields $T$ along $\gamma$,
  
  $$
  \langle R(\gamma',J)\gamma',T\rangle=K\{\langle\gamma',\gamma'\rangle\langle J,T\rangle-\langle\gamma',T\rangle\langle J,\gamma'\rangle\}=K\langle J,T\rangle.
  $$
  
  As a result, the Jacobi equation can be written as
  
  $$
  \frac{D^2 J}{dt^2}+KJ=0.\qquad(2)
  $$
  
  Let $w(t)$ be a parallel field along $\gamma$ with $\langle\gamma'(t),w(t)\rangle=0$
  and $|w(t)|=1$. It is easy to verify that
  
  $$
  J(t)=\begin{cases}\dfrac{\sin(t\sqrt{K})}{\sqrt{K}}\,w(t),&\text{if }K>0,\\[1ex] t\,w(t),&\text{if }K=0,\\[1ex]\dfrac{\sinh(t\sqrt{-K})}{\sqrt{-K}}\,w(t),&\text{if }K<0,\end{cases}
  $$
  
  is a solution of (2) with initial conditions $J(0)=0$, $J'(0)=w(0)$.
  
  As we saw previously, given $p\in M$, $v\in T_p M$, and $w\in T_v(T_p M)$, we can
  construct a Jacobi field along the geodesic $\gamma:[0,1]\to M$ given by
  $\gamma(t)=\exp_p tv$. For that, consider $f(t,s)=\exp_p tv(s)$, where $v(s)$ is a
  curve in $T_p M$ with $v(0)=v$, $v'(0)=w$, and take
  $J(t)=\frac{\partial f}{\partial s}(t,0)$. Notice that $J(0)=0$. We are going to
  show that this is essentially the only way of constructing Jacobi fields along
  $\gamma(t)$ with $J(0)=0$.
\end{example}

\begin{proposition}[Proposition 2.4]
  \label{prop:dc-ch5-2-4}
  \dcref{ch5:2.4}
  Let $\gamma:[0,a]\to M$ be a geodesic and let $J$ be a Jacobi field along
  $\gamma$ with $J(0)=0$. Put $\frac{DJ}{dt}(0)=w$ and $\gamma'(0)=v$. Using the
  natural identification $T_{av}(T_{\gamma(0)}M)\simeq T_{\gamma(0)}M$, construct a
  curve $v(s)$ in $T_{\gamma(0)}M$ with $v(0)=av$, $v'(0)=aw$. Put
  $f(t,s)=\exp_p(\frac{t}{a}v(s))$,
  $p=\gamma(0)$, and define a Jacobi field $\bar J$ by
  $\bar J(t)=\frac{\partial f}{\partial s}(t,0)$. Then $\bar J=J$ on $[0,a]$.
\end{proposition}
\begin{proof}
  For $s=0$,
  
  $$
  \frac{D}{dt}\frac{\partial f}{\partial s}=\frac{D}{\partial t}\big((d\exp_p)_{tv}(tw)\big)
  =\frac{D}{\partial t}\big(t(d\exp_p)_{tv}(w)\big)
  =(d\exp_p)_{tv}(w)+t\frac{D}{\partial t}\big((d\exp_p)_{tv}(w)\big).
  $$
  
  Therefore, for $t=0$,
  
  $$
  \frac{D\bar J}{dt}(0)=\frac{D}{\partial t}\frac{\partial f}{\partial s}(0,0)=(d\exp_p)_o(w)=w.
  $$
  
  Since $J(0)=\bar J(0)=0$ and $\frac{DJ}{dt}(0)=\frac{D\bar J}{dt}(0)=w$, we
  conclude, from the uniqueness theorem, that $J=\bar J$. $\square$
\end{proof}

\begin{corollary}[Corollary 2.5]
  \label{cor:dc-ch5-2-5}
  \dcref{ch5:2.5}
  Let $\gamma:[0,a]\to M$ be a geodesic and put $p=\gamma(0)$. Then a Jacobi field
  $J$ along $\gamma$ with $J(0)=0$ is given by
  
  $$
  J(t)=(d\exp_p)_{t\gamma'(0)}(tJ'(0)),\quad t\in[0,a].
  $$
\end{corollary}

\begin{remark}[Remark 2.6]
  \label{rem:dc-ch5-2-6}
  \dcref{ch5:2.6}
  \uses{prop:dc-ch5-2-4}
  It is possible to obtain a construction analogous to \cref{prop:dc-ch5-2-4} for Jacobi
  fields that do not satisfy the condition $J(0)=0$. Since we shall not use this
  fact, we leave its proof as an exercise (Exercise 2).
  
  From now on, for simplicity of notation, we put $\frac{DJ}{dt}=J'$,
  $\frac{D^2 J}{dt^2}=J''$, etc.
\end{remark}

\begin{proposition}[Proposition 2.7]
  \label{prop:dc-ch5-2-7}
  \dcref{ch5:2.7}
  Let $p\in M$ and $\gamma:[0,a]\to M$ be a geodesic with $\gamma(0)=p$,
  $\gamma'(0)=v$. Let $w\in T_v(T_p M)$ with $|w|=1$ and let $J$ be a Jacobi field
  along $\gamma$ given by
  
  $$
  J(t)=(d\exp_p)_{tv}(tw),\quad 0\le t\le a.
  $$
  
  Then the Taylor expansion of $|J(t)|^2$ about $t=0$ is given by
  
  $$
  |J(t)|^2=t^2-\tfrac{1}{3}\langle R(v,w)v,w\rangle t^4+R(t),\qquad(3)
  $$
  
  where $\lim_{t\to 0}\frac{R(t)}{t^4}=0$.
\end{proposition}
\begin{proof}
  Since $J(0)=0$ and $J'(0)=w$, the first three coefficients are
  
  $$
  \langle J,J\rangle(0)=0,\quad\langle J,J\rangle'(0)=2\langle J,J'\rangle(0)=0,\quad\langle J,J\rangle''(0)=2\langle J',J'\rangle(0)+2\langle J'',J\rangle(0)=2.
  $$
  
  Since $J''(0)=-R(\gamma',J)\gamma'(0)=0$, we have
  $\langle J,J\rangle'''(0)=6\langle J',J''\rangle(0)+2\langle J''',J\rangle(0)=0$. Now
  we need the fact
  
  $$
  \nabla_{\gamma'}(R(\gamma',J)\gamma')(0)=R(\gamma',J')\gamma'(0).\qquad(4)
  $$
  
  To prove (4), note that for any $W$, at $t=0$,
  
  $$
  \Big\langle\frac{D}{dt}(R(\gamma',J)\gamma'),W\Big\rangle
  =\frac{d}{dt}\langle R(\gamma',W)\gamma',J\rangle-\langle R(\gamma',J)\gamma',W'\rangle
  =\Big\langle\frac{D}{dt}(R(\gamma',W)\gamma'),J\Big\rangle+\langle R(\gamma',W)\gamma',J'\rangle
  =\langle R(\gamma',J')\gamma',W\rangle,
  $$
  
  which implies (4). It follows from (4) and the Jacobi equation that
  $J'''(0)=-R(\gamma',J')\gamma'(0)$. Therefore,
  
  $$
  \langle J,J\rangle''''(0)=8\langle J',J'''\rangle(0)+6\langle J'',J''\rangle(0)+2\langle J'''',J\rangle(0)=-8\langle J',R(\gamma',J')\gamma'\rangle(0)=-8\langle R(v,w)v,w\rangle.
  $$
  
  Putting together the calculation above, we obtain (3). $\square$
\end{proof}

\begin{remark}[Remark 2.8]
  \label{rem:dc-ch5-2-8}
  \dcref{ch5:2.8}
  The expression (4) can also be obtained using the (covariant) derivation of
  tensors described in Section 5 of Chap. 4 (see Exercise 5 of Chap. 4).
\end{remark}

\begin{corollary}[Corollary 2.9]
  \label{cor:dc-ch5-2-9}
  \dcref{ch5:2.9}
  If $\gamma:[0,\ell]\to M$ is parametrized by arc length (i.e. $|v|=1$) and
  $\langle w,v\rangle=0$, the expression $\langle R(v,w)v,w\rangle$ is the sectional
  curvature at $p$ with respect to the plane $\sigma$ generated by $v$ and $w$.
  Therefore, in this situation,
  
  $$
  |J(t)|^2=t^2-\tfrac{1}{3}K(p,\sigma)t^4+R(t).\qquad(5)
  $$
\end{corollary}

\begin{corollary}[Corollary 2.10]
  \label{cor:dc-ch5-2-10}
  \dcref{ch5:2.10}
  With the same conditions as in the previous corollary,
  
  $$
  |J(t)|=t-\tfrac{1}{6}K(p,\sigma)t^3+\tilde R(t),\quad\lim_{t\to 0}\frac{\tilde R(t)}{t^3}=0.\qquad(6)
  $$
\end{corollary}

\begin{remark}[Remark 2.11]
  \label{rem:dc-ch5-2-11}
  \dcref{ch5:2.11}
  The expression (6) essentially contains the relation between geodesics and
  curvature mentioned at the beginning of this chapter. Considering the
  parametrized surface
  
  $$
  f(t,s)=\exp_p tv(s),\quad t\in[0,\delta],\quad s\in(-\varepsilon,\varepsilon),
  $$
  
  where $\delta$ is chosen so small that $\exp_p tv(s)$ is defined, and $v(s)$ is a
  curve in $T_p M$ with $|v(s)|=1$, $v(0)=v$, $v'(0)=w$, we see that the rays
  $t\mapsto tv(s)$, $t\in[0,\delta]$, that start from the origin $0$ of $T_p M$,
  deviate from the ray $t\mapsto tv(0)$ with velocity
  
  $$
  \Big|\Big(\frac{\partial}{\partial s}tv(s)\Big)(0)\Big|=|tw|=t.
  $$
  
  On the other hand, (6) tells us that the geodesics $t\mapsto\exp_p(tv(s))$
  deviate from the geodesic $\gamma(t)=\exp_p tv(0)$ with a velocity that differs
  from $t$ by a term of third order in $t$, given by $-\tfrac{1}{6}K(p,\sigma)t^3$.
  This tells us that, locally, the geodesics spread apart less than the rays in
  $T_p M$ if $K_p(\sigma)>0$, and more than the rays in $T_p M$ if $K_p(\sigma)<0$.
  For $t$ small, the value $K(p,\sigma)t^3$ furnishes an approximation for the
  extent of this spread with an error of order $t^3$.
\end{remark}

\section{Conjugate points}

We now turn to the relationship between the singularities of the exponential map
and Jacobi fields.

\begin{definition}[conjugate point, multiplicity]
  \label{def:dc-ch5-3-1}
  \dcref{ch5:3.1}
  Let $\gamma:[0,a]\to M$ be a geodesic. The point $\gamma(t_0)$ is said to be
  \emph{conjugate} to $\gamma(0)$ along $\gamma$, $t_0\in(0,a]$, if there exists a Jacobi
  field $J$ along $\gamma$, not identically zero, with $J(0)=0=J(t_0)$. The maximum
  number of such linearly independent fields is called the \emph{multiplicity} of the
  conjugate point $\gamma(t_0)$.
  
  Observe that if $\gamma(t_0)$ is conjugate to $\gamma(0)$, then $\gamma(0)$ is
  conjugate to $\gamma(t_0)$.
\end{definition}

\begin{remark}[Remark 3.2]
  \label{rem:dc-ch5-3-2}
  \dcref{ch5:3.2}
  \uses{rem:dc-ch5-2-2}
  If $\dim M=n$, there exist exactly $n$ linearly independent Jacobi fields along
  the geodesic $\gamma:[0,a]\to M$ which are zero at $\gamma(0)$. This follows from
  the fact, easily checked, that the Jacobi fields $J_1,\dots,J_k$ with $J_i(0)=0$
  are linearly independent if and only if $J_1'(0),\dots,J_k'(0)$ are linearly
  independent. In addition, the Jacobi field $J(t)=t\gamma'(t)$ never vanishes for
  $t\neq 0$ (see \cref{rem:dc-ch5-2-2}). From this we deduce that the multiplicity of a
  conjugate point never exceeds $n-1$.
\end{remark}

\begin{example}[Example 3.3]
  \label{ex:dc-ch5-3-3}
  \dcref{ch5:3.3}
  \uses{ex:dc-ch5-2-3}
  Let $S^n=\{x\in\mathbb{R}^{n+1};|x|=1\}$. Assuming the fact (to be proved in the
  next chapter) that the sectional curvatures of $S^n$ are all equal to one, the
  Jacobi field on $S^n$ given in \cref{ex:dc-ch5-2-3}, namely $J(t)=(\sin t)w(t)$, satisfies
  $J(0)=J(\pi)=0$. Therefore, along any geodesic $\gamma$ of $S^n$, the antipodal
  point $\gamma(\pi)$ to $\gamma(0)$ is conjugate to $\gamma(0)$. There exist $n-1$
  such fields that are linearly independent, that is, the multiplicity of
  $\gamma(\pi)$ as a conjugate point of $\gamma(0)$ is $n-1$.
\end{example}

\begin{definition}[conjugate locus]
  \label{def:dc-ch5-3-4}
  \dcref{ch5:3.4}
  The set of (first) conjugate points to the point $p\in M$, for all the geodesics
  that start at $p$, is called the \emph{conjugate locus} of $p$ and is denoted by
  $C(p)$.
  
  On $S^n$, $C(p)=\{-p\}$, for all $p$. The case of $S^n$, however, is not typical.
  A more typical example is given by the ellipsoid, where $C(p)$ is, in general, a
  curve with four singular points (see Fig. 4 of Chap. 13; cf. Braunmühl, A.,
  "Geodätische Linien auf dreiachsigen Flächen 2-Grades", Math. Ann. 20 (1882),
  557-586).
\end{definition}

\begin{proposition}[Proposition 3.5]
  \label{prop:dc-ch5-3-5}
  \dcref{ch5:3.5}
  \uses{cor:dc-ch5-2-5}
  Let $\gamma:[0,a]\to M$ be a geodesic and put $\gamma(0)=p$. The point
  $q=\gamma(t_0)$, $t_0\in(0,a]$, is conjugate to $p$ along $\gamma$ if and only if
  $v_0=t_0\gamma'(0)$ is a critical point of $\exp_p$. In addition, the multiplicity
  of $q$ as a conjugate point of $p$ is equal to the dimension of the kernel of the
  linear map $(d\exp_p)_{v_0}$.
\end{proposition}
\begin{proof}
  The point $q=\gamma(t_0)$ is a conjugate point of $p$ along $\gamma$ if
  and only if there exists a non-zero Jacobi field $J$ along $\gamma$ with
  $J(0)=J(t_0)=0$. Let $v=\gamma'(0)$ and $w=J'(0)$. From \cref{cor:dc-ch5-2-5},
  $J(t)=(d\exp_p)_{tv}(tw)$, $t\in[0,a]$. Observe that $J$ is non-zero if and only
  if $w\neq 0$. Therefore, $q=\gamma(t_0)$ is conjugate to $p$ if and only if
  
  $$
  0=J(t_0)=(d\exp_p)_{t_0 v}(t_0 w),\quad w\neq 0,
  $$
  
  that is, if and only if $t_0 v$ is a critical point of $\exp_p$. The first
  assertion is therefore proved.
  
  The multiplicity of $q$ is equal to the number of linearly independent Jacobi
  fields $J_1,\dots,J_k$ which are zero at $0$ and at $t_0$. The fields
  $J_1,\dots,J_k$ are linearly independent if and only if
  $J_1'(0),\dots,J_k'(0)$ are linearly independent in $T_p M$. From the
  construction above, the multiplicity of $q$ is equal to the dimension of the
  kernel of $(d\exp_p)_{t_0 v}$. $\square$
\end{proof}

\begin{proposition}[Proposition 3.6]
  \label{prop:dc-ch5-3-6}
  \dcref{ch5:3.6}
  Let $J$ be a Jacobi field along the geodesic $\gamma:[0,a]\to M$. Then
  
  $$
  \langle J(t),\gamma'(t)\rangle=\langle J'(0),\gamma'(0)\rangle t+\langle J(0),\gamma'(0)\rangle,\quad t\in[0,a].
  $$
\end{proposition}
\begin{proof}
  Omitting the $t$ for notation, we have from the Jacobi equation,
  
  $$
  \langle J',\gamma'\rangle'=\langle J'',\gamma'\rangle=-\langle R(\gamma',J)\gamma',\gamma'\rangle=0.
  $$
  
  Therefore, $\langle J',\gamma'\rangle=\langle J'(0),\gamma'(0)\rangle$. In
  addition,
  
  $$
  \langle J,\gamma'\rangle'=\langle J',\gamma'\rangle=\langle J'(0),\gamma'(0)\rangle.
  $$
  
  Integrating this last equation in $t$, we obtain finally
  $\langle J,\gamma'\rangle=\langle J'(0),\gamma'(0)\rangle t+\langle J(0),\gamma'(0)\rangle$.
  $\square$
\end{proof}

\begin{corollary}[Corollary 3.7]
  \label{cor:dc-ch5-3-7}
  \dcref{ch5:3.7}
  If $\langle J,\gamma'\rangle(t_1)=\langle J,\gamma'\rangle(t_2)$, $t_1,t_2\in[0,a]$,
  $t_1\neq t_2$, then $\langle J,\gamma'\rangle$ does not depend on $t$; in
  particular, if $J(0)=J(a)=0$, then $\langle J,\gamma'\rangle(t)\equiv 0$.
\end{corollary}

\begin{corollary}[Corollary 3.8]
  \label{cor:dc-ch5-3-8}
  \dcref{ch5:3.8}
  Suppose that $J(0)=0$. Then $\langle J'(0),\gamma'(0)\rangle=0$ if and only if
  $\langle J,\gamma'\rangle(t)\equiv 0$; in particular, the space of Jacobi fields
  $J$ with $J(0)=0$ and $\langle J,\gamma'\rangle(t)\equiv 0$ has dimension equal to
  $n-1$.
\end{corollary}

\begin{proposition}[Proposition 3.9]
  \label{prop:dc-ch5-3-9}
  \dcref{ch5:3.9}
  Let $\gamma:[0,a]\to M$ be a geodesic. Let $V_1\in T_{\gamma(0)}M$ and
  $V_2\in T_{\gamma(a)}M$. If $\gamma(a)$ is not conjugate to $\gamma(0)$ there
  exists a unique Jacobi field $J$ along $\gamma$, with $J(0)=V_1$ and $J(a)=V_2$.
\end{proposition}
\begin{proof}
  Let $\mathcal{J}$ be the space of Jacobi fields $J$ with $J(0)=0$. Define
  a mapping $\Theta:\mathcal{J}\to T_{\gamma(a)}M$ by $\Theta(J)=J(a)$,
  $J\in\mathcal{J}$. Since $\gamma(a)$ is not conjugate to $\gamma(0)$, $\Theta$ is
  injective. Indeed, if $J_1\neq J_2$ with $J_1(a)=J_2(a)$, we should have
  $J_1-J_2$, a non-zero Jacobi field with $(J_1-J_2)(0)=0$, which contradicts the
  fact that $\gamma(a)$ is not conjugate to $\gamma(0)$.
  
  Since $\Theta$ is linear, it follows from injectivity and the fact that
  $\dim\mathcal{J}=\dim T_{\gamma(a)}M$ that $\Theta$ is an isomorphism. Hence there
  exists $\bar J_1\in\mathcal{J}$ with $\bar J_1(0)=0$ and $\bar J_1(a)=V_2$. By an
  analogous argument, there exists a Jacobi field $\bar J_2$ along $\gamma$ with
  $\bar J_2(a)=0$, $\bar J_2(0)=V_1$. The desired field is now given by
  $J=\bar J_1+\bar J_2$. Uniqueness is clear. $\square$
\end{proof}

\begin{corollary}[Corollary 3.10]
  \label{cor:dc-ch5-3-10}
  \dcref{ch5:3.10}
  Let $\gamma:[0,a]\to M$ be a geodesic in $M$, $\dim M=n$, and let
  $\mathcal{J}^\perp$ be the space of Jacobi fields with $J(0)=0$,
  $J'(0)\perp\gamma'(0)$. Let $\{J_1,\dots,J_{n-1}\}$ be a basis of
  $\mathcal{J}^\perp$. If $\gamma(t)$, $t\in(0,a]$, is not conjugate to $\gamma(0)$,
  then $\{J_1(t),\dots,J_{n-1}(t)\}$ is a basis for the orthogonal complement
  $\{\gamma'(t)\}^\perp\subset T_{\gamma(t)}M$ of $\gamma'(t)$.
\end{corollary}
