\chapter{The Sphere Theorem}
\label{chap:dc-sphere-theorem}

Faithful transcription of do Carmo, \emph{Riemannian Geometry}. Each numbered
statement keeps do Carmo's number, kind and (name); proofs keep his explicit
cross-references ("by \cref{prop:dc-ch13-2-2}", "from Rauch's Theorem", "Chap. 11") so
the dependency graph is recoverable from the text.

\section{Introduction}

One of the most beautiful theorems of Global Differential Geometry is the Sphere
Theorem.

\begin{theorem}[Sphere Theorem]
  \label{thm:dc-ch13-1-1}
  \dcref{ch13:1.1}
  Let $M^n$ be a compact simply connected Riemannian manifold, whose sectional
  curvature $K$ satisfies
  
  $$
  0<hK_{\max}<K\leq K_{\max}.\qquad(1)
  $$
  
  Then if $h=1/4$, $M$ is homeomorphic to a sphere.
  
  The number $h$ is called the \emph{pinching} of $M$. Multiplying the metric by a
  constant, we can suppose that $K_{\max}=1$, and (1) can be written, without loss
  of generality, as
  
  $$
  0<h<K\leq 1.\qquad(1')
  $$
  
  The Sphere Theorem was proved for the first time by Rauch [R 1] for
  $h\sim 3/4$. A fundamental contribution was made by Klingenberg [Kℓ 1], who
  introduced into the problem the consideration of the "cut locus" (see Section 2).
  For $M$ of even dimension Klingenberg obtained in [Kℓ 1] an estimate for the
  distance from a point to its cut locus and proved the theorem for $h\sim 0.55$.
  Using Toponogov's theorem and that estimate, Berger [Br 1] obtained the theorem,
  still in even dimension, with $h=1/4$; the use of Toponogov's theorem is
  unnecessary, as shown by Tsukamoto [Ts]. Finally, Klingenberg [Kℓ 2] extended his
  estimate from even to odd dimension, which together with the work of Berger
  [Br 1] yielded the theorem as stated.
  
  In even dimension, the theorem is false if (1') is replaced by
  
  $$
  0<1/4\leq K\leq 1.\qquad(2)
  $$
  
  The complex projective space $P^n(\mathbb{C})$, $n>1$, is simply connected, has a
  metric whose sectional curvature satisfies (2) and is not homeomorphic to a
  sphere (Exercise 12, Chap. 8). Berger [Br 1] proved that if (2) holds then either
  $\operatorname{diam}(M)>\pi$ and $M$ is homeomorphic to a sphere, or
  $\operatorname{diam}(M)=\pi$ and $M$ is isometric to a symmetric space. In the odd
  dimensional case the theorem is still true substituting (1') by (2), but it is not
  known whether this can be improved. For dimensions two and three the theorem holds
  supposing only $h\geq 0$: a compact simply connected $M^n$, $n=2,3$, with positive
  sectional curvature is homeomorphic to $S^n$ (for $n=2$ by Gauss–Bonnet, for
  $n=3$ by a theorem of Hamilton [H]).
  
  This chapter presents the Sphere Theorem without Toponogov's theorem. Section 2
  introduces the "cut locus"; Section 3 estimates the distance from a point to its
  cut locus (using elements of Morse Theory, omissible in even dimension); Section 4
  proves the theorem; Section 5 mentions further developments. Unless otherwise
  stated, the manifolds are complete and the geodesics normalized.
\end{theorem}

\section{The cut locus}

Let $M$ be complete, $p\in M$, and $\gamma:[0,\infty)\to M$ a normalized geodesic
with $\gamma(0)=p$. For small $t>0$, $d(\gamma(0),\gamma(t))=t$, that is,
$\gamma([0,t])$ is minimizing; and if $\gamma([0,t_1])$ is not minimizing the same
holds for all $t>t_1$. By continuity, the set of $t>0$ with
$d(\gamma(0),\gamma(t))=t$ is of the form $[0,t_0]$ or $[0,\infty)$. In the first
case $\gamma(t_0)$ is called the *cut point of $p$ along $\gamma$*; in the second
no cut point exists. The *cut locus of $p$*, denoted $C_m(p)$, is the union of the
cut points of $p$ along all geodesics starting from $p$. If $M$ is compact, a cut
point exists along every geodesic from $p$; the converse holds (Cor. 2.11).

\begin{remark}[Remark 2.1]
  \label{rem:dc-ch13-2-1}
  \dcref{ch13:2.1}
  The cut locus, for surfaces, was introduced as the "ligne de partage" by
  H. Poincaré in 1905 [Po]. For Riemannian manifolds it was introduced by
  J. H. C. Whitehead in 1935 [Wh], who proved several of the properties presented
  here (pp. 700–704 of [Wh]). Interest revived in 1959 when Klingenberg [Kℓ 1]
  showed its usefulness in improving the "pinching" of the sphere theorem.
\end{remark}

\begin{proposition}[characterization of the cut point]
  \label{prop:dc-ch13-2-2}
  \dcref{ch13:2.2}
  \uses{prop:dc-ch13-2-9}
  Suppose that $\gamma(t_0)$ is the cut point of $p=\gamma(0)$ along $\gamma$. Then:
  - (a) either $\gamma(t_0)$ is the first conjugate point of $\gamma(0)$ along
    $\gamma$,
  - (b) or there exists a geodesic $\sigma\neq\gamma$ from $p$ to $\gamma(t_0)$ such
    that $\ell(\sigma)=\ell(\gamma)$.
  
  Conversely, if (a) or (b) holds, then there exists $\tilde t\in(0,t_0]$ such that
  $\gamma(\tilde t)$ is the cut point of $p$ along $\gamma$.
\end{proposition}
\begin{proof}
  Take a sequence $\{t_0+\varepsilon_i\}$, $\varepsilon_i>0$,
  $\varepsilon_i\to 0$, and minimizing geodesics $\sigma_i$ joining $p$ to
  $\gamma(t_0+\varepsilon_i)$. The unit sphere in $T_pM$ being compact, pass to a
  subsequence with $\sigma_i'(0)\to\sigma'(0)$; by continuity $\sigma$ is a
  minimizing geodesic from $p$ to $\gamma(t_0)$, so $\ell(\sigma)=\ell(\gamma)$. If
  $\sigma\neq\gamma$, (b) holds. If $\sigma=\gamma$, we show $d\exp_p$ is singular at
  $t_0\gamma'(0)$: otherwise $\exp_p$ is a diffeomorphism on a neighborhood $U$ of
  $t_0\gamma'(0)$, and writing $\gamma(t_0+\varepsilon_j)=\sigma_j(t_0+\varepsilon_j')$
  with $\varepsilon_j'\leq\varepsilon_j$ gives
  $\exp_p(t_0+\varepsilon_j)\gamma'(0)=\exp_p(t_0+\varepsilon_j')\sigma_j'(0)$, hence
  $\gamma'(0)=\sigma_j'(0)$, contradicting the definition of $t_0$. Conversely, if
  (a) holds, since a geodesic does not minimize after the first conjugate point
  (\cref{cor:dc-ch11-2-9} to the Index Theorem, Chap. 11), the cut point occurs at
  $\gamma(\tilde t)$, $\tilde t\leq t_0$. If (b) holds, joining $\sigma(t_0-\varepsilon)$
  to $\gamma(t_0+\varepsilon)$ by the unique minimizing geodesic $\tau$ inside a
  totally normal neighborhood of $\gamma(t_0)$ yields a curve of length strictly less
  than $t_0+\varepsilon$, so again the cut point occurs at $\gamma(\tilde t)$,
  $\tilde t\leq t_0$. $\square$
\end{proof}

\begin{example}[the sphere]
  \label{ex:dc-ch13-2-3}
  \dcref{ch13:2.3}
  If $M^n=S^n$ and $p\in S^n$, the cut locus of $p$ is its antipodal point. Here the
  cut locus of $p$ coincides with the locus of conjugate points to $p$.
\end{example}

\begin{example}[real projective space]
  \label{ex:dc-ch13-2-4}
  \dcref{ch13:2.4}
  If $M^n=P^n(\mathbb{R})$, obtained by identifying antipodal points of $S^n$, the
  cut locus of $[p,-p]$ is the subset $P^{n-1}(\mathbb{R})\subset P^n(\mathbb{R})$
  obtained by identifying the antipodal points of the "equator" of $p$. The conjugate
  locus of $[p,-p]$ is $[p,-p]$ itself.
\end{example}

\begin{example}[flat torus, cylinder]
  \label{ex:dc-ch13-2-5}
  \dcref{ch13:2.5}
  On the flat torus $T$ obtained by identifying opposite sides of a quadrilateral
  $ABCD$, the cut locus of $A$ is the set formed by the medians of the segments $BA$
  and $BC$; the conjugate locus is empty. Similarly, the cut locus of a point $p$ of
  a cylinder in $\mathbb{R}^3$ is the "opposite generator" to that through $p$.
\end{example}

\begin{remark}[Remark 2.6]
  \label{rem:dc-ch13-2-6}
  \dcref{ch13:2.6}
  \uses{ex:dc-ch13-2-3, ex:dc-ch13-2-4, ex:dc-ch13-2-5, prop:dc-ch13-2-2}
  In \cref{ex:dc-ch13-2-3} the situations (a) and (b) of \cref{prop:dc-ch13-2-2} occur simultaneously;
  in \cref{ex:dc-ch13-2-4} situation (b) occurs before (a); in \cref{ex:dc-ch13-2-5} only situation (b)
  occurs. For the ellipsoid, except for two directions, (b) occurs before (a).
\end{remark}

\subsection{Remark (wiedersehen manifold)}
A compact Riemannian manifold $M$ for which the cut locus $C_m(p)$ of every
$p\in M$ reduces to a single point is called a \emph{wiedersehen manifold}. By a theorem
of L. Green [Gn], wiedersehen surfaces ($\dim M=2$) are isometric to the sphere;
this was generalized to even $\dim M$ by Berger and Kazdan [Be, pp. 236–246] and to
odd $\dim M$ by C. T. Yang [Yg]. See Kazdan [Ka].

\begin{corollary}[symmetry of cut points]
  \label{cor:dc-ch13-2-7}
  \dcref{ch13:2.7}
  \uses{prop:dc-ch13-2-2}
  If $q$ is a cut point of $p$ along $\gamma$, then $p$ is the cut point of $q$ along
  $-\gamma$; in particular $q\in C_m(p)$ if and only if $p\in C_m(q)$.
\end{corollary}
\begin{proof}
  By \cref{prop:dc-ch13-2-2}, either $q$ is conjugate to $p$, or there is a geodesic
  $\sigma\neq\gamma$ from $p$ to $q$ with $\ell(\sigma)=\ell(\gamma)=d(p,q)$. In both
  cases the cut point of $q$ along $-\gamma$ does not occur before $p$; since
  $\ell(-\gamma)=d(p,q)$, $p$ is that cut point. $\square$
\end{proof}

\begin{corollary}[injectivity radius]
  \label{cor:dc-ch13-2-8}
  \dcref{ch13:2.8}
  If $q\in M-C_m(p)$, there exists a unique minimizing geodesic joining $p$ to $q$.
  
  Corollary 2.8 shows that $\exp_p$ is injective on the open ball $B_r(p)$ if and only
  if $r\leq d(p,C_m(p))$. For this reason
  
  $$
  i(M)=\inf_{p\in M}d(p,C_m(p))
  $$
  
  is called the *injectivity radius of $M$*. Corollary 2.8 also shows that
  $M-C_m(p)$ is homeomorphic to an open ball of Euclidean space; in a sense the
  topology of $M$ is contained in its cut locus.
\end{corollary}

\begin{proposition}[continuity of the cut-distance]
  \label{prop:dc-ch13-2-9}
  \dcref{ch13:2.9}
  Define $f:T_1M\to\mathbb{R}\cup\{\infty\}$ by $f(\gamma(0),\gamma'(0))=t_0$ if
  $\gamma(t_0)$ is the cut point of $\gamma(0)$ along $\gamma$, and
  $f(\gamma(0),\gamma'(0))=\infty$ if no cut point exists, where $\mathbb{R}\cup\{\infty\}$
  carries the topology with base the open intervals $(a,b)$ together with the sets
  $(a,\infty]=(a,\infty)\cup\{\infty\}$. Then $f$ is continuous.
\end{proposition}
\begin{proof}
  Let $\gamma_i(0)\to\gamma(0)$, $\gamma_i'(0)\to\gamma'(0)$, with cut points
  $\gamma_i(t_o^i)$ and $\gamma(t_o)$. We show $\lim t_o^i=t_o$. First
  $\limsup t_o^i\leq t_o$: if $t_o<\infty$ and $\varepsilon>0$, only finitely many $j$
  can have $t_o+\varepsilon<t_o^j$, else $d(\gamma_j(0),\gamma_j(t_o+\varepsilon))=t_o+\varepsilon$
  gives in the limit $d(\gamma(0),\gamma(t_o+\varepsilon))=t_o+\varepsilon$,
  contradicting that $\gamma(t_o)$ is a cut point. Now let
  $\bar t=\liminf t_o^i\leq t_o$; suppose $\bar t<\infty$ and pass to a subsequence
  $t_o^j\to\bar t$. If for a subsequence the $\gamma_j(t_o^j)$ are conjugate to
  $\gamma_j(0)$, then (accumulation of conjugate points is conjugate) $\gamma(\bar t)$
  is conjugate to $\gamma(0)$, so $\bar t\geq t_o$. Otherwise, by \cref{prop:dc-ch13-2-2},
  geodesics $\sigma_j\neq\gamma_j$ exist with $\sigma_j(0)=\gamma_j(0)$,
  $\sigma_j(t_o^j)=\gamma_j(t_o^j)$, $\ell(\sigma_j)=\ell(\gamma_j)$; a subsequence
  $\sigma_j\to\sigma$ joins $\gamma(0)$ to $\gamma(\bar t)$. If $\sigma\neq\gamma$,
  \cref{prop:dc-ch13-2-2} gives $t_o\leq\bar t$; if $\sigma=\gamma$, a similar argument shows
  $\gamma(\bar t)$ conjugate to $\gamma(0)$, so $\bar t\geq t_o$. $\square$
\end{proof}

\begin{corollary}[the cut locus is closed]
  \label{cor:dc-ch13-2-10}
  \dcref{ch13:2.10}
  \uses{prop:dc-ch13-2-9}
  For all $p\in M$, $C_m(p)$ is closed; in particular, if $M$ is compact, $C_m(p)$ is
  compact.
\end{corollary}
\begin{proof}
  One verifies $C_m(p)=\{\gamma(t);t=f(p,\gamma'(0))\}$, where $f$ is the
  function of \cref{prop:dc-ch13-2-9}. If $q$ is an accumulation point of $C_m(p)$, take
  $\gamma_j(t_j)\to q$ with $t_j=f(p,\gamma_j'(0))$ and $\gamma_j'(0)\to v$. With
  $\gamma(0)=p$, $\gamma'(0)=v$, continuity of $f$ gives
  $q=\exp_p(f(p,\gamma'(0))\gamma'(0))=\gamma(f(p,\gamma'(0)))\in C_m(p)$. $\square$
\end{proof}

\begin{corollary}[cut point everywhere implies compact]
  \label{cor:dc-ch13-2-11}
  \dcref{ch13:2.11}
  Suppose $M$ is complete and there exists $p\in M$ which has a cut point for every
  geodesic starting from $p$. Then $M$ is compact.
\end{corollary}
\begin{proof}
  Since $M=\bigcup\{\gamma(t);t\leq f(p,\gamma'(0))\}$ over all geodesics
  $\gamma$ with $\gamma(0)=p$, and $f$ is continuous and (by hypothesis) bounded, $M$
  is bounded; the result follows from the Hopf–Rinow Theorem. $\square$
\end{proof}

\begin{proposition}[geometry of a nearest cut point]
  \label{prop:dc-ch13-2-12}
  \dcref{ch13:2.12}
  \uses{prop:dc-ch13-2-2}
  Let $p\in M$ and suppose there exists $q\in C_m(p)$ realizing the distance from $p$
  to $C_m(p)$. Then:
  - (a) either there exists a minimizing geodesic $\gamma$ from $p$ to $q$ along which
    $q$ is conjugate to $p$,
  - (b) or there exist exactly two minimizing geodesics $\gamma$ and $\sigma$ from $p$
    to $q$, with $\gamma'(\ell)=-\sigma'(\ell)$, $\ell=d(p,q)$.
\end{proposition}
\begin{proof}
  Let $\gamma$ be minimizing from $p$ to $q$. By \cref{prop:dc-ch13-2-2} either $q$
  is conjugate to $p$ along $\gamma$ — case (a) — or there is another minimizing
  $\sigma\neq\gamma$ with $\ell(\sigma)=\ell(\gamma)$. Suppose $q$ is not conjugate to
  $p$ along $\gamma$ or $\sigma$, and $\gamma'(\ell)\neq-\sigma'(\ell)$; we derive a
  contradiction (which also shows there can be only two such geodesics). Choose
  $V\in T_qM$ with $\langle V,\gamma'(\ell)\rangle<0$ and
  $\langle V,\sigma'(\ell)\rangle<0$. Since $q$ is not conjugate along $\gamma$, build
  a variation $\gamma_s$ of $\gamma$ with
  $\frac{d}{ds}\ell(\gamma_s)|_{s=0}=\langle V,\gamma'(\ell)\rangle<0$ by the formula
  for the first variation, and similarly $\sigma_s$ of $\sigma$ with
  $\frac{d}{ds}\ell(\sigma_s)|_{s=0}=\langle V,\sigma'(\ell)\rangle<0$ and
  $\sigma_s(\ell)=\tau(s)$. For small $s>0$, $\ell(\gamma_s)<\ell(\gamma)$ and
  $\ell(\sigma_s)<\ell(\sigma)$. If $\ell(\gamma_s)=\ell(\sigma_s)$, by \cref{prop:dc-ch13-2-2}
  $\tau(s)=\gamma_s(\ell)$ is a cut point with
  $d(p,\gamma_s(\ell))=\ell(\gamma_s)<d(p,C_m(p))$, a contradiction; if
  $\ell(\gamma_s)<\ell(\sigma_s)$, $\sigma_s$ is not minimizing so a cut point
  $\sigma_s(\tilde t)$, $\tilde t<\ell$, contradicts again; the case
  $\ell(\gamma_s)>\ell(\sigma_s)$ is similar. $\square$
\end{proof}

\begin{proposition}[injectivity radius vs. closed geodesics]
  \label{prop:dc-ch13-2-13}
  \dcref{ch13:2.13}
  \uses{prop:dc-ch13-2-9, prop:dc-ch10-2-4, prop:dc-ch13-2-12}
  If the sectional curvature $K$ of a complete Riemannian manifold $M$ satisfies
  $0<K_{\min}\leq K\leq K_{\max}$, then:
  - (a) $i(M)\geq\pi/\sqrt{K_{\max}}$, or
  - (b) there exists a closed geodesic $\gamma$ in $M$, whose length is less than that
    of any other closed geodesic in $M$, with $i(M)=\tfrac{1}{2}\ell(\gamma)$.
\end{proposition}
\begin{proof}
  By Bonnet–Myers (Theor. 3.1 of Chap. 9), $M$ is compact. Since $T_1M$ is
  compact, \cref{prop:dc-ch13-2-9} gives $p\in M$ with $d(p,C_m(p))=\inf_{r\in M}d(r,C_m(r))$,
  and $C_m(p)$ compact gives $q\in C_m(p)$ realizing this distance. If $q$ is conjugate
  to $p$, then $d(p,q)\geq\pi/\sqrt{K_{\max}}$ by \cref{prop:dc-ch10-2-4} — case
  (a). If $q$ is not conjugate to $p$, by \cref{prop:dc-ch13-2-12} there are two minimizing
  geodesics $\mu,\sigma$ from $p$ to $q$ with $\mu'(\ell)=-\sigma'(\ell)$,
  $\ell=d(p,q)$. Since $q\in C_m(p)$ we have $p\in C_m(q)$ and $p$ realizes the
  distance from $q$ to $C_m(q)$, so $\mu'(0)=-\sigma'(0)$; thus $\mu$ and $\sigma$ form
  a closed geodesic $\gamma$ satisfying (b). $\square$
\end{proof}

\section{The estimate of the injectivity radius}

The next proposition is a crucial step in the Sphere Theorem; its proof uses Morse
Theory. A reader who wishes only the even-dimensional Sphere Theorem can omit it and
pass to \cref{prop:dc-ch13-3-4}.

\begin{proposition}[Klingenberg [Kℓ 2]]
  \label{prop:dc-ch13-3-1}
  \dcref{ch13:3.1}
  Let $M^n$, $n\geq 3$, be a simply connected, compact Riemannian manifold with
  $1/4<K\leq 1$. Then $i(M)\geq\pi$.
\end{proposition}

\begin{lemma}[index comparison]
  \label{lem:dc-ch13-3-2}
  \dcref{ch13:3.2}
  Let $M$ and $\bar M$ be Riemannian manifolds of the same dimension with sectional
  curvatures satisfying $\sup\bar K\leq\inf K$. Consider a geodesic
  $\sigma:[0,\ell]\to M$, $\sigma(0)=p$, fix $\bar p\in\bar M$, let
  $i:T_pM\to T_{\bar p}\bar M$ be a linear isometry, and set
  $\bar\sigma(t)=\exp_{\bar p}t(i\sigma'(0))$. Then
  $\operatorname{index}\sigma\geq\operatorname{index}\bar\sigma$.
\end{lemma}
\begin{proof}
  For a piecewise differentiable field $\bar W$ along $\bar\sigma$, define
  $W$ along $\sigma$ by $W(t)=P_t\circ i^{-1}\circ\bar P_t^{-1}(\bar W(t))$ (parallel
  transports). Then $\langle W,\sigma'\rangle=\langle\bar W,\bar\sigma'\rangle$,
  $|W(t)|=|\bar W(t)|$, $\langle W',W'\rangle=\langle\bar W',\bar W'\rangle$. Since
  $\sup\bar K\leq\inf K$, $I(W,W)\leq I(\bar W,\bar W)$; so $I(\bar W,\bar W)<0$ forces
  $I(W,W)<0$. $\square$
\end{proof}

For the next lemma we need the following facts. Let $f:M^n\to\mathbb{R}$ be a
differentiable function on a differentiable manifold $M$. A point $p\in M$ is a
critical point of $f$ if $df(p)=0$; $f(p)$ is then called a critical value of $f$.
If $p$ is a critical point and $(x_1,\dots,x_n)$ is a coordinate system on $M$
around $p$, the hessian of $f$, defined by the matrix
$(\partial^2f/\partial x_i\partial x_j)(p)$, represents a symmetric bilinear form
on $T_pM$ that does not depend on the chosen system of coordinates. We say that
the critical point is non-degenerate if the determinant of this matrix is
different from zero. Such a critical point is isolated, and it is possible to
choose a coordinate system $(x_1,\dots,x_{n-\lambda},y_1,\dots,y_\lambda)$ in such
a way that, in the coordinate neighborhood considered,

$$
f(x_1,\dots,x_{n-\lambda},y_1,\dots,y_\lambda)
=f(p)+x_1^2+\cdots+x_{n-\lambda}^2-y_1^2-\cdots-y_\lambda^2.
$$

The integer $\lambda$ is called the index of the non-degenerate critical point
$p$. (For more details see Milnor [Mi].)

\begin{lemma}[Morse-theoretic deformation]
  \label{lem:dc-ch13-3-3}
  \dcref{ch13:3.3}
  \uses{prop:dc-ch13-3-1, prop:dc-ch13-2-13, prop:dc-ch13-2-12, lem:dc-ch13-3-2, cor:dc-ch13-2-10}
  Let $M^n$ be a differentiable manifold and $f:M^n\to\mathbb{R}$ a function whose
  critical points are all non-degenerate. Given $p,q\in M$ and a differentiable curve
  $\gamma:[0,1]\to M$ with $\gamma(0)=p$, $\gamma(1)=q$, let $a=\max\{f(p),f(q)\}$,
  $M^a=\{x\in M;f(x)\leq a\}$, and $b=\max_{t\in[0,1]}(f\circ\gamma(t))$. Assume
  $f^{-1}([a,b])$ is compact and contains no critical points of index zero or one.
  Then for all $\delta>0$, $\gamma$ is homotopic, keeping endpoints fixed, to a curve
  $\bar\gamma$ with $\bar\gamma([0,1])\subset M^{a+\delta}$.
\end{lemma}
\begin{proof}
  (Sketch, after Milnor [Mi].) If $f^{-1}([a_1,a_2])$ is compact and
  critical-point-free, the gradient flow of $f$ gives a deformation retract
  $M^{a_2}\to M^{a_1}$. Let $p_1,\dots,p_k$ be the critical points in $M^b-M^a$ with
  distinct values $b\geq c_1>\dots>c_k>a$. At a critical value $c_1$ with index
  $\lambda\geq 2$, write $f=c_1+x_1^2+\dots+x_{n-\lambda}^2-y_1^2-\dots-y_\lambda^2$ on
  a neighborhood $H$, retract $M^{c_1+\varepsilon}\to M^{c_1-\varepsilon}\cup H$, and
  let $L=\{y_1=\dots=y_\lambda=0\}$. Since
  $\dim\bar\gamma_1'+\dim L\leq 1+n-2<\dim M$, transversality lets one perturb
  $\bar\gamma_1$ off $L$ within $H$ fixing endpoints; the gradient flow then retracts
  into $M^{c_1-\varepsilon}$. Proceeding inductively yields
  $\bar\gamma\subset M^{a+\delta}$. $\square$
  
  (Remark after Lemma 3.3: if there are critical points of index zero or one in
  $f^{-1}([a,b])$ and $c$ is the largest value of such critical points, then
  $\gamma$ is homotopic to $\bar\gamma$ with
  $\bar\gamma([0,1])\subset M^{c+\delta}$.)
  
  \emph{Proof of \cref{prop:dc-ch13-3-1}.} Suppose $i(M)<\pi$. By \cref{prop:dc-ch13-2-13} there is a
  closed geodesic $\gamma$ of length $\ell=\ell(\gamma)<2\pi$. By \cref{cor:dc-ch11-2-10} of
  Chap. 11 the conjugate points to $\gamma(0)$ along $\gamma$ are discrete; choose
  $\varepsilon>0$ with: (1) $\gamma(\ell-\varepsilon)$ not conjugate to $p=\gamma(0)$;
  (2) $\exp_p$ a diffeomorphism on $B_{2\varepsilon}(p)$;
  (3) $3\varepsilon<2\pi-\pi/\sqrt{\bar K}$, $\bar K=\inf_M K$;
  (4) $3\varepsilon<2\pi-\ell$; (5) $5\varepsilon<2\pi$. By Sard's Theorem pick a
  regular value $q\in B_\varepsilon(\gamma(\ell-\varepsilon))$ of $\exp_p$, with
  $q=\gamma_1(t)$ a geodesic from $p$, $3\varepsilon<\ell(\gamma_1)<\ell$. A
  minimizing geodesic $\gamma_o$ from $p$ to $q$ has
  $\ell(\gamma_o)\leq d(p,\gamma(\ell-\varepsilon))+d(\gamma(\ell-\varepsilon),q)\leq 2\varepsilon$,
  so $\gamma_o\neq\gamma_1$.
  
  Consider $\Omega_{p,q}^c$ (\cref{rem:dc-ch11-2-12} of Chap. 11) and its finite-dimensional
  approximation $B$. Since $q$ is a regular value, all critical points in
  $\Omega_{p,q}^c$ are non-degenerate. As $M$ is simply connected, there is a homotopy
  $\gamma_s$ between $\gamma_o$ and $\gamma_1$ fixing $p,q$; deforming into $B$ and
  applying the remark after Lemma 3.3, for all $\delta>0$ the larger-length curve
  $\bar\gamma$ of the deformed $\bar\gamma_s$ satisfies $\ell(\bar\gamma)<a+\delta$,
  $a=\max\{\ell(\gamma_o),\ell(\gamma_1),\ell(\sigma)\}$, with $\sigma$ the largest
  geodesic of index $<2$. Put $\delta=\varepsilon$. Here $\ell(\gamma_o)\leq2\varepsilon$
  and $\ell(\gamma_1)<\ell<2\pi-3\varepsilon$ by (4). To estimate $\ell(\sigma)$, apply
  \cref{lem:dc-ch13-3-2} with $\bar M$ a sphere of curvature $\bar K=\inf_M K>1/4$, getting
  $\operatorname{index}\bar\sigma\leq\operatorname{index}\sigma<2$. Since $\bar\sigma$
  is a geodesic on $S^n$, the Morse Index Theorem gives: if
  $\ell(\sigma)>\pi/\sqrt{\bar K}$ then $\operatorname{index}\bar\sigma\geq2$;
  therefore by (3) $\ell(\sigma)\leq\pi/\sqrt{\bar K}<2\pi-3\varepsilon$. Hence by (5)
  $\ell(\bar\gamma)<a+\varepsilon<2\pi-3\varepsilon+\varepsilon=2\pi-2\varepsilon$. On
  the other hand, the Lemma of Klingenberg (Exercise 1 of Chap. 10) gives in the
  homotopy a curve $\bar\gamma_{s_o}$ with $\ell(\gamma_o)+\ell(\bar\gamma_{s_o})\geq2\pi$,
  so $\ell(\bar\gamma)\geq2\pi-\ell(\gamma_o)\geq2\pi-2\varepsilon$, contradicting
  $\ell(\bar\gamma)<2\pi-2\varepsilon$. Thus $i(M)<\pi$ is untenable. $\square$
\end{proof}

\begin{proposition}[even-dimensional estimate]
  \label{prop:dc-ch13-3-4}
  \dcref{ch13:3.4}
  \uses{prop:dc-ch13-2-12, ex:dc-ch13-2-4}
  If the sectional curvature $K$ of a compact orientable even-dimensional Riemannian
  manifold $M^n$ satisfies $0<K\leq 1$, then $i(M)\geq\pi$.
\end{proposition}
\begin{proof}
  By compactness there are $p,q\in M$ with $q\in C_m(p)$ and $d(p,q)=i(M)$.
  Suppose $d(p,q)<\pi$. If $p$ is conjugate to $q$, then $d(p,q)\geq\pi$ by
  \cref{prop:dc-ch10-2-4} of Chap. 10, a contradiction. So $p$ is not conjugate to $q$; by
  Propositions 2.12 and 2.13 there is a closed geodesic $\gamma$ through $p=\gamma(0)$
  and $q$ with $\ell(\gamma)<2\pi$. Parallel transport along $\gamma$ leaves invariant
  a vector $V$ orthogonal to $\gamma$ (Lemma 3.8 of Chap. 9); computing the second
  variation $E_V''(0)$ for this $V$, it is strictly negative (here even dimension is
  crucial). Hence there is a variation through regular closed curves $\gamma_s$ with
  $\ell(\gamma_s)<\ell(\gamma)$ for $s\neq0$. Let $q_s$ be a point of $\gamma_s$ at
  maximum distance from $\gamma_s(0)$; the minimizing geodesic $\sigma_s$ from
  $q_s=\sigma_s(0)$ to $\gamma_s(0)$ has $\sigma_s(0)\to q$, and an accumulation point
  $w$ of $\sigma_s'(0)$ gives a minimizing geodesic $\sigma(t)=\exp_q tw$ from $q$ to
  $p$. By the first-variation formula $\sigma_s'(0)\perp\gamma_s'$ at $q_s$, so
  $\sigma'(0)\perp\gamma'$ at $q$. This yields three minimizing geodesics from $p$ to
  $q$; since $p$ is not conjugate to $q$, this contradicts \cref{prop:dc-ch13-2-12}. $\square$
\end{proof}

\begin{remark}[Remark 3.5]
  \label{rem:dc-ch13-3-5}
  \dcref{ch13:3.5}
  \uses{prop:dc-ch13-3-4}
  By Synge's theorem (Chap. 9, Cor. 3.10) the manifolds satisfying the hypotheses of
  \cref{prop:dc-ch13-3-4} are simply connected. Conversely a simply connected manifold is
  orientable (else it has a two-sheeted covering, Exercise 12 of Chap. 0,
  contradicting simple connectivity). Hence the orientability hypothesis in
  \cref{prop:dc-ch13-3-4} is equivalent to assuming $M$ simply connected.
\end{remark}

\section{The Sphere Theorem}

In what follows, whether $\dim M$ is even or odd plays no role.

\begin{lemma}[Berger [Br 1]]
  \label{lem:dc-ch13-4-1}
  \dcref{ch13:4.1}
  Let $M$ be a compact Riemannian manifold and $p,q\in M$ with
  $d(p,q)=\operatorname{diam}M$. Then for all $W\in T_pM$ there exists a minimizing
  geodesic $\gamma$ from $p=\gamma(0)$ to $q$ with $\langle\gamma'(0),W\rangle\geq0$.
\end{lemma}
\begin{proof}
  Let $\lambda(t)=\exp_p tW$ and let $\gamma_t$ be a minimizing geodesic with
  $\gamma_t(0)=\lambda(t)$, $\gamma_t(\ell(\gamma_t))=q$. If for each $n$ there is
  $t_n\in[0,1/n]$ with $\langle\gamma_{t_n}'(0),\lambda'(t_n)\rangle\geq0$, a
  subsequence $\gamma_{t_n}\to\gamma$ gives
  $0\leq\langle\gamma'(0),\lambda'(0)\rangle=\langle\gamma'(0),W\rangle$. Otherwise
  some $n$ has $\langle\gamma_t'(0),\lambda'(t)\rangle<0$ for all $t\in[0,1/n]$. Taking
  a totally normal neighborhood $U$ of $\lambda(t)$, $q_o\in U$ on $\gamma_t$, and
  $\sigma_s$ minimizing from $q_o$ to $\lambda(s)$, the first variation of energy gives
  $\tfrac{1}{2}\frac{d}{ds}E(\sigma_s)|_{s=t}=-\langle\gamma_t'(0),\lambda'(t)\rangle>0$,
  so for $s<t$, $d(q_o,\lambda(s))<d(q_o,\lambda(t))$, hence
  $d(q,\lambda(s))<d(q,\lambda(t))$; but $p$ is at maximum distance from $q$, so
  $d(q,\lambda(t))\leq d(q,\lambda(0))$ — a contradiction. $\square$
\end{proof}

\begin{lemma}[covering by two balls; Tsukamoto [Ts]]
  \label{lem:dc-ch13-4-2}
  \dcref{ch13:4.2}
  \uses{ex:dc-ch13-2-5}
  Let $M^n$ be a compact, simply connected Riemannian manifold whose sectional
  curvature satisfies $\tfrac14<\delta\leq K\leq 1$, and let $p,q\in M$ with
  $d(p,q)=\operatorname{diam}(M)$. Then
  
  $$
  M=B_\rho(p)\cup B_\rho(q),
  $$
  
  where $B_r(p)$ denotes the open geodesic ball of radius $r$ and center $p$, and
  $\rho$ is such that $\pi/(2\sqrt\delta)<\rho<\pi$.
\end{lemma}
\begin{proof}
  By the injectivity-radius estimate of Section 3, $B_\rho(p)$ contains no
  points of $C_m(p)$, so it is diffeomorphic via $\exp$ to a Euclidean ball, and
  likewise $B_\rho(q)$. (Here $i(M)\geq\pi$ enters crucially.) Suppose some $r\in M$
  has $d(p,r)\geq\rho$ and $d(q,r)\geq\rho$, with $d(p,r)\geq d(q,r)\geq\rho$. A
  minimizing geodesic from $q$ to $r$ meets $\partial B_\rho(q)$ at a point
  $q'\notin B_\rho(p)$ (else $d(r,q')>d(r,B_\rho(p))\geq d(r,B_\rho(q))=d(r,q')$,
  absurd). By Bonnet–Myers, $\operatorname{diam}(M)\leq\pi/\sqrt\delta<2\rho$, so the
  intersection $q''$ of the minimizing geodesic from $q$ to $p$ with $\partial B_\rho(q)$
  satisfies $d(p,q'')=d(p,q)-d(q,q'')<2\rho-\rho=\rho$, i.e. $q''\in B_\rho(p)$. As
  $\partial B_\rho(q)$ is path connected, $\partial B_\rho(p)\cap\partial B_\rho(q)\neq\emptyset$,
  giving $r_o$ with $d(r_o,p)=d(r_o,q)=\rho$. Take $\lambda$ minimizing from $p$ to
  $r_o$; by Berger's Lemma there is $\gamma$ minimizing from $p$ to $q$ with
  $\langle\gamma'(0),\lambda'(0)\rangle\geq0$. Let $s\in\gamma$ with $d(p,s)=\rho$.
  Applying Rauch's Theorem (\cref{prop:dc-ch10-2-5} of Chap. 10), comparing $M^n$ with a
  sphere $S^n$ of curvature $\delta$: since the angle
  $\sphericalangle r_o p s\leq\pi/2$, $d(r_o,s)\leq\pi/(2\sqrt\delta)$. As
  $d(r_o,p)=d(r_o,q)=\rho$ and some $s$ has $d(r_o,s)<\rho$, the distance from $r_o$ to
  $\gamma$ is realized by an interior point $s_o$, with the minimizing geodesic from
  $r_o$ to $s_o$ orthogonal to $\gamma$ and
  $d(r_o,\gamma)=d(r_o,s_o)\leq\pi/(2\sqrt\delta)$. Since $d(p,q)\leq\pi/\sqrt\delta$,
  either $d(p,s_o)\leq\pi/(2\sqrt\delta)$ or $d(q,s_o)\leq\pi/(2\sqrt\delta)$; in the
  former case, since $d(r_o,s_o)\leq\pi/(2\sqrt\delta)$ and
  $\sphericalangle p s_o r_o=\pi/2$, Rauch gives $d(p,r_o)\leq\pi/(2\sqrt\delta)<\rho$,
  contradicting $d(p,r_o)=\rho$ (the other case is analogous). $\square$
\end{proof}

\begin{lemma}[the equator]
  \label{lem:dc-ch13-4-3}
  \dcref{ch13:4.3}
  \uses{lem:dc-ch13-4-2}
  Under the conditions of \cref{lem:dc-ch13-4-2}, on each geodesic of length $\rho$ starting from
  $p$ there exists a unique point $m$ such that $d(p,m)=d(q,m)<\rho$. Similarly, on
  each geodesic of length $\rho$ starting from $q$ there exists a unique point $n$
  equidistant from $p$ and $q$.
\end{lemma}
\begin{proof}
  For a geodesic $\gamma$ with $\gamma(0)=p$, set
  $f(s)=d(q,\gamma(s))-d(p,\gamma(s))$. Then $f$ is continuous and $f(0)=d(q,p)>0$. If
  $\gamma(s_o)$ is the cut point of $p$ along $\gamma$, the estimate of Klingenberg
  (Prop. 3.1, or 3.4 in even dimension) gives $d(p,\gamma(s_o))\geq\pi>\rho$, and
  \cref{lem:dc-ch13-4-2} gives $d(q,\gamma(s_o))<\rho$, so $f(s_o)<0$. Hence $f(s_1)=0$ for some
  $s_1\in(0,s_o)$, giving $m=\gamma(s_1)$. For uniqueness, suppose $m_1\neq m_2$ both
  equidistant with $m_1$ between $p$ and $m_2$; then
  $d(q,m_2)=d(p,m_2)=d(p,m_1)+d(m_1,m_2)=d(q,m_1)+d(m_1,m_2)$, so the unique minimizing
  geodesic $\sigma_1$ from $q$ to $m_1$ coincides with $\gamma$, forcing $q\in\gamma$
  and then $p=q$, a contradiction. $\square$
\end{proof}

\begin{remark}[Remark 4.4]
  \label{rem:dc-ch13-4-4}
  \dcref{ch13:4.4}
  \uses{thm:dc-ch13-1-1, lem:dc-ch13-4-2, lem:dc-ch13-4-3}
  By its uniqueness, the point $m$ depends continuously on the (initial direction of
  the) geodesic which contains it.
  
  \emph{Proof of \cref{thm:dc-ch13-1-1}.} Let $p,q\in M$ with $\operatorname{diam}M=d(p,q)$. Let $D_1$,
  $D_2$ be the subsets of $M$ formed by all geodesic segments $\overline{pm}$ and
  $\overline{qn}$ respectively, with $m,n$ from \cref{lem:dc-ch13-4-3}. By continuity of $m$ (Rem.
  4.4), $D_1,D_2$ are closed; we claim $D_1\cup D_2=M$ and
  $\partial D_1=\partial D_2=D_1\cap D_2$. Indeed, for $r\in M$, by \cref{lem:dc-ch13-4-2} either
  $d(p,r)<\rho$ or $d(q,r)<\rho$; in the first case, since $d(p,C_m(p))\geq\pi>\rho$,
  there is a unique minimizing $\gamma$ through $p,r$ and a unique $m$ on $\gamma$ with
  $d(p,m)=d(q,m)<\rho$, and $r\in\overline{pm}=D_1$ if $d(p,r)<d(q,r)$, $r\in\partial D_1$
  if $d(p,r)=d(q,r)$ (then $r=m$), and $r\in D_2$ if $d(p,r)>d(q,r)$; so
  $r\in D_1\cup D_2$, and $r\in D_1\cap D_2$ forces $d(p,r)=d(q,r)$, $r=m=n$.
  
  Define $\varphi:S^n\to M^n$ as follows. Associate to a fixed $N\in S^n$ the point
  $p$, and to its antipode $S$ the point $q$. Choose a linear isometry
  $i:T_NS^n\to T_pM$. For each $e$ in the equator $E$ of $S^n$ relative to $N$, take
  the geodesic $\gamma(s)$ of $S^n$, $0\leq s\leq\pi$, with $\gamma(0)=N$,
  $\gamma(\pi/2)=e$; let $m$ be the point of \cref{lem:dc-ch13-4-3} on the geodesic of $M$ through
  $p$ with velocity $i\gamma'(0)$. Define
  
  $$
  \varphi(\gamma(s))=\exp_p\!\Big(s\tfrac{2}{\pi}d(p,m)\,(i\circ\gamma'(0))\Big),\quad 0<s\leq\tfrac{\pi}{2},
  $$
  
  $$
  \varphi(\gamma(s))=\exp_q\!\Big(\big(2-\tfrac{2s}{\pi}\big)d(q,m)\,V\Big),\quad\tfrac{\pi}{2}\leq s<\pi,
  $$
  
  where $V$ is the unit tangent at $q$ of the unique minimizing geodesic from $q$ to
  $m$. Then $\varphi$ maps the closed northern hemisphere bijectively onto $D_1$, the
  closed southern hemisphere bijectively onto $D_2$, and $E$ onto
  $\partial D_1=\partial D_2=D_1\cap D_2$. By uniqueness of $m,n$ (\cref{lem:dc-ch13-4-3}),
  $\varphi$ is continuous; since $M=D_1\cup D_2$, $\varphi$ is surjective; since
  $D_1\cap D_2=\partial D_1=\varphi(E)$, $\varphi$ is injective. As $S^n$ is compact,
  $\varphi$ is a homeomorphism. $\square$
\end{remark}

\section{Some further developments}

Survey section; no numbered statements. The Sphere Theorem has produced an intense
and continuing research activity. The main lines of development mentioned by do
Carmo are:

- \textbf{Homeomorphism vs. diffeomorphism.} The homeomorphism here glues two discs along
  their boundaries, which may produce a differentiable structure on $M$ distinct from
  the standard sphere; so this proof does not give a diffeomorphism. In 1966 Gromoll
  [Gℓ] and Calabi (unpublished) proved a \emph{differentiable} sphere theorem for a
  pinching $h_n$ sufficiently small (depending on $n$); $h_n$ has since been improved
  ([Sh], [SSK]), the best estimate at the time being $\lim h_n\sim0.68$ as
  $n\to\infty$ [IHR].

- **The case $h<1/4$ in even dimension.** Following Rauch [R2], one compares $M$ with
  a symmetric space (compact, simply connected, rank one); after results of do Carmo
  [dC 1], Klingenberg [Kl 3], Kobayashi [Ko], Cheeger [Cr 1], a satisfactory theorem
  was obtained by Min-Oo and Ruh [MR]. Berger [Br 3] showed that for $h=(1/4)-\varepsilon$
  a compact simply connected $h$-pinched manifold is still homeomorphic to a sphere or
  one of those symmetric spaces (no estimate for $\varepsilon$ is known).

- **The odd-dimensional case $h<1/4$** appears difficult: the estimate $i(M)\geq\pi$
  (valid in even dimension for any $h$ by Prop. 3.4) cannot hold in odd dimension
  without restriction on $h$. By an example of Berger (Cheeger–Ebin [CE] pp. 70–71)
  the best one can hope for is $h=1/9$, not yet proved.

- \textbf{Diameter sphere theorem (Grove–Shiohama [GS], 1977).} If $M$ is compact,
  $K\geq1$, and $\operatorname{diam}(M)>\pi/2$, then $M$ is homeomorphic to a sphere.
  The case $\operatorname{diam}(M)=\pi/2$ (where it fails, e.g. real projective space)
  was essentially classified by Gromoll and Grove [Gℓ-G].

- \textbf{Almost-constant negative curvature (Gromov [Gv 0], 1978).} If
  $-1-\varepsilon\leq K\leq-1$ and the diameter is bounded above by $D$, then for
  $\varepsilon=\varepsilon(n,D)$ small, $M$ admits a metric of constant curvature
  $-1$.

- \textbf{Betti number bound (Gromov [Gv 2]).} For compact connected manifolds of
  non-negative curvature there is a constant $C(n)$ with $\sum_i b_i\leq C(n)$ (Betti
  numbers over any field); hence connected sums of sufficiently many copies of
  products $S^p\times S^{n-p}$ admit no non-negatively curved metric.

- **Finiteness and compactness (Cheeger [Cr 2]; Weinstein [We 1], Peters [Pe]; Gromov
  [Gv 1]).** For fixed dimension there are finitely many homeomorphism types among
  compact Riemannian manifolds with volume bounded below (or injectivity radius
  bounded below), diameter bounded above, and curvature bounded in absolute value; the
  set of such manifolds has compact closure in a suitable topology, with applications
  including Berger's $(\tfrac14-\varepsilon)$-theorem. See Pansu's Bourbaki seminar
  [Pa], the survey of Berger [Br 3] on the role of the Sphere Theorem, the references
  in Sakai [Sa], and another proof of the sphere theorem (after Gromov) in Eschenburg
  [E].
