% Complete bibliography from the Cao--Zhu source.
\providecommand{\bysame}{\leavevmode\hbox to 3em{\hrulefill}\thinspace}
\begin{thebibliography}{999}

\bibitem{AM}U. Abresch and W.T. Meyer,
{\it Injectivity radius estimates and sphere theorems,} Comparison
Geometry, MSRI Publications, \textbf{30} (1997), 1--48.

\bibitem{AW}S. Aloff and N. Wallach,
{\it An infinite family of distinct $7$-manifolds admitting
positively curved Riemannian structure}, Bull. Amer. Math. Soc.,
\textbf{81} (1975), 93--97.

\bibitem{As53}W. Ambrose and I.M. Singer,
{\it A theorem on holonomy,} Trans. Amer. Math. Soc., \textbf{75}
(1953), 428--443.

\bibitem{An}M. Anderson,
{\it Geometrization of $3$-manifolds via the Ricci flow}, Notices
Amer. Math. Soc., {\bf 51}({\bf 2}) (2004), 184--193.

\bibitem{Bando}S. Bando,
{\it On the classification of three-dimensional compact K\"ahler
manifolds of nonnegative bisectional curvature,} J. Differential
Geom., \textbf{19}({\bf 2}) (1984), 283--297.

\bibitem{BSY}J. Bartz, M. Struwe, and R. Ye,
{\it A new approach to the Ricci flow on $S\sp 2$}, Ann. Scuola
Norm. Sup. Pisa Cl. Sci. (4), \textbf{21}({\bf 3}) (1994),
475--482.

\bibitem{Ber}M. Berger,
{\it Sur les groupes d'holonomie des vari\'{e}t\'{e}s \`{a}
connexion affine et les vari\'{e}t\'{e}s riemanniennes}, Bull.
Soc. Math. France, \textbf{83} (1955), 279-330.

\bibitem{BW}   C. B$\ddot{o}$hm and B. Wilking,  {\it Manifolds with positive curvature
operators are space forms}, math.DG/0606187.

\bibitem{BGP}Y. Burago, M. Gromov, and G. Perelman,
{\it A.D. Alexandrov spaces with curvatures bounded below},
Russian Math. Surveys, {\bf 47} (1992), 1--58.

\bibitem{BBI}D. Burago, Y. Burago, and S. Ivanov,
{\it A Course in Metric Geometry,} Graduate Studies in
Mathematics, \textbf{33}, Amer. Math. Soc. Providence, Rhode
Island.

\bibitem{BR}R. Bryant,
{\it Local existence of gradient Ricci solitons}, preprint.

\bibitem{Cao85}H.-D. Cao,
{\it Deformation of K\"{a}hler metrics to K\"{a}hler-Einstein
metrics on compact K\"{a}hler manifolds}, Invent. Math.,
\textbf{81}({\bf 2}) (1985), 359--372.

\bibitem{Cao92} \bysame,
%Cao, H. D.,
{\it On Harnack's inequalities for the K$\ddot{a}$hler-Ricci
flow,} Invent. Math., \textbf{109}({\bf 2}) (1992), 247--263.

\bibitem{Cao94} \bysame,
%Cao, H. D.,
{\it Existence of gradient K\"{a}hler-Ricci solitons, Elliptic and
Parabolic Methods in Geometry} (Minneapolis, MN, 1994), 1--16, A K
Peters, Wellesley, MA, 1996.

\bibitem{Cao97} \bysame,
%Cao, H. D.,
{\it Limits of Solutions to the K\"{a}hler-Ricci flow}, J.
Differential Geom., \textbf{45 } (1997), 257--272.

\bibitem{CCZ}H.-D. Cao, B.L. Chen, and X.P. Zhu,
{\it Ricci flow on compact K\"{a}hler manifolds of positive
bisectional curvature,} C.R. Math. Acad. Sci. Paris,
\textbf{337}({\bf 12}) (2003), 781--784.

\bibitem{CC99}H.-D. Cao and B. Chow,
{\it Recent developments on the Ricci flow}, Bull. Amer. Math. Soc
(N.S.), {\bf 36}({\bf 1}) (1999), 59--74.

\bibitem{CCCY}H.-D. Cao, B. Chow, S.C. Chu,  and S.-T. Yau,
{\it Collected Papers on Ricci Flow}, Series in Geometry and
Topology, \textbf{37}, International Press, 2003.

\bibitem{Cao04} H.-D. Cao, R.S. Hamilton, and T. Ilmanen,
{\it Gaussian densities and stability for some Ricci solitons,}
math.DG/0404165.

\bibitem{Chee}J. Cheeger,
{\it Finiteness theorems for Riemannian manifolds}, Amer. J.
Math., \textbf{92} (1970), 61--74.

\bibitem{CC96}J. Cheeger and T.H. Colding,
{\it Lower bounds on Ricci curvature and the almost rigidity of
warped products}, Ann. of Math. (2), \textbf{144}({\bf 1}) (1996),
189--237.

\bibitem{CC} \bysame,
%Cheeger, J. and Colding, T. H.,
{\it On the structure of the spaces with Ricci curvature bounded
below}, I, J. Differential Geom., {\bf 46} (1997), 406--480.

\bibitem{CE} J. Cheeger and D. Ebin,
{\it Comparison theorems in Riemannian geometry,} North-Holland,
1975.

\bibitem{CG} J. Cheeger and D. Gromoll,
{\it On the structure of complete manifolds of nonnegative
curvature,} Ann. of Math., \textbf{46} (1972), 413--433.

\bibitem{ChG86}J. Cheeger and M. Gromov,
{\it Collapsing Riemannian manifolds while keeping their curvature
bounded}, I, J. Differential Geom., \textbf{23}({\bf 3}) (1986),
309--346.

\bibitem{ChG90} \bysame,
%Cheeger, J. and  Gromov, M.,
{\it Collapsing Riemannian manifolds while keeping their curvature
bounded}, II, J. Differential Geom., \textbf{32}({\bf 1}) (1990),
269--298.

\bibitem{CGF92} J. Cheeger, M. Gromov, and K. Fukaya,
{\it  Mikhael Nilpotent structures and invariant metrics on
collapsed manifolds,} J. Amer. Math. Soc., \textbf{5}({\bf 2}
(1992), 327--372.

\bibitem{CGT} J. Cheeger, M. Gromov, and M. Taylor,
{\it Finite propagation speed, kernel estimates for functions of
the Laplace operator, and the geometry of complete Riemannian
manifolds,} J. Differential Geom., {\bf 17} (1982), 15--53.

\bibitem{CY}J. Cheeger and S.-T. Yau,
{\it A lower bound for the heat kernel}, Comm. on Pure and Appl.
Math.,\textbf{ 34} (1981), 465--480.

\bibitem{CTZ04}B.L. Chen, S.H. Tang, and X. P. Zhu,
{\it A uniformization theorem of complete noncompact K\"ahler
surfaces with positive bisectional curvature}, J. Differential
Geometry, {\bf 67} (2004), 519--570.

\bibitem{CZ00}B.L. Chen and X.P. Zhu,
{\it Complete Riemannian manifolds with pointwise pinched
curvature}, Invent. Math., \textbf{140}({\bf 2}) (2000), 423--452.

\bibitem{CZ03} \bysame,
%Chen, B. L. and Zhu, X. P.,
{\it On complete noncompact K\"ahler manifolds with positive
bisectional curvature}, Math. Ann., {\bf 327} (2003), 1--23.

\bibitem{CZ04} \bysame,
%Chen, B. L. and Zhu, X. P.,
{\it Volume growth and curvature decay of positively curved
K\"ahler manifolds}, Quarterly Journal of Pure and Applied
Mathematics, \textbf{1}({\bf 1}) (2005), 68--108.

 \bibitem{CZ05U} \bysame,
%Chen, B. L. and Zhu, X. P.,
{\it Uniqueness of the Ricci Flow on Complete Noncompact
Manifolds,} arXiv:DG (AP) math. DG/0505447, May 2005.

\bibitem{CZ05F}  \bysame,
%Chen, B. L. and Zhu, X. P.,
{\it Ricci Flow with Surgery on Four-manifolds with Positive
Isotropic Curvature,} math.DG/0504478, April 2005.

\bibitem{CLY}S.Y. Cheng, P. Li, and S.-T. Yau,
{\it On the upper estimate of the heat kernel of complete
Riemannian manifold,} Amer. J. Math., {\bf 103}({\bf 5}) (1981),
1021--1063.

\bibitem{Chow}B. Chow,
{\it The Ricci flow on $2$-sphere}, J. Differential Geom., {\bf
33} (1991), 325--334.

\bibitem{ChowE} \bysame,
%Chow, B.,
{\it On the entropic estimate for the Ricci flow on compact
$2$-orbifolds}, J. Differential Geom., {\bf 33} (1991), 597--600.

\bibitem{ChCh95}B. Chow and S.C. Chu,
{\it A geometric interpretation of Hamilton's Harnack inequality
for the Ricci flow,} Math. Res. Lett., \textbf{2} (1995),
701--718.

\bibitem{CH}B. Chow and R.S. Hamilton,
{\it Constrained and linear Harnack inequalities for parabolic
equation,} Invent. Math. {\bf 128} (1997), 213-238.

\bibitem{CK04}B. Chow and D. Knopf,
{\it The Ricci flow: An introduction}, Mathematical Surveys and
Monographs, Amer. Math. Soc., Providence, RI, 2004.

\bibitem{Cl04}B. Chow and P. Lu,
{\it The maximum principle for systems of parabolic equations
subject to an avoidance set,} Pacific J. Math., \textbf{214 }({\bf
2}) (2004), 201--222.

\bibitem{CLN} B. Chow, P. Lu, and L. Ni,
{\it Hamilton's Ricci flow}, American Mathematical Society and
Science Press, to appear.

\bibitem{CM} T.H. Colding and W.P. Minicozzi, II,
{\it Estimates for the extinction time for the Ricci flow on
certain $3$-manifolds and a question of Perelman,} J. Amer. Math.
Soc., \textbf{18}({\bf 3}) (2005), 561--569.

\bibitem{De}D. De Turck,
{\it Deforming metrics in the direction of their Ricci tensors},
J. Differential Geom., {\bf 18} (1983), 157--162.

\bibitem{Dr}G. Drees,
{\it Asymptotically flat manifold of nonnegative curvature}, Diff.
Geom. Appl., \textbf{4} (1994), 77--90.

\bibitem{ESS}J. Eschenburg, V. Schroeder, and M. Strake,
{\it Curvature at infinity of open nonnegatively curved manifold},
J. Differential Geom., \textbf{30} (1989), 155--166.

\bibitem{FIM}M. Feldman, T. Ilmanen, and D. Knopf,
{\it Rotationally symmetric shrinking and expanding K\"ahler-Ricci
solitons}, J. Differential Geom., {\bf 65} (2003), 169--209.

\bibitem{Ft} T. Frankel,
{\it Manifolds with positive curvature}, Pacific J. Math.,
\textbf{11} (1961), 165--174.

\bibitem{GT} D. Gilbarg and N.S. Trudinger,
{\it Elliptic partial differential equations of second order,}
Reprint of the 1998 edition, Classics in Mathematics,
Springer-Verlag, Berlin, 2001.

\bibitem{GW82}R.E. Greene and H. Wu,
{\it On a new gap phenomonon in Riemannian geometry}, Proc. Nat.
Acad. Sci. U.S.A., \textbf{79}({\bf 2}) (1982), 714--715.

\bibitem{GW} \bysame,
%Greene, R. E. and Wu, H.,
{\it Gap theorems for noncompact Riemannian manifolds}, Duke Math.
J., \textbf{49} (1982), 731--756.

\bibitem{Gre} \bysame,
%Greene, R. E. and Wu, H.,
{\it Lipschitz convergence of Riemannian manifolds}, Pacific J.
Math., \textbf{131} (1988), 119--141.

\bibitem{GM}D. Gromoll and W. Meyer,
{\it On complete open manifolds of positive curvature}, Ann. of
Math., \textbf{90} (1969), 75--90.

\bibitem{Gro}M. Gromov,
{\it Metric Structures for Riemannian and Non-Riemannian Spaces},
Edited by J. LaFontaine and P. Pansu, English translation by Sean
Michael Bates, Birkh\"{a}user, 1998.

\bibitem{Gro78} \bysame,
%Gromov, M.,
{\it Almost flat manifolds}, J. Differential Geom., \textbf{13}
(1978), 231--241.

\bibitem{Gro79} \bysame,
%Gromov, M.,
{\it Hyperbolic manifolds} (according to Thurston and
J$\phi$rgensen), Bourbaki Seminar, Vol. 1979/80, 40--53, Lecture
Notes in Math., \textbf{842}, Springer, Berlin-New York, 1981.

\bibitem{GL80}M. Gromov and H.B. Lawson, Jr.,
{\it The classification of simply connected manifolds of positive
scalar curvature,} Ann. of Math. (2), \textbf{111}({\bf 3})
(1980), 423--434.

\bibitem{GL83} \bysame,
%Gromov, M. and Lawson, H. B. Jr.,
{\it Positive scalar curvature and the Dirac operator on complete
Riemannian manifolds,} Inst. Hautes \'{E}tudes Sci. Publ. Math.,
\textbf{58} (1983), 83--196.

\bibitem{Ha82} R.S. Hamilton,
{\it Three manifolds with positive Ricci curvature}, J.
Differential Geom., {\bf 17} (1982), 255--306.

\bibitem{Ha86} \bysame,
%Hamilton, R. S.,
{\it Four--manifolds with positive curvature operator}, J.
Differential Geom., {\bf 24} (1986), 153--179.

\bibitem{Ha88} \bysame,
%Hamilton, R. S.,
{\it The Ricci flow on surfaces}, Contemporary Mathematics, {\bf
71} (1988), 237--261.

\bibitem{Ha93} \bysame,
%Hamilton, R. S.,
{\it The Harnack estimate for the Ricci flow}, J. Differential
Geom., {\bf 37} (1993), 225-243.

\bibitem{Ha95} \bysame,
%Hamilton, R. S.,
{\it A compactness property for solution of the Ricci flow}, Amer.
J. Math., {\bf 117} (1995), 545--572.

\bibitem{Ha95F} \bysame,
%Hamilton, R. S.,
{\it The formation of singularities in the Ricci flow}, Surveys in
Differential Geometry, {\bf 2}, 7-136, International Press, 1995.

\bibitem{Ha97} \bysame,
%Hamilton, R. S.,
{\it Four manifolds with positive isotropic curvature}, Comm. in
Analysis and Geometry, {\bf 5} (1997), 1--92.

\bibitem{Ha99} \bysame,
%Hamilton, R. S.,
{\it Non-singular solutions to the Ricci flow on three manifolds},
Comm. Anal. Geom., {\bf 7} (1999), 695-729.

\bibitem{Ha93E} \bysame,
%Hamilton, R. S.,
{\it Eternal solutions to the Ricci flow}, J. Differential Geom.,
\textbf{38 }({\bf 1}) (1993), 1--11.

\bibitem{HaL} \bysame,
%Hamilton, R. S.,
{\it Lectures on the Ricci flow at Columbia University}, Fall,
2003.

\bibitem{He} S. Helgason,
{\it Differential Geometry and Symmetric Spaces}, Academic Press,
1962 (Second edition, 1978).

\bibitem{Hempel} J. Hempel,
{\it 3-manifolds,} Reprint of the 1976 original, AMS Chelsea
Publishing, Providence, RI, 2004.

\bibitem{Hir} M.W. Hirsch,
{\it Differential Topology}, Springer-Verlag, 1976.

\bibitem{Hu} G. Huisken,
{\it Ricci deformation of the metric on a Riemanian manifold,} J.
Differential Geometry, {\bf 21 } (1985), 47--62.

\bibitem{Hu90} \bysame
{\it Asymptotic behavior for singularities of the mean curvature
flow.} J. Differential Geometry, {\bf 31 } (1990), 285--299.

\bibitem{Iv} T. Ivey,
{\it Ricci solitons on compact three-manifolds}, Diff. Geom.
Appl., \textbf{3} (1993), 301--307.

\bibitem{Jaco} W. Jaco and P. Shalen,
{\it Seifert fibered spaces in $3$-manifolds}, Men. Amer. Math.
Soc., \textbf{220} (1979).

\bibitem{Joha} K. Johannson,
{\it Homotopy equivalences of $3$-manifolds with boundaries},
Lecture Notes in Math., \textbf{761} (1979).

\bibitem{KM} D.A. Kazdan and G. Margulis, {\it A simple proof of Selberg's
hypothesis,} Math. Sb. (117), \textbf{75} (1968), 163--168.

\bibitem{KW}J.L. Kazdan and F.W. Warner,
{\it Curvature functions for compact $2$-manifolds}, Ann. of Math.
(2),  \textbf{99} (1974), 14--47.

\bibitem{KL} B. Kleiner and J. Lott,
{\it Note on Perelman's paper}, http://www.math.lsa.umich.edu
/research /ricciflow /perelman.html, June 2003.

\bibitem{KL2} B. Kleiner and J. Lott,
{\it Notes on Perelman's papers}, preprint, math.DG/0605667.

\bibitem{Kn} H. Kneser,
{\it Geschlossene Fl\"{a}chen in dreidmensionalen
mannigfaltigkeiten,} Jahresbericht der Deut. Math. Verein.,
\textbf{38} (1929), 248--260.

\bibitem{Ko} N. Koiso,
{\it On rotationally symmmetric Hamilton's equation for
K\"{a}hler-Einstein metrics}, Recent Topics in Diff. Anal. Geom.
Adv. Studies Pure Math., \textbf {18-I} (1990), 327--337.

\bibitem{LSU} O.A. Ladyzenskaja, V.A. Solonnikov, and N.N. Uralceva,
{\it Linear and quasilinear equations of parabolic type,} Transl.
Amer. Math. Soc., {\bf 23} (1968).

\bibitem{LY} P. Li and S.-T. Yau,
{\it On the parabolic kernel of the Schr\"odinger operator,} Acta
Math., \textbf{156}({\bf 3}-{\bf 4}) (1986), 153--201.

\bibitem{Mar94} C. Margerin,
{\it Une caract\'erisation optimale de la structure
diff\'erentielle standard de la sph\`ere en terme de courbure pour
$($presque$)$ toutes les dimensions. I. Les \'enonc\'es}, (French)
[A sharp curvature-type characterization of the smooth sphere in
all but a few dimensions. I. Statement of the theorems] C.R. Acad.
Sci. Paris S\'er. I Math., \textbf{319}({\bf 6}) (1994), 605--607.

\bibitem{Mar98} \bysame,
%Margerin, C.,
{\it A sharp characterization of the smooth $4$-sphere in
curvature terms}, Comm. Anal. Geom., \textbf{6}({\bf 1}) (1998),
21--65.

\bibitem{Mc} C. McMullen, {\it Renormalization and 3-manifolds
which fiber over the circle}, Annals of Mathematics Studies {\bf
12}, Princeton U. Press, 1996.


\bibitem{MY} W.W. Meeks, III and S.-T. Yau,
{\it  The existence of embedded minimal surfaces and the problem
of uniqueness,} Math. Z., \textbf{179}({\bf 2}) (1982), 151--168.

\bibitem{MSY} W.W. Meeks, III, L. Simon, and S.-T. Yau,
{\it Embedded minimal surfaces, exotic spheres, and manifolds with
positive Ricci curvature}, Ann. of Math., \textbf{116} (1982),
621--659.

\bibitem{MiMo} M. Micallef and J.D. Moore, J
{\it Minimal two-spheres and the topology of manifolds with
positive curvature on totally isotropic two-planes}, Ann. of Math.
(2), {\bf 127} (1988), 199--227.

\bibitem{Mi} A.D. Milka,
{\it Metric structure of some class of spaces containing straight
lines}, Ukrain. Geometrical. Sbornik, \textbf{4} (1967), 43--48.

\bibitem{Mil62} J. Milnor,
{\it A unique factorization theorem for $3$-manifolds}, Amer. J.
Math., \textbf{84} (1962), 1--7.

\bibitem{Mil} \bysame,
%Milnor, J.,
{\it Towards the Poincar\'e conjecture and the classification of
$3$-manifolds}, Notices Amer. Math. Soc., {\bf 50}({\bf 10})
(2003), 1226--1233.

\bibitem{Mo88} N. Mok,
{\it The uniformization theorem for compact K\"ahler manifolds of
nonnegative holomorphic bisectional curvature}, J. Differential
Geom., {\bf 27} (1988), 179--214.

\bibitem{MoSY} N. Mok, Y.T. Siu, and S.-T. Yau,
{\it The Poincar\'e--Lelong equation on complete K\"ahler
manifolds}, Compositio Math., {\bf 44} (1981), 183--218.

\bibitem{Morgan} J.W. Morgan,
{\it On Thurston's uniformization theorem for three-dimensional
manifolds}, The Smith Conjecture (Bass and Morgan, eds.), Pure
Appl. Math., \textbf{112}, Academic Press, (1984), 37--125.

\bibitem{Mor} \bysame,
%Morgan, J. W.,
{\it Recent progress on the Poincar\'{e} conjecture and the
classification of $3$-manifolds}, Bull.of the A.M.S., {\bf
42}({\bf 1}) (2005), 57--78.

\bibitem{MT} J. Morgan and G. Tian,
{\it  Ricci flow and the Poincar\'e conjecture}, preprint,
math.DG/0607607.

\bibitem{Mori} S. Mori,
{\it Projective manifolds with ample tangent bundles}, Ann. of
Math., {\bf 100} (1979), 593--606.

\bibitem{Mos} G.D. Mostow,
{\it  Strong rigidity of locally symmetric spaces,} Annals of
Mathematics Studies, \textbf{78}, Princeton University Press,
Princeton, N.J.; University of Tokyo Press, Tokyo, 1973.

\bibitem{Ni} L. Ni,
{\it Ancient solution to Kahler-Ricci flow,} math.DG.0502494,
February 2005, preprint.

\bibitem{Ni06} \bysame,
%Ni, L.,
{\it A note on Perelman's LYH inequality,} math.DG.0602337,
February 2006, preprint.

\bibitem{NT} L. Ni and L.F. Tam,
{\it Plurisubharmonic functions and the structure of complete
K\"ahler manifolds with nonnegative curvature}, J. Differential
Geom., \textbf{64}({\bf 3}) (2003), 457--524.

\bibitem{Ns} S. Nishikawa,
{\it Deformation of Riemannian metrics and manifolds with bounded
curvature ratios}, Geometric measure theory and the calculus of
variations (Arcata, Calif., 1984), 343--352, Proc. Sympos. Pure
Math., {\bf 44}, Amer. Math. Soc., Providence, RI, 1986.

\bibitem{Ot} J.-P. Otal, {\it Le th\'eor\'eme d'hyperbolisation pir les
vari\'et\'es fibr\'ees de dimension 3},  At\'erisque {\bf 235},
1996.


\bibitem{P1} G. Perelmann,
{\it The entropy formula for the Ricci flow and its geometric
applications}, arXiv:math. DG/0211159 v1 November 11, 2002,
preprint.

\bibitem{P2} \bysame,
%Perelmann, G.,
{\it Ricci flow with surgery on three manifolds}, math.DG/0303109
v1 March 10, 2003, preprint.

\bibitem{P3} \bysame,
%Perelmann, G.,
{\it Finite extinction time to the solutions to the Ricci flow on
certain three manifolds}, math.DG/0307245 July 17, 2003, preprint.

\bibitem{Pe} S. Peters,
{\it Convergence of Riemannian mainfolds}, Comp. Math.,
\textbf{62} (1987), 3--16.

\bibitem{Pr} G. Prasad,
{\it Strong rigidity of $\mathbf{Q}$-rank $1$ lattices,} Invent.
Math., \textbf{21} (1973), 255--286.

\bibitem{Ro} O.S. Rothaus,
{\it Logarithmic Sobolev inequalities and the
 spectrum of Schr$\ddot{o}$dinger opreators,}
J. Funct. Anal., \textbf{42}({\bf 1}) (1981), 110--120.

\bibitem{ScY79m} R. Schoen and S.-T. Yau,
{\it On the structure of manifolds with positive scalar
curvature,} Manuscripta math., \textbf{28} (1979), 159--183.

\bibitem{ScY79} \bysame,
%Schoen, R. and Yau, S. T.,
{\it Existence of incompressible minimal surfaces and the topology
of three-dimensional manifolds with nonnegative scalar curvature,}
Ann. of Math. (2), \textbf{110}({\bf 1}) (1979), 127--142.

\bibitem{ScY82} \bysame,
%Schoen, R. and Yau, S. T.,
{\it Complete three dimensional manifolds with positive Ricci
curvature and scalar curvature}, Seminor on Differential Geometry,
Princeton University Press, 1982, 209-228.

\bibitem{ScY} \bysame,
%Schoen, R. and Yau, S. T.,
{\it Lectures on differential geometry}, in `Conference
Proceedings and Lecture Notes in Geometry and Topology', {\bf 1},
International Press Publications, 1994.

\bibitem{Se} N. Sesum,
{\it Limiting behaviour of the Ricci flow,} math.DG/0402194,
February 2004.

\bibitem{Sh89} W.X.  Shi,
{\it Deforming the metric on complete Riemannian manifold}, J.
Differential Geometry,  {\bf 30} (1989), 223--301.

\bibitem{Sh90} \bysame,
%Shi, W. X.,
{\it Ricci deformation of the metric on complete noncompact
K\"{a}hler manifolds,} Ph.D. Thesis, Harvard University, 1990.

 \bibitem{Sh90B} \bysame,
%Shi, W. X.,
{\it Complete noncompact K\"ahler manifolds with positive
holomorphic bisectional curvature}, Bull. Amer Math. Soc., {\bf
23} (1990), 437--440.

\bibitem{Sh97} \bysame,
%Shi, W. X.,
{\it Ricci flow and the uniformization on complete noncompact
K\"{a}hler manifolds,} J. Differential Geom., \textbf{45}({\bf 1})
(1997), 94--220.

\bibitem{ShY} T. Shioya and T. Yamaguchi,
{\it Volume collapsed three-manifolds with a lower curvature
bound,} Math. Ann., \textbf{333} (2005), 131--155.

\bibitem{Sin} I.M. Singer,
{\it Infinitesimallly homogeneous spaces}, Comm. Pure Appl. Math.,
\textbf{13} (1960), 685--697.

\bibitem{SiY} Y.T. Siu and S.-T. Yau,
{\it Complete K\"ahler manifolds with non--positive curvature of
faster than quadratic decay}, Ann. of Math., {\bf 105} (1977),
225--264.

\bibitem{St} M. Struwe,
{\it Curvature flows on surfaces}, Ann. Sc. Norm. Super. Pisa Cl.
Sci. (5), \textbf{1}({\bf 2}) (2002), 247--274.

\bibitem{Th82} W.P. Thurston,
{\it Three-dimensional manifolds,
 Kleinian groups and hyperbolic geometry,} Bull. Amer. Math. Soc. (N.S.),
 \textbf{6} (1982), no. 3, 357--381.

\bibitem{Th86} \bysame,
%Thurston, W. P.,
{\it Hyperbolic structures on $3$-manifolds. {\rm I:} Deformation
of acylindrical manifolds,} Ann. of Math. (2), \textbf{124}({\bf
2}) (1986), 203--246.

\bibitem{Th88} \bysame,
%Thurston, W. P.,
{\it Hyperbolic structures on $3$-manifolds {\rm II:} Surface
groups and $3$-manifolds which fiber over the circle,} preprint,
1988.

\bibitem{Th97} \bysame,
%Thurston, W. P.,
{\it Three-dimensional geometry and topology,} Vol. 1, Edited by
Silvio Levy, Princeton Mathematical Series, {\bf 35}, Princeton
University Press, Princeton, NJ, 1997.

\bibitem{W} F. Waldhausen,
{\it Eine Klasse von $3$-dimensionalen Mannigfaltigkeiten} I, II,
Invent. Math., \textbf{3} (1967), 308--333; \textbf{4} (1967),
87--117.

\bibitem{WZ} X.J. Wang, and X.H. Zhu,
{\it K\"{a}hler-Ricci solitons on toric manifolds with positive
first Chern class}, Adv. Math., \textbf{188}({\bf 1}) (2004),
87--103.

\bibitem{Y76}S.-T. Yau,
{\it Some function-theoretic properties of complete Riemannian
manifold and their applications to geometry}, Indiana Univ. Math.
J., \textbf{25} (1976), 659--670.

\bibitem{Y78} \bysame,
%Yau, S. T.,
{\it The role of partial differential equations in differential
geometry}, in `Proceedings of the International Congress of
Mathematicians', Helsinki, 1978, 237--250.

\bibitem{Y06} \bysame,
{\it Perspectives on geometric analysis}, to appear in Survey in
Differential Geometry, Vol. X. (math.DG/0602363, February, 2006)


\end{thebibliography}
