% PDF page 504
\source{gilbarg-trudinger:page-0504}

\begin{verbatim}
[BA 2]    The Dirichlet problem for equations of Monge-Ampère type and their n-dimensional
          analogues. Dokl. Akad. Nauk SSSR 126, 923-926 (1959) [Russian].
[BA 3]    Theory of quasilinear elliptic equations. Sibirsk. Mat. Z. 2, 179-186 (1961) [Russian].
[BA 4]    Geometrical methods for solving elliptic equations. Moscow: Izdat. Nauka 1965 [Russian].
[BA 5]    Mean curvature and quasilinear elliptic equations. Sibirsk. Mat. Z. 9, 1014-1040 (1968)
          [Russian]. English Translation in Siberian Math. J. 9, 752-771 (1968).
[BA 6]    Geometric problems in quasilinear elliptic equations. Uspehi Mat. Nauk. 25, no. 3, 49-112
          (1970) [Russian]. English Translation in Russian Math. Surveys 25, no. 3, 45-109 (1970).
[BA 7]    The Dirichlet problem for the elliptic n-dimensional Monge-Ampère equations and related
          problems in the theory of quasilinear equations. Proceedings of Seminar on Monge-Ampère
          Equations and Related Topics, Istituto Nazionale di Alta Matematica, Rome 1-78 (1982).

Bernstein, S.
[BE 1]    Sur la généralisation du problème de Dirichlet. I. Math. Ann. 62, 253-271 (1906).
[BE 2]    Méthode générale pour la résolution du problème de Dirichlet. C. R. Acad. Sci. Paris
          144, 1025-1027 (1907).
[BE 3]    Sur la généralisation du problème de Dirichlet II. Math. Ann. 69, 82-136 (1910).
[BE 4]    Conditions nécessaires et suffisantes pour la possibilité du problème de Dirichlet. C. R.
          Acad. Sci. Paris 150, 514-515 (1910).
[BE 5]    Sur les surfaces définies au moyen de leur courbure moyenne et totale. Ann. Sci. École
          Norm. Sup. 27, 233-256 (1910).
[BE 6]    Sur les équations du calcul des variations. Ann. Sci. École Norm. Sup. 29, 431-485 (1912).

Bers, L., and L. Nirenberg
[BN]      On linear and non-linear elliptic boundary value problems in the plane. In: Convegno
          Internazionale sulle Equazioni Lineari alle Derivate Parziali, Trieste, pp. 141-167. Rome:
          Edizioni Cremonese 1955.

Bers, L., and M. Schechter
[BS]      Elliptic equations. In: Partial Differential Equations, pp. 131-299. New York: Interscience
          1964.

Bliss, G. A.
[BL]      An integral inequality. J. London Math. Soc. 5, 40-46 (1930).

Bombieri, E.
[BM]      Variational problems and elliptic equations. In: Proceedings of the International Congress
          of Mathematicians, Vancouver 1974. Vol. 1, 53-63. Vancouver: Canadian Mathematical
          Congress 1975.

Bombieri, E., E. De Giorgi, and E. Giusti
[BDG]     Minimal cones and the Bernstein problem. Invent. Math. 7, 243-268 (1969).

Bombieri, E., E. De Giorgi, and M. Miranda
[BDM]     Una maggiorazione a priori relativa alle ipersuperfici minimali non parametrische. Arch.
          Rational Mech. Anal. 32, 255-267 (1969).

Bombieri, E., and E. Giusti
[BG 1]    Harnack's inequality for elliptic differential equations on minimal surfaces. Invent. Math.
          15, 24-46 (1972).
[BG 2]    Local estimates for the gradient of non-parametric surfaces of prescribed mean curvature.
          Comm. Pure Appl. Math. 26, 381-394 (1973).

Bony, J. M.
[BY]      Principe du maximum dans les espaces de Sobolev. C. R. Acad. Sci. Paris 265, 333-336
          (1967).

Bouligand, G., G. Giraud, and P. Delens
[BGD]     Le problème de la dérivée oblique en théorie du potentiel. Actualités Scientifiques et
          Industrielles 219. Paris: Hermann 1935.

Brandt, A.
[BR 1]    Interior estimates for second-order elliptic differential (or finite-difference) equations via
          the maximum principle. Israel J. Math. 7, 95-121 (1969).
[BR 2]    Interior Schauder estimates for parabolic differential- (or difference-) equations via the
          maximum principle. Israel J. Math. 7, 254-262 (1969).
\end{verbatim}

% PDF page 505
\source{gilbarg-trudinger:page-0505}
\begin{thebibliography}{99}

\bibitem{BV}
Brezis, H., and L. C. Evans
[BV] \quad A variational inequality approach to the Bellman-Dirichlet equation for two elliptic
operators. Arch. Rational Mech. Anal. \textbf{71}, 1--13 (1979).

\bibitem{BW1}
Browder, F. E.
[BW 1] \quad Strongly elliptic systems of differential equations. In: Contributions to the Theory of Partial
Differential Equations, pp. 15--51. Princeton, N. J.: Princeton University Press 1954.

\bibitem{BW2}
\quad On the regularity properties of solution of elliptic differential equations. Comm. Pure Appl.
Math. \textbf{9}, 351--361 (1956).

\bibitem{BW3}
\quad Apriori estimates for solutions of elliptic boundary value problems, I, II, III. Neder. Akad.
Wetensch. Indag. Math. \textbf{22}, 149--159, 160--169 (1960), \textbf{23}, 404--410 (1961).

\bibitem{BW4}
\quad Probl\`emes non-lin\'eaires. S\'eminaire de Math\'ematiques Sup\'erieures, No. 15 (\'Et\'e, 1965).
Montreal, Que.: Les Presses de l'Universit\'e de Montreal 1966.

\bibitem{BW5}
\quad Existence theorems for nonlinear partial differential equations. In: Proceedings of Symposia
in Pure Mathematics, Volume XVI. pp. 1--60. Providence, R. I.: American Mathematical
Society 1970.

\bibitem{CA1}
Caccioppoli, R.
[CA 1] \quad Sulle equazioni ellittiche a derivate parziali con $n$ variabili indipendenti. Atti Accad.
Naz. Lincei Rend. Cl. Sci. Fis. Mat. Natur. (6) \textbf{19}, 83--89 (1934).

\bibitem{CA2}
\quad Sulle equazioni ellittiche a derivate parziali con due variabili indipendenti, e sui problemi
regolari di calcolo delle variazioni. I. Atti Accad. Naz. Lincei Rend. Cl. Sci. Fis. Mat.
Natur. (6) \textbf{22}, 305--310 (1935).

\bibitem{CA3}
\quad Sulle equazioni ellittiche a derivate parziali con due variabili indipendenti, e sui problemi
regolari di calcolo delle variazioni. II. Atti Accad. Naz. Lincei Rend. Cl. Sci. Fis. Mat.
Natur. (6) \textbf{22}, 376--379 (1935).

\bibitem{CA4}
\quad Limitazioni integrali per le soluzioni di un'equazione lineare ellittica a derivate parziali.
Giorn. Mat. Battaglini (4) \textbf{4} (80), 186--212 (1951).

\bibitem{CNS}
Caffarelli, L., L. Nirenberg, and J. Spruck
[CNS] \quad The Dirichlet problem for nonlinear second order elliptic equations, I. Monge-Amp\`ere
equations. Comm. Pure Appl. Math. \textbf{37}, 369--402 (1984).

\bibitem{CKNS}
Caffarelli, L., J. Kohn, L. Nirenberg, and J. Spruck
[CKNS] \quad The Dirichlet problem for nonlinear second-order elliptic equations II. Comm. Pure Appl.
Math. \textbf{38}, 209--252 (1985).

\bibitem{CL}
Calabi, E.
[CL] \quad Improper affine hyperspheres of convex type and a generalization of a theorem by K.
J\"orgens. Michigan Math. J. \textbf{5}, 105--126 (1958).

\bibitem{CZ}
Calderon, A. P., and A. Zygmund
[CZ] \quad On the existence of certain singular integrals. Acta Math. \textbf{88}, 85--139 (1952).

\bibitem{CM1}
Campanato, S.
[CM 1] \quad Propriet\`a di inclusione per spazi di Morrey. Ricerche Mat. \textbf{12}, 67--86 (1963).

\bibitem{CM2}
\quad Equazioni ellittiche del II$^\circ$ ordine e spazi $L^{(2,\lambda)}$. Ann. Mat. Pura Appl. (4) \textbf{69}, 321--381
(1965).

\bibitem{CM3}
\quad Sistemi ellittici in forma divergenza. Regolarit\`a all' interno. Quaderni della Scuola Norm.
Sup. di Pisa (1980).

\bibitem{CS}
Campanato, S., and G. Stampacchia
[CS] \quad Sulle maggiorazioni in $L^p$ nella teoria delle equazioni ellittiche. Boll. Un. Mat. Ital. (3) \textbf{20},
393--399 (1965).

\bibitem{CY1}
Cheng, S. Y., and S. T. Yau
[CY 1] \quad On the regularity of the Monge-Amp\`ere equation det $(d^2u/dx_i\,dx_j) = F(x, u)$. Comm. Pure
Appl. Math. \textbf{30}, 41--68 (1977).

\bibitem{CY2}
\quad On the regularity of the solution of the $n$-dimensional Minkowski problem. Comm. Pure
Appl. Math. \textbf{19}, 495--516 (1976).

\bibitem{CY3}
\quad On the existence of a complete K\"ahler metric on non-compact complex manifolds and the
regularity of Fefferman's equation. Comm. Pure Appl. Math. \textbf{33}, 507--544 (1980).

\bibitem{CY4}
\quad The real Monge-Amp\`ere equation and affine flat structures. Proc. 1980 Beijing Symposium
on Differential Geometry and Differential Equations. Vol. I, 339--370 (1982). Editors,
S. Chern, W. T. Wu.

\bibitem{CI1}
Chicco, M.
[CI 1] \quad Principio di massimo forte per sottosoluzioni di equazioni ellittiche di tipo variazionale.
Boll. Un. Mat. Ital. (3) \textbf{22}, 368--372 (1967).

\end{thebibliography}

% PDF page 506
\source{gilbarg-trudinger:page-0506}

\begin{flushleft}
{}[CI 2]\quad Semicontinuit\`a delle sottosoluzioni di equazioni ellittiche di tipo variazionale. Boll. Un.\ Mat.\ Ital.\ (4) 1, 548--553 (1968).\\
{}[CI 3]\quad Principio di massimo per soluzioni di problemi al contorno misti per equazioni ellittiche\\
di tipo variazionale. Boll. Un.\ Mat.\ Ital.\ (4) 3, 384--394 (1970).\\
{}[CI 4]\quad Solvability of the Dirichlet problem in $H^2 \cap Q(\Omega)$ for a class of linear second order elliptic\\
partial differential equations. Boll. Un.\ Mat.\ Ital.\ (4) 4, 374--387 (1971).\\
{}[CI 5]\quad Sulle equazioni ellittiche del secondo ordine a coefficienti continui. Ann.\ Mat.\ Pura Appl.\\
(4) 88, 123--133 (1971).\\[1ex]
Cordes, H.\ O.\\
{}[CO 1]\quad \`Uber die erste Randwertaufgabe bei quasilinearen Differentialgleichungen zweiter\\
Ordnung in mehr als zwei Variablen. Math.\ Ann.\ 131, 278--312 (1956).\\
{}[CO 2]\quad Zero order a priori estimates for solutions of elliptic differential equations. In: Proceedings\\
of Symposia in Pure Mathematics, Volume IV. pp.\ 157--166. Providence, R.I.: American\\
Mathematical Society 1961.\\[1ex]
Courant, R., and D.\ Hilbert\\
{}[CH]\quad Methods of Mathematical Physics. Volumes I, II. New York: Interscience 1953, 1962.\\[1ex]
De Giorgi, E.\\
{}[DG 1]\quad Sulla differenziabilit\`a e l'analitic\`a delle estremali degli integrali multipli regolari. Mem.\\
Accad.\ Sci.\ Torino Cl.\ Sci.\ Fis.\ Mat.\ Natur.\ (3) 3, 25--43 (1957).\\
{}[DG 2]\quad Una estensione del teorema di Bernstein. Ann.\ Scuola Norm.\ Sup.\ Pisa (3) 19, 79--85 (1965).\\[1ex]
Delano\={e}, P.\\
{}[DE 1]\quad \'Equations du type Monge-Amp\`ere sur les vari\'et\'es Riemanniennes compactes I. J.\ Funct.\\
Anal.\ 40, 358--386 (1981).\\
{}[DE 2]\quad \'Equations du type Monge-Amp\`ere sur les vari\'et\'es Riemanniennes compactes II. J.\ Funct.\\
Anal.\ 41, 341--353 (1981).\\[1ex]
Dieudonn\'e, J.\\
{}[DI]\quad Foundations of Modern Analysis. New York: Academic Press, 1960.\\[1ex]
Douglas, J., T.\ Dupont, and J.\ Serrin\\
{}[DDS]\quad Uniqueness and comparison theorems for nonlinear elliptic equations in divergence form.\\
Arch.\ Rational Mech.\ Anal.\ 42, 157--168 (1971).\\[1ex]
Douglis, A., and L.\ Nirenberg\\
{}[DN]\quad Interior estimates for elliptic systems of partial differential equations. Comm.\ Pure Appl.\\
Math.\ 8, 503--538 (1955).\\[1ex]
Dunford, N., and J.\ T.\ Schwartz\\
{}[DS]\quad Linear Operators, Part I. New York: Interscience 1958.\\[1ex]
Edmunds, D.\ E.\\
{}[ED]\quad Quasilinear second order elliptic and parabolic equations. Bull.\ London Math.\ Soc.\ 2.\\
5--28 (1970).\\[1ex]
Edwards, R.\ E.\\
{}[EW]\quad Functional Analysis: Theory and Applications. New York: Holt, Rinehart and Winston\\
1965.\\[1ex]
Egorov, Yu.\ V., and V.\ A.\ Kondrat'ev\\
{}[EK]\quad The oblique derivative problem. Mat.\ Sb.\ (N.S.) \textbf{78} (120), 148--176 (1969) [Russian].\\
English Translation in Math USSR Sb.\ 7, 139--169 (1969).\\[1ex]
Emmer, M.\\
{}[EM]\quad Esistenza, unicit\`a e regolarit\`a nelle superficie di equilibrio nei capillari. Ann.\ Univ.\ Ferrara\\
\textbf{18}, 79--94 (1973).\\[1ex]
Evans, L.\ C.\\
{}[EV 1]\quad A convergence theorem for solutions of non-linear second order elliptic equations. Indiana\\
University Math.\ J.\ 27, 875--887 (1978).\\
{}[EV 2]\quad Classical solutions of fully nonlinear, convex, second order elliptic equations. Comm.\ Pure\\
Appl.\ Math.\ 25, 333--363 (1982).\\
{}[EV 3]\quad Classical solutions of the Hamilton-Jacobi Bellman equation for uniformly elliptic opera-\\
tors. Trans.\ Amer.\ Math.\ Soc.\ 275, 245--255 (1983).\\
{}[EV 4]\quad Some estimates for nondivergence structure, second order equations. Trans.\ Amer.\ Math.\\
Soc.\ 287, 701--712 (1985).\\[1ex]
Evans, L.\ C., and A.\ Friedman\\
{}[EF]\quad Optimal stochastic switching and the Dirichlet problem for the Bellman equation. Trans.\\
Amer.\ Math.\ Soc.\ 253, 365--389 (1979).
\end{flushleft}

% PDF page 507
\source{gilbarg-trudinger:page-0507}

\begin{thebibliography}{99}

\bibitem[EL 1]{EL1}
Evans, L. C., and P. L. Lions
Résolution des équations de Hamilton-Jacobi-Bellman pour des opérateurs uniformément
elliptiques.C. R. Acad. Sci. Paris, \textbf{290}, 1049--1052 (1980).

\bibitem[EL 2]{EL2}
Fully non-linear second order elliptic equations with large zeroth order coefficient. Ann.
Inst. Fourier (Grenoble) \textbf{31}, fasc. 2, 175--191 (1981).

\bibitem[FKS]{FKS}
Fabes, E. B., C. E. Kenig, and R. P. Serapioni
The local regularity of solutions of degenerate elliptic equations. Comm. Part. Diff. Equats.
\textbf{7}, 77--116 (1982).

\bibitem[FS]{FS1}
Fabes, E. B., and D. W. Stroock
The $L^p$-integrability of Green's functions and fundamental solutions for elliptic and
parabolic equations. Duke Math. J. \textbf{51}, 997--1016 (1984).

\bibitem[FE]{FE}
Federer, H.
Geometric Measure Theory. Berlin--Heidelberg--New York: Springer-Verlag 1969.

\bibitem[FS]{FS2}
Fefferman, C., and E. Stein
$H^p$ spaces of several variables. Acta Math. \textbf{129}, 137--193 (1972).

\bibitem[FN 1]{FN1}
Finn, R.
A property of minimal surfaces. Proc. Nat. Acad. Sci. U.S.A., \textbf{39}, 197--201 (1953).

\bibitem[FN 2]{FN2}
On equations of minimal surface type. Ann. of Math. (2) \textbf{60}, 397--416 (1954).

\bibitem[FN 3]{FN3}
New estimates for equations of minimal surface type. Arch. Rational Mech. Anal. \textbf{14},
337--375 (1963).

\bibitem[FN 4]{FN4}
Remarks relevant to minimal surfaces, and to surfaces of prescribed mean curvature.
J. Analyse Math. \textbf{14}, 139--160 (1965).

\bibitem[FG 1]{FG1}
Finn, R., and D. Gilbarg
Asymptotic behavior and uniqueness of plane subsonic flows. Comm. Pure Appl. Math.
\textbf{10}, 23--63 (1957).

\bibitem[FG 2]{FG2}
Three-dimensional subsonic flows, and asymptotic estimates for elliptic partial differential
equations. Acta Math. \textbf{98}, 265--296 (1957).

\bibitem[FS]{FS3}
Finn, R., and J. Serrin
On the Hölder continuity of quasi-conformal and elliptic mappings. Trans. Amer. Math.
Soc. \textbf{89}, 1--15 (1958).

\bibitem[FI 1]{FI1}
Fiorenza, R.
Sui problemi di derivata obliqua per le equazioni ellittiche. Ricerche Mat. \textbf{8}, 83--110 (1959).

\bibitem[FI 2]{FI2}
Sui problemi di derivata obliqua per le equazioni ellittiche quasi lineari. Ricerche Mat. \textbf{15},
74--108 (1966).

\bibitem[FL]{FL}
Fleming, W.
On the oriented Plateau problem. Rend. Circ. Mat. Palermo (2) \textbf{11}, 69--90 (1962).

\bibitem[FR]{FR}
Friedman, A.
Partial Differential Equations. New York: Holt, Rinehart and Winston 1969.

\bibitem[FD 1]{FD1}
Friedrichs, K. O.
The identity of weak and strong extensions of differential operators. Trans. Amer. Math.
Soc. \textbf{55}, 132--151 (1944).

\bibitem[FD 2]{FD2}
On the differentiability of the solutions of linear elliptic differential equations. Comm.
Pure Appl. Math. \textbf{6}, 299--326 (1953).

\bibitem[GA]{GA}
Gårding, L.
Dirichlet's problem for linear elliptic partial differential equations. Math. Scand. \textbf{1}, 55--72
(1953).

\bibitem[GZ 1]{GZ1}
Gariepy, R., and W. P. Ziemer
Behaviour at the boundary of solutions of quasilinear elliptic equations. Arch. Rational
Mech. Anal. \textbf{56}, 372--384 (1974).

\bibitem[GZ 2]{GZ2}
A regularity condition at the boundary for solutions of quasilinear elliptic equations. Arch.
Rational Mech. Anal. \textbf{67}, 25--39 (1977).

\bibitem[GE 1]{GE1}
Gerhardt, C.
Existence, regularity, and boundary behaviour of generalized surfaces of prescribed mean
curvature. Math. Z. \textbf{139}, 173--198 (1974).

\bibitem[GE 2]{GE2}
On the regularity of solutions to variational problems in BV ($\Omega$). Math. Z. \textbf{149}, 281--286
(1976).

\end{thebibliography}

% PDF page 508
\source{gilbarg-trudinger:page-0508}
\begin{thebibliography}{99}

\bibitem{GE3}
Global regularity of the solutions to the capillarity problem. Ann. Scuola Norm. Sup.
Pisa (4) 3, 157--175 (1976).

\bibitem{GE4}
Boundary value problems for surfaces of prescribed mean curvature. J. Math. Pures et
Appl. \textbf{58}, 75--109 (1979).

\bibitem{GI1}
Giaquinta, M.
On the Dirichlet problem for surfaces of prescribed mean curvature. Manuscripta Math.
\textbf{12}, 73--86 (1974).

\bibitem{GI2}
Multiple Integrals in the Calculus of Variations and Nonlinear Elliptic Systems, Annals of
Math. Studies 105, Princeton Univ. Press, Princeton, 1983.

\bibitem{GL1}
Gilbarg, D.
Some local properties of elliptic equations. In: Proceedings of Symposia in Pure Mathe-
matics, Volume IV, pp. 127--141. Providence, R.I.: American Mathematical Society 1961.

\bibitem{GL2}
Boundary value problems for nonlinear elliptic equations in $n$ variables. In: Nonlinear
Problems, pp. 151--159. Madison, Wis.: University of Wisconsin Press 1963.

\bibitem{GH}
Gilbarg, D., and L. Hörmander
Intermediate Schauder estimates. Arch. Rational Mech. Anal. \textbf{74}, 297--318 (1980).

\bibitem{GS}
Gilbarg, D., and J. Serrin
On isolated singularities of solutions of second order elliptic differential equations. J.
Analyse Math. \textbf{4}, 309--340 (1955/56).

\bibitem{GR1}
Giraud, G.
Sur le problème de Dirichlet généralisé (deuxième mémoire). Ann. Sci. École Norm. Sup.
(3) \textbf{46}, 131--245 (1929).

\bibitem{GR2}
Généralisation des problèmes sur les opérations du type elliptiques. Bull. Sci. Math. (2)
\textbf{56}, 248--272, 281--312, 316--352 (1932).

\bibitem{GR3}
Sur certains problèmes non linéaires de Neumann et sur certains problèmes non linéaires
mixtes. Ann. Sci. École Norm. Sup. (3) \textbf{49}, 1--104, 245--309 (1932).

\bibitem{GR4}
Équations à integrales principales d’ordre quelconque. Ann. Sci. Ecole Norm. Sup. \textbf{53}
(36) 1--40 (1936).

\bibitem{GT1}
Giusti, E.
Superfici cartesiane di area minima. Rend. Sem. Mat. Fis. Milano \textbf{40}, 3--21 (1970).

\bibitem{GT2}
Boundary value problems for non-parametric surfaces of prescribed mean curvature.
Ann. Scuola Norm. Sup. Pisa (4) \textbf{3}, 501--548 (1976).

\bibitem{GT3}
On the equation of surfaces of prescribed mean curvature. Invent. Math. \textbf{46}, 111--137 (1978).

\bibitem{GT4}
Equazioni elliptiche del secondo ordine. Bologna: Pitagora Editrice 1978.

\bibitem{GT5}
Minimal Surfaces and Functions of Bounded Variation. Boston: Birkhäuser, 1984.

\bibitem{GC}
Greco, D.
Nuove formule integrali di maggiorazione per le soluzioni di un'equazione lineare di
tipo ellittico ed applicazioni alla teoria del potenziale. Ricerche Mat. \textbf{5}, 126--149 (1956).

\bibitem{GU}
Günter, N. M.
Potential Theory and its Applications to Basic Problems of Mathematical Physics. New
York : Frederick Ungar Publishing Co. 1967.

\bibitem{GN}
Gustin, N.
A bilinear integral identity for harmonic functions. Amer. J. Math. \textbf{70}, % [illegible]

\bibitem{HL}
Hardy, G. H., and J. E. Littlewood
Some properties of fractional integrals. I. Math. Z. \textbf{27}, 565--606 (1928).

\bibitem{HLP}
Hardy, G. H., J. E. Littlewood, and G. Polya
Inequalities. 2nd ed., Cambridge: Cambridge University Press 1952.

\bibitem{HA}
Hartman, P.
On the bounded slope condition. Pacific J. Math. \textbf{18}, 495--511 (1966).

\bibitem{HN}
Hartman, P., and L. Nirenberg
On spherical image maps whose Jacobians do not change sign. Amer. J. Math. \textbf{81}, 901--
920 (1959).

\bibitem{HS}
Hartman, P., and G. Stampacchia
On some non-linear elliptic differential-functional equations. Acta Math. \textbf{115}, 271--310
(1966).

\end{thebibliography}

% PDF page 509
\source{gilbarg-trudinger:page-0509}

\begin{thebibliography}{99}

\bibitem[HE 1]{HE1}
Heinz, E.
Über die Lösungen der Minimalflächengleichung. Nachr. Akad. Wiss. Göttingen. Math.-
Phys. Kl. IIa, 51--56 (1952).

\bibitem[HE 2]{HE2}
On elliptic Monge-Ampère equations and Weyl's imbedding problem. Analyse Math. 7,
1--52 (1959).

\bibitem[HE 3]{HE3}
Interior gradient estimates for surfaces $z = f(x, y)$ with prescribed mean curvature. J.
Differential Geometry \textbf{5}, 149--157 (1971).

\bibitem[HL]{HL}
Helms, L. L.
Introduction to potential theory. New York: Wiley-Interscience 1969.

\bibitem[HR]{HR}
Hervé, R.-M.
Recherches axiomatiques sur la théorie des fonctions surharmoniques et du potentiel.
Ann. Inst. Fourier (Grenoble) \textbf{12}, 415--571 (1962).

\bibitem[HH]{HH}
Hervé, R.-M., and M. Hervé
Les fonctions surharmoniques associées à un opérateur elliptique du second ordre à
coefficients discontinus. Ann. Inst. Fourier (Grenoble) \textbf{19}, fasc. 1, 305--359 (1969).

\bibitem[HI]{HI}
Hilbert, D.
Über das Dirichletsche Prinzip. Jber. Deutsch. Math.-Verein. \textbf{8}, 184--188 (1900).

\bibitem[HD]{HD}
Hildebrandt, S.
Maximum principles for minimal surfaces and for surfaces of continuous mean curvature.
Math. Z. \textbf{128}, 253--269 (1972).

\bibitem[HO 1]{HO1}
Hopf, E.
Elementare Bemerkungen über die Lösungen partieller Differentialgleichungen zweiter
Ordnung vom elliptischen Typus. Sitz. Ber. Preuss. Akad. Wissensch. Berlin. Math.-Phys.
Kl. \textbf{19}, 147--152 (1927).

\bibitem[HO 2]{HO2}
Zum analytischen Charakter der Lösungen regulärer zweidimensionaler Variationspro-
bleme. Math. Z. \textbf{30}, 404--413 (1929).

\bibitem[HO 3]{HO3}
Über den funktionalen, insbesondere den analytischen Charakter der Lösungen elliptischer
Differentialgleichungen zweiter Ordnung. Math. Z. \textbf{34}, 194--233 (1932).

\bibitem[HO 4]{HO4}
On S. Bernstein's theorem on surfaces $z(x, y)$ of nonpositive curvature. Proc. Amer. Math.
Soc. \textbf{1}, 80--85 (1950).

\bibitem[HO 5]{HO5}
A remark on linear elliptic differential equations of second order. Proc. Amer. Math. Soc.
\textbf{3}, 791--793 (1952).

\bibitem[HO 6]{HO6}
On an inequality for minimal surfaces $z = z(x, y)$. J. Rational Mech. Anal. \textbf{2}, 519--522 (1953).

\bibitem[HM 1]{HM1}
Hörmander, L.
Pseudo-differential operators and non-elliptic boundary problems. Ann. of Math. (2)
\textbf{83}, 129--209 (1966).

\bibitem[HM 2]{HM2}
Hypoelliptic second order differential equations. Acta Math. \textbf{119}, 147--161 (1967).

\bibitem[HM 3]{HM3}
The boundary problems of physical geodesy. Arch. Rational Mech. Anal. \textbf{62}, 1--52 (1976).

\bibitem[IV 1]{IV1}
Ivanov, A. V.
Interior estimates of the first derivatives of the solutions of quasi-linear nonuniformly
elliptic and nonuniformly parabolic equations of general form. Zap. Nauchn. Sem. Lenin-
grad. Otdel. Mat. Inst. Steklov. (LOMI) \textbf{14}, 24--47 (1969) [Russian]. English Translation
in Sem. Math. V. A. Steklov Math. Inst. Leningrad \textbf{14}, 9--21 (1971).

\bibitem[IV 2]{IV2}
The Dirichlet problem for second-order quasi-linear nonuniformly elliptic equations.
Zap. Nauchn. Sem. Leningrad. Otdel. Mat. Inst. Steklov. (LOMI) \textbf{19}, 79--94 (1970) [Russian].
English Translation in Sem. Math. V. A. Steklov Math. Inst. Leningrad \textbf{19}, 43--52 (1972).

\bibitem[IV 3]{IV3}
A priori estimates for solutions of nonlinear second order elliptic equations. Zap. Nauchn.
Sem. Leningrad. Otdel. Mat. Inst. Steklov. (LOMI) \textbf{59}, 31--59 (1976) [Russian]. English
translation in J. Soviet Math. \textbf{10}, 217--240 (1978).

\bibitem[IV 4]{IV4}
A priori estimates of the second derivatives of solutions to nonlinear second order equations
on the boundary of the domain. Zap. Nauchn. Sem. Leningrad. Otdel. Mat. Inst. Steklov.
(LOMI) \textbf{69}, 65--76 (1977) [Russian]. English translation in J. Soviet Math. \textbf{10}, 44--53 (1978).

\bibitem[IC 1]{IC1}
Ivočkina, N. M.
The integral method of barrier functions and the Dirichlet problem for equations with
operators of Monge-Ampère type. Mat. Sb. (N. S.) \textbf{112}, 193--206 (1980) [Russian]. English
translation in Math. USSR Sb. \textbf{40}, 179--192 (1981).

\bibitem[IC 2]{IC2}
An a priori estimate of $\|u\|_{C^2(\Omega)}$ for convex solutions of Monge-Ampère equations, Zap.
Nauchn. Sem. Leningr. Otdel. Mat. Inst. Steklova (LOMI) \textbf{96}, 69--79 (1980).

\end{thebibliography}

% PDF page 510
\source{gilbarg-trudinger:page-0510}
\begin{quote}
\textbf{498}\hfill \textbf{Bibliography}

\noindent [IC 3]\quad Classical solvability of the Dirichlet problem for the Monge-Amp\`ere equation, Zap.\ Nauch.\ Sem.\ Leningr.\ Otdel.\ Mat.\ Inst.\ Steklova (LOMI) 131, 72--79 (1983).

\noindent Ivo\v{c}kina, N. M., and A. P. Oskolkov

\noindent [IO]\quad Nonlocal estimates for the first derivatives of solutions of the first boundary problem for certain classes of nonuniformly elliptic and nonuniformly parabolic equations and systems.
Trudy Mat.\ Inst.\ Steklov 110, 65--101 (1970) [Russian]. English Translation in Proc.\ Steklov Inst.\ Math.\ 110, 72--115 (1970).

\noindent Jenkins, H.

\noindent [JE]\quad On 2-dimensional variational problems in parametric form.\ Arch.\ Rational Mech.\ Anal.\ 8, 181--206 (1961).

\noindent Jenkins, H., and J. Serrin

\noindent [JS 1]\quad Variational problems of minimal surface type I.\ Arch.\ Rational Mech.\ Anal.\ 12, 185--212 (1963).

\noindent [JS 2]\quad The Dirichlet problem for the minimal surface equation in higher dimensions.\ J.\ Reine Angew.\ Math.\ 229, 170--187 (1968).

\noindent John, F., and L. Nirenberg

\noindent [JN]\quad On functions of bounded mean oscillation.\ Comm.\ Pure Appl.\ Math.\ 14, 415--426 (1961).

\noindent Kamynin L. I., and B. N. Him\v{c}enko

\noindent [KH]\quad Investigations on the maximum principle.\ Dokl.\ Akad.\ Nauk SSSR \textbf{240}, 774--777 (1978) [Russian]. English Translation in Soviet Math.\ Dokl.\ 19, 677--681 (1978).

\noindent Kellogg, O. D.

\noindent [KE 1]\quad On the derivatives of harmonic functions on the boundary.\ Trans.\ Amer.\ Math.\ Soc.\ \textbf{33}, 486--510 (1931).

\noindent [KE 2]\quad Foundations of Potential Theory.\ New York: Dover 1954.

\noindent Kinderlehrer, D., and G. Stampacchia

\noindent [KST]\quad An introduction to variational inequalities and their applications.\ New York: Academic Press (1980).

\noindent Kohn, J. J.

\noindent [KJ]\quad Pseudo-differential operators and hypoellipticity.\ In: Proceedings of Symposia in Pure Mathematics, Volume XXIII, pp.\ 61--69.\ Providence, R. I.: American Mathematical Society (1973).

\noindent Kondrachov, V. I.

\noindent [KN]\quad Sur certaines propri\'et\'es des fonctions dans l'espace $L^p$.\ C. R. (Doklady) Acad.\ Sci.\ URSS (N. S.) \textbf{48}, 535--538 (1945).

\noindent Korn, A.

\noindent [KR 1]\quad \v{U}ber Minimalfl\"achen, deren Randkurven wenig von ebenen Kurven abweichen.\ Abhandl.\ K\"onigl.\ Preuss.\ Akad.\ Wiss., Berlin 1909: Anhang, Abh.\ 2.

\noindent [KR 2]\quad Zwei Anwendungen der Methode der sukzessiven Ann\"aherungen.\ Schwarz Festschrift.\ Berlin 1914, 215--229.

\noindent Ko\v{s}elev, A. E.

\noindent [KO]\quad On boundedness in $L^p$ of derivatives of solutions of elliptic differential equations.\ Mat.\ Sb.\ (N. S.) \textbf{38} (80), 359--372 (1956) [Russian].

\noindent Krylov, N. V.

\noindent [KV 1]\quad On the first boundary value problem for second order elliptic equations, Differents.\ Uravn.\ \textbf{3}, 315--325 (1967) [Russian]. English Translation in Differential Equations \textbf{3}, 158--164 (1967).

\noindent [KV 2]\quad Controlled diffusion processes.\ Berlin--Heidelberg--New York: Springer-Verlag (1980).

\noindent [KV 3]\quad On control of a diffusion process up to the time of first exit from a region.\ Izvestia Akad.\ Nauk.\ SSSR \textbf{45}, 1029--1048 (1981) [Russian]. English Translation in Math.\ USSR Izv.\ \textbf{19}, 297--313 (1982).

\noindent [KV 4]\quad Boundedly inhomogeneous elliptic and parabolic equations.\ Izvestia Akad.\ Nauk.\ SSSR \textbf{46}, 487--523 (1982) [Russian]. English Translation in Math.\ USSR Izv.\ \textbf{20} (1983).

\noindent [KV 5]\quad Boundedly inhomogeneous elliptic and parabolic equations in a domain.\ Izvestia Akad.\ Nauk.\ SSSR \textbf{47}, 75--108 (1983) [Russian]. English translation in Math.\ USSR Izv.\ \textbf{22}, 67--97 (1984).

\noindent [KV 6]\quad On degenerate nonlinear elliptic equations.\ Mat.\ Sb.\ (N. S.) \textbf{120}, 311--330 (1983) [Russian]. English translation in Math.\ USSR Sbornik \textbf{48}, 307--326 (1984).
\end{quote}

% PDF page 511
\source{gilbarg-trudinger:page-0511}

\begin{quote}
\textbf{Bibliography}

\medskip

\textbf{Krylov, N. V. and M. V. Safonov}
\newline
{}[KS 1]\quad An estimate of the probability that a diffusion process hits a set of positive measure. Dokl.\newline
Akad. Nauk. SSSR \textbf{245}, 253--255 (1979) [Russian]. English translation in Soviet Math.\newline
Dokl. \textbf{20}, 253--255 (1979).
\newline
{}[KS 2]\quad Certain properties of solutions of parabolic equations with measurable coefficients. Izvestia\newline
Akad. Nauk. SSSR \textbf{40}, 161--175 (1980) [Russian]. English translation in Math. USSR Izv.\newline
\textbf{16}, 151--164 (1981).

\medskip

\textbf{Ladyzhenskaya, O. A.}
\newline
{}[LA]\quad Solution of the first boundary problem in the large for quasi-linear parabolic equations.\newline
Trudy Moskov. Mat. Ob\v{s}. \textbf{7}, 149--177 (1958) [Russian].

\medskip

\textbf{Ladyzhenskaya, O. A., and N. N. Ural'tseva}
\newline
{}[LU 1]\quad On the smoothness of weak solutions of quasilinear equations in several variables and of\newline
variational problems. Comm. Pure Appl. Math. \textbf{14}, 481--495 (1961).
\newline
{}[LU 2]\quad Quasilinear elliptic equations and variational problems with several independent variables.\newline
Uspehi Mat. Nauk \textbf{16}, no. 1, 19--90 (1961) [Russian]. English Translation in Russian Math.\newline
Surveys \textbf{16}, no. 1, 17--92 (1961).
\newline
{}[LU 3]\quad On H\"older continuity of solutions and their derivatives of linear and quasilinear elliptic\newline
and parabolic equations. Trudy Mat. Inst. Steklov. \textbf{73}, 172--220 (1964) [Russian]. English\newline
Translation in Amer. Math. Soc. Transl. (2) \textbf{61}, 207--269 (1967).
\newline
{}[LU 4]\quad Linear and Quasilinear Elliptic Equations. Moscow: Izdat. ``Nauka'' 1964 [Russian].\newline
English Translation: New York: Academic Press 1968. 2nd Russian ed. 1973.
\newline
{}[LU 5]\quad Global estimates of the first derivatives of the solutions of quasi-linear elliptic and parabolic\newline
equations. Zap. Nauch. Sem. Leningrad. Otdel. Mat. Inst. Steklov. (LOMI) \textbf{14}, 127--155\newline
(1969) [Russian]. English Translation in Sem. Math. V. A. Steklov Math. Inst. Leningrad\newline
\textbf{14}, 63--77 (1971).
\newline
{}[LU 6]\quad Local estimates for gradients of solutions of non-uniformly elliptic and parabolic equations.\newline
Comm. Pure Appl. Math. \textbf{23}, 677--703 (1970).
\newline
{}[LU 7]\quad H\"older estimates for solutions of second order quasilinear elliptic equations in general\newline
form. Zap. Nauch. Sem. Leningrad. Otdel. Mat. Inst. Steklova (LOMI) \textbf{96}, 161--168 (1980)\newline
{}[Russian]. English translation in J. Soviet Math. \textbf{21}, 762--768 (1983).
\newline
{}[LU 8]\quad On Sounds for max $|u_x|$ for solutions of quasilinear elliptic and parabolic equations and\newline
existence theorems. Zap. Nauch. Sem. Leningr. Otdel. Mat. Inst. Steklova (LOMI) \textbf{138},\newline
90--107 (1984).

\medskip

\textbf{Landis, E. M.}
\newline
{}[LN 1]\quad Harnack's inequality for second order elliptic equations of Cordes type. Dokl. Akad.\newline
Nauk SSSR \textbf{179}, 1272--1275 (1968) [Russian]. English Translation in Soviet Math. Dokl.\newline
\textbf{9}, 540--543 (1968).
\newline
{}[LN 2]\quad Second Order Equations of Elliptic and Parabolic Types. Moscow: Nauka, 1971 [Russian].

\medskip

\textbf{Landkof, N. S.}
\newline
{}[LK]\quad Foundations of Modern Potential Theory. New York--Heidelberg--Berlin: Springer-Verlag\newline
(1972).

\medskip

\textbf{Lax, P. D.}
\newline
{}[LX]\quad On Cauchy's problem for hyperbolic equations and the differentiability of solutions of\newline
elliptic equations. Comm. Pure Appl. Maths. \textbf{8}, 615--633 (1955).

\medskip

\textbf{Lax, P. D., and A. M. Milgram}
\newline
{}[LM]\quad Parabolic equations. In: Contributions to the Theory of Partial Differential Equations,\newline
pp. 167--190. Princeton, N.J.: Princeton University Press 1954.

\medskip

\textbf{Lebesgue, H.}
\newline
{}[LE]\quad Sur le probl\`eme de Dirichlet. Rend. Circ. Mat. Palermo \textbf{24}, 371--402 (1907).

\medskip

\textbf{Leray, J.}
\newline
{}[LR]\quad Discussion d'un probl\`eme de Dirichlet. J. Math. Pures Appl. \textbf{18}, 249--284 (1939).

\medskip

\textbf{Leray, J., et J.-L. Lions}
\newline
{}[LL]\quad Quelques r\'esultats de Vis\v{i}k sur les probl\`emes elliptiques non lin\'eaires par les m\'ethodes de\newline
Minty-Browder. Bull. Soc. Math. France \textbf{93}, 97--107 (1965).

\medskip

\textbf{Leray, J., et J. Schauder}
\newline
{}[LS]\quad Topologie et \'{e}quations fonctionnelles. Ann. Sci. \'{E}cole Norm. Sup. \textbf{51}, 45--78 (1934).
\end{quote}

% PDF page 512
\source{gilbarg-trudinger:page-0512}

\begin{thebibliography}{99}

\bibitem{LW1}
Lewy, H.
\newblock A priori limitations for solutions of Monge-Ampère equations I, II.
\newblock Trans. Amer. Math.
Soc. {\bf 37}, 417--434 (1935); {\bf 41}, 365--374 (1937).

\bibitem{LST}
Lewy, H., and G. Stampacchia
\newblock On existence and smoothness of solutions of some noncoercive variational inequalities.
\newblock Arch. Rational Mech. Anal. {\bf 41}, 241--253 (1971).

\bibitem{LC}
Lichtenstein, L.
\newblock Neuere Entwicklung der Theorie partieller Differentialgleichungen zweiter Ordnung
\newblock vom elliptischen Typus. In: Enc. d. Math. Wissenschaft. 2.3.2, 1277--1334 (1924). Leipzig:
\newblock B. G. Teubner 1923--27.

\bibitem{LB1}
Lieberman, G. M.
\newblock The quasilinear Dirichlet problem with decreased regularity at the boundary. Comm.
\newblock Partial Differential Equations 6, 437--497 (1981).

\bibitem{LB2}
\newblock Solvability of quasilinear elliptic equations with nonlinear boundary conditions. Trans.
\newblock Amer. Math. Soc. 273, 753--765 (1982).

\bibitem{LB3}
\newblock The nonlinear oblique derivative problem for quasilinear elliptic equations. Journal of
\newblock Nonlinear Analysis 8, 49--65 (1984).

\bibitem{LB4}
\newblock The conormal derivative problem for elliptic equations of variational type. J. Differential
\newblock Equations 49, 218--257 (1983).

\bibitem{LT}
Lieberman, G. M., and N. S. Trudinger
\newblock Nonlinear oblique boundary value problems for nonlinear elliptic equations. Trans. Amer.
\newblock Math. Soc. 296, 509--546 (1986).

\bibitem{LP1}
Lions, P.-L.
\newblock Problèmes elliptiques du 2ème ordre non sous forme divergence. Proc. Roy. Soc. Edinburgh
\newblock 84A, 263--271 (1979).

\bibitem{LP2}
\newblock A remark on some elliptic second order problems. Boll. Un. Mat. Ital. (5) 17A, 267--270
\newblock (1980).

\bibitem{LP3}
\newblock Résolution de problèmes elliptiques quasilineaires. Arch. Rational Mech. Anal. 74, 335--353
\newblock (1980).

\bibitem{LP4}
\newblock Résolution analytique des problèmes de Bellman-Dirichlet. Acta Math., 146, 151--166
\newblock (1981).

\bibitem{LP5}
\newblock Generalized solutions of Hamilton-Jacobi equations. London: Pitman (1982).

\bibitem{LP6}
\newblock Sur les équations de Monge-Ampère I. Manuscripta Math. 41, 1--44 (1983).

\bibitem{LP7}
\newblock Sur les équations de Monge-Ampère 2. Arch. Rational Mech. Anal. (1984).

\bibitem{LP8}
\newblock Optimal control of diffusion processes and Hamilton-Jacobi Bellman equations I, II.
\newblock Comm. Part. Diff. Equats. 8, 1101--1174, 1229--1276 (1983).

\bibitem{LP9}
\newblock Optimal control of diffusion processes and Hamilton-Jacobi Bellman equations III. In:
\newblock Collège de France Seminar IV, London: Pitman (1984).

\bibitem{LP10}
\newblock Hamilton-Jacobi-Bellman equations and the optimal control of stochastic systems. Proc.
\newblock Int. Cong. Math. Warsaw, 1983.

\bibitem{LD}
Lions, P.-L., and J. L. Menaldi
\newblock Problèmes de Bellman avec les contrôles dans les coefficients de plus haut degré. C. R. Acad.
\newblock Sci. Paris 287, 409--412 (1978).

\bibitem{LPT}
Lions, P.-L., and N. S. Trudinger
\newblock Linear oblique derivative problems for the uniformly elliptic Hamilton-Jacobi Bellman
\newblock equation. Math. Zeit. 191, 1--15 (1986).

\bibitem{LTU}
Lions P.-L., N. S. Trudinger and J. I. E. Urbas
\newblock The Neumann problem for equations of Monge-Ampère type. Comm. Pure Appl. Math. 39,
\newblock 539--563 (1986).

\bibitem{LSW}
Littman, W., G. Stampacchia, and H. F. Weinberger
\newblock Regular points for elliptic equations with discontinuous coefficients. Ann. Scuola Norm.
\newblock Sup. Pisa (3) 17, 43--77 (1963).

\bibitem{MV1}
Mamedov, I. T.
\newblock On an a priori estimate of the Hölder norm of solutions of quasilinear parabolic equations
\newblock with discontinuous coefficients. Dokl. Akad. Nauk. SSSR 252, (1980) [Russian]. English
\newblock translation in Soviet Math. Dokl. 21, 872--875 (1980).

\bibitem{MV2}
\newblock Behavior near the boundary of solutions of degenerate second order elliptic equations.
\newblock Mat. Zametki 30, 343--352 (1981) [Russian]. English Translation in Math. Notes 30,
\newblock 661--666 (1981).

\end{thebibliography}

% PDF page 513
\source{gilbarg-trudinger:page-0513}

\begin{thebibliography}{99}

\bibitem{MA}
Marcinkiewicz, J.
Sur l’interpolation d’opérations. C. R. Acad. Sci. Paris \textbf{208}, 1272--1273 (1939).

\bibitem{MZ}
Maz'ya, V. G.
Behavior near the boundary of the solution of the Dirichlet problem for a second order
elliptic equation in divergence form. Mat. Zametki \textbf{2}, 209--220 (1967) [Russian]. English
Translation in Math. Notes \textbf{2}, 610--619 (1967).

\bibitem{MSJ}
Melin, A., and J. Sj\"ostrand
Fourier integral operators with complex phase functions and parametrics for an interior
boundary value problem. Comm. Part. Diff. Equats. \textbf{1}, 313--400 (1976).

\bibitem{ME1}
Meyers, N. G.
An $L^p$-estimate for the gradient of solutions of second order elliptic divergence equations.
Ann. Scuola Norm. Sup. Pisa (3) \textbf{17}, 189--206 (1963).

\bibitem{ME2}
An example of non-uniqueness in the theory of quasilinear elliptic equations of second order.
Arch. Rational Mech. Anal. \textbf{14}, 177--179 (1963).

\bibitem{MS1}
Meyers, N. G., and J. Serrin
The exterior Dirichlet problem for second order elliptic partial differential equations.
J. Math. Mech. \textbf{9}, 513--538 (1960).

\bibitem{MS2}
$H = W$. Proc. Nat. Acad. Sci. U.S.A. \textbf{51}, 1055--1056 (1964).

\bibitem{MI1}
Michael, J. H.
A general theory for linear elliptic partial differential equations. J. Differential Equations \textbf{23},
1--29 (1977).

\bibitem{MI2}
Barriers for uniformly elliptic equations and the exterior cone condition. J. Math. Anal.
and Appl. \textbf{79}, 203--217 (1981).

\bibitem{MSI}
Michael, J. H., and L. M. Simon
Sobolev and mean-value inequalities on generalized submanifolds of $R^n$. Comm. Pure
Appl. Math. \textbf{26}, 361--379 (1973).

\bibitem{ML1}
Miller, K.
Barriers on cones for uniformly elliptic operators. Ann. Mat. Pura Appl. (4) \textbf{76}, 93--105
(1967).

\bibitem{ML2}
Exceptional boundary points for the nondivergence equation which are regular for the
Laplace equation and vice-versa. Ann. Scuola Norm. Sup. Pisa (3) \textbf{22}, 315--330 (1968).

\bibitem{ML3}
Extremal barriers on cones with Phragm\`en-Lindel\"of theorems and other applications.
Ann. Mat. Pura Appl. (4) \textbf{90}, 297--329 (1971).

\bibitem{ML4}
Nonequivalence of regular boundary points for the Laplace and nondivergence equations,
even with continuous coefficients. Ann. Scuola Norm. Sup. Pisa (3) \textbf{24}, 159--163 (1970).

\bibitem{MR1}
Miranda, C.
Sul problema misto per le equazioni lineari ellittiche. Ann. Mat. Pura Appl. (4) \textbf{39}, 279--
303 (1955).

\bibitem{MR2}
Partial Differential Equations of Elliptic Type. 2nd ed., Berlin-Heidelberg-New York.
Springer-Verlag 1970.

\bibitem{MD1}
Miranda, M.
Diseguaglianze di Sobolev sulle ipersuperfici minimali. Rend. Sem. Math. Univ. Padova
\textbf{38}, 69--79 (1967).

\bibitem{MD2}
Una maggiorazione integrale per le curvature delle ipersuperfici minimali. Rend. Sem.
Mat. Univ. Padova \textbf{38}, 91--107 (1967).

\bibitem{MD3}
Comportamento delle successioni convergenti di frontiere minimali. Rend. Sem. Mat.
Univ. Padova \textbf{38}, 238--257 (1967).

\bibitem{MD4}
Un principio di massimo forte per le frontiere minimali e una sua applicazione alla risco-
luzione del problema al contorno per l'equazione delle superfici di area minima. Rend.
Sem. Mat. Univ. Padova \textbf{45}, 355--366 (1971).

\bibitem{MD5}
Dirichlet problem with $L^1$ data for the non-homogeneous minimal surface equation.
Indiana Univ. Math. J. \textbf{24}, 227--241 (1974).

\bibitem{MY1}
Morrey, C. B., Jr.
On the solutions of quasi-linear elliptic partial differential equations. Trans. Amer. Math.
Soc. \textbf{43}, 126--166 (1938).

\bibitem{MY2}
Functions of several variables and absolute continuity, II. Duke Math. J. \textbf{6}, 187--215
(1940).

\end{thebibliography}

% PDF page 514
\source{gilbarg-trudinger:page-0514}

\begin{thebibliography}{99}

\bibitem{MY3}
Multiple integral problems in the calculus of variations and related topics. Univ. California
Publ. Math. (N. S.) 1, 1--130 (1943).

\bibitem{MY4}
Second order elliptic equations in several variables and Hölder continuity. Math. Z. 72,
146--164 (1959).

\bibitem{MY5}
Multiple Integrals in the Calculus of Variations. Berlin-Heidelberg-New York: Springer-
Verlag 1966.

Moser, J.
\bibitem{MJ1}
A new proof of de Giorgi's theorem concerning the regularity problem for elliptic differential
equations. Comm. Pure Appl. Math. 13, 457--468 (1960).

\bibitem{MJ2}
On Harnack's theorem for elliptic differential equations. Comm. Pure Appl. Math. 14,
577--591 (1961).

Motzkin, T., and W. Wasow
\bibitem{MW}
On the approximations of linear elliptic differential equations by difference equations with
positive coefficients. J. Math. and Phys. 31, 253--259 (1952).

Nadirashvili, N. S.
\bibitem{ND}
A lemma on the inner derivative, and the uniqueness of the solution of the second boundary
value problem for second order elliptic equations. Dokl. Akad. Nauk SSSR 261, 804--808
(1981) [Russian]. English Translation in Soviet Math. Dokl. 24, 598--601 (1981).

Nash, J.
\bibitem{NA}
Continuity of solutions of parabolic and elliptic equations. Amer. J. Math. 80, 931--954
(1958).

Nečas, J.
\bibitem{NE}
Les méthodes directes en théorie des équations elliptiques. Prague: Éditeurs Academia 1967.

Neumann, J. von
\bibitem{NU}
Über einen Hilfssatz der Variationsrechnung. Abh. Math. Sem. Univ. Hamburg 8, 28--31
(1931).

Nirenberg, L.
\bibitem{NI1}
On nonlinear elliptic partial differential equations and Hölder continuity. Comm. Pure
Appl. Math. 6, 103--156 (1953).

\bibitem{NI2}
Remarks on strongly elliptic partial differential equations. Comm. Pure Appl. Math. 8,
649--675 (1955).

\bibitem{NI3}
On elliptic partial differential equations. Ann. Scuola Norm. Sup. Pisa (3) 13, 115--162
(1959).

\bibitem{NI4}
Elementary remarks on surfaces with curvature of fixed sign. In: Proceedings of Symposia
in Pure Mathematics, Volume III, pp. 181--185. Providence, R.I.: American Mathematical
Society 1961.

Nitsche, J. C. C.
\bibitem{NT}
Vorlesungen über Minimalflächen. Berlin-Heidelberg-New York: Springer-Verlag 1975.

Novrusov, A. A.
\bibitem{NO1}
Regularity of boundary points for elliptic equations with continuous coefficients. Vestn.
Mosk. Univ. Ser. Mat. Mekh., no. 6, 18--25 (1971) [Russian].

\bibitem{NO2}
The modulus of continuity of the solution of the Dirichlet problem at a regular boundary
point. Mat. Zametki 12, 67--72 (1972) [Russian]. English translation in Math. Notes 12,
no. 1, 472--475 (1973).

\bibitem{NO3}
Estimates of the Hölder norm of solutions to quasilinear elliptic equations with discon-
tinuous coefficients. Dokl. Akad. Nauk. SSSR 253, 31--33 (1980) [Russian]. English
translation in Soviet Math. Dokl. 22, 25--28 (1980).

\bibitem{NO4}
Regularity of the boundary points for second order degenerate linear elliptic equations.
Mat. Zametki 30, 353--362 (1981) [Russian]. English Translation in Math. Notes 30,
666--671 (1982).

Oddson, J. K.
\bibitem{OD}
On the boundary point principle for elliptic equations in the plane. Bull. Amer. Math.
Soc. 74, 666--670 (1968).

Oleinik, O. A.
\bibitem{OL}
On properties of solutions of certain boundary problems for equations of elliptic type.
Mat. Sb. (N. S.) 30, 695--702 (1952) [Russian].

Oleinik, O. A., and Radkevich, E. V.
\bibitem{OR}
Second order equations with non-negative characteristic form. (Mathematical Analysis
1969). Moscow: Itogi Nauki 1971 [Russian]. English Translation: Providence, R.I.,
Amer. Math. Soc. 1973.

\end{thebibliography}

% PDF page 515
\source{gilbarg-trudinger:page-0515}
\begin{thebibliography}{99}

\bibitem[OS 1]{OS1}
On the solution of a boundary value problem for linear elliptic equations in an unbounded domain. Vestnik Leningrad. Univ. 16, no. 7, 38--50 (1961) [Russian].

\bibitem[OS 2]{OS2}
A priori estimates of first derivatives of solutions of the Dirichlet problem for nonuniformly elliptic differential equations. Trudy Mat. Inst. Steklov 102, 105--127 (1967) [Russian]. English Translation in Proc. Steklov Inst. Math. 102, 119--144 (1967).

\bibitem[OM 1]{OM1}
On the Gauss curvature of minimal surfaces. Trans. Amer. Math. Soc. 96, 115--128 (1960).

\bibitem[OM 2]{OM2}
A Survey of Minimal Surfaces. New York: Van Nostrand 1969.

\bibitem[PN]{PN}
On the theory of solvability of a problem with oblique derivative. Mat. Sb. (N. S.) 114 (156), 226--268 (1981) [Russian]. English Translation in Math. USSR Sb. 42, 197--235 (1981).

\bibitem[PA]{PA}
Operatori Nelinari. Bucharest: Editura Academiei Republicii Socialiste România 1974.

\bibitem[PE]{PE}
Eine neue Behandlung der Randwertaufgabe für $\Delta u = 0$. Math. Z. 18, 42--54 (1923).

\bibitem[PS]{PS}
On the Hölder continuity of solutions of second order elliptic equations in two variables. Ann. Scuola Norm. Sup. Pisa (3) 26, 391--402 (1972).

\bibitem[PG 1]{PG1}
Monge-Ampère equations of elliptic type. Groningen: Noordhoff (1964).

\bibitem[PG 2]{PG2}
On a regular solution of the $n$-dimensional Minowski problem. Dokl. Akad. Nauk SSSR 119, 785--788 (1971) [Russian]. English Translation in Soviet Math. Dokl. 12, 1192--1196 (1971).

\bibitem[PG 3]{PG3}
On the regularity of generalized solutions of the equation $\det(\partial^2 u/\partial x_i\partial x_j)=\phi(x_1,\dots,x_n)$ $> 0$. Dokl. Akad. Nauk SSSR 200, 543--547 (1971) [Russian]. English Translation in Soviet Math. Dokl. 12, 1436--1440 (1971).

\bibitem[PG 4]{PG4}
The Dirichlet problem for the $n$-dimensional analogue of the Monge-Ampère equation. Dokl. Akad. Nauk SSSR 201, 790--793 (1971) [Russian]. English Translation in Soviet Math. Dokl. 12, 1727--1731 (1971).

\bibitem[PG 5]{PG5}
The Minkowski multidimensional problem. New York: J. Wiley (1978).

\bibitem[PW]{PW}
Maximum Principles in Differential Equations. Englewood Cliffs, N.J.: Prentice-Hall 1967.

\bibitem[PU 1]{PU1}
Su una limitazione per soluzioni di equazioni ellittiche. Boll. Un. Mat. Ital. (3) 21, 228--233 (1966).

\bibitem[PU 2]{PU2}
Operatori ellittici estremanti. Ann. Mat. Pura Appl. (4) 72, 141--170 (1966).

\bibitem[PU 3]{PU3}
Limitazioni per soluzioni di equazioni ellittiche. Ann. Mat. Pura Appl. (4) 74, 15--30 (1966).

\bibitem[RA]{RA}
Geometrische Betrachtungen über zweidimensionale reguläre Variationsprobleme. Acta Litt. Sci. Univ. Szeged 2, 228--253 (1924--26).

\bibitem[RE]{RE}
Ein Satz über mittlere Konvergenz. Nachr. Akad. Wiss. Göttingen Math.-Phys. Kl. 30--35 (1930).

\bibitem[RZ]{RZ}
Sur les fonctions conjuguées. Math. Z. 27, 218--244 (1927).

\bibitem[RO]{RO}
The Sobolev inequalities with best possible constants. In: Analysis Seminar at California Institute of Technology, 1966.

\bibitem[RY]{RY}
Real Analysis. 2nd ed., London: Macmillan 1970.

\bibitem[SF]{SF}
Harnack inequalities for elliptic equations and Hölder continuity of their solutions. Zap. Nauch. Sem. Leningrad. Otdel. Mat. Inst. Steklov. (LOMI) 96, 272--287 (1980) [Russian]. English translation in J. Soviet Math. 21, 851--863 (1983).

\end{thebibliography}

% PDF page 516
\source{gilbarg-trudinger:page-0516}

\begin{thebibliography}{99}

\bibitem{SH}
Schaefer, H.
Über die Methode der a priori Schranken. Math. Ann. \textbf{129}, 415--416 (1955).

\bibitem{SC1}
Schauder, J.
Der Fixpunktsatz in Funktionalräumen. Studia Math. \textbf{2}, 171--180 (1930).

\bibitem{SC2}
Über den Zusammenhang zwischen der Eindeutigkeit und Lösbarkeit partieller Differentialgleichungen zweiter Ordnung vom elliptischen Typus. Math. Ann. \textbf{106}, 661--721 (1932).

\bibitem{SC3}
Über das Dirichletsche Problem im Großen für nicht-lineare elliptische Differentialgleichungen. Math. Z. \textbf{37}, 623--634, 768 (1933).

\bibitem{SC4}
Über lineare elliptische Differentialgleichungen zweiter Ordnung. Math. Z. \textbf{38}, 257--282 (1934).

\bibitem{SC5}
Numerische Abschätzungen in elliptischen linearen Differentialgleichungen. Studia Math. \textbf{5}, 34--42 (1935).

\bibitem{SZ1}
Schulz, F.
Über die Beschränktheit der zweiten Ableitungen der Lösungen nichtlinearer elliptischer Differentialgleichungen. Math. Z. \textbf{175}, 181--188 (1980).

\bibitem{SZ2}
A remark on fully nonlinear, concave elliptic equations. Proc. Centre for Math. Anal. Aust. Nat. Univ. \textbf{8}, 202--207 (1984).

\bibitem{SE1}
Serrin, J.
On the Harnack inequality for linear elliptic equations. J. Analyse Math. \textbf{4}, 292--308 (1955/56).

\bibitem{SE2}
Local behavior of solutions of quasi-linear elliptic equations. Acta Math. \textbf{111}, 247--302 (1964).

\bibitem{SE3}
The problem of Dirichlet for quasilinear elliptic differential equations with many independent variables. Philos. Trans. Roy. Soc. London Ser. A \textbf{264}, 413--496 (1969).

\bibitem{SE4}
Gradient estimates for solutions of nonlinear elliptic and parabolic equations. In: Contributions to Nonlinear Functional Analysis, pp. 565--601. New York: Academic Press 1971.

\bibitem{SE5}
Nonlinear elliptic equations of second order. Lecture notes (unpublished), Symposium on Partial Differential Equations at Berkeley, 1971.

\bibitem{SI1}
Simon, L. M.
Global estimates of Hölder continuity for a class of divergence-form elliptic equations. Arch. Rational Mech. Anal. \textbf{56}, 253--272 (1974).

\bibitem{SI2}
Boundary regularity for solutions of the non-parametric least area problem. Ann. of Math. \textbf{103}, 429--455 (1976).

\bibitem{SI3}
Interior gradient bounds for non-uniformly elliptic equations. Indiana Univ. Math. J. \textbf{25}, 821--855 (1976).

\bibitem{SI4}
Equations of mean curvature type in 2 independent variables. Pacific J. Math. \textbf{69}, 245--268 (1977).

\bibitem{SI5}
Remarks on curvature estimates for minimal hypersurfaces. Duke Math. J. \textbf{43}, 545--553 (1976).

\bibitem{SI6}
A Hölder estimate for quasiconformal mappings between surfaces in Euclidean space, with application to graphs having quasiconformal Gauss map. Acta Math. (1977).

\bibitem{SI7}
Survey lectures on minimal submanifolds. In: Seminar on Minimal Submanifolds. Annals of Math. Studies \textbf{103}, Princeton (1983).

\bibitem{SI8}
Lectures on geometric measure theory. Proc. Centre for Mathematical Analysis. Australian Nat. Univ. 3(1983).

\bibitem{SM}
Simons, J.
Minimal varieties in riemannian manifolds. Ann. of Math. (2) \textbf{88}, 62--105 (1968).

\bibitem{SS}
Simon, L. M., and J. Spruck
Existence and regularity of a capillary surface with prescribed contact angle. Arch. Rational Mech. Anal. \textbf{61}, 19--34 (1976).

\bibitem{SJ}
Sjöstrand, J.
Operators of principal type with interior boundary conditions. Acta Math. \textbf{130}, 1--51 (1973).

\bibitem{SL}
Slobodeckii, L. M.
Estimate in L$_p$ of solutions of elliptic systems. Dokl. Akad. Nauk SSSR \textbf{123}, 616--619 (1958)
{}[Russian].

\bibitem{SO1}
Sobolev, S. L.
On a theorem of functional analysis. Mat. Sb. (N.S.) \textbf{4} (46), 471--497 (1938) [Russian].
English Translation in Amer. Math. Soc. Transl. (2) \textbf{34}, 39--68 (1963).

\end{thebibliography}

% PDF page 517
\source{gilbarg-trudinger:page-0517}

\begin{flushleft}
Bibliography \hfill 505

{}[SO 2]\quad Applications of Functional Analysis in Mathematical Physics. Leningrad: Izdat. Leningrad.\\
Gos. Univ. 1950 [Russian]. English Translation: Translations of Mathematical Mono-\\
graphs, Vol. 7. Providence, R.I.: American Mathematical Society 1963.

Spruck, J.\\
{}[SP]\quad Gauss curvature estimates for surfaces of constant mean curvature. Comm. Pure Appl.\\
Math. 27, 547--557 (1974).

Stampacchia, G.\\
{}[ST 1]\quad Contributi alla regolarizzazione delle soluzioni dei problemi al contorno per equazioni\\
del secondo ordine ellittiche. Ann. Scuola Norm. Sup. Pisa (3) 12, 223--245 (1958).\\
{}[ST 2]\quad I problemi al contorno per le equazioni differenziali di tipo ellittico. In: Atti VI Congr.\\
Un. Mat. Ital. (Naples, 1959), pp. 21--44. Rome: Edizioni Cremonese 1960.\\
{}[ST 3]\quad Problemi al contorno ellittici, con dati discontinui, dotati di soluzioni hölderiane. Ann.\\
Mat. Pura Appl. (4) 51, 1--37 (1960).\\
{}[ST 4]\quad Le problème de Dirichlet pour les équations elliptiques du second ordre à coefficients\\
discontinues. Ann. Inst. Fourier (Grenoble) 15, fasc. 1, 189--258 (1965).\\
{}[ST 5]\quad Équations elliptiques du second ordre à coefficients discontinues. Séminaire de Mathé-\\
matiques Supérieures, No. 16 (Été, 1965). Montreal, Que.: Les Presses de l'Université de\\
Montréal 1966.

Stein, E.\\
{}[SN]\quad Singular integrals and differentiability properties of functions. Princeton: Princeton Uni-\\
versity Press (1970.

Sternberg, S.\\
{}[SB]\quad Lectures on Differential Geometry. Englewood Cliffs, N.J.: Prentice-Hall 1964.

Talenti, G.\\
{}[TA 1]\quad Equazioni lineari ellittiche in due variabili. Le Matematiche 21, 339--376 (1966).\\
{}[TA 2]\quad Best constant in Sobolev inequality. Ann. Mat. Pura Appl. 110, 353--372 (1976).\\
{}[TA 3]\quad Elliptic equations and rearrangements. Ann. Scuola Norm. Sup. Pisa (4) 3, 697--718 (1976).\\
{}[TA 4]\quad Some estimates of solutions to Monge-Ampère type equations in dimension two. Ann.\\
Scuola Norm. Sup. Pisa (4) 8, 183--230 (1981).

Temam, R.\\
{}[TE]\quad Solutions généralisées de certaines équations du type hypersurface minima. Arch. Rational\\
Mech. Anal. 44, 121--156 (1971).

Trudinger, N. S.\\
{}[TR 1]\quad On Harnack type inequalities and their application to quasilinear elliptic equations.\\
Comm. Pure Appl. Math. 20, 721--747 (1967).\\
{}[TR 2]\quad On imbeddings into Orlicz spaces and some applications. J. Math. Mech. 17, 473--483\\
(1967).\\
{}[TR 3]\quad Some existence theorems for quasi-linear, non-uniformly elliptic equations in divergence\\
form. J. Math. Mech. 18, 909--919 (1968/69).\\
{}[TR 4]\quad On the regularity of generalized solutions of linear, non-uniformly elliptic equations.\\
Arch. Rational Mech. Anal. 42, 50--62 (1971).\\
{}[TR 5]\quad The boundary gradient estimate for quasilinear elliptic and parabolic differential equations.\\
Indiana Univ. Math. J. 21, 657--670 (1972).\\
{}[TR 6]\quad A new proof of the interior gradient bound for the minimal surface equation in $n$ dimensions.\\
Proc. Nat. Acad. Sci. U.S.A. 69, 821--823 (1972).\\
{}[TR 7]\quad Linear elliptic operators with measurable coefficients. Ann. Scuola Norm. Sup. Pisa (3)\\
27, 265--308 (1973).\\
{}[TR 8]\quad Gradient estimates and mean curvature. Math. Z. 131, 165--175 (1973).\\
{}[TR 9]\quad A sharp inequality for subharmonic functions on two-dimensional manifolds. Math. Z.\\
133, 75--79 (1973).\\
{}[TR 10]\quad On the comparison principle for quasilinear divergence structure equations. Arch. Rational\\
Mech. Anal. 57, 128--133 (1974).\\
{}[TR 11]\quad Maximum principles for linear, non-uniformly elliptic operators with measurable\\
coefficients. Math. Z. 156, 291--301 (1977).\\
{}[TR 12]\quad Local estimates for subsolutions and supersolutions of general second order elliptic quasi-\\
linear equations. Invent. Math. 61, 67--79 (1980).\\
{}[TR 13]\quad Fully nonlinear, uniformly elliptic equations under natural structure conditions. Trans.\\
Amer. Math. Soc. 28, 217--231 (1983).
\end{flushleft}

% PDF page 518
\source{gilbarg-trudinger:page-0518}
\begin{center}
506\hfill Bibliography
\end{center}

\noindent [TR 14]\quad Boundary value problems for fully nonlinear elliptic equations. Proc. Centre for Math. Anal. Aust. Nat. Univ. 8, 65--83 (1984).

\noindent [TR 15]\quad On an interpolation inequality and its application to nonlinear elliptic equations. Proc. Amer. Math. Soc. 95, 73--78 (1985).

\noindent Trudinger, N. S., and J. Urbas

\noindent [TU]\quad On the Dirichlet problem for the prescribed Gauss curvature equation, Bull. Australian Math. Soc. 28, 751--770 (1983).

\noindent Ural'tseva, N.

\noindent [UR]\quad The solvability of the capillarity problem. Vestnik Leningrad. Univ. no. 19, 54--64 (1973) [Russian]. English Translation in Vestnik Leningrad Univ. 6, 363--375 (1979).

\noindent Wahl, W. von

\noindent [WA]\quad \c{U}ber quasilineare elliptische Differentialgleichungen in der Ebene. Manuscripta Math. 8, 59--67 (1973).

\noindent Weinberger, H. F.

\noindent [WE]\quad Symmetrization in uniformly elliptic problems. In: Studies in Mathematical Analysis and Related Topics, pp. 424--428. Stanford, Calif.: Stanford University Press 1962.

\noindent Widman, K.-O.

\noindent [WI 1]\quad Inequalities for the Green function and boundary continuity of the gradient of solutions of elliptic differential equations. Math. Scand. 21, 17--37 (1967).

\noindent [WI 2]\quad A quantitative form of the maximum principle for elliptic partial differential equations with coefficients in $L_\infty$. Comm. Pure Appl. Math. 21, 507--513 (1968).

\noindent [WI 3]\quad On the H\"older continuity of solutions of elliptic partial differential equations in two variables with coefficients in $L_\infty$. Comm. Pure Appl. Math. 22, 669--682 (1969).

\noindent Wiener, N.

\noindent [WN]\quad The Dirichlet problem. J. Math. and Phys. 3, 127--146 (1924).

\noindent Williams, G. H.

\noindent [WL]\quad Existence of solutions for nonlinear obstacle problems. Math. Z. 154, 51--65 (1977).

\noindent Winzell, B.

\noindent [WZ 1]\quad The oblique derivative problem, I. Math. Ann. 229, 267--278 (1977).

\noindent [WZ 2]\quad The oblique derivative problem, II. Ark. Mat. 17, 107--122 (1979).

\noindent Yau, S. T.

\noindent [YA]\quad On the Ricci curvature of a compact K\"ahler manifold and the complex Monge-Amp\`ere equation. Comm. Pure Appl. Math. 31, 339--411 (1978).

\noindent Yosida, K.

\noindent [YO]\quad Functional Analysis. 4th ed., Berlin-Heidelberg-New York: Springer-Verlag 1974.

% PDF page 519
\source{gilbarg-trudinger:page-0519}
\begin{center}
Epilogue
\end{center}

\noindent This book has been devoted to the theory of elliptic, second order, partial differential equations with emphasis on the Dirichlet problem for linear and quasilinear equations. Its second edition in 1983 included an introductory chapter on fully nonlinear elliptic equations as the Krylov-Safonov H\"older estimates had recently opened up the higher dimensional theory. This was analogous to the role of the De Giorgi-Nash H\"older estimates in the higher dimensional quasilinear theory about a quarter century earlier. It should not be surprising that the fully nonlinear theory, with its rich applications to stochastic optimization and geometry, has blossomed since our second edition appeared.

\noindent We comment briefly on some of the main developments.

\noindent \textbf{Viscosity solutions.} The notion of viscosity solution, introduced for first order equations by Crandall and Lions, ([LP 5], [CL], [CIL]), was extended to second order equations, with dramatic consequences in the wake of a breakthrough by Jensen [JEN] enabling approximation by semi-convex or semi-concave functions. The concept of viscosity subsolution relates to that of subharmonic function, as introduced in Sections 2.8 and 6.3. Using the terminology of Chapter 17, if $F \in C^0(\Gamma)$ is monotone increasing with respect to $r \in R^{n \times n}$, we call a function $u \in C^0(\Omega)$ a viscosity subsolution (supersolution) of equation (17.1) in $\Omega$, if for every point $y \in \Omega$ and function $v \in C^2(\Omega)$ satisfying $u \leq v (\geq v)$ in $\Omega$ and $u(y)=v(y)$, we have $F[v](y) \geq 0,\ (\leq 0)$. It is readily seen that for linear elliptic equations, $Lu = f$, this notion coincides with that of Section 6.3. Moreover, Ishii [IS] showed that the Perron process could be used to infer existence of viscosity solutions of the Dirichlet problem, with the aid of comparison principles extending those of Jensen [JEN]. Various aspects of this theory and its widespread applications are described in the expository works [CIL], [FLS].

\noindent \textbf{Uniformly elliptic equations.} The second derivative H\"older estimates in Sections 17.4 and 17.8 were improved by Safonov [SE 2], [SF 4] and Caffarelli [CAF] by perturbation arguments from the special case (17.32). As a by-product, simpler proofs of the Schauder estimates for linear equations were obtained by various authors, including in particular an ``$L^\infty$-Campanato'' method [SF 4], [KV 9]. Caffarelli also deduced $L^p$-estimates for second derivatives for $p>n$; (see [CC]). The basic theory is also covered in [KV 7], [TR 16].

% PDF page 520
\source{gilbarg-trudinger:page-0520}


Non-uniformly elliptic equations. The Monge-Ampère and Gauss curvature equations are special cases of Hessian and curvature equations determined by the elementary symmetric functions. The classical Dirichlet problem is treated in the works [CNS 2, 3], [IC 4], [KV 7, 8], [TR 17].

Quasilinear equations. The special treatment of the two dimensional case stems from Morrey's gradient estimate, Theorem 12.4. By showing that the exponent in the Hölder estimate, Corollary 9.24, can be arbitrarily small, Safonov [SF 3], confirmed that this approach is not extendible to higher dimensions.

Finally, we note that Korevaar [KOR] showed that the interior gradient bound for the minimal surface and prescribed mean curvature equations can be deduced from the maximum principle along the lines of Section 15.3. The resultant bound is not as precise as that in Theorem 16.5.

\bigskip

\textbf{Bibliography}

\begin{description}
\item[CC] Cabré, X., and L. Caffarelli, Fully nonlinear elliptic equations. Amer. Math. Soc. Colloquium Publications 43 (1995).
\item[CAF] Caffarelli, L., Interior a priori estimates for solutions of fully non-linear equations. Ann. Math. 130, 189--213 (1989).
\item[CNS 2] Caffarelli, L., L. Nirenberg and J. Spruck, The Dirichlet problem for nonlinear second order equations III, Functions of the eigenvalues of the Hessian. Acta Math. 155, 261--301 (1985).
\item[CNS 3] Caffarelli, L., L. Nirenberg and J. Spruck, The Dirichlet problem for nonlinear second order equations V, The Dirichlet problem for Weingarten surfaces. Comm. Pure Appl. Math. 41, 47--70 (1988).
\item[CIL] Crandall, M. G., H. Ishii and P.-L. Lions, User's guide to viscosity solutions of second order partial differential equations. Bull. Amer. Math. Soc. 27, 1--67 (1992).
\item[CL] Crandall, M. G., and P.-L. Lions, Viscosity solutions of Hamilton-Jacobi equations. Trans. Amer. Math. Soc. 277, 1--42 (1983).
\item[FLS] Fleming, W. H., and H. M. Soner, Controlled Markov processes and viscosity solutions. New York: Springer 1993.
\item[IS] Ishii, H., On uniqueness and existence of viscosity solutions of fully nonlinear second order PDE's. Comm. Pure Appl. Math. 42, 14--45 (1989).
\item[IC 4] Ivo\v{c}kina, N. M., The Dirichlet problem for the curvature equation of order m, Algebra; Analiz 2, 192--217 (1990) [Russian]. English translation: Leningrad Math. J. 2, 631--654 (1991).
\item[JEN] Jensen, R., The maximum principle for viscosity solutions of fully nonlinear second order partial differential equations. Arch. Rat. Mech. Anal. 101, 1--27 (1988).
\item[KOR] Korevaar, N., An easy proof of the interior gradient bound for solutions to the prescribed mean curvature equation, In: Nonlinear Functional Analysis and its Applications. Proc. Symp. Pure Math. 45, (2), 81--90. Providence: Amer. Math. Soc. 1986.
\item[KV 7] Krylov, N. V., Nonlinear elliptic and parabolic equations of the second order. Moscow: Nauka 1985 [Russian]. English translation: Dordrecht: Reidel 1987.
\item[KV 8] Krylov, N. V., On the general notion of fully nonlinear second-order elliptic equations. Trans. Amer. Math. Soc. 347, 857--895 (1995).
\item[KV 9] Krylov, N. V., Lectures on elliptic and parabolic equations in H\"older spaces. Providence: Amer. Math. Soc. 1996.
\item[SF 2] Safonov, M. V., On the classical solution of Bellman's elliptic equation. Dokl. Akad. Nauk SSSR 278, 810--813 (1984) [Russian]. English translation: Soviet Math. Dokl. 30, 482--485 (1984).
\end{description}


% PDF page 521
\source{gilbarg-trudinger:page-0521}
Epilogue

\begin{thebibliography}{99}

\bibitem[SF 3]{SF3} Safonov, M. V., Unimprovability of estimates of Hölder constants for solutions of linear elliptic equations with measurable coefficients. Mat. Sb. \textbf{132}, 275--288 (1987) [Russian]. English translation: Math. USSR Sbornik \textbf{60}, 269--281 (1988).

\bibitem[SF 4]{SF4} Safonov, M. V., Nonlinear elliptic equations of second order. Lecture Notes, Univ. Firenze 1991.

\bibitem[TR 16]{TR16} Trudinger, N. S., Lectures on nonlinear elliptic equations of the second order. Lectures in Mathematical Sciences, Univ. of Tokyo 1995.

\bibitem[TR 17]{TR17} Trudinger, N. S., On the Dirichlet problem for Hessian equations. Acta Math. \textbf{175}, 151--164 (1995).

\end{thebibliography}

% PDF page 522
\source{gilbarg-trudinger:page-0522}
\begin{flushleft}
\textbf{Subject Index}

\vspace{2.4em}

adjoint \hspace{1em} 79, 84\\
formal- \hspace{1em} 182\\
apriori estimates \hspace{1em} 3\\
\hspace*{1.5em} see gradient estimates, Hölder estimates,\\
\hspace*{1.5em} linear elliptic equations, maximum prin-\\
\hspace*{1.5em} ciples, quasilinear elliptic equations,\\
\hspace*{1.5em} Schauder estimates\\
area magnification factor \hspace{1em} 414\\[1.8em]

Banach space \hspace{1em} 73\\
barrier\\
\hspace*{1.5em} for Laplace's equation \hspace{1em} 25\\
\hspace*{1.5em} for linear equations \hspace{1em} 104\\
\hspace*{1.5em} local- \hspace{1em} 25, 106\\
\hspace*{1.5em} for non-uniformly elliptic equations \hspace{1em} 118\\
\hspace*{1.5em} for quasilinear equations \hspace{1em} 333ff., 344\\
\hspace*{1.5em} upper-(lower-) \hspace{1em} 104, 334\\
Bellman equation \hspace{1em} 443\\
Bernstein's theorem\\
\hspace*{1.5em} for equations of mean curvature type \hspace{1em} 412\\
\hspace*{1.5em} for minimal surface equation \hspace{1em} 438\\
bilinear form\\
\hspace*{1.5em} bounded- \hspace{1em} 83\\
\hspace*{1.5em} coercive- \hspace{1em} 83\\
boundary manifold \hspace{1em} 283\\
boundary point lemma \hspace{1em} 34, 46\\
bounded linear mapping \hspace{1em} 74\\
bounded slope condition \hspace{1em} 283, 309, 314, 317\\
Brower fixed point theorem \hspace{1em} 279\\
Calderon-Zygmund inequality \hspace{1em} 230\\
capacity \hspace{1em} 27\\
capillarity, equation of \hspace{1em} 262, 485\\
Cauchy's inequality \hspace{1em} 145\\
co-area formula \hspace{1em} 416\\
compact imbedding of Banach spaces \hspace{1em} 167\\
compact mapping \hspace{1em} 75, 280\\
\hspace*{1.5em} spectrum of- \hspace{1em} 78\\
comparison principles \hspace{1em} 263, 268, 347\\
completely continuous mapping\\
\hspace*{1.5em} see compact mapping\\
conductor potential \hspace{1em} 27\\[1.8em]

cone condition\\
\hspace*{1.5em} exterior- \hspace{1em} 29, 203\\
\hspace*{1.5em} uniform exterior- \hspace{1em} 205\\
\hspace*{1.5em} uniform interior- \hspace{1em} 158\\
conormal derivative problem \hspace{1em} 485\\
constant mean curvature equation\\
\hspace*{1.5em} Dirichlet problem, solvability \hspace{1em} 409\\
\hspace*{1.5em} see also quasilinear elliptic equations\\
contact set \hspace{1em} 220\\
continuous imbedding \hspace{1em} 158\\
contraction mapping \hspace{1em} 74\\
cube decomposition \hspace{1em} 226\\[1.8em]

difference quotient \hspace{1em} 109, 168\\
Dini continuity \hspace{1em} 70\\
Dirichlet integral \hspace{1em} 295, 418\\
Dirichlet problem\\
\hspace*{1.5em} see constant mean curvature equation, Euler-\\
\hspace*{1.5em} Lagrange equation, Laplace's equation,\\
\hspace*{1.5em} linear elliptic equations, equations of\\
\hspace*{1.5em} mean curvature type, minimal surface\\
\hspace*{1.5em} equation, Poisson's equation, prescribed\\
\hspace*{1.5em} mean curvature equation, quasilinear el-\\
\hspace*{1.5em} liptic equations, fully nonlinear equations,\\
\hspace*{1.5em} Monge-Ampère equation, prescribed\\
\hspace*{1.5em} Gauss curvature equation\\
distance function \hspace{1em} 354\\
distribution function \hspace{1em} 227\\
divergence theorem \hspace{1em} 13\\
domain\\
\hspace*{1.5em} with $C^{k,\alpha}$ boundary portion \hspace{1em} 94\\
\hspace*{1.5em} of class $C^{k,\alpha}$ \hspace{1em} 94\\
dual space \hspace{1em} 79\\[1.8em]

eigenvalue \hspace{1em} 78, 212\\
\hspace*{1.5em} minimum \hspace{1em} 214\\
\hspace*{1.5em} multiplicity \hspace{1em} 78\\
eigenvector \hspace{1em} 78\\
elliptic equations (operators),\\
\hspace*{1.5em} see linear elliptic equations (operators), quasi-\\
\hspace*{1.5em} linear elliptic equations (operators), fully\\
\hspace*{1.5em} nonlinear equations
\end{flushleft}

% PDF page 523
\source{gilbarg-trudinger:page-0523}
\begin{quote}
512 \hfill Subject Index

\vspace{1em}

\begin{tabular}{@{}p{0.44\textwidth}p{0.46\textwidth}@{}}
\textbf{elliptic parametric functional} \hfill 435 &
\textbf{Harnack inequality} \\
\textbf{extension of a function} \hfill 136f. &
\quad for equations in two variables \hfill 41 \\
\textbf{exterior boundary value problem} \hfill 141 &
\quad for harmonic functions \hfill 16, 29 \\
\textbf{Euler-Lagrange equation} \hfill 288 &
\quad for linear equations \hfill 199 \\
\quad \textbf{Dirichlet problem} \hfill 289ff. &
\quad for strong solutions \hfill 250 \\
\quad \textbf{global gradient estimate} \hfill 368 &
\quad weak- \hfill 194, 217 \\
\quad \emph{see also} quasilinear elliptic equations &
\textbf{Hessian matrix} \hfill 354 \\
&
\textbf{Hilbert space} \hfill 80 \\
\textbf{Fréchet derivative} \hfill 446 &
\textbf{Hölder coefficient} \hfill 52 \\
\textbf{Fredholm alternative} \hfill 76, 84 &
\textbf{Hölder conjugate} \hfill 147 \\
\textbf{fully nonlinear equations} \hfill 441 &
\textbf{Hölder continuity} \hfill 52 \\
\quad \textbf{Dirichlet problem} \hfill 463, 464, 466 &
\quad \textbf{De Giorgi estimate} \hfill 200 \\
\quad \textbf{Hölder estimates for second derivatives} \hfill 453, 461 &
\quad \textbf{for weak solutions} \hfill 200 \\
\quad \textbf{in two variables} \hfill 450 &
\textbf{Hölder estimates} \\
\textbf{fundamental solution} \hfill 17 &
\quad for equations in two variables \hfill 302ff. \\
&
\quad for Gauss map \hfill 429 \\
\textbf{Gauss curvature} \hfill 424 &
\quad for Newtonian potential \hfill 57 \\
\textbf{Gauss map} \hfill 423 &
\quad for Poisson's equation \hfill 60, 62, 123ff., 125 \\
\quad \textbf{Hölder estimates} \hfill 429 &
\quad for quasiconformal maps \hfill 299f. \\
\textbf{generalized Dirichlet problem} \hfill 4, 178 &
\quad on hypersurface \hfill 422 \\
\quad \textbf{solvability} \hfill 144 &
\quad for quasilinear equations \hfill 281, 287, 320, 323, \\
\quad \textbf{uniqueness} \hfill 180 &
\quad 328, 331, 378 \\
\quad \emph{see also} Dirichlet problem &
\quad for strong solutions \hfill 250 \\
\textbf{generalized solution} \hfill 3, 144, 177 &
\quad \emph{see also} gradient estimates, Schauder estimates \\
\quad \emph{see also} weak solution &
\textbf{Hölder's inequality} \hfill 145 \\
\textbf{gradient estimates} &
\quad \textbf{generalized} \hfill 146 \\
\quad for equations of mean curvature type \hfill 431 &
\textbf{Hölder spaces} \hfill 52 \\
\quad for Euler-Lagrange equation \hfill 368 &
\quad \textbf{norms and seminorms} \hfill 53, 61, 66, 90, 96 \\
\quad for harmonic functions \hfill 23, 29 &
\textbf{hypersurface} \hfill 388 \\
\quad for linear equations \hfill 48 &
\\
\quad for minimal surface equation \hfill 339, 341, 406 &
\textbf{implicit function theorem in Banach spaces} \hfill 447 \\
\quad for Poisson's equation \hfill 41 &
\textbf{inner product} \hfill 80 \\
\quad for prescribed mean curvature equation \hfill 340, 342, 345, 367, 406f. &
\textbf{inner product space} \hfill 80 \\
\quad for quasilinear equations &
\textbf{interpolation inequalities} \\
\quad \quad in divergence form \hfill 320, 323, 374, 379, 386 &
\quad in Hölder spaces \hfill 89, 95, 130ff., 143 \\
\quad \quad in general form \hfill 328, 331, 337, 338ff. &
\quad in $L^p$ spaces \hfill 146 \\
\quad \quad in two variables \hfill 302, 323f. &
\quad in Sobolev spaces \hfill 171, 173 \\
\textbf{Green's function} &
\\
\quad for a ball \hfill 19 &
\textbf{John-Nirenberg inequality} \hfill 166 \\
\quad \textbf{Dirichlet-} \hfill 19 &
\\
\quad for oblique derivative problem \hfill 121 &
\textbf{Kellogg's theorem} \hfill 66 \\
\textbf{Green's identities} \hfill 17 &
\textbf{Kelvin transform} \hfill 67 \\
\textbf{Green's operator} \hfill 183 &
\textbf{Kondrachov compactness theorem} \hfill 167 \\
\textbf{Green's representation formula} \hfill 18 &
\\
&
\emph{$L^p$ spaces} \hfill 145 \\
\textbf{harmonic function} \hfill 13 &
\quad \textbf{embedding theorem} \hfill 159 \\
\quad \textbf{convergence theorems} \hfill 21f. &
\quad \textbf{norm} \hfill 145 \\
\quad \textbf{derivative estimates} \hfill 23, 29 &
\quad \textbf{$L^p$ estimates} \hfill 219, 225, 235 \\
\quad \textbf{on hypersurface} \hfill 393 &
\textbf{Laplace-Beltrami operator} \hfill 391 \\
\quad \textbf{mean value property} \hfill 14, 21 &
\textbf{Laplace's equation} \hfill 1, 13ff. \\
\quad \textbf{weakly-} \hfill 29 &
\quad \textbf{Dirichlet problem} \\
\quad \textbf{harmonic lifting} \hfill 24 &
\quad \quad \textbf{solvability} \hfill 20, 26 \\
\textbf{Harnack's convergence theorem} \hfill 22 &
\quad \quad \textbf{uniqueness} \hfill 15 \\
&
\quad \textbf{fundamental solution} \hfill 17
\end{tabular}
\end{quote}

% PDF page 524
\source{gilbarg-trudinger:page-0524}
\begin{itemize}
\item[] \textbf{Subject Index}

\item[] maximum principle \hfill 15
\item[] \quad see also harmonic functions
\item[] Laplacian \hfill 13, 391
\item[] Lax-Milgram theorem \hfill 83
\item[] Leray-Schauder fixed point theorem \hfill 280, 286
\item[] linear elliptic equations (operators) \hfill 1, 31f., 87ff., 177ff.
\item[] \quad classical solutions
\item[] \qquad boundary regularity \hfill 111, 140
\item[] \qquad Dirichlet problem
\item[] \qquad \quad Fredholm alternative \hfill 107
\item[] \qquad \quad solvability \hfill 100ff., 106, 112, 115f., 142, 188
\item[] \qquad \quad uniqueness \hfill 33, 37
\item[] \qquad global bound \hfill 36
\item[] \qquad global regularity \hfill 111
\item[] \qquad interior regularity \hfill 109f.
\item[] \qquad maximum principles \hfill 32ff., 143
\item[] \qquad Neumann problem \hfill 35
\item[] \qquad oblique derivative problem \hfill 125ff.
\item[] \qquad Schauder estimates \hfill 90, 93, 95ff., 112, 141f.
\item[] \quad in divergence form \hfill 3, 45f., 177ff.
\item[] \quad see below weak solutions
\item[] \quad strong solutions \hfill 219ff.
\item[] \quad in two variables
\item[] \qquad Hölder estimates \hfill 302f.
\item[] \quad weak solutions \hfill 3, 177, 188
\item[] \qquad boundary estimates \hfill 202ff.
\item[] \qquad Dirichlet problem
\item[] \qquad \quad Fredholm alternative \hfill 182
\item[] \qquad \quad solvability \hfill 181, 206
\item[] \qquad \quad uniqueness \hfill 180
\item[] \qquad global bound \hfill 188, 191
\item[] \qquad global regularity \hfill 186f.
\item[] \qquad Hölder continuity \hfill 200, 202
\item[] \qquad interior regularity \hfill 183ff.
\item[] \qquad local bound \hfill 194
\item[] \qquad maximum principles \hfill 45, 179, 198, 216
\item[] linear mapping, bounded \hfill 74
\item[] Lipschitz continuity \hfill 52
\item[] Liouville's theorem \hfill 30, 44

\item[] $M^p$ spaces \hfill 164
\item[] Marcinkiewicz interpolation theorem \hfill 228
\item[] maximum principles
\item[] \quad for fully nonlinear equations \hfill 443
\item[] \quad for Laplace's equation \hfill 15
\item[] \quad for linear equations
\item[] \qquad classical solutions \hfill 32ff., 143
\item[] \qquad strong solutions \hfill 220
\item[] \qquad weak solutions \hfill 45, 179, 198, 216
\item[] \quad for quasilinear equations \hfill 264, 271ff., 277f.
\item[] \qquad gradient- \hfill 362
\item[] mean curvature \hfill 354, 386, 390
\item[] \qquad generalized- \hfill 342
\item[] mean curvature type, equations of \hfill 410, 429ff.
\item[] \qquad curvature estimate \hfill 430
\item[] \qquad gradient estimate \hfill 431
\item[] mean value inequality \hfill 14, 29, 71
\item[] \quad on hypersurface \hfill 393, 396, 439

\item[] mean value theorem \hfill 14
\item[] method of continuity \hfill 75, 448
\item[] minimal surface \hfill 357
\item[] minimal surface equation (operator) \hfill 1
\item[] \qquad curvature estimate \hfill 437
\item[] \qquad Dirichlet problem \hfill 352, 407
\item[] \qquad gradient estimates \hfill 339, 341, 406
\item[] \quad see also quasilinear elliptic equations (operators)
\item[] minimal surface type, equations of
\item[] \quad see also mean curvature type, equations of \hfill 437
\item[] mixed boundary condition \hfill 47
\item[] mixed boundary value problem \hfill 215
\item[] mollifier \hfill 147
\item[] Monge-Ampere equation \hfill 441, 467
\item[] \qquad Dirichlet problem \hfill 471, 473, 474
\item[] \qquad gradient estimates \hfill 472
\item[] \qquad second derivative estimates \hfill 470, 471, 476
\item[] Morrey's lemma \hfill 298
\item[] Moser iteration technique \hfill 191, 195

\item[] Newtonian potential \hfill 18, 51ff.
\item[] \qquad differentiability properties \hfill 54f.
\item[] \qquad generalized \hfill 68
\item[] \qquad Hölder estimates \hfill 57, 64
\item[] nonlinear boundary value problems \hfill 481
\item[] non-uniformly elliptic equations (operators)
\item[] \qquad 111ff.
\item[] \quad see also linear elliptic equations (operators)
\item[] norm \hfill 73
\item[] \qquad boundary- \hfill 95
\item[] \qquad on Hölder spaces \hfill 53, 61, 66, 90, 96
\item[] \qquad on $L^p$ spaces \hfill 145
\item[] normal mapping \hfill 221
\item[] normed linear space \hfill 73

\item[] oblique derivative problem \hfill 120ff.
\item[] \quad for linear equations
\item[] \qquad Schauder estimates \hfill 125f.
\item[] \qquad solvability \hfill 128
\item[] \quad for nonlinear equations \hfill 481
\item[] \quad non-regular- \hfill 141
\item[] \quad for Poisson's equation \hfill 120ff.
\item[] \qquad Hölder estimates \hfill 123, 125
\item[] \qquad regular- \hfill 121
\item[] oblique derivative boundary condition \hfill 120, 126, 481
\item[] \quad regular- \hfill 47, 121, 126
\item[] orthogonality \hfill 81
\item[] orthogonal projection \hfill 82

\item[] parallelogram law \hfill 80
\item[] parametric functional \hfill 435
\item[] \quad elliptic- \hfill 435
\item[] partition of unity \hfill 136
\item[] perpendicular elements \hfill 81
\item[] Perron method \hfill 24, 102
\end{itemize}

% PDF page 525
\source{gilbarg-trudinger:page-0525}

Perron solution 25

Poincar\'e inequalities 164

Poisson's equation 51ff.

\quad boundary H\"older estimates 64ff.

\quad compactness of solutions 61

\quad Dirichlet problem 15, 56, 62, 67, 71

\quad gradient estimates 41

\quad interior H\"older estimates 60, 62

\quad oblique derivative problem 120ff.

Poisson integral 20

Poisson integral formula 20

Poisson kernel 20

prescribed Gauss curvature equation 442

prescribed mean curvature equation 261, 388f.

\quad boundary gradient estimate 340, 345

\quad gradient estimate 367

\quad maximum principle 276f.

Dirichlet problem

\quad non-solvability 352

\quad solvability 407, 440

\quad uniqueness 407

\emph{see also} quasilinear elliptic equations

principal coordinate system 354

principal curvature 354

principal directions 354

Pucci equation 442

\vspace{1em}

quasiconformal mapping 294ff.

\quad H\"older estimates 299

\quad on hypersurface 416

\quad H\"older estimates 422

quasilinear elliptic equations (operators) 1, 257ff.

\quad comparison principles 263, 268, 347

\quad in divergence form 5, 260

\qquad Dirichlet problem, solvability 283, 331, 385

\qquad global gradient estimates 323, 374, 380

\qquad interior gradient estimates 320, 379, 387

\quad regularity 379

\quad equivalent 259

\quad in general form

\qquad boundary gradient estimates 337, 383ff., 344ff., 358

\qquad continuity estimate 353

\qquad Dirichlet problem

\qquad\quad non-solvability 350, 352

\qquad\quad non-uniqueness 267

\qquad\quad solvability 281, 287, 331, 380ff., 385

\qquad\quad uniqueness 264

\qquad global gradient estimates 331, 367

\qquad interior gradient estimates 328, 372

\qquad regularity 281, 287

\quad maximum principles 262, 264, 271ff., 277ff.

\qquad for gradient 362

\quad in two variables

\qquad Dirichlet problem, solvability 283, 305, 312, 331

\qquad gradient estimates 311, 323f.

quasisolution 291

Rayleigh quotient 213

reflexive space 79

regular boundary point 25, 139, 206

\quad Wiener criterion 28, 206

regular functional 289

regular oblique derivative problem

\quad \emph{see} oblique derivative problem

regularity

\quad of classical solutions 109, 140

\quad of weak solutions 183ff.

\quad for quasilinear equations 281, 287, 379

\quad \emph{see also} gradient estimates, H\"older estimates,

\qquad Schauder estimates

regularization of a function 147

resolvent operator 78

Riesz representation theorem 82

\vspace{1em}

scalar product 80

Schauder estimates

\quad boundary 95ff., 100

\quad global 98, 142

\quad interior 90, 93, 112, 141

\quad for oblique derivative problem 126f.

Schauder fixed point theorem 279

Schwarz inequality 80, 146

Schwarz reflection principle 28

segment condition 155

Sobolev inequalities 155, 158, 164, 439

\quad best constant 158

Sobolev spaces 153

\quad density theorems 154

\quad embedding theorems 155, 158, 162, 167

\quad norm 153

\quad scalar product 154

sphere condition

\quad enclosing- 339

\quad exterior- 27

\quad interior- 33

Stokes' theorem 413

straightening the boundary 94

strict exterior plane condition 117

strictly elliptic equations (operators) 31, 87

\quad \emph{see also} linear elliptic equations (operators),

\qquad quasilinear elliptic equations (operators)

strong derivative 150

strong maximum principle

\quad \emph{see maximum principles}

strong solution 219

\quad Dirichlet problem 241

\quad Harnack inequality 250

\quad H\"older estimates 250

structure conditions 188, 194, 271, 305, 335ff.,

\quad 366ff.

\quad natural- 367, 375

structural inequalities

\quad \emph{see structure conditions}

subfunction 24, 102

subharmonic function 13, 23

\quad on hypersurface 393


% PDF page 526
\source{gilbarg-trudinger:page-0526}

\begin{quote}
Subject Index
\end{quote}

\noindent
\begin{tabular}{@{}p{0.48\textwidth}p{0.48\textwidth}@{}}
\textbf{weakly-} 29 & \emph{see also} linear elliptic equations (operators),\\
\textbf{subsolution} 33, 45, 102 & quasilinear elliptic equations (operators)\\
\textbf{weak-} 188 & \\
\textbf{superfunction} 24, 102 & \\
\textbf{superharmonic function} 13, 23 & variational problems 288ff.\\
\textbf{weakly-} 29 & \\
\textbf{supersolution} 33, 45, 102 & \\
\textbf{weak-} 188 & \\
 & \\
\textbf{tangential gradient} 389 & weak convergence 85\\
\textbf{test function} 178 & weak derivative 149\\
\textbf{three-point condition} 310, 314 & \hspace{1em}chain rule 151\\
\textbf{total variation} 175 & \hspace{1em}product rule 150\\
\textbf{triangle inequality} 80 & weak maximum principle\\
 & \hspace{1em}\emph{see maximum principles}\\
 & weak solution 3, 177\\
 & \hspace{1em}\emph{see also} linear elliptic equations\\
 & Wiener criterion 28, 206\\
 & Wirtinger's inequality 297\\
 & \\
\textbf{uniformly convex domain} 283, 339 & Young's inequality 145\\
\textbf{uniformly elliptic equations (operators)} 1, 31, 259 & \\
\end{tabular}

% PDF page 527
\source{gilbarg-trudinger:page-0527}
\begin{center}
\textbf{Notation Index}
\end{center}

\[
\begin{array}{@{}ll@{\qquad}ll@{}}
\textit{Spaces and their norms and seminorms} & &
\| \cdot \|_{W^{k,p}(\Omega)},\ \| \cdot \|_{k,p;\Omega} & 153\\
\mathrm{BV}(\Omega) & 175 &
\mathbb{R}^n \times \mathbb{R}^n & 441\\
C^k(\Omega),\ C^k(\overline{\Omega}),\ C_0^k(\Omega) & 10 &
\textit{Other notation (see also page 9)} & \\
C^{\alpha}(\Omega),\ C^{\alpha}(\overline{\Omega}),\ C^{k,\alpha}(\Omega),\ C^{k,\alpha}(\overline{\Omega}) & 52 &
A(\mathcal{G}_{R}) & 393\\
C^{k,\alpha}(\partial\Omega),\ C^{k,\alpha}(T)\ (T=\partial\Omega) & 95 &
a^{ij}_0,\ a^{ij}_x & 342\\
C_*^k(\Omega) & 86,\,133 &
a^{ij}_* & 359,\,364,\,430\\
H^k(\Omega),\ H_0^k(\Omega) & 154 &
b_0,\ b_\infty & 342\\
L^p(\Omega) & 145 &
b_* & 430\\
L^p_{\mathrm{loc}}(\Omega) & 147 &
c_i & 359,\,364\\
M^p(\Omega) & 164 &
\operatorname{cap}\Omega & 27\\
W^k(\Omega) & 150 &
\mathcal{D},\ \mathcal{D}^* & 220\\
W^{k,p}(\Omega) & 153 &
\mathcal{D}(r,z),\ \mathcal{D}(r) & 295\\
W^{k,p}_0(\Omega) & 154 &
\mathcal{D}(\rho,Z) & 418\\
W^{k,p}_{\mathrm{loc}}(\Omega) & 154 &
\Delta & 13,\,391\\
\| \cdot \|_{k,\Omega},\ \| \cdot \|_{k,\alpha;\Omega} & 53 &
\Delta^h,\ \Delta_i^h & 109\\
\| \cdot \|^*_{k,\Omega},\ \| \cdot \|^*_{k,\alpha;\Omega} & 53 &
d_x,\ d_{x,y} & 61\\
\| \cdot \|^*_{k,\Omega},\ \| \cdot \|^*_{k,\alpha;\Omega} & 61 &
\tilde d_x,\ \tilde d_{x,y} & 66\\
\| \cdot \|_{k,\Omega\cup T},\ \| \cdot \|_{k,\alpha;\Omega\cup T} & 66,\,96 &
d(x) & 354\\
\| \cdot \|^{(\sigma)}_{k,\Omega},\ \| \cdot \|^{(\sigma)}_{k,\alpha;\Omega} & 90 &
\delta & 361,\,389\\
\| \cdot \|^{(k)}_{0,\alpha;\Omega} & 61 &
\delta_k & 319\\
\| \cdot \|^{(k)}_{0,\alpha;\Omega\cup T} & 66 &
\bar{\delta} & 363\\
{}[\,\cdot\,]_{x_0} & 52 &
\delta_i\delta_i & 391\\
{}[\,\cdot\,]_{D} & 52 &
\partial_i & 369\\
{}[\,\cdot\,]_{k;\Omega},\ [\,\cdot\,]_{k,\alpha;\Omega} & 53 &
\mathcal{E} & 259\\
{}[\,\cdot\,]^*_{k;\Omega},\ [\,\cdot\,]^*_{k,\alpha;\Omega} & 61 &
\mathcal{E}^* & 347\\
{}[\,\cdot\,]^*_{k;\Omega\cup T},\ [\,\cdot\,]^*_{k,\alpha;\Omega\cup T} & 66,\,96 &
e_i & 109\\
{}[\,\cdot\,]^{(\sigma)}_{k;\Omega},\ [\,\cdot\,]^{(\sigma)}_{k,x;\Omega} & 90 &
\mathcal{F} & 334\\
\|\,\cdot\,\|_{k;\bar{\Omega}},\ \|\,\cdot\,\|_{k,\alpha;\bar{\Omega}} & 53 &
F_{ij} & 441\\
\|\,\cdot\,\|_{C^k(\overline{\Omega})},\ \|\,\cdot\,\|_{C^{k,\alpha}(\overline{\Omega})} & 53 &
G(x,y) & 19,\,121\\
\|\,\cdot\,\|_{C^k(\overline{\Omega})},\ \|\,\cdot\,\|_{C^{k,\alpha}(\overline{\Omega})} & 53 &
G & 423\\
\|\cdot\| & 73,\,74 &
\Gamma(x-y) & 17\\
\|\cdot\|_{\mathrm{BV}(\Omega)} & 175 &
\Gamma^+ & 220\\
\|\,\cdot\,\|_{L^p(\Omega)} & 145 &
H & 354,\,401
\end{array}
\]

% PDF page 528
\source{gilbarg-trudinger:page-0528}

\begin{quote}
Notation Index
\end{quote}

\noindent
\begin{tabular}{@{}p{0.48\textwidth}p{0.48\textwidth}@{}}
$J(\varphi)$ 414 & $\mathcal{G}_R,\ \mathcal{G}_R(y)$ 394\\
$\mathcal{K}^+,\ \mathcal{K}^-$ 342 & $\mathcal{T}$ 339\\
$k,\ k(R)$ 194 & $\overline{\mathcal{T}}$ 361\\
$\bar{k},\ \bar{k}(R)$ 200 & $u^+,\ u^-$ 33, 152\\
$\kappa_i$ 354 & $u_M^+,\ u_M^-$ 303\\
$\mathcal{L}_a$ 442 & $u_h$ 147\\
$\mathcal{L}(u, v)$ 177 & $u^{ij}$ 467\\
$\mathfrak{M}$ 339 & $V_\mu$ 159\\
$\chi(v),\ \chi_u(v)$ 221 & $(\partial\Omega,\ \varphi)$ 283\\
$v$ 13, 354 & \\
\end{tabular}

% PDF page 529
\source{gilbarg-trudinger:page-0529}
\begin{quote}
\textit{Grundlehren der mathematischen Wissenschaften}

\textit{A Series of Comprehensive Studies in Mathematics}

\textit{A Selection}

210. Gihman/Skorohod: The Theory of Stochastic Processes I
211. Comfort/Negrepontis: The Theory of Ultrafilters
212. Switzer: Algebraic Topology - Homotopy and Homology
215. Schaefer: Banach Lattices and Positive Operators
217. Stenstr\"om: Rings of Quotients
218. Gihman/Skorohod: The Theory of Stochastic Processes II
219. Duvaut/Lions: Inequalities in Mechanics and Physics
220. Kirillov: Elements of the Theory of Representations
221. Mumford: Algebraic Geometry I: Complex Projective Varieties
222. Lang: Introduction to Modular Forms
223. Bergh/L{\"o}fstr{\"o}m: Interpolation Spaces. An Introduction
224. Gilbarg/Trudinger: Elliptic Partial Differential Equations of Second Order
225. Sch{\"u}tte: Proof Theory
226. Karoubi: K-Theory. An Introduction
227. Grauert/Remmert: Theorie der Steinschen R{\"a}ume
228. Segal/Kunze: Integrals and Operators
229. Hasse: Number Theory
230. Klingenberg: Lectures on Closed Geodesics
231. Lang: Elliptic Curves. Diophantine Analysis
232. Gihman/Skorohod: The Theory of Stochastic Processes III
233. Stroock/Varadhan: Multidimensional Diffusion Processes
234. Aigner: Combinatorial Theory
235. Dynkin/Yushkevich: Controlled Markov Processes
236. Grauert/Remmert: Theory of Stein Spaces
237. K\"othe: Topological Vector Spaces II
238. Graham/McGehee: Essays in Commutative Harmonic Analysis
239. Elliott: Probabilistic Number Theory I
240. Elliott: Probabilistic Number Theory II
241. Rudin: Function Theory in the Unit Ball of C$^n$
242. Huppert/Blackburn: Finite Groups II
243. Huppert/Blackburn: Finite Groups III
244. Kubert/Lang: Modular Units
245. Cornfeld/Fomin/Sinai: Ergodic Theory
246. Naimark/Stern: Theory of Group Representations
247. Suzuki: Group Theory I
248. Suzuki: Group Theory II
249. Chung: Lectures from Markov Processes to Brownian Motion
250. Arnold: Geometrical Methods in the Theory of Ordinary Differential Equations
251. Chow/Hale: Methods of Bifurcation Theory
252. Aubin: Nonlinear Analysis on Manifolds. Monge-Amp\`ere Equations
253. Dwork: Lectures on $p$-adic Differential Equations
254. Freitag: Siegelsche Modulfunktionen
255. Lang: Complex Multiplication
256. H\"ormander: The Analysis of Linear Partial Differential Operators I
257. H\"ormander: The Analysis of Linear Partial Differential Operators II
258. Smoller: Shock Waves and Reaction-Diffusion Equations
259. Duren: Univalent Functions
260. Freidlin/Wentzell: Random Perturbations of Dynamical Systems
261. Bosch/G\"untzer/Remmert: Non Archimedean Analysis -- A System Approach
to Rigid Analytic Geometry
262. Doob: Classical Potential Theory and Its Probabilistic Counterpart
\end{quote}

% PDF page 530
\source{gilbarg-trudinger:page-0530}
263. Krasnosel'skii/Zabre\u{\i}ko: Geometrical Methods of Nonlinear Analysis
264. Aubin/Cellina: Differential Inclusions
265. Grauert/Remmert: Coherent Analytic Sheaves
266. de Rham: Differentiable Manifolds
267. Arbarello/Cornalba/Griffiths/Harris: Geometry of Algebraic Curves, Vol. I
268. Arbarello/Cornalba/Griffiths/Harris: Geometry of Algebraic Curves, Vol. II
269. Schapira: Microdifferential Systems in the Complex Domain
270. Scharlau: Quadratic and Hermitian Forms
271. Ellis: Entropy, Large Deviations, and Statistical Mechanics
272. Elliott: Arithmetic Functions and Integer Products
273. Nikol'skii: Treatise on the Shift Operator
274. H\"ormander: The Analysis of Linear Partial Differential Operators III
275. H\"ormander: The Analysis of Linear Partial Differential Operators IV
276. Liggett: Interacting Particle Systems
277. Fulton/Lang: Riemann-Roch Algebra
278. Barr/Wells: Toposes, Triples and Theories
279. Bishop/Bridges: Constructive Analysis
280. Neukirch: Class Field Theory
281. Chandrasekharan: Elliptic Functions
282. Lelong/Gruman: Entire Functions of Several Complex Variables
283. Kodaira: Complex Manifolds and Deformation of Complex Structures
284. Finn: Equilibrium Capillary Surfaces
285. Burago/Zalgaller: Geometric Inequalities
286. Andrianov: Quadratic Forms and Hecke Operators
287. Maskit: Kleinian Groups
288. Jacod/Shiryayev: Limit Theorems for Stochastic Processes
289. Manin: Gauge Field Theory and Complex Geometry
290. Conway/Sloane: Sphere Packings, Lattices and Groups
291. Hahn/O'Meara: The Classical Groups and K-Theory
292. Kashiwara/Schapira: Sheaves on Manifolds
293. Revuz/Yor: Continuous Martingales and Brownian Motion
294. Knus: Quadratic and Hermitian Forms over Rings
295. Dierkes/Hildebrandt/K{\"u}ster/Wohlrab: Minimal Surfaces I
296. Dierkes/Hildebrandt/K{\"u}ster/Wohlrab: Minimal Surfaces II
297. Pastur/Figotin: Spectra of Random and Almost-Periodic Operators
298. Berline/Getzler/Vergne: Heat Kernels and Dirac Operators
299. Pommerenke: Boundary Behaviour of Conformal Maps
300. Orlik/Terao: Arrangements of Hyperplanes
301. Loday: Cyclic Homology
302. Lange/Birkenhake: Complex Abelian Varieties
303. DeVore/Lorentz: Constructive Approximation
304. Lorentz/v. Golitschek/Makovoz: Constructive Approximation. Advanced Problems
305. Hiriart-Urruty/Lemar\'echal: Convex Analysis and Minimization Algorithms I.
Fundamentals
306. Hiriart-Urruty/Lemar\'echal: Convex Analysis and Minimization Algorithms II.
Advanced Theory and Bundle Methods
307. Schwarz: Quantum Field Theory and Topology
308. Schwarz: Topology for Physicists
309. Adem/Milgram: Cohomology of Finite Groups
310. Giaquinta/Hildebrandt: Calculus of Variations I: The Lagrangian Formalism
311. Giaquinta/Hildebrandt: Calculus of Variations II: The Hamiltonian Formalism
312. Chung/Zhao: From Brownian Motion to Schr\"odinger's Equation
313. Malliavin: Stochastic Analysis
314. Adams/Hedberg: Function Spaces and Potential Theory
315. B\"urgisser/Clausen/Shokrollahi: Algebraic Complexity Theory
316. Saff/Totik: Logarithmic Potentials with External Fields
317. Rockafellar/Wets: Variational Analysis
318. Kobayashi: Hyperbolic Complex Spaces

% PDF page 531
\source{gilbarg-trudinger:page-0531}

\begin{quote}
\textbf{M. Aigner} Combinatorial Theory \ \textsc{isbn} 978-3-540-61787-7
\textbf{A. L. Besse} Einstein Manifolds \ \textsc{isbn} 978-3-540-74120-6
\textbf{N. P. Bhatia, G. P. Szegö} Stability Theory of Dynamical Systems \ \textsc{isbn} 978-3-540-42748-3
\textbf{J. W. S. Cassels} An Introduction to the Geometry of Numbers \ \textsc{isbn} 978-3-540-61788-4
\textbf{R. Courant, F. John} Introduction to Calculus and Analysis I \ \textsc{isbn} 978-3-540-65058-4
\textbf{R. Courant, F. John} Introduction to Calculus and Analysis II/1 \ \textsc{isbn} 978-3-540-66569-4
\textbf{R. Courant, F. John} Introduction to Calculus and Analysis II/2 \ \textsc{isbn} 978-3-540-66570-0
\textbf{P. Dembowski} Finite Geometries \ \textsc{isbn} 978-3-540-61786-0
\textbf{A. Dold} Lectures on Algebraic Topology \ \textsc{isbn} 978-3-540-58660-9
\textbf{J. L. Doob} Classical Potential Theory and Its Probabilistic Counterpart \ \textsc{isbn} 978-3-540-41206-9
\textbf{R. S. Ellis} Entropy, Large Deviations, and Statistical Mechanics \ \textsc{isbn} 978-3-540-29059-9
\textbf{H. Federer} Geometric Measure Theory \ \textsc{isbn} 978-3-540-60656-7
\textbf{S. Flügge} Practical Quantum Mechanics \ \textsc{isbn} 978-3-540-65035-5
\textbf{L. D. Faddeev, L. A. Takhtajan} Hamiltonian Methods in the Theory of Solitons
\hfill \textsc{isbn} 978-3-540-69843-2
\textbf{I. I. Gikhman, A. V. Skorokhod} The Theory of Stochastic Processes I \ \textsc{isbn} 978-3-540-20284-4
\textbf{I. I. Gikhman, A. V. Skorokhod} The Theory of Stochastic Processes II \ \textsc{isbn} 978-3-540-20285-1
\textbf{I. I. Gikhman, A. V. Skorokhod} The Theory of Stochastic Processes III \ \textsc{isbn} 978-3-540-49940-4
\textbf{D. Gilbarg, N. S. Trudinger} Elliptic Partial Differential Equations of Second Order
\hfill \textsc{isbn} 978-3-540-41160-4
\textbf{H. Grauert, R. Remmert} Theory of Stein Spaces \ \textsc{isbn} 978-3-540-00373-1
\textbf{H. Hasse} Number Theory \ \textsc{isbn} 978-3-540-42749-0
\textbf{F. Hirzebruch} Topological Methods in Algebraic Geometry \ \textsc{isbn} 978-3-540-58663-0
\textbf{L. Hörmander} The Analysis of Linear Partial Differential Operators I -- Distribution Theory
and Fourier Analysis \ \textsc{isbn} 978-3-540-00662-6
\textbf{L. Hörmander} The Analysis of Linear Partial Differential Operators II -- Differential
Operators with Constant Coefficients \ \textsc{isbn} 978-3-540-22516-4
\textbf{L. Hörmander} The Analysis of Linear Partial Differential Operators III -- Pseudo-
Differential Operators \ \textsc{isbn} 978-3-540-49937-4
\textbf{L. Hörmander} The Analysis of Linear Partial Differential Operators IV -- Fourier
Integral Operators \ \textsc{isbn} 978-3-642-00117-8
\textbf{K. It\^o, H. P. McKean, Jr.} Diffusion Processes and Their Sample Paths \ \textsc{isbn} 978-3-540-60629-1
\textbf{T. Kato} Perturbation Theory for Linear Operators \ \textsc{isbn} 978-3-540-58661-6
\textbf{S. Kobayashi} Transformation Groups in Differential Geometry \ \textsc{isbn} 978-3-540-58659-3
\textbf{K. Kodaira} Complex Manifolds and Deformation of Complex Structures \ \textsc{isbn} 978-3-540-22614-7
\textbf{Th. M. Liggett} Interacting Particle Systems \ \textsc{isbn} 978-3-540-22617-8
\textbf{J. Lindenstrauss, L. Tzafriri} Classical Banach Spaces I and II \ \textsc{isbn} 978-3-540-60628-4
\textbf{R. C. Lyndon, P. E. Schupp} Combinatorial Group Theory \ \textsc{isbn} 978-3-540-41158-1
\textbf{S. Mac Lane} Homology \ \textsc{isbn} 978-3-540-58662-3
\textbf{C. B. Morrey Jr.} Multiple Integrals in the Calculus of Variations \ \textsc{isbn} 978-3-540-69915-6
\textbf{D. Mumford} Algebraic Geometry I -- Complex Projective Varieties \ \textsc{isbn} 978-3-540-58657-9
\textbf{O. T. O'Meara} Introduction to Quadratic Forms \ \textsc{isbn} 978-3-540-66564-9
\textbf{G. Pólya, G. Szegö} Problems and Theorems in Analysis I -- Series, Integral Calculus,
Theory of Functions \ \textsc{isbn} 978-3-540-63640-3
\textbf{G. Pólya, G. Szegö} Problems and Theorems in Analysis II -- Theory of Functions. Zeros.
Polynomials. Determinants. Number Theory. Geometry
\hfill \textsc{isbn} 978-3-540-63686-1
\textbf{W. Rudin} Function Theory in the Unit Ball of C$^{\mathrm{n}}$ \ \textsc{isbn} 978-3-540-68272-1
\textbf{S. Sakai} C$^*$-Algebras and W$^*$-Algebras \ \textsc{isbn} 978-3-540-63633-5
\textbf{C. L. Siegel, J. K. Moser} Lectures on Celestial Mechanics \ \textsc{isbn} 978-3-540-58656-2
\textbf{T. A. Springer} Jordan Algebras and Algebraic Groups \ \textsc{isbn} 978-3-540-63632-8
\textbf{D. W. Stroock, S. R. S. Varadhan} Multidimensional Diffusion Processes \ \textsc{isbn} 978-3-540-28998-2
\textbf{R. R. Switzer} Algebraic Topology: Homology and Homotopy \ \textsc{isbn} 978-3-540-42750-6
\textbf{A. Weil} Basic Number Theory \ \textsc{isbn} 978-3-540-58655-5
\textbf{A. Weil} Elliptic Functions According to Eisenstein and Kronecker \ \textsc{isbn} 978-3-540-65036-2
\textbf{K. Yosida} Functional Analysis \ \textsc{isbn} 978-3-540-58654-8
\textbf{O. Zariski} Algebraic Surfaces \ \textsc{isbn} 978-3-540-58658-6
\end{quote}
