% Source: Lee, "Introduction to Riemannian Manifolds" (2nd ed.), Chapter 4 "Connections", PDF pages 98-126 of the cited source.
% Transcribed from the cited source, with manual checks for a few pages, and organized according to leanblueprint
% conventions (semantic \label, bracketed titles, \uses dependency edges, \dcref to Lee's own numbering) below.

\section{The Problem of Differentiating Vector Fields}

For a smooth curve $\gamma:I\to\mathbb R^n$,
\begin{align}
\gamma'(t)&=\dot\gamma^i(t)\frac{\partial}{\partial x^i}\bigg|_{\gamma(t)},\label{eq:4.1}\\
\gamma''(t)&=\ddot\gamma^i(t)\frac{\partial}{\partial x^i}\bigg|_{\gamma(t)}.\label{eq:4.2}
\end{align}
Dots denote derivatives in $t$. A curve is a straight line exactly when it admits a parametrization with $\gamma''=0$.

For $Y=Y^i\partial_i$ on $\mathbb R^n$ and $v=v^i\partial_i|_p$, the Euclidean directional derivative is
\[
\bar\nabla_vY=v(Y^i)\partial_i|_p,
\qquad v(Y^i)=v^j\frac{\partial Y^i(p)}{\partial x^j}.
\]
For a vector field $X$, this gives
\begin{equation}
\bar\nabla_XY=X(Y^i)\frac{\partial}{\partial x^i}.\tag{4.3}
\end{equation}

If $M\subset\mathbb R^n$ is an embedded submanifold, the tangential acceleration of a curve is $\gamma''{}^\top=\pi^\top(\gamma'')$. For a local field $Y$ on $M$, extend it to $\widetilde Y$ near $M$ and define
\begin{equation}
\nabla_v^\top Y=\pi^\top(\bar\nabla_v\widetilde Y).\tag{4.4}
\end{equation}
The projection is onto $T_pM$; the value is independent of the extension by Problem 4-1.
\uses{prop:normal-bundle}

On an abstract manifold, velocity still has the coordinate-independent expression \eqref{eq:4.1}, but $\gamma'(t+h)$ and $\gamma'(t)$ lie in different tangent spaces, so their difference quotient is undefined. Coordinate expressions can therefore give incompatible accelerations (the unit circle has Cartesian acceleration $-\cos t\,\partial_x-\sin t\,\partial_y$ but polar expression $r''\partial_r+\theta''\partial_\theta=0$). A connection supplies a coordinate-independent rule for comparing and differentiating such vectors.

\section{Connections}

\begin{definition}[Connection]
\label{def:connection}
\lean{LeeLib.Ch04.Connection, LeeLib.Ch04.covariantDeriv}
\leanok
Let $\pi:E\to M$ be a smooth vector bundle over a smooth manifold with or without boundary. A connection is a map
\[
\nabla:\mathfrak X(M)\times\Gamma(E)\to\Gamma(E)
\]
such that
\begin{enumerate}
\item[(i)] $\nabla_{f_1X_1+f_2X_2}Y=f_1\nabla_{X_1}Y+f_2\nabla_{X_2}Y$;
\item[(ii)] $\nabla_X(a_1Y_1+a_2Y_2)=a_1\nabla_XY_1+a_2\nabla_XY_2$;
\item[(iii)] $\nabla_X(fY)=f\nabla_XY+(Xf)Y$.
\end{enumerate}
Here $f,f_i\in C^\infty(M)$, $a_i\in\mathbb R$, and the fields have the indicated domains.
\end{definition}

\begin{lemma}[Locality]
\label{lem:connection-is-local-operator}
\dcref{ch4:4.1}
\uses{def:connection}
\lean{LeeLib.Ch04.covariantDeriv_local}
\leanok
If $\nabla$ is a connection, then $\nabla_XY|_p$ depends only on the germs of $X,Y$ near $p$: if $X=\widetilde X$ and $Y=\widetilde Y$ near $p$, then $\nabla_XY|_p=\nabla_{\widetilde X}\widetilde Y|_p$.
\end{lemma}
\begin{proof}
\leanok
\sketch If $Y$ vanishes near $p$, choose a bump function $\varphi$ supported there with $\varphi(p)=1$. Since $\varphi Y=0$, the product rule gives
\begin{equation}
0=\nabla_X(\varphi Y)=(X\varphi)Y+\varphi\nabla_XY.\tag{4.5}
\end{equation}
Both terms except $\nabla_XY|_p$ vanish at $p$. The argument for changing $X$ is analogous, using $C^\infty$-linearity in the first slot.
\end{proof}

\begin{exercise}\label{exer:lemma-4-1-comparison}\dcref{ch4:4.2}
\uses{lem:connection-is-local-operator}
Complete the proof of the preceding lemma for the first argument.
\end{exercise}

\begin{proposition}[Restriction of a Connection]\label{prop:connection-restriction}\dcref{ch4:4.3}
\uses{lem:connection-is-local-operator}
If $\nabla$ is a connection on $E\to M$ and $U\subseteq M$ is open, there is a unique connection $\nabla^U$ on $E|_U$ satisfying
\begin{equation}
\nabla^U_{X|_U}(Y|_U)=(\nabla_XY)|_U.\tag{4.6}
\end{equation}
\end{proposition}
\begin{proof}
\sketch Extend local fields and sections near each $p\in U$ to global ones with bump functions and define
\begin{equation}
\nabla^U_XY|_p=(\nabla_{\widetilde X}\widetilde Y)|_p.\tag{4.7}
\end{equation}
Lemma \ref{lem:connection-is-local-operator} makes this independent of extensions and smooth; the connection axioms follow locally.
\end{proof}

\begin{exercise}\label{exer:connection-restriction-is-connection}\dcref{ch4:4.4}
\uses{prop:connection-restriction}
Complete the proof that $\nabla^U$ is a connection.
\end{exercise}

\begin{proposition}\label{prop:covariant-deriv-dependence-at-point}\dcref{ch4:4.5}
\uses{lem:connection-is-local-operator}
\lean{LeeLib.Ch04.covariantDeriv_dependence_at_point}
\leanok
For a connection, $\nabla_XY|_p$ depends only on the germ of $Y$ near $p$ and the single value $X_p$.
\end{proposition}
\begin{proof}
\leanok
\sketch Locality handles $Y$. If $X_p=0$, restrict to coordinates and write $X=X^i\partial_i$ with $X^i(p)=0$; restriction to the coordinate neighborhood and $C^\infty$-linearity give $\nabla_XY|_p=X^i(p)\nabla_{\partial_i}Y|_p=0$.
\end{proof}

Thus for $v\in T_pM$ and a local section $Y$, define $\nabla_vY$ by extending $v$ to a local field $X$ and setting $\nabla_vY=\nabla_XY|_p$; the preceding proposition makes this well defined.

\section{Connections in the Tangent Bundle}

A connection in $TM$ is a connection on $M$ and is a map $\mathfrak X(M)^2\to\mathfrak X(M)$ satisfying (i)--(iii). It is not a $(1,2)$-tensor because the second slot obeys the product rule rather than $C^\infty(M)$-linearity.

For a local frame $(E_i)$, define connection coefficients by
\begin{equation}
\nabla_{E_i}E_j=\Gamma^k_{ij}E_k.\tag{4.8}
\end{equation}
\begin{proposition}
\label{prop:connection-coefficients-frame}
\dcref{ch4:4.6}
\uses{def:connection}
\lean{LeeLib.Ch04.connectionCoeff, LeeLib.Ch04.covariantDeriv_frame_eq_sum_connectionCoeff, LeeLib.Ch04.covariantDeriv_eq_connectionCoeff_formula}
\leanok
If $X=X^iE_i$ and $Y=Y^jE_j$ on a local frame domain, then
\begin{equation}
\nabla_XY=\left(X(Y^k)+X^iY^j\Gamma^k_{ij}\right)E_k.\tag{4.9}
\end{equation}
\end{proposition}
\begin{proof}
\sketch Expand $\nabla_X(Y^jE_j)$ using the product rule and $C^\infty$-linearity in $X$, substitute (4.8), and rename the dummy index in the derivative term.
\end{proof}

\begin{proposition}[Transformation Law for Connection Coefficients]
\label{prop:transformation-law-connection-coefficients}
\dcref{ch4:4.7}
\uses{def:connection,prop:connection-coefficients-frame}
\lean{LeeLib.Ch04.connectionCoeff_transformation_law}
\leanok
If $\widetilde E_i=A^j_iE_j$ are two local frames, their coefficients satisfy
\[
\widetilde\Gamma^k_{ij}=(A^{-1})^k_pA^q_iA^r_j\Gamma^p_{qr}+(A^{-1})^k_pA^q_iE_q(A^p_j).
\]
\end{proposition}
\begin{proof}
\leanok
\sketch Expand $\nabla_{\widetilde E_i}(A^r_jE_r)$ using the product rule from Definition \ref{def:connection}; substitute (4.8) and Proposition \ref{prop:connection-coefficients-frame}, then write $E_p=(A^{-1})^k_p\widetilde E_k$ and $\widetilde E_i(A^p_j)=A^q_iE_q(A^p_j)$. Reusing Proposition \ref{prop:connection-coefficients-frame} gives the stated formula; the derivative term shows that coefficients are not tensor components.
\end{proof}

\section{Existence of Connections}

\begin{example}[The Euclidean Connection]
\label{ex:euclidean-connection}
\dcref{ch4:4.8}
On $T\mathbb R^n$, the formula (4.3) defines the Euclidean connection $\bar\nabla$; its standard-frame coefficients are zero.
\end{example}

\begin{example}[The Tangential Connection on a Submanifold of $\mathbb{R}^n$]
\label{ex:tangential-connection-submanifold}
\dcref{ch4:4.9}
\uses{prop:normal-bundle,ex:euclidean-connection,exer:extension-smooth-function,prob:embedded-submanifold-covariant-derivative,prop:adapted-orthonormal-frames}
For an embedded $M\subseteq\mathbb R^n$, define
\[
\nabla_X^TY=\pi^T\bigl(\widetilde\nabla_{\widetilde X}\widetilde Y|_M\bigr),
\]
where extensions are taken near $M$, $\widetilde\nabla$ is the Euclidean connection, and $\pi^T$ is the tangential projection. This is independent of extensions by (4.4) and is a connection.

\sketch Linearity in $X,Y$ is immediate. For an extension $\widetilde f$ of $f$,
\[
\nabla_X^T(fY)=\pi^T\bigl(\widetilde\nabla_{\widetilde X}(\widetilde f\widetilde Y)|_M\bigr)=(Xf)Y+f\nabla_X^TY.
\]
\end{example}

\begin{lemma}
\label{lem:connections-from-connection-coefficients}
\dcref{ch4:4.10}
\uses{prop:connection-coefficients-frame}
\lean{LeeLib.Ch04.connOfCoeff, LeeLib.Ch04.connectionCoeff_connOfCoeff, LeeLib.Ch04.covariantDeriv_eq_connectionCoeff_formula}
If a smooth $n$-manifold with or without boundary has a global frame $(E_i)$, then (4.9) gives a one-to-one correspondence between connections on $TM$ and $n^3$ smooth functions $\Gamma^k_{ij}$.
\end{lemma}
\begin{proof}
\sketch A connection supplies $\Gamma^k_{ij}$ through (4.8). Conversely, (4.9) defines a smooth operation satisfying linearity; the product rule is the direct expansion of $X(Y^j)$.
\end{proof}

\begin{exercise}
\label{exer:complete-proof-lemma-4.10}
\dcref{ch4:4.11}
\uses{lem:connections-from-connection-coefficients}
\lean{LeeLib.Ch04.connOfCoeff_isCovariantDerivativeOn, LeeLib.Ch04.connectionCoeff_connOfCoeff}
\leanok
Complete the proof of Lemma 4.10.
\end{exercise}

\begin{proposition}
\label{prop:tangent-bundle-admits-connection}
\dcref{ch4:4.12}
\uses{lem:connections-from-connection-coefficients}
Every smooth manifold with or without boundary admits a connection on its tangent bundle.
\end{proposition}
\begin{proof}
\sketch Choose local connections $\nabla^\alpha$ on a coordinate cover and a subordinate partition of unity $(\varphi_\alpha)$. Set
\begin{equation}
\nabla_XY=\sum_\alpha\varphi_\alpha\nabla_X^\alpha Y.\tag{4.11}
\end{equation}
Local finiteness gives smoothness. The partition identity $\sum_\alpha\varphi_\alpha=1$ yields
\[
\nabla_X(fY)=\sum_\alpha\varphi_\alpha((Xf)Y+f\nabla_X^\alpha Y)=(Xf)Y+f\nabla_XY.
\]
\end{proof}

\begin{proposition}[The Difference Tensor]
\label{prop:difference-tensor}
\dcref{ch4:4.13}
\uses{def:connection}
\lean{LeeLib.Ch04.differenceTensor, LeeLib.Ch04.differenceTensor_apply}
\leanok
For connections $\nabla^0,\nabla^1$ on $TM$, define $D(X,Y)=\nabla_X^1Y-\nabla_X^0Y$. Then $D$ is $C^\infty(M)$-bilinear and hence a $(1,2)$-tensor field.
\end{proposition}
\begin{proof}
\leanok
\sketch First-slot linearity is inherited. Expanding $D(X,fY)$ with the two product rules cancels the two $(Xf)Y$ terms, leaving $D(X,fY)=fD(X,Y)$.
\end{proof}

\begin{theorem}
\label{thm:connections-affine-space}
\dcref{ch4:4.14}
\uses{prop:difference-tensor}
\lean{LeeLib.Ch04.addOneForm, LeeLib.Ch04.eq_addOneForm_differenceTensor}
\leanok
For any connection $\nabla^0$ on a smooth manifold, the set of all connections is the affine space
\[
\mathcal A(TM)=\{\nabla^0+D:D\in\Gamma(T^{(1,2)}TM)\},
\]
where $(\nabla^0+D)_XY=\nabla^0_XY+D(X,Y)$.
\end{theorem}
\begin{proof}
\leanok
\uses{prop:difference-tensor}
\sketch Proposition \ref{prop:difference-tensor} writes any $\nabla$ as $\nabla^0+D$. Conversely, adding a tensor $D$ preserves the two linearities, and $C^\infty$-linearity of $D$ in $Y$ leaves the product rule of $\nabla^0$ unchanged.
\end{proof}
\section{Covariant Derivatives of Tensor Fields}

\begin{proposition}
\label{prop:connection-on-tensor-bundles}
\dcref{ch4:4.15}
\lean{LeeLib.Ch04.covDerivCovector, LeeLib.Ch04.covDerivCovector_apply, LeeLib.Ch04.covDeriv_pairing_leibniz}
Let $M$ be a smooth manifold with or without boundary and let $\nabla$ be a
connection in $TM$. There is a unique induced connection on every tensor
bundle $T^{(k,l)}TM$, denoted again by $\nabla$, satisfying:
\begin{enumerate}
\item[(i)] on $TM$, it is the given connection;
\item[(ii)] on $T^{(0,0)}TM=M\times\mathbb R$, $\nabla_Xf=Xf$;
\item[(iii)] $\nabla_X(F\otimes G)=(\nabla_XF)\otimes G+F\otimes(\nabla_XG)$;
\item[(iv)] $\nabla_X(\operatorname{tr}F)=\operatorname{tr}(\nabla_XF)$ for every
contraction of one covariant and one contravariant index.
\end{enumerate}
It also satisfies
\begin{enumerate}
\item[(a)]
\[
\nabla_X\langle\omega,Y\rangle
=\langle\nabla_X\omega,Y\rangle+\langle\omega,\nabla_XY\rangle;
\]
\item[(b)] for $F\in\Gamma(T^{(k,l)}TM)$, smooth $1$-forms
$\omega^1,\ldots,\omega^k$, and smooth vector fields $Y_1,\ldots,Y_l$,
\begin{align}
(\nabla_XF)(\omega^1,\ldots,\omega^k,Y_1,\ldots,Y_l)
&=X\big(F(\omega^1,\ldots,\omega^k,Y_1,\ldots,Y_l)\big)\nonumber\\
&\quad-\sum_{i=1}^kF(\omega^1,\ldots,\nabla_X\omega^i,\ldots,\omega^k,Y_1,\ldots,Y_l)\nonumber\\
&\quad-\sum_{j=1}^lF(\omega^1,\ldots,\omega^k,Y_1,\ldots,\nabla_XY_j,\ldots,Y_l).
\label{eq:4.12}\tag{4.12}
\end{align}
\end{enumerate}
\end{proposition}
\begin{proof}
\sketch Pairing is the contraction of $\omega\otimes Y$, so (i)--(iv) give
(a). Repeatedly expressing the evaluation of $F$ as contractions of
$F\otimes\omega^1\otimes\cdots\otimes\omega^k\otimes Y_1\otimes\cdots\otimes
Y_l$ gives (b). Formula (a) with functions determines the derivative of every
1-form:
\begin{equation}
(\nabla_X\omega)(Y)=X(\omega(Y))-\omega(\nabla_XY).
\label{eq:4.13}\tag{4.13}
\end{equation}
Then (4.12) determines all higher tensor derivatives, proving uniqueness.
Conversely, define the derivative of $1$-forms by (4.13) and all other
derivatives by (4.12). The terms obtained by differentiating a coefficient
$f$ cancel when an argument is replaced by $f\omega^i$ or $fY_j$, so the
result is tensorial and smooth. Linearity and the tensor-product rule follow
from the corresponding rules for differentiation of functions.
\end{proof}

\begin{proposition}
\label{prop:covariant-derivative-components}
\dcref{ch4:4.16}
Let $M$ be a smooth manifold with or without boundary, let $\nabla$ be a
connection in $TM$, and let $(E_i)$ be a local frame with dual coframe
$(\epsilon^j)$ and coefficients $\nabla_{E_i}E_j=\Gamma^k_{ij}E_k$.
If $X=X^jE_j$, then for
$\omega=\omega_i\epsilon^i$,
\[
\nabla_X\omega=\bigl(X(\omega_k)-X^j\omega_i\Gamma^i_{jk}\bigr)\epsilon^k.
\]
For a tensor
\[
F=F^{i_1\cdots i_k}_{j_1\cdots j_l}
E_{i_1}\otimes\cdots\otimes E_{i_k}\otimes
\epsilon^{j_1}\otimes\cdots\otimes\epsilon^{j_l},
\]
\[
\begin{aligned}
\nabla_XF={}&\Bigl(X(F^{i_1\cdots i_k}_{j_1\cdots j_l})
+\sum_{s=1}^kX^mF^{i_1\cdots p\cdots i_k}_{j_1\cdots j_l}\Gamma^{i_s}_{mp}\\
&\qquad-\sum_{s=1}^lX^mF^{i_1\cdots i_k}_{j_1\cdots p\cdots j_l}\Gamma^p_{ms}\Bigr)
E_{i_1}\otimes\cdots\otimes E_{i_k}\otimes
\epsilon^{j_1}\otimes\cdots\otimes\epsilon^{j_l}.
\end{aligned}
\]
\end{proposition}

\begin{proposition}[The Total Covariant Derivative]
\label{prop:total-covariant-derivative}
\dcref{ch4:4.17}
Let $M$ be a smooth manifold with or without boundary and $\nabla$ a
connection. For $F\in\Gamma(T^{(k,l)}TM)$ define
\[
\nabla F\colon\underbrace{\Omega^1(M)\times\cdots\times\Omega^1(M)}_{k}
\times\underbrace{\mathfrak{X}(M)\times\cdots\times\mathfrak{X}(M)}_{l+1}
\longrightarrow C^\infty(M)
\]
by
\begin{equation}
(\nabla F)(\omega^1,\ldots,\omega^k,Y_1,\ldots,Y_l,X)
=(\nabla_XF)(\omega^1,\ldots,\omega^k,Y_1,\ldots,Y_l).
\tag{4.14}
\end{equation}
This is a smooth $(k,l+1)$-tensor field, called the total covariant derivative.
\end{proposition}
\begin{proof}
\sketch The tensor characterization lemma applies because $\nabla_XF$ is
tensorial in its $k+l$ arguments and a connection is $C^\infty(M)$-linear in
$X$.
\end{proof}

Writing the differentiation index last, if $Y=Y^iE_i$ and
$\omega=\omega_i\epsilon^i$, then
\[
\nabla Y=Y^i_{;j}E_i\otimes\epsilon^j,
\qquad Y^i_{;j}=E_jY^i+Y^k\Gamma^i_{jk},
\]
and
\[
\nabla\omega=\omega_{i;j}\epsilon^i\otimes\epsilon^j,
\qquad \omega_{i;j}=E_j\omega_i-\omega_k\Gamma^k_{ji}.
\]

\begin{proposition}
\label{prop:total-covariant-derivative-components}
\dcref{ch4:4.18}
Let $M$ be a smooth manifold with or without boundary, let $\nabla$ be a
connection in $TM$, and let $(E_i)$ be a smooth local frame with connection
coefficients $\Gamma^k_{ij}$. For a $(k,l)$-tensor $F$,
\[
\begin{aligned}
F^{i_1\cdots i_k}_{j_1\cdots j_l;m}
={}&E_m\bigl(F^{i_1\cdots i_k}_{j_1\cdots j_l}\bigr)
+\sum_{s=1}^kF^{i_1\cdots p\cdots i_k}_{j_1\cdots j_l}\Gamma^{i_s}_{mp}\\
&-\sum_{s=1}^lF^{i_1\cdots i_k}_{j_1\cdots p\cdots j_l}\Gamma^p_{mj_s}.
\end{aligned}
\]
\end{proposition}

\begin{exercise}
\label{exer:prove-proposition-4-18}
\dcref{ch4:4.19}
\uses{prop:total-covariant-derivative-components}
Prove Proposition \ref{prop:total-covariant-derivative-components}.
\end{exercise}

\begin{exercise}
\label{exer:tensor-product-covariant-derivative}
\dcref{ch4:4.20}
For smooth $(k,l)$- and $(r,s)$-tensor fields $F,G$,
\[
\begin{aligned}
(\nabla(F\otimes G))^{j_1\cdots j_kp_1\cdots p_r}_{i_1\cdots i_lq_1\cdots q_s;m}
={}&F^{j_1\cdots j_k}_{i_1\cdots i_l;m}G^{p_1\cdots p_r}_{q_1\cdots q_s}\\
&+F^{j_1\cdots j_k}_{i_1\cdots i_l}G^{p_1\cdots p_r}_{q_1\cdots q_s;m}.
\end{aligned}
\]
Equivalently, the semicolon derivative of the component product obeys the
ordinary product rule. This does not assert
$\nabla(F\otimes G)=(\nabla F)\otimes G+F\otimes(\nabla G)$, because the
differentiation index in the first term would not be in the final lower slot.
\end{exercise}

\section{Second Covariant Derivatives}

For a smooth manifold $M$ with or without boundary and a connection $\nabla$,
set $\nabla^2F=\nabla(\nabla F)$ for $F\in\Gamma(T^{(k,l)}TM)$ and define
\[
\nabla^2_{X,Y}F(\ldots)=\nabla^2F(\ldots,Y,X).
\]
The reversal of $X,Y$ records that the differentiation index is appended last;
$\nabla^2_{X,Y}$ differentiates first in the $Y$ direction and then in the
$X$ direction. It is $C^\infty(M)$-linear in $Y$, unlike
$\nabla_X(\nabla_YF)$.

\begin{proposition}
\label{prop:second-covariant-derivative-formula}
\dcref{ch4:4.21}
\uses{prop:connection-on-tensor-bundles}
\lean{LeeLib.Ch04.secondCovariantDeriv_function}
For every smooth vector or tensor field $F$,
\[
\nabla^2_{X,Y}F=\nabla_X(\nabla_YF)-\nabla_{\nabla_XY}F.
\]
\end{proposition}
\begin{proof}
\sketch Contract $\nabla F\otimes Y$ in its final two slots to obtain
$\nabla_YF$, and contract $\nabla^2F\otimes X$ and then $Y$ to obtain
$\nabla^2_{X,Y}F$. Applying $\nabla_X$, using compatibility with contraction
and the tensor-product rule, gives
\[
\nabla_X(\nabla_YF)=\nabla^2_{X,Y}F+\nabla_{\nabla_XY}F.
\]
Rearrange to obtain the formula.
\end{proof}

\begin{example}[The Covariant Hessian]
\label{ex:covariant-hessian}
\dcref{ch4:4.22}
\uses{prop:second-covariant-derivative-formula}
\lean{LeeLib.Ch04.covariantHessianTensor, LeeLib.Ch04.covariantHessianTensor_apply, LeeLib.Ch04.covariantHessian_apply}
\leanok
For $u\in C^\infty(M)$, $\nabla u=du$ as a $1$-form, since
$\nabla_Xu=Xu=du(X)$. The tensor $\nabla^2u=\nabla(du)$ is the covariant
Hessian, and
\[
\nabla^2u(Y,X)=\nabla^2_{X,Y}u
=\nabla_X(\nabla_Yu)-\nabla_{\nabla_XY}u
=X(Yu)-(\nabla_XY)u.
\]
In local coordinates,
\[
\nabla^2u=u_{ij}\,dx^i\otimes dx^j,
\qquad u_{ij}=\partial_j\partial_i u-\Gamma^k_{ji}\partial_ku.
\]
\end{example}

\section{Vector and Tensor Fields Along Curves}

Let $M$ be a smooth manifold with or without boundary and let
$\gamma\colon I\to M$ be smooth. A vector field along $\gamma$ is a
continuous map $V\colon I\to TM$ with $V(t)\in T_{\gamma(t)}M$; it is smooth
when smooth as a map to $TM$. Write $\mathfrak{X}(\gamma)$ for the smooth
fields. Pointwise operations make it a real vector space and a
$C^\infty(I)$-module:
\[
(fX)(t)=f(t)X(t).
\]
The velocity $\gamma'$ is a smooth field along $\gamma$. In $\mathbb R^2$, if
$R$ is rotation by $\pi/2$, then $N=R\gamma'$ has coordinates
$N(t)=(-\dot\gamma^2(t),\dot\gamma^1(t))$.

If $\overline V$ is a smooth vector field on a neighborhood of $\gamma(I)$,
then $V(t)=\overline V_{\gamma(t)}$ is smooth and is called \emph{extendible}.
If $\gamma(t_1)=\gamma(t_2)$ but $\gamma'(t_1)\ne\gamma'(t_2)$, then $\gamma'$
is not extendible. A tensor field along $\gamma$ is defined analogously in
$T^{(k,l)}TM$, and extendibility means restriction of a smooth tensor field
defined near $\gamma(I)$.

\begin{example}
\label{ex:figure-eight-curve}
\dcref{ch4:4.23}
The curve
\[
\gamma\colon(-\pi,\pi)\to\mathbb R^2,
\qquad \gamma(t)=(\sin 2t,\sin t),
\]
is an injective smooth immersion, but its velocity field is not extendible.
\end{example}

\section{Covariant Derivatives Along Curves}

\begin{theorem}[Covariant Derivative Along a Curve]
\label{thm:covariant-derivative-along-curve}
\dcref{ch4:4.24}
\uses{ex:figure-eight-curve}
\lean{LeeLib.Ch04.covariantDerivAlongChart, LeeLib.Ch04.covariantDerivAlongChart_add, LeeLib.Ch04.covariantDerivAlongChart_smul}
For a connection $\nabla$ on a smooth manifold $M$ with or without boundary and
a smooth curve $\gamma\colon I\to M$, there is a unique operator
\[
D_t\colon\mathfrak{X}(\gamma)\to\mathfrak{X}(\gamma)
\]
such that
\begin{enumerate}
\item[(i)] $D_t(aV+bW)=aD_tV+bD_tW$ for $a,b\in\mathbb R$;
\item[(ii)] $D_t(fV)=f'V+fD_tV$ for $f\in C^\infty(I)$;
\item[(iii)] if $V$ is the restriction of a smooth field $\widetilde V$ near
$\gamma(I)$, then
\[
D_tV(t)=\nabla_{\gamma'(t)}\widetilde V.
\]
\end{enumerate}
The same construction applies to smooth tensor fields of every type along
$\gamma$.
\end{theorem}
\begin{proof}
\sketch At $t_0\in I$, locality reduces the construction to a coordinate
neighborhood of $\gamma(t_0)$. Writing
$V(t)=V^i(t)\partial_i|_{\gamma(t)}$ gives the forced formula
\begin{equation}
D_tV(t)=\bigl(\dot V^k(t)+\dot\gamma^i(t)V^j(t)\Gamma^k_{ij}(\gamma(t))\bigr)
\partial_k|_{\gamma(t)}.\tag{4.15}
\end{equation}
This proves uniqueness. Define $D_t$ by (4.15) in a chart containing the curve;
the coordinate formula satisfies (i)--(iii), and uniqueness makes definitions
agree on overlaps. At an endpoint, extend the coordinate curve to a slightly
larger interval before restricting.
\end{proof}

\begin{exercise}
\label{exer:covariant-derivative-properties}
\dcref{ch4:4.25}
\uses{thm:covariant-derivative-along-curve}
Verify directly that (4.15) satisfies properties (i)--(iii) of Theorem
\ref{thm:covariant-derivative-along-curve}.
\end{exercise}

\begin{proposition}
\label{prop:covariant-derivative-depends-on-curve}
\dcref{ch4:4.26}
\uses{prop:covariant-deriv-dependence-at-point,thm:covariant-derivative-along-curve}
Let $M$ be a smooth manifold with or without boundary, let $p\in M$,
$v\in T_pM$, and let $Y,\overline Y$ be smooth vector fields
that agree along a smooth curve $\gamma$ with $\gamma(t_0)=p$ and
$\gamma'(t_0)=v$. Then
\[
\nabla_vY=\nabla_v\overline Y.
\]
\end{proposition}
\begin{proof}
\sketch The common restriction $Z(t)=Y_{\gamma(t)}=\overline Y_{\gamma(t)}$
is a field along $\gamma$. Both $Y$ and $\overline Y$ extend $Z$, so property
(iii) of Theorem \ref{thm:covariant-derivative-along-curve} identifies both
derivatives with $D_tZ(t_0)$.
\end{proof}
\section{Geodesics}

Let $M$ be a smooth manifold, possibly with boundary, with a connection $\nabla$
in $TM$. The acceleration of a smooth curve $\gamma\colon I\to M$ is
$D_t\gamma'$. A curve is a geodesic when this field vanishes. In coordinates
$\gamma(t)=(x^1(t),\ldots,x^n(t))$, this is equivalent to
\begin{equation}
\ddot x^k(t)+\dot x^i(t)\dot x^j(t)\Gamma^k_{ij}(x(t))=0. \tag{4.16}
\end{equation}

\begin{definition}[Geodesic]
\label{def:geodesic}
\uses{thm:covariant-derivative-along-curve}
\lean{LeeLib.Ch04.chartAcceleration, LeeLib.Ch04.IsGeodesicInChart, LeeLib.Ch04.isGeodesicInChart_iff}
\leanok
The \emph{acceleration} of $\gamma$ is $D_t\gamma'$. The curve is a
\emph{geodesic} when $D_t\gamma'\equiv0$, equivalently when its velocity is
parallel. Its coordinate functions satisfy (4.16).
\end{definition}

\begin{theorem}[Existence and Uniqueness of Geodesics]
\label{thm:geodesic-existence-uniqueness}
\dcref{ch4:4.27}
\lean{LeeLib.Ch04.IsSprayGeodesicAt, LeeLib.Ch04.exists_sprayGeodesic_initial, LeeLib.Ch04.sprayGeodesic_eventuallyEq, LeeLib.Ch04.isSprayGeodesicAt_eventuallyEq_of_lift_eq, LeeLib.Ch04.IsSprayGeodesicOnWithInitial.chartAcceleration_zero, LeeLib.Ch04.IsSprayGeodesicOnWithInitial.hasDerivAt_extChartAt_zero}
Let $M$ be a smooth manifold without boundary and $\nabla$ a connection in
$TM$. For every $p\in M$, $w\in T_pM$, and $t_0\in\mathbb R$, there are an open
interval $I\ni t_0$ and a geodesic $\gamma\colon I\to M$ such that
\[
\gamma(t_0)=p,
\qquad \gamma'(t_0)=w.
\]
Any two geodesics with these initial data agree on their common domain.
\end{theorem}

\begin{proof}
\sketch In coordinates near $p$, introduce $v^i=\dot x^i$ and rewrite (4.16) as
\begin{align}
\dot x^k(t)&=v^k(t),\\
\dot v^k(t)&=-v^i(t)v^j(t)\Gamma^k_{ij}(x(t)). \tag{4.17}
\end{align}
This is the integral-curve equation for
\begin{equation}
G_{(x,v)}=v^k\frac{\partial}{\partial x^k}\bigg|_{(x,v)}
-v^iv^j\Gamma^k_{ij}(x)\frac{\partial}{\partial v^k}\bigg|_{(x,v)}. \tag{4.18}
\end{equation}
Local flow existence and uniqueness for $G$ give the required geodesic and local
uniqueness. For two solutions, the set of times at which their positions and
velocities agree is nonempty and locally open by the same coordinate argument;
continuity makes it closed in the common interval. Connectedness of the interval
gives equality on the full common domain.
\end{proof}

A geodesic is \emph{maximal} if it has no extension to a strictly larger interval.
A \emph{geodesic segment} has compact interval domain.

\begin{corollary}[Existence of maximal geodesics]
\label{cor:existence-of-maximal-geodesic}
\dcref{ch4:4.28}
\uses{thm:geodesic-existence-uniqueness}
Let $M$ be a smooth manifold and $\nabla$ a connection in $TM$. For every
$p\in M$ and $v\in T_pM$, there is a unique maximal geodesic
$\gamma\colon I\to M$ on an open interval $I\ni0$ satisfying
$\gamma(0)=p$ and $\gamma'(0)=v$.
\lean{LeeLib.Ch04.maximalSprayGeodesicInterval, LeeLib.Ch04.maximalSprayGeodesic, LeeLib.Ch04.maximalSprayGeodesicInterval_isOpen, LeeLib.Ch04.maximalSprayGeodesic_structure_of_footInSource}
\end{corollary}

\begin{proof}
\sketch Take the union of all open intervals containing $0$ on which a geodesic
with the prescribed data exists. Theorem \ref{thm:geodesic-existence-uniqueness}
makes the local geodesics agree on overlaps, so they glue to the unique maximal
one.
\end{proof}

\begin{exercise}[Geodesics in Euclidean space]
\label{exer:euclidean-geodesics}
\dcref{ch4:4.29}
\uses{prop:connection-restriction}
Show that the maximal geodesics of the Euclidean connection on $\mathbb R^n$ are
exactly constant curves and constant-speed parametrizations of straight lines.
\end{exercise}

Write $\gamma_v$ for the maximal geodesic with initial velocity $v$ and initial
point $\pi(v)$, where $\pi\colon TM\to M$ is the bundle projection.

\section{Parallel Transport}

A smooth vector or tensor field $V$ along $\gamma$ is \emph{parallel} when
$D_tV=0$. Thus $\gamma$ is geodesic exactly when $\gamma'$ is parallel.

\begin{exercise}
\label{exer:parallel-vector-field-euclidean-connection}
\dcref{ch4:4.30}
Let $\gamma\colon I\to\mathbb R^n$ be smooth. Show that a smooth vector field
along $\gamma$ is parallel for the Euclidean connection if and only if its
components in the standard basis are constant.
\end{exercise}

In a coordinate representation $\gamma(t)=(\gamma^1(t),\ldots,\gamma^n(t))$,
a vector field is parallel precisely when
\begin{equation}
\dot V^k(t)=-V^j(t)\dot\gamma^i(t)\Gamma^k_{ij}(\gamma(t)),
\qquad k=1,\ldots,n. \tag{4.19}
\end{equation}
Tensor fields satisfy the analogous linear systems from Proposition 4.18.

\begin{theorem}[Existence, Uniqueness, and Smoothness for Linear ODEs]
\label{thm:linear-ode-existence-uniqueness}
\dcref{ch4:4.31}
\lean{LeeLib.Ch04.exists_hasDerivWithinAt_Icc, LeeLib.Ch04.IsSolOn, LeeLib.Ch04.flowMap, LeeLib.Ch04.flowMap_injective}
Let $I\subseteq\mathbb R$ be an open interval and let
$A_i^k\colon I\to\mathbb R$ be smooth for $1\leq i,k\leq n$. For every
$t_0\in I$ and $c=(c^1,\ldots,c^n)\in\mathbb R^n$, the problem
\begin{equation}
\dot V^k(t)=A_i^k(t)V^i(t) \tag{4.20}
\end{equation}
with $V^k(t_0)=c^k$ has a unique smooth solution on all of $I$. The solution
depends smoothly on $(t,c)\in I\times\mathbb R^n$.
\end{theorem}

\begin{proof}
\sketch After translating the parameter, take $t_0=0$. On
$I\times\mathbb R^n$ define
\[
Y=\frac{\partial}{\partial x^0}
+\sum_{k=1}^n A_j^k(x^0)x^j\frac{\partial}{\partial x^k}.
\]
Solutions correspond to integral curves $\eta(t)=(t,V(t))$ with
\begin{equation}
\eta(0)=(0,c^1,\ldots,c^n). \tag{4.21}
\end{equation}
The flow theorem gives local existence, uniqueness, and smooth dependence. To
show the maximal solution fills $I=(a,b)$, suppose its right endpoint is
$b_0<b$. On $[0,b_0]$, choose $M$ with $|A(t)|\leq M$ in Frobenius norm. Then
\[
\frac d{dt}|V|^2=2V^TA(t)V\leq2M|V|^2,
\]
so
\[
\frac d{dt}\left(e^{-2Mt}|V(t)|^2\right)\leq0,
\qquad
|V(t)|\leq e^{Mt}|V(0)|.
\]
Thus $(t,V(t))$ remains in the compact set
$[0,b_0]\times\overline B_R(0)$, with $R=e^{Mb_0}|V(0)|$, contradicting the
escape lemma. Apply the same argument to $-Y$ for the left endpoint, and undo
the parameter translation.
\end{proof}

\begin{theorem}[Existence and Uniqueness of Parallel Transport]
\label{thm:existence-uniqueness-parallel-transport}
\dcref{ch4:4.32}
\uses{thm:covariant-derivative-along-curve,thm:linear-ode-existence-uniqueness}
\lean{LeeLib.Ch04.exists_isParallelAlongChart_Icc, LeeLib.Ch04.isParallelAlongChart_eqOn_Icc, LeeLib.Ch04.chartParallelTransport}
Let $M$ be a smooth manifold, possibly with boundary, with a connection in $TM$.
Given a smooth curve $\gamma\colon I\to M$, $t_0\in I$, and either
$v\in T_{\gamma(t_0)}M$ or $v\in T^{(k,l)}_{\gamma(t_0)}M$, there is a unique
parallel vector or tensor field $V$ along $\gamma$ satisfying $V(t_0)=v$.
\end{theorem}

\begin{proof}
\sketch If $\gamma(I)$ lies in one chart, solve (4.19), or the analogous tensor
system, on all of $I$ using Theorem \ref{thm:linear-ode-existence-uniqueness}.
In general, extend the solution through successive chart-covered subintervals.
If the supremum of reachable parameters were an interior point of $I$, a chart
around its image and the single-chart result would extend the field farther.
The same argument to the left of $t_0$ gives existence on all of $I$; local
uniqueness propagates across overlaps.
\end{proof}

The resulting field is the \emph{parallel transport} of $v$ along $\gamma$. For
$t_0,t_1\in I$, define
\begin{equation}
P_{t_0t_1}^\gamma\colon T_{\gamma(t_0)}M\longrightarrow T_{\gamma(t_1)}M,
\qquad P_{t_0t_1}^\gamma(v)=V(t_1). \tag{4.22}
\end{equation}
This map is linear, and $P_{t_1t_0}^\gamma$ is its inverse.

For an admissible curve $\gamma\colon[a,b]\to M$, a \emph{piecewise smooth vector
field along $\gamma$} is a continuous $V\colon[a,b]\to TM$, with
$V(t)\in T_{\gamma(t)}M$, that is smooth on every subinterval of some admissible
partition. It is parallel when $D_tV=0$ wherever both fields are smooth.

\begin{corollary}[Parallel Transport Along Piecewise Smooth Curves]
\label{cor:parallel-transport-piecewise-smooth-curves}
\dcref{ch4:4.33}
\lean{LeeLib.Ch04.exists_isPiecewiseParallelAlongChart_Icc, LeeLib.Ch04.isPiecewiseParallelAlongChart_eqOn_Icc}
Let $M$ be a smooth manifold, possibly with boundary, with a connection in $TM$.
For an admissible curve $\gamma\colon[a,b]\to M$ and initial vector
$v\in T_{\gamma(a)}M$ or tensor $v\in T^{(k,l)}(T_{\gamma(a)}M)$, there is a
unique piecewise smooth parallel field $V$ along $\gamma$ with $V(a)=v$.
Moreover, $V$ is smooth wherever $\gamma$ is smooth.
\end{corollary}

\begin{proof}
\sketch On an admissible partition $(a_0,\ldots,a_k)$, transport $v$ along the
first smooth segment and use its endpoint value as the initial value on the next
segment. Induction gives existence, and segmentwise uniqueness gives global
uniqueness.
\end{proof}

Transporting a basis of $T_{\gamma(t_0)}M$ produces a \emph{parallel frame}
$(E_1,\ldots,E_n)$ along $\gamma$. Since parallel transport is an isomorphism,
this is a basis at each time. If $V=V^iE_i$, then
\begin{equation}
D_tV(t)=\dot V^i(t)E_i(t) \tag{4.23}
\end{equation}
wherever $V$ and $\gamma$ are smooth. Hence $V$ is parallel exactly when its
components in a parallel frame are constant.

\begin{theorem}[Parallel Transport Determines Covariant Differentiation]
\label{thm:parallel-transport-determines-covariant-diff}
\dcref{ch4:4.34}
\uses{cor:parallel-transport-piecewise-smooth-curves}
For a smooth curve $\gamma\colon I\to M$, a smooth vector field $V$ along it,
and $t_0\in I$,
\begin{equation}
D_tV(t_0)=\lim_{t_1\to t_0}
\frac{P_{t_1t_0}^\gamma V(t_1)-V(t_0)}{t_1-t_0}. \tag{4.24}
\end{equation}
\end{theorem}

\begin{proof}
\sketch Write $V(t)=V^i(t)E_i(t)$ in a parallel frame. Equation (4.23) gives
$D_tV(t_0)=\dot V^i(t_0)E_i(t_0)$, while
\[
P_{t_1t_0}^\gamma V(t_1)=V^i(t_1)E_i(t_0).
\]
The difference quotient has the same limit.
\end{proof}

\begin{corollary}[Parallel Transport Determines the Connection]
\label{cor:parallel-transport-determines-connection}
\dcref{ch4:4.35}
\uses{thm:parallel-transport-determines-covariant-diff}
Let $X,Y$ be smooth vector fields. For $p\in M$ and any smooth curve
$\gamma\colon I\to M$ satisfying $\gamma(0)=p$ and $\gamma'(0)=X_p$,
\begin{equation}
\nabla_XY|_p=\lim_{h\to0}
\frac{P_{h0}^\gamma Y(\gamma(h))-Y_p}{h}. \tag{4.25}
\end{equation}
\end{corollary}

\begin{proof}
\sketch Restrict $Y$ to $\gamma$. The defining property of covariant
differentiation along curves gives $\nabla_XY|_p=D_t(Y\circ\gamma)(0)$; apply
Theorem \ref{thm:parallel-transport-determines-covariant-diff}.
\end{proof}

A smooth vector or tensor field on $M$ is \emph{parallel} when it is parallel
along every smooth curve.

\begin{proposition}
\label{prop:parallel-field-characterization}
\dcref{ch4:4.36}
Let $M$ be a smooth manifold, possibly with boundary, with a connection $\nabla$
in $TM$. A smooth vector or tensor field $A$ is parallel on $M$ if and only if
$\nabla A\equiv0$.
\end{proposition}

\section{Pullback Connections}

For a diffeomorphism $\varphi\colon M\to\widetilde M$, write $\varphi_*X$ for
the vector field satisfying
$d\varphi_p(X_p)=(\varphi_*X)_{\varphi(p)}$.

\begin{lemma}[Pullback Connections]
\label{lem:pullback-connection}
\dcref{ch4:4.37}
\uses{lem:pushforward-vector-field}
\lean{LeeLib.Ch04.TangentConnection.pullback}
\leanok
Let $M,\widetilde M$ be smooth manifolds, possibly with boundary, let
$\widetilde\nabla$ be a connection in $T\widetilde M$, and let
$\varphi\colon M\to\widetilde M$ be a diffeomorphism. Then
\begin{equation}
(\varphi^*\widetilde\nabla)_XY
=(\varphi^{-1})_*
\left(\widetilde\nabla_{\varphi_*X}(\varphi_*Y)\right) \tag{4.26}
\end{equation}
defines a connection in $TM$, called the \emph{pullback connection}.
\end{lemma}

\begin{proof}
\sketch Real linearity in $Y$ is immediate. For $f\in C^\infty(M)$, set
$\widetilde f=f\circ\varphi^{-1}$. Since
$\varphi_*(fX)=\widetilde f\,\varphi_*X$, equation (4.26) gives
\[
(\varphi^*\widetilde\nabla)_{fX}Y
=f(\varphi^*\widetilde\nabla)_XY.
\]
Also $(\varphi_*X)(\widetilde f)=(Xf)\circ\varphi^{-1}$, so the Leibniz rule for
$\widetilde\nabla$ yields
\[
(\varphi^*\widetilde\nabla)_X(fY)
=f(\varphi^*\widetilde\nabla)_XY+(Xf)Y.
\]
Thus (4.26) satisfies the connection axioms.
\end{proof}

\begin{proposition}[Properties of Pullback Connections]
\label{prop:pullback-connection-properties}
\dcref{ch4:4.38}
\uses{lem:pullback-connection}
\lean{LeeLib.Ch04.covariantDeriv_pullback_naturality, LeeLib.Ch04.pullback_covariantDeriv_apply}
Let $\nabla=\varphi^*\widetilde\nabla$ and let $\gamma\colon I\to M$ be smooth.
\begin{enumerate}
\item[(a)] For every smooth vector field $V$ along $\gamma$,
\[
d\varphi\circ D_tV=\widetilde D_t(d\varphi\circ V),
\]
where the two derivatives are taken along $\gamma$ and $\varphi\circ\gamma$.
\item[(b)] If $\gamma$ is a $\nabla$-geodesic, then $\varphi\circ\gamma$ is a
$\widetilde\nabla$-geodesic.
\item[(c)] For $t_0,t_1\in I$,
\[
d\varphi_{\gamma(t_1)}\circ P_{t_0t_1}^\gamma
=P_{t_0t_1}^{\varphi\circ\gamma}\circ d\varphi_{\gamma(t_0)}.
\]
\end{enumerate}
\end{proposition}
\section{Problems}

\begin{exercise}
\label{prob:embedded-submanifold-covariant-derivative}
\dcref{ch4:4-1}
\uses{prop:tangent-space-submanifold}
Let $M\subseteq\mathbb R^n$ be an embedded submanifold, let
$Y\in\mathfrak X(M)$, and define $\nabla_v^\top Y$ by (4.4) for
$p\in M$ and $v\in T_pM$.
\begin{enumerate}
\item[(a)] Prove that $\nabla_v^\top Y$ is independent of the chosen extension
$\widetilde Y$ of $Y$. You may use Proposition A.28.
\item[(b)] Prove invariance under rigid motions: if $F\in E(n)$ and
$\widetilde M=F(M)$, then
\[
dF_p(\nabla_v^\top Y)=\nabla_{dF_p(v)}^\top(F_*Y).
\]
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:lie-derivative-not-connection}
\dcref{ch4:4-2}
\uses{prop:lie-derivative-equals-bracket}
Let $M$ be a positive-dimensional smooth manifold and
$\mathcal L_XY=[X,Y]$.
\begin{enumerate}
\item[(a)] Prove that
$\mathcal L\colon\mathfrak X(M)\times\mathfrak X(M)\to\mathfrak X(M)$ is not a
connection.
\item[(b)] Find smooth vector fields $X,Y$ on $\mathbb R^2$ such that
$X=Y=\partial_1$ along the $x^1$-axis, but
$\mathcal L_X(\partial_2)$ and $\mathcal L_Y(\partial_2)$ differ there.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:transformation-law-connection}
\dcref{ch4:4-3}
\uses{prop:transformation-law-connection-coefficients}
\lean{LeeLib.Ch04.connectionCoeff_transformation_law}
\leanok
Prove Proposition 4.7, the transformation law for connection coefficients.
\end{exercise}

\begin{exercise}
\label{prob:characterizing-connections}
\dcref{ch4:4-4}
\uses{thm:connections-affine-space}
Prove Theorem 4.14 characterizing the affine space of connections.
\end{exercise}

\begin{exercise}
\label{prob:tensor-field-formulas}
\dcref{ch4:4-5}
\uses{prop:covariant-derivative-components}
Prove Proposition 4.16, the local component formulas for covariant derivatives of
tensor fields.
\end{exercise}

\begin{exercise}
\label{prob:torsion-tensor}
\dcref{ch4:4-6}
\uses{ex:covariant-hessian}
\lean{LeeLib.Ch04.torsion, LeeLib.Ch04.torsion_apply, LeeLib.Ch04.isSymmetric_iff}
Let $\nabla$ be a connection in $TM$ and define
\[
\tau(X,Y)=\nabla_XY-\nabla_YX-[X,Y].
\]
\begin{enumerate}
\item[(a)] Prove that $\tau$ is a $(1,2)$-tensor field, called the
\emph{torsion tensor} of $\nabla$.
\item[(b)] Call $\nabla$ \emph{symmetric} when $\tau\equiv0$. Prove that this is
equivalent to
\[
\Gamma^k_{ij}=\Gamma^k_{ji}
\]
in every coordinate frame. \emph{Warning.} The coefficients need not be symmetric
in a noncoordinate frame.
\item[(c)] Prove that $\nabla$ is symmetric if and only if the covariant Hessian
$\nabla^2u$ is a symmetric $2$-tensor for every $u\in C^\infty(M)$.
\item[(d)] Prove that the Euclidean connection on $\mathbb R^n$ is symmetric.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:figure-eight-immersion}
\dcref{ch4:4-7}
\uses{ex:figure-eight-curve}
Let $\gamma\colon(-\pi,\pi)\to\mathbb R^2$ be the figure-eight curve of
Example 4.23. Prove that $\gamma$ is an injective smooth immersion whose velocity
field is not extendible.
\end{exercise}

\begin{exercise}
\label{prob:vector-field-extendibility}
\dcref{ch4:4-8}
Let $M$ be a smooth manifold without boundary, let $I\subseteq\mathbb R$ be any
interval, bounded or unbounded and with or without endpoints, and let
$\gamma\colon I\to M$ be smooth.
\begin{enumerate}
\item[(a)] If $t_0\in I$ and $\gamma'(t_0)\ne0$, prove that some connected
neighborhood $J$ of $t_0$ in $I$ has the property that every smooth vector field
along $\gamma|_J$ is extendible.
\item[(b)] If $I$ is open or compact and $\gamma$ is a smooth embedding, prove
that every smooth vector field along $\gamma$ is extendible.
\end{enumerate}
\end{exercise}

\begin{exercise}[Two connections and their difference tensor]
\label{prob:connections-difference-tensor}
\dcref{ch4:4-9}
\uses{prob:torsion-tensor}
\lean{LeeLib.Ch04.sameTorsion_iff_differenceTensor_symm, LeeLib.Ch04.sameChartAcceleration_iff_chartGamma_diff_antisymm, LeeLib.Ch04.isGeodesicInChart_congr_of_diff_antisymm, LeeLib.Ch04.chartGamma_diff_antisymm_iff_differenceTensor_frame_antisymm, LeeLib.Ch04.differenceTensor_chartFrame_eq_sum}
Let $\nabla^0,\nabla^1$ be connections on $TM$, with difference tensor $D$.
\begin{enumerate}
\item[(a)] Prove that the two connections have the same torsion if and only if
$D$ is symmetric:
\[
D(X,Y)=D(Y,X).
\]
\item[(b)] Prove that they determine the same geodesics if and only if $D$ is
antisymmetric:
\[
D(X,Y)=-D(Y,X).
\]
\end{enumerate}
\end{exercise}

\begin{exercise}[Parallel vector fields and reparametrization]
\label{prob:parallel-reparametrization}
\dcref{ch4:4-10}
\lean{LeeLib.Ch04.covariantDerivAlongChart_comp, LeeLib.Ch04.isParallelAlongChart_comp}
Let $M$ carry a connection, let $\gamma\colon I\to M$ be smooth, and let
$Y\in\mathfrak X(\gamma)$. Prove that if $Y$ is parallel along $\gamma$, then its
induced field is parallel along every reparametrization of $\gamma$.
\end{exercise}

\begin{exercise}[Connection on Lie groups]
\label{prob:lie-group-connection}
\dcref{ch4:4-11}
\uses{prob:torsion-tensor}
Let $G$ be a Lie group.
\begin{enumerate}
\item[(a)] Prove that there is a unique connection $\nabla$ in $TG$ for which
every left-invariant vector field is parallel.
\item[(b)] Prove that the torsion of $\nabla$ vanishes if and only if the identity
component of $G$ is abelian.
\end{enumerate}
\end{exercise}

\begin{exercise}[Parallel tensor fields]
\label{prob:parallel-tensor-fields}
\dcref{ch4:4-12}
\uses{prop:parallel-field-characterization}
Prove Proposition 4.36 on the characterization of parallel fields: a vector or
tensor field $A$ is parallel if and only if $\nabla A=0$.
\end{exercise}

\begin{exercise}[Pullback connections]
\label{prob:pullback-connections}
\dcref{ch4:4-13}
\uses{prop:pullback-connection-properties}
Prove Proposition 4.38 on the naturality properties of pullback connections.
\end{exercise}

\begin{exercise}[Connection 1-forms and Cartan's first structure equation]
\label{prob:connection-1forms-cartan}
\dcref{ch4:4-14}
\uses{prob:torsion-tensor}
Let $M$ be a smooth $n$-manifold, let $\nabla$ be a connection in $TM$, let
$(E_i)$ be a local frame on an open $U\subseteq M$, and let $(\epsilon^i)$ be its
dual coframe.
\begin{enumerate}
\item[(a)] Prove that there is a unique matrix of smooth $1$-forms
$(\omega_i{}^j)$ on $U$, called the \emph{connection $1$-forms} of the frame,
such that
\[
\nabla_XE_i=\omega_i{}^j(X)E_j
\qquad(X\in\mathfrak X(U)).
\]
\item[(b)] Prove Cartan's first structure equation
\[
d\epsilon^j=\epsilon^i\wedge\omega_i{}^j+\tau^j,
\]
where the torsion $2$-forms $\tau^1,\ldots,\tau^n\in\Omega^2(M)$ are defined by
\[
\tau(X,Y)=\tau^j(X,Y)E_j.
\]
\end{enumerate}
\end{exercise}
