% Source: Lee, "Introduction to Riemannian Manifolds" (2nd ed.), Chapter 3 "Model Riemannian Manifolds", PDF pages 69-97 of the cited source.
% Transcribed from the cited source, with manual checks for a few pages, and organized according to leanblueprint
% conventions (semantic \label, bracketed titles, \uses dependency edges, \dcref to Lee's own numbering) below.

\section{Symmetries of Riemannian Manifolds}

For a Riemannian manifold $(M,g)$, $\operatorname{Iso}(M,g)$ acts on $M$ by composition. The manifold is \emph{homogeneous} if this action is transitive. Each isometry $\varphi$ has differentials $d\varphi_p:T_pM\to T_{\varphi(p)}M$ that are linear isometries. The isotropy subgroup $\operatorname{Iso}_p(M,g)$ fixes $p$ and acts on $T_pM$ through $I_p(\varphi)=d\varphi_p$. The manifold is \emph{isotropic at $p$} if this representation is transitive on the unit vectors of $T_pM$; it is isotropic if this holds at every point.

Let
\[
O(M)=\bigsqcup_{p\in M}\{\text{orthonormal bases of }T_pM\}.
\]
The induced action is
\begin{equation}
\varphi\cdot(b_1,\ldots,b_n)=\bigl(d\varphi_p(b_1),\ldots,d\varphi_p(b_n)\bigr).\tag{3.1}
\end{equation}
The manifold is \emph{frame-homogeneous} when this action is transitive. The converses below fail in general: the induced paraboloid is isotropic at one point only, Berger metrics are homogeneous but not isotropic, and Fubini--Study metrics are homogeneous and isotropic but not frame-homogeneous.

\begin{proposition}\label{prop:frame-homogeneous-manifolds}\dcref{ch3:3.1}
Let $(M,g)$ be a Riemannian manifold.
\begin{enumerate}
\item[(a)] If $M$ is isotropic at one point and homogeneous, then it is isotropic.
\item[(b)] If $M$ is frame-homogeneous, then it is homogeneous and isotropic.
\end{enumerate}
\end{proposition}
\uses{prob:homogeneous-isotropic-riemannian}

\section{Euclidean Spaces}

Euclidean space is $(\mathbb R^n,\bar g)$ with the metric from (2.8). An $n$-dimensional inner product space $V$ carries the constant metric $g(v,w)=\langle v,w\rangle$ under $T_pV\cong V$; an orthonormal basis identifies it isometrically with $(\mathbb R^n,\bar g)$.

The Euclidean group is
\[
E(n)=\mathbb R^n\rtimes O(n),\qquad (b,A)(b',A')=(b+Ab',AA'),
\]
with faithful representation and action
\[
\rho(b,A)=\begin{pmatrix}A&b\\0&1\end{pmatrix}.
\]
\begin{equation}
(b,A)\cdot x=b+Ax.\tag{3.2}
\end{equation}
Translations and orthogonal linear maps preserve $\bar g$, and this action is transitive on $O(\mathbb R^n)$; Euclidean space is frame-homogeneous.

\section{Spheres}

For $R>0$, let $S^n(R)\subset\mathbb R^{n+1}$ be the radius-$R$ sphere with round metric $\bar g_R$ induced by the Euclidean metric; write $\bar g=\bar g_1$. The restriction of the $O(n+1)$ action is isometric.

\begin{proposition}
\label{prop:frame-homogeneous-sphere}
\dcref{ch3:3.2}
The group $O(n+1)$ acts transitively on $O(S^n(R))$, and thus each round sphere is frame-homogeneous.
\end{proposition}
\begin{proof}
\sketch Given $p\in S^n(R)$ and an orthonormal basis $(b_i)$ of $T_pS^n(R)$, the radial vector $p/R$ is orthogonal to all $b_i$. The matrix with columns $(b_1,\ldots,b_n,p/R)$ lies in $O(n+1)$ and sends the north-pole frame and north pole to $(b_i)$ and $p$.
\end{proof}

Metrics $g_1,g_2$ are \emph{conformally related} if $g_2=fg_1$ for positive smooth $f$. A diffeomorphism $\varphi:(M,g)\to(\widetilde M,\widetilde g)$ is conformal if $\varphi^*\widetilde g=fg$; conformal equivalence is existence of such a map. A manifold is \emph{locally conformally flat} if every point has a neighborhood conformally equivalent to an open Euclidean set.

\begin{exercise}\label{exer:conformality-equivalence-relation}\dcref{ch3:3.3}
\begin{enumerate}
\item[(a)] Show that conformality is an equivalence relation on Riemannian metrics on a fixed smooth manifold.
\item[(b)] Show that conformal equivalence is an equivalence relation on Riemannian manifolds.
\end{enumerate}
\end{exercise}

\begin{exercise}\label{exer:conformal-volume-forms}\dcref{ch3:3.4}
If $g_2=fg_1$ on an oriented $n$-manifold, prove
\[
dV_{g_2}=f^{n/2}dV_{g_1}.
\]
\end{exercise}

Stereographic projection from $N=(0,R)$ is determined by
\begin{align*}
u^i&=\lambda\xi^i,\\
-R&=\lambda(\tau-R).
\tag{3.3}
\end{align*}
Thus
\begin{equation}
\sigma(\xi,\tau)=\frac{R\xi}{R-\tau},\tag{3.4}
\end{equation}
and solving back gives
\begin{equation}
\xi^i=\frac{u^i}{\lambda},\qquad \tau=R\frac{\lambda-1}{\lambda},\tag{3.5}
\end{equation}
\begin{equation}
\sigma^{-1}(u)=\left(\frac{2R^2u}{|u|^2+R^2},R\frac{|u|^2-R^2}{|u|^2+R^2}\right).\tag{3.6}
\end{equation}

\begin{proposition}[Stereographic projection is a conformal diffeomorphism between $S^n(R) \setminus \{N\}$ and $\mathbb{R}^n$]
\label{prop:stereographic-conformal-diffeomorphism}
\dcref{ch3:3.5}
The map $\sigma:S^n(R)\setminus\{N\}\to\mathbb R^n$ is a diffeomorphism and
\[
(\sigma^{-1})^*\bar g_R=\frac{4R^4}{(|u|^2+R^2)^2}\bar g.
\]
\end{proposition}
\begin{proof}
\sketch Formula (3.6) gives a smooth inverse. Differentiating its components and substituting into the Euclidean metric cancels the mixed terms and yields the displayed positive conformal factor.
\end{proof}

\begin{corollary}[Each sphere with a round metric is locally conformally flat]
\label{cor:sphere-round-metric-locally-conformally-flat}
\dcref{ch3:3.6}
\uses{prop:stereographic-conformal-diffeomorphism}
Every round sphere is locally conformally flat.
\end{corollary}
\begin{proof}
\sketch Stereographic projection handles neighborhoods away from the north pole; rotate the sphere, or project from the south pole, to handle the north pole.
\end{proof}

\section{Hyperbolic Spaces}

\begin{theorem}
\label{thm:hyperbolic-space-models}
\dcref{ch3:3.7}
Let $n>1$ and $R>0$. The following models are mutually isometric:
\begin{enumerate}
\item[(a)] The hyperboloid $\mathbb H^n(R)=\{(\xi,\tau):|\xi|^2-\tau^2=-R^2,\ \tau>0\}\subset\mathbb R^{n,1}$ with
\begin{equation}
\bar g_R=\iota^*\bar g,\qquad \bar g=\sum_{i=1}^n(d\xi^i)^2-(d\tau)^2.\tag{3.8}
\end{equation}
\item[(b)] The Beltrami--Klein ball $\mathbb K^n(R)=\{|w|<R\}$ with
\begin{equation}
\bar g_R=R^2\frac{\sum_i(dw^i)^2}{R^2-|w|^2}+R^2\frac{(\sum_iw^idw^i)^2}{(R^2-|w|^2)^2}.\tag{3.9}
\end{equation}
\item[(c)] The Poincare ball $\mathbb B^n(R)=\{|u|<R\}$ with
\[
\bar g_R=4R^4\frac{\sum_i(du^i)^2}{(R^2-|u|^2)^2}.
\]
\item[(d)] The Poincare half-space $\mathbb U^n(R)=\{(x,y):y>0\}$ with
\[
\bar g_R=R^2\frac{\sum_{i=1}^{n-1}(dx^i)^2+dy^2}{y^2}.
\]
\end{enumerate}
\end{theorem}
\begin{proof}
\sketch For $f(\xi,\tau)=|\xi|^2-\tau^2$, the Minkowski gradient is
\begin{equation}
\operatorname{grad}f=2\sum_i\xi^i\frac{\partial}{\partial\xi^i}+2\tau\frac{\partial}{\partial\tau},\tag{3.10}
\end{equation}
with squared norm $-4R^2$ on the hyperboloid; Corollary 2.71 makes its induced metric Riemannian.

Central projection $c:\mathbb H^n(R)\to\mathbb K^n(R)$ and its inverse are
\[
c(\xi,\tau)=\frac{R\xi}{\tau},\tag{3.11}
\]
\[
c^{-1}(w)=\left(\frac{Rw}{\sqrt{R^2-|w|^2}},\frac{R^2}{\sqrt{R^2-|w|^2}}\right).\tag{3.12}
\]
Differentiation shows that $c$ pulls the Klein metric back to the hyperboloid metric.

Hyperbolic stereographic projection $\pi:\mathbb H^n(R)\to\mathbb B^n(R)$ from $(0,-R)$ satisfies
\begin{align}
u^i&=\lambda\xi^i,\label{eq:3.13a}\\
R&=\lambda(\tau+R).\label{eq:3.13b}
\end{align}
Therefore
\[
\pi(\xi,\tau)=\frac{R\xi}{R+\tau},\qquad
|\pi(\xi,\tau)|^2=R^2\frac{\tau^2-R^2}{(\tau+R)^2}<R^2,
\]
\[
\pi^{-1}(u)=\left(\frac{2R^2u}{R^2-|u|^2},\frac{R(R^2+|u|^2)}{R^2-|u|^2}\right),
\]
and
\[
(\pi^{-1})^*\bar g_R=\frac{4R^4\sum_j(du^j)^2}{(R^2-|u|^2)^2}.
\]

The generalized Cayley map $\kappa:\mathbb U^n(R)\to\mathbb B^n(R)$ is, in dimension two,
\begin{equation}
\kappa(z)=iR\frac{z-iR}{z+iR},\tag{3.14}
\end{equation}
and in real coordinates, with $x\in\mathbb R^{n-1}$,
\begin{equation}
\kappa(x,y)=\left(\frac{2R^2x}{|x|^2+(y+R)^2},R\frac{|x|^2+|y|^2-R^2}{|x|^2+(y+R)^2}\right).\tag{3.15}
\end{equation}
Its inverse is
\[
\kappa^{-1}(u,v)=\left(\frac{2R^2u}{|u|^2+(v-R)^2},R\frac{R^2-|u|^2-v^2}{|u|^2+(v-R)^2}\right),
\]
and direct substitution gives $\kappa^*\bar g_R^{\mathbb B}=\bar g_R^{\mathbb U}$.
\end{proof}

We use $\mathbb H^n(R)$ for any model in the theorem and $\bar g_R$ for its corresponding metric; $R=1$ is denoted $(\mathbb H^n,\bar g)$.

\begin{corollary}[Each hyperbolic space is locally conformally flat]
\label{cor:hyperbolic-locally-conformally-flat}
\dcref{ch3:3.8}
Every hyperbolic space is locally conformally flat.
\end{corollary}
\begin{proof}
\sketch In the Poincare ball and half-space models, the displayed metrics are positive smooth factors times the Euclidean metric.
\end{proof}

Let $O(n,1)$ preserve the Minkowski form, and let $O^+(n,1)$ preserve the upper hyperboloid sheet. Its restriction acts isometrically on the hyperboloid model.

\begin{proposition}
\label{prop:lorentz-group-transitive}
\dcref{ch3:3.9}
\uses{prop:frame-homogeneous-sphere}
The group $O^+(n,1)$ acts transitively on $O(\mathbb H^n(R))$, and therefore $\mathbb H^n(R)$ is frame-homogeneous.
\end{proposition}
\begin{proof}
\sketch At $p\in\mathbb H^n(R)$, $p/R$ is Minkowski-unit and normal to the tangent space by (3.10). Appending it to an orthonormal tangent frame gives a Minkowski-orthonormal basis; the matrix with these columns lies in $O^+(n,1)$ and sends the north-pole frame to the prescribed frame.
\end{proof}
\section{Invariant Metrics on Lie Groups}

Let $G$ be a Lie group. A Riemannian metric $g$ on $G$ is \emph{left-invariant} if
$L_\varphi^*g=g$ for every $\varphi\in G$; right-invariance is defined analogously
using right translations, and \emph{bi-invariant} means both left- and
right-invariant.

\begin{lemma}
\label{lem:left-invariant-metrics-lie-group}
\dcref{ch3:3.10}
Let $\mathfrak g$ be the Lie algebra of $G$.
\begin{enumerate}[label=(\alph*)]
\item A Riemannian metric $g$ is left-invariant if and only if $g(X,Y)$ is
constant on $G$ for all $X,Y\in\mathfrak g$.
\item Under $T_eG\cong\mathfrak g$, the map $g\mapsto g_e\in\Sigma^2(T_e^*G)$
is a bijection from left-invariant Riemannian metrics on $G$ to inner products
on $\mathfrak g$.
\end{enumerate}
\end{lemma}

\begin{exercise}
\label{exer:prove-preceding-lemma}
\dcref{ch3:3.11}
Prove Lemma \ref{lem:left-invariant-metrics-lie-group}.
\end{exercise}

Consequently, every inner product on $\mathfrak g$ extends uniquely to a
left-invariant metric, and similarly for right-invariant metrics. Either kind
is homogeneous.

For $\varphi\in G$, let $C_\varphi(\psi)=\varphi\psi\varphi^{-1}$ and
$\operatorname{Ad}(\varphi)=(C_\varphi)_*\colon\mathfrak g\to\mathfrak g$.

\begin{proposition}[Bi-invariance and the Adjoint Representation]
\label{prop:bi-invariant-metric-adjoint}
\dcref{ch3:3.12}
\uses{lem:left-invariant-metrics-lie-group}
Let $g$ be a left-invariant metric on $G$, with corresponding inner product
$\langle\cdot,\cdot\rangle$ on $\mathfrak g$. Then $g$ is bi-invariant if and
only if
\[
\langle\operatorname{Ad}(\varphi)X,\operatorname{Ad}(\varphi)Y\rangle
=\langle X,Y\rangle
\]
for all $\varphi\in G$ and $X,Y\in\mathfrak g$.
\end{proposition}
\begin{proof}
\sketch For $\varphi,\psi\in G$ and $X,Y\in\mathfrak g$, left-invariance gives
$(R_{\varphi^{-1}})_*X=\operatorname{Ad}(\varphi)X$ and hence
\[
((R_{\varphi^{-1}})^*g)_\psi(X_\psi,Y_\psi)
=\langle\operatorname{Ad}(\varphi)X,
\operatorname{Ad}(\varphi)Y\rangle.
\]
Thus all right translations preserve $g$ exactly when the inner product is
$\operatorname{Ad}(G)$-invariant.
\end{proof}

For a subgroup $H\subseteq\operatorname{GL}(V)$, an inner product on $V$ is
\emph{$H$-invariant} if every element of $H$ is orthogonal for that product.

\begin{lemma}[Compactness Criterion for Invariant Inner Products]
\label{lem:invariant-inner-product-compact}
\dcref{ch3:3.13}
Let $V$ be a finite-dimensional real vector space and
$H\subseteq\operatorname{GL}(V)$ a subgroup. An $H$-invariant inner product on
$V$ exists if and only if the closure of $H$ in $\operatorname{GL}(V)$ is
compact.
\end{lemma}
\begin{proof}
\sketch If an $H$-invariant inner product exists, then $H$ lies in its compact
orthogonal group, so its closure is compact. Conversely, let $K=\overline H$ be
compact, choose an inner product $(\cdot,\cdot)_0$, and let $\mu$ be a
right-invariant volume form on $K$. Define
\[
(x,y)=\int_K(kx,ky)_0\,\mu(k).
\]
This is symmetric, bilinear, and positive definite. For $k_0\in K$, right
invariance of $\mu$ yields
\[
(k_0x,k_0y)=\int_K(kk_0x,kk_0y)_0\,\mu(k)=(x,y),
\]
so the product is $K$-invariant and therefore $H$-invariant.
\end{proof}

\begin{theorem}[Existence of Bi-invariant Metrics]
\label{thm:existence-bi-invariant-metrics}
\dcref{ch3:3.14}
A Lie group $G$ admits a bi-invariant metric if and only if
$\operatorname{Ad}(G)$ has compact closure in
$\operatorname{GL}(\mathfrak g)$.
\end{theorem}
\begin{proof}
\sketch Combine Proposition \ref{prop:bi-invariant-metric-adjoint} with Lemma
\ref{lem:invariant-inner-product-compact}.
\end{proof}

\begin{corollary}
\label{cor:compact-lie-group-bi-invariant-metric}
\dcref{ch3:3.15}
Every compact Lie group admits a bi-invariant Riemannian metric.
\end{corollary}
\begin{proof}
\sketch The continuous image $\operatorname{Ad}(G)$ of compact $G$ is compact;
apply Theorem \ref{thm:existence-bi-invariant-metrics}.
\end{proof}

If $\operatorname{Ad}(G)$ has compact closure, every adjoint orbit is bounded.
An unbounded orbit therefore obstructs a bi-invariant metric.

\begin{example}[Invariant Metrics on Lie Groups]
\label{ex:invariant-metrics-lie-groups}
\dcref{ch3:3.16}
\uses{ex:n-torus-riemannian-submanifold,ex:adjoint-representations-lie-groups,prob:berger-sphere,prob:orthogonal-group-bi-invariant-metric,thm:hyperbolic-space-models}
\begin{enumerate}[label=(\alph*)]
\item On an abelian Lie group the adjoint representation is trivial, so every
left-invariant metric is bi-invariant. This includes the Euclidean metric on
$\mathbb R^n$ and the flat metric on $\mathbb T^n$.

\item The metric induced on a Lie subgroup by a left-invariant, respectively
bi-invariant, metric is again left-invariant, respectively bi-invariant.

\item The group $\operatorname{SL}(2,\mathbb R)$ has left-invariant metrics but
no bi-invariant metric. In
$\mathfrak{sl}(2,\mathbb R)$, take
\[
X_0=\begin{pmatrix}0&1\\0&0\end{pmatrix},\qquad
A_c=\begin{pmatrix}c&0\\0&1/c\end{pmatrix}\quad(c>0).
\]
Then
\[
\operatorname{Ad}(A_c)X_0=A_cX_0A_c^{-1}
=\begin{pmatrix}0&c^2\\0&0\end{pmatrix},
\]
so the orbit is unbounded. The same obstruction applies to
$\operatorname{SL}(n,\mathbb R)$ for $n\ge2$, and the subgroup inheritance in
(b) excludes bi-invariant metrics on $\operatorname{GL}(n,\mathbb R)$ for
$n\ge2$. The group $\operatorname{GL}(1,\mathbb R)$ is abelian.

\item The map
\[
(w,z)\longmapsto
\begin{pmatrix}w&z\\-\overline z&\overline w\end{pmatrix} \tag{3.16}
\]
identifies $S^3\subseteq\mathbb C^2$ diffeomorphically with
$\operatorname{SU}(2)$; under this identification the round metric is
bi-invariant.

\item On the skew-symmetric matrices $\mathfrak o(n)$,
\[
\langle A,B\rangle=\operatorname{tr}(A^TB)
\]
is $\operatorname{Ad}(O(n))$-invariant and induces a bi-invariant metric on
$O(n)$.

\item The upper half-space $\mathbb U^n$ is a Lie group under the matrix
realization
\[
(x,y)\longleftrightarrow
\begin{pmatrix}I_{n-1}&0\\x^T&y\end{pmatrix},
\qquad x\in\mathbb R^{n-1},\quad y>0.
\]
Its hyperbolic metric $\mathfrak g_h$ is left-invariant but not right-invariant.

\item For $n\ge1$, the $(2n+1)$-dimensional Heisenberg group is
\[
H_n=\left\{
\begin{pmatrix}1&x^T&z\\0&I&y\\0&0&1\end{pmatrix}:
x,y\in\mathbb R^n,\ z\in\mathbb R
\right\}.
\]
It is nilpotent: the lower central series
$G\supseteq[G,G]\supseteq[G,[G,G]]\supseteq\cdots$ terminates at the trivial
group, where $[G_1,G_2]$ is generated by the commutators
$x_1x_2x_1^{-1}x_2^{-1}$. The group $H_n$ has left-invariant metrics but no
bi-invariant metric.

\item The group
\[
\mathrm{Sol}=\left\{
\begin{pmatrix}e^z&0&x\\0&e^{-z}&y\\0&0&1\end{pmatrix}:
x,y,z\in\mathbb R
\right\}
\]
is solvable: its derived series
$G\supseteq[G,G]\supseteq[[G,G],[G,G]]\supseteq\cdots$ terminates at the
trivial group. It is nonnilpotent and has left-invariant metrics but no
bi-invariant metric.
\end{enumerate}
\end{example}

The Lie groups admitting bi-invariant metrics are precisely the direct products
of compact groups and abelian groups.

\section{Other Homogeneous Riemannian Manifolds}

A \emph{homogeneous $G$-space} is a smooth manifold with a smooth transitive
action of a Lie group $G$.

\begin{theorem}[Existence of Invariant Metrics on Homogeneous Spaces]
\label{thm:existence-invariant-metrics-homogeneous-spaces}
\dcref{ch3:3.17}
\uses{lem:invariant-inner-product-compact}
\uses{lem:invariant-inner-product-compact}
\uses{thm:equivariant-rank-theorem}
\uses{thm:local-section-theorem}
Let $M$ be a homogeneous $G$-space, $p_0\in M$, and
$I_{p_0}\colon G_{p_0}\to\operatorname{GL}(T_{p_0}M)$ the isotropy
representation. A $G$-invariant Riemannian metric on $M$ exists if and only if
$I_{p_0}(G_{p_0})$ has compact closure in
$\operatorname{GL}(T_{p_0}M)$.
\end{theorem}
\begin{proof}
\sketch A $G$-invariant metric restricts at $p_0$ to an
$I_{p_0}(G_{p_0})$-invariant inner product, so Lemma
\ref{lem:invariant-inner-product-compact} gives necessity. Conversely, choose
such an invariant inner product $g_{p_0}$ and, for $p\in M$, choose
$\varphi\in G$ with $\varphi(p)=p_0$ and set
\[
g_p=(d\varphi_p)^*g_{p_0}.
\]
Isotropy invariance makes this independent of $\varphi$ and $G$-invariant.

For the orbit map $\pi(\psi)=\psi\cdot p_0$ and action map
$\theta_\varphi(p)=\varphi\cdot p$,
\begin{equation}
\pi\circ L_\varphi=\theta_\varphi\circ\pi. \tag{3.17}
\end{equation}
The equivariant rank theorem makes $\pi$ a submersion. The pullback
$\tau=\pi^*g$ is left-invariant, hence smooth because its coefficients in a
left-invariant frame are constant. A local section $\sigma\colon U\to G$ of
$\pi$ then gives $g|_U=\sigma^*\tau$, proving smoothness.
\end{proof}

\begin{corollary}[Compact Isotropy Gives an Invariant Metric]
\label{cor:invariant-metric-compact-isotropy}
\dcref{ch3:3.18}
\uses{thm:existence-invariant-metrics-homogeneous-spaces}
If a Lie group $G$ acts smoothly and transitively on a smooth manifold $M$ and
all isotropy groups are compact, then $M$ admits a $G$-invariant Riemannian
metric.
\end{corollary}

\begin{exercise}[Invariant Metrics and Isotropy Inner Products]
\label{exer:invariant-metrics-bijection}
\dcref{ch3:3.19}
Suppose a homogeneous $G$-space $M$ admits a $G$-invariant metric. For every
$p\in M$, prove that $g\mapsto g_p$ bijects the $G$-invariant metrics on $M$
with the $I_p(G_p)$-invariant inner products on $T_pM$.
\end{exercise}

\section{Locally Homogeneous Riemannian Manifolds}

A Riemannian manifold $(M,g)$ is \emph{locally homogeneous} if, for every
$p,q\in M$, some isometry between neighborhoods of $p$ and $q$ maps $p$ to
$q$. It is \emph{locally frame-homogeneous} if the isometry can additionally be
chosen to map any prescribed orthonormal basis of $T_pM$ to any prescribed
orthonormal basis of $T_qM$. Homogeneity and frame homogeneity imply their local
counterparts, which also pass to open submanifolds.

\begin{proposition}[Riemannian Coverings Preserve Local Homogeneity]
\label{prop:covering-locally-homogeneous}
\dcref{ch3:3.20}
Let $(\widetilde M,\widetilde g)$ be homogeneous and let
$\pi\colon(\widetilde M,\widetilde g)\to(M,g)$ be a Riemannian covering. Then
$(M,g)$ is locally homogeneous. If $\widetilde M$ is frame-homogeneous, then
$M$ is locally frame-homogeneous.
\end{proposition}

\begin{exercise}
\label{exer:prove-locally-homogeneous-covering}
\dcref{ch3:3.21}
Prove Proposition \ref{prop:covering-locally-homogeneous}.
\end{exercise}

\begin{theorem}[Uniformization of Compact Surfaces]
\label{thm:uniformization-compact-surfaces}
\dcref{ch3:3.22}
Every compact connected smooth $2$-manifold admits a locally
frame-homogeneous Riemannian metric and a Riemannian covering by
$\mathbb R^2$, the hyperbolic plane, or the round unit sphere.
\end{theorem}
\begin{proof}
\sketch By the classification of compact surfaces, it suffices to treat the
sphere, connected sums of tori, and connected sums of projective planes.
The round $S^2$ covers itself, the flat torus is covered by $\mathbb R^2$, and
$\mathbb{RP}^2$ has a two-sheeted round covering by $S^2$.

For an orientable surface of genus $n\ge2$, use the side word
$a_1b_1a_1^{-1}b_1^{-1}\cdots a_nb_na_n^{-1}b_n^{-1}$ on a regular
$4n$-gon in the Poincar\'e disk, with all interior angles $\pi/(2n)$. The
orientation-preserving isometry group contains
\begin{equation}
G=\left\{
\begin{pmatrix}\alpha&\beta\\\overline\beta&\overline\alpha\end{pmatrix}:
\alpha,\beta\in\mathbb C,\ |\alpha|^2-|\beta|^2>0
\right\}. \tag{3.18}
\end{equation}
The side-pairing transformations generate a discrete subgroup $\Gamma_n$ for
which $\mathbb B^2/\Gamma_n$ is the genus-$n$ surface.

The connected sum of two projective planes is the Klein bottle, represented by
\begin{equation}
\begin{aligned}
(x,0)&\sim(x,1) &&(0\le x\le1),\\
(0,y)&\sim(1,1-y) &&(0\le y\le1).
\end{aligned}
\tag{3.19}
\end{equation}
It is also $\mathbb R^2/\Gamma$, where
\begin{equation}
\Gamma=\left\{(b,A)\in E(2):
b=(k,\ell),\ k,\ell\in\mathbb Z,\quad
A=\begin{pmatrix}1&0\\0&(-1)^k\end{pmatrix}
\right\}. \tag{3.20}
\end{equation}
This action is free, proper, and isometric, so the quotient is flat. For a
connected sum of $n\ge3$ projective planes, use a regular hyperbolic $2n$-gon
with side word $a_1a_1a_2a_2\cdots a_na_n$ and interior angles $\pi/n$; the
side pairings lie in the full hyperbolic isometry group because the quotient is
nonorientable. Details are given in \cite{Ive92}.

Each quotient is locally frame-homogeneous by Proposition
\ref{prop:covering-locally-homogeneous}. Finally, uniqueness of smooth
structures on surfaces up to diffeomorphism allows the constructed metric to
be pulled back to the given smooth structure \cite{Mun56}.
\end{proof}

Every compact orientable $3$-manifold is a connected sum of compact manifolds,
each of which either admits a Riemannian covering by a homogeneous Riemannian
manifold or can be cut along embedded tori into pieces carrying finite-volume
locally homogeneous metrics. The simply connected homogeneous Riemannian
$3$-manifolds admitting finite-volume quotients are exactly
\[
\mathbb R^3,\quad S^3,\quad\mathbb H^3,\quad
S^2\times\mathbb R,\quad\mathbb H^2\times\mathbb R,
\]
with their standard product metrics where applicable, together with $H_1$,
$\mathrm{Sol}$, and the universal covering group of
$\operatorname{SL}(2,\mathbb R)$, each with a left-invariant metric; see
with references \cite{Thu97}, \cite{Sco83}, and
\cite[KL08,MF10,MT14]{BBBMP}.

\subsection{Symmetric Spaces}

A \emph{point reflection at $p$} is an isometry $\varphi\colon M\to M$ such
that $\varphi(p)=p$ and $d\varphi_p=-\operatorname{Id}_{T_pM}$. A connected
Riemannian manifold is a \emph{Riemannian symmetric space} if it has a point
reflection at every point. It is \emph{locally symmetric} if every point has a
neighborhood admitting a point reflection at that point. Every symmetric space
is locally symmetric, and every Riemannian symmetric space is homogeneous.

\begin{lemma}
\label{lem:homogeneous-point-reflection-symmetric}
\dcref{ch3:3.23}
A connected homogeneous Riemannian manifold with a point reflection at one
point is symmetric.
\end{lemma}
\begin{proof}
\sketch Let $\varphi$ reflect at $p$. For $q\in M$, choose an isometry $\psi$
with $\psi(p)=q$. Then $\widetilde\varphi=\psi\circ\varphi\circ\psi^{-1}$ fixes
$q$ and
\[
d\widetilde\varphi_q
=d\psi_p\circ(-\operatorname{Id}_{T_pM})\circ d(\psi^{-1})_q
=-\operatorname{Id}_{T_qM}.
\]
\end{proof}

\begin{example}[Riemannian Symmetric Spaces]
\label{ex:riemannian-symmetric-spaces}
\dcref{ch3:3.24}
\uses{lem:homogeneous-point-reflection-symmetric,ex:fubini-study-metric,prob:grassmann-stiefel-manifolds,prob:fubini-study-metric-homogeneous,prob:grassmannian-metric-homogeneous}
\begin{enumerate}[label=(\alph*)]
\item Every connected frame-homogeneous Riemannian manifold is symmetric:
frame homogeneity supplies an isometry fixing $p$ and taking an orthonormal
basis $(b_i)$ to $(-b_i)$. Hence Euclidean, spherical, and hyperbolic spaces are
symmetric.

\item If a connected Lie group $G$ has a bi-invariant metric, inversion
$\Phi(x)=x^{-1}$ satisfies $d\Phi_e=-\operatorname{Id}$ and is an isometry.
Indeed, bi-invariance and
$\Phi=R_{p^{-1}}\circ\Phi\circ L_{p^{-1}}$ give
\[
(\Phi^*g)_p
=d(L_{p^{-1}})_p^*\,d\Phi_e^*\,d(R_{p^{-1}})_e^*g_p=g_p.
\]
Thus inversion reflects at $e$, and Lemma
\ref{lem:homogeneous-point-reflection-symmetric} applies.

\item Complex projective spaces and Grassmann manifolds, with their standard
metrics, are symmetric spaces.

\item Products of symmetric spaces are symmetric for the product metric. A
symmetric space is \emph{irreducible} if it is not isometric to a product of
positive-dimensional symmetric spaces.
\end{enumerate}
\end{example}
\section{Model Pseudo-Riemannian Manifolds}

For $r,s\geq0$ and $R>0$, define the pseudosphere and pseudohyperbolic space by
\[
S^{r,s}(R)=\left\{(\xi,\tau)\in\mathbb R^{r+1,s}:
\sum_{i=1}^{r+1}(\xi^i)^2-\sum_{j=1}^{s}(\tau^j)^2=R^2\right\},
\]
\[
\mathbb H^{r,s}(R)=\left\{(\xi,\tau)\in\mathbb R^{r,s+1}:
\sum_{i=1}^{r}(\xi^i)^2-\sum_{j=1}^{s+1}(\tau^j)^2=-R^2\right\}.
\]
Equip them with the induced metrics
\[
\bar g_R^{(r,s)}=\iota^*\bar g^{(r+1,s)}
\quad\text{on }S^{r,s}(R),
\qquad
\bar g_R^{(r,s)}=\iota^*\bar g^{(r,s+1)}
\quad\text{on }\mathbb H^{r,s}(R).
\]

\begin{theorem}
\label{thm:pseudo-riemannian-signature}
\dcref{ch3:3.25}
For all $r,s\geq0$ and $R>0$, both $S^{r,s}(R)$ and
$\mathbb H^{r,s}(R)$ are pseudo-Riemannian manifolds of signature $(r,s)$.
\end{theorem}

For a pseudo-Riemannian manifold of signature $(r,s)$, call an orthonormal basis
standard-ordered when the positive vectors precede the negative vectors, so that
the dual-basis expression of the metric is
\[
(e^1)^2+\cdots+(e^r)^2-(e^{r+1})^2-\cdots-(e^{r+s})^2.
\]
Let $O(M)$ be the set of all standard-ordered orthonormal bases of tangent spaces.
The manifold is \emph{frame-homogeneous} when its isometry group acts transitively
on $O(M)$.

\begin{theorem}
\label{thm:frame-homogeneous-pseudo-riemannian}
\dcref{ch3:3.26}
All pseudo-Euclidean spaces, pseudospheres, and pseudohyperbolic spaces are
frame-homogeneous.
\end{theorem}

In signature $(n,1)$, $(S^{n,1}(R),\bar g_R^{(n,1)})$ is de Sitter space of
radius $R$, and $(\mathbb H^{n,1}(R),\bar g_R^{(n,1)})$ is anti-de Sitter space
of radius $R$.

\section{Problems}

\begin{exercise}
\label{prob:isometric-action-e-n}
\dcref{ch3:3-1}
Show that (3.2) defines a smooth isometric action of $\operatorname{E}(n)$ on
$(\mathbb R^n,\bar g)$ and that the induced action on $O(\mathbb R^n)$ is
transitive.
\end{exercise}

\begin{exercise}
\label{prob:rp-n-frame-homogeneous}
\dcref{ch3:3-2}
\uses{ex:real-projective-space-metric}
Prove that the metric on $\mathbb{RP}^n$ described in Example 2.34 is
frame-homogeneous.
\end{exercise}

\begin{exercise}
\label{prob:homogeneous-isotropic-riemannian}
\dcref{ch3:3-3}
\uses{prop:frame-homogeneous-manifolds}
Prove Proposition 3.1 concerning homogeneous and isotropic Riemannian manifolds.
\end{exercise}

\begin{exercise}
\label{prob:theorem-3-7-proof}
\dcref{ch3:3-4}
\uses{thm:hyperbolic-space-models}
Complete the proof of Theorem 3.7 by proving
\[
\kappa^*\bar g_R^3=\bar g_R^4.
\]
\end{exercise}

\begin{exercise}
\label{prob:sphere-hyperbolic-scaling}
\dcref{ch3:3-5}
\begin{enumerate}
\item[(a)] Prove that $(S^n(R),\bar g_R)$ is isometric to
$(S^n,R^2\bar g)$ for every $R>0$.
\item[(b)] Prove that $(\mathbb H^n(R),\bar g_R)$ is isometric to
$(\mathbb H^n,R^2\bar g)$ for every $R>0$.
\item[(c)] Explain why defining metrics on $\mathbb R^n$ by
$\bar g_R=R^2\bar g$ is unnecessary.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:conformal-metrics}
\dcref{ch3:3-6}
Show that Riemannian metrics $g_1,g_2$ are conformal if and only if they induce
the same angles, though not necessarily the same lengths. Show also that a
diffeomorphism is a conformal equivalence if and only if it preserves angles.
For the forward computation, use a $g_1$-orthonormal frame $(E_i)$ and compare
the $g_2$-angle between $E_i$ and
$(\cos\theta)E_i+(\sin\theta)E_j$.
\end{exercise}

\begin{exercise}
\label{prob:sl2r-upper-half-plane}
\dcref{ch3:3-7}
Let $\mathbb U^2$ be the upper half-plane with
$\bar g=(dx^2+dy^2)/y^2$, and let
\[
A\cdot z=\frac{az+b}{cz+d},
\qquad
A=\begin{pmatrix}a&b\\c&d\end{pmatrix}\in\operatorname{SL}(2,\mathbb R).
\]
Prove that this is a smooth, transitive, orientation-preserving isometric action
of $\operatorname{SL}(2,\mathbb R)$ on $(\mathbb U^2,\bar g)$. Determine whether
the induced action on $O(\mathbb U^2)$ is transitive.
\end{exercise}

\begin{exercise}
\label{prob:poincare-disk}
\dcref{ch3:3-8}
Let $\mathbb B^2$ be the Poincare disk with
$\bar g=(du^2+dv^2)/(1-u^2-v^2)^2$, and let $G\subseteq\operatorname{GL}(2,\mathbb C)$
be the subgroup defined by (3.18). For $w=u+iv$, define
\[
\begin{pmatrix}\alpha&\beta\\\bar\beta&\bar\alpha\end{pmatrix}\cdot w
=\frac{\alpha w+\beta}{\bar\beta w+\bar\alpha}.
\]
Prove that this gives a smooth, transitive, orientation-preserving isometric
action on $(\mathbb B^2,\bar g)$. One route is to conjugate the action on the
upper half-plane by the Cayley transform (3.14) with $R=1$, and use Problem 3-7.
\end{exercise}

\begin{exercise}
\label{prob:compact-lie-group-volume-form-bi-invariant}
\dcref{ch3:3-9}
Let $G$ be a connected compact Lie group with a left-invariant metric $g$ and
left-invariant orientation. Prove that its Riemannian volume form $dV_g$ is bi-invariant. You may
show that $dV_g$ equals the volume form of a bi-invariant metric.
\end{exercise}

\begin{exercise}
\label{prob:berger-sphere}
\dcref{ch3:3-10}
\uses{ex:invariant-metrics-lie-groups}
In $\mathfrak{su}(2)$ let
\[
X=\begin{pmatrix}0&1\\-1&0\end{pmatrix},\quad
Y=\begin{pmatrix}0&i\\i&0\end{pmatrix},\quad
Z=\begin{pmatrix}i&0\\0&-i\end{pmatrix}.
\]
For $a>0$, let $g_a$ be the left-invariant metric on $\operatorname{SU}(2)$ for
which $X,Y,aZ$ are orthonormal.
\begin{enumerate}
\item[(a)] Prove that $g_a$ is bi-invariant exactly when $a=1$.
\item[(b)] Prove that the map (3.16) is an isometry from
$(S^3,\bar g)$ to $(\operatorname{SU}(2),g_1)$.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:orthogonal-group-bi-invariant-metric}
\dcref{ch3:3-11}
\uses{ex:invariant-metrics-lie-groups}
Prove that
\[
\langle A,B\rangle=\operatorname{tr}(A^TB)
\]
defines a bi-invariant Riemannian metric on $O(n)$.
\end{exercise}

\begin{exercise}
\label{prob:hyperbolic-space-metrics}
\dcref{ch3:3-12}
\uses{ex:invariant-metrics-lie-groups}
Regard the upper half-space $\mathbb H^n$ as the Lie group of Example 3.16(f).
\begin{enumerate}
\item[(a)] Prove that $\bar g_R$ is left-invariant for every $R>0$.
\item[(b)] Prove that $\mathbb H^n$ admits no bi-invariant metric.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:heisenberg-group-metrics}
\dcref{ch3:3-13}
\uses{ex:invariant-metrics-lie-groups}
Write an explicit formula in global coordinates
$(x^1,\ldots,x^n,y^1,\ldots,y^n,z)$ for an arbitrary left-invariant metric on
the Heisenberg group $H_n$ of Example 3.16(g), and prove that $H_n$ has no
bi-invariant metric.
\end{exercise}

\begin{exercise}
\label{prob:sol-group-metrics}
\dcref{ch3:3-14}
\uses{prob:heisenberg-group-metrics,ex:invariant-metrics-lie-groups}
Repeat Problem 3-13 for the group Sol of Example 3.16(h).
\end{exercise}

\begin{exercise}
\label{prob:minkowski-hyperboloid}
\dcref{ch3:3-15}
Let $\mathbb R^{n,1}$ have coordinates $(\xi,\tau)$ and metric
$\bar g^{(n,1)}=\sum_i d(\xi^i)^2-d\tau^2$, and set
\[
S=\{(\xi,\tau):(\xi^1)^2+\cdots+(\xi^n)^2=\tau=1\}.
\]
\begin{enumerate}
\item[(a)] Prove that $S$ is a smooth submanifold diffeomorphic to
$S^{n-1}$ and that its induced metric $g=\iota_S^*\bar g^{(n,1)}$ makes it isometric
to the round unit $(n-1)$-sphere.
\item[(b)] For $A\in O^+(n,1)$ and $p\in S$, let $\langle p\rangle$ be the line
spanned by $p$. Prove that $A(\langle p\rangle)$ meets $S$ in exactly one point,
and define $A\cdot p$ to be that point. Show that this is a smooth transitive
action on $S$.
\item[(c)] Prove that the action in (b) is by conformal diffeomorphisms of
$(S,g)$.
\end{enumerate}
\end{exercise}

\begin{exercise}
\label{prob:no-riemannian-metric-invariant-lorentz}
\dcref{ch3:3-16}
\uses{prob:minkowski-hyperboloid}
Prove that no Riemannian metric on the sphere is invariant under the action in
Problem 3-15.
\end{exercise}

\begin{exercise}
\label{prob:lie-group-adjoint-representation}
\dcref{ch3:3-17}
\uses{thm:existence-bi-invariant-metrics}
For a Lie group $G$, let $G\times G$ act on $G$ by
$(\varphi_1,\varphi_2)\cdot\psi=\varphi_1\psi\varphi_2^{-1}$. Prove that the
action is transitive and that the isotropy group of the identity is
\[
\Delta=\{(\varphi,\varphi):\varphi\in G\}.
\]
Then prove commutativity of
\[
\begin{array}{ccc}
\Delta & \cong & G \\
I_e \downarrow & & \downarrow \mathrm{Ad} \\
\mathrm{GL}(T_eG) & \cong & \mathrm{GL}(\mathfrak g)
\end{array}
\]
and use it to give an alternative proof of Theorem 3.14.
\end{exercise}

\begin{exercise}
\label{prob:klein-bottle-flat-covering}
\dcref{ch3:3-18}
Let $\Gamma\subset E(2)$ be the subgroup defined by (3.20). Prove that $\Gamma$
acts freely and properly on $\mathbb R^2$, that the orbit space is homeomorphic
to the Klein bottle, and that the Klein bottle consequently has a flat metric and
a Riemannian covering by the Euclidean plane.
\end{exercise}

\begin{exercise}
\label{prob:fubini-study-metric-homogeneous}
\dcref{ch3:3-19}
\uses{ex:fubini-study-metric}
Show that the Fubini--Study metric on $\mathbb{CP}^n$ is homogeneous, isotropic,
and symmetric.
\end{exercise}

\begin{exercise}
\label{prob:grassmannian-metric-homogeneous}
\dcref{ch3:3-20}
\uses{prob:grassmann-stiefel-manifolds}
Show that the metric on $G_k(\mathbb R^n)$ defined in Problem 2-7 is homogeneous
and symmetric.
\end{exercise}

\begin{exercise}
\label{prob:riemannian-quotients-conjugate-subgroups}
\dcref{ch3:3-21}
\uses{prop:riemannian-quotient-metric}
Let $(\widetilde M,\widetilde g)$ be simply connected, and let
$\Gamma_1,\Gamma_2\subset\operatorname{Iso}(\widetilde M,\widetilde g)$ be
countable subgroups acting smoothly, freely, and properly (with the discrete
topology). For $i=1,2$, put $M_i=\widetilde M/\Gamma_i$ and let $g_i$ make the
quotient maps $\pi_i$ Riemannian coverings. Prove that $(M_1,g_1)$ and
$(M_2,g_2)$ are isometric if and only if $\Gamma_1$ and $\Gamma_2$ are conjugate
subgroups of $\operatorname{Iso}(\widetilde M,\widetilde g)$.
\end{exercise}

\begin{exercise}
\label{prob:pseudospheres-pseudo-riemannian}
\dcref{ch3:3-22}
\uses{thm:pseudo-riemannian-signature,thm:hyperbolic-space-models}
Prove Theorem 3.25 on the pseudo-Riemannian signatures of pseudospheres and
pseudohyperbolic spaces. Mimic the proof that $\mathbb H^n(R)$ is Riemannian.
\end{exercise}

\begin{exercise}
\label{prob:frame-homogeneity-pseudo-spaces}
\dcref{ch3:3-23}
\uses{thm:frame-homogeneous-pseudo-riemannian}
Prove Theorem 3.26, asserting frame homogeneity of pseudo-Euclidean spaces,
pseudospheres, and pseudohyperbolic spaces.
\end{exercise}

\begin{exercise}
\label{prob:diffeomorphism-pseudohyperbolic-pseudosphere}
\dcref{ch3:3-24}
For positive integers $r,s$ and $R>0$, prove that both
$\mathbb H^{r,s}(R)$ and $S^{s,r}(R)$ are diffeomorphic to
$\mathbb R^r\times S^s$. Consider
\[
\varphi(x,y)=\left(Rx,\sqrt{1+|x|^2}\,Ry\right),
\qquad
\psi(x,y)=\left(\sqrt{1+|x|^2}\,Ry,Rx\right)
\]
from $\mathbb R^r\times S^s$ into $\mathbb R^{r+s+1}$.
\end{exercise}

\begin{exercise}
\label{prob:universal-anti-de-sitter}
\dcref{ch3:3-25}
\uses{prob:diffeomorphism-pseudohyperbolic-pseudosphere}
Let $K^r(R)$ be the radius-$R$ ball with Beltrami--Klein metric (3.9), and set
$\bar{\mathbb H}^{r,1}(R)=K^r(R)\times\mathbb R$ with
\[
q=\bar g_R^2\oplus(-f^2dt^2),
\qquad
f(w)=\frac{R^2}{\sqrt{R^2-|w|^2}}.
\]
Define
\[
F(w,t)=\frac{(Rw,R^2\cos t,R^2\sin t)}{\sqrt{R^2-|w|^2}}
\]
into $\mathbb R^{r,2}$. Prove that its image is
$\mathbb H^{r,1}(R)$ and that $F$ is a pseudo-Riemannian covering of
$(\mathbb H^{r,1}(R),q_R^{(r,1)})$ by $(\bar{\mathbb H}^{r,1}(R),q)$.
The constructions of warped-product metrics and coverings are understood in the
pseudo-Riemannian setting. Since Problem 3-24 gives simple connectivity for
$s\geq2$ but not for $s=1$, this covering is the universal anti-de Sitter space.
\end{exercise}
