\begin{thebibliography}{20}
\bibitem{Akutagawa-Ishida-Lebrun} K. Akutagawa, M. Ishida and C. Lebrun,
``Perelman's Invariant, Ricci Flow, and the Yamabe Invariants of Smooth Manifolds'',
Arch. Math. 88, p. 71-76 (2007)

\bibitem{Anderson3} M. Anderson,
``Scalar Curvature and Geometrization Conjectures for $3$-Manifolds'',
in \underline{Comparison Geometry}, MSRI Publ. 30,
Cambridge University Press, Cambridge, p. 49-82 (1997)

\bibitem{Anderson} M. Anderson,
``Scalar Curvature and the Existence of Geometric Structures on
$3$-Manifolds. I'', J. Reine Angew. Math. 553, p. 125-182 (2002)

\bibitem{Anderson2} M. Anderson,
``Canonical Metrics on 3-Manifolds and 4-Manifolds'',
Asian Journal of Math. 10, p. 127-164 (2006)

\bibitem{Anderson-Chow} G. Anderson and B. Chow,
``A Pointwise Bound for Solutions of the Linearized Ricci Flow
Equation on 3-Manifolds'', Calculus of Variations 23, p. 1-12 (2005)

\bibitem{Angenent-Knopf}  S. Angenent and D. Knopf,
``Precise Asymptotics of the Ricci Flow
Neckpinch'', Comm. Anal. Geom. 15, p. 773-844 (2007)

\bibitem{Ballmann-Gromov-Schroeder}
W. Ballmann, M. Gromov and V. Schroeder,
\underline{Manifolds of Nonpositive Curvature},
Progress in Mathematics 61,
Birkh\"auser, Boston (1985)

\bibitem{Beckner}
W. Beckner,
``Geometric Asymptotics and the Logarithmic Sobolev Inequality'',
Forum Math. 11, p. 105-137 (1999)

\bibitem{Besson} L. Bessi\`eres, G. Besson, M. Boileau, S. Maillot and
J. Porti, ``Weak Collapsing and Geometrisation of Aspherical 3-Manifolds'',
preprint, http://arxiv.org/abs/0706.2065 (2007)

\bibitem{Bourguignon} J.-P. Bourguignon, ``Une Stratification de l'Espace
des Structures Riemanniennes'',
Compositio Math. 30, p. 1-41 (1975)

\bibitem{Burago-Burago-Ivanov}
D. Burago, Y. Burago and S. Ivanov,
\underline{A Course in Metric Geometry},
Graduate Studies in Mathematics 33,
American Mathematical Society, Providence (2001)

\bibitem{Burago-Gromov-Perelman}
Y. Burago, M. Gromov and G. Perelman,
``A. D. Aleksandrov Spaces with Curvatures Bounded Below'',
Russian Math. Surveys 47, p. 1-58 (1992)

\bibitem{Calabi} E. Calabi,
``An Extension of E. Hopf's Maximum Principle with an Application to
Riemannian Geometry'',
Duke Math. J. 25, p. 45-56 (1957)

\bibitem{Cao-Chow} H-D. Cao and B. Chow,
``Recent Developments on the Ricci Flow'', Bull. of the AMS 36, p. 59-74 (1999)

\bibitem{Cao-Zhu} H.-D. Cao and X.-P. Zhu,
``A Complete Proof of the Poincar\'e and Geometrization Conjectures - Application of the Hamilton-Perelman theory of the Ricci Flow'',
Asian Journal of Math. 10, p. 165-492 (2006) \\
Erratum to
``A Complete Proof of the Poincar\'e and Geometrization Conjectures - Application of the Hamilton-Perelman theory of the Ricci Flow'',  Asian Journal of Math. 10, p. 663-664 (2006)

\bibitem{Chau-Tam-Yu (2007)} A. Chau, L.-F. Tam and C. Yu,
``Pseudolocality for the Ricci Flow and Applications'',
preprint, http://arxiv.org/abs/math/0701153 (2007)

\bibitem{Chavel} I. Chavel,
\underline{Riemannian Geometry - a Modern Introduction},
Cambridge Tracts in Mathematics 108,
Cambridge University Press, Cambridge (1993)

\bibitem{Cheeger-Ebin} J. Cheeger and D. Ebin,
\underline{Comparison Theorems in Riemannian Geometry}.
North-Holland Publishing Co., Amsterdam-Oxford (1975)

\bibitem{Cheeger-Gromoll} J. Cheeger and D. Gromoll,
``On the Structure of Complete Manifolds of Nonnegative Curvature'',
Ann. of Math. 96, p. 413-443 (1972)

\bibitem{Cheeger-Gromov-Taylor} J. Cheeger, M. Gromov and M. Taylor,
``Finite Propagation Speed, Kernel Estimates for Functions of the
Laplace Operator, and the Geometry of Complete Riemannian Manifolds'',
J. Diff. Geom. 17, p. 15-53 (1982)

\bibitem{Chen-Zhu} B. Chen and X. Zhu,
``Uniqueness of the Ricci Flow on Complete
Noncompact Manifolds'', J. of Diff. Geom. 74, p. 119-154
(2006)

\bibitem{Chow-Knopf} B. Chow and D. Knopf,
\underline{The Ricci Flow : An Introduction}, AMS Mathematical Surveys and
Monographs, Amer. Math. Soc., Providence (2004)

\bibitem{Chow-Lu-Ni} B. Chow, P. Lu and L. Ni, \underline{Hamilton's Ricci Flow},
Graduate Studies in Mathematics 77, AMS, Providence (2006)

\bibitem{Colding-Minicozzi} T. Colding and W. Minicozzi,
``Estimates for the Extinction Time for the Ricci Flow on Certain 3-Manifolds and a
Question of Perelman'',
J. Amer. Math. Soc. 18, p.  561-569 (2005)

\bibitem{Colding-Minicozzi2} T. Colding and W. Minicozzi,
``Width and Finite Extinction Time of Ricci Flow'',
preprint, http://arxiv.org/abs/0707.0108 (2007)

\bibitem{Dodziuk} J. Dodziuk,  ``Maximum Principle for Parabolic
Inequalities and the Heat Flow on Open Manifolds'',
Indiana Univ. Math. J. 32, p. 703-716 (1983)

\bibitem{Eells-Sampson} J. Eells and J. Sampson,
``Harmonic Mappings of Riemannian Manifolds'',
Amer. J. Math. 86, p. 109-160 (1964)

\bibitem{Eschenburg} J.-H. Eschenburg, ``Local Convexity and Nonnegative
Curvature---Gromov's Proof of the Sphere Theorem'',
Inv. Math. 84, p. 507-522 (1986)

\bibitem{Hamiltonnnn} R. Hamilton,
``Four-Manifolds with Positive Curvature Operator'',
J. Diff. Geom. 24, p. 153-179 (1986)

\bibitem{Hamsurface} R. Hamilton,
``The Ricci Flow on Surfaces'', in
\underline{Mathematics and General Relativity},
Contemp. Math. 71,
AMS, Providence, RI, p. 237-262 (1988).

\bibitem{harnack} R. Hamilton,
``The Harnack Estimate for the Ricci Flow'',
J. Diff. Geom. 37, p. 225-243 (1993)

\bibitem{Hamiltonnn} R. Hamilton,
``A Compactness Property for Solutions of the Ricci Flow''
Amer. J. Math. 117, p. 545-572 (1995)

\bibitem{Hamilton} R. Hamilton,
``The Formation of Singularities in the Ricci Flow'', in
\underline{Surveys in Differential Geometry}, Vol. II,
Internat. Press, Cambridge, p. 7-136 (1995)

\bibitem{Hamiltonn} R. Hamilton,
``Nonsingular Solutions of the Ricci Flow on Three-Manifolds'',
Comm. Anal. Geom. 7, p. 695-729 (1999)

\bibitem{Hamiltonnnnn} R. Hamilton,
``Three-Manifolds with Positive Ricci Curvature'',
J. Diff. Geom. 17, p. 255-306 (1982)

\bibitem{Hamiltonnnnnn} R. Hamilton,
``Four-Manifolds with Positive Isotropic Curvature'',
Comm. Anal. Geom. 5, p. 1-92 (1997)

\bibitem{Ishii} H. Ishii,
``On the Equivalence of Two Notions of Weak Solutions, Viscosity Solutions
and Distribution Solutions'',
Funkcial. Ekvac. 38, p. 101-120 (1995)

\bibitem{Kapovitch} V. Kapovitch, ``Perelman's Stability
Theorem,'' in \underline{Surveys in Differential Geometry},
vol. XI, Metric and Comparison Geometry, eds. J. Cheeger and
K. Grove, International Press, Somerville, MA, p. 103-136

\bibitem{KleinerLottwebsite}
B. Kleiner and J. Lott, webpage at
http://www.math.lsa.umich.edu/\~{}lott/ricciflow/perelman.html

\bibitem{Kleiner-Lott} B. Kleiner and J. Lott, ``Locally Collapsed
$3$-Manifolds'', to appear

\bibitem{Lu-Tian} P. Lu and G. Tian,
``Uniqueness of Standard Solutions in the Work of Perelman'',
preprint, http://www.math.lsa.umich.edu/\~{}lott/ricciflow/perelman.html  (2005)

\bibitem{Matveev}
S. Matveev,
\underline{Algorithmic Topology and Classification of 3-Manifolds},
Springer-Verlag, Berlin (2003)

\bibitem{meeksyau}
W. Meeks III and S.-T. Yau,
``Topology of Three-Dimensional Manifolds and the Embedding Problems in
Minimal Surface Theory'',
Ann. of Math. 112, p. 441-484 (1980)

\bibitem{Morgan} J. Morgan,
``Recent Progress on the Poincar\'e Conjecture and the Classification of $3$-Manifolds'',
Bull. Amer. Math. Soc. 42, p. 57-78 (2005)

\bibitem{Morgan-Tian} J. Morgan and G. Tian,
\underline{Ricci Flow and the Poincar\'e Conjecture},
Clay Mathematics Monographs 3, Amer. Math. Soc., Providence (2007)

\bibitem{Morgan-Tian2} J. Morgan and G. Tian,
``Completion of Perelman's Proof of the Geometrization Conjecture'',
to appear

\bibitem{Mostow} G. Mostow,
\underline{Strong Rigidity of Locally Symmetric Spaces},
Annals of Math. Studies No. 78,
Princeton University Press, Princeton (1973)

\bibitem{Ni3} L. Ni, ``The Entropy Formula for Linear Heat Equation'',
J. Geom. Anal. 14, p. 87-100 (2004)

\bibitem{Ni2} L. Ni, ``Ricci Flow and Nonnegativity of Sectional
Curvature'', Math. Res. Let. 11, p. 883-904 (2004)

\bibitem{Ni} L. Ni, ``A Note on Perelman's LYH Inequality'',
Comm. Anal. Geom. 14, p. 883-905 (2006)

\bibitem{Perelman}
G. Perelman,
``The Entropy Formula for the Ricci Flow and its Geometric
Applications'', http://arXiv.org/abs/math.DG/0211159

\bibitem{Perelman2}
G. Perelman,
``Ricci Flow with Surgery on Three-Manifolds'',
http://arxiv.org/abs/math.DG/0303109

\bibitem{Perelman3}
G. Perelman,
``Finite Extinction Time for the Solutions to the Ricci Flow
on Certain Three-Manifolds'',
http://arxiv.org/abs/math.DG/0307245


\bibitem{Prasad} G. Prasad, ``Strong Rigidity of $\Q$-Rank $1$ Lattices'',
Invent. Math. 21, p. 255-286 (1973)

\bibitem{Reed-Simon} M. Reed and B. Simon,
\underline{Methods of Modern Mathematical Physics IV, Analysis of
Operators}, Academic Press, New York (1978)

\bibitem{Rothaus (1981)} O. Rothaus,
``Logarithmic Sobolev Inequalities and the Spectrum of Schr\"odinger
Operators'', J. Funct. Anal. 42, p. 110-120 (1981)

\bibitem{schoen} R. Schoen,
``Estimates for Stable Minimal Surfaces in Three-Dimensional Manifolds'',
in \underline{Seminar on Minimal Submanifolds},
Ann. of Math. Stud. 103, Princeton Univ. Press, Princeton, N.J., p. 111-126 (1983).

\bibitem{schoenyau}
R. Schoen and S.-T. Yau,
``Complete Three-Dimensional Manifolds with Positive Ricci Curvature
  and Scalar Curvature'', in
\underline{Seminar on Differential Geometry}, Ann. of
  Math. Stud. 102,  Princeton Univ. Press, Princeton, N.J., p.  209-228 (1982)

\bibitem{Scott} P. Scott,
``The Geometries of $3$-Manifolds'',
Bull. London Math. Soc. 15, p. 401-487 (1983)

\bibitem{Sharafutdinov} V. Sharafutdinov,
``The Pogorelov-Klingenberg Theorem for Manifolds that are Homeomorphic to
$\R^n$'',
Siberian Math. J. 18, p. 649-657 (1977)

\bibitem{Shi (1989)} W.-X. Shi, ``Complete Noncompact Three-Manifolds
with Nonnegative Ricci Curvature'',
J. Diff. Geom. 29, p. 353-360 (1989)

\bibitem{Shi2 (1989)} W.-X. Shi, ``Deforming the Metric on Complete
Riemannian Manifolds'',
J. Diff. Geom. 30, p. 223-301 (1989)

\bibitem{Shi3 (1989)} W.-X. Shi, `` Ricci Deformation of the Metric on
Complete Noncompact Riemannian Manifolds'',
J. Diff Geom. 30, p. 303-394 (1989)

\bibitem{Shioya-Yamaguchi} T. Shioya and T. Yamaguchi,
``Volume-Collapsed Three-Manifolds with a Lower Curvature Bound'',
Math. Ann. 333, p. 131-155 (2005)

\bibitem{Thurston} W. Thurston,
``Three-Dimensional Manifolds, Kleinian Groups and Hyperbolic Geometry'',
Bull. Amer. Math. Soc. 6, p. 357-381 (1982)

\bibitem{Topping} P. Topping, \underline{Lectures on the Ricci Flow},
LMS Lecture Notes 325, London Mathematical Society and Cambridge University
Press,
http://www.maths.warwick.ac.uk/~topping/RFnotes.html

\bibitem{Ye1} R. Ye, ``On the $l$-Function and the Reduced Volume of
Perelman I,II'', Trans. Amer. Math. Soc. 360, p. 507-531, 533-544 (2008)

\bibitem{Ye2} R. Ye, ``On the Uniqueness of 2-Dimensional
$\kappa$-Solutions'',
http://www.math.ucsb.edu/\~{}yer/ricciflow.html
\end{thebibliography}
