\chapter{The Poincare Conjecture}

This blueprint is the repository's primary mathematical specification. It is
organized by the dependency structure of the proof rather than by the chapter
order of any one reference. The source-based projects under
\texttt{formalized-sources/} provide detailed supporting developments.

\begin{definition}[Poincare three-manifold]
\label{def:poincare-three-manifold}
A \emph{Poincare three-manifold} is a closed, connected topological
$3$-manifold with trivial fundamental group.
\end{definition}

\begin{theorem}[Poincare conjecture]
\label{thm:poincare-conjecture}
Every Poincare three-manifold is homeomorphic to the $3$-sphere.
\uses{def:poincare-three-manifold,thm:finite-time-extinction,thm:surgery-topology}
\end{theorem}

\chapter{Proof architecture}

The formal proof will separate the analytic control of Ricci flow from the
topological reconstruction after surgery. Each statement below is a milestone
whose internal dependencies will be refined as the custom formalization grows.

\begin{theorem}[Ricci flow with controlled surgery]
\label{thm:controlled-surgery}
A normalized initial metric on a closed $3$-manifold admits a Ricci flow with
surgery satisfying the required noncollapsing and canonical-neighborhood
estimates.
\end{theorem}

\begin{theorem}[Finite-time extinction]
\label{thm:finite-time-extinction}
For a Poincare three-manifold, the controlled Ricci flow with surgery becomes
extinct in finite time.
\uses{def:poincare-three-manifold,thm:controlled-surgery}
\end{theorem}

\begin{theorem}[Topology of the surgery decomposition]
\label{thm:surgery-topology}
The surgery history and finite-time extinction determine a connected-sum
decomposition into the standard topological pieces arising in the proof.
\uses{thm:controlled-surgery,thm:finite-time-extinction}
\end{theorem}
