\chapter{Bounded curvature at bounded distance}
\label{chap:bounded-curvature-bounded-distance}
This chapter is
devoted to Perelman's result about bounded curvature at bounded
distance for blow-up limits. Crucial to the argument is that each
member of the sequence of generalized Ricci flows has curvature
pinched toward positive and also has strong canonical neighborhoods.


\section{Pinching toward positive: the definitions}\label{10.1}

In this section we give the definition of what it means for a
generalized Ricci flow to have curvature pinched toward
positive. This is the
obvious generalization of the corresponding notion for Ricci flows.

\begin{definition}[Curvature pinched toward positive]
\label{def:curvature-pinched-toward-positive}
\label{pinchdef}
\uses{def:generalized-ricci-flow,def:curvature-operator}
Let $({\mathcal M},G)$ be a generalized three-dimensional Ricci flow
whose domain of definition is contained in $[0,\infty)$. For each
$x\in {\mathcal M}$, let $\nu(x)$ be the smallest eigenvalue of
$\Rm(x)$ on $\wedge^2T_xM_{{\bf t}(x)}$, as measured with
respect to a $G(x)$-orthonormal basis for the horizontal space at
$x$, and set $X(x)=\max(0,-\nu(x))$. We say that $({\mathcal
M},G)$ has curvature {\em pinched toward positive} if, for all $x\in
{\mathcal M}$, if the following two inequalities hold:
\begin{enumerate}
\item[(1)] $$R(x)\ge \frac{-6}{1+4{\bf t}(x)},$$
\item[(2)] $$R(x)\geq 2X(x)\left(\log X(x)+\log
(1+{\bf t}(x))-3\right),
$$ whenever
$0<X(x)$.
\end{enumerate}
\end{definition}

According to Theorem~\ref{pincha} if $(M,g(t)),\ 0\le a\le t<T$, is
Ricci flow with $M$  a compact three-manifold, and if the two
conditions given in the definition hold at the initial time $a$,
then they hold for all $t\in [a,T)$. In particular, if $a=0$ and if
$|\Rm(p,0)|\le 1$ for all $p\in M$, then the curvature of the
flow is pinched toward positive.

Next we fix $\epsilon_0>0$ sufficiently small such that for any
$0<\epsilon\le \epsilon_0$ all the results of the Appendix hold for
$2\epsilon$ and $\alpha=10^{-2}$, and Proposition~\ref{narrows}
holds for $2\epsilon$.


\section{The statement of the theorem}

Here is the statement of the main theorem of this chapter, the
theorem that establishes bounded curvature at bounded distance for
blow-up limits.

\begin{theorem}[Bounded curvature at bounded distance]
\label{thm:bounded-curvature-bounded-distance}
\label{bcbd}
\uses{def:curvature-pinched-toward-positive,def:strong-epsilon-neck-canonical-neighborhood,nocones}
Fix $0<\epsilon\le \epsilon_0$ and  $C<\infty$. Then for each
$A<\infty$ there are $D_0<\infty$ and $D<\infty$ depending on $A$,
$\epsilon$ and $C$ such that the following holds. Suppose that
$({\mathcal M},G)$ is a generalized three-dimensional Ricci flow
whose interval of definition is contained in $[0,\infty)$, and
suppose that $x\in {\mathcal M}$. Set $t={\bf t}(x)$. We suppose
that these data satisfy the following:
\begin{enumerate}
\item[(1)] $({\mathcal M},G)$ has curvature pinched toward positive.
\item[(2)] Every point $y\in{\mathcal  M}$ with
$R(y)\ge 4R(x)$  and ${\bf t}(y)\le t$ has a strong
$(C,\epsilon)$-canonical neighborhood.
\end{enumerate}
If $R(x)\ge D_0$, then $R(y)\le DR(x)$ for all $y\in
B(x,t,AR(x)^{-1/2})$.
\end{theorem}


This chapter is devoted to the proof of this theorem. The proof is
by contradiction. Suppose that there is some $A_0<\infty$ for which
the result fails. Then there are a sequence of generalized
three-dimensional Ricci flows $({\mathcal M}_n,G_n)$  whose
intervals of definition are contained in $[0,\infty)$ and whose
curvatures are pinched toward positive. Also, there are points
$x_n\in {\mathcal M}_n$ satisfying the second condition given in the
theorem and points $y_n\in {\mathcal M}_n$ such that for all $n$ we
have:
\begin{enumerate}
\item[(1)] $\lim_{n\rightarrow\infty}R(x_n)=\infty$.
\item[(2)] ${\bf t}(y_n)={\bf t}(x_n)$,
\item[(3)] $d(x_n,y_n)< A_0R(x_n)^{-1/2}$,
\item[(4)] $$\lim_{n\rightarrow\infty}\frac{R(y_n)}{R(x_n)}=\infty.$$
\end{enumerate}
For the rest of this chapter we assume that such a sequence of
generalized Ricci flows exists. We shall eventually derive a
contradiction.

Let us sketch how the argument goes. We show that there is a
(partial) geometric blow-up limit of the sequence $({\mathcal
M}_n,G_n)$ based at the $x_n$. We shall see that the following hold
for this limit. It is an incomplete manifold $U_\infty$
diffeomorphic to $S^2\times (0,1)$ with the property that the
diameter of $U_\infty$ is finite and the curvature goes to infinity
at one end of $U_\infty$, an end denoted ${\mathcal E}$, while
remaining bounded at the other end. (The non-compact manifold in
question is diffeomorphic to $S^2\times (0,1)$ and, consequently, it
has two ends.)
 Every point of $U_\infty$
sufficiently close to ${\mathcal E}$ is the center of a
$2\epsilon$-neck in $U_\infty$. In fact, there is a partial
geometric limiting flow on $U_\infty$ so that these points are
centers of evolving $2\epsilon$-necks. Having constructed this
incomplete blow-up limit of the original sequence we then consider
further blow-up limits about the end ${\mathcal E}$, the end where
the scalar curvature goes to infinity. On the one hand, a direct
argument shows that a sequence of rescalings of $U_\infty$ around
points converging to the end ${\mathcal E}$ converge in the
Gromov-Hausdorff sense to a cone. On the other hand, a slightly
different sequence of rescalings at the same points converges
geometrically to a limiting non-flat Ricci flow. Since both limits
are non-degenerate three-dimensional spaces, we show that the ratio
of the rescaling factors used to construct them converges to a
finite, non-zero limit. This means that the two limits differ only
by an overall constant factor. That is to say the geometric blow-up
limit is isometric to an open subset of a non-flat cone. This
contradicts Hamilton's result (Theorem~\ref{nocones}) which says
that it is not possible to flow under the Ricci flow to an open
subset of a non-flat cone. Now we carry out all the steps in this
argument.

\section{The incomplete geometric limit}\label{sectincomp}

We fix a sequence $({\mathcal M}_n,G_n,x_n)$ of generalized Ricci
flows as above. The first step  is to shift and rescale this
sequence of generalized Ricci flows so that we can form an
(incomplete) geometric limit which will be a tube of finite length
with scalar curvature going to infinity at one end.

We shift the time parameter of $({\mathcal M}_n,G_n)$ by $-{\bf
t}(x_n)$. We change notation and denote these shifted flows by
$({\mathcal M}_n,G_n)$. This allows us to arrange that ${\bf
t}(x_n)=0$ for all $n$. Since shifting leaves the curvature
unchanged, the shifted flows  satisfy a weaker version of curvature
pinched toward positive. Namely, for the shifted flows we have
\begin{eqnarray}\label{weakerpos}
R(x) & \ge & -6 \nonumber \\
R(x) & \ge & 2X(x)\left(\log(X(x))-3\right).
\end{eqnarray}

 We set $Q_n=R(x_n)$, and we denote by $M_n$ the $0$-time-slice of
${\mathcal M}_n$. We rescale $({\mathcal M}_n,G_n)$ by $Q_n$. Denote
by $({\mathcal M}'_n,G_n')$ the rescaled (and shifted) generalized
flows. For the rest of this argument we implicity use the metrics
$G'_n$. If we are referring to $G_n$ we mention it explicitly.

\subsection{The sequence of tubes}

 Let
$\gamma_n$ be a smooth path in $B_{G_n}(x_n,0,A_0Q_n^{-1/2})$ from
$x_n$ to $y_n$. For all $n$ sufficiently large we have
$R_{G'_n}(y_n) \gg 1$. Thus, there is a point $z_n\in \gamma_n$ such
that $R_{G'_n}(z_n)=4$ and such that on the sub-path
$\gamma_n|_{[z_n,y_n]}$ we have $R_{G'_n}\ge 4$. We replace
$\gamma_n$ by this sub-path. Now, with this replacement, according
to the second condition in the statement of the theorem, every point
of $\gamma_n$ has a strong $(C,\epsilon)$ canonical
neighborhood. As $n$ tends to infinity
the ratio of $R(y_n)/R(z_n)$ tends to infinity. This means that for
all $n$ sufficiently large, no point of  $\gamma_n$ can be contained
in an $\epsilon$-round component
or a $C$-component, because if it were then all
of $\gamma_n$ would be contained in that component, contradicting
the fact that the curvature ratio is arbitrarily large for large
$n$. Hence, for $n$ sufficiently large, every point of $\gamma_n$ is
either contained in the core of a
$(C,\epsilon)$-cap or is the center of a
strong $\epsilon$-neck. According to
Proposition~\ref{Xcontainedin}, for all $n$ sufficiently large
$\gamma_n$ is contained an open submanifold $X_n$ of the zero
time-slice of ${\mathcal M}'_n$ that is one of the following:
\begin{enumerate}
\item[(1)] an $\epsilon$-tube  and both endpoints of $\gamma_n$ are centers
of $\epsilon$-necks
contained in $X_n$,
\item[(2)] a $C$-capped $\epsilon$-tube with cap ${\mathcal C}$,
and each endpoint of $\gamma_n$ either is contained in the core $Y$
of ${\mathcal C}$ or is the center of an $\epsilon$-neck contained
in $X_n$,
\item[(3)] a double $C$-capped $\epsilon$-tube, or finally
\item[(4)] the union of two $(C,\epsilon)$-caps.
\end{enumerate}

The fourth possibility is incompatible with the fact that the ratio
of the curvatures at the endpoints of $\gamma_n$ grows arbitrarily
large as $n$ tends to infinity. Hence, this fourth possibility
cannot occur for $n$ sufficiently large. Thus, for all $n$
sufficiently large $X_n$ is one of the first three types listed
above.

\begin{claim}[Existence of a minimizing geodesic in $X_n$]
\label{claim:minimizing-geodesic-in-Xn}
There is a geodesic $\hat\gamma_n$ in $X_n$ with endpoints $z_n$ and
$y_n$. This geodesic is minimizing among all paths in $X_n$ from
$z_n$ to $y_n$.
\end{claim}

\begin{proof}
This is clear in the third case since $X_n$ is a closed manifold.

Let us consider the first case. There are $\epsilon$-necks $N(z_n)$
and $N(y_n)$ centered at $z_n$ and $y_n$ and contained in $X_n$.
Suppose first that the central $2$-spheres $S(z_n)$ and $S(y_n)$ of
these necks are disjoint. Then they are the boundary of a compact
sumanifold $X_n'$ of $X_n$. It follows easily from
Lemma~\ref{curveshort} that any sequence of minimizing paths from
$z_n$ to $y_n$ is contained in the union of $X_n'$ with the  middle
halves of $N(z_n)$ and $N(y_n)$. Since this manifold has compact
closure in $X_n$, the usual arguments show that one can extract a
limit of a subsequence which is a minimizing geodesic in $X_n$ from
$z_n$ to $y_n$. If $S(z_n)\cap S(y_n)\not=\emptyset$, then $y_n$ is
contained in the middle half of $N(z_n)$, and again it follows
immediately from Lemma~\ref{curveshort} that there is a minimizing
geodesic in $N(z_n)$ between these points.

Now let us consider the second case. If each of $z_n$ and $y_n$ is
the center of an $\epsilon$-neck in $X_n$, the argument as in the
first case applies. If both points are contained in the core of
${\mathcal C}$ then, since that core has compact closure in $X_n$,
the result is again immediate. Lastly, suppose that one of the
points, we can assume by the symmetry of the roles of the points
that it is $z_n$, is the center of an $\epsilon$-neck $N(z_n)$ in
$X_n$ and the other is contained in the core of ${\mathcal C}$.
Suppose that the central $2$-sphere $S(z_n)$ of $N(z_n)$  meets the
core $Y$ of ${\mathcal C}$. Then $z_n$ lies in the  half of the neck
$N={\mathcal C}\setminus \overline Y$ whose closure contains the
frontier of $Y$. Orient $s_N$ so that this half is the positive
half. Thus, by Lemma~\ref{curveshort} any minimizing sequence of
paths from $z_n$ to $y_n$ is eventually contained in the union of
the core of ${\mathcal C}$ and the the positive three-quarters of
this neck. Hence, as before we can pass to a limit and construct  a
minimizing geodesic in $X_n$ connecting $z_n$ to $y_n$. On the other
hand, if $S(z_n)$ is disjoint from $Y$, then $S(z_n)$ separates
$X_n$ into a compact complementary component and a non-compact
complementary component and the compact complementary component
contains $Y$. Orient the $s_N$-direction so that the compact
complementary component lies on the positive side of $S(z_n)$. Then
any minimizing sequence of paths in $X_n$ from $z_n$ to $y_n$ is
eventually contained in the union of the compact complementary
component of $N(z_n)$ and the positive $3/4$'s of $N(z_n)$. As
before, this allows us to pass to a limit to obtain a minimizing
geodesic in $X_n$.
\end{proof}

This claim allows us to assume (as we now shall) that $\gamma_n$ is
a minimizing geodesic in $X_n$ from $z_n$ to $y_n$.


\begin{claim}[Balanced chain of necks along a sub-geodesic]
\label{claim:balanced-chain-subgeodesic}
For every $n$ sufficiently large, there is a sub-geodesic
$\gamma'_n$ of $\gamma_n$ with end points $z_n'$ and $y_n'$ such
that the following hold:
\begin{enumerate}
\item[(1)] The length of $\gamma'_n$ is bounded independent of $n$.
\item[(2)] $R(z_n')$ is bounded independent of $n$.
\item[(3)] $R(y_n')$ tends to infinity as $n$ tends to infinity.
\item[(4)] $\gamma_n'$ is contained in a strong $\epsilon$-tube
$T_n$ that is the union of a balanced
chain of strong
$\epsilon$-necks centered at points of $\gamma'_n$. The first
element in this chain is a strong $\epsilon$-neck $N(z'_n)$ centered
at $z'_n$. The last element is a strong $\epsilon$-neck containing
$y'_n$.
\item[(5)] For every  $x\in T_n$, we have $R(x)>3$ and $x$ is the center of a strong
$\epsilon$-neck in the flow $({\mathcal M}'_n,G'_n)$.
\end{enumerate}
\end{claim}

\begin{proof}
The first item is clear since, for all $n$, the geodesic $\gamma_n$
has $G_n$-length at most $A_0Q_n^{-1/2}$ and hence $G'_n$-length at
most $A_0$. Suppose that we have a $(C,\epsilon)$-cap ${\mathcal C}$
whose core $Y$ contains a point of $\gamma_n$. Let $N$ be the
$\epsilon$-neck that is the complement of the closure of $Y$ in
${\mathcal C}$, and let $\widehat Y$ be the union of $Y$ and the
closed negative half of $N$. We claim that $\widehat Y$ contains
either $z_n$ or $y_n$. By Corollary~\ref{coreint}, since $Y$
contains a point of $\gamma_n$, the intersection of $\widehat Y$
with $\gamma_n$ is a subinterval containing one of the end points of
$\gamma_n$, i.e., either $z_n$ or $y_n$. This means that any point
$w$  which is contained in a $(C,\epsilon)$-cap whose core contains
a point of $\gamma_n$ must satisfy one of the following:
$$R(w)< CR(z'_n)\ \ \ \text{or}\ \ \ R(w)> C^{-1}R(y'_n).$$

We pass to a subsequence so that $R(y_n)/R(z_n)>4C^2$ for all $n$,
and we pass to a subinterval $\gamma'_n$ of $\gamma_n$ with
endpoints $z_n'$ and $y_n'$ such that:
\begin{enumerate}
\item[(1)] $R(z'_n)=2CR(z_n)$
\item[(2)] $R(y'_n)=(2C)^{-1}R(y_n)$
\item[(3)] $R(z'_n)\le R(w)\le R(y'_n)$ for all $w\in \gamma'_n$.
\end{enumerate}
Clearly, with these choices $R(z'_n)$ is bounded independent of $n$
and $R(y'_n)$ tends to infinity as $n$ tends to infinity. Also, no
point of $\gamma'_n$ is contained in the core of a
$(C,\epsilon)$-cap. Since every point of $\gamma'_n$ has a strong
$(C,\epsilon)$-canonical neighborhood, it follows that every point
of $\gamma'_n$ is the center of a strong $\epsilon$-neck. It now
follows from Proposition~\ref{epschains} that there is a balanced
$\epsilon$-chain consisting of strong $\epsilon$-necks centered at
points of $\gamma'_n$ whose union contains $\gamma'_n$. (Even if the
$2$-spheres of these necks do not separate the zero time-slice of
${\mathcal M}'_n$, as we build the balanced $\epsilon$-chain as
described in Proposition~\ref{epschains} the new necks we add can
not meet the negative end of $N(z'_n)$ since the geodesic
$\gamma'_n$ is minimal.) We can take the first element in the
balanced chain to be a strong $\epsilon$-neck $N(z'_n)$ centered at
$z'_n$, and the last element to be a strong $\epsilon$-neck $N^+_n$
containing $y'_n$. The union of this chain is $T_n$.

Next, we show that every point of $T_n$ is the center of a strong
$\epsilon$-neck in $({\mathcal M}_n,G_n)$. We must rule out the
possibility that there is a
 point of $T_n$ that is contained in the core of a $(C,\epsilon)$-cap.
Since $T_n$ is a union of $\epsilon$-necks centered at points of
$\gamma_n'$ we see that every point  $w\in T_n$ has
$$(3C/2)R(z_n)<R(w)<(2/3C)R(y_n).$$
This tells us that no point of $T_n$ is contained in a
$(C,\epsilon)$-cap whose core contains a point of $\gamma_n$. Thus,
to complete the argument we need only see that if there is a point
of $T_n$ contained in the core of a $(C,\epsilon)$-cap then the core
of that $(C,\epsilon)$-cap also contains a point of $\gamma_n$.
 The
scalar curvature inequality implies that both $z_n$ and $y_n$ are
outside $T_n$. This means that $\gamma_n$ traverses $T_n$ from one
end to the other. Let $w_-$, resp. $w_+$, be the point of $\gamma_n$
that lies in the frontier of $T_n$ contained in the closure of the
$N(z_n')$, resp. $N^+_n$. Since the scalar curvatures at these two
points of $\gamma$ satisfy the weak version of the above
inequalities, we see that there are strong $\epsilon$-necks $N(w_-)$
and $N(w_+)$ centered at them. Let $\widehat T_n$ be the union of
$T_n$, $N(w_-)$ and $N(w_+)$. It is also a strong $\epsilon$-tube,
and every point $\hat w$ of $\widehat T_n$ satisfies
$$(1.1)CR(z_n)<R(\hat w)<(0.9)C^{-1}R(y_n).$$
Thus, $z_n$ and $y_n$ are disjoint from $\widehat T_n$ and hence
$\gamma$ crosses $\widehat T_n$ from one end to the other.

 Now
suppose that $T_n$ meets the core $Y$ of a $(C,\epsilon)$-cap
${\mathcal C}$.
 Consider the boundary $S$ of the closure of $Y$.
If it is disjoint from $T_n$ then $T_n$ is contained in the core
$Y$. For large $n$ this is inconsistent with the fact that the ratio
of the scalar curvature at the endpoints of $\gamma'_n$ goes to
infinity. Thus, we are left to consider the case when
 $S$ contains a point of the tube $T_n$. In this case $S$ is
completely contained in $\widehat T_n$ and by
Corollary~\ref{s2isotopic} $S$ is isotopic to the $2$-spheres of the
product decomposition of $\widehat T_n$. Hence, $S$ meets a point of
$\gamma_n$ and consequently the core $Y$ contains a point of
$\gamma_n$. But we have already seen that this is not possible.

Lastly, we must show that $R(x)>3$ for every $x\in T_n$. We have
just seen that every $x\in T_n$ is the center of an $\epsilon$-neck.
If $x$ is contained in the $\epsilon$-neck centered at $z'_n$ or
$y'_n$, then since $R(z'_n)\ge 4$ and $R(y'_n)\ge 4$, clearly
$R(x)>3$. We must consider the case when $x$ is not contained in
either of these $\epsilon$-necks. In this case the central
$2$-sphere $S_x$of the $\epsilon$-neck centered at $x$ is contained
in the compact submanifold of $T_n'$ bounded by the central
$2$-spheres of the necks centered at $z'_n$ and $y'_n$. These
$2$-spheres are disjoint and by Condition 4 in
Proposition~\ref{S2intersection} each is a homotopically non-trivial
$2$-sphere in $T_n'$. Hence, the compact manifold with their
disjoint union as boundary is diffeomorphic to $S^2\times [0,1]$
and, again according to Condition 4 of
Proposition~\ref{S2intersection}, $S_x$ is isotopic to the
$2$-sphere factor in this product decomposition. Since the
intersection of $\gamma'_n$ with this submanifold is an arc spanning
from one boundary component to the other, $S_x$ must meet
$\gamma'_n$, in say $w$.
 By construction, since $w\in \gamma'_n$ we have $R(w)\ge 4$.
  This implies that $R(x)>3$.
 This completes the proof of the claim.
\end{proof}


\subsection{Extracting a limit of a subsequence of the tubes}

Passing to a subsequence we arrange that the $R(z_n')$ converge. Now
consider the subset ${\mathcal A}\subset \Ar$ consisting of all
$A>0$ such that there is a uniform bound, independent of $n$, for
the curvature on $B(z'_n,A)\cap T_n$. The set ${\mathcal A}$ is
non-empty since $R(z'_n)$ is bounded independent of $n$ and for
every $n$ there is a strong $\epsilon$-neck $N(z'_n)$ centered at
$z'_n$ contained in $T_n$. On the other hand, since
$d_{G_n'}(z_n',y_n')$ is uniformly bounded and $R(y_n')\rightarrow
\infty$,  there is a finite upper bound for ${\mathcal A}$. Let
$A_1$ be the least upper bound of ${\mathcal A}$. We set
$U_n=T_n\cap B(z'_n,A_1)$. This is an open subset of $T_n$
containing $z'_n$. We let $g_n'=G_n'|U_n$.

\begin{claim}[$U_n$ contains the neck centered at $z_n'$]
\label{claim:neck-contained-in-Un}
For all $n$ sufficiently large,  $3R(z'_n)^{-1/2}\epsilon^{-1}/2$ is
less than $A_1$, and hence $U_n$ contains the strong $\epsilon$-neck
$N(z_n')$ centered at $z'_n$.
\end{claim}

\begin{proof}
The curvature on  $N(z'_n)$ is bounded independent of $n$. Consider
a point $w$ near the end of $N(z'_n)$ that separates $y_n'$ from
$z_n'$. It is also the center of a strong $\epsilon$-neck $N(w)$. By
Proposition~\ref{S2intersection} and our assumption that
$\epsilon\le \bar\epsilon(10^{-2})$, the scalar curvature on
$N(z'_n)\cup N(w)$ is between $(0.9)R(z'_n)$ and $(1.1)R(z'_n)$.
Since, by construction, the negative end of $N(z'_n)$ contains an
end of $T_n$, this implies that
$$N(z'_n)\cup N(w)\supset B(z'_n,7R(z'_n)^{-1/2}\epsilon^{-1}/4)\cap T_n,$$
 so that we see that
$A_1\ge 7\epsilon^{-1}\lim_{n\rightarrow\infty}
R(z'_n)^{-1/2}/4$. Thus, $A_1>3R(z'_n)^{-1}\epsilon^{-1}/2$ for all
$n$ sufficiently large. Obviously then $U_n$ contains $N(z'_n)$.
\end{proof}

The next claim uses terminology from Definition~\ref{deltareg}.

\begin{claim}[Uniform curvature bound on $\operatorname{Reg}_\delta(U_n,g_n')$]
\label{claim:uniform-curvature-bound-reg-delta}
For any $\delta>0$ there is a uniform bound, independent of $n$, for
the curvature on $\operatorname{Reg}_\delta(U_n,g'_n)$.
\end{claim}

\begin{proof}
To prove this it suffices to show that given $\delta>0$ there is
$A<A_1$ such that $\operatorname{Reg}_\delta(U_n,g'_n)\subset B(z'_n,A)$ for
all $n$ sufficiently large. Of course, if we establish this for
every $\delta>0$ sufficiently small, then it follows for all
$\delta>0$. First of all, by Corollary~\ref{neckgeo} and
Lemma~\ref{neckcurv}, the fact that $\epsilon\le
\bar\epsilon(10^{-2})$ implies that any point $w$ with the property
that the strong $\epsilon$-neighborhood centered at $w$ contains
$z_n'$ is contained in the ball of radius
$(1.1)R(z'_n)^{-1/2}\epsilon^{-1}<A_1$ centered at $z_n'$. Thus, it
suffices to consider points $w_n$ in $\operatorname{Reg}_\delta(U_n,g'_n)$
with the property that the strong $\epsilon$-neck centered at $w_n$
does not contain $z_n'$. Fix such a $w_n$. Take a path $\mu_n(s)$
starting at $w_n$ moving in the $s$-direction at unit speed measured
in the $s$-coordinate of the $\epsilon$-neck centered at $w_n$ away
from $z'_n$ and ending at the frontier of this neck. Let $u_1$ be
the final point of this path. The rescaled version of
Lemma~\ref{neckdistcomp} implies that the forward difference
quotient for the distance from $z'_n$ satisfies
$$(0.99)R(w_n)^{-1/2}\le \frac{d}{ds}d(z'_n,\mu_n(s))
\le (1.01)R(w_n)^{-1/2}.$$ Of course, since we are working in an
$\epsilon$-neck we also have
$$(1-\epsilon)R(w_n)^{-1/2}\le \frac{d(d(w_n,\mu_n(s)))}{ds}\le
(1+\epsilon)R(w_n)^{-1/2}.$$  We continue the path $\mu_n$ moving in
the $s$-direction of a neck centered at $u_1$. Applying
Lemma~\ref{neckdistcomp} again both to the distance from $w_n$ and
the distance from $z'_n$ yields:
$$(0.99)R(u_1)^{-1/2}\le \frac{d(d(z'_n,\mu_n(s)))}
{ds}\le (1.01)R(u_1)^{-1/2}$$
$$(0.99)R(u_1)^{-1/2}\le \frac{d(d(w_n,\mu_n(s)))}
{ds}\le (1.01)R(u_1)^{-1/2}$$ on this part of the path $\mu_n$. We
repeat this process as many times as necessary until we reach a
point $w'_n\in U_n$ at distance $\delta/2$ from $w_n$. This is
possible since the ball of radius $\delta$ centered at $w_n$ is
contained in $U_n$. By the difference quotient inequalities, it
follows that $d(z'_n,w'_n)-d(z'_n,w_n)>\delta/4$.
 Since $w_n'\in U_n$ and
consequently that $d(z_n',w_n')<A_1$. It follows that
$d(z_n',w_n)\le A_1-\delta/4$. This proves that, for all $n$
sufficiently large, $\operatorname{Reg}_\delta(U_n,g'_n)\subset
B(z_n',A_1-\delta/4)$, and consequently that the curvature on
$\operatorname{Reg}_\delta(U_n,g'_n)$ is bounded independent of $n$.
\end{proof}

By Shi's theorem (Theorem~\ref{thm:shi-derivative-estimates}), the fact that each point of
$U_n$ is the center of a strong $\epsilon$-neck means that there is
a bound, independent of $n$, on all covariant derivatives of the
curvature at any point of $U_n$ in terms of the bound on the
curvature at the center point. In particular, because of the
previous result, we see that for any $\epsilon>0$ and any $\ell\ge
0$ there is a uniform bound for $|\nabla^\ell \Rm|$ on
$\operatorname{Reg}_\delta(U_n,g'_n)$. Clearly, since the base point $z_n'$ has
bounded curvature it lies in $\operatorname{Reg}_\delta(U_n,g'_n)$ for
sufficiently small $\delta$ (how small being independent of $n$).
Lastly, the fact that every point in $U_n$ is the center of an
$\epsilon$-neighborhood implies that $(U_n,g'_n)$ is $\kappa$
non-collapsed on scales $\le r_0$ where both $\kappa$ and $r_0$ are
universal. Since the $\gamma'_n$ have uniformly bounded lengths, the
$\epsilon$-tubes $T_n'$ have uniformly bounded diameter. Also, we
have seen that their have curvatures are bounded from below by $3$.
It follows that their volumes are uniformly bounded. Now invoking
Theorem~\ref{basicconv} we see that after passing to a subsequence
we have a geometric limit $(U_\infty,g_\infty,z_\infty)$ of a
subsequence of $(U_n,g'_n,z'_n)$.


\subsection{Properties of the limiting tube}


Now we come to a result establishing all the properties we need for
the limiting manifold.


\begin{proposition}[Properties of the incomplete geometric limit tube]
\label{prop:incomplete-limit-tube-properties}
\label{Uinfty}
\uses{def:generalized-ricci-flow,def:evolving-epsilon-neck,def:epsilon-neck,claim:balanced-chain-subgeodesic,claim:uniform-curvature-bound-reg-delta,basicconv,thm:shi-derivative-estimates,smoothconv,canonvary,Xcontainedin,partialflowlimit,blowupposcurv}
The geometric limit $(U_\infty,g_\infty,z_\infty)$ is an incomplete
Riemannian $3$-manifold of finite diameter. There is a
diffeomorphism $\psi\colon U_\infty\to S^2\times (0,1)$. There is a
$2\epsilon$-neck centered at $z_\infty$ whose central $2$-sphere
$S^2(z_\infty)$ maps under $\psi$ to a $2$-sphere isotopic to a
$2$-sphere  factor in the product decomposition. The scalar
curvature is bounded at one end of $U_\infty$ but tends to infinity
at the other end, the latter end  which is denoted ${\mathcal E}$.
Let ${\mathcal U}_\infty\subset U_\infty\times(-\infty,0]$ be the
open subset consisting of all $(x,t)$ for which $-R(x)^{-1}<t\le 0$.
We have a generalized Ricci flow on ${\mathcal U}_\infty$ which is a
partial geometric limit of a subsequence of the generalized Ricci
flows $({\mathcal M}'_n,G'_n,z'_n)$. In particular, the zero-time
slice of the limit flow is $(U_\infty,g_\infty)$. The Riemannian
curvature is non-negative at all points of the limiting smooth flow
on ${\mathcal U}_\infty$. Every point $x\in U_\infty\times\{0\}$
which is not separated from ${\mathcal E}$ by $S^2(z_\infty)$ is the
center of an evolving $2\epsilon$-neck $N(x)$ defined for an
interval of normalized time of length $1/2$. Furthermore, the
central $2$-sphere of $N(x)$ is isotopic to the $2$-sphere factor of
$U_\infty$ under the diffeomorphism $\psi$.
\end{proposition}


The proof of this proposition occupies the rest of
Chapter~\ref{sectincomp}.

\begin{proof}
Let $V_1\subset V_2 \subset \cdots \subset U_\infty$ be the open
subsets and $\varphi_n\colon V_n\to U_n$ be the maps having all the
properties stated in Definition~\ref{smoothconv} so as to exhibit
$(U_\infty,g_\infty,z_\infty)$ as the  geometric limit of the
$(U_n,g_n',z_n')$.

Since the $U_n$ are all contained in $B(z_n',A_1)$, it follows that any point
of $U_\infty$ is within $A_1$ of the limiting base point $z_\infty$. This
proves that the diameter of $U_\infty$ is bounded.




For each $n$ there is the  $\epsilon$-neck $N(z_n')$ centered at
$z_n'$ contained in $U_n$. The middle two-thirds, $N_n'$, of this
neck has closure contained in $\operatorname{Reg}_\delta(U_n,g_n)$ for some
$\delta>0$ independent of $n$ (in fact, restricting to $n$
sufficiently large, $\delta$ can be taken to be approximately equal
to $R(z_\infty)^{-1/2}\epsilon^{-1}/3$). This means that for some
$n$ sufficiently large and for all $m\ge n$ the image
$\varphi_m(V_n)\subset U_m$ contains $N'_m$. For any fixed $n$ as
$m$ tends to infinity  the metrics $\varphi_m^*g_m|_{V_n}$ converge
uniformly in the $C^\infty$-topology to $g_\infty|_{V_n}$. Thus, it
follows from Proposition~\ref{canonvary} that for all $m$
sufficiently large, $\varphi^{-1}_m(N'_m)$ is a $3\epsilon/2$-neck
centered at $z_\infty$. We fix such a neck $N'(z_\infty)\subset
U_\infty$. Let $S(z_\infty)$ be the central $2$-sphere of
$N'(z_\infty)$. For each $n$ sufficiently large,
$\varphi_n(S(z_\infty))$ separates $U_n$ into two components, one,
say $W^-_n$ contained in $N(z'_n)$ and the other, $W^+_n$ containing
all of $U_n\setminus N(z'_n)$. It follows that $S(z_\infty)$
separates $U_\infty$ into two components, one, denoted $W^-_\infty$,
where the curvature is bounded (and where, in fact, the curvature is
close to $R(z_\infty)$) and the other, denoted $W^+_\infty$, where
it is unbounded.

\begin{claim}[Points of $W^+_\infty$ are neck centers]
\label{claim:points-in-Wplus-are-neck-centers}
Any point $q\in W^+_\infty$ is the center of a $2\epsilon$-neck in
$U_\infty$.
\end{claim}

\begin{proof}
Fix a point $q\in W^+_\infty$.
 For all $n$ sufficiently large denote by $q_n=\varphi_n(q)$.
 Then for all $n$ sufficiently large, $q_n\in W^+_n$ and
$\lim_{n\rightarrow\infty}R(q_n)=R(q)$. This means that for all
$n$ sufficiently large $R(y_n')>>R(q_n))$, and hence
 the $3\epsilon/2$-neck centered at
 $q_n\in U_n$ is disjoint from $N(y'_n)$. Thus, by the rescaled version of
 Corollary~\ref{neckgeo}, we see that  the distance from
 the $3\epsilon/2$-neck centered at $q_n$ to $N(y'_n)$ is bounded below by
 $(0.99)\epsilon^{-1}R(q_n)^{-1/2}/4\ge \epsilon^{-1}R(q_\infty)^{-1/2}/12$.
 Also, since $q_n\in W_n$, this $3\epsilon/2$-neck $N'(q_n)$ centered at
 $q_n$ does not extend past the $2$-sphere at $s^{-1}(-3\epsilon^{-1}/4)$ in the
 $ \epsilon$-neck $N(z'_n)$. It follows that for all $n$ sufficiently large that
 this $3\epsilon/2$-neck has
 compact closure contained in $\operatorname{Reg}_\delta(U_n,g_n)$ for some $\delta$
 independent of $n$, and hence there is $m$ such that for all $n$ sufficiently large $N'(q_n)$ is
 contained in the image $\varphi_n(V_m)$. Again using the fact that
 $\varphi_n^*(g_n|_{V_m})$ converges in the $C^\infty$-topology to
 $g_\infty|_{V_m}$ as $n$ tends to infinity, we see, by Proposition~\ref{canonvary}
 that for all $n$
 sufficiently large $\varphi_n^{-1}(N_m)$ contains a $2\epsilon$-neck in $U_\infty$ centered
 at $q$.
\end{proof}


It now follows from Proposition~\ref{Xcontainedin} that $W^+_\infty$
is contained in an $2\epsilon$-tube $T_\infty$ that is contained in
$U_\infty$. Furthermore, the frontier of $W^+_\infty$ in $T_\infty$
is the $2$-sphere $S(z_\infty)$ which is isotopic to the central
$2$-spheres of the $2\epsilon$-necks making up $T_\infty$. Hence,
the closure $\overline W^+_\infty$ of $W^+_\infty$ is a
$2\epsilon$-tube with boundary $S(z_\infty)$. In particular,
$\overline W^+_\infty$ is diffeomorphic to $S^2\times [0,1)$.

Now we consider the closure $\overline W^-_\infty$ of  $W^-_\infty$.
Since the closure of each $W^-_n$ is the closed negative half of the
$\epsilon$-neck $N(z_n')$ and the curvatures of the $z_n'$ have a
finite, positive limit, the limit $\overline W^-_\infty$ is
diffeomorphic to a product $S^2\times (-1,0]$. Hence, $U_\infty$ is
the union of  $\overline W^+_\infty$ and  $\overline W^-_\infty$
along their common boundary. It follows immediately that $U_\infty$
is diffeomorphic to $S^2\times (0,1)$.


\begin{claim}[Curvature bounded at one end, unbounded at the other]
\label{claim:curvature-bounded-one-end-infinite-other}
The curvature is bounded in a neighborhood of one end of $U_\infty$
and goes to infinity at the other end.
\end{claim}

\begin{proof}
 A neighborhood of one end of $U_\infty$, the end $\overline W^-_\infty$,
 is the limit of the
 negative halves of $\epsilon$-necks centered at $z_n'$.
 Thus, the curvature is bounded on this neighborhood, and in fact is
approximately equal to $R(z_\infty)$. Let $x_k$ be any sequence of
points in $U_\infty$ tending to the other end. We show that $R(x_k)$
tends to $\infty$ as $k$ does. The point is that since the sequence
is tending to the end, the distance from $x_k$ to the end of
$U_\infty$ is going to zero. Yet, each $x_k$ is the center of an
$\epsilon$-neck in $U_\infty$. The only way this is possible is if
the scales of these $\epsilon$-necks are converging to zero as $k$
goes to infinity. This is equivalent to the statement that $R(x_k)$
tends to $\infty$ as $k$ goes to infinity.
\end{proof}


The next step in the proof of Proposition~\ref{prop:incomplete-limit-tube-properties} is to extend
the flow backwards a certain amount. As stated in the proposition,
the amount of backward time that we can extend the flow is not
uniform over all of $U_\infty$, but rather depends on the curvature
of the point at time zero.



\begin{claim}[Existence of a backward flowline]
\label{claim:flowline-exists-for-x}
For each $x\in U_n\subset M_n$ there is a flowline $\{x\}\times
(-R(x)^{-1},0]$ in ${\mathcal M}_n$. Furthermore, the scalar
curvature at any point of this flow line is less than or equal to
the scalar curvature at $x$.
\end{claim}


\begin{proof}
Since $x\in U_n\subset T_n$, there is a strong $\epsilon$-neck in
${\mathcal M}_n$ centered at $x$. Both statements follow immediately
from that.
\end{proof}


Let $X\subset U_\infty$ be an open submanifold with compact closure
and set
$$t_0(X)= \sup_{x\in X}(-R_{g_\infty}(x)^{-1}).$$ Then for
all $n$ sufficiently large $\varphi_n$ is defined on $X$ and the
scalar curvature of the flow $g_n(t)$ on $\varphi_n(X)\times
(t_0,0]$ is uniformly bounded independent of $n$. Thus, according to
Proposition~\ref{partialflowlimit} by passing to a subsequence we
can arrange that there is a limiting flow defined on $X\times
(t_0,0]$. Let ${\mathcal U}_\infty\subset U_\infty\times
(-\infty,0]$ consist of all pairs $(x,t)$ with the property that
$-R_{g_\infty}(x,0)^{-1}<t\le 0$.
 Cover ${\mathcal U}_\infty$ by countably many such boxes
of the type $X\times (-t_0(X),0]$ as described above, and take a
diagonal subsequence. This allows us to pass to a subsequence so
that the limiting flow exists (as a generalized Ricci flow) on
${\mathcal U}_\infty$.



\begin{claim}[Non-negativity of the limiting curvature]
\label{claim:limiting-flow-nonnegative-curvature}
 The curvature of the generalized Ricci flow on ${\mathcal
U}_\infty$ is non-negative.
\end{claim}

\begin{proof}
 This claim follows from the fact that the
original sequence $({\mathcal M}_n,G_n)$ consists of generalized
flows whose curvatures are pinched toward positive in the weak sense
given in Equation~\ref{weakerpos} and the fact that $Q_n\rightarrow
\infty$ as $n\rightarrow \infty$. (See Theorem~\ref{blowupposcurv}.)
\end{proof}

This completes the proof that all the properties claimed in
Proposition~\ref{prop:incomplete-limit-tube-properties} hold for the geometric limit
$(U_\infty,g_\infty,z_\infty)$. This  completes the proof of that
proposition. \end{proof}





\section{Cone limits near the end ${\mathcal E}$ for rescalings of $U_\infty$}\label{GromHaus}

The next step is to study the nature  of the limit $U_\infty$ given
in Proposition~\ref{prop:incomplete-limit-tube-properties}. We shall show that an appropriate
blow-up limit (limit in the Gromov-Hausdorff sense) around the end
is a cone.

Let $(X,d_X)$ be a metric space. Recall that the  cone on $X$,
denoted $C(X)$, is the quotient space $X\times [0,\infty)$ under the
identification $(x,0)\cong (y,0)$ for all $x,y\in X$. The image of
$X\times\{0\}$ is the {\sl cone point} of the cone. The metric on
$C(X)$ is given by
\begin{equation}\label{conemetric}
d((x,s_1),(y,s_2))=s_1^2+s_2^2-2s_1s_2\cos(\mathrm{min}(d_X(x,y),\pi)). \end{equation} The open cone $C'(X)$ is the
complement of the cone point in $C(X)$ with the induced metric.


The purpose of this section is to prove the following result.

\begin{proposition}[Rescaled limits converge to a cone at the end]
\label{prop:cone-limit-gromov-hausdorff}
\label{statement}
\uses{prop:incomplete-limit-tube-properties}
Let $(U_\infty,g_\infty,z_\infty)$  be as in the conclusion of
Proposition~\ref{prop:incomplete-limit-tube-properties}. Let $Q_\infty=R_{g_\infty}(z_\infty)$ and
let ${\mathcal E}$ be the end of $U_\infty$ where the scalar
curvature is unbounded. Let $\lambda_n$ be any sequence of positive
numbers with $\lim_{n\rightarrow\infty}\lambda_n=+\infty$. Then
there is a sequence $x_n$ in $U_\infty$ such that for each $n$ the
distance from $x_n$ to ${\mathcal E}$ is $\lambda_n^{-1/2}$, and
such that the pointed Riemannian manifolds
$(U_\infty,\lambda_ng_\infty,x_n)$ converge in the Gromov-Hausdorff
sense to an open cone, an open cone not homeomorphic to
an open ray (i.e., not homeomorphic to the open cone on a point).
\end{proposition}

The rest of this section is devoted to the proof of this result.

\subsection{Directions at ${\mathcal E}$}

 We orient the direction down the tube $U_\infty$ so that
${\mathcal E}$ is at the positive end. This gives an $s_N$-direction
for each $2\epsilon$-neck $N$ contained in $U_\infty$.




Fix a point $x\in U_\infty$. We say a ray $\gamma$ with endpoint $x$
limiting to ${\mathcal E}$ is {\sl a minimizing geodesic ray} if for
every $y\in \gamma$ the segment on $\gamma$ from $x$ to $y$ is a
minimizing geodesic segment; i.e., the length of this geodesic
segment is equal to $d(x,y)$.

\begin{claim}[Existence of minimizing rays to the end]
\label{claim:minimizing-ray-to-end-exists}
\uses{prop:incomplete-limit-tube-properties,genint}
There is a minimizing geodesic ray to ${\mathcal E}$ from each $x\in
U_\infty$ with $R(x)\ge 2Q_\infty$.
\end{claim}

\begin{proof}
Fix $x$ with $R(x)\ge 2Q_\infty$ and fix a $2\epsilon$-neck $N_x$
centered at $x$. Let $S_x^2$ be the central $2$-sphere of this neck.
Take a sequence of points $q_n$ tending to the end ${\mathcal E}$,
each being closer to the end than $x$ in the sense that $S^2_x$ does
not separate any $q_n$ from the end ${\mathcal E}$. We claim that
there is a minimizing geodesic from $x$ to each $q_n$. The reason is
that by Lemma~\ref{genint} any minimizing sequence of arcs from $x$
to $q_n$ cannot exit from the minus end of $N_x$ nor the plus end of
a $2\epsilon$-neck centered at $q_n$. Consider a sequence of paths
from $x$ to $q_n$ minimizing the distance.  Hence these paths all
lie in a fixed compact subset of $U_n$. After replacing the sequence
by a subsequence, we can pass to a limit, which is clearly a
minimizing geodesic from $x$ to $q_n$. Consider minimizing geodesics
$\mu_n$ from $x$ to $q_n$. The same argument shows that, after
passing to a subsequence, the $\mu_n$ converge to a minimizing
geodesic ray from $x$ to ${\mathcal E}$.
\end{proof}


\begin{claim}[Minimizing rays realize the distance to the end]
\label{claim:shortest-ray-equivalence}
(1) Any minimizing geodesic ray from $x$ to the end ${\mathcal E}$
is a shortest ray from $x$ to the end ${\mathcal E}$, and conversely
any shortest ray from $x$ to the end ${\mathcal E}$ is a minimizing
geodesic ray.

(2) The length of a shortest ray from $x$ to ${\mathcal E}$ is the
distance (see Section~\ref{ENDS}) from $x$ to ${\mathcal E}$.
\end{claim}


\begin{proof}
The implication in (1) in one direction is clear: If $\gamma$ is a
ray from $x$ to the end ${\mathcal E}$, and for some $y\in \gamma$
the segment on $\gamma$ from $x$ to $y$ is not minimizing, then
there is a shorter geodesic segment $\mu$ from $x$ to $y$. The union
of this together with the ray on $\gamma$ from  $y$ to the end is a
shorter ray from $x$ to the end.

Let us establish the opposite implication. Suppose that $\gamma$ is
a minimizing geodesic ray from $x$ to the end ${\mathcal E}$ and
that there is a $\delta>0$ and a shortest geodesic ray $\gamma'$
from $x$ to the end ${\mathcal E}$ with $|\gamma'|=|\gamma|-\delta$.
As we have just seen, $\gamma'$ is a minimizing geodesic ray. Take a
sequence of points $q_i$ tending to the end ${\mathcal E}$ and let
$S^2_i$ be the central $2$-sphere in the $2\epsilon$-neck centered
at $q_i$. Of course, for all $i$ sufficiently large, both $\gamma'$
and $\gamma$ must cross $S^2_i$. Since the scalar curvature tends to
infinity at the end ${\mathcal E}$, it follows from
Lemma~\ref{directions} for all $i$ sufficiently large, the extrinsic
diameter of $S^2_i$ is less than $\delta/3$. Let $p_i$ be a point of
intersection of $\gamma$ with $S^2_i$. For all $i$ sufficiently
large the length $d_i$ of the sub-ray in $\gamma$ from $p_i$ to the
end ${\mathcal E}$ is at most $\delta/3$. Let $p'_i$ be a point of
intersection of $\gamma'$ with $S^2_i$ and let $d'_i$ be the length
of the ray in $\gamma'$ from $p'_i$ to the end ${\mathcal E}$. Let
$\lambda$ be the sub-geodesic of $\gamma$ from $x$ to $p_i$ and
$\lambda'$ the sub-geodesic of $\gamma'$ from $x$ to $p'_i$. Let
$\beta$ be a minimizing geodesic from $p'_i$ to $p_i$. Of course,
$|\beta|<\delta/3$ so that by the minimality of $\lambda$ and
$\lambda'$ we have
$$-\delta/3<|\lambda|-|\lambda'|<\delta/3.$$ Since
$|\lambda'|+d'_i=|\lambda|+d_i-\delta$, we have
$$2\delta/3\le d_i-d'_i.$$
This is absurd since $d'_i>0$ and $d_i<\delta/3$.

(2) follows immediately from (1) and the definition.
\end{proof}

Given this result, the usual arguments show:

\begin{corollary}[Uniqueness of shortest continuations]
\label{cor:unique-shortest-subray}
\label{distlim}
\uses{claim:shortest-ray-equivalence}
If $\gamma$ is a minimizing geodesic ray from $x$ to the end
${\mathcal E}$, then for any $y\in \gamma\setminus\{x\}$ the sub-ray
of $\gamma$ from $y$ to the end, is the unique shortest geodesic
from $y$ to the end.
\end{corollary}

Also, we have a version of the triangle inequality for distances to
${\mathcal E}$.

\begin{lemma}[Triangle inequality for distance to the end]
\label{lem:triangle-inequality-distance-to-end}
Let $x$ and $y$ be points of $M$. Then the three distances $d(x,y)$,
$d(x,{\mathcal E})$ and $d(y,{\mathcal E})$ satisfy the triangle
inequality.
\end{lemma}


\begin{proof}
From the definitions it is clear that $d(x,y)+d(y,{\mathcal E})\ge
d(x,{\mathcal E})$, and symmetrically, reversing the roles of $x$
and $y$. The remaining inequality that we must establish is the
following: $d(x,{\mathcal E})+d(y,{\mathcal E})\ge d(x,y)$. Let
$q_n$ be any sequence of points converging to ${\mathcal E}$. Since
the end is at finite distance, it is clear that $d(x,{\mathcal E})=
\lim_{n\rightarrow\infty}d(x,q_n)$. The remaining inequality
follows from this and  the usual triangle inequality applied to
$d(x,q_n)$, $d(y,q_n)$ and $d(x,y)$.
\end{proof}


\begin{definition}[Direction at the end]
\label{def:direction-at-end}
\uses{cor:unique-shortest-subray}
We say that two minimizing geodesic rays limiting to ${\mathcal E}$
are {\em equivalent} if one is contained in the other. From the
unique continuation of geodesics it is easy to see that this
generates an equivalence relation. An equivalence class is a {\sl
direction at ${\mathcal E}$}, and the set of equivalence classes is
the {\sl set of directions at ${\mathcal E}$}.
\end{definition}

\begin{lemma}[Multiple directions at the end]
\label{lem:multiple-directions-at-end}
\label{morethanone}
\uses{def:direction-at-end}
There is more than one direction at ${\mathcal E}$.
\end{lemma}

\begin{proof}
Take a minimal geodesic ray $\gamma$ from a point $x$ limiting to
the end and let $y$ be a point closer to ${\mathcal E}$ than $x$ and
not lying on $\gamma$. Then a minimal geodesic ray from $y$ to
${\mathcal E}$ gives a direction at ${\mathcal E}$ distinct from the
direction determined by $\gamma$.
\end{proof}

\begin{remark}[Space of directions is homeomorphic to $S^2$]
\label{rem:space-of-directions-topology}
In fact, the general theory of positively curved spaces implies that
the space of directions is homeomorphic to $S^2$. Since we do not
need this stronger result we do not prove it.
\end{remark}

\subsection{The Metric on the space of directions at ${\mathcal
E}$}



\begin{definition}[Comparison angle between two directions]
\label{def:angle-between-directions-theta}
\label{defntheta}
\uses{def:direction-at-end}
Let $\gamma$ and $\mu$ be minimizing geodesic rays limiting to
${\mathcal E}$, of lengths $a$ and $b$, parameterized by the
distance from the end. For $0<s\le a$ and $0<s'\le b$ construct a
triangle $\alpha_se\beta_{s'}$ in the Euclidean plane with
$|\alpha_se|=s, |e\beta_{s'}|=s'$ and
$|\alpha_s\beta_{s'}|=d(\gamma(s),\mu(s'))$. We define
$\theta(\gamma,s,\mu,s')$ to be the angle at $e$ of the triangle
$\alpha_se\beta_{s'}$.
\end{definition}

\begin{lemma}[Monotone convergence of comparison angles]
\label{lem:theta-monotone-convergence}
\label{thetaconv}
\uses{def:angle-between-directions-theta,thm:length-comparison}
For all $\gamma,s,\mu,s'$ as in the previous definition we have
$$0\le \theta(\gamma,s,\mu,s')\le \pi.$$ Furthermore,  $\theta(\gamma,s,\mu,s')$
is a non-increasing function of $s$  when $\gamma,\mu,s'$ are held
fixed, and symmetrically it is a non-increasing function of $s'$
when $\gamma,s,\mu$ are held fixed. In particular, fixing $\gamma$
and $\mu$, the function $\theta(\gamma,s,\mu,s')$ is non-decreasing
as $s$ and $s'$ tend to zero. Thus, there is a well-defined limit as
$s$ and $s'$ go to zero, denoted $ \theta(\gamma,\mu)$. This limit
is greater than or equal to $\theta(\gamma,s,\mu,s')$ for all $s$
and $s'$ for which the latter is defined. We have
$0\le\theta(\gamma,\mu)\le \pi$. The angle $\theta(\gamma,\mu)=0$ if
and only if $\gamma$ and $\mu$ are equivalent. Furthermore, if
$\gamma$ is equivalent to $\gamma'$ and $\mu$ is equivalent to
$\mu'$, then $\theta(\gamma,\mu)=\theta(\gamma',\mu')$.
\end{lemma}


\begin{proof}
By restricting $\gamma$ and $\mu$ to slightly smaller rays, we can
assume that each is the unique shortest ray from its endpoint to the
end ${\mathcal E}$. Let $x$, resp., $y$ be the endpoint of $\gamma$,
resp., $\mu$. Now let $q_n$ be any sequence of points in $U_\infty$
limiting to the end ${\mathcal E}$, and consider minimizing geodesic
rays $\gamma_n$ from $q_n$ to $x$ and $\mu_n$ from $q_n$ to $y$,
each parameterized by the distance from $q_n$. By passing to a
subsequence we can assume that each of the sequences $\{\gamma_n\}$
and $\{\mu_n\}$ converge to a minimizing geodesic ray, which  by
uniqueness, implies that the first sequence limits to $\gamma$ and
the second to $\mu$. For $s,s'$ sufficiently small, let
$\theta_n(s,s')$ be the angle at $\tilde q_n$ of the Euclidean
triangle $\alpha_n\tilde q_n\beta_n$, where $|\alpha_n\tilde
q_n|=d(\gamma_n(s),q_n)$, $|\beta_n\tilde q_n|=d(\mu_n(s'),q_n)$ and
$|\alpha_n\beta_n|=d(\gamma_n(s),\mu_n(s'))$. Clearly, for fixed $s$
and $s'$ sufficiently small, $\theta_n(s,s')$ converges as
$n\rightarrow\infty$ to $\theta(\gamma,s,\mu,s')$. By the Toponogov
property (Theorem~\ref{thm:length-comparison}) for manifolds with
non-negative curvature, for each $n$ the function $\theta_n(s,s')$
is a non-increasing function of each variable, when the other is
held fixed. This property then passes to the limit, giving the first
statement in the lemma.

By the monotonicity, $\theta(\gamma,\mu)=0$ if and only if for all
$s,s'$ sufficiently small we have $\theta(\gamma,s,\mu,s')=0$, which
means one of $\gamma$ and $\mu$ is contained in the other.

It is obvious that the last statement holds.
\end{proof}


It follows that $\theta(\gamma,\mu)$ yields a well-defined function
on the set of pairs of directions at ${\mathcal E}$. It is clearly a
symmetric, non-negative function which is positive off of the
diagonal. The next lemma shows that it is a metric by establishing
the triangle inequality for $\theta$.

\begin{lemma}[Triangle inequality for the angle metric]
\label{lem:theta-triangle-inequality}
\uses{def:angle-between-directions-theta}
If $\gamma,\mu,\nu$ are minimizing geodesic rays limiting to
${\mathcal E}$, then
$$\theta(\gamma,\mu)+\theta(\mu,\nu)\ge \theta(\gamma,\nu).$$
\end{lemma}


\begin{proof}
\uses{cor:unique-shortest-subray,cor:angle-comparison}
By Corollary~\ref{cor:unique-shortest-subray}, after replacing
$\gamma, \mu,\nu$ by
equivalent, shorter geodesic arcs, we can assume that they are the
unique minimizing geodesics from their end points, say $x,y,z$
respectively, to ${\mathcal E}$. Let $q_n$ be a sequence of points
limiting to ${\mathcal E}$, and let $\gamma_n,\mu_n,\nu_n$ be
minimizing geodesics from $x,y,z$ to $q_n$. Denote by
$\theta_n(x,y), \theta_n(y,z)$, and $\theta_n(x,z)$, respectively,
the angles  at $\tilde q_n$ of the Euclidean triangles with the
following edge lengths: $\{d(x,y), d(x,q_n), d(y,q_n)\}$,
$\{d(y,z),d(y,q_n),d(z,q_n)\}$, and $\{d(z,x),d(z,q_n),d(x,q_n)\}$.
According to Corollary~\ref{cor:angle-comparison} we have
$\theta_n(x,y)+\theta_n(y,z)\ge \theta_n(x,z)$. Passing to the limit
as $n$ goes to $\infty$ and then the limit as $x$, $y$ and $z$ tend
to ${\mathcal E}$, gives the result.
\end{proof}


\begin{definition}[Cone of directions at the end]
\label{def:cone-of-directions}
\uses{def:angle-between-directions-theta,def:direction-at-end}
Let $X({\mathcal E})$ denote the set of directions at ${\mathcal
E}$. We define the metric on $X({\mathcal E})$ by setting
$d([\gamma],[\mu])=\theta(\gamma,\mu)$. We call this the {\sl
(metric) space of realized directions at} ${\mathcal E}$. The {\sl
metric space of directions at} ${\mathcal E}$ is the completion
$\bar X({\mathcal E})$ of $X({\mathcal E})$ with respect to the
given metric. We denote by $(C_{\mathcal E},g_{\mathcal E})$ the
cone on $\bar X({\mathcal E})$ with the cone metric as given in
Equation~(\ref{conemetric}).
\end{definition}


\begin{proposition}[The cone of directions is not a ray]
\label{prop:cone-not-a-ray}
\label{notray}
\uses{def:cone-of-directions,lem:multiple-directions-at-end}
$(C_{\mathcal E},g_{\mathcal E})$ is a metric cone that is not
homeomorphic to a ray.
\end{proposition}


\begin{proof}
\uses{lem:multiple-directions-at-end}
By construction $(C_{\mathcal E},g_{\mathcal E})$ is a metric cone.
That it is not homeomorphic to a ray follows immediately from
Lemma~\ref{lem:multiple-directions-at-end}.
\end{proof}


\subsection{Comparison results for distances}

\begin{lemma}[Cone distance bounds the limit distance]
\label{lem:cone-distance-upper-bound}
\label{distcomp}
\uses{def:cone-of-directions}
Suppose that $\gamma$ and $\mu$ are unique shortest geodesic rays
from points $x$ and $y$ to the end ${\mathcal E}$. Let $[\gamma]$
and $[\mu]$ be the points of $X({\mathcal E})$ represented by these
two geodesics rays. Let $a$, resp. $b$, be the distance from $x$,
resp. $y$, to ${\mathcal E}$. Denote by  $x'$, resp. $y'$, the image
in $C_{\mathcal E}$ of the point $([\gamma],a)$, resp. $([\mu],b)$,
of $X({\mathcal E})\times [0,\infty)$. Then
$$d_{g_\infty}(x,y)\le d_{g_{\mathcal E}}(x',y').$$
\end{lemma}

\begin{proof}
\uses{def:angle-between-directions-theta,lem:theta-monotone-convergence}
By the definition of the cone metric we have
$$d_{g_{\mathcal E}}(x',y')=a^2+b^2-2ab\,\cos(\theta(\gamma,\mu)).$$
On the other hand by Definition~\ref{def:angle-between-directions-theta} and the law of
cosines for Euclidean triangles, we have
$$d_{g_\infty}(x,y)=a^2+b^2-2ab\,\cos(\theta(\gamma,a,\mu,b)).$$
The result is now immediate from the fact, proved in
Lemma~\ref{lem:theta-monotone-convergence} that
$$0\le \theta(\gamma,a,\mu,b)\le \theta(\gamma,\mu)\le \pi,$$
and the fact that the cosine is a monotone decreasing function on
the interval $[0,\pi]$.
\end{proof}

\begin{corollary}[Rescaled cone distance bound]
\label{cor:rescaled-cone-distance-bound}
\label{refdistcomp}
\uses{lem:cone-distance-upper-bound}
Let $\gamma,\mu,x,y$ be as in the previous lemma. Fix $\lambda>0$.
Let $a=d_{\lambda g_\infty}(x,{\mathcal E})$ and $b=d_{\lambda
g_\infty}(y,{\mathcal E})$. Set $x'_\lambda$ and $y'_\lambda$ equal
to the points in the cone $([\gamma],a)$ and $([\mu],b)$. Then we
have
$$d_{\lambda g_\infty}(x,y)\le d_{g_{\mathcal E}}(x'_\lambda,y'_\lambda).$$
\end{corollary}


\begin{proof}
\uses{lem:cone-distance-upper-bound}
This is immediate by applying the previous lemma to the rescaled
manifold $(U_\infty,\lambda g_\infty)$, and noticing that rescaling
does not affect the cone $C_{\mathcal E}$ nor its metric.
\end{proof}

\begin{lemma}[Nets of directions contain close pairs]
\label{lem:directions-net-cardinality-bound}
\uses{def:direction-at-end,def:angle-between-directions-theta}
For any $\delta>0$ there is $K=K(\delta)<\infty$ so that for any set
of realized directions at ${\mathcal E}$ of cardinality $K$,
$\ell_1,\ldots,\ell_{K}$, it must be the case that there are $j$ and
$j'$ with $j\not=j'$ such that $\theta(\ell_j,\ell_{j'})<\delta$.
\end{lemma}

\begin{proof}
Let $K$ be such that, given $K$ points in the central $2$-sphere of
any $2\epsilon$-tube of scale $1$, at least two are within distance
$\delta/2$ of each other. Now suppose that we have $K$ directions
$\ell_1,\ldots,\ell_K$ at ${\mathcal E}$. Let
$\gamma_1,\ldots,\gamma_K$ be minimizing geodesic rays limiting to
${\mathcal E}$ that represent these directions. Choose a point $x$
sufficiently close to the end ${\mathcal E}$ so that all the
$\gamma_j$ cross the central $2$-sphere $S^2$ of the
$2\epsilon$-neck centered at $x$. By replacing the $\gamma_j$ with
sub-rays we can assume that for each $j$ the endpoint $x_j$ of
$\gamma_j$ lies in $S^2$. Let $d_j$ be the length of $\gamma_j$. By
taking $x$ sufficiently close to ${\mathcal E}$ we can also assume
the following. For each $j$ and $j'$, the angle at $e$ of the
Euclidean triangle $\alpha_je\alpha_{j'}$, where $|\alpha_je|=d_j;
|\alpha_{j'}e|=d_{j'}$ and $|\alpha_j\alpha_{j'}|=d(x_j,x_{j'})$ is
within $\delta/2$ of $\theta(\ell_j,\ell_{j'})$. Now there must be
$j\not= j'$ with $d(x_i,x_j)<(\delta/2)r_i$ where $r_i$ is the scale
of $N_i$. Since $d_j,d_{j'}>\epsilon^{-1}r_i/2$, it follows that the
angle at $e$ of $\alpha_je\alpha_{j'}$ is less than $\delta/2$.
Consequently, $\theta(\ell_j,\ell_{j'})<\delta$.
\end{proof}

Recall that a $\delta$-net in a metric space $X$ is a finite set of
points such that $X$ is contained in the union of the
$\delta$-neighborhoods of these points. The above lemma immediately
yields:

\begin{corollary}[Compactness and nets for the space of directions]
\label{cor:directions-space-compact-net}
\label{deltanet}
\uses{lem:directions-net-cardinality-bound,def:cone-of-directions}
The metric completion $\bar X({\mathcal E})$ of the space of
directions at ${\mathcal E}$ is a compact space. For every
$\delta>0$ this space has a  $\delta$-net consisting of realized
directions. For every $0<r<R<\infty$ the annular region $A_{\mathcal
E}(r,R)=\bar{X}({\mathcal E})\times [r,R]$ in $C_{\mathcal E}$ has a
$\delta$-net consisting of points $(\ell_i,s_i)$ where for each $i$
we have $\ell_i$ is a realizable direction and $r<s_i<R$.
\end{corollary}


\subsection{Completion of the proof of a cone limit at ${\mathcal E}$}

Now we are ready to prove Proposition~\ref{prop:cone-limit-gromov-hausdorff}. In fact, we
prove a version of the proposition that identifies the sequence of
points $x_n$ and also identifies the cone to which the
rescaled manifolds converge.

\begin{proposition}[Explicit Gromov--Hausdorff convergence to the cone]
\label{prop:gromov-hausdorff-limit-is-cone}
\label{secondstatment}
\uses{def:cone-of-directions,cor:directions-space-compact-net,prop:incomplete-limit-tube-properties}
Let $(U_\infty,g_\infty)$  be an incomplete Riemannian $3$-manifold
of non-negative curvature with an end ${\mathcal E}$ as in the
hypothesis of Proposition~\ref{prop:cone-limit-gromov-hausdorff}. Fix a minimizing geodesic
ray $\gamma$ limiting to ${\mathcal E}$. Let $\lambda_n$ be any
sequence of positive numbers tending to infinity. For each $n$
sufficiently large let $x_n\in \gamma$ be the point at distance
$\lambda_n^{-1/2}$ from the end ${\mathcal E}$. Then the based
metric spaces $(U_\infty,\lambda_ng_\infty,x_n)$ converge in the
Gromov-Hausdorff sense to
$\left( C'_{\mathcal E},g_{\mathcal E},([\gamma],1)\right)$. Under
this convergence the distance function from the end ${\mathcal E}$
in $(U_\infty,\lambda_ng_\infty)$ converges to the distance function
from the cone point in the open cone.
\end{proposition}


\begin{proof}
\uses{cor:directions-space-compact-net,cor:rescaled-cone-distance-bound,lem:theta-monotone-convergence}
It suffices to prove that given any subsequence of the original
sequence, the result holds for a further subsequence. So let us
replace the given sequence by a subsequence. Recall that for each
$0<r<R<\infty$ we have $A_{\mathcal E}(r,R)\subset C'_{\mathcal E}$,
the compact annulus which is the image of $\bar X({\mathcal
E})\times [r,R]$. The statement about the non-compact spaces
converging in the Gromov-Hausdorff topology, means that for each
compact subspace $K$ of $ C'_{\mathcal E}$ containing the base
point, for all $n$ sufficiently large, there are compact subspaces
$K_n\subset (U_\infty,\lambda_ng_\infty)$ containing $x_n$ with the
property that the $(K_n,x_n)$ converge in the Gromov-Hausdorff
topology to $(K,x)$ (see Section D of Chapter 3, p. 39, of
\cite{Gromov}).

Because of this, it suffices to  fix $0<r<1<R<\infty$ arbitrarily
and prove the convergence result for $A_{\mathcal E}(r,R)$. Since
the Gromov-Hausdorff distance from a compact pointed metric space to
a $\delta$-net in it containing the base point is at most $\delta$,
it suffices to prove that for $\delta>0$ there is a $\delta$-net
$({\mathcal N},h)$ in $A_{\mathcal E}(r,R)$, with $([\gamma],1)\in
{\mathcal N}$ such that for all $n$ sufficiently large there are
embeddings $\varphi_n$ of ${\mathcal N}$ into $A_n(r,R)=\bar
B_{\lambda_ng_\infty}({\mathcal E},R)\setminus
B_{\lambda_ng_\infty}({\mathcal E},r)$  with the following four
properties:
\begin{enumerate}
\item[(1)]  $\varphi_n^*(\lambda_ng_\infty)$ converge to $h$ as $n\rightarrow\infty$,
\item[(2)] $\varphi_n([\gamma],1)=x_n$,
and
\item[(3)]  $\varphi_n({\mathcal N})$ is a $\delta$-net in $A_n(r,R)$, and
\item[(4)] denoting the cone point by $c\in C_{\mathcal E}$,
if $d(p,c)=r$ then $d(\varphi_n(p),{\mathcal E})=r$.
\end{enumerate}

According to Corollary~\ref{cor:directions-space-compact-net} there is a $\delta$-net
${\mathcal N}\subset A_{\mathcal E}(r,R)$ consisting of points
$(\ell_i,s_i)$ where the $\ell_i$ are realizable directions and
$r<s_i< R$. Add $([\gamma],1)$ to ${\mathcal N}$ if necessary so
that we can assume that $([\gamma],1)\in {\mathcal N}$. Let
$\gamma_i$ be a minimizing geodesic realizing $\ell_i$ and let $d_i$
be its length.

Fix $n$ sufficiently large so that $\lambda_n^{-1/2}R\le d_i$ for all $i$. We
define $\varphi_n\colon {\mathcal N}\to A_n(r,R)$ as follows. For any
$a_i=([\gamma_i],s_i)\in {\mathcal N}$ we let
$\varphi_n(a_i)=\gamma_i(\lambda_n^{-1/2}s_i)$. (Since $\lambda_n^{-1/2}s\le
\lambda_n^{-1/2}R\le d_i$, the geodesic $\gamma_i$ is defined at
$\lambda_n^{-1/2}s_i$.) This defines the embeddings $\varphi_n$ for all $n$
sufficiently large. Notice that
$$d_{g_\infty}\left(\varphi_n(\ell_i,s_i),\varphi_n(\ell_j,s_j)\right)=\lambda_n^{-1}s_i^2+\lambda_n^{-1}
s_j^2-2\lambda_n^{-1}s_is_j\theta(\gamma_i,\lambda_n^{-1/2}s_i,\gamma_j,\lambda_n^{-1/2}s_j),$$
or equivalently
$$d_{\lambda_ng_\infty}\left(\varphi_n(\ell_i,s_i),\varphi_n(\ell_j,s_j)\right)=s_i^2+s_j^2-2s_is_j
\theta(\gamma_i,\lambda_n^{-1/2}s_i,\gamma_j,\lambda_n^{-1/2}s_j).$$
Because of the convergence result on angles (Lemma~\ref{lem:theta-monotone-convergence}),
for all $i$ and $j$ we have \begin{eqnarray*} \mathrm{lim}_{n\rightarrow\infty}d_{\lambda_ng_\infty}\left(\varphi_n(\ell_i,s_i),\varphi_n(\ell_j,s_j)\right)
 & = & s_i^2+s_j^2-2s_is_j\mathrm{cos}(\theta(\gamma_i,\gamma_j)) \\ & = & d_{g_{\mathcal
E}}\left((\ell_i,s_i),(\ell_j,s_j)\right).\end{eqnarray*}
 This
establishes the existence of the $\varphi_n$ for all $n$
sufficiently large satisfying the first condition. Clearly, from the
definition $\varphi_n([\gamma],1)=x_n$, and for all $p\in {\mathcal
N}$ we have $d(\varphi_n(p),{\mathcal E})=d(p,c)$.

It remains to check that for all $n$ sufficiently large
$\varphi_n({\mathcal N})$ is a $\delta$-net in $A_n(r,R)$. For $n$
sufficiently large let $z\in A_n(r,R)$ and let $\gamma_z$ be a
minimizing geodesic ray from $z$ to ${\mathcal E}$ parameterized by
the distance from the end. Set
$d_n=d_{\lambda_ng_\infty}(z,{\mathcal E})$, so that $r\le d_n\le
R$. Fix $n$ sufficiently large so that $\lambda_n^{-1/2}R<d_i$ for
all $i$. The point $([\gamma_z],d_n)\in C_{\mathcal E}$ is contained
in $A_{\mathcal E}(r,R)$ and hence there is an element
$a=([\gamma_i],s_i)\in {\mathcal N}$ within distance $\delta$ of
$([\gamma_z],d_n)$ in $C_{\mathcal E}$. Since $s_i\le R$,
$\lambda_n^{-1/2}s_i\le d_i$ and hence
$x=\gamma_i(\lambda_n^{-1/2}s_i)$ is defined. By
Corollary~\ref{cor:rescaled-cone-distance-bound} we have
$$d_{\lambda_ng_\infty}(x,z)\le
d_{g_{\mathcal E}}\left(([\gamma],d_n),([\gamma_i],s_i)\right)\le
\delta.$$ This completes the proof that for $n$ sufficiently large
the image $\varphi_n({\mathcal N})$ is a $\delta$-net in $A_n(r,R)$.

This shows that the $(U_\infty,\lambda_ng_\infty,x_n)$ converge in
the Gromov-Hausdorff topology to $(C'_{\mathcal E},g_{\mathcal
E},([\gamma],1))$.
\end{proof}

\begin{remark}[Non-uniqueness of the incomplete Gromov--Hausdorff limit]
\label{rem:noncomplete-multiple-limits}
Notice that since the manifolds $(U_\infty,\lambda_ng_\infty,x_n)$
are not complete, there can be more than one Gromov-Hausdorff limit.
For example we could take the full cone as a limit. Indeed, the cone
is the only Gromov-Hausdorff limit that is complete as a metric
space.
\end{remark}


\section{Comparison of the Gromov-Hausdorff limit and the smooth
limit}

Let us recap the progress to date. We constructed an incomplete
geometric blow-up limit $({\mathcal U}_\infty,G_\infty,z_\infty)$
for our original sequence. It has non-negative Riemann curvature. We
showed that the zero time-slice $U_\infty$ of the limit is
diffeomorphic to a tube $S^2\times (0,1)$ and that at one end of the
tube the scalar curvature goes to infinity. Also, any point
sufficiently near this end is the center of an evolving
$2\epsilon$-neck defined for an interval of normalized  time of
length $1/2$ in the limiting flow. Then we took a further blow-up
limit. We chose a sequence of points $x_n\in U_\infty$ tending to
the end ${\mathcal E}$ where the scalar curvature goes to infinity.
Then we formed $(U_\infty,\lambda_ng_\infty,x_n)$ where the distance
from $x_n$ to the end ${\mathcal E}$ is $\lambda_n^{-1/2}$. By
fairly general principles (in fact it is a general theorem about
manifolds of non-negative curvature) we showed that this sequence
converges in the Gromov-Hausdorff
sense to a cone.

The next step is to show that this second blow-up limit also exists
as a geometric limit away from the cone point. Take a sequence of
points $x_n\in U_\infty$ tending to ${\mathcal E}$. We let
$\lambda'_n=R(x_n)$, and we consider the based Riemannian manifolds
$(U_\infty,\lambda_n'g_\infty(0),x_n)$. Let $B_n\subset U_n$ be the
metric ball of radius $\epsilon^{-1}/3$ centered at $x_n$ in
$(U_\infty,\lambda'_ng_\infty(0))$. Since this ball is contained in
a $2\epsilon$-neck centered at $x_n$, the curvature on this ball is
bounded, and this ball has compact closure in $U_\infty$. Also, for
each $y\in B_n$, there is a rescaled flow $\lambda'g(t)$  defined on
$\{y\}\times (-1/2,0]$ whose  curvature on $B_n\times (-1/2,0]$ is
bounded. Hence, by Theorem~\ref{sliceconv} we can pass to a
subsequence and extract a geometric limit. In fact, by
Proposition~\ref{partialflowlimit} there is even a geometric
limiting flow defined on the time interval $(-1/2,0]$.

We must compare the zero time-slice of this geometric limiting flow
with the corresponding open subset of the Gromov-Hausdorff
limit constructed in the previous
section. Of course, one obvious difference is that we have used
different blow-up factors: $d(x_n,{\mathcal E})^{-2}$ in the first
case and $R(x_n)$ in the second case. So one important ingredient in
comparing the limits will be to compare these factors, at least in
the limit.


\subsection{Comparison of the blow-up factors}


Now let us compare the two limits: (i) the Gromov-Hausdorff limit of
the sequence $(U_\infty,\lambda_ng_\infty,x_n)$ and (ii) the
geometric limit of the sequence $(U_\infty,\lambda_n'g_\infty,x_n)$
 constructed above.

\begin{claim}[The blow-up ratios are bounded]
\label{claim:blowup-ratio-bounded}
\uses{prop:incomplete-limit-tube-properties,prop:gromov-hausdorff-limit-is-cone,S2intersection,directions}
The ratio $\rho_n=\lambda_n'/\lambda_n$ is bounded above and below
by positive constants.
\end{claim}


\begin{proof}
First of all, since there is a $2\epsilon$-neck centered at $x_n$,
by Proposition~\ref{S2intersection} we see that the distance
$\lambda_n^{-1/2}$ from $x_n$ to ${\mathcal E}$ is at least
$R(x_n)^{-1/2}\epsilon^{-1}/2=(\lambda'_n)^{-1/2}\epsilon^{-1}/2$.
Thus,
$$\rho_n^{-1}=\lambda_n/\lambda'_n\le 4\epsilon^2.$$

On the other hand, suppose that
$\rho_n=\lambda_n'/\lambda_n\rightarrow \infty$ as $n\rightarrow
\infty$. Rescale by $\lambda'_n$ so that $R(x_n)=1$. The distance
from $x_n$ to ${\mathcal E}$ is $\sqrt{\rho_n}$. Then by
Lemma~\ref{directions} with respect to this metric there is a sphere
of diameter at most $2\pi$  through  $x_n$ that separates all points
at distance at most $\sqrt{\rho_n}-\epsilon^{-1}$ from ${\mathcal
E}$ from all points at distance at least
$\sqrt{\rho_n}+\epsilon^{-1}$ from ${\mathcal E}$. Now rescale the
metric by $\rho_n$. In the rescaled metric there is a $2$-sphere of
diameter at most $2\pi/\sqrt{\rho_n}$ through $x_n$ that separates
all points at distance at most $1-\epsilon^{-1}/\sqrt{\rho_n}$ from
${\mathcal E}$ from all points at distance at least
$1+\epsilon^{-1}/\sqrt{\rho_n}$ from ${\mathcal E}$. Taking the
Gromov-Hausdorff limit of these spaces, we see that the base point
$x_\infty$ separates all points of distance less than one from
${\mathcal E}$ from all points of distance greater than one from
${\mathcal E}$. This is impossible since the Gromov-Hausdorff limit
is a cone that is not the cone on a single point.
\end{proof}


\subsection{Completion of the comparison of the blow-up limits}

Once we know that the $\lambda_n/\lambda_n'$ are bounded above and
below by positive constants, we can pass to a subsequence so that
these ratios converge to a finite positive limit. This means that
the Gromov-Hausdorff limit of the sequence of based metric spaces
$(U_\infty,\lambda_n'g_\infty,x_n)$ is a  cone, namely the
Gromov-Hausdorff limiting cone constructed is Section~\ref{GromHaus}
rescaled by $\lim_{n\rightarrow\infty}\rho_n$. In particular,
the balls of radius $\epsilon^{-1}/2$ around the base points in this
sequence converge in the Gromov-Hausdorff sense to the ball of
radius $\epsilon^{-1}/2$ about the base point of a cone.


But we have already seen that the balls of radius $\epsilon^{-1}/2$
centered at the base points converge geometrically to a limiting
manifold. That is to say, on every   ball of radius less than
$\epsilon^{-1}/2$ centered at the base point the metrics converge
uniformly in the $C^\infty$-topology to a limiting smooth metric.
Thus, on every ball of radius less than $\epsilon^{-1}/2$ centered
at the base point the limiting smooth metric is isometric to the
metric of the Gromov-Hausdorff limit. This means that the limiting
smooth metric on the ball $B_\infty$ of radius $\epsilon^{-1}/2$
centered at the base point is isometric to an open subset of a cone.
Notice that the scalar curvature of the limiting smooth metric at
the base point is $1$, so that this cone is a non-flat cone.




\section{The final contradiction}

We have now shown that the smooth limit of the balls of radius
$\epsilon^{-1}/2$ centered at the base points of
$(U_\infty,\lambda_n'g_\infty,x_n)$ is isometric to an open subset
of a non-flat cone, and is also the zero time-slice of a
Ricci flow defined for the time interval $(-1/2,0]$. This
contradicts Proposition~\ref{nocones},  one of the consequences of
the maximum principle established by Hamilton. The contradiction
shows that the limit $(U_\infty,g_\infty,x_\infty)$ cannot exist.
The only assumption that we made in order to construct this limit
was that Theorem~\ref{thm:bounded-curvature-bounded-distance} did not hold for some $A_0<\infty$.
Thus, we have established Theorem~\ref{thm:bounded-curvature-bounded-distance}  by contradiction.
